---
title: "Serving large CalCOFI CTD profiles on ERDDAP via DuckDB-over-Parquet"
subtitle: "Memory, latency, and partitioning/ordering — with notes for the CoastWatch ERDDAP"
author: "CalCOFI (B. Best) — for Lynn DeWitt & the ERDDAP community"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: kable
editor_options:
  chunk_output_type: console
---

## Summary

CalCOFI's CTD profile data is large in *long* (tidy) format — one row per
cast × depth × `measurement_type` (~104 sensor types): `ctd_measurement` is
**~216 M rows / ~16 GB** of Parquet (partitioned by `cruise_key`), plus an
adaptively-thinned headline table `ctd_thin` (~5.5 M rows). We wanted to serve
these on ERDDAP without converting to NetCDF and without the memory blow-up that
`EDDTableFromParquetFiles` hits on large files.

**What we found / did:**

1. `EDDTableFromParquetFiles` reads each matching file *fully into the JVM heap*
   before filtering. A single 935 MB `ctd_wide.parquet` OOM'd our 4 GB heap; the
   216 M-row table can't load at all on a 2 GB heap (times out at ~1.98 GB).
2. **`EDDTableFromDatabase` + the DuckDB JDBC driver**, pointed at a tiny `.db`
   of `CREATE VIEW … read_parquet(…)` views (joining the long tables to
   `ctd_cast` for `time`/`lat`/`lon`), serves the **full 216 M-row table within a
   ~63 MB load heap** and streams filtered queries. This is now live at
   <https://erddap.calcofi.io> (`calcofi_ctd_thin`, `calcofi_ctd_measurement`).
3. On the partitioning question you raised (Lynn): for the typical *"one variable
   in a lat/lon/depth box"* query, **partitioning the long table by
   `measurement_type`** (so the query touches ~1/N of the data) and **ordering
   rows spatially within each partition** gives the biggest wins — quantified
   below. DuckDB already does predicate pushdown + Parquet row-group skipping, so
   layout changes translate directly into less I/O.

This note documents the benchmark and the layout experiment so others weighing
Parquet-on-ERDDAP (cf. Roy Mendelssohn's posts, ERDDAP
[#447](https://github.com/ERDDAP/erddap/discussions/447) /
[#448](https://github.com/ERDDAP/erddap/issues/448)) have concrete numbers.

```{r}
#| label: setup
#| message: false
#| warning: false
library(dplyr); library(readr); library(tidyr); library(knitr); library(stringr)
res <- read_csv("data/bench_erddap/results_all.csv", show_col_types = FALSE) |>
  mutate(approach = factor(approach, levels = c("duckdb","parquet","csv")),
         table = factor(table, levels = c("thin","measurement")))
lay_path <- "data/bench_erddap/layout_results.csv"
lay <- if (file.exists(lay_path)) read_csv(lay_path, show_col_types = FALSE) else NULL
```

## 1 · Background

CalCOFI has occupied the California Current quarterly since 1949. The CTD cast
files (<https://calcofi.org/data/oceanographic-data/ctd-cast-files/>) are ~1 GB
as a single wide Parquet (one column per sensor). We ingest them into a tidy
**long** schema (`ingest_calcofi_ctd-cast`,
<https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html>): each sensor reading
on its own row keyed by `measurement_type`, which is convenient for analysis and
for adding new sensors, but multiplies row count. Tables (release v2026.06.08 /
supplemental v2026.04.08):

| table | rows | Parquet | note |
|---|--:|--:|---|
| `ctd_cast` | 5.55 M | 128 MB | one row per cast; carries `time`,`lat`,`lon` |
| `ctd_thin` | 5.55 M | 155 MB | adaptively-thinned headline table |
| `ctd_measurement` | 216 M | ~16 GB | full long table (partitioned by `cruise_key`) |

The long tables are keyed by `ctd_cast_uuid` and **do not carry coordinates** —
`time`/`latitude`/`longitude` live in `ctd_cast`. Any ERDDAP representation must
re-join them.

## 2 · The memory problem with `EDDTableFromParquetFiles`

`EDDTableFromParquetFiles` (ERDDAP ≥ 2.27) reads each *matching file* fully into
the JVM heap, then filters. That is fine for small files but:

- The single **935 MB** `ctd_wide.parquet` OOM'd our 4 GB heap and dropped the
  dataset from every `LoadDatasets` run.
- Even partitioned and at a deliberately small 2 GB heap, the 216 M-row table
  **could not load** (heap peaked ~1.98 GB, then timed out building the file
  table / reading files).
- Every *query* re-reads the whole matching file into heap, so latency scales
  with file size, not result size.

## 3 · The approach: `EDDTableFromDatabase` + DuckDB views

ERDDAP has no native DuckDB type yet (`EDDTableFromDuckDb`,
[#448](https://github.com/ERDDAP/erddap/issues/448), is a proposal; the ADBC/S3
idea is [#447](https://github.com/ERDDAP/erddap/discussions/447)). But ERDDAP
*already* speaks JDBC via `EDDTableFromDatabase`, and DuckDB ships a JDBC driver.
So we:

1. Build a **tiny `.db`** (274 KB — view definitions only) of
   `CREATE VIEW ctd_*_erddap AS SELECT … FROM read_parquet('…/ctd_*/**/*.parquet')
   t JOIN read_parquet('…/ctd_cast.parquet') c USING (ctd_cast_uuid)` — the join
   denormalizes `time`/`lat`/`lon` *logically* (no physical 16 GB rewrite).
2. Add the **DuckDB JDBC jar** to ERDDAP's `WEB-INF/lib` via a 4-line custom
   image (`FROM erddap/erddap:latest`). Pin the jar to the DuckDB engine that
   built the `.db` (we use 1.5.2).
3. Point `EDDTableFromDatabase` at `jdbc:duckdb:/…/calcofi_ctd.db` with
   connection properties `duckdb.read_only=true, memory_limit, threads,
   temp_directory`, and `sourceCanOrderBy/sourceCanDoDistinct = yes` so DuckDB
   (not the heap) does any ORDER BY / DISTINCT.

DuckDB then does predicate pushdown + Parquet partition pruning + row-group
skipping, and streams the JDBC `ResultSet` back to ERDDAP — so the whole table is
never on the heap.

::: {.callout-important title="Gotcha: serve time as a TIMESTAMP, not epoch double"}
ERDDAP reads any `time` variable (units `seconds since 1970-…`) via
`ResultSet.getTimestamp()`. If the DuckDB view exposes `time` as an epoch
**DOUBLE**, the DuckDB JDBC driver throws `NullPointerException: … varlen_data is
null` on *every* query that selects time. Expose `time` as a real **TIMESTAMP**
in the view (ERDDAP converts it to epoch seconds) and run the container in UTC.
Also: `EDDTableFromDatabase` does **not** auto-compute variable ranges, so set
`actual_range` (+ `geospatial_*`/`time_coverage_*`) on `time`/`lat`/`lon`/`depth`
or the form shows no min/max or sliders.
:::

## 4 · Benchmark: DuckDB vs Parquet vs CSV

Isolated ERDDAP (custom image) on our host, **2 GB heap** (deliberately small to
expose the differences), serving identical denormalized columns three ways.
Queries (median of repeats): **q1** `.das`; **q2** one cruise + one
`measurement_type`; **q3** one type, depth 0–200 m, all cruises; **q4**
whole-cruise dump.

```{r}
#| label: matrix
res |>
  transmute(table, approach,
            load = load_status, `load s` = load_secs,
            `load heap MB` = load_peak_heap_mb,
            `query heap MB` = query_peak_heap_mb,
            `q2 cruise ms` = q2_cruise_ms, `q3 type+depth ms` = q3_type200_ms,
            `q4 dump ms` = q4_dump_ms, `q3 rows` = q3_rows) |>
  arrange(table, approach) |> kable()
```

**Takeaways.** DuckDB loads *both* tables (incl. 216 M rows) with a flat ~63 MB
heap and streams queries; Parquet is 7–19× slower on the thin table (it re-reads
the file per query) and **cannot load** the 216 M-row table at 2 GB; CSV cannot
even load the thin table in 200 s. The DuckDB query-time heap (~1.1–1.6 GB on the
big dump) is ERDDAP buffering the *response*, not the table.

## 5 · Partitioning & ordering for long-format bbox queries

> *"I wonder how organizing each sensor reading on its own row affects a query for
> data within a lat/lon/depth box."* — Lynn DeWitt

This is the key question for long format. We measured (DuckDB, the serving engine)
the same bounding-box query — **`temperature_ave`, 32–34°N, 119–121°W, 0–200 m
(~393 k rows)** — against four physical layouts of the 216 M-row denormalized
table:

- **A** — production view: raw Parquet partitioned by `cruise_key`, joined to
  `ctd_cast` at query time (what's live now).
- **B** — denormalized, partitioned by `cruise_key` (isolates layout from the join).
- **C** — denormalized, partitioned by `measurement_type` (one variable ⇒ ~1/N files).
- **D** — denormalized, partitioned by `measurement_type` **and** ordered by
  `latitude, longitude, depth` within each partition (row-group skipping for the box).

```{r}
#| label: layout
#| eval: !expr "!is.null(lay)"
lay |>
  transmute(layout, `build s` = build_secs, size, files = nfiles,
            `query s (median)` = query_secs_median, `result rows` = result_rows) |>
  kable()
```

```{r}
#| label: layout-missing
#| echo: false
#| output: asis
#| eval: !expr "is.null(lay)"
cat("*(layout_results.csv not yet present — run scripts/bench_parquet_layout.sh)*")
```

**How to read this.** Partition pruning (C, D) drops the data DuckDB must open
from all ~95 cruise files to just the one `measurement_type` partition. Spatial
ordering (D) additionally lets DuckDB skip Parquet **row groups** whose
lat/lon/depth min–max can't intersect the box. The trade-off: partitioning by
`measurement_type` *helps single-variable queries* (the common ERDDAP case) but a
query for *all* variables in a region must then open all type partitions — so the
"best" layout depends on expected access patterns (exactly your point). For the
adaptively-thinned `ctd_thin` (155 MB) the question is moot; for the full
`ctd_measurement` it matters, and the layout is a cheap one-time re-write.

**Takeaway.** This ~393 k-row box query runs in **0.075–0.875 s across all four
layouts — every one sub-second**, so DuckDB + SSD already handle long-format bbox
queries comfortably; layout is an optimization, not a rescue (reassuring for the
"sensor-reading-per-row" worry). Decomposing the ~12× spread:

- **Removing the query-time join ≈ 4×** (A→B, 0.875 → 0.223 s): materializing
  `time`/`lat`/`lon` into the table both skips the join *and* lets the lat/lon
  predicate prune at the Parquet scan.
- **Partitioning by `measurement_type` ≈ 2×** (B→C, → 0.104 s): a one-variable
  query opens 1 of 53 partitions instead of all 95 cruise files.
- **Spatial ordering within each partition ≈ 1.4×** (C→D, → 0.075 s): DuckDB
  skips Parquet row groups whose `latitude/longitude/depth` min–max miss the box.

Costs & caveats: the denormalized layouts add ~17 GB, and the spatial sort is a
one-time global sort (here ~11 min with heavy temp spill) — pay once at ingest,
save on every query. And partitioning by `measurement_type` *slows* a query that
wants *all* variables in a region (it must open every type partition), so the
optimum depends on whether single-variable or all-variable access dominates. Our
live datasets currently keep the **zero-extra-storage logical-join view** (layout
A, 0.2–0.9 s here); if heavy single-variable bbox traffic emerges, the
`measurement_type`-partition + spatial-sort layout is a cheap ~12× upgrade.

## 6 · Recommendations & open questions

- **Serving path.** `EDDTableFromDatabase` + DuckDB JDBC works on stock ERDDAP
  today (only the driver jar is added) and is the only option here that serves the
  216 M-row table within a small heap. It complements — and is a pragmatic bridge
  to — native Parquet/DuckDB support (#447/#448). We're happy to share the custom
  image + `.db` builder.
- **Layout.** For the full long table, partition by `measurement_type` and order
  spatially within; keep `cruise_key` only if cruise-scoped access dominates.
- **NetCDF alternative.** Rasmus's NetCDF-per-cast route (à la Trinidad Head) is
  fully viable too and may be the better fit if CoastWatch hosts; the long-format
  DuckDB approach is attractive when you want the tidy table queryable as-is.
- **Open:** for hosting on the CoastWatch ERDDAP, would you prefer (a) NetCDF
  files we generate, (b) Parquet + a DuckDB `.db` (this approach), or (c) waiting
  on native DuckDB/ADBC? We can produce any of these from the same pipeline.

## 7 · Reproduce

All on our host (`CalCOFI/workflows`, `CalCOFI/server`, `CalCOFI/erddap`):

- `server/erddap/Dockerfile` — ERDDAP + `duckdb_jdbc`.
- `libs/erddap_duckdb.R` — `.db` view builder + `EDDTableFromDatabase` XML
  generator (reuses `libs/erddap.R`).
- `scripts/gen_prod_datasets.R` — production dataset blocks (metadata-enriched,
  with `actual_range`).
- `scripts/bench_erddap.sh` — the DuckDB/Parquet/CSV matrix → `results_all.csv`.
- `scripts/bench_parquet_layout.sh` — the partitioning/ordering experiment →
  `layout_results.csv`.

Live: <https://erddap.calcofi.io/erddap/tabledap/calcofi_ctd_thin.html> ·
<https://erddap.calcofi.io/erddap/tabledap/calcofi_ctd_measurement.html>
