---
title: "Serving large CalCOFI CTD profiles on ERDDAP: format × schema × granularity"
subtitle: "NetCDF vs DuckDB vs Parquet vs CSV — memory, latency, and layout, for the CoastWatch ERDDAP"
author: "CalCOFI (B. Best) — for Lynn DeWitt & the ERDDAP community"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: bench_erddap_ctd
  workflow_type: reference
  workflow_url: https://calcofi.io/workflows/bench_erddap_ctd.html
editor_options:
  chunk_output_type: console
---

## Summary

CalCOFI's CTD profile data is large in *long* (tidy) format — one row per
cast × depth × `measurement_type` (~53 sensor types): `ctd_measurement` is
**~216 M rows / ~16 GB** of Parquet, plus an adaptively-thinned headline table
`ctd_thin` (~5.5 M rows). We want to serve it on ERDDAP without the memory
blow-up that disabled an earlier single-file attempt. To get an *apples-to-apples*
read, we benchmarked a controlled grid of **four axes** at a deliberately small
**2 GB JVM heap**:

- **format** — NetCDF (`EDDTableFromNcCFFiles`), DuckDB (`EDDTableFromDatabase`),
  Parquet (`EDDTableFromParquetFiles`), CSV (`EDDTableFromAsciiFiles`)
- **schema** — **wide** (canonical sensors as columns) vs **long** (tidy)
- **granularity** — **split** (one file per cruise, 96 files) vs **lumped** (one
  file for the whole table) — *for the file formats; DuckDB is granularity-agnostic*
- **query** — whole-dataset (load) vs subset (filtered)

**What the grid shows (the four findings):**

1. **Granularity is the memory lever, not format.** ERDDAP's file backends read
   *whole matching files onto the heap*, so a single **lumped 2.1 GB** wide NetCDF
   can't be served at a 2 GB heap (ERDDAP sheds requests, then times out), while the
   **same data split into 96 per-cruise files loads at ~1.1 GB and serves fast**.
   This — not the file format — is what sank the original single-file `ctd_wide`.
2. **Lumped "loads" ≠ usable.** A smaller lumped file can bind (the 598 MB wide
   Parquet loaded at 1.6 GB) but then *every query re-reads the whole file*: a
   cruise dump took **74 s lumped vs 1.2 s split**. Split wins on latency even when
   lumped fits.
3. **DuckDB is the memory-safe baseline (~65 MB any config)** because it *streams*
   from the underlying Parquet instead of loading files onto the heap — but its
   *on-the-fly wide pivot view* is expensive for broad cross-cruise queries (76 s),
   so keep the **long** view live and materialize wide only if needed.
4. **NetCDF, the native path, is per-cruise.** CalCOFI has **~5.55 M casts**, so
   one file *per cast* (à la Trinidad Head) is infeasible; bundling **by cruise = 96
   files**, served as **one aggregated ERDDAP dataset** (one form), loads small and
   serves cruise-scoped queries fastest.

Live now at <https://erddap.calcofi.io> (`calcofi_ctd_thin`, `calcofi_ctd_measurement`,
DuckDB-long). Everything needed to reproduce is in §8.

```{r}
#| label: setup
#| message: false
#| warning: false
library(dplyr); library(readr); library(tidyr); library(knitr); library(stringr); library(ggplot2); library(forcats)
res <- read_csv("data/bench_erddap/results_all.csv", show_col_types = FALSE) |>
  mutate(format = factor(format, levels = c("duckdb","netcdf","parquet","csv")),
         schema = factor(schema, levels = c("wide","long")),
         granularity = factor(granularity, levels = c("na","split","lumped"),
                              labels = c("view (DB)","split (per-cruise)","lumped (1 file)")),
         table = factor(table, levels = c("thin","measurement"),
                        labels = c("ctd_thin (5.5M rows)","ctd_measurement (216M rows)")),
         loaded = load_status == "loaded")
lay_path <- "data/bench_erddap/layout_results.csv"
lay <- if (file.exists(lay_path)) read_csv(lay_path, show_col_types = FALSE) else NULL
PAL <- c("split (per-cruise)" = "#1b9e77", "lumped (1 file)" = "#d95f02", "view (DB)" = "#7570b3")
```

## 1 · Background

CalCOFI has occupied the California Current quarterly since 1949. The CTD cast
files (<https://calcofi.org/data/oceanographic-data/ctd-cast-files/>) are ~1 GB
as a single wide Parquet. We ingest them into a tidy **long** schema
(`ingest_calcofi_ctd-cast`,
<https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html>): each sensor reading
on its own row keyed by `measurement_type`, convenient for analysis but
multiplying row count. Tables (release v2026.06.08 / supplemental v2026.04.08):

| table | rows | casts | Parquet | note |
|---|--:|--:|--:|---|
| `ctd_cast` | 5.55 M | 5.55 M | 128 MB | one row per cast; carries `time`,`lat`,`lon` |
| `ctd_thin` | 5.55 M | 0.42 M | 155 MB | adaptively-thinned headline table |
| `ctd_measurement` | 216 M | 5.36 M | ~16 GB | full long table (96 `cruise_key` partitions) |

The long tables are keyed by `ctd_cast_uuid` and carry no coordinates —
`time`/`latitude`/`longitude` live in `ctd_cast`, so every representation re-joins
them. The **5.55 M casts** figure is what makes per-cast NetCDF infeasible (§3).

## 2 · How ERDDAP serves files — and the NetCDF "per-cruise = one dataset" point

A persistent point of confusion: **per-cruise files do *not* mean a dataset (or
form) per cruise.** Every ERDDAP `…FromFiles` type points at a *directory*
(`fileDir` + `fileNameRegex`) and aggregates all matching files into **one**
`datasetID` — one Data Access Form, one `.das`, with sliders/min-max **rolled up
across the whole collection**. The per-cruise split is purely *storage*: it's how
we go from "5.55 M per-cast files ERDDAP can't index" to "96 files it can," while
the user still sees a single dataset. ERDDAP prunes by `cruise_key` (a
`subsetVariable`) and by each file's time/lat/lon min-max, so a cruise-scoped query
opens one file and a broad spatial query opens many.

The flip side is the memory model: for **file** backends ERDDAP reads each
*matching file fully onto the JVM heap* before filtering. That is the entire
origin of the memory problem — a single **935 MB** `ctd_wide.parquet` OOM'd a 4 GB
heap and dropped the dataset from every `LoadDatasets` run. The benchmark below
isolates that as the **granularity** axis. **DuckDB** is the exception: via JDBC it
*streams* a filtered `ResultSet` from the underlying Parquet (predicate pushdown +
partition pruning + disk spill), so the whole table is never on the heap.

::: {.callout-important title="Two DuckDB gotchas"}
Serve `time` as a real **TIMESTAMP** in the view (not epoch double) or the DuckDB
JDBC driver NPEs on every query that selects time. And `EDDTableFromDatabase` does
**not** auto-compute variable ranges — set `actual_range` (+ `geospatial_*`) or the
form shows no sliders. (NetCDF computes these from the files automatically.)
:::

## 3 · The benchmark grid

Isolated ERDDAP (custom image with the DuckDB JDBC jar) on our host, **2 GB heap**.
We serve the **same canonical data** every way and run the **same-intent** queries:
**q1** `.das`; **q2** one cruise + temperature (subset); **q3** one variable
(salinity) over depth 0–200 m across all cruises (subset, cross-file); **q4**
whole-cruise dump. In *wide* schema a query selects a named column; in *long* it
filters `measurement_type` — chosen so q2/q3 return the **same rows** either way.

```{r}
#| label: matrix
res |>
  arrange(table, schema, format, granularity) |>
  transmute(table, format, schema, granularity, load = load_status,
            `load heap MB` = load_peak_heap_mb,
            `q2 cruise ms` = na_if(q2_cruise_ms, -1),
            `q3 var+depth ms` = na_if(q3_var200_ms, -1),
            `q4 dump ms` = na_if(q4_dump_ms, -1)) |>
  kable()
```

### Finding 1 & 3 — load memory: granularity (files) vs streaming (DuckDB)

```{r}
#| label: fig-mem
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Peak JVM heap during dataset load at a 2 GB cap, by format and granularity. DuckDB (purple) stays ~65 MB whatever the schema/size because it streams. The file formats rise with what ERDDAP must read; the lumped 2.1 GB measurement NetCDF blows past the cap and times out (labeled). Bars are labeled with peak MB, or the failure mode."
res |>
  mutate(lab = ifelse(loaded, paste0(load_peak_heap_mb), toupper(load_status))) |>
  ggplot(aes(format, load_peak_heap_mb, fill = granularity)) +
  geom_col(position = position_dodge2(preserve = "single"), width = 0.85) +
  geom_text(aes(label = lab, colour = loaded),
            position = position_dodge2(width = 0.85, preserve = "single"),
            vjust = -0.3, size = 2.6, show.legend = FALSE) +
  geom_hline(yintercept = 2048, linetype = "dashed", colour = "grey45") +
  facet_grid(schema ~ table) +
  scale_fill_manual(values = PAL, name = "granularity") +
  scale_colour_manual(values = c(`TRUE` = "grey20", `FALSE` = "#d95f02")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "peak load heap (MB); dashed = 2 GB cap") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

### Finding 2 — query latency: split serves subsets fast, lumped re-reads everything

```{r}
#| label: fig-latency
#| fig-width: 9
#| fig-height: 4.6
#| fig-cap: "Median latency (log scale) for loaded cells. SPLIT files serve cruise-scoped queries (q2, q4) fastest — one cruise = one file. LUMPED files re-read the whole file on every query (q4 cruise-dump up to 74 s). DuckDB shows TWO purple bars — its wide-pivot and long views; the wide view is slow on the broad cross-cruise scan (q3: 76.5 s wide vs 23.9 s long) because it aggregates the whole table on the fly, while it's faster on the cruise dump (q4: 3 s wide vs 99 s long, fewer rows)."
lbl_ms <- function(x) ifelse(x >= 1000, paste0(round(x/1000, 1), "s"), paste0(round(x), "ms"))
res |>
  filter(loaded) |>
  transmute(table, format, granularity,
            `q2 cruise` = na_if(q2_cruise_ms, -1),
            `q3 var+depth` = na_if(q3_var200_ms, -1),
            `q4 cruise dump` = na_if(q4_dump_ms, -1)) |>
  pivot_longer(starts_with("q"), names_to = "query", values_to = "ms") |>
  filter(!is.na(ms)) |>
  ggplot(aes(format, ms, fill = granularity)) +
  geom_col(position = position_dodge2(preserve = "single"), width = 0.8) +
  geom_text(aes(label = lbl_ms(ms)),
            position = position_dodge2(width = 0.8, preserve = "single"),
            vjust = -0.3, size = 2.2) +
  facet_grid(query ~ table) +
  scale_fill_manual(values = PAL, name = "granularity") +
  scale_y_log10(expand = expansion(mult = c(0, 0.25))) +
  labs(x = NULL, y = "median latency (log scale)") +
  theme_minimal(base_size = 10) + theme(legend.position = "top")
```

**Reading the grid.**

- **Load memory (Fig 1).** **DuckDB ≈ 65 MB in every cell** — wide or long, thin or
  216 M rows — because it never holds a file on-heap. The file formats scale with
  what ERDDAP reads: per-cruise **split** measurement loads at ~1.1–1.8 GB (it
  indexes 96 files), and the **lumped 2.1 GB** measurement NetCDF **exceeds the cap
  and times out** (ERDDAP starts shedding requests under memory pressure). The
  smaller 598 MB lumped Parquet squeaks in at 1.6 GB — *granularity bites once the
  single file is large relative to the heap.*
- **Query latency (Fig 2).** **Split** files serve cruise-scoped q2/q4 in well under
  a second (one `cruise_key` → one file). **Lumped** files have no such pruning, so
  every query re-reads the whole file — the lumped-Parquet cruise dump took **74 s**
  vs **1.2 s** split, and lumped CSV/Parquet q2 are seconds, not milliseconds.
  **DuckDB-wide's** broad q3 is **76 s** because the pivot view aggregates the whole
  216 M-row table on the fly; the **DuckDB-long** view does the same q3 in 24 s and
  is what we serve live.
- **Apples-to-apples format (wide, split).** With schema and granularity held fixed,
  **NetCDF-split** and **Parquet-split** are the fast file formats (sub-second
  cruise queries); DuckDB is lightest on memory but slowest on broad aggregation;
  CSV works only at thin scale.
- **Schema (wide vs long).** For the materialized file formats it's close. For
  **DuckDB** the on-the-fly wide pivot is *slower* than long on cross-cruise scans —
  a reason to keep the tidy long view live and reach for wide only behind a
  materialization.

## 4 · Partitioning & ordering within the long table (Lynn's question)

> *"I wonder how organizing each sensor reading on its own row affects a query for
> data within a lat/lon/depth box."* — Lynn DeWitt

Beyond split-vs-lumped, *within* the long table we measured (in DuckDB, the serving
engine) the same bounding-box query — **`temperature_ave`, 32–34°N, 119–121°W,
0–200 m (~393 k rows)** — across four physical layouts:

- **A** raw Parquet partitioned by `cruise_key`, joined to `ctd_cast` at query time
  (live now); **B** denormalized, partitioned by `cruise_key`; **C** denormalized,
  partitioned by `measurement_type`; **D** = C **plus** rows sorted by
  `latitude,longitude,depth` within each partition.

```{r}
#| label: layout
#| eval: !expr "!is.null(lay)"
lay |>
  transmute(layout, `build s` = build_secs, size, files = nfiles,
            `query s (median)` = query_secs_median, `result rows` = result_rows) |>
  kable()
```

```{r}
#| label: fig-layout
#| eval: !expr "!is.null(lay)"
#| fig-width: 7.5
#| fig-height: 3
#| fig-cap: "The same ~393k-row bbox query across four physical layouts of the long table (DuckDB on SSD). ~12× spread, but every layout is already sub-second."
lay |>
  mutate(layout = fct_rev(factor(layout))) |>
  ggplot(aes(layout, query_secs_median)) +
  geom_col(width = 0.7, fill = "#7570b3") +
  geom_text(aes(label = paste0(formatC(query_secs_median, format="f", digits=3), " s")),
            hjust = -0.15, size = 3) +
  coord_flip() + scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "median query (s)") + theme_minimal(base_size = 11)
```

```{r}
#| label: layout-missing
#| echo: false
#| output: asis
#| eval: !expr "is.null(lay)"
cat("*(layout_results.csv not present — run scripts/bench_parquet_layout.sh)*")
```

Decomposing the ~12× spread: removing the query-time join ≈ 4× (A→B);
partitioning by `measurement_type` ≈ 2× (B→C, a one-variable query opens 1 of 53
partitions); spatial ordering ≈ 1.4× (C→D, row-group skipping). The trade-off is
the one you raised: partitioning by `measurement_type` helps single-variable
queries but slows "all variables in a region." Every layout is already sub-second,
so for the long table this is an optimization, not a rescue.

## 5 · Recommendations

- **The memory question is "lumped vs split," not "which format."** Never serve the
  big table as one file via a file backend — split it per cruise (96 files). DuckDB
  sidesteps the question entirely by streaming.
- **For CoastWatch hosting, NetCDF is the most native fit** — per-cruise CF profile
  files served as one `EDDTableFromNcCFFiles` dataset; small load heap, fast
  cruise-scoped queries, auto ranges/sliders, source `.nc` downloadable. We generate
  them from the same pipeline (96 files/table). Costs: canonical sensors only,
  ~2.3 GB duplicated, regenerate on update.
- **To keep the tidy long table queryable as-is, DuckDB-over-Parquet** is the
  lightest path: a 274 KB `.db` of views on stock ERDDAP (only the driver jar
  added), serving 216 M rows within ~65 MB, always reflecting the live Parquet. Keep
  it **long** — the wide pivot view is slow on broad scans. This runs live today.
- **Avoid `EDDTableFromParquetFiles` on a single large file** (whole-file on-heap
  reads), and **avoid lumped anything** for the big table.
- **Open for you, Lynn:** (a) per-cruise **NetCDF** we generate, (b) **Parquet + the
  small DuckDB `.db`**, or (c) wait on native Parquet/DuckDB
  ([#447](https://github.com/ERDDAP/erddap/discussions/447) /
  [#448](https://github.com/ERDDAP/erddap/issues/448))? Any is producible from the
  same pipeline.

## 6 · Reproduce

Everything runs in the CalCOFI server containers (`../server/`), reading from and
writing to the shared **`/share/data/erddap-duckdb`** tree, mounted identically into
`rstudio` (build), the live `erddap`, and the bench `erddap_bench` — no
host-specific paths.

**Code** (all in [CalCOFI/workflows](https://github.com/CalCOFI/workflows)):

| file | role |
|---|---|
| [`libs/erddap_netcdf.R`](https://github.com/CalCOFI/workflows/blob/main/libs/erddap_netcdf.R) | per-cruise + lumped CF NetCDF builders + `EDDTableFromNcCFFiles` XML |
| [`libs/erddap_duckdb.R`](https://github.com/CalCOFI/workflows/blob/main/libs/erddap_duckdb.R) | `.db` view builder + `EDDTableFromDatabase` XML |
| [`libs/erddap.R`](https://github.com/CalCOFI/workflows/blob/main/libs/erddap.R) | shared dataVariable / type / units helpers |
| [`scripts/gen_ctd_netcdf.R`](https://github.com/CalCOFI/workflows/blob/main/scripts/gen_ctd_netcdf.R) · [`gen_ctd_wide.R`](https://github.com/CalCOFI/workflows/blob/main/scripts/gen_ctd_wide.R) · [`gen_ctd_lumped.R`](https://github.com/CalCOFI/workflows/blob/main/scripts/gen_ctd_lumped.R) | build the split/lumped, wide/long serving artifacts |
| [`scripts/gen_bench_datasets.R`](https://github.com/CalCOFI/workflows/blob/main/scripts/gen_bench_datasets.R) | emit the per-cell `<dataset>` blocks (datasetID = cell) |
| [`scripts/bench_erddap.sh`](https://github.com/CalCOFI/workflows/blob/main/scripts/bench_erddap.sh) | the format×schema×granularity matrix → `results_all.csv` |
| [`scripts/bench_parquet_layout.sh`](https://github.com/CalCOFI/workflows/blob/main/scripts/bench_parquet_layout.sh) | the partitioning/ordering experiment → `layout_results.csv` |
| [`server/erddap/Dockerfile`](https://github.com/CalCOFI/server/blob/main/erddap/Dockerfile) · [`docker-compose.bench.yml`](https://github.com/CalCOFI/server/blob/main/docker-compose.bench.yml) | ERDDAP + `duckdb_jdbc`; isolated bench instance |

**Data & artifacts** are browsable / downloadable via the Caddy file server at
[**file.calcofi.io/erddap-ctd**](https://file.calcofi.io/erddap-ctd/). Live datasets:
<https://erddap.calcofi.io/erddap/tabledap/calcofi_ctd_thin.html> ·
<https://erddap.calcofi.io/erddap/tabledap/calcofi_ctd_measurement.html>.

```{r}
#| label: data-files
read_csv("data/bench_erddap/results_all.csv", show_col_types = FALSE) |> kable(caption = "results_all.csv (the full grid)")
if (!is.null(lay)) kable(lay, caption = "layout_results.csv")
```

The per-cruise NetCDF builder and the benchmark harness, in full:

```{r}
#| label: embed-netcdf
#| eval: false
#| file: libs/erddap_netcdf.R
```

```{r}
#| label: embed-bench
#| eval: false
#| file: scripts/bench_erddap.sh
```
