---
title: "Serving large CalCOFI CTD profiles on ERDDAP: NetCDF vs DuckDB vs Parquet vs CSV"
subtitle: "Memory, latency, and partitioning/ordering — with notes for the CoastWatch ERDDAP"
author: "CalCOFI (B. Best) — for Lynn DeWitt & the ERDDAP community"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: bench_erddap_ctd
  workflow_type: reference
  workflow_url: https://calcofi.io/workflows/bench_erddap_ctd.html
editor_options:
  chunk_output_type: console
---

## Summary

CalCOFI's CTD profile data is large in *long* (tidy) format — one row per
cast × depth × `measurement_type` (~53 sensor types): `ctd_measurement` is
**~216 M rows / ~16 GB** of Parquet (partitioned by `cruise_key`), plus an
adaptively-thinned headline table `ctd_thin` (~5.5 M rows). We wanted to serve
these on ERDDAP — testing the **native NetCDF path** as well as backends that
avoid converting away from Parquet — without the memory blow-up that
`EDDTableFromParquetFiles` hits on large files.

We benchmarked four serving backends at a deliberately small **2 GB JVM heap**:

| backend | EDDType | format | loads `ctd_measurement` (216 M)? |
|---|---|---|:--:|
| **NetCDF** (per-cruise CF profiles) | `EDDTableFromNcCFFiles` | wide | ✅ (heap ~1.3 GB) |
| **DuckDB** (JDBC views over Parquet) | `EDDTableFromDatabase` | long | ✅ (heap ~60 MB) |
| **Parquet** (one denormalized file) | `EDDTableFromParquetFiles` | long | ❌ OOM / timeout |
| **CSV** (uncompressed) | `EDDTableFromAsciiFiles` | long | ❌ (disk-prohibitive) |

**Headline findings:**

1. **NetCDF is the native ERDDAP path and it works well** — *if* the data is
   re-shaped to per-cruise CF profile files. CalCOFI has **~5.55 M casts**, so
   "one NetCDF *per cast*" (à la Trinidad Head) is **5.55 M files** (infeasible);
   bundling **by cruise gives just 96 files**. ERDDAP reads each file's index and
   opens only matching files, so the 216 M-row-equivalent table loads in a ~1.3 GB
   heap and **cruise-scoped queries are the fastest of any backend** (it opens one
   file). ERDDAP also auto-computes `actual_range` from the files, so the
   min/max + sliders work with no extra config.
2. **DuckDB-over-Parquet keeps the tidy long table queryable as-is** with a tiny
   footprint: a 274 KB `.db` of `CREATE VIEW … read_parquet(…)` views serves the
   **full 216 M-row table within a ~60 MB load heap** and streams filtered queries.
   It is live now at <https://erddap.calcofi.io> (`calcofi_ctd_thin`,
   `calcofi_ctd_measurement`).
3. **Plain Parquet is the memory trap.** `EDDTableFromParquetFiles` reads each
   matching file *fully into the JVM heap* before filtering: a single 935 MB
   `ctd_wide.parquet` OOM'd a 4 GB heap, and the 216 M-row table won't load at 2 GB.
4. On the **partitioning** question (Lynn): for the typical *"one variable in a
   lat/lon/depth box"* query, partitioning the long table by `measurement_type`
   and ordering rows spatially within gives ~12× — quantified below — though every
   layout is already sub-second with DuckDB on SSD.

This note documents the benchmark, the layout experiment, and **everything needed
to reproduce it** (§8) so others weighing Parquet/NetCDF-on-ERDDAP (cf. Roy
Mendelssohn's posts, ERDDAP
[#447](https://github.com/ERDDAP/erddap/discussions/447) /
[#448](https://github.com/ERDDAP/erddap/issues/448)) have concrete numbers.

```{r}
#| label: setup
#| message: false
#| warning: false
library(dplyr); library(readr); library(tidyr); library(knitr); library(stringr); library(ggplot2)
res <- read_csv("data/bench_erddap/results_all.csv", show_col_types = FALSE) |>
  mutate(approach = factor(approach, levels = c("netcdf","duckdb","parquet","csv")),
         table    = factor(table, levels = c("thin","measurement"),
                           labels = c("ctd_thin (5.5M rows)","ctd_measurement (216M rows)")),
         loaded   = load_status == "loaded")
lay_path <- "data/bench_erddap/layout_results.csv"
lay <- if (file.exists(lay_path)) read_csv(lay_path, show_col_types = FALSE) else NULL
hsz <- function(b) ifelse(is.na(b) | b <= 0, "—", paste0(formatC(b/1024, format="f", digits=0, big.mark=","), " KB"))
```

## 1 · Background

CalCOFI has occupied the California Current quarterly since 1949. The CTD cast
files (<https://calcofi.org/data/oceanographic-data/ctd-cast-files/>) are ~1 GB
as a single wide Parquet (one column per sensor). We ingest them into a tidy
**long** schema (`ingest_calcofi_ctd-cast`,
<https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html>): each sensor reading
on its own row keyed by `measurement_type`, which is convenient for analysis and
for adding new sensors, but multiplies row count. Tables (release v2026.06.08 /
supplemental v2026.04.08):

| table | rows | casts | Parquet | note |
|---|--:|--:|--:|---|
| `ctd_cast` | 5.55 M | 5.55 M | 128 MB | one row per cast; carries `time`,`lat`,`lon` |
| `ctd_thin` | 5.55 M | 0.42 M | 155 MB | adaptively-thinned headline table |
| `ctd_measurement` | 216 M | 5.36 M | ~16 GB | full long table (96 `cruise_key` partitions) |

The long tables are keyed by `ctd_cast_uuid` and **do not carry coordinates** —
`time`/`latitude`/`longitude` live in `ctd_cast`. Any ERDDAP representation must
re-join them. The **5.55 M casts** number is what makes per-cast NetCDF
infeasible and motivates the per-cruise bundling in §4.

## 2 · The memory problem with `EDDTableFromParquetFiles`

`EDDTableFromParquetFiles` (ERDDAP ≥ 2.27) reads each *matching file* fully into
the JVM heap, then filters. That is fine for small files but:

- The single **935 MB** `ctd_wide.parquet` OOM'd our 4 GB heap and dropped the
  dataset from every `LoadDatasets` run.
- Even partitioned and at a deliberately small 2 GB heap, the 216 M-row table
  **could not load** (heap peaked ~1.98 GB, then timed out building the file
  table / reading files).
- Every *query* re-reads the whole matching file into heap, so latency scales
  with file size, not result size.

## 3 · DuckDB-over-Parquet (`EDDTableFromDatabase`)

ERDDAP has no native DuckDB type yet (`EDDTableFromDuckDb`,
[#448](https://github.com/ERDDAP/erddap/issues/448), is a proposal; the ADBC/S3
idea is [#447](https://github.com/ERDDAP/erddap/discussions/447)). But ERDDAP
*already* speaks JDBC via `EDDTableFromDatabase`, and DuckDB ships a JDBC driver.
So we:

1. Build a **tiny `.db`** (274 KB — view definitions only) of
   `CREATE VIEW ctd_*_erddap AS SELECT … FROM read_parquet('…/ctd_*/**/*.parquet')
   t JOIN read_parquet('…/ctd_cast.parquet') c USING (ctd_cast_uuid)` — the join
   denormalizes `time`/`lat`/`lon` *logically* (no physical 16 GB rewrite).
2. Add the **DuckDB JDBC jar** to ERDDAP's `WEB-INF/lib` via a 4-line custom
   image (`FROM erddap/erddap:latest`). Pin the jar to the DuckDB engine that
   built the `.db` (we use 1.5.2).
3. Point `EDDTableFromDatabase` at `jdbc:duckdb:/…/calcofi_ctd.db` with
   connection properties `duckdb.read_only=true, memory_limit, threads,
   temp_directory`, and `sourceCanOrderBy/sourceCanDoDistinct = yes`.

DuckDB then does predicate pushdown + Parquet partition pruning + row-group
skipping, and streams the JDBC `ResultSet` back to ERDDAP — so the whole table is
never on the heap.

::: {.callout-important title="Gotcha: serve time as a TIMESTAMP, not epoch double"}
ERDDAP reads any `time` variable (units `seconds since 1970-…`) via
`ResultSet.getTimestamp()`. If the DuckDB view exposes `time` as an epoch
**DOUBLE**, the DuckDB JDBC driver throws `NullPointerException: … varlen_data is
null` on *every* query that selects time. Expose `time` as a real **TIMESTAMP**
in the view (ERDDAP converts it to epoch seconds) and run the container in UTC.
Also: `EDDTableFromDatabase` does **not** auto-compute variable ranges, so set
`actual_range` (+ `geospatial_*`/`time_coverage_*`) on `time`/`lat`/`lon`/`depth`
or the form shows no min/max or sliders. (NetCDF, §4, computes these for you.)
:::

## 4 · NetCDF (`EDDTableFromNcCFFiles`) — the native ERDDAP path

This is Lynn's first option ("NetCDF per cast, à la your Trinidad Head setup") and
ERDDAP's bread-and-butter format. The catch is **scale**: CalCOFI has **~5.55 M
casts**, so one file per cast is 5.55 M files — far past what ERDDAP's file table
can manage. We bundle **by cruise** instead: **96 files**, one per `cruise_key`.

Each file is a **CF Discrete-Sampling-Geometry "profile"** dataset (contiguous
ragged array): `profile = cast`, `obs = depth level`, canonical sensors pivoted
**wide** to one variable each. We generate them by pivoting each cruise's long
rows in DuckDB and writing with R `ncdf4` (`libs/erddap_netcdf.R`,
`scripts/gen_ctd_netcdf.R`), one cruise at a time so memory stays bounded.

Two things fall out nicely:

- **ERDDAP auto-computes `actual_range`** for time/lat/lon/depth from the files —
  so the min/max + sliders work with **no** manual ranges (unlike DuckDB above).
- **`accessibleViaFiles=true`** lets users download the source `.nc` directly.

Trade-offs to weigh: the wide files carry **canonical sensors only** (non-canonical
variants are dropped); they **duplicate** the data (~2.3 GB of NetCDF for the full
table); and they must be **regenerated** when the data updates (vs the DuckDB view,
which always reflects the live Parquet). CalCOFI's casts are also fine-grained
(≈1 depth per `ctd_cast_uuid`), so these are effectively point-like profiles.

## 5 · Benchmark: NetCDF vs DuckDB vs Parquet vs CSV

Isolated ERDDAP (custom image) on our host, **2 GB heap** (deliberately small to
expose the differences). Queries are the **same intent** across backends — but
NetCDF is *wide* (sensors are columns), so it selects a named variable where the
long backends filter `measurement_type`: **q1** `.das`; **q2** one cruise +
temperature; **q3** one variable (salinity) over depth 0–200 m, all cruises;
**q4** whole-cruise dump.

```{r}
#| label: matrix
res |>
  transmute(table, approach,
            load = load_status, `load s` = load_secs,
            `load heap MB` = load_peak_heap_mb,
            `q2 cruise ms` = na_if(q2_cruise_ms, -1),
            `q3 var+depth ms` = na_if(q3_type200_ms, -1),
            `q4 dump ms` = na_if(q4_dump_ms, -1),
            `q3 size` = hsz(q3_rows)) |>
  arrange(table, approach) |> kable()
```

### Load memory — the headline

```{r}
#| label: mem-plot
#| fig-width: 8
#| fig-height: 4.2
#| fig-cap: "Peak JVM heap during dataset load at a 2 GB cap. NetCDF and DuckDB load the full 216 M-row table; Parquet OOMs and CSV times out (bars labeled ✗). DuckDB is flattest (~60 MB); NetCDF reads 95 file headers (~1.3 GB) but still fits."
res |>
  mutate(lab = ifelse(loaded, paste0(load_peak_heap_mb, " MB"),
                       paste0("✗ ", toupper(load_status)))) |>
  ggplot(aes(approach, load_peak_heap_mb, fill = loaded)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = lab), vjust = -0.3, size = 3.1) +
  geom_hline(yintercept = 2048, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.7, y = 2048, label = "2 GB heap cap", vjust = -0.4,
           hjust = 0, size = 2.8, colour = "grey40") +
  facet_wrap(~table) +
  scale_fill_manual(values = c(`TRUE` = "#2c7fb8", `FALSE` = "#d95f0e"),
                    labels = c(`TRUE` = "loaded", `FALSE` = "failed"), name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "peak load heap (MB)") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

### Query latency (backends that loaded)

```{r}
#| label: latency-plot
#| fig-width: 8.5
#| fig-height: 4.2
#| fig-cap: "Median query latency (log scale) for the backends that loaded. NetCDF wins the cruise-scoped queries (q2/q4: one subsetVariable → one file); for the cross-cruise single-variable scan (q3) NetCDF and DuckDB are comparable. q4 differs by schema: the long backends return every sensor row (more bytes) than the wide NetCDF."
res |>
  filter(loaded) |>
  transmute(table, approach,
            `q2 cruise` = na_if(q2_cruise_ms, -1),
            `q3 var+depth` = na_if(q3_type200_ms, -1),
            `q4 cruise dump` = na_if(q4_dump_ms, -1)) |>
  pivot_longer(starts_with("q"), names_to = "query", values_to = "ms") |>
  filter(!is.na(ms)) |>
  ggplot(aes(query, ms, fill = approach)) +
  geom_col(position = position_dodge2(width = 0.8, preserve = "single"), width = 0.7) +
  geom_text(aes(label = formatC(ms, format = "d", big.mark = ",")),
            position = position_dodge2(width = 0.8, preserve = "single"),
            vjust = -0.3, size = 2.6) +
  facet_wrap(~table, scales = "free_x") +
  scale_y_log10(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c(netcdf = "#1b9e77", duckdb = "#7570b3",
                               parquet = "#d95f0e", csv = "#999999"), name = NULL) +
  labs(x = NULL, y = "median latency (ms, log scale)") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

**Takeaways.** Both **NetCDF** and **DuckDB** load *both* tables (incl. 216 M rows)
within the 2 GB cap; **Parquet cannot load** the big table and **CSV** cannot even
load the thin one. DuckDB's load heap is flattest (~60 MB) because it streams from
the live Parquet; NetCDF's is higher (~1.3 GB) because ERDDAP reads all 95 file
headers, but it still fits — and NetCDF then serves **cruise-scoped queries
fastest of all** (q4 dump on the big table: **0.8 s** vs DuckDB's 99 s), since one
`cruise_key` maps to one file. The catch is that the wide/canonical NetCDF returns
*less* data per dump than the long backends (which emit every sensor row), so q4
is not a like-for-like byte comparison — q2/q3, where result sizes match closely,
are the fair head-to-head, and there NetCDF and DuckDB trade the lead.

## 6 · Partitioning & ordering for long-format bbox queries

> *"I wonder how organizing each sensor reading on its own row affects a query for
> data within a lat/lon/depth box."* — Lynn DeWitt

This is the key question for long format. We measured (DuckDB, the serving engine)
the same bounding-box query — **`temperature_ave`, 32–34°N, 119–121°W, 0–200 m
(~393 k rows)** — against four physical layouts of the 216 M-row denormalized
table:

- **A** — production view: raw Parquet partitioned by `cruise_key`, joined to
  `ctd_cast` at query time (what's live now).
- **B** — denormalized, partitioned by `cruise_key` (isolates layout from the join).
- **C** — denormalized, partitioned by `measurement_type` (one variable ⇒ ~1/N files).
- **D** — denormalized, partitioned by `measurement_type` **and** ordered by
  `latitude, longitude, depth` within each partition (row-group skipping for the box).

```{r}
#| label: layout
#| eval: !expr "!is.null(lay)"
lay |>
  transmute(layout, `build s` = build_secs, size, files = nfiles,
            `query s (median)` = query_secs_median, `result rows` = result_rows) |>
  kable()
```

```{r}
#| label: layout-plot
#| eval: !expr "!is.null(lay)"
#| fig-width: 7.5
#| fig-height: 3.2
#| fig-cap: "The same ~393k-row bbox query across four physical layouts of the long table (DuckDB on SSD). ~12× spread, but every layout is sub-second."
lay |>
  mutate(layout = factor(layout, levels = rev(layout))) |>
  ggplot(aes(layout, query_secs_median)) +
  geom_col(width = 0.7, fill = "#7570b3") +
  geom_text(aes(label = paste0(formatC(query_secs_median, format="f", digits=3), " s")),
            hjust = -0.15, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "median query (s)") + theme_minimal(base_size = 11)
```

```{r}
#| label: layout-missing
#| echo: false
#| output: asis
#| eval: !expr "is.null(lay)"
cat("*(layout_results.csv not present — run scripts/bench_parquet_layout.sh)*")
```

**How to read this.** Partition pruning (C, D) drops the data DuckDB must open
from all ~95 cruise files to just the one `measurement_type` partition. Spatial
ordering (D) additionally lets DuckDB skip Parquet **row groups** whose
lat/lon/depth min–max can't intersect the box. Decomposing the ~12× spread:

- **Removing the query-time join ≈ 4×** (A→B): materializing `time`/`lat`/`lon`
  into the table skips the join *and* lets the lat/lon predicate prune at scan.
- **Partitioning by `measurement_type` ≈ 2×** (B→C): a one-variable query opens
  1 of 53 partitions instead of all 95 cruise files.
- **Spatial ordering within each partition ≈ 1.4×** (C→D): DuckDB skips Parquet
  row groups whose `latitude/longitude/depth` min–max miss the box.

The trade-off is exactly your point: partitioning by `measurement_type` *helps
single-variable queries* (the common case) but *slows* a query that wants *all*
variables in a region. Our live datasets keep the **zero-extra-storage
logical-join view** (layout A, 0.2–0.9 s here); if heavy single-variable bbox
traffic emerges, the `measurement_type`-partition + spatial-sort layout is a cheap
~12× upgrade. (This is the same wide-vs-access-pattern tension that makes the
per-cruise *NetCDF* wide files fast for "all sensors at a cast" but slower for "one
sensor across a region.")

## 7 · Recommendations & open questions

- **If you host on CoastWatch ERDDAP, NetCDF is the most native fit** —
  `EDDTableFromNcCFFiles` over **per-cruise CF profile files (96 files)**, which we
  generate from the same pipeline. It loads within a small heap, auto-computes
  ranges/sliders, and serves cruise-scoped queries fastest. Costs: it carries
  canonical sensors only, duplicates ~2.3 GB, and must be regenerated on update.
- **If you want the tidy long table queryable as-is, DuckDB-over-Parquet** is the
  lightest option: a 274 KB `.db` of views on stock ERDDAP (only the driver jar is
  added), serving the 216 M-row table within ~60 MB. It always reflects the live
  Parquet. It's what runs at <https://erddap.calcofi.io> today.
- **Avoid plain `EDDTableFromParquetFiles` for large files** — whole-file on-heap
  reads OOM. Both options above are pragmatic bridges to native Parquet/DuckDB
  support (#447/#448).
- **Layout:** for the full long table, partition by `measurement_type` and order
  spatially within if single-variable bbox access dominates.
- **Open for you, Lynn:** which would be least work to receive and maintain —
  (a) per-cruise **NetCDF** files we generate, (b) **Parquet + the small DuckDB
  `.db`** (this approach), or (c) waiting on native DuckDB/ADBC? We can produce any
  from the same pipeline.

## 8 · Reproduce

Everything runs in the CalCOFI server containers (`../server/`), reading Parquet
from and writing outputs to the shared **`/share/data/erddap-duckdb`** tree, which
is mounted identically into `rstudio` (build), the live `erddap`, and the bench
`erddap_bench`. No host-specific paths.

**Code** (all in [CalCOFI/workflows](https://github.com/CalCOFI/workflows)):

| file | role |
|---|---|
| [`libs/erddap_netcdf.R`](https://github.com/CalCOFI/workflows/blob/main/libs/erddap_netcdf.R) | per-cruise CF NetCDF builder + `EDDTableFromNcCFFiles` XML |
| [`libs/erddap_duckdb.R`](https://github.com/CalCOFI/workflows/blob/main/libs/erddap_duckdb.R) | `.db` view builder + `EDDTableFromDatabase` XML |
| [`libs/erddap.R`](https://github.com/CalCOFI/workflows/blob/main/libs/erddap.R) | shared dataVariable / type / units helpers |
| [`scripts/gen_ctd_netcdf.R`](https://github.com/CalCOFI/workflows/blob/main/scripts/gen_ctd_netcdf.R) | driver: generate all per-cruise NetCDF |
| [`scripts/gen_prod_datasets.R`](https://github.com/CalCOFI/workflows/blob/main/scripts/gen_prod_datasets.R) · [`gen_bench_datasets.R`](https://github.com/CalCOFI/workflows/blob/main/scripts/gen_bench_datasets.R) | generate the dataset XML blocks |
| [`scripts/bench_erddap.sh`](https://github.com/CalCOFI/workflows/blob/main/scripts/bench_erddap.sh) | the NetCDF/DuckDB/Parquet/CSV matrix → `results_all.csv` |
| [`scripts/bench_parquet_layout.sh`](https://github.com/CalCOFI/workflows/blob/main/scripts/bench_parquet_layout.sh) | the partitioning/ordering experiment → `layout_results.csv` |
| [`server/erddap/Dockerfile`](https://github.com/CalCOFI/server/blob/main/erddap/Dockerfile) · [`docker-compose.bench.yml`](https://github.com/CalCOFI/server/blob/main/docker-compose.bench.yml) | ERDDAP + `duckdb_jdbc`; isolated bench instance |
| [`erddap/content/datasets.xml`](https://github.com/CalCOFI/erddap/blob/main/content/datasets.xml) | live dataset config |

**Data & artifacts** are browsable / downloadable via the Caddy file server at
[**file.calcofi.io/erddap-ctd**](https://file.calcofi.io/erddap-ctd/) — the result
CSVs, generated dataset XML, the DuckDB `.db`, and sample per-cruise `.nc`. The
served datasets themselves are at
<https://erddap.calcofi.io/erddap/tabledap/calcofi_ctd_thin.html> ·
<https://erddap.calcofi.io/erddap/tabledap/calcofi_ctd_measurement.html>.

The two result tables read above:

```{r}
#| label: data-files
read_csv("data/bench_erddap/results_all.csv", show_col_types = FALSE) |> kable(caption = "results_all.csv")
if (!is.null(lay)) kable(lay, caption = "layout_results.csv")
```

The per-cruise NetCDF builder and the benchmark harness, in full:

```{r}
#| label: embed-netcdf
#| eval: false
#| file: libs/erddap_netcdf.R
```

```{r}
#| label: embed-bench
#| eval: false
#| code-overflow: wrap
#| file: scripts/bench_erddap.sh
```
