---
title: "Deploy consumers"
subtitle: "Bring every server-side consumer onto the promoted release"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: deploy_consumers
  workflow_type: deploy
  dependency:
    - test_release
  output: _output/deploy_consumers.html
  workflow_url: https://calcofi.io/workflows/deploy_consumers.html
  description: >
    Reports which release each consumer is actually serving, and — when
    CALCOFI_DEPLOY=true — brings the server-side ones (db-viz-hex, db-viz-cruise,
    the h3t tile API) onto the promoted release. Runs after test_release, so it
    can only ever deploy a version that passed the consumer-contract suite.
editor_options:
  chunk_output_type: console
---

## Why this is a target

Every consumer that follows a release has drifted at least once, and always the
same way: `latest.txt` was correct and something *derived* from it lagged, with
nothing erroring. The releases index was a static page nobody regenerated; the
h3t API held its old database file open across a symlink flip; db-viz-hex's tile
tag was hardcoded and sat three releases behind.

The common cause is that the deploy lived only as prose in a runbook. Making it
a **target** rather than a step inside `test_release.qmd` means:

- it is visible in `tar_visnetwork()` / `tar_outdated()`, so "the release shipped
  but consumers were never updated" is a state you can *see*
- it can be re-run on its own (`tar_make(names = tidyselect::any_of("deploy_consumers"))`)
  without re-running the query suite
- it depends on `test_release`, so it can only ever deploy a version that passed
  the consumer-contract checks

GitHub-hosted consumers (`db-query`, `db-viz-station`) redeploy themselves on
dispatch from the promote step and are not handled here. ERDDAP has its own
target, `publish_to_erddap`, because its ~1.6 GB server-side parquet pull is the
slowest leg and is worth running independently.

```{r}
#| label: setup
#| message: false
#| warning: false
librarian::shelf(dplyr, glue, here, knitr, tibble, quiet = TRUE)
here <- here::here

DEPLOY   <- identical(Sys.getenv("CALCOFI_DEPLOY"), "true")
SSH_HOST <- Sys.getenv("CALCOFI_SSH_HOST", "calcofi")
`%||%`   <- function(x, y) if (is.null(x) || length(x) == 0 || !nzchar(x)) y else x

RELEASE <- tryCatch(
  trimws(readLines(
    "https://storage.googleapis.com/calcofi-db/ducklake/releases/latest.txt",
    warn = FALSE)[1]),
  error = function(e) NA_character_)

cat(glue("promoted release : {RELEASE}\n"))
cat(glue("deploy enabled   : {DEPLOY} (CALCOFI_DEPLOY)\n"))
cat(glue("ssh host         : {SSH_HOST}\n"))
```

## Which release is each consumer actually serving? ----

```{r}
#| label: status
# Read the state consumers actually expose, not what we believe we deployed.
# Each probe below is chosen because the OBVIOUS check is the one that lies:
#   * the h3t API's table list looks current even when it is serving an old file
#     — only `db_mtime` reveals which inode it has open (this is how a stale
#     v2026.08.03 was caught while the symlink already said v2026.08.04)
#   * an app returning HTTP 200 says nothing about which database it opened
ssh_lines <- function(cmd) {
  out <- suppressWarnings(system2(
    "ssh", c(SSH_HOST, shQuote(cmd)), stdout = TRUE, stderr = FALSE))
  out[nzchar(out)]
}
ssh_out <- function(cmd) paste(ssh_lines(cmd), collapse = " ")
# -L: several of these endpoints redirect (ERDDAP's info index 302s), and a
# redirect is a healthy response, not a failure.
http_code <- function(url) suppressWarnings(system2(
  "curl", c("-sL", "-o", "/dev/null", "-w", "%{http_code}", "--max-time", "45",
            shQuote(url)), stdout = TRUE))

# Which file does the h3t API actually have open?
#
# Ask it for the mtime, then match that against the mtimes on disk — IN R. The
# obvious version does the comparison inside the ssh command, but that nests
# quotes through glue -> ssh -> shell and silently produces no match, which is
# indistinguishable from "no file matched". Fetch the list, compare here.
h3t_serving <- function() {
  h <- suppressWarnings(system2("curl",
    c("-s", "--max-time", "45", "https://h3t.calcofi.io/h3t/health"),
    stdout = TRUE))
  h <- paste(h, collapse = "")
  mt <- sub('.*"mtime":"([0-9.]+)".*', "\\1", h)
  if (!grepl("^[0-9.]+$", mt)) return("(unreachable)")
  disk <- ssh_lines(
    "cd /share/github/CalCOFI/db-viz-hex/data && stat -c '%.6Y %n' calcofi_v*.duckdb")
  m <- regmatches(disk, regexpr("^[0-9.]+", disk))
  f <- sub("^[0-9.]+ +", "", disk)
  # compare to 3 decimals: the API reports the mtime it saw at open time, which
  # can differ from stat's in the last digits of float formatting
  hit <- which(substr(m, 1, nchar(mt) - 3) == substr(mt, 1, nchar(mt) - 3))
  if (length(hit)) f[hit[1]] else glue("(unmatched mtime {mt})")
}

consumer_status <- function() {
  tibble::tribble(
    ~consumer,        ~serving,                                                       ~probe,
    "db-viz-hex",     ssh_out("readlink /share/github/CalCOFI/db-viz-hex/data/calcofi_latest.duckdb"),
                      "calcofi_latest.duckdb symlink",
    "h3t API",        h3t_serving(),
                      "/h3t/health db_mtime -> file on disk",
    "ERDDAP",         ssh_out("grep -om1 'release v2026[0-9.]*' /share/github/CalCOFI/erddap/content/datasets.xml"),
                      "datasets.xml summary text",
    "db-viz-cruise",  ssh_out("stat -c %y /share/data/db-viz-cruise/db-viz-cruise.duckdb 2>/dev/null | cut -c1-16"),
                      "database mtime (carries no version)") |>
    mutate(http = c(http_code("https://app.calcofi.io/db-viz-hex/"),
                    http_code("https://h3t.calcofi.io/h3t/health"),
                    http_code("https://erddap.calcofi.io/erddap/info/index.json"),
                    http_code("https://app.calcofi.io/db-viz-cruise/")))
}

before <- consumer_status()
kable(before, caption = glue("Consumer state BEFORE (promoted release: {RELEASE})"))
```

## Deploy ----

```{r}
#| label: deploy
#| results: asis
if (!DEPLOY) {
  cat(glue(
    "**Not deployed.** `CALCOFI_DEPLOY` is not `\"true\"`, so this target ",
    "reported state and changed nothing.\n\n",
    "The server-side consumers above may still be serving a previous release. ",
    "To deploy, set `CALCOFI_DEPLOY=true` (in `~/.Renviron` to make it the ",
    "default for every release) and re-run:\n\n",
    "```r\n",
    "targets::tar_invalidate(names = tidyselect::any_of(\"deploy_consumers\"))\n",
    "targets::tar_make(names = tidyselect::any_of(\"deploy_consumers\"))\n",
    "```\n\n",
    "or run the script directly: `bash scripts/deploy_consumers.sh`.\n"))
} else {
  cat("Running `scripts/deploy_consumers.sh`.\n\n```\n")
  out <- system2("bash", c(here("scripts", "deploy_consumers.sh"),
                           "--release", RELEASE),
                 stdout = TRUE, stderr = TRUE)
  cat(paste(out, collapse = "\n"), "\n```\n")
  rc <- attr(out, "status") %||% 0L
  if (!identical(as.integer(rc), 0L))
    stop(glue(
      "consumer deploy FAILED for {RELEASE}. latest.txt is promoted and correct, ",
      "so calcofi4r and the hosted sites are fine — but db-viz-hex, db-viz-cruise ",
      "or the h3t API may still be serving the previous release."))
}
```

```{r}
#| label: status-after
#| eval: !expr DEPLOY
after <- consumer_status()
kable(after, caption = glue("Consumer state AFTER (promoted release: {RELEASE})"))

# The h3t API is the one that fails silently, so assert on it rather than
# trusting the script's own exit status.
h3t_now <- after$serving[after$consumer == "h3t API"]
if (nzchar(RELEASE) && nzchar(h3t_now) && !grepl(RELEASE, h3t_now, fixed = TRUE))
  warning(glue(
    "h3t API is serving {h3t_now}, not {RELEASE}. It holds its database file ",
    "open, so a symlink flip alone does not move it — the container restart in ",
    "step 3 is what does."))
```

```{r}
#| label: write-output
#| include: false
# `output:` in the calcofi block is this HTML; quarto writes it. Nothing else to do.
```
