---
title: "The CalCOFI grid from the official station positions"
subtitle: "Sites on their line, no slivers: today's grid against a Voronoi of the official stations and two repairs, measured on v2026.10.01"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    fig-align: center
editor_options:
  chunk_output_type: console
---

## Why

Ben, 2026-10-02 ([CalCOFI/workflows#130](https://github.com/CalCOFI/workflows/issues/130)):
a grid point, i.e. a **site** (x: station, y: line), must land on its line, and the grid has
**slivers** (the [calcofi4r "CalCOFI grid"
figure](https://calcofi.io/calcofi4r/articles/calcofi4r.html#calcofi-grid); related
[#86](https://github.com/CalCOFI/workflows/issues/86) and
[CalCOFI/db-viz-station#19](https://github.com/CalCOFI/db-viz-station/issues/19)). Rather than
patch, start from the [official station
positions](https://calcofi.org/sampling-info/station-positions/) and build the cells as a Voronoi
tessellation of them.

This is an **exploration**, not an ingest. It has no `calcofi:` block, like every other
`explore_*.qmd`, so `build_targets_list()` never sees it: it is not a target, not a release
input, and it writes nothing to `cc_stage_dir()/parquet`. It changes no `grid_key`. It reads
the published release v2026.10.01 and writes only the downloaded station table (under
`cc_stage_dir()/reference/`) and this page.

::: {.callout-tip title="Short answer"}
See [Recommendation](#sec-recommendation) and the headline table
@tbl-headline.
:::

```{r}
#| label: setup
librarian::shelf(
  calcofi4db, calcofi4r, DBI, dplyr, tidyr, ggplot2, glue, purrr, readr, sf, DT, quiet = T)
options(dplyr.summarise.inform = FALSE)
suppressMessages(sf_use_s2(FALSE))  # planar on lon/lat, the same predicate DuckDB's ST_Intersects uses in assign_grid_key()

ver      <- "v2026.10.01"       # the promoted release on 2026-10-02
CRS_M    <- 3310                # NAD83 / California Albers (m): every area, length and Voronoi below
KM_LINE  <- 22.2124             # km per line unit    (+proj=calcofi, measured below)
KM_STA   <- 7.3864              # km per station unit (+proj=calcofi, 4 nmi)
OFF_KM   <- 1                   # a site more than this from its line is "off its line"
SLIVER_W <- 2                   # km: a part narrower than this (mean width 2A/P) is a sliver ...
NOM_5    <- 5 * KM_STA * (10/3) * KM_LINE   # nominal area of a 5-unit nearshore cell, km²
SLIVER_A <- 0.25 * NOM_5        # km²: ... or smaller than a quarter of a 5-unit cell
CLIM_YR  <- c(1993, 2013)       # climatology window (build_climatology() default)

source(here::here("libs/download_station_positions.R"))
source(here::here("libs/reference_layers.R"))   # read_osm_land(): the release's land mask

# helpers ----
# polygon parts of a keyed sf, with area (km²) and mean width 2A/P (km)
split_parts <- function(x, key = "grid_key") {
  p <- x |> select(all_of(key)) |> st_cast("MULTIPOLYGON") |> st_cast("POLYGON", warn = FALSE)
  p$area_km2 <- as.numeric(st_area(p)) / 1e6
  p$width_km <- 2 * p$area_km2 / (as.numeric(st_length(st_cast(st_geometry(p), "MULTILINESTRING"))) / 1000)
  p |> group_by(across(all_of(key))) |>
    mutate(part = row_number(), n_parts = n(), largest = area_km2 == max(area_km2)) |> ungroup()
}
# a part is a sliver when it does not hold its cell's generator (the station it is the cell of)
# and it is thin or small
flag_slivers <- function(parts, gen, key = "grid_key") {
  gk  <- gen[[key]][match(parts[[key]], gen[[key]])]
  hit <- mapply(function(p, k) {
    gi <- which(gen[[key]] == k); length(gi) && st_intersects(p, st_geometry(gen)[gi], sparse = FALSE)[1, 1]
  }, st_geometry(parts), parts[[key]])
  parts$holds_gen <- hit
  parts$sliver    <- !hit & (parts$width_km < SLIVER_W | parts$area_km2 < SLIVER_A)
  # a sliver touching another cell is the visible defect; one touching none is a pocket of
  # water the land mask cuts off (a lagoon, a harbour) and stays with its cell
  nb <- st_is_within_distance(parts, parts, 50)
  parts$touches_other <- mapply(function(i, k) any(parts[[key]][i] != k), nb, parts[[key]])
  parts
}
# absorb every sliver into the neighbouring cell it shares the longest edge with,
# smallest first; a sliver with no neighbour keeps its key (reported)
absorb_slivers <- function(parts, key = "grid_key", tol_m = 50) {
  parts$key_new  <- parts[[key]]
  parts$shared_m <- NA_real_
  for (pid in order(parts$area_km2)) {
    if (!parts$sliver[pid]) next
    me     <- parts[pid, ]
    others <- which(parts$key_new != me$key_new)
    cand   <- others[st_intersects(st_buffer(st_geometry(me), tol_m), st_geometry(parts)[others], sparse = FALSE)[1, ]]
    if (!length(cand)) next
    bnd <- st_boundary(st_geometry(me))
    len <- vapply(cand, function(j) {
      x <- st_intersection(bnd, st_buffer(st_geometry(parts)[j], tol_m))
      if (!length(x) || all(st_is_empty(x))) 0 else sum(as.numeric(st_length(x)))
    }, numeric(1))
    if (max(len) <= 0) next
    parts$key_new[pid]  <- parts$key_new[cand[which.max(len)]]
    parts$shared_m[pid] <- max(len)
  }
  parts
}
# the first cell a point falls in (assign_grid_key()'s ST_Intersects ... LIMIT 1)
first_hit <- function(pts, cells, key = "grid_key") {
  ix <- st_intersects(pts, cells)
  vapply(ix, function(i) if (length(i)) cells[[key]][i[1]] else NA_character_, "")
}
# a site's distance from its line, km (+proj=calcofi via calcofi4db, never re-implemented)
off_line_km <- function(lon, lat, line) {
  p <- cc_lonlat_to_calcofi(lon, lat)
  abs(p$line - line) * KM_LINE
}
fmt <- function(x, d = 0) format(round(x, d), big.mark = ",", nsmall = d)
pct <- function(x, d = 1) paste0(format(round(100 * x, d), nsmall = d), "%")
```

## How today's grid is built {#sec-today-build}

The release's `grid` table (218 cells) is `calcofi4db::build_grid_reference()`, which reads two
datasets bundled in **calcofi4r**, `cc_grid` (cells) and `cc_grid_ctrs` (sites), and names each
cell `st{station}-ln{line}` (`_hist` for the historical pattern). Every ingest then calls
`calcofi4db::assign_grid_key()`, a point-in-polygon `ST_Intersects … LIMIT 1` of the sample's
position against those cells, so `grid_key` on `sample` and `obs` is wherever the position falls.

`cc_grid` comes from `calcofi4r/data-raw/cc_grid.R`, which:

1. lays three **idealized lattices** in `+proj=calcofi` (line, station) coordinates, not the
   occupied positions: 5-unit spacing for stations ≤ 60 on the standard lines, 10-unit spacing
   offshore to station 120, and a 20-unit historical lattice out to the convex hull of every
   CTD cast;
2. takes a Voronoi tessellation of each lattice (in line/station units), cuts the finer ones
   out of the coarser, and hand-merges six historical pairs (e.g. 55/80 from 50/80 + 60/80);
3. keeps the cells listed in `data-raw/cc_grid_sta-keys.csv` and erases Natural Earth 10 m land
   (`st_difference`), which leaves gaps along the coast;
4. glues five of those gaps back onto line 93.3 cells by hand (`cc_grid_box4slivers.geojson`,
   `st_union` onto `93.3,30` … `93.3,110`), which is where the line-93.3 cells pick up second,
   detached parts;
5. makes the site of each cell `st_centroid(geom, of_largest_polygon = TRUE)` on lon/lat
   (`cc_grid_ctrs`; the release's `grid.geom_ctr`).

So a site is the centroid of whatever polygon survived steps 2–4, never a station position:
that is why sites leave their line wherever land, a merge or a glued gap reshaped the cell. And
the lattice is nominal (35, 40, 45, 50 …) while the occupied stations are not (90.0 37, 90.0 53,
80.0 51, 83.3 39.4 …), which is why an inshore cell holds several real stations.

```{r}
#| label: read-release
cat_ <- cc_catalog(ver)
con  <- suppressMessages(dbConnect(duckdb::duckdb()))
invisible(dbExecute(con, "INSTALL httpfs; LOAD httpfs; INSTALL spatial; LOAD spatial"))
rp <- function(tbl) cc_read_parquet_sql(cc_release_sources(cat_, tbl))

gr <- dbGetQuery(con, glue("
  SELECT grid_key, line, station, shore, pattern, spacing, area_km2,
         ST_AsWKB(geom) AS wkb, ST_AsWKB(geom_ctr) AS wkb_ctr
  FROM {rp('grid')}"))
grid <- st_sf(
  select(gr, -wkb, -wkb_ctr),
  geom = st_as_sfc(structure(gr$wkb, class = "WKB"), crs = 4326))
grid_ctr <- st_as_sfc(structure(gr$wkb_ctr, class = "WKB"), crs = 4326)

# the release grid is calcofi4r::cc_grid, cell for cell
cg <- calcofi4r::cc_grid
cg_key <- with(cg, glue(
  "st{sub('.*,', '', sta_key)}-ln{sub(',.*', '', sta_key)}{ifelse(sta_pattern == 'historical', '_hist', '')}"))
stopifnot(setequal(cg_key, grid$grid_key))
n_same_geom <- sum(mapply(function(a, b) isTRUE(all.equal(a, b)),
                          st_geometry(grid), st_geometry(cg)[match(grid$grid_key, cg_key)]))

# sample: one local copy (26 MB)
invisible(dbExecute(con, glue("
  CREATE TABLE sample AS
  SELECT sample_key, dataset_key, grid_key, site_key, cruise_key, datetime, longitude, latitude
  FROM {rp('sample')}")))
clim_src <- cc_release_sources(cat_, "climatology")
invisible(dbExecute(con, glue(
  "CREATE TABLE climatology AS SELECT dataset_key, site_key, grid_key, month FROM {cc_read_parquet_sql(clim_src)}")))
n_sample <- dbGetQuery(con, "SELECT count(*) n FROM sample")$n
cat(glue("{ver}: grid {nrow(grid)} cells ({n_same_geom} identical to calcofi4r::cc_grid {packageVersion('calcofi4r')}); ",
         "sample {fmt(n_sample)} rows; climatology datasets ",
         "{paste(dbGetQuery(con, 'SELECT DISTINCT dataset_key FROM climatology ORDER BY 1')$dataset_key, collapse = ', ')}"), "\n")
```

## The official station positions {#sec-official}

`libs/download_station_positions.R` fetches the page and its linked files into
`cc_stage_dir()/reference/calcofi.org_station-positions/` and stops if the page's table, the
linked `CalCOFIStationOrder.csv` and `CalCOFI_113StationMap.kml` disagree.

```{r}
#| label: official
st <- download_station_positions()
fetched <- format(attr(st, "fetched"), "%Y-%m-%d")
n_ros <- sum(st$sta_type == "ROS"); n_sccoos <- sum(st$sta_type == "SCCOOS")
```

What the page lists (fetched `r fetched`):

- **Station Position (Lat/Lon), Depth, and Type**: `r nrow(st)` stations, `r n_ros` typed ROS
  and `r n_sccoos` SCCOOS, on `r n_distinct(st$line)` lines (60 to 93.4), with order occupied
  and estimated depth. This is the **113-station pattern**.
- The **75-station pattern** (standard since 1984; summer and fall): "66 stations from San Diego
  to Avila Beach" on the 6 lines 93.3 to 76.7, plus "9 additional coastal ~20 m SCCOOS stations".
  In the table that is every row with line ≥ 76.7: `r sum(st$in_75)` rows.
- The **104-station pattern** (winter and spring) adds 29 stations on lines 73.3 to 60.0; the 113
  pattern adds 9 more by sampling every 20 nm on line 66.7. The page does not say which 9, so
  the 104 subset is not marked.
- **SCCOOS Station Positions**: the same 9 SCCOOS stations (its CSV link returns 404).
- **Navy Stations**: `r sum(!is.na(st$navy_ops_area))` standard-pattern stations Navy operations
  may close, with the operations area (its CSV link returns 404).
- Not listed: any historical station (e.g. 90.0 32, occupied 1959–84), any line north of 60 or
  south of 93.4, or the Baja and Gulf of California lines.

### Each position against `+proj=calcofi`

`calcofi4db::cc_lonlat_to_calcofi()` projects each official lon/lat to continuous (line,
station); the residual is that minus the station's label, and the distance is from the official
position to `cc_calcofi_to_lonlat()` of the label.

```{r}
#| label: tbl-residuals
#| tbl-cap: "Official positions against `+proj=calcofi`: residuals in line and station units and the distance (m) between the listed position and the projected label, by station type."
pr <- cc_lonlat_to_calcofi(st$longitude, st$latitude)
nm <- cc_calcofi_to_lonlat(st$line, st$station)
st <- st |> mutate(
  d_line = pr$line - line, d_sta = pr$station - station,
  dist_m = as.numeric(st_distance(
    st_as_sf(st, coords = c("longitude", "latitude"), crs = 4326),
    st_as_sf(nm, coords = c("longitude", "latitude"), crs = 4326), by_element = TRUE)))
# km per unit, measured (the constants in setup)
u <- cc_calcofi_to_lonlat(c(90, 91, 90), c(60, 60, 61))
suppressMessages(sf_use_s2(TRUE))
km_unit <- as.numeric(st_distance(st_as_sf(u, coords = c("longitude", "latitude"), crs = 4326))[1, 2:3]) / 1000
suppressMessages(sf_use_s2(FALSE))
stopifnot(abs(km_unit[1] - KM_LINE) < 0.01, abs(km_unit[2] - KM_STA) < 0.01)
st |> group_by(sta_type) |>
  summarize(
    n = n(),
    `max |Δ line|` = signif(max(abs(d_line)), 2), `max |Δ station|` = signif(max(abs(d_sta)), 2),
    `median distance (m)` = round(median(dist_m), 1), `max distance (m)` = round(max(dist_m), 1)) |>
  knitr::kable()
```

Every official position is its label under `+proj=calcofi` to within
**`r round(max(st$dist_m))` m** (`r signif(max(abs(st$d_sta)), 2)` station units at most): the
table is the projection, rounded. One line unit is `r round(km_unit[1], 2)` km and one station
unit `r round(km_unit[2], 3)` km (4 nmi). So an official position *is* on its line, and a
Voronoi built from them inherits that.

## Today's grid {#sec-today}

```{r}
#| label: today
gm <- st_transform(grid, CRS_M)
nom_ll <- cc_calcofi_to_lonlat(grid$line, grid$station)
nom <- st_as_sf(cbind(st_drop_geometry(grid)[c("grid_key", "line", "station", "pattern")], nom_ll),
                coords = c("longitude", "latitude"), crs = 4326)
ctr_xy <- st_coordinates(grid_ctr)
ctr_c  <- cc_lonlat_to_calcofi(ctr_xy[, 1], ctr_xy[, 2])
grid <- grid |> mutate(
  ctr_line = ctr_c$line, ctr_sta = ctr_c$station,
  off_km   = abs(ctr_line - line) * KM_LINE,
  along_km = (ctr_sta - station) * KM_STA,
  nom_in_cell = mapply(function(p, c) st_intersects(p, c, sparse = FALSE)[1, 1],
                       st_geometry(nom), st_geometry(grid)))

# the coastline the release masks land with (OSM, ingest_spatial.qmd's reference layer)
bb   <- st_bbox(grid)
# the zip ingest_spatial.qmd downloads (~926 MB); a private stage dir reads the shared copy
land_zip <- c(cc_stage_path("reference", OSM_LAND_ZIP),
              path.expand(file.path("~/_big/calcofi/reference", OSM_LAND_ZIP)))
land_zip <- land_zip[file.exists(land_zip)][1]
stopifnot("no OSM land-polygons zip: run ingest_spatial.qmd's reference layers first" = !is.na(land_zip))
land <- read_osm_land(
  land_zip,
  bbox = c(xmin = bb[["xmin"]] - .5, ymin = bb[["ymin"]] - .5, xmax = bb[["xmax"]] + .5, ymax = bb[["ymax"]] + .5))
land_m <- st_union(st_transform(st_make_valid(land), CRS_M))

# outer hull of today's grid: the union with interior holes filled
hull_m  <- st_union(st_make_valid(gm))
outer_m <- st_union(st_sfc(
  lapply(st_geometry(st_cast(hull_m, "POLYGON")), function(p) st_polygon(list(p[[1]]))), crs = CRS_M))
gaps <- st_difference(st_difference(outer_m, hull_m), land_m) |> st_cast("POLYGON") |> st_sf()
gaps$area_km2 <- as.numeric(st_area(gaps)) / 1e6
gaps$at_coast <- st_is_within_distance(gaps, land_m, 100, sparse = FALSE)[, 1]

# slivers
parts_now <- split_parts(gm) |> left_join(st_drop_geometry(grid)[c("grid_key", "spacing", "pattern")], by = "grid_key")
parts_now <- flag_slivers(parts_now, st_transform(nom, CRS_M))

# official stations per cell
stp <- st_transform(st_as_sf(st, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE), CRS_M)
stp$grid_key_now <- first_hit(stp, gm)
n_per_cell <- table(stp$grid_key_now)
```

```{r}
#| label: tbl-today
#| tbl-cap: "Today's grid (v2026.10.01 `grid`, 218 cells): sites off their line, slivers, gaps and cells holding several official stations."
today <- tribble(
  ~measure, ~value,
  "cells", fmt(nrow(grid)),
  glue("sites (geom_ctr) more than {OFF_KM} km off their line"), glue("{sum(grid$off_km > OFF_KM)} ({sum(grid$off_km > OFF_KM & grid$pattern != 'historical')} outside the historical pattern)"),
  "sites more than 0.1 km off their line", fmt(sum(grid$off_km > 0.1)),
  "largest distance of a site from its line (km)", fmt(max(grid$off_km), 1),
  "median distance of a site from its line (km)", fmt(median(grid$off_km), 2),
  "cells whose own station (label) lies outside the cell", fmt(sum(!grid$nom_in_cell)),
  "cells in more than one piece", fmt(sum(parts_now$part == 1 & parts_now$n_parts > 1)),
  glue("sliver parts (no station of their own; mean width < {SLIVER_W} km or area < {fmt(SLIVER_A)} km²)"),
    glue("{sum(parts_now$sliver)} of {nrow(parts_now)} parts ({sum(parts_now$sliver & parts_now$n_parts == 1)} whole cells, {sum(parts_now$sliver & parts_now$n_parts > 1)} pieces of multi-part cells; {sum(parts_now$sliver & parts_now$touches_other)} touch another cell)"),
  glue("  of which thinner than {SLIVER_W} km"), fmt(sum(parts_now$sliver & parts_now$width_km < SLIVER_W)),
  "gaps inside the outer hull that are neither a cell nor OSM land (> 1 km²)",
    glue("{sum(gaps$area_km2 > 1)} ({sum(gaps$area_km2 > 1 & !gaps$at_coast)} away from the coast), {fmt(sum(gaps$area_km2))} km² in all"),
  "cells holding ≥ 2 official stations", glue("{sum(n_per_cell >= 2)} (holding {sum(n_per_cell[n_per_cell >= 2])} of {nrow(st)} stations)"),
  "official stations outside every cell", fmt(sum(is.na(stp$grid_key_now))))
knitr::kable(today, col.names = c("", ""))
```

The sub-kilometre offsets are built into the lattice: the lines are 3.3 and 3.4 units apart in
turn (80, 83.3, 86.7, 90 …), and a Voronoi edge sits at the midpoint, so the centre of every cell
on lines 76.7, 83.3, 86.7 and 93.3 is 0.025 line units (about 0.55 km) off its line, while lines
80 and 90 sit almost on it (the median offset of a site on a 10-unit cell is
`r fmt(median(grid$off_km[grid$spacing == 10 & grid$line %in% c(76.7, 83.3, 86.7, 93.3)]), 2)` km
on those four lines and
`r fmt(median(grid$off_km[grid$spacing == 10 & grid$line %in% c(80, 90)]), 2)` km on 80 and 90).
The large offsets are the cells that land, a hand-merge or a glued gap reshaped (@tbl-offline).

```{r}
#| label: tbl-offline
#| tbl-cap: "Sites more than 1 km off their line, outside the historical pattern: the site's continuous (line, station) under `+proj=calcofi`, its distance across the line and along it, the cell's area and whether its own station label is inside the cell."
grid |> st_drop_geometry() |>
  filter(off_km > OFF_KM, pattern != "historical") |>
  arrange(desc(off_km)) |>
  transmute(grid_key, pattern, area_km2 = round(area_km2),
            site = sprintf("%.2f / %.2f", ctr_line, ctr_sta),
            `off line (km)` = round(off_km, 1), `along line (km)` = round(along_km, 1),
            `station in cell` = nom_in_cell) |>
  knitr::kable()
```

```{r}
#| label: tbl-multi
#| tbl-cap: "Cells of today's grid that hold two or more official stations (113-station table)."
stp |> st_drop_geometry() |> filter(!is.na(grid_key_now)) |>
  group_by(grid_key = grid_key_now) |>
  summarize(n = n(), stations = paste(sprintf("%.1f %.1f%s", line, station, ifelse(sta_type == "SCCOOS", " (SCCOOS)", "")), collapse = "; ")) |>
  filter(n >= 2) |> arrange(desc(n), grid_key) |> knitr::kable()
```

```{r}
#| label: fig-today
#| warning: false
#| fig-cap: "Today's grid in the Southern California Bight and Point Conception: cells (grey; slivers red), sites (`grid.geom_ctr`; orange where more than 1 km off their line), official stations (blue; SCCOOS hollow) and the CalCOFI lines (thin)."
#| fig-height: 8
#| fig-width: 9
lines_ll <- map_dfr(sort(unique(st$line)), function(l) {
  s <- seq(20, 125, by = 0.5); ll <- cc_calcofi_to_lonlat(l, s); tibble(line = l, lon = ll$longitude, lat = ll$latitude)
}) |> st_as_sf(coords = c("lon", "lat"), crs = 4326) |> group_by(line) |> summarize(do_union = FALSE) |> st_cast("LINESTRING")
zoom <- coord_sf(xlim = c(-122.6, -117), ylim = c(32.3, 35.2), expand = FALSE)
site_pts <- st_sf(grid_key = grid$grid_key, off = grid$off_km > OFF_KM, geom = grid_ctr)
ggplot() +
  geom_sf(data = land, fill = "grey85", color = NA) +
  geom_sf(data = grid, fill = "grey97", color = "grey45", linewidth = 0.3) +
  geom_sf(data = st_transform(filter(parts_now, sliver), 4326), fill = "red", color = "red", linewidth = 0.6) +
  geom_sf(data = lines_ll, color = "steelblue", linewidth = 0.2, alpha = 0.6) +
  geom_sf(data = stp |> st_transform(4326), aes(shape = sta_type), color = "steelblue4", size = 1.8) +
  geom_sf(data = site_pts, aes(color = off), size = 1.4) +
  geom_sf_text(data = filter(site_pts, off), aes(label = grid_key), size = 2.3, nudge_y = 0.05, color = "darkorange3") +
  scale_shape_manual(values = c(ROS = 16, SCCOOS = 1), name = "official") +
  scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "darkorange"), name = "site > 1 km off line") +
  zoom + theme_minimal() + labs(x = NULL, y = NULL)
```

## Option A: Voronoi of the official positions {#sec-option-a}

The generators are the `r nrow(st)` official positions. Outside the official pattern (beyond
the convex hull of the official stations buffered by 20 nm, half the offshore spacing) today's
cells are kept as generators at their labelled (line, station) under `+proj=calcofi`, so the
historical and Baja extent keeps a cell for every sample it holds today; a generator on land is
dropped. The Voronoi is taken in metres (California Albers), not in line/station units, which
weigh a line unit (22.2 km) and a station unit (7.4 km) alike. It is clipped to today's outer
hull (with interior holes filled) and to the OSM land mask, and then the coastline's leftovers
(pieces cut off by land that hold no station: the same sliver rule) are absorbed into the
neighbour they share the longest edge with.

**How offshore cells are bounded.** An official station on the edge of the pattern (station 100
to 120) is bounded seaward by the next historical generator, i.e. the 20-unit lattice, and
finally by today's outer hull, so an outer official cell reaches about halfway to the historical
lattice beyond it. The alternative (a fixed buffer around the official stations, with
everything beyond keyed only to the historical cells) is a question below.

```{r}
#| label: option-a
dom <- st_buffer(st_convex_hull(st_union(stp)), 5 * KM_STA * 1000)
nom_m <- st_transform(nom, CRS_M)
on_land  <- st_intersects(nom_m, land_m, sparse = FALSE)[, 1]
keep_now <- !st_intersects(nom_m, dom, sparse = FALSE)[, 1] & !on_land
gen_a <- bind_rows(
  stp |> transmute(grid_key = glue("st{station}-ln{line}"), line, station,
                   gen_source = "official", sta_type),
  nom_m[keep_now, ] |> transmute(grid_key, line, station, gen_source = "today's cell", sta_type = NA_character_))
stopifnot(!anyDuplicated(gen_a$grid_key))
vor <- st_collection_extract(
  st_voronoi(st_union(st_geometry(gen_a)), envelope = st_as_sfc(st_bbox(st_buffer(outer_m, 5e5)))), "POLYGON")
idx <- vapply(st_intersects(gen_a, vor), function(i) i[1], integer(1))
stopifnot(!anyNA(idx), !anyDuplicated(idx))
clip_m <- st_difference(outer_m, land_m)
a_raw  <- st_sf(st_drop_geometry(gen_a), geom = st_intersection(st_geometry(vor)[idx], clip_m))
stopifnot(nrow(a_raw) == nrow(gen_a))

parts_a_raw <- flag_slivers(split_parts(a_raw), gen_a)
parts_a     <- absorb_slivers(parts_a_raw)
A <- parts_a |> group_by(grid_key = key_new) |> summarize() |>
  left_join(st_drop_geometry(gen_a), by = "grid_key")
A$area_km2 <- as.numeric(st_area(A)) / 1e6
parts_a_after <- flag_slivers(split_parts(A), gen_a)
# the site of an A cell is its generator: on its line by construction
A_site <- st_transform(gen_a, 4326)
A_off  <- off_line_km(st_coordinates(A_site)[, 1], st_coordinates(A_site)[, 2], A_site$line)
cat(glue(
  "A: {nrow(gen_a)} generators ({sum(gen_a$gen_source == 'official')} official, {sum(keep_now)} today's cells kept, ",
  "{sum(on_land)} of today's labels on land dropped); {nrow(A)} cells; ",
  "coastline slivers absorbed {sum(parts_a$key_new != parts_a$grid_key)} of {sum(parts_a_raw$sliver)} ",
  "({fmt(sum(parts_a_raw$area_km2[parts_a_raw$sliver & parts_a_raw$touches_other]), 1)} km²); ",
  "slivers left {sum(parts_a_after$sliver)}, of which {sum(parts_a_after$sliver & parts_a_after$touches_other)} touch another cell ",
  "and {sum(parts_a_after$sliver & !parts_a_after$touches_other)} are cut-off water pockets ",
  "({fmt(sum(parts_a_after$area_km2[parts_a_after$sliver & !parts_a_after$touches_other]), 1)} km² in all)"), "\n")
```

```{r}
#| label: tbl-option-a-cells
#| tbl-cap: "Option A: area of the official stations' cells by station type and pattern (km²)."
A |> st_drop_geometry() |> filter(gen_source == "official") |>
  left_join(st_drop_geometry(stp)[c("line", "station", "in_75")], by = c("line", "station")) |>
  group_by(sta_type, pattern = ifelse(in_75, "75-station", "113 only")) |>
  summarize(n = n(), min = round(min(area_km2)), median = round(median(area_km2)), max = round(max(area_km2))) |>
  knitr::kable()
```

```{r}
#| label: fig-option-a
#| fig-cap: "Option A in the Southern California Bight and Point Conception: Voronoi cells of the official stations (blue; SCCOOS hollow), today's cell edges dashed for comparison."
#| fig-height: 8
#| fig-width: 9
ggplot() +
  geom_sf(data = land, fill = "grey85", color = NA) +
  geom_sf(data = st_transform(A, 4326), aes(fill = gen_source), color = "grey30", linewidth = 0.3, alpha = 0.35) +
  geom_sf(data = grid, fill = NA, color = "darkorange", linewidth = 0.25, linetype = "dashed") +
  geom_sf(data = lines_ll, color = "steelblue", linewidth = 0.2, alpha = 0.6) +
  geom_sf(data = filter(A_site, gen_source == "official"), aes(shape = sta_type), color = "steelblue4", size = 1.8) +
  scale_shape_manual(values = c(ROS = 16, SCCOOS = 1), name = "official") +
  scale_fill_manual(values = c(official = "lightskyblue", `today's cell` = "grey80"), name = "generator") +
  zoom + theme_minimal() + labs(x = NULL, y = NULL)
```

## Option B: keep the cells, absorb the slivers {#sec-option-b}

Every sliver part of today's grid (the rule in @tbl-today) joins the neighbouring cell it shares
the longest edge with, smallest first. A whole-cell sliver retires its key. Sites are recomputed
as today's build does (centroid of the largest polygon).

```{r}
#| label: option-b
parts_b <- absorb_slivers(parts_now)
B <- parts_b |> group_by(grid_key = key_new) |> summarize()
B_site <- suppressWarnings(st_centroid(st_transform(B, 4326), of_largest_polygon = TRUE)) |>
  left_join(st_drop_geometry(grid)[c("grid_key", "line", "pattern")], by = "grid_key")
B_off <- off_line_km(st_coordinates(B_site)[, 1], st_coordinates(B_site)[, 2], B_site$line)
retired_b <- setdiff(grid$grid_key, B$grid_key)
parts_b_after <- flag_slivers(split_parts(B), st_transform(nom, CRS_M))
cat(glue("B: {sum(parts_b$key_new != parts_b$grid_key)} of {sum(parts_b$sliver)} slivers absorbed; {nrow(B)} cells; ",
         "retired: {paste(retired_b, collapse = ', ')}"), "\n")
```

```{r}
#| label: tbl-option-b
#| tbl-cap: "Option B: each sliver of today's grid, outside the historical pattern, and the cell it joins."
parts_b |> st_drop_geometry() |>
  filter(key_new != grid_key, pattern != "historical") |>
  arrange(grid_key, part) |>
  transmute(grid_key, part, n_parts, `area (km²)` = signif(area_km2, 3), `mean width (km)` = signif(width_km, 2),
            `joins` = key_new, `shared edge (km)` = round(shared_m / 1000, 1)) |>
  knitr::kable()
```

## Option C: keep the cells, move each site onto its line {#sec-option-c}

Each site moves to the nearest point of its line (in `+proj=calcofi`, the line is `x = line`)
that lies inside its own cell. Cells, and so every `grid_key`, are unchanged; only
`grid.geom_ctr` moves. A cell its own line does not cross cannot be fixed this way.

```{r}
#| label: option-c
gc <- st_transform(grid, st_crs("+proj=calcofi"))
c_sta <- vapply(seq_len(nrow(gc)), function(i) {
  ln  <- st_sfc(st_linestring(cbind(gc$line[i], c(-300, 400))), crs = st_crs(gc))
  seg <- st_intersection(ln, st_geometry(gc)[i])
  if (!length(seg) || all(st_is_empty(seg))) return(NA_real_)
  y <- vapply(st_geometry(st_cast(seg, "LINESTRING")), function(p) {
    r <- range(st_coordinates(p)[, "Y"]); min(max(grid$ctr_sta[i], r[1]), r[2]) }, numeric(1))
  y[which.min(abs(y - grid$ctr_sta[i]))]
}, numeric(1))
c_ll <- cc_calcofi_to_lonlat(grid$line, c_sta)
c_ok <- !is.na(c_sta)
suppressMessages(sf_use_s2(TRUE))
c_move_km <- rep(NA_real_, nrow(grid))
c_move_km[c_ok] <- as.numeric(st_distance(
  grid_ctr[c_ok], st_as_sfc(st_as_sf(c_ll[c_ok, ], coords = c("longitude", "latitude"), crs = 4326)), by_element = TRUE)) / 1000
suppressMessages(sf_use_s2(FALSE))
cat(glue("C: {sum(c_ok)} sites moved onto their line (median {round(median(c_move_km, na.rm = TRUE), 2)} km, ",
         "{sum(c_move_km > OFF_KM, na.rm = TRUE)} by more than {OFF_KM} km, max {round(max(c_move_km, na.rm = TRUE), 1)} km); ",
         "{sum(!c_ok)} cells their line does not cross: {paste(grid$grid_key[!c_ok], collapse = ', ')}"), "\n")
```

```{r}
#| label: fig-options-bc
#| fig-cap: "Options B and C in the Southern California Bight: B's cells after absorbing slivers (grey; absorbed pieces red) and C's sites moved onto their line (orange arrows from today's site to the new one; black crosses where the line misses the cell)."
#| fig-height: 8
#| fig-width: 9
mv <- tibble(grid_key = grid$grid_key, x0 = ctr_xy[, 1], y0 = ctr_xy[, 2],
             x1 = c_ll$longitude, y1 = c_ll$latitude, km = c_move_km)
ggplot() +
  geom_sf(data = land, fill = "grey85", color = NA) +
  geom_sf(data = st_transform(B, 4326), fill = "grey97", color = "grey45", linewidth = 0.3) +
  geom_sf(data = st_transform(filter(parts_b, key_new != grid_key), 4326), fill = "red", color = "red", alpha = 0.5) +
  geom_sf(data = lines_ll, color = "steelblue", linewidth = 0.2, alpha = 0.6) +
  geom_segment(data = filter(mv, km > 0.5), aes(x = x0, y = y0, xend = x1, yend = y1),
               color = "darkorange", arrow = arrow(length = unit(0.08, "cm"))) +
  geom_point(data = filter(mv, is.na(km)), aes(x0, y0), shape = 4, size = 2) +
  zoom + theme_minimal() + labs(x = NULL, y = NULL)
```

## What each option does to `sample` and the climatology {#sec-effect}

Every distinct `sample` position in `r ver` is assigned a cell under today's grid (recomputed
from the release's `grid`, the check that the method reproduces the release), Option A and
Option B; Option C keeps every cell, so it changes no `grid_key`. A row "changes" when its
cell's key differs from today's recomputed key; "moves" counts only the rows whose new cell is
not today's cell under a new name (each new cell is matched to the today's cell it overlaps
most).

```{r}
#| label: assign
pos <- dbGetQuery(con, "
  SELECT longitude, latitude, count(*) AS n
  FROM sample WHERE longitude IS NOT NULL AND latitude IS NOT NULL
    AND isfinite(longitude) AND isfinite(latitude)
  GROUP BY ALL")
pos_sf <- st_as_sf(pos, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
pos_m  <- st_transform(pos_sf, CRS_M)
pos$key_now <- first_hit(pos_sf, grid)
pos$key_a   <- first_hit(pos_m, A)
pos$key_b   <- first_hit(pos_m, B)
# the today's cell each A cell overlaps most: a key change into it is a rename, not a move
ov <- suppressWarnings(st_intersection(select(A, key_a = grid_key), select(gm, key_now = grid_key)))
ov$a <- as.numeric(st_area(ov))
xw_a <- st_drop_geometry(ov) |> group_by(key_a) |> slice_max(a, n = 1, with_ties = FALSE) |> select(key_a, key_now_main = key_now)
pos <- pos |> left_join(xw_a, by = "key_a")
dbWriteTable(con, "pos_key", pos |> select(longitude, latitude, key_now, key_a, key_b, key_now_main), overwrite = TRUE)
invisible(dbExecute(con, "
  CREATE OR REPLACE TABLE s_key AS
  SELECT s.*, p.key_now, p.key_a, p.key_b, p.key_now_main
  FROM sample s LEFT JOIN pos_key p USING (longitude, latitude)"))
agree <- dbGetQuery(con, "
  SELECT count(*) FILTER (WHERE grid_key IS NOT DISTINCT FROM key_now) AS same,
         count(*) FILTER (WHERE grid_key IS NOT NULL AND key_now IS NULL) AS rel_only,
         count(*) FILTER (WHERE grid_key IS NULL AND key_now IS NOT NULL) AS now_only,
         count(*) AS n FROM s_key")
cat(glue("recomputed today's key = released grid_key on {fmt(agree$same)} of {fmt(agree$n)} rows ({pct(agree$same / agree$n, 2)}); ",
         "{fmt(agree$rel_only)} released keys without a position match, {fmt(agree$now_only)} positions the release left ungridded"), "\n")
```

```{r}
#| label: tbl-agree
#| tbl-cap: "Where the recomputed key and the released `sample.grid_key` differ, by dataset: rows the release left without a `grid_key` although their position falls in a cell, and rows keyed to a different cell."
dbGetQuery(con, "
  SELECT dataset_key,
         count(*) FILTER (WHERE grid_key IS NULL AND key_now IS NOT NULL) AS released_null_but_in_a_cell,
         count(*) FILTER (WHERE grid_key IS NOT NULL AND key_now IS NOT NULL AND grid_key <> key_now) AS different_cell
  FROM s_key GROUP BY 1
  HAVING released_null_but_in_a_cell + different_cell > 0 ORDER BY 2 DESC") |>
  knitr::kable(format.args = list(big.mark = ","))
```

```{r}
#| label: tbl-per-dataset
#| tbl-cap: "Share of `sample` rows per dataset whose `grid_key` would change under each option (against today's grid recomputed from the release). A: key changes, and moves (new cell is not today's cell renamed). B: key changes. C changes none."
per_ds <- dbGetQuery(con, "
  SELECT dataset_key,
         count(*)                                                               AS n,
         count(key_now)                                                         AS n_gridded,
         count(*) FILTER (WHERE key_a IS DISTINCT FROM key_now)                 AS a_changed,
         count(*) FILTER (WHERE key_now_main IS DISTINCT FROM key_now)          AS a_moved,
         count(*) FILTER (WHERE key_b IS DISTINCT FROM key_now)                 AS b_changed
  FROM s_key GROUP BY 1 ORDER BY n DESC")
tot <- per_ds |> summarize(dataset_key = "all", across(-dataset_key, sum))
bind_rows(per_ds, tot) |>
  transmute(dataset_key, rows = fmt(n), gridded = fmt(n_gridded),
            `A: key changes` = pct(a_changed / n), `A: moves` = pct(a_moved / n),
            `B: key changes` = pct(b_changed / n), `C` = pct(0)) |>
  knitr::kable(align = "lrrrrrr")
```

The climatology (`build_climatology()`, calcofi4db ≥ 4.8.0) is keyed on the **station**
(`sample.site_key`), not the cell, so no option changes a climatology mean; `grid_key` rides on
each row only as the station's modal cell. Two measures, on the climatology's own inputs
(`calcofi_bottle`, `calcofi_ctd-cast`, `calcofi_dic` samples with a `site_key`,
`r CLIM_YR[1]`–`r CLIM_YR[2]`):

- **cells × month with ≥ 3 cruises**, the grain a cell-keyed baseline (or a map consumer) has;
- **climatology rows whose `grid_key` label would change**: the station's modal cell under each
  option, joined to the release's `climatology` rows.

```{r}
#| label: tbl-climatology
#| tbl-cap: "Effect on the climatology grain: cell × calendar month with at least 3 distinct cruises (1993–2013, climatology input datasets), and climatology rows whose modal-cell label would change."
clim_in <- "dataset_key IN ('calcofi_bottle', 'calcofi_ctd-cast', 'calcofi_dic') AND site_key IS NOT NULL
            AND year(datetime) BETWEEN {CLIM_YR[1]} AND {CLIM_YR[2]}"
cells_month <- function(k) dbGetQuery(con, glue("
  SELECT count(*) n FROM (
    SELECT {k}, month(datetime) m FROM s_key WHERE {k} IS NOT NULL AND {glue(clim_in)}
    GROUP BY ALL HAVING count(DISTINCT cruise_key) >= 3)"))$n
modal <- function(k) glue("
  SELECT dataset_key, site_key, mode({k}) AS cell FROM s_key WHERE {glue(clim_in)} GROUP BY ALL")
clim_change <- function(k) dbGetQuery(con, glue("
  WITH now AS ({modal('key_now')}), opt AS ({modal(k)})
  SELECT count(*) AS n_rows, count(*) FILTER (WHERE now.cell IS DISTINCT FROM opt.cell) AS changed,
         count(DISTINCT (c.dataset_key, c.site_key)) FILTER (WHERE now.cell IS DISTINCT FROM opt.cell) AS stations
  FROM climatology c JOIN now USING (dataset_key, site_key) JOIN opt USING (dataset_key, site_key)"))
# the release's own label against the recomputed modal (they differ only where obs weighting differs)
lab <- dbGetQuery(con, glue("
  WITH now AS ({modal('key_now')})
  SELECT count(*) n, count(*) FILTER (WHERE c.grid_key IS NOT DISTINCT FROM now.cell) same
  FROM climatology c JOIN now USING (dataset_key, site_key)"))
ca <- clim_change("key_a"); cb <- clim_change("key_b")
clim_tbl <- tibble(
  option = c("today", "A", "B", "C"),
  `cells × month with ≥ 3 cruises` = c(cells_month("key_now"), cells_month("key_a"), cells_month("key_b"), cells_month("key_now")),
  `climatology rows relabelled` = c("", glue("{fmt(ca$changed)} of {fmt(ca$n_rows)} ({ca$stations} stations)"),
                                    glue("{fmt(cb$changed)} of {fmt(cb$n_rows)} ({cb$stations} stations)"), "0"))
knitr::kable(clim_tbl)
```

The recomputed modal cell matches the released `climatology.grid_key` on
`r fmt(lab$same)` of `r fmt(lab$n)` rows (the release takes the mode over `obs` rows, this over
samples).

### Cells that hold several stations, in the data

```{r}
#| label: tbl-stations-per-cell
#| tbl-cap: "Cells holding two or more distinct CTD stations (`calcofi_ctd-cast` `site_key`, each occupied on at least 3 cruises since 2004 and counted in the cell most of its casts fall in), under today's grid and each option."
# a station's cell is the cell most of its casts fall in (casts scatter within ~2 nmi of the
# position, so a station near an edge has casts on both sides)
sk <- dbGetQuery(con, "
  SELECT site_key, mode(key_now) AS key_now, mode(key_a) AS key_a, mode(key_b) AS key_b
  FROM s_key WHERE dataset_key = 'calcofi_ctd-cast' AND site_key IS NOT NULL AND year(datetime) >= 2004
  GROUP BY site_key HAVING count(DISTINCT cruise_key) >= 3")
# normalize the station label: source forms differ ('093.3 026.4', '93.3    26.4', '090.0 27.76')
sk <- sk |> mutate(
  st_line = as.numeric(sub("^\\s*(\\S+)\\s+.*$", "\\1", site_key)),
  st_sta  = as.numeric(sub("^.*\\s+(\\S+)\\s*$", "\\1", site_key)),
  station = sprintf("%.1f %.1f", round(st_line, 1), round(st_sta, 1))) |>
  distinct(station, .keep_all = TRUE)
multi <- function(k) sk |> filter(!is.na(.data[[k]])) |> count(cell = .data[[k]]) |> filter(n >= 2)
tibble(
  option = c("today", "A", "B", "C"),
  `cells with ≥ 2 stations` = c(nrow(multi("key_now")), nrow(multi("key_a")), nrow(multi("key_b")), nrow(multi("key_now"))),
  `stations in such cells` = c(sum(multi("key_now")$n), sum(multi("key_a")$n), sum(multi("key_b")$n), sum(multi("key_now")$n)),
  `most stations in one cell` = c(max(multi("key_now")$n), max(c(0, multi("key_a")$n)), max(multi("key_b")$n), max(multi("key_now")$n))) |>
  knitr::kable()
```

In Option A a cell still holds two stations in the data where a station that is not in the
official table (†), or casts labelled with an official station but taken nearer another, was occupied on 3 or more cruises since 2004 (@tbl-a-multi): those stations key to
the nearest official station, which is question 3 below.

```{r}
#| label: tbl-a-multi
#| tbl-cap: "Option A cells holding two or more CTD stations in the data (each occupied on ≥ 3 cruises since 2004, in the cell most of its casts fall in); stations not in the official table are marked †."
off_keys <- sprintf("%.1f %.1f", st$line, st$station)
sk |> filter(!is.na(key_a)) |>
  group_by(key_a) |> filter(n() >= 2) |>
  summarize(n = n(), stations = paste(ifelse(station %in% off_keys, station, paste0(station, " †")), collapse = "; ")) |>
  arrange(desc(n), key_a) |> rename(grid_key = key_a) |> knitr::kable()
```

## Headline {#sec-headline}

```{r}
#| label: tbl-headline
#| tbl-cap: "The three options against today's grid, on v2026.10.01."
row_share <- function(col) pct(sum(per_ds[[col]]) / sum(per_ds$n))
hl <- tibble(
  ` ` = c("cells", glue("sites > {OFF_KM} km off their line"), "sliver parts touching another cell",
          "cut-off water pockets (no neighbour)", "cells holding ≥ 2 official stations",
          "sample rows whose grid_key changes", "  of which a real move (not a rename)",
          "cells × month with ≥ 3 cruises (1993–2013)", "climatology rows relabelled"),
  today = c(nrow(grid), sum(grid$off_km > OFF_KM), sum(parts_now$sliver & parts_now$touches_other),
            sum(parts_now$sliver & !parts_now$touches_other), sum(n_per_cell >= 2),
            "", "", clim_tbl[[2]][1], ""),
  A = c(nrow(A), sum(A_off > OFF_KM), sum(parts_a_after$sliver & parts_a_after$touches_other),
        sum(parts_a_after$sliver & !parts_a_after$touches_other), sum(table(first_hit(stp, A)) >= 2),
        row_share("a_changed"), row_share("a_moved"), clim_tbl[[2]][2], clim_tbl[[3]][2]),
  B = c(nrow(B), sum(B_off > OFF_KM), sum(parts_b_after$sliver & parts_b_after$touches_other),
        sum(parts_b_after$sliver & !parts_b_after$touches_other),
        sum(table(first_hit(stp, B)) >= 2), row_share("b_changed"), row_share("b_changed"), clim_tbl[[2]][3], clim_tbl[[3]][3]),
  C = c(nrow(grid), sum(!c_ok), sum(parts_now$sliver & parts_now$touches_other),
        sum(parts_now$sliver & !parts_now$touches_other), sum(n_per_cell >= 2), "0.0%", "0.0%", clim_tbl[[2]][4], "0"))
hl <- mutate(hl, across(everything(), as.character))
pocket_km2_a <- sum(parts_a_after$area_km2[parts_a_after$sliver & !parts_a_after$touches_other])
knitr::kable(hl)
```

## Recommendation {#sec-recommendation}

**Option A**, with a key crosswalk. It is the only option that fixes both defects at the source:
every modern site *is* an official position, which `+proj=calcofi` reproduces to within
`r round(max(st$dist_m))` m, so all `r nrow(A)` sites sit on their line by construction; no cell
is a remnant of an idealized lattice, so the slivers left are `r hl$A[3]` between cells plus
`r hl$A[4]` water pockets the OSM coastline cuts off (`r fmt(pocket_km2_a)` km² in all); and no
cell holds two official stations (today `r hl$today[5]` cells do), which ends the station mixing
that #86 and the climatology work routed around with `site_key`. Option B removes the slivers
but keeps the nominal lattice (`r hl$B[2]` sites still more than `r OFF_KM` km off their line,
`r hl$B[5]` cells still holding several official stations); Option C moves the dots and nothing
else, and cannot place `r sum(!c_ok)` sites at all because their own line misses their cell.

The price of A is `r hl$A[6]` of `sample` rows changing `grid_key`, `r hl$A[7]` moving to a
different cell rather than a renamed one (@tbl-per-dataset), and `r fmt(ca$changed)` of
`r fmt(ca$n_rows)` climatology rows (`r ca$stations` stations) taking a new modal-cell label; the climatology means do not move (they are keyed
on `site_key`), and the cell-month grain with three or more cruises grows from
`r clim_tbl[[2]][1]` to `r clim_tbl[[2]][2]` because the inshore stations stop sharing cells. So
the release that ships A needs a `grid_key` crosswalk (old to new, by largest overlap, as
computed here) beside the per-dataset diff in `RELEASES.md`, and the answers to the questions
below first.

## Questions for Erin and Betty {#sec-questions}

1. **Naming of new cells.** Name a cell by its official station (`st37-ln90`, `st27.7-ln90`),
   which renames the inshore cells, or keep today's names where a new cell is mostly today's
   cell (`st35-ln90` for the cell of 90.0 37)? Decimal stations (`st27.7`, `st26.4-ln93.4`) in a
   key: acceptable?
2. **Inshore SCCOOS stations.** Each SCCOOS station gets its own small cell in Option A
   (@tbl-option-a-cells; against about 2,700 km² for a 5-unit cell today). Is that the intent, or should the SCCOOS
   stations share the cell of the nearest ROS station (90.0 27.7 with 90.0 28, 93.4 26.4 with
   93.3 26.7), so a cell is a ROS station and its SCCOOS neighbour?
3. **A historical off-grid cast.** In Option A a cast at a station no longer occupied keys to
   the nearest official station's cell when it lies within 20 nm of the official pattern
   (90.0 32, 1959–84, keys to `r first_hit(st_transform(st_as_sf(cc_calcofi_to_lonlat(90, 32), coords = c("longitude", "latitude"), crs = 4326), CRS_M), A)`), and to today's historical cell
   beyond that (the Baja lines). Is that right, or should historical stations keep cells of
   their own (generators from every station occupied on N or more cruises, not only the
   official table)? @tbl-a-multi lists the stations this already affects since 2004.
4. **Offshore bound.** Should an outer official station's cell stop at a fixed distance (20 nm,
   half the offshore spacing) with the historical lattice beyond it, rather than reach halfway
   to the next historical generator?
5. **The 104 vs 113 pattern.** Which 9 stations of line 66.7 are the 113-only infill (every 20
   nm)? The page does not say; Option A gives all 11 line-66.7 stations a cell.
6. **The Navy stations** are ordinary cells here. Anything about them a map should show?

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
