---
title: "explore_db"
---

## Species

While working on [+ larvae into oceano app · Issue #23 · CalCOFI/apps](https://github.com/CalCOFI/apps/issues/23), noticing some funkiness with `species_codes` and `species_codes_new` tables.

```{r}
#| code-fold: true

librarian::shelf(
  calcofi/calcofi4r,
  dplyr, DT, glue, here, 
  quiet = T)

source(here("../apps_dev/libs/db.R")) # con: database connection 
```

### `species_codes`

```{r}
d_spp <- tbl(con, "species_codes") |> 
  collect()
dim(d_spp)
table(d_spp$taxon_rank, useNA="ifany")
```

### `species_codes_new`

```{r}
d_spp_new <- tbl(con, "species_codes_new") |> 
  collect()
dim(d_spp_new)
table(d_spp_new$taxon_rank, useNA="ifany")
```

### `species_codes` vs `species_codes_new`

```{r}
d_spp |> 
  left_join(d_spp_new, by="scientific_name") |>
  filter(is.na(taxon_rank.x) | is.na(taxon_rank.y)) |> 
  # names() |> paste(collapse=', ') |> cat()
  select(
    scientific_name, 
    id.x, id.y, 
    spccode.x, spccode.y, 
    itis_tsn.x, itis_tsn.y, 
    common_name.x, common_name.y,
    taxon_rank.x, taxon_rank.y) |> 
  datatable()
```

## Questions

- Why `is.na(taxon_rank)` (n=34) in `species_codes_new` but not `species_codes`?
- What is `taxon_rank == 'None'`?

## Rename suggestions

- Drop `species_codes_new` since missing `taxon_rank` for 34 rows that are populated in `species_codes`
- Rename `species_codes` to `species`
- Drop field `id` <int>, since confusing with `spccode` <int>
- Rename field `spccode` to `sp_id` since <int>

