---
title: "Join Larval and Bottle Data"
format: 
  html:
    toc: true
    number-sections: true
editor_options: 
  chunk_output_type: console
---

## Load packages and Connect to Database

With [`cc_db_connect()`](https://calcofi.io/calcofi4r/reference/cc_db_connect.html), note that you must either be running this from the CalCOFI server or have a tunnelled SSH connection with a user account on the server.

```{r}
#| warning: false

if (!require("librarian"))
  install.packages("librarian")
# use librarian::shelf() to load packages, installing if needed
librarian::shelf(
  calcofi/calcofi4r,
  dplyr, DBI, DT, glue, readr, sf, skimr,
  quiet = T)

tbl_head <- function(tbl, n=10){
  dbGetQuery(
    con, 
    glue("SELECT * FROM {tbl} LIMIT {n}")) %>% 
  datatable()
}
tbl_fields <- function(tbl){
  dbListFields(con, tbl) %>% paste(collapse=", ")
}

con <- cc_db_connect()
```

## Create function `get_nearest_bottle()`

```{r}
# create function to form query and return results
get_nearest_bottle <- function(
    fld,
    datetime, max_hrs,
    lon, lat, max_kms,
    depth_min, depth_max,
    verbose = FALSE){
  
  message(glue("{datetime}: {lon}, {lat}"))

  sql <- glue(
    "SELECT 
      b.{fld},
      c.datetime, c.longitude, c.latitude,
      b.depthm,
      ST_Distance(c.geom, ST_GeomFromText('POINT({lon} {lat})', 4326)::GEOGRAPHY)/1000
        AS diff_kms,
      EXTRACT(EPOCH FROM c.datetime - '{datetime}'::TIMESTAMP)/3600
        AS diff_hrs
    FROM ctd_bottles AS b
    LEFT JOIN
      ctd_casts AS c USING (cast_count)
    WHERE
      ST_DWithin(
        c.geom::GEOGRAPHY,
        ST_GeomFromText('POINT({lon} {lat})', 4326)::GEOGRAPHY,
        {max_kms} * 1000) AND
      c.datetime <= ('{datetime}'::TIMESTAMP + INTERVAL '{max_hrs} HOUR') AND
      c.datetime >= ('{datetime}'::TIMESTAMP - INTERVAL '{max_hrs} HOUR') AND
    depthm <= {depth_max} AND
    depthm >= {depth_min}")
  
  if (verbose)
    message(sql)
  
  dbGetQuery(con, sql) %>% 
    tibble()
}

# run with "most recent" larval data attributable in space and time
x <- get_nearest_bottle(
  fld       = "t_degc",
  datetime  = "1999-12-06 20:32:00 UTC", 
  max_hrs   = 800,
  lon       = -117.8483,
  lat       =  29.11833, 
  max_kms   = 1000,
  depth_min = 0,
  depth_max = 20,
  verbose   = TRUE)
datatable(x)

# render temperature using simple mean
mean(x$t_degc)
# compare weighted means using seperation difference in space (kilometers)...
weighted.mean(x$t_degc, x$diff_kms)
# ... or time (hours)
weighted.mean(x$t_degc, x$diff_hrs)
```

#### Added `ctd_casts.datetime`

```{r}
# add a single field with datetime to make querying easier
dbSendQuery(
  con,
  "ALTER TABLE ctd_casts ADD COLUMN IF NOT EXISTS datetime TIMESTAMP")
dbSendQuery(
  con,
  "UPDATE ctd_casts 
  SET datetime = (date::text || ' ' || time::text || ' America/Los_Angeles')::timestamp")
```

## Get Space & Time for Larval Data 

Use `*_new` tables from Marina:

> I’ve uploaded some new tables into Postgres, containing the updated data that Ed had sent me earlier.  Some tables are meant to replace existing ones, and have names ending with _new (e.g. larvae_counts_new, scrippscasts_new).  There are also tables called ships, cruises and nets.  The hierarchy is:
> 
> cruise -> station -> tow -> net -> eggs/larvae
> 
> The species_codes_new table has some entries missing in the taxon_rank column, I’ll fill them in ASAP.

Questions:
- [ ] How does `scrippscasts_new` fit into above scheme (cruise -> station -> tow -> net -> eggs/larvae)?

```{r}
# cruise -> station -> tow -> net -> eggs/larvae
d <- tbl(con, "cruises") %>%        # tbl_fields("cruises")
  # cruiseid, cruise, ship
  left_join(
    tbl(con, "stations_new"),       # tbl_fields("stations_new")
    # cruiseid: stationid, orderocc, latitude, longitude, line, station
    by = "cruiseid") %>% 
  left_join(
    tbl(con, "tows_new"),           # tbl_fields("tows_new")
    # stationid: towid, towtype, townumber, starttime
    by = "stationid") %>%
  left_join(
    tbl(con, "nets"),               # tbl_fields("nets")
    # towid: netid, netside, shf, volsampled, propsorted
    by = "towid") %>%
  left_join(
    tbl(con, "larvae_counts_new"),  # tbl_fields("larvae_counts_new")
    # netid: spccode, tally
    by = "netid") %>% 
  left_join(
    tbl(con, "species_codes_new"),  # tbl_fields("species_codes_new")
    # spccode: id, scientific_name, itis_tsn, common_name, taxon_rank
    by = "spccode") %>% 
  arrange(desc(starttime)) %>% 
  # show_query()
  collect()

# nrow(d)   # 398,142
d_l <- d %>% 
  filter(
    !is.na(tally),
    !is.na(latitude), !is.na(longitude),
    !is.na(starttime))
# nrow(d_l) # 365,213

# look at first 1,000 records
d_l %>% 
  head(1000) %>% 
  datatable()
```


## Run `get_nearest_bottle()` on one record

```{r}
# dbGetQuery(con, "SELECT MAX(datetime) FROM ctd_casts")
# 2020-01-26 01:51:16

d_l1 <- d_l %>% 
  filter(as.Date(starttime) < as.Date("2020-01-26")) %>% 
  slice(1)
# str(d_l1)
# tibble [1 × 25] (S3: tbl_df/tbl/data.frame)
#  $ cruiseid       : int 683
#  $ cruise         : chr "202001"
#  $ ship           : chr "RL"
#  $ stationid      : int 60417
#  $ orderocc       : int 103
#  $ latitude       : num 36.9
#  $ longitude      : num -125
#  $ line           : num 60
#  $ station        : num 80
#  $ towid          : int 75794
#  $ towtype        : chr "CB"
#  $ townumber      : int 1
#  $ starttime      : POSIXct[1:1], format: "2020-01-25 21:01:00"
#  $ netid          : int 88135
#  $ netside        : chr "S"
#  $ shf            : num 4.43
#  $ volsampled     : num 478
#  $ propsorted     : num 0.5
#  $ spccode        : int 288
#  $ tally          : int 1
#  $ id             : int 148
#  $ scientific_name: chr "Protomyctophum crockeri"
#  $ itis_tsn       : chr "162685"
#  $ common_name    : chr "California flashlightfish                    "
#  $ taxon_rank     : chr "Species"

# run with "most recent" larval data attributable in space and time
x <- get_nearest_bottle(
  fld       = "t_degc",
  datetime  = d_l1$starttime, 
  max_hrs   = 2,
  lon       = d_l1$longitude,
  lat       = d_l1$latitude, 
  max_kms   = 10,
  depth_min = 0,
  depth_max = 20,
  verbose   = F)
x
datatable(x)

# render temperature using simple mean
mean(x$t_degc)
# compare weighted means using seperation difference in space (kilometers)...
weighted.mean(x$t_degc, x$diff_kms)
# ... or time (hours)
weighted.mean(x$t_degc, x$diff_hrs)
```

## Run `get_nearest_bottle()` on many records using `purrr`

* [R for Data Science (2e) - 25  Hierarchical data](https://r4ds.hadley.nz/rectangling.html#introduction)

- R menu Help > Cheat Sheets > List manipulation with `purrr`

```{r}
librarian::shelf(
  purrr)

d10 <- d_l %>% 
  filter(as.Date(starttime) < as.Date("2020-01-26")) %>% 
  slice(1:10) %>% 
  mutate(
    temp_data = pmap(
      list(starttime, longitude, latitude),
      function(starttime, longitude, latitude, ...){
        get_nearest_bottle(
          fld       = "t_degc",
          datetime  = starttime, 
          max_hrs   = 2,
          lon       = longitude,
          lat       = latitude, 
          max_kms   = 10,
          depth_min = 0,
          depth_max = 20) } ),
    temp_nrow = map_int(temp_data, nrow),
    temp_avg  = map_dbl(
      temp_data, 
      function(x){
        y <- x %>% 
          filter(depthm == 0)
        weighted.mean(y$t_degc, y$diff_kms) } ) )
d10 %>% 
  select(1:7, towid, netid, starttime, temp_nrow, temp_avg) %>% 
  datatable()
```


## Show Tables

```{r}
# show all tables
dbListTables(con)
```

### Environmental Tables

`ctd_casts` -[`cast_count`]-> `ctd_bottles`

#### `ctd_casts`

```{r}
tbl_head("ctd_casts")
```

#### `ctd_bottles`

```{r}
tbl_head("ctd_bottles")
```

### Larval Tables

`cruises` -[`cruiseid`]-> `stations_new` -[`stationid`]-> `tows_new` -[`towid`]-> `nets` -[`netid`]-> `larvae_counts_new` -[`spccode`]-> `species_codes_new`

#### `cruises`

```{r}
tbl_head("cruises")
```

#### `stations_new`

```{r}
tbl_head("stations_new")
```

#### `tows_new`

```{r}
tbl_head("tows_new")
```

#### `nets`

```{r}
tbl_head("nets")
```

#### `larvae_counts_new`

```{r}
tbl_head("larvae_counts_new")
```

#### `species_codes_new`

```{r}
tbl_head("species_codes_new")
```


## OLD: Get Space & Time for Larval Data

```{r}
d_larvae <- tbl(
  con, "larvae_counts") %>%  # tbl_fields("larvae_counts")
  # cruise, ship, orderocc, towtype, townum, netloc: spccode, count
  left_join(
    tbl(con, "stations"),    # tbl_fields("stations")    
    # cruise, ship, orderocc: longitude, latitude, line, station, geom
    by = c("cruise", "ship", "orderocc")) %>% 
  left_join(
    tbl(con, "tows"),        # tbl_fields("tows")    
    # cruise, ship, orderocc, towtype, townum, netloc: shf, volsampled, percsorted, datetime, date
    by = c("cruise", "ship", "orderocc", "towtype", "townum", "netloc")) %>% 
  arrange(desc(date), desc(datetime)) %>% 
  # show_query() # uncomment to show SQL
  collect()
# nrow(d_larvae) # 397,496
d_larvae_miss <- d_larvae %>% 
  filter(is.na(date) | is.na(longitude)) 
d_larvae_miss %>% 
  write_csv("data/larvae-count_missing-location-or-date.csv")
nrow(d_larvae_miss) # 90,131 without attributable date or location

# show cruise-ship in larvae_counts missing stations with location
d_larvae_miss_stations <- tbl(
  con, "larvae_counts") %>%
  anti_join(
    tbl(con, "stations"),
    by = c("cruise", "ship", "orderocc")) %>% 
  group_by(cruise, ship) %>% 
  summarize(
    n_orderocc = n(),
    .groups = "drop") %>% 
  # show_query() # uncomment to show SQL
  collect()
d_larvae_miss_stations %>% 
  write_csv("data/larvae-count_cruise-ship_missing-in_stations.csv")
d_larvae_miss_stations %>% 
  datatable(
    caption = "The cruise and ship values from the larvae_counts table without a match in the stations table.")

# show cruise-ship in larvae_counts missing tows with datetime
d_larvae_miss_tows <- tbl(
  con, "larvae_counts") %>%
  anti_join(
    tbl(con, "tows"),        # tbl_fields("tows")    
    # cruise, ship, orderocc, towtype, townum, netloc: shf, volsampled, percsorted, datetime, date
    by = c("cruise", "ship", "orderocc", "towtype", "townum", "netloc")) %>% 
  group_by(cruise, ship) %>% 
  summarize(
    n_orderocc = n(),
    .groups = "drop") %>% 
  # show_query() # uncomment to show SQL
  collect()
d_larvae_miss_tows %>% 
  write_csv("data/larvae-count_cruise-ship_missing-in_tows.csv")
d_larvae_miss_tows %>% 
  datatable(
    caption = "The cruise and ship values from the larvae_counts table without a match in the tows table.")

# get larvae with valid space and time attribution
d_larvae_st <- d_larvae %>% 
  filter(
    !is.na(date),
    !is.na(longitude)) 
nrow(d_larvae_st) # 307,365 rows with valid space and time attribution
d_larvae_st1 <- d_larvae_st %>% 
  slice(1) %>% 
  select(-geom)
d_larvae_st1 %>% 
  datatable()
```

Yes, above is a bad merge. Compare lines 104-116 to below which returns 397,496 records, all with times and geoms.
BUT let's get my updated data structure on here so that we don't have to join by composite primary keys with six fields. I think that will make this much shorter and less error prone. 

```{r}
dbGetQuery(con,
   '
    select * from larvae_counts
    left join tows on larvae_counts.cruise=tows.cruise and
        larvae_counts.ship=tows.ship and
        larvae_counts.orderocc=tows.orderocc and
        larvae_counts.towtype=tows.towtype and
        larvae_counts.townum=tows.townum and
        larvae_counts.netloc=tows.netloc
    left join stations on tows.cruise=stations.cruise and
        tows.ship=stations.ship and
        tows.orderocc=stations.orderocc
    ')

# check
dbGetQuery(con,
   '
    select count(*) from
    (select * from larvae_counts
    left join tows on larvae_counts.cruise=tows.cruise and
        larvae_counts.ship=tows.ship and
        larvae_counts.orderocc=tows.orderocc and
        larvae_counts.towtype=tows.towtype and
        larvae_counts.townum=tows.townum and
        larvae_counts.netloc=tows.netloc
    left join stations on tows.cruise=stations.cruise and
        tows.ship=stations.ship and
        tows.orderocc=stations.orderocc
    where datetime=null or geom=null) A
    ')
```

So the most recent larval data with attributable space and time is:
- `r d_larvae_st1$datetime` (longitude: `r d_larvae_st1$longitude`; latitude: `r d_larvae_st1$latitude`)

We're clearly mismatching something since the most recent date for the ERDDAP subset is `2020-01-18T04:57:00Z`:

- [ERDDAP - CalCOFI Larvae Counts, Scientific Names A to AM - Data Access Form](https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIlrvcntAtoAM.html)

Here are some of the files output above highlighting larvae_count data without expected matches:

* [larvae-count\_cruise-ship\_missing-in\_stations.csv](https://github.com/CalCOFI/scripts/blob/main/data/larvae-count_cruise-ship_missing-in_stations.csv)
* [larvae-count\_cruise-ship\_missing-in\_tows.csv](https://github.com/CalCOFI/scripts/blob/main/data/larvae-count_cruise-ship_missing-in_tows.csv)
* [larvae-count\_missing-location-or-date.csv](https://github.com/CalCOFI/scripts/blob/main/data/larvae-count_missing-location-or-date.csv)

#### Get Space & Time for Larval Data with `*_new` Tables -- no results

```{r}
# try with *_new tables:
tbl(con, "larvae_counts") %>%   # tbl_fields("larvae_counts")
# cruise, ship, orderocc, towtype, townum, netloc: spccode, count
  left_join(
    tbl(con, "cruises"),        # tbl_fields("cruises")
    # cruise, ship: cruiseid
    by = c("cruise", "ship")) %>% 
  left_join(
    tbl(con, "stations_new"),   # tbl_fields("stations_new")
    # cruiseid, orderocc: stationid, latitude, longitude, line, station
    by = c("cruiseid", "orderocc")) %>% 
  left_join(
    tbl(con, "tows_new"),       # tbl_fields("tows_new")
    # stationid, towtype, townumber: towid, starttime
    by = c("stationid", "towtype", "townum" = "townumber")) %>%
  filter(!is.na(starttime), !is.na(longitude)) %>% 
  arrange(desc(starttime))
# yikes, no rows!
```




