---
title: "Summer CTD temperature anomalies, 1993–2026"
subtitle: "Worked example: pulling CalCOFI CTD temperature from R and differencing it against a 1993–2013 baseline"
author: "Ben Best (EcoQuants) for Rasmus Swalethorp (SIO)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    df-print: paged
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---

## Why this notebook

Rasmus asked, for CCIEA:

> I am hoping to try and connect to your database via R and extract all the
> summer cruise temperature data to generate some anomaly plots using **1993–2013
> as the baseline** (or 1998–2013 if you are still working on getting the older
> data ingested).

**1993–2013 is now available.** Until the August 2026 release, the 1 m-binned CTD
record jumped straight from **1998 to 2003** — calcofi.org publishes *Final*
1 m-binned products for 1998 and for 2003 onward, and only raw cast files before
that. Jim Wilkinson's archive filled the gap: 45 cruises covering **1993-08
through 2002** are now in the database, so a 1993–2013 baseline means what it
says instead of quietly meaning "1998, plus 2003–2013".

This notebook is meant to be **copied and edited**, not just read. Every step
runs against the public release with no credentials, and the transect,
climatology and anomaly functions are exported from `calcofi4r` so nothing here
is bespoke to this document.

## Setup

`cc_get_db()` opens a read-only DuckDB connection to the current public release
and caches it locally on first use. Nothing is downloaded twice.

```{r}
#| label: setup

librarian::shelf(
  calcofi4r, dplyr, DT, ggplot2, glue, purrr, scales, stringr, tidyr,
  quiet = TRUE)

# `version = "latest"` resolves the promoted release; pin a version string here
# to freeze an analysis against one snapshot.
con <- cc_get_db()

# which release "latest" resolved to, stated so a rerun is comparable
release <- readLines(
  "https://storage.googleapis.com/calcofi-db/ducklake/releases/latest.txt",
  warn = FALSE) |> trimws()

cat(glue("release: {release}\n",
         "tables : {length(DBI::dbListTables(con))}\n"))
```

### Two knobs, stated once

Everything downstream reads these. They are deliberately at the top rather than
buried in a call, because the choice of baseline is the single assumption a
reader of an anomaly plot most needs to see.

```{r}
#| label: knobs

BASELINE   <- c(1993, 2013)   # inclusive
SUMMER     <- 7:8             # CalCOFI's summer occupation runs Jul–Aug
LINE       <- 93.3            # the line off San Diego; every core line works
VAR        <- "temperature_ave"
DEPTH_MAX  <- 500
MIN_N      <- 3               # baseline cells thinner than this are dropped
```

## What "summer" covers

CalCOFI occupies the grid roughly quarterly. Before treating a month filter as
"summer", look at what is actually there — the pattern is not uniform across
three decades, and a couple of years have no summer occupation at all.

```{r}
#| label: cruise-inventory

cruises <- tbl(con, "sample") |>
  filter(dataset_key == "calcofi_ctd-cast", sample_type == "cast") |>
  mutate(yr = as.integer(strftime(datetime, "%Y")),
         mo = as.integer(strftime(datetime, "%m"))) |>
  count(cruise_key, yr, mo, data_stage, name = "n_casts") |>
  collect() |>
  arrange(desc(cruise_key))

summer <- cruises |> filter(mo %in% SUMMER)

cat(glue(
  "{nrow(cruises)} CTD cruises, {min(cruises$yr)}–{max(cruises$yr)}\n",
  "{nrow(summer)} of them in months {paste(SUMMER, collapse = '/')} ",
  "({length(unique(summer$yr))} distinct years)\n",
  "baseline window {BASELINE[1]}–{BASELINE[2]}: ",
  "{sum(summer$yr >= BASELINE[1] & summer$yr <= BASELINE[2])} summer cruises\n"))
```

```{r}
#| label: summer-gaps

yrs_all    <- seq(min(summer$yr), max(summer$yr))
yrs_missed <- setdiff(yrs_all, summer$yr)

cat(if (length(yrs_missed))
      glue("years with NO summer CTD occupation: ",
           "{paste(yrs_missed, collapse = ', ')}\n")
    else "every year in range has a summer occupation\n")
```

### Processing stage matters for the most recent cruises

The database records which CTD product each cruise came from, and the two
preliminary tiers are not interchangeable. A **`preliminary_without_bottle`**
cruise has not been through the bottle merge, so its bottle-corrected salinity,
oxygen and chlorophyll *do not exist* — but its **temperature is unaffected**,
because temperature is not bottle-corrected. That is what makes temperature the
right variable to run on the newest data.

```{r}
#| label: stage-table

summer |>
  count(data_stage, name = "n_cruises") |>
  arrange(desc(n_cruises)) |>
  datatable(
    rownames = FALSE,
    caption  = "Processing stage of summer CTD cruises",
    options  = list(dom = "t", pageLength = 10))
```

## The one-cruise case: a section and its anomaly

Start with a single line and cruise, since everything larger is this repeated.
Five `calcofi4r` calls, no SQL:

```{r}
#| label: one-section

cruise_recent <- summer |> filter(yr == max(yr)) |> slice(1) |> pull(cruise_key)

sta <- cc_transect_stations(con, LINE, cruise_recent)
sec <- cc_transect_section(
  con, LINE, cruise_recent, variables = VAR, depth_max = DEPTH_MAX)

clim <- cc_climatology(
  con, variables = VAR, years = BASELINE, depth_max = DEPTH_MAX, min_n = MIN_N)

anom <- cc_anomaly(sec, clim, sta)

cat(glue(
  "cruise {cruise_recent}: {nrow(sta)} stations, {nrow(sec)} binned values\n",
  "baseline: {nrow(clim)} (station, depth, month) cells\n",
  "{round(100 * mean(!is.na(anom$anomaly)))}% of this section has a baseline\n"))
```

```{r}
#| label: fig-section-anomaly
#| fig-height: 4.5
#| fig-cap: !expr glue("Temperature anomaly on line {LINE}, cruise {cruise_recent}, relative to the {BASELINE[1]}–{BASELINE[2]} mean for the same station, depth and calendar month.")

# The diverging scale is pinned so zero is the neutral colour. Left to autoscale,
# a uniformly warm section would paint its least-warm part blue.
lim <- max(abs(anom$anomaly), na.rm = TRUE)

ggplot(anom, aes(dist_km, depth_m, fill = anomaly)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_gradient2(
    low = "#0d366b", mid = "#f0efec", high = "#7d1b28",
    midpoint = 0, limits = c(-lim, lim), na.value = "transparent",
    name = "Δ °C") +
  scale_y_reverse(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = "Distance offshore (km)", y = "Depth (m)",
       title = glue("Line {LINE} · {cruise_recent} · temperature anomaly")) +
  theme_minimal(base_size = 11)
```

## All summer cruises: an upper-ocean anomaly time series

The headline CCIEA-style figure. Rather than sectioning every cruise, difference
**every summer observation** against the same baseline and average over the upper
ocean, so each year contributes one number per depth layer.

Two choices worth stating:

- **Average the anomalies, not the values.** Averaging temperatures first and
  differencing after would let a year that sampled more nearshore stations look
  cold purely from where it sampled. Differencing per station and depth removes
  the station's own climate before any averaging happens.
- **Only cells with a real baseline count.** A station newly added to the grid has
  no 1993–2013 history, and treating its absence as zero would drag every mean
  toward zero by an amount nobody can see.

```{r}
#| label: summer-anomaly-series

# every summer observation, binned exactly as the climatology is
obs_summer <- tbl(con, "obs") |>
  filter(
    dataset_key      == "calcofi_ctd-cast",
    measurement_type == VAR,
    !is.na(measurement_value), !is.na(depth_min_m), !is.na(datetime),
    depth_min_m      <= DEPTH_MAX) |>
  mutate(
    yr  = as.integer(strftime(datetime, "%Y")),
    mon = as.integer(strftime(datetime, "%m"))) |>
  filter(mon %in% SUMMER) |>
  mutate(depth_m = round(depth_min_m / 5) * 5) |>
  select(grid_key, cruise_key, yr, mon, depth_m,
         variable = measurement_type, value = measurement_value) |>
  collect()

anom_summer <- obs_summer |>
  inner_join(clim, by = c("grid_key", "mon" = "month", "depth_m", "variable")) |>
  mutate(anomaly = value - clim_mean)

cat(glue(
  "{nrow(obs_summer)} summer observations; ",
  "{nrow(anom_summer)} ({round(100*nrow(anom_summer)/nrow(obs_summer))}%) ",
  "have a {BASELINE[1]}–{BASELINE[2]} baseline\n"))
```

```{r}
#| label: fig-anomaly-series
#| fig-height: 4
#| fig-cap: !expr glue("Summer (month {paste(SUMMER, collapse='/')}) temperature anomaly by year and depth layer, relative to the {BASELINE[1]}–{BASELINE[2]} mean for each station, depth and calendar month. Grey band marks the baseline window.")

layers <- tribble(
  ~layer,          ~lo, ~hi,
  "0–50 m",     0,  50,
  "50–100 m",  50, 100,
  "100–200 m", 100, 200,
  "200–500 m", 200, 500)

series <- anom_summer |>
  mutate(layer = cut(depth_m, c(-1, 50, 100, 200, 500),
                     labels = layers$layer)) |>
  filter(!is.na(layer)) |>
  summarize(anomaly = mean(anomaly), n = n(), .by = c(yr, layer))

ggplot(series, aes(yr, anomaly, colour = layer)) +
  annotate("rect", xmin = BASELINE[1] - .5, xmax = BASELINE[2] + .5,
           ymin = -Inf, ymax = Inf, fill = "grey70", alpha = .16) +
  geom_hline(yintercept = 0, linewidth = .4, colour = "grey40") +
  geom_line(linewidth = .7) +
  geom_point(size = 1.6) +
  scale_colour_manual(values = c("#e34948", "#3987e5", "#256abf", "#0d366b"),
                      name = NULL) +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(x = NULL, y = "Temperature anomaly (°C)",
       title = glue("CalCOFI summer CTD temperature anomaly, ",
                    "{BASELINE[1]}–{BASELINE[2]} baseline")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r}
#| label: series-table

series |>
  mutate(anomaly = round(anomaly, 2)) |>
  pivot_wider(id_cols = yr, names_from = layer, values_from = anomaly) |>
  arrange(desc(yr)) |>
  datatable(
    rownames = FALSE,
    caption  = glue("Summer temperature anomaly (°C) by depth layer, ",
                    "{BASELINE[1]}–{BASELINE[2]} baseline"),
    options  = list(pageLength = 15, dom = "tip"))
```

## Reading it honestly

Four caveats that belong beside any use of these numbers.

**The baseline is 21 years, not a 30-year WMO normal.** CalCOFI's 1 m-binned CTD
record does not reach back far enough for one without splicing in ship
hydrocast data of a different vintage. 1993–2013 does span both the 1997–98
El Niño and the 1998–99 La Niña, and it ends before the 2014–16 marine heatwave —
so the heatwave and everything since read as departures rather than being folded
into the normal.

**Coverage is not constant.** Line 93.3 has not been sampled past station 90 since
January 2025, though 113 of the 130 cruises before that reached station 120. A
spatial mean over "the line" therefore covers different ocean in different years.
The per-station differencing above removes each station's own climate, which is
most of the problem — but a year that skipped the offshore stations still has no
offshore signal in it.

**Some cells have no baseline.** They are dropped, not zero-filled; the percentage
retained is printed above so you can see how much was dropped rather than
inferring it.

**Preliminary cruises are fine for temperature and not for the corrected series.**
Temperature is not bottle-corrected, so `preliminary_without_bottle` cruises carry
the same temperature they always will. Salinity, oxygen and chlorophyll on those
cruises are a different matter — the corrected forms do not exist yet, and the
uncorrected sensor series (`salinity_1`, `oxygen_ml_l_1`) are what is available.

**The two-sensor averages are recomputed, not taken as shipped.** The source CSVs
carry pre-computed averages — `TempAve`, `SaltAve_Corr`, `OxAve_StaCorr`,
`OxAveuM_StaCorr` — and where one of the two sensors is missing or has failed,
those sometimes fold the bad sensor in rather than falling back to the good one.
The obvious form averages in the `-99` missing marker: `(20.90 + −99)/2 = −39.05`.
The dangerous form does not: on 2607SH `Temp2` reads −32 to +33 °C while `Temp1`
sits at a steady ~20 °C, so `TempAve` comes out at **0.77 °C** — arithmetically
correct, inside any plausible range, and worth an 18 °C cold anomaly.

The ingest applies one rule to all four: **a sensor outside its declared physical
bounds is treated as absent, and the average is taken over whatever remains** —
one good sensor averages to itself, no good sensor gives null rather than a
number. The `-99` sentinel is filtered before anything is averaged. So the
numbers below are the repaired ones; reading the source CSVs directly gives you
the unrepaired values. See
[Q21](https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html#questions-for-data-providers).

## Doing this for another line, variable or baseline

Everything above is parameterised. To rerun for a different line, change `LINE`;
for a different baseline, change `BASELINE` — including to `c(1998, 2013)` if you
want to compare against what would have been possible before the backfill.

```{r}
#| label: sensitivity-1998

clim_98 <- cc_climatology(
  con, variables = VAR, years = c(1998, 2013),
  depth_max = DEPTH_MAX, min_n = MIN_N)

cmp <- anom_summer |>
  select(grid_key, mon, depth_m, variable, yr, value) |>
  inner_join(clim_98, by = c("grid_key", "mon" = "month", "depth_m", "variable"),
             suffix = c("", "_98")) |>
  mutate(anomaly_98 = value - clim_mean) |>
  filter(depth_m <= 50) |>
  summarize(anomaly_98 = mean(anomaly_98), .by = yr)

comparison <- series |>
  filter(layer == "0–50 m") |>
  select(yr, anomaly_9313 = anomaly) |>
  inner_join(cmp, by = "yr") |>
  mutate(difference = round(anomaly_9313 - anomaly_98, 3))

cat(glue(
  "0–50 m summer anomaly, 1993–2013 vs 1998–2013 baseline:\n",
  "  median absolute difference {round(median(abs(comparison$difference)), 3)} °C\n",
  "  maximum                    {round(max(abs(comparison$difference)), 3)} °C\n"))
```

```{r}
#| label: sensitivity-table

comparison |>
  mutate(across(c(anomaly_9313, anomaly_98), \(x) round(x, 2))) |>
  arrange(desc(yr)) |>
  datatable(
    rownames  = FALSE,
    colnames  = c("Year", "1993–2013", "1998–2013", "Difference"),
    caption   = "Sensitivity of the 0–50 m summer anomaly to the baseline window",
    options   = list(pageLength = 12, dom = "tip"))
```

## Where the definitions live

Nothing in this notebook is private to it:

| What | Where |
|---|---|
| `cc_get_db()`, `cc_transect_*()`, `cc_climatology()`, `cc_anomaly()` | [`calcofi4r/R/transect.R`](https://github.com/CalCOFI/calcofi4r/blob/main/R/transect.R) |
| The same climatology in SQL, for the static app | [`ctd-transects/scripts/build_sections.sql`](https://github.com/CalCOFI/ctd-transects/blob/main/scripts/build_sections.sql) |
| How the CTD data is ingested | [`ingest_calcofi_ctd-cast.qmd`](https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html) |
| Interactive sections and anomalies | [calcofi.io/ctd-transects](https://calcofi.io/ctd-transects/) |
| Table and column reference | [calcofi.io/schema](https://calcofi.io/schema/) |

```{r}
#| label: cleanup
#| include: false

DBI::dbDisconnect(con, shutdown = TRUE)
```
