---
title: "Depth Profile Plot Explore"
author: "Ben Best"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

## Packages & setup

```{r}
# packages
if (!require("librarian")){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  calcofi/calcofi4r, 
  # marmap,
  cmocean, dplyr, DT, glue, here, interp,
  mapview, plotly, proj4, purrr, sf, skimr, tidyr)
mapviewOptions(fgb = F)

source(here("../apps/libs/db.R"))
```

```{r}
calcofi4r::stations %>% 
  st_drop_geometry() %>% 
  skimr::skim()
nrow(calcofi4r::stations) # 2,634

mapview(calcofi4r::stations, zcol="sta_id_line")
```

## Weber & Moore (2013)

Source: Appendix of Weber & Moore (2013) [Corrected conversion algorithms for the CalCOFI station grid and their implementation in several computer languages](http://calcofi.org/publications/calcofireports/v54/Vol_54_Weber.pdf). _California Cooperative Oceanic Fisheries Investigations Reports_

```{r}
deg2rad <- function(deg) deg * pi / 180
rad2deg <- function(rad) rad * 180 / pi
inverse_mercator <- function(mercatorlat, iterations = 3){ 
  
  approxlat <- mercatorlat
  
  iterlatitude <- function(mercatorlat, approxlat){
    approxlat <- 2 * (atan(exp(deg2rad(mercatorlat) + 0.00676866 * sin(deg2rad(approxlat)))) * 180 / pi - 45)
    approxlat
  }
  
  for (i in 1:iterations)
    approxlat <  iterlatitude(mercatorlat, approxlat)
  
  approxlat
}

to_mercator <- function(latitude) { 
  y <- rad2deg(log(tan(deg2rad(45 + latitude / 2))) - 0.00676866 * sin(deg2rad(latitude)))
  y
}

station_to_latlon <- function(x, roundlines = true) {
  
  if (length(x) == 2 & class(x) != 'matrix'){
    x <- matrix(x, 1, 2)
  }
  line <- x[, 1]
  station <- x[, 2]
  
  reflatitude <- 34.15 - 0.2 * (line - 80) * cos(deg2rad(30))
  latitude <- reflatitude - (station - 60) * sin(deg2rad(30)) / 15
  l1 <- (to_mercator(latitude) - to_mercator(34.15)) * tan(deg2rad(30))
  l2 <- (to_mercator(reflatitude) - to_mercator(latitude)) / (cos(deg2rad(30)) * sin(deg2rad(30)))
  longitude <- -1 * (l1 + l2 + 121.15)
  cbind(lon = longitude, lat = latitude)
}
# https://proj.org/operations/projections/calcofi.html # -121.15 34.15   80.00   60.00
station_to_latlon(c(80.0, 60.0))
#          lon   lat
# [1,] -121.15 34.15

lonlat_to_station <- function(x){
  # x = c(-121.15, 34.15)
  if (length(x) == 2 & class(x) != 'matrix'){
    x <- matrix(x, 1, 2)
  }
  longitude <- x[, 1]
  latitude  <- x[, 2]
  # longitude <- -121.15 
  # latitude  <- 34.15
  
  # assume we're in the western hemispere
  longitude[longitude > 180] <- -1 * (longitude[longitude > 180] - 360)
  longitude[longitude < 0]   <- longitude[longitude < 0] * -1
  
  l1 <- (to_mercator(latitude) - to_mercator(34.15)) * tan(deg2rad(30))
  l2 <- longitude - l1 - 121.15
  mercreflatitude <- l2 * cos(deg2rad(30)) * sin(deg2rad(30)) + to_mercator(latitude)
  reflatitude     <- inverse_mercator(mercreflatitude)
  
  line    <- 80 - (reflatitude - 34.15) * 5 / cos(deg2rad(30))
  station <- 60 + (reflatitude - latitude) * 15 / sin(deg2rad(30))
  
  cbind(line = line, station = station)
}
# https://proj.org/operations/projections/calcofi.html # 80.0 60.0   -121.15 34.15
lonlat_to_station(c(-121.15, 34.15))
# 68.42567 120.142 # Doh! Wrong answer
```

## `proj-bin`

For Debian Linux, install `proj` binary:

```{bash}
sudo apt-get install proj-bin
which proj
```

## f() with proj-bin

Define functions:

```{r}
lonlat_to_stationid <- function(lon, lat){
  # convert station ID to lon, lat using the proj library
  
  system(glue("echo {lon} {lat} | proj +proj=calcofi +epsg=4326 -f '%05.1f'"), intern=T) %>%
    stringr::str_replace("\t", " ")
}
lonlat_to_stationid(-121.15, 34.15) # "080.0 060.0"

stationid_to_lonlat <- function(stationid){
  # using 5th decimal place, a la CCE_Stations.txt
  
  system(glue("echo {stationid} | proj +proj=calcofi +epsg=4326  -I -d 5"), intern=T) %>%
    stringr::str_replace("\t", " ")
}
stationid_to_lonlat("080.0 060.0") # "-121.15000 34.15000"

# A: offshore/1, line/1
sta_a.1.1 <- calcofi4r::stations %>% 
  st_drop_geometry() %>% 
  mutate(
    x = round(sta_id_station),
    y = round(sta_id_line)) %>% 
  group_by(x, y) %>% 
  summarize(
    n = n(),
    .groups = "drop") %>% 
  mutate(
    z = glue("{y} {x}")) %>%
  mutate(
    lonlat = map_chr(z, stationid_to_lonlat)) %>% 
  separate(lonlat, c("lon", "lat"), sep = " ", convert=T) %>%
  st_as_sf(
    coords = c("lon", "lat"), crs=4326, remove = F) %>% 
  rename(
    n_stations = n,
    sta_id = z)
nrow(sta_a.1.1) # 2,270
mapview(sta_a.1.1, zcol="y")

# B: offshore/5, line/2
sta_b.5.2 <- calcofi4r::stations %>% 
  st_drop_geometry() %>% 
  mutate(
    x = round(sta_id_station/5) * 5,
    y = round(sta_id_line/2) * 2) %>% 
  group_by(x, y) %>% 
  summarize(
    n = n(),
    .groups = "drop") %>% 
  mutate(
    z = glue("{y} {x}")) %>%
  mutate(
    lonlat = map_chr(z, stationid_to_lonlat)) %>% 
  separate(lonlat, c("lon", "lat"), sep = " ", convert=T) %>%
  st_as_sf(
    coords = c("lon", "lat"), crs=4326, remove = F) %>% 
  rename(
    n_stations = n,
    sta_id = z)
nrow(sta_b.5.2) # 1,128
mapview(sta_b.5.2, zcol="y")

# C: offshore/10, line/3
sta_c.10.3 <- calcofi4r::stations %>% 
  st_drop_geometry() %>% 
  mutate(
    x = round(sta_id_station/10) * 10,
    y = round(sta_id_line/3) * 3) %>% 
  group_by(x, y) %>% 
  summarize(
    n = n(),
    .groups = "drop") %>% 
  mutate(
    z = glue("{y} {x}")) %>%
  mutate(
    lonlat = map_chr(z, stationid_to_lonlat)) %>% 
  separate(lonlat, c("lon", "lat"), sep = " ", convert=T) %>%
  st_as_sf(
    coords = c("lon", "lat"), crs=4326, remove = F) %>% 
  rename(
    # sta_offshore = x,
    # sta_alongshore = y,
    n_stations = n,
    sta_id = z)
nrow(sta_c.10.3) # 702
mapview(sta_c.10.3, zcol="y")

# D: offshore/10, line/4
sta_d.10.4 <- calcofi4r::stations %>% 
  st_drop_geometry() %>% 
  mutate(
    x = round(sta_id_station/10) * 10,
    y = round(sta_id_line/4) * 4) %>% 
  group_by(x, y) %>% 
  summarize(
    n = n(),
    .groups = "drop") %>% 
  mutate(
    z = glue("{y} {x}")) %>%
  mutate(
    lonlat = map_chr(z, stationid_to_lonlat)) %>% 
  separate(lonlat, c("lon", "lat"), sep = " ", convert=T) %>%
  st_as_sf(
    coords = c("lon", "lat"), crs=4326, remove = F) %>% 
  rename(
    # sta_offshore = x,
    # sta_alongshore = y,
    n_stations = n,
    sta_id = z)
nrow(sta_d.10.4) # 611
mapview(sta_d.10.4, zcol="y")

hull <- st_convex_hull(st_union(calcofi4r::stations)) %>% 
  st_transform(leaflet:::epsg3857) %>% 
  st_buffer(1000)

sta_d.10.4_v <- sta_d.10.4 %>% 
  st_transform(leaflet:::epsg3857) %>% 
  st_union() %>% 
  st_voronoi(hull)

sta_d.10.4_vp <- sta_d.10.4_v %>% 
  st_collection_extract(
  type = "POLYGON")
sta_d.10.4_vpi <- st_intersection(sta_d.10.4_vp, hull)
sta_d.10.4_vpig <- sta_d.10.4_vpi %>% 
  st_transform(4326)

mapview(sta_d.10.4_vpi) +
  mapview(sta_d.10.4, zcol="y")

librarian::shelf(
  rnaturalearth, rnaturalearthdata)
# devtools::install_github("ropensci/rnaturalearthhires")

land_l <- ne_countries(scale = "large", returnclass = "sf") %>% 
  st_combine() %>% 
  st_make_valid()
# land_m <- ne_countries(scale = "medium", returnclass = "sf")
# land_s <- ne_countries(scale = "small", returnclass = "sf")

# sta_d.10.4_vpigl_0 <- sta_d.10.4_vpigl
sta_d.10.4_vpigl <- st_difference(sta_d.10.4_vpig, land_l)
mapview(sta_d.10.4_vpigl, alpha.regions = 0.5) +
  mapview(sta_d.10.4, zcol="y", cex=2)
```

## NEW: `sf::st_crs("+proj=calcofi")`

```{r}
st_crs("+proj=calcofi")

# 80.00 60.00 -> -121.15 34.15   
pt_cc <- tribble(
  ~alongshore, ~offshore,
  80.00, 60.00) %>% 
  st_as_sf(
    coords = c("alongshore", "offshore"), 
    crs = st_crs("+proj=calcofi"))
pt_cc # 80 60

pt_gcs <- pt_cc %>% 
  st_transform(4326) # -121.15 34.15

pt_gcs %>% 
  st_transform(st_crs("+proj=calcofi")) # 80, 60

# freq of alongshore
y <- stations %>% 
  group_by(sta_id_line) %>% 
  summarize(n = n())

y %>% st_drop_geometry() %>% datatable()
  

y %>% st_drop_geometry() %>% arrange(desc(n)) %>% head()
# 1        90      82
# 2        93.3    65
# 3       100      53
# 4       120      53
# 5        83.3    47
# 6       110      47

dbListFields(con, "ctd_casts")

lns <- tbl(con, "ctd_casts") %>% 
  group_by(stline) %>% 
  summarize(cnt = n()) %>% 
  arrange(desc(cnt)) %>% 
  collect()
View(lns)
lns10 <- head(lns, n=10)

lns10
#  1   90      4222
#  2   93.3    3611
#  3   86.7    2609
#  4   83.3    2387
#  5   80      2233
#  6   76.7    1629
#  7  100      1243
#  8  120      1216
#  9  110      1084
# 10   70      1025
lns10 <- lns10 %>% pull(stline)


seq(90, 93.3, by=0.3)
lnsv <- seq(min(lns10), max(lns10), by=0.3)
sum(lns10 %in% lnsv) # 2

lnsv2 <- c(
  seq(90, max(lns10), by=0.3),
  seq(90, min(lns10), by=-0.3)[-1]) %>% 
  sort()
sum(lns10 %in% lnsv2) # 4

hist(stations$sta_id_line)

sta <- tbl(con, "ctd_casts") %>% 
  group_by(ststa) %>% 
  summarize(cnt = n()) %>% 
  arrange(desc(cnt)) %>% 
  collect() # %>% 
  # head(n=10)

View(sta)

sta_odd <- sta %>% 
  mutate(
    mod5 = ststa %% 5) %>% 
  filter(mod5 != 0)

# ∆ 0.3, 10 (alongshore, offshore); ctr: 90, 60

sta_cc <- stations %>% 
  st_transform(st_crs("+proj=calcofi"))
yx <- st_coordinates(sta_cc)

# alongshore
min(yx[,1])
max(yx[,1])
# offshore
min(yx[,2])
max(yx[,2])

# ∆ 3.3, 10 (alongshore, offshore); ctr: 90, 60
cy <- 90
cx <- 60
dy <- 3.3
dx <- 10
vy <- c(
  seq(cy, max(yx[,1]), by = dy),
  seq(cy, min(yx[,1]), by = -1*dy)[-1]) %>% 
  sort()
vx <- c(
  seq(cx, max(yx[,2]), by = dx),
  seq(cx, min(yx[,2]), by = -1*dx)[-1]) %>% 
  sort()


g <- expand_grid(
  y = vy,
  x = vx) %>% 
  mutate(
    geom = map2(x, y, function(x, y){
      dx2 <- dx/2
      dy2 <- dy/2
      
      # y = 90; x = 60
      p <- tribble(
        ~y, ~x, 
        y - dy2, x - dx2,      # lower left
        y + dy2, x + dx2) %>%  # upper right
        st_as_sf(
          coords = c("y", "x"), 
          crs    = st_crs("+proj=calcofi")) %>% 
        st_bbox() %>% 
        st_as_sfc() %>% 
        st_transform(4326) %>% 
        .[[1]]
    })) %>% 
  st_as_sf()
    
# g_0 <- g 
mapview(g)

shelf(leaflet)

leaflet() %>% 
  addProviderTiles(providers$Esri.OceanBasemap) %>% 
  addPolygons(data=g)

gc <- g %>% 
  filter(
    y <= cy + dy*5,
    y >= cy - dy*5,
    x <= cx + dx*5,
    x >= cx - dx*5)

leaflet() %>% 
  addProviderTiles(providers$Esri.OceanBasemap) %>% 
  addPolygons(data = gc)


r <- terra::rast(
  xmin = min(vy), xmax = max(vy), 
  ymin = min(vx), ymax = max(vx), 
  ncols = length(vy), nrows = length(vx),
  nlyrs = 1, vals = NA,
  crs = "+proj=calcofi")
```
"

### init `postgis_raster` w/ `proj=calcofi` SRID 880001

```{r}
shelf(rgdal, rpostgis)

q("CREATE EXTENSION postgis_raster;")

crs_cc <- sp::CRS("+proj=calcofi")
pgSRID(con, crs_cc, create.srid = TRUE)
# No matches were found in spatial_ref_sys. New SRID created (880001)

pgPostGIS(con)
pgListGeom(con, geog = TRUE)
pgListRast(con)
```

### `test_rast`

```{r}
q("CREATE TABLE test_rast(rid serial primary key, rast raster);")
q("
INSERT INTO test_rast(rid, rast) 
  VALUES(1, ST_MakeEmptyRaster(100, 100, 0.0005, 0.0005, 1, 1, 0, 0, 4326) );
")
# Add another band of type 8 bit unsigned integer with pixels initialized to 200
q(" 
UPDATE test_rast
  SET rast = ST_AddBand(rast,'8BUI'::text, 200)
  WHERE rid = 1;")
# once done populating table, create spatial index on raster column
q("CREATE INDEX test_rast_convexhull_idx ON test_rast USING gist( ST_ConvexHull(rast) );")

pgListRast(con)

r <- pgGetRast(con, "test_rast")
r
```

### `grd_a` in calcofi projection

```{r}
sta_cc <- stations %>% 
  st_transform(st_crs("+proj=calcofi"))
yx <- st_coordinates(sta_cc)

# alongshore
min(yx[,1])
max(yx[,1])
# offshore
min(yx[,2])
max(yx[,2])

# ∆ 3.3, 10 (alongshore, offshore); ctr: 90, 60
cy <- 90
cx <- 60
dy <- 3.3
dx <- 10
vy <- c(
  seq(cy, max(yx[,1]), by = dy),
  seq(cy, min(yx[,1]), by = -1*dy)[-1]) %>% 
  sort()
vx <- c(
  seq(cx, max(yx[,2]), by = dx),
  seq(cx, min(yx[,2]), by = -1*dx)[-1]) %>% 
  sort()

max(vx)

# ST_MakeEmptyRaster(integer width, integer height, float8 upperleftx, float8 upperlefty, float8 scalex, float8 scaley, float8 skewx, float8 skewy, integer srid=unknown)
# ST_MakeEmptyRaster(3.3 width, 10 height, {max(vx)} upperleftx, {max(vy)} upperlefty, 1, 1, 0, 0, 880001)

q("DROP TABLE IF EXISTS grd_a;")
q("CREATE TABLE grd_a(rid serial primary key, rast raster);")
q(glue("
INSERT INTO grd_a(rid, rast) 
  VALUES(
    1, 
    ST_MakeEmptyRaster(
      width:={length(vy)}, 
      height:={length(vx)}, 
      upperleftx:={min(vy)}, 
      upperlefty:={max(vx)}, 
      scalex:=dy, 
      scaley:=dx, 
      skewx:=0, 
      skewy:=0, 
      srid:=880001) );
"))
# Add another band of type 8 bit unsigned integer with pixels initialized to 200
q(" 
UPDATE grd_a
  SET rast = ST_AddBand(rast,'8BUI'::text, 0)
  WHERE rid = 1;")

# once done populating table, create spatial index on raster column
q("CREATE INDEX test_rast_convexhull_idx ON test_rast USING gist( ST_ConvexHull(rast) );")

pgListRast(con)

r_cc <- pgGetRast(con, "grd_a")
r_cc

ext_gcs <- raster::projectExtent(r_cc, crs = sp::CRS(leaflet:::epsg4326))
ext_gcs

mapview(st_bbox(ext_gcs))
ext_mer <- raster::projectExtent(r_cc, crs = sp::CRS(leaflet:::epsg3857))

raster::projectRaster(4326)



r_gcs <- r_cc %>% 
  terra::project(4326)

shelf(leaflet)

r_mer <-leaflet::projectRasterForLeaflet(r_cc)

  st_transform(leaflet:::epsg3857)
```



### `grd_mer` in web mercator projection

```{r}
crs_mer <- 3857
sta_mer <- stations %>% 
  st_transform(crs_mer)
xy <- st_coordinates(sta_mer)
apply(xy, 2, range)
#              X       Y
# [1,] -17050244 1968402
# [2,] -11925222 6097048

# 1 km square grid 
# grd = "grd_mer"; rid = 1
# dx <- 1000
# dy <- 1000

# 100 km square grid 
grd = "r_mer100km"; rid = 1
dx <- 1000*100
dy <- 1000*100

# passing through ctr: -13440906, 3736349
cx <- mean(xy[,1]) # -13440906
cy <- mean(xy[,2]) #   3736349
vx <- c(
  seq(cx, max(xy[,1]), by = dy),
  seq(cx, min(xy[,1]), by = -1*dy)[-1]) %>% 
  sort()
vy <- c(
  seq(cy, max(xy[,2]), by = dx),
  seq(cy, min(xy[,2]), by = -1*dx)[-1]) %>% 
  sort()
crs = crs_mer
# ST_MakeEmptyRaster(integer width, integer height, float8 upperleftx, float8 upperlefty, float8 scalex, float8 scaley, float8 skewx, float8 skewy, integer srid=unknown)
# ST_MakeEmptyRaster(3.3 width, 10 height, {max(vx)} upperleftx, {max(vy)} upperlefty, 1, 1, 0, 0, 880001)
q(glue("DROP TABLE IF EXISTS {grd};"))
q(glue("CREATE TABLE {grd}(rid serial primary key, rast raster);"))
q(glue("
INSERT INTO {grd} (rid, rast) 
  VALUES(
    {rid}, 
    ST_MakeEmptyRaster(
      width:={length(vx)}, 
      height:={length(vy)}, 
      upperleftx:={min(vx)}, 
      upperlefty:={max(vy)}, 
      scalex:={dx}, 
      scaley:={dy}, 
      skewx:=0, 
      skewy:=0, 
      srid:={crs}) );
"))
# Add another band of type 8 bit unsigned integer with pixels initialized to 200
q(glue(" 
UPDATE {grd}
  SET rast = ST_AddBand(rast,'8BUI'::text, 0)
  WHERE rid = {rid};"))

# once done populating table, create spatial index on raster column
q(glue("CREATE INDEX {grd}_convexhull_idx ON {grd} USING gist( ST_ConvexHull(rast) );"))

pgListRast(con)

shelf(raster)
r <- pgGetRast(con, grd)
v <- values(r)
v <- c(1, v[2:(length(v) - 1)], 10)
values(r) <- v
mapview(r)
raster::crs(r_mer)
# Deprecated Proj.4 representation:
#  +proj=merc +a=6378137 +b=6378137 +lat_ts=0 +lon_0=0 +x_0=0 +y_0=0 +k=1 +units=m
#  +nadgrids=@null +wktext +no_defs 
sp::CRS(leaflet:::epsg3857)
# Deprecated Proj.4 representation:
#  +proj=merc +a=6378137 +b=6378137 +lat_ts=0 +lon_0=0 +x_0=0
#  +y_0=0 +k=1 +units=m +nadgrids=@null +wktext +no_defs
r_cc

ext_gcs <- raster::projectExtent(r_cc, crs = sp::CRS(leaflet:::epsg4326))
ext_gcs

mapview(st_bbox(ext_gcs))
ext_mer <- raster::projectExtent(r_cc, crs = sp::CRS(leaflet:::epsg3857))

raster::projectRaster(4326)



r_gcs <- r_cc %>% 
  terra::project(4326)

shelf(leaflet)

r_mer <-leaflet::projectRasterForLeaflet(r_cc)

  st_transform(leaflet:::epsg3857)
```


## rasters in R 

```{r}
# gcs
b <- st_bbox(stations)
r_gcs <- raster(
  xmn = b$xmin, xmx = b$xmax, 
  ymn = b$ymin, ymx = b$ymax,
  resolution = c(1,1),
  crs = 4326)
r_sta_gcs <- rasterize(stations, r_gcs, fun="count")

mapview(r_sta_gcs)
```

## make `grd_mer_*km`

```{r}
librarian::shelf(
  calcofi/calcofi4r,
  glue, raster, sf, st)

resolutions_km = c(5, 20, 50)

stations_mer <- stations %>% 
  st_transform(3857) # web mercator 
b <- stations_mer %>% 
  st_bbox()

for (res_km in resolutions_km){ # res_km = resolutions_km[2]
  r_tmpl <- raster(
    xmn = b$xmin, xmx = b$xmax, 
    ymn = b$ymin, ymx = b$ymax,
    resolution = res_km * 1000,
    crs = 3857)
  r_sta <- rasterize(stations_mer, field = 1, r_tmpl, fun="count")
  # plot(r_sta)
  writeRaster(r_sta, glue("data/r_sta-cnt_{res_km}km-mer.tif"), overwrite = T)
}
```


## cluster marker

```{r}
shelf(leaflet, raster)
leaflet() %>% 
  addProviderTiles(providers$Esri.OceanBasemap, group="Ocean") %>% 
  addProviderTiles(providers$Stamen.Toner, group="B&W") %>% 
  #addRasterImage(r_1km  , project=F, group="1km") %>% 
  addMarkers(
    data  = stations,
    popup =~as.character(sta_id),
    group = "stations",
    clusterOptions = markerClusterOptions(
      removeOutsideVisibleBounds = F),
    labelOptions = labelOptions(
      noHide = F,
      direction = 'auto'))

stations_crs <- stations %>% 
  st_transform(crs)
b <- stations_crs %>% 
  st_bbox()
r <- raster(
  xmn = b$xmin, xmx = b$xmax, 
  ymn = b$ymin, ymx = b$ymax,
  resolution = c(dx, dy),
  crs = crs)
r_sta <- rasterize(stations_crs, r, fun="count")

plot(raster(r_sta, "ID"))
leaflet() %>% 
  addProviderTiles(providers$Esri.OceanBasemap) %>% 
  addRasterImage(raster(r_sta, "ID"), project = F)
#mapview(r_sta)
```

```{r}
shelf(rpostgis)

system.time({
  pgWriteRast(con, "r_sta_cnt", r_sta, overwrite = T)
}) # 12.219 secs

system.time({
  writeRaster(r_sta, "data/r_sta_cnt.tif", overwrite = T)
}) # 0.033 secs # 9.7 KB

system.time({
  r_db <- pgGetRast(con, "r_sta_cnt")
}) # 9.446 secs

d <- dbGetQuery(con, "SELECT * FROM r_sta_cnt")
names(d)
d$rid

r_db_tiff <- dbGetQuery(con, "
  SELECT ST_AsTIFF(ST_Union(rast), 'LZW') as tiff
  FROM r_sta_cnt
  WHERE rid=1")
# Error: Failed to fetch row: 
#   ERROR:  
#     rt_raster_to_gdal: Could not load the output GDAL driver
#   CONTEXT:
#     PL/pgSQL function st_astiff(raster,text[],integer)

# try raw
r_db <- dbGetQuery(con, "
  SELECT rast
  FROM r_sta_cnt
  WHERE rid=1")
r_db_x <- raster(r_db$rast)
plot(r_db$rast)

# try using raw code from rpostgis::pgGetRast()
info <- dbGetQuery(conn, glue("
  select 
    st_xmax(st_envelope(rast)) as xmx,\n              st_xmin(st_envelope(rast)) as xmn,\n              st_ymax(st_envelope(rast)) as ymx,\n              st_ymin(st_envelope(rast)) as ymn,\n              st_width(rast) as cols,\n              st_height(rast) as rows\n              from\n              (select st_union(", 
                aq[1], rastque, ",", aq[2], aq[3], b, ") rast from ", 
                nameque, " ", clauses, ") as a;"))
            vals <- dbGetQuery(conn, paste0("select\n            unnest(st_dumpvalues(rast, 1)) as vals \n            from\n            (select st_union(", 
                aq[1], rastque, ",", aq[2], aq[3], b, ") rast from ", 
                nameque, " ", clauses, ") as a;"))$vals
            rout <- raster::raster(nrows = info$rows, ncols = info$cols, 
                xmn = info$xmn, xmx = info$xmx, ymn = info$ymn, 
                ymx = info$ymx, crs = sp::CRS(p4s), val = vals)

leaflet() %>% 
  addProviderTiles(providers$Esri.OceanBasemap, group="Ocean") %>% 
  addProviderTiles(providers$Stamen.Toner, group="B&W") %>% 
  #addRasterImage(r_1km  , project=F, group="1km") %>% 
  addCircleMarkers(
    data  = stations,
    radius = 4,
    popup =~as.character(sta_id),
    group = "stations")

```

