---
title: "explore_duckdb"
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

Goals:

-   [ ] Write database tables to duckdb

-   [ ] Upload to S3 bucket publicly available online

-   [ ] View tables

```{r}
librarian::shelf(
  calcofi/calcofi4db, 
  DBI, dplyr, DT, duckdb, fs, glue, here, tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F)

# set up paths
is_server <-  Sys.info()[["sysname"]] == "Linux"
dir_data  <- ifelse(
  is_server,
  "/share/data",
  "~/My Drive/projects/calcofi/data")
dk_db     <- glue("{dir_data}/swfsc.noaa.gov/calcofi.duckdb")

# get database connection
con_dk <- dbConnect(duckdb(), dbdir = dk_db, read_only = F)
# dbListTables(con_dk)
```

## Load Data from PostgreSQL

```{r pg_to_duckdb}
#| eval: false

con_pg <- get_db_con(c("dev"))

tbls = dbListTables(con_pg)
#  [1] "cruise"   "egg"      "grid"     "larva"    "net"      "ship"     "site"     "site_seg"
#  [9] "species"  "tow"      "tow_type"

for (tbl in tbls){
  if (!(dbExistsTable(con_dk, tbl)))
    dbReadTable(con_pg, tbl) %>% 
      dbWriteTable(con_dk, tbl, .)
}
```

## Add H3 reference

-   [h3 – DuckDB Community Extensions](https://duckdb.org/community_extensions/extensions/h3.html)
-   [Tables of Cell Statistics Across Resolutions \| H3](https://h3geo.org/docs/core-library/restable/)
-   [Terminology](https://h3geo.org/docs/library/terminology/)
    -   `H3Index`: an unsigned 64-bit integer representing any H3 object (hexagon, pentagon, directed edge ...); often represented as a 15-character (or 16-character) hexadecimal string, like '8928308280fffff'
-   h3 res 9: sqrt(105332.513): \~324.5 m width
-   h3 res 10: sqrt( 15047.502): \~122.7 m width

```{r}
dbExecute(con_dk, "INSTALL h3 FROM community; LOAD h3;")
# test: dbGetQuery(con_dk, "SELECT to_hex(h3_latlng_to_cell(37.7887987, -122.3931578, 10)) AS h3_res10;") # 4

for (res in 1:10){
  fld <- glue("hex_h3_res{res}")
  
  dbExecute(con_dk, glue(
    "ALTER TABLE site ADD COLUMN IF NOT EXISTS {fld} BIGINT;"))
  
  dbExecute(con_dk, glue(
    "UPDATE site SET {fld} = h3_latlng_to_cell(latitude, longitude, {res});"))
}

# tbl(con_dk, "site")

# update h3_res10 column with h3 index
# 61,220 rows affected
```

## Close Connection

```{r cleanup}
dbDisconnect(con_dk, shutdown = T); duckdb_shutdown(duckdb()); rm(con_dk)
```
