---
title: "Ingest CDFW Dungeness Crab Megalopae"
calcofi:
  # underscores: target_name must be a valid R symbol, unlike the filename
  target_name: ingest_cdfw_dungeness_crab
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/cdfw_dungeness-crab/manifest.json
  provider: cdfw
  dataset: dungeness-crab
  # HELD OUT OF THE RELEASE. This ingest runs in the pipeline and writes its full
  # parquet outputs — including its slice of the consolidated core — but every
  # release-side discovery step in release_database.qmd skips it
  # (calcofi4db::release_excluded_datasets()). The ONLY thing still holding it out
  # is Q08: no licence, no citation, and permission to publish unconfirmed. Flip to
  # `true` — or delete the key, since the flag is opt-out — once Q08 is answered,
  # and move measurement_type_new.csv into the shared registry in the same change.
  in_release: false
  workflow_url: https://calcofi.io/workflows/ingest_cdfw_dungeness-crab.html
  questions_file: metadata/cdfw/dungeness-crab/questions.csv
  dataset_meta:
    dataset_name: CDFW Dungeness Crab Megalopae
    description: >
      Dungeness crab (Metacarcinus magister) megalopae and larval decapods
      picked from archived CalCOFI oblique-tow plankton samples, examined for the
      California Department of Fish and Wildlife. Two efforts: a 2008-2014 time
      series of 310 sorted samples (A. Klemmedson, 2015) yielding 24 M. magister
      megalopae from 14 samples, and a 1949-2009 master sorting log of 2,011
      archived samples (E. Jones, as of 2012-02-15) recording which had been
      examined. Spring cruises, CalBOBL oblique tows, lines < 80 / stations < 90.
      Counts are occurrence in the examined aliquot, not standardized density.
      Sorting was performed at the SIO Pelagic Invertebrate Collection, where the
      removed megalopae are archived in vials.
    citation_main: ""
    link_calcofi_org: ""
    link_data_source: ""
    link_others: []
    coverage_temporal: 1949 to 2014
    coverage_spatial: "California Current, San Francisco north to Washington"
    license: ""
    pi_names: Laura Rogers-Bennett
  # publishes the consolidated core AND NOTHING ELSE. The source shape is wrangled
  # in the notebook and projected into sample / obs / obs_attribute /
  # sample_measurement + the taxa refs via the generic append_* helpers. No source
  # table is written to parquet: the workbooks are archived, so a per-dataset copy
  # is redundant and would give consumers two representations to choose between.
  # dungeness_* survive as compat VIEWs inside the notebook only.
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (310 examined subsamples + 2,011 sorting-log tows)"}
    - {table: obs, shared: true, note: "core occurrence headline: counts x 4 resolved taxa, plus 216 zero-valued absences from the sorting log"}
    - {table: obs_attribute, shared: true, note: "per-individual M. magister carapace lengths"}
    - {table: sample_measurement, shared: true, note: "event-level effort (prop_sorted, settled volume)"}
    - {table: taxon, shared: true, note: "shared taxa reference"}
    - {table: dataset_taxon, shared: true, note: "raw count label -> taxon_key crosswalk"}
  erd:
    color: "#f5d5b8"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: three hand-maintained Excel workbooks plus correspondence, under
`data-public/cdfw/dungeness-crab/`. Provider `cdfw` (California Department of Fish
and Wildlife), which commissioned the work; the sorting itself was done at the
SIO Pelagic Invertebrate Collection.

| Workbook | Becomes | Rows |
|---|---|---|
| `Dungeness Time Series - Angela Klemmedson.xlsx` (`Compiled Cruises`) | `dungeness_sample` + `dungeness_measurement` + `dungeness_carapace` | 310 samples |
| `Cancer Magister sorting update as of 02-15-12_EJones.xlsx` | `dungeness_sorting_log` | 2,011 samples |
| `ScrippsArchivedMegalopae_sent_5_10_2012.xlsx` | `dungeness_specimen` | 5 specimens |

The two PDFs and the `.docx` are the provenance for what the numbers mean and are
archived alongside the data, not parsed.

**Why this is interesting**: 24 *M. magister* megalopae in 310 samples over seven
years — a near-absence record. The 2012 correspondence establishes that historic
CalCOFI samples labelled *Cancer magister* are probably *C. productus*, so the
`dungeness_specimen` verification table is load-bearing rather than trivia.

**Held out of the release** — see `in_release: false` in the frontmatter and
"Not yet done" at the end. This notebook writes its complete parquet outputs;
`release_database.qmd` ignores them.

```{mermaid}
graph LR
  X1[Compiled Cruises] --> S[dungeness_sample<br/>station + effort]
  X1 --> M[dungeness_measurement<br/>counts, long]
  X1 --> C[dungeness_carapace<br/>per individual]
  X2[sorting log] --> L[dungeness_sorting_log<br/>sorted / unsorted]
  X3[archived megalopae] --> V[dungeness_specimen<br/>ID verification]
  M -.dungeness_sample_id.-> S
  C -.dungeness_sample_id.-> S
  M -.measurement_type.-> T[measurement_type]
  S -.site_sample_key.-> I[(swfsc_ichthyo<br/>sample / ship / grid)]
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))
source(here("libs/ingest.R"))
source(here("libs/parse_dungeness_crab.R"))

cc           <- read_calcofi_meta(here("ingest_cdfw_dungeness-crab.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_src      <- path_expand(glue("{dir_data}/cdfw/dungeness-crab"))

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

# the tables this ingest owns
dcr_tbls <- c("dungeness_sample", "dungeness_measurement", "dungeness_carapace",
              "dungeness_sorting_log", "dungeness_specimen")

# Other ingests archive their sources to gs://calcofi-files-public and their
# outputs to gs://calcofi-db unconditionally. This one does NOT by default:
# whether these files may be published is itself an open blocker (Q08 — no
# licence, no citation, and CDFW commissioned the work), and both buckets are
# world-readable. Uploading is a one-way door, so it is opt-in until Q08 is
# answered. Flip alongside `in_release: true`.
publish_to_gcs <- FALSE
```

## Read Source Workbooks

The workbook parsing lives in `libs/parse_dungeness_crab.R` so it is reviewable
and re-runnable. Three quirks it absorbs, each of which silently corrupts the
data if missed:

- the time series workbook is saved with Excel's **1904 date epoch**
  (`date1904="1"`), so every date read as text is four years early — the cruise
  labels are what caught it (`1404SH` reading as 2010);
- the sorting log lost the leading month from most `SampleDate` cells
  (`/5/1988`), recoverable from its separate `Month` column;
- the sorting log pads numbers with U+00A0, which defeats `as.numeric()`.

```{r}
#| label: read-source

stopifnot("source dir not found" = dir_exists(dir_src))
# archive the sources (including the PDF/DOCX provenance) — see publish_to_gcs
if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_src, gcs_prefix = glue("archive/{provider}/{dataset}"),
              bucket = "calcofi-files-public", exclude = c(".DS_Store", "*.tmp"))
} else {
  cat("publish_to_gcs is FALSE — sources NOT archived to gs://calcofi-files-public",
      "(pending Q08: licence + permission to publish)\n")
}

f_ts   <- file.path(dir_src, "Dungeness Time Series - Angela Klemmedson.xlsx")
f_log  <- file.path(dir_src, "Cancer Magister sorting update as of 02-15-12_EJones.xlsx")
f_spec <- file.path(dir_src, "ScrippsArchivedMegalopae_sent_5_10_2012.xlsx")
stopifnot(file_exists(f_ts), file_exists(f_log), file_exists(f_spec))

d_ts   <- read_dungeness_timeseries(f_ts)
d_log  <- read_dungeness_sorting_log(f_log)
d_spec <- read_dungeness_specimens(f_spec)
d_note <- read_dungeness_notes(f_ts)

cat(glue("time series {nrow(d_ts)} samples ({format(min(d_ts$date))} to ",
         "{format(max(d_ts$date))}); sorting log {nrow(d_log)} samples; ",
         "specimens {nrow(d_spec)}"), "\n")

# the workbook's own "Plots" sheet summarises 310 samples / 24 M. magister; hold
# the parse to that, so a re-export that changes the sheet fails loudly here
stopifnot(
  "expected 310 samples in 'Compiled Cruises'" = nrow(d_ts) == 310,
  "expected 24 M. magister megalopae in total" =
    sum(d_ts$n_mega_magister, na.rm = TRUE) == 24,
  "1904 epoch not applied — dates predate the 2008-2014 series" =
    min(d_ts$date) >= as.Date("2008-01-01"))
```

## Methods, in the Sorter's Own Words

The workbook's `Notes` sheet is **empty to every spreadsheet reader** — its
content is a floating text box, which lives in `xl/drawings/` rather than in any
cell, so `read_excel()` returns 0×0. It is the most valuable single artifact here:
Angela Klemmedson's dated statement of gear, magnification, measurement
landmarks, station bounds, results, and her reasoning about aliquots. Recovering
it answered three questions this ingest had open, so it is reproduced verbatim
below rather than summarized.

Three things it settles:

1. **Counts are occurrence, not density, by the author's explicit choice** —
   *"I am ignoring 'Aliq.' and 'Volume' because I am not measuring the density of
   megalopae in a volume of water, simply the occurrence of megalopae in
   samples… therefore I will not adjust the numbers of megalopae."* So the raw
   counts this ingest stores are what the author intended, and the aliquot column
   is context, not a correction factor.
2. **"14 samples (15 total removal vials)"** — independent confirmation of the
   15-rows / 14-occupations discrepancy found below, and its explanation: the
   repeated station occupation is a second *removal vial*, not a duplicate row.
3. **The non-Dungeness counts are lower-confidence** — *"Data beyond Dungeness
   crab megalopae should be used more as a guideline… there is potential for
   error in some of the identifications."* Carried into the column descriptions
   so it reaches consumers.

```{r}
#| label: methods-note
#| echo: false
#| results: asis

cat("::: {.callout-note appearance='simple' title='Notes sheet, ",
    "\"Dungeness Time Series - Angela Klemmedson.xlsx\" (recovered from ",
    "xl/drawings/drawing1.xml)'}\n\n", sep = "")
cat(paste0("> ", gsub("\n", "\n> ", d_note)), "\n\n")
cat(":::\n")
```

## Resolve Cruise + Position from the CalCOFI Reference

The time series records a cruise label and a CalCOFI line/station/order-of-
occupation, but **no coordinates**. Both are recoverable from the CalCOFI
reference tables:

- `cruise_key`: the label is `YYMM` + the two-letter CalCOFI ship abbreviation,
  which is `ship.ship_key` — so `1404SH` → `2014-04-3322` (Bell M. Shimada).
- position: join the resulting `(cruise_key, site_key, order_occ)` to the station
  occupations in the `swfsc_ichthyo` core `sample` table. That triple identifies
  a physical station occupation, so the match also links each sorted sample back
  to the tows and nets it came from via `site_sample_key`.

```{r}
#| label: resolve-keys

load_prior_tables(
  con, parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables = c("ship", "cruise", "grid", "sample"),
  geom_tables = c("grid", "sample"), as_view = TRUE)

d_ship  <- dbGetQuery(con, "SELECT ship_key, ship_nodc, ship_name FROM ship")
valid_ck <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM cruise")$cruise_key

d_cruise_lbl <- parse_cruise_label(unique(d_ts$cruise_orig)) |>
  left_join(d_ship, by = "ship_key") |>
  mutate(cruise_key = if_else(
    is.na(year) | is.na(ship_nodc), NA_character_,
    sprintf("%04d-%02d-%s", year, month, ship_nodc))) |>
  mutate(cruise_key = if_else(cruise_key %in% valid_ck, cruise_key, NA_character_))

stopifnot("every cruise label must resolve to a known cruise_key" =
            !any(is.na(d_cruise_lbl$cruise_key)))
d_cruise_lbl |>
  select(cruise_orig, ship_key, ship_name, cruise_key) |>
  datatable(caption = "Cruise label → cruise_key", options = list(dom = "t"),
            rownames = FALSE)
```

```{r}
#| label: match-sites

# CalCOFI station occupations: one row per (cruise, station, order of occupation)
d_site <- dbGetQuery(con, "
  SELECT cruise_key, site_key, order_occ,
         sample_key AS site_sample_key, latitude, longitude, datetime
  FROM sample WHERE sample_type = 'site'")

d_ts_keyed <- d_ts |>
  left_join(select(d_cruise_lbl, cruise_orig, ship_key, cruise_key),
            by = "cruise_orig") |>
  left_join(d_site, by = c("cruise_key", "site_key", "order_occ")) |>
  mutate(position_source = if_else(
    is.na(site_sample_key), "unmatched", "swfsc_ichthyo:site"))

n_matched <- sum(d_ts_keyed$position_source != "unmatched")
cat(glue("station match: {n_matched}/{nrow(d_ts_keyed)} samples ",
         "({round(100 * n_matched / nrow(d_ts_keyed), 1)}%)"), "\n")

# an independent check on the match AND on the unconfirmed time zone: the source
# date should agree with the matched occupation's own timestamp
d_chk <- d_ts_keyed |>
  filter(position_source != "unmatched") |>
  mutate(d_days = as.integer(date - as.Date(datetime)))
cat(glue("date agreement with matched occupation: ",
         "{sum(d_chk$d_days == 0, na.rm = TRUE)}/{nrow(d_chk)} same day, ",
         "{sum(abs(d_chk$d_days) <= 1, na.rm = TRUE)} within 1 day"), "\n")

d_ts_keyed |>
  filter(position_source == "unmatched") |>
  select(cruise_orig, cruise_key, line, station, order_occ, date) |>
  datatable(caption = "Samples with no matching CalCOFI station occupation (Q09)",
            options = list(dom = "t"), rownames = FALSE)
```

## Build Sample + Measurement Tables

Counts go long, one measurement per row, with the sorter's free-text
identification note kept beside the count it describes. Per-individual carapace
lengths are their own table: the source cell holds one length per megalopa found
(`"6.0mm, 6.5mm, 7.0mm, 7.0mm"`), which is sub-occurrence detail, not a scalar.

Counts are recorded **as found in the examined aliquot** and are deliberately
*not* standardized — which matches the author's stated choice (see the Notes
sheet above): the aliquot is context for interpreting an absence, not a
correction factor, and no filtered-water volume exists in the source to make a
density anyway.

```{r}
#| label: build-sample

d_sample <- d_ts_keyed |>
  mutate(dungeness_sample_id = row_number(), .before = 1) |>
  transmute(
    dungeness_sample_id, cruise_orig, cruise_key, ship_key,
    line, station, site_key, order_occ = as.integer(order_occ),
    date,
    datetime_start_utc = as_datetime(date) + dseconds(coalesce(time_sec, 0L)),
    latitude, longitude, position_source, site_sample_key,
    aliquot, volume_ml,
    sorted_by = "Angela Klemmedson",
    comments)
dbWriteTable(con, "dungeness_sample", as.data.frame(d_sample), overwrite = TRUE)
cat(glue("dungeness_sample: {nrow(d_sample)} rows"), "\n")
```

```{r}
#| label: build-measurement

# (count column, free-text description column) -> canonical measurement_type
meas_map <- tribble(
  ~measurement_type,    ~n_col,             ~desc_col,
  "megalopae_magister", "n_mega_magister",  NA_character_,
  "megalopae_other",    "n_mega_other",     "mega_other_desc",
  "zoea_cancer",        "n_zoea_cancer",    "zoea_cancer_desc",
  "zoea_other",         "n_zoea_other",     "zoea_other_desc")

d_meas <- pmap_dfr(meas_map, function(measurement_type, n_col, desc_col) {
  tibble(
    dungeness_sample_id = d_sample$dungeness_sample_id,
    measurement_type    = measurement_type,
    measurement_value   = d_ts_keyed[[n_col]],
    taxa_description    = if (is.na(desc_col)) NA_character_
                          else d_ts_keyed[[desc_col]])
}) |>
  # a blank source cell is "not recorded", which is not the same as a counted
  # zero — keep the row and flag it rather than dropping or zero-filling it
  mutate(measurement_qual = if_else(
    is.na(measurement_value), "blank in source", NA_character_)) |>
  arrange(dungeness_sample_id, measurement_type) |>
  mutate(dungeness_measurement_id = row_number(), .before = 1)
dbWriteTable(con, "dungeness_measurement", as.data.frame(d_meas), overwrite = TRUE)

# nothing may be lost or invented in the pivot
stopifnot(
  "measurement rows must be samples x types" =
    nrow(d_meas) == nrow(d_sample) * nrow(meas_map),
  "pivoted counts must sum to the source totals" =
    sum(d_meas$measurement_value, na.rm = TRUE) ==
    sum(d_ts_keyed$n_mega_magister, d_ts_keyed$n_mega_other,
        d_ts_keyed$n_zoea_cancer,   d_ts_keyed$n_zoea_other, na.rm = TRUE))
cat(glue("dungeness_measurement: {nrow(d_meas)} rows ",
         "({sum(!is.na(d_meas$measurement_qual))} blank in source)"), "\n")
```

```{r}
#| label: build-carapace

d_carapace <- parse_carapace_lengths(d_ts_keyed$carapace_len_txt) |>
  mutate(dungeness_sample_id = d_sample$dungeness_sample_id[row_index]) |>
  select(dungeness_sample_id, individual_num, carapace_length_mm) |>
  arrange(dungeness_sample_id, individual_num) |>
  mutate(dungeness_carapace_id = row_number(), .before = 1)
dbWriteTable(con, "dungeness_carapace", as.data.frame(d_carapace), overwrite = TRUE)

# one measured individual per megalopa counted — this is the check that the
# carapace cell was split correctly rather than plausibly
n_mag <- sum(d_ts_keyed$n_mega_magister, na.rm = TRUE)
stopifnot("one carapace length per M. magister megalopa counted" =
            nrow(d_carapace) == n_mag)
cat(glue("dungeness_carapace: {nrow(d_carapace)} individuals, ",
         "{min(d_carapace$carapace_length_mm)}-",
         "{max(d_carapace$carapace_length_mm)} mm"), "\n")
```

## Build Sorting Log + Specimen Verification

The sorting log is **effort, not counts**: which archived samples had been looked
at as of 2012-02-15 and which had not. It carries its own coordinates, so no
station matching is needed — but it reaches far north and offshore of the CalCOFI
grid, so most rows get no `grid_key`.

It records no ship, so `cruise_key` is assigned only where the parsed year-month
matches exactly one cruise in the reference.

```{r}
#| label: build-sorting-log

d_ym_lookup <- dbGetQuery(con, "
  SELECT strftime(date_ym, '%Y-%m') AS cruise_ym, cruise_key
  FROM cruise") |>
  count(cruise_ym, cruise_key) |>
  add_count(cruise_ym, name = "n_cruises") |>
  filter(n_cruises == 1) |>
  select(cruise_ym, cruise_key)

d_sorting <- d_log |>
  mutate(
    # expedition labels are inconsistent ("8803", "CALCOFI 0404", "BODEGA BAY");
    # take a trailing YYMM where one exists, otherwise no cruise at all
    yymm      = str_match(expedition_orig, "(\\d{2})(\\d{2})[A-Z]?$")[, 1],
    yy        = as.integer(str_sub(yymm, 1, 2)),
    mm        = as.integer(str_sub(yymm, 3, 4)),
    # this log spans 1949-2009: resolve the 2-digit year against the row's own
    # date rather than a fixed pivot
    cruise_ym = if_else(
      is.na(yymm) | is.na(date), NA_character_,
      sprintf("%04d-%02d", 100L * (year(date) %/% 100L) + yy, mm)),
    # one row records longitude +138.483 where every other is negative; flag it
    # rather than silently negating (Q07)
    longitude = if_else(!is.na(longitude) & longitude > 0, NA_real_, longitude)) |>
  left_join(d_ym_lookup, by = "cruise_ym") |>
  transmute(
    sorting_status, expedition_orig, cruise_ym, cruise_key,
    line, station, site_key, latitude, longitude,
    date,
    datetime_start_utc = as_datetime(date) + dseconds(coalesce(time_start_sec, 0L)),
    datetime_end_utc   = as_datetime(date) + dseconds(coalesce(time_end_sec, 0L)),
    max_depth_m, net, mesh_mm, tow_type, preservative,
    sorted_by = "Emily Jones") |>
  arrange(date, site_key) |>
  mutate(dungeness_sorting_log_id = row_number(), .before = 1)
dbWriteTable(con, "dungeness_sorting_log", as.data.frame(d_sorting), overwrite = TRUE)

cat(glue("dungeness_sorting_log: {nrow(d_sorting)} rows ",
         "({sum(d_sorting$sorting_status == 'sorted')} sorted, ",
         "{sum(d_sorting$sorting_status == 'unsorted')} unsorted); ",
         "cruise_key resolved for {sum(!is.na(d_sorting$cruise_key))}; ",
         "dates {format(min(d_sorting$date))} to {format(max(d_sorting$date))}"), "\n")
```

```{r}
#| label: build-specimen

d_specimen <- d_spec |>
  transmute(
    specimen_num, cruise_orig, line, station, site_key, date,
    datetime_start_utc = as_datetime(date) + dseconds(coalesce(time_sec, 0L)),
    is_magister, carapace_length_mm, notes)
dbWriteTable(con, "dungeness_specimen", as.data.frame(d_specimen), overwrite = TRUE)
cat(glue("dungeness_specimen: {nrow(d_specimen)} specimens, ",
         "{sum(d_specimen$is_magister)} confirmed M. magister"), "\n")
```

## Enforce Column Types

Before any geometry exists: `enforce_column_types()` rewrites columns, and a
table carrying a `GEOMETRY` column is the one thing DuckDB ≥ 1.5.1 will not let
you rewrite (see the known-bug note in `CLAUDE.md`).

```{r}
#| label: enforce-types

d_flds_rd <- read_csv(here(glue("metadata/{provider}/{dataset}/flds_redefine.csv")))
enforce_column_types(con, d_flds_rd = d_flds_rd, tables = dcr_tbls)
```

## Spatial

`geom` and `grid_key` come last, so every other column value is already settled
in R and nothing has to `UPDATE` a table that has geometry on it.

```{r}
#| label: spatial

for (tbl in c("dungeness_sample", "dungeness_sorting_log")) {
  add_point_geom(con, tbl, lon_col = "longitude", lat_col = "latitude")
  assign_grid_key(con, tbl)
}

d_spatial_cov <- dbGetQuery(con, "
  SELECT 'dungeness_sample'      AS tbl, COUNT(*) AS n,
         COUNT(latitude)         AS n_positioned,
         COUNT(grid_key)         AS n_in_grid FROM dungeness_sample
  UNION ALL
  SELECT 'dungeness_sorting_log', COUNT(*), COUNT(latitude), COUNT(grid_key)
  FROM dungeness_sorting_log")

# the borrowed swfsc_ichthyo reference VIEWs have done their job (cruise/ship
# resolution, station match, grid assignment). Drop them so validation and the
# metadata sidecar see only this dataset's tables — otherwise the checks below
# report NULLs in ichthyo's 61k-row `sample`, which is not this ingest's business.
for (v in c("sample", "ship", "cruise", "grid"))
  dbExecute(con, glue("DROP VIEW IF EXISTS {v}"))

d_spatial_cov |>
  datatable(caption = "Position + grid coverage", options = list(dom = "t"),
            rownames = FALSE)
```

## Measurement Vocabulary + Dataset Metadata

The source's four count columns are **not** measurement types. Baking the taxon
into the type name (`megalopae_magister`) is exactly the anti-pattern the core
model exists to remove: in `obs` the quantity is `abundance` and the organism is
`taxon_key`. So the four raw labels became entries in
`metadata/measurement_taxon.csv`, which maps each to
`(measurement_type, taxon_key, life_stage)` — see "Emit Core Tables" below.

That leaves two genuinely new canonical quantities, staged in
`metadata/cdfw/dungeness-crab/measurement_type_new.csv` rather than appended to
the shared `metadata/measurement_type.csv`: that registry is loaded wholesale
into the release, so adding them now would publish types with no observations
into a release that excludes this dataset. **Move them into the shared registry
in the same change that flips `in_release: true`.** The aliquot fraction maps onto
the existing `prop_sorted`, so it needs nothing new.

```{r}
#| label: meas-types

d_meas_type_shared <- read_measurement_type(here("metadata/measurement_type.csv"))
d_meas_type_new    <- read_csv(here(glue(
  "metadata/{provider}/{dataset}/measurement_type_new.csv")))

stopifnot(
  "staged measurement types must not already be in the shared registry" =
    !any(d_meas_type_new$measurement_type %in% d_meas_type_shared$measurement_type))

d_meas_type <- bind_rows(d_meas_type_shared, d_meas_type_new)
dbWriteTable(con, "measurement_type", as.data.frame(d_meas_type), overwrite = TRUE)

this_provider <- provider; this_dataset <- dataset   # `provider`/`dataset` are also column names
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot("this dataset must appear in the ingest YAML registry" = nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)

d_meas_type_new |>
  select(measurement_type, description, units, grain) |>
  datatable(caption = "New canonical measurement types (staged, not in the shared registry)",
            options = list(dom = "t"), rownames = FALSE)
```

## Emit Core Tables

Project this dataset into the shared consolidated core model (`RUNBOOK.md` §3b,
`design_env-bio-consolidation.md`). These core tables **are** this ingest's
published output — `release_database.qmd` concatenates per-dataset shards rather
than re-deriving them — and the per-dataset tables survive as compat VIEWs.

This uses the generic `append_*` helpers with plain SQL, so **no `calcofi4db`
change is needed to add a dataset**. `emit_core_tables()` is a convenience wrapper
whose `switch(dataset_key, …)` arms happen to hold the already-migrated datasets'
SQL; a dataset's projection belongs in the notebook that owns it.

Two modelling decisions worth review:

- **`sample_type = 'subsample'` is new vocabulary** (existing values: `bottle`,
  `cast`, `net`, `region_pool`, `site`, `tow`, `transect`, `underway`). The grain
  genuinely is a lab subsample of an archived net catch — finer than the tow,
  since three station occupations were examined twice as separate removal vials —
  so none of the existing values fits. Flagged as Q11.
- **`parent_sample_key` points at the matched `swfsc_ichthyo` site occupation**, a
  deliberate cross-dataset edge in the `sample` adjacency list. This is the payoff
  of the station match: it makes crab larvae joinable to CTD and bottle
  environment through the shared physical station occupation instead of stranding
  them on their own island.

The 2,011-row sorting log becomes `sample` rows (`sample_type = 'tow'`), and the
sorted/unsorted distinction becomes the presence or absence of an `obs` row:

- **216 sorted** → one `obs` row for *M. magister* with `measurement_value = 0`.
  "Examined, none found" is a real observation, and a zero is how you record it.
- **1,795 unsorted** → `sample` row, no `obs`. We have not looked, which is not
  the same as looking and finding nothing.

That distinction is the entire scientific content of the log, and encoding it this
way means it needs no `sorting_status` column of its own — which is what lets this
ingest publish the core and nothing else (see "Write Outputs"). Note the zeros are
safe with respect to the open Q03: that question doubts the *positive* Reilly-era
"Cancer magister" labels, not Jones's report of finding none.

```{r}
#| label: emit-core

ds_key <- glue("{provider}_{dataset}")

# the taxon crosswalk resolves each raw count label to (measurement_type,
# taxon_key, life_stage). Restricted to this dataset so the helpers below build
# only our slice of the shared taxa references.
# worms_id/itis_id MUST be integer: read as double, CAST(... AS VARCHAR) yields
# "440388.0" and every taxon_key silently fails to resolve against `taxon`
mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"))
stopifnot("crosswalk must cover all four raw count labels" =
            setequal(mt_taxon$raw_measurement_type[mt_taxon$target == "obs"],
                     c("megalopae_magister", "megalopae_other",
                       "zoea_cancer", "zoea_other")))
dbWriteTable(con, "_measurement_taxon", as.data.frame(mt_taxon), overwrite = TRUE)

# taxa references — generic helpers, driven by the crosswalk rather than by a
# per-dataset vocabulary table
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- build_dataset_taxon(con,   mt_taxon, tx_over)
n_tx_group <- build_taxon_group(con,     mt_taxon, tx_over)

sub_key <- glue("'{ds_key}:subsample:' || CAST(dungeness_sample_id AS VARCHAR)")

# 1. sample — examined subsamples, parented to the matched station occupation
append_sample(con, glue("
  SELECT {sub_key}, 'subsample', site_sample_key,
         COALESCE(site_sample_key, {sub_key}),
         '{ds_key}', grid_key, site_key, cruise_key, order_occ,
         latitude, longitude, CAST(datetime_start_utc AS TIMESTAMP),
         NULL::DOUBLE, NULL::DOUBLE, NULL::VARCHAR
  FROM dungeness_sample"))

# 1b. sample — sorting-log tows (effort/absence; no obs attach to these)
append_sample(con, glue("
  SELECT '{ds_key}:tow:' || CAST(dungeness_sorting_log_id AS VARCHAR),
         'tow', NULL::VARCHAR,
         '{ds_key}:tow:' || CAST(dungeness_sorting_log_id AS VARCHAR),
         '{ds_key}', grid_key, site_key, cruise_key, NULL::INTEGER,
         latitude, longitude, CAST(datetime_start_utc AS TIMESTAMP),
         0::DOUBLE, max_depth_m, tow_type
  FROM dungeness_sorting_log"))

# 2. obs — occurrence headline. Because the taxon now lives in taxon_key, the
#    four source columns collapse to ONE measurement_type across four
#    taxon_key/life_stage combinations.
append_obs(con, glue("
  SELECT 'bio', '{ds_key}',
         '{ds_key}:subsample:' || CAST(s.dungeness_sample_id AS VARCHAR),
         s.grid_key, s.cruise_key, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP),
         NULL::DOUBLE, NULL::DOUBLE,
         'worms:' || CAST(CAST(mx.worms_id AS BIGINT) AS VARCHAR),
         mx.life_stage, mx.measurement_type, m.measurement_value,
         m.measurement_qual, NULL::DOUBLE
  FROM dungeness_measurement m
  JOIN dungeness_sample s USING (dungeness_sample_id)
  JOIN _measurement_taxon mx
    ON mx.raw_measurement_type = m.measurement_type AND mx.target = 'obs'"))

# 2b. obs — the sorting log's ABSENCES. A sorted sample with no M. magister is a
#     zero-valued occurrence, not a missing row; an unsorted sample gets nothing,
#     because "not examined" and "examined, none found" are different facts.
append_obs(con, glue("
  SELECT 'bio', '{ds_key}',
         '{ds_key}:tow:' || CAST(l.dungeness_sorting_log_id AS VARCHAR),
         l.grid_key, l.cruise_key, l.latitude, l.longitude,
         CAST(l.datetime_start_utc AS TIMESTAMP),
         0::DOUBLE, l.max_depth_m,
         'worms:' || CAST(CAST(mx.worms_id AS BIGINT) AS VARCHAR),
         mx.life_stage, mx.measurement_type, 0::DOUBLE,
         'examined, none found (E. Jones sorting log, as of 2012-02-15)',
         NULL::DOUBLE
  FROM dungeness_sorting_log l
  CROSS JOIN (SELECT * FROM _measurement_taxon
              WHERE raw_measurement_type = 'megalopae_magister') mx
  WHERE l.sorting_status = 'sorted'"))

# 3. obs_attribute — per-individual carapace lengths within the M. magister
#    occurrence: one row per measured megalopa, bin_value IS the length
append_obs_attribute(con, glue("
  SELECT '{ds_key}',
         '{ds_key}:subsample:' || CAST(c.dungeness_sample_id AS VARCHAR),
         'worms:' || CAST(CAST(mx.worms_id AS BIGINT) AS VARCHAR), mx.life_stage,
         mx.measurement_type, c.carapace_length_mm,
         NULL::VARCHAR, 1::BIGINT, NULL::VARCHAR
  FROM dungeness_carapace c
  CROSS JOIN (SELECT * FROM _measurement_taxon
              WHERE target = 'attribute' AND measurement_type = 'carapace_length') mx"))

# 4. sample_measurement — event-level effort/context for the examined subsamples
append_sample_measurement(con, glue("
  SELECT '{ds_key}:subsample:' || CAST(dungeness_sample_id AS VARCHAR),
         '{ds_key}', mt, mv, NULL::VARCHAR
  FROM (
    SELECT dungeness_sample_id, 'prop_sorted' AS mt, aliquot AS mv FROM dungeness_sample
    UNION ALL
    SELECT dungeness_sample_id, 'settled_volume_ml', volume_ml FROM dungeness_sample)
  WHERE mv IS NOT NULL"))

core_n <- list(
  sample             = dbGetQuery(con, "SELECT COUNT(*) FROM sample")[[1]],
  obs                = dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]],
  obs_attribute      = dbGetQuery(con, "SELECT COUNT(*) FROM obs_attribute")[[1]],
  sample_measurement = dbGetQuery(con, "SELECT COUNT(*) FROM sample_measurement")[[1]],
  taxon              = n_taxon, dataset_taxon = n_ds_taxon, taxon_group = n_tx_group)
cat(glue("core projection — ",
         "{paste(names(core_n), unlist(core_n), sep = '=', collapse = '  ')}"), "\n")
```

The projection must be lossless and must not invent anything, so assert it rather
than eyeball it:

```{r}
#| label: core-parity

n_sub <- dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE sample_type='subsample'")[[1]]
n_tow <- dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE sample_type='tow'")[[1]]
stopifnot(
  "one core sample per examined subsample" = n_sub == nrow(d_sample),
  "one core sample per sorting-log tow"    = n_tow == nrow(d_sorting),
  "sample_key must be unique within the shard" =
    dbGetQuery(con, "SELECT COUNT(*) FROM (SELECT sample_key FROM sample
                     GROUP BY 1 HAVING COUNT(*) > 1)")[[1]] == 0,
  # the time series' counts, plus one zero-valued absence per sorted log sample
  "obs must carry every source count row plus the sorting-log absences" =
    core_n$obs == nrow(d_meas) + sum(d_sorting$sorting_status == "sorted"),
  "obs totals must equal the source totals (absences add zero)" =
    dbGetQuery(con, "SELECT SUM(measurement_value) FROM obs")[[1]] ==
    sum(d_meas$measurement_value, na.rm = TRUE),
  "every sorted log sample must have exactly one absence obs" =
    dbGetQuery(con, "
      SELECT COUNT(*) FROM obs o JOIN sample s USING (sample_key)
      WHERE s.sample_type = 'tow'")[[1]] ==
      sum(d_sorting$sorting_status == "sorted"),
  "unsorted log samples must carry no obs" =
    dbGetQuery(con, glue("
      SELECT COUNT(*) FROM obs o
      WHERE o.sample_key IN (
        SELECT '{ds_key}:tow:' || CAST(dungeness_sorting_log_id AS VARCHAR)
        FROM dungeness_sorting_log WHERE sorting_status = 'unsorted')"))[[1]] == 0,
  "sorting-log absences must all be zero" =
    dbGetQuery(con, "
      SELECT COUNT(*) FROM obs o JOIN sample s USING (sample_key)
      WHERE s.sample_type = 'tow' AND o.measurement_value <> 0")[[1]] == 0,
  "every obs must resolve a taxon_key" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs WHERE taxon_key IS NULL")[[1]] == 0,
  # NOT-NULL is not enough: a numeric-typed worms_id renders as "440388.0", which
  # is non-NULL and joins to nothing. Assert the FK actually resolves.
  "every obs.taxon_key must resolve in taxon" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs o LEFT JOIN taxon t USING (taxon_key)
                     WHERE t.taxon_key IS NULL")[[1]] == 0,
  "every obs_attribute.taxon_key must resolve in taxon" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs_attribute a
                     LEFT JOIN taxon t USING (taxon_key)
                     WHERE t.taxon_key IS NULL")[[1]] == 0,
  "every obs.measurement_type must resolve in measurement_type" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs o
                     LEFT JOIN measurement_type t USING (measurement_type)
                     WHERE t.measurement_type IS NULL")[[1]] == 0,
  "every obs.sample_key must resolve in sample" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key)
                     WHERE s.sample_key IS NULL")[[1]] == 0,
  "one obs_attribute row per measured individual" =
    core_n$obs_attribute == nrow(d_carapace),
  # the M. magister headline and its carapace attribution must agree per sample
  "carapace individuals must match the magister count per sample" =
    dbGetQuery(con, "
      WITH a AS (SELECT sample_key, SUM(count) s FROM obs_attribute
                 WHERE measurement_type='carapace_length' GROUP BY 1),
           o AS (SELECT sample_key, SUM(measurement_value) v FROM obs
                 WHERE taxon_key='worms:440388' AND measurement_type='abundance'
                 GROUP BY 1)
      SELECT COUNT(*) FROM a JOIN o USING (sample_key) WHERE a.s <> o.v")[[1]] == 0,
  # the headline count must come from the time series alone, not the absences
  "the 24 megalopae must all come from the examined subsamples" =
    dbGetQuery(con, "
      SELECT SUM(o.measurement_value) FROM obs o JOIN sample s USING (sample_key)
      WHERE s.sample_type = 'subsample' AND o.taxon_key = 'worms:440388'")[[1]] == 24)
cat("core parity: all assertions passed\n")

dbGetQuery(con, "
  SELECT COALESCE(t.scientific_name, o.taxon_key) AS taxon, o.life_stage,
         o.measurement_type, COUNT(*) AS n_obs, SUM(o.measurement_value) AS total
  FROM obs o LEFT JOIN taxon t USING (taxon_key)
  GROUP BY 1,2,3 ORDER BY total DESC") |>
  datatable(caption = "obs by resolved taxon — the four source columns, collapsed",
            options = list(dom = "t"), rownames = FALSE)
```

The per-dataset tables the core fully models become VIEWs over it, so the previews
below and any ad-hoc query keep working against the names this notebook built. The
originals are kept as `*_src` for the source ERD.

These VIEWs are a **convenience inside this notebook only** — they are not
published. Nothing per-dataset is: see "Write Outputs".

```{r}
#| label: compat-views

for (t in c("dungeness_measurement", "dungeness_carapace")) {
  invisible(dbExecute(con, glue("DROP TABLE IF EXISTS {t}_src")))
  invisible(dbExecute(con, glue("ALTER TABLE {t} RENAME TO {t}_src")))
}

# scoped to the examined subsamples: the 216 sorting-log absence rows are also
# `obs` for this dataset, but they were never part of dungeness_measurement and
# their sample_key resolves to a sorting-log id, not a dungeness_sample_id
invisible(dbExecute(con, glue("
  CREATE OR REPLACE VIEW dungeness_measurement AS
  SELECT obs_id AS dungeness_measurement_id,
         CAST(split_part(sample_key, ':', 3) AS INTEGER) AS dungeness_sample_id,
         measurement_type, measurement_value, measurement_qual
  FROM obs
  WHERE dataset_key = '{ds_key}' AND split_part(sample_key, ':', 2) = 'subsample'")))
invisible(dbExecute(con, glue("
  CREATE OR REPLACE VIEW dungeness_carapace AS
  SELECT obs_attribute_id AS dungeness_carapace_id,
         CAST(split_part(sample_key, ':', 3) AS INTEGER) AS dungeness_sample_id,
         ROW_NUMBER() OVER (PARTITION BY sample_key ORDER BY obs_attribute_id)
           AS individual_num,
         bin_value AS carapace_length_mm
  FROM obs_attribute
  WHERE dataset_key = '{ds_key}' AND measurement_type = 'carapace_length'")))

n_view <- dbGetQuery(con, "SELECT COUNT(*) FROM dungeness_measurement")[[1]]
n_cara <- dbGetQuery(con, "SELECT COUNT(*) FROM dungeness_carapace")[[1]]
stopifnot(
  "compat view must expose every source measurement row" = n_view == nrow(d_meas),
  "compat view must expose every obs_attribute row" = n_cara == core_n$obs_attribute)
cat(glue("compat views over core — dungeness_measurement {n_view} rows ",
         "(the {core_n$obs - n_view} sorting-log absences are obs but not part of ",
         "this view), dungeness_carapace {n_cara} rows"), "\n")
```

## Schema

Two diagrams, because they answer different questions: what the workbooks look
like, and what this ingest actually publishes.

### What gets published — the core

This is the schema a consumer sees. It is the only thing written to parquet.

```{r}
#| label: schema-core

core_tbls <- core_output_tables(con, extra = "measurement_type")
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls),
       colors = list(
         lightblue   = c("sample", "obs"),
         lightgreen  = c("obs_attribute", "sample_measurement"),
         lightyellow = c("taxon", "dataset_taxon", "measurement_type")))
```

### How the source maps onto it

The workbooks are three unrelated spreadsheets; the core is five related tables.
Neither diagram alone shows how one becomes the other, so state it explicitly —
this is the part a reader actually needs in order to trust the numbers.

```{r}
#| label: source-to-core
#| echo: false

tribble(
  ~source, ~`source field`, ~`core table`, ~`core field`, ~transform,
  "Compiled Cruises", "Cruise", "sample", "cruise_key",
    "YYMM + 2-letter ship abbrev resolved through ship.ship_key",
  "Compiled Cruises", "Line / Station", "sample", "site_key",
    "formatted 'LLL.L SSS.S'; position from the matched ichthyo occupation",
  "Compiled Cruises", "Order_Occ", "sample", "order_occ",
    "matched to swfsc_ichthyo site occupation -> parent_sample_key",
  "Compiled Cruises", "Date + Time", "sample", "datetime",
    "1904-epoch date + day-fraction time (tz unconfirmed, Q05)",
  "Compiled Cruises", "Dungeness Crab Megalopae", "obs",
    "measurement_value", "measurement_type='abundance', taxon_key=worms:440388, life_stage='megalopa'",
  "Compiled Cruises", "Other Megalopae", "obs", "measurement_value",
    "as above, taxon_key=worms:1130 (Decapoda), life_stage='megalopa'",
  "Compiled Cruises", "Cancer Zoea", "obs", "measurement_value",
    "as above, taxon_key=worms:106749 (Cancridae), life_stage='zoea'",
  "Compiled Cruises", "Other Zoea", "obs", "measurement_value",
    "as above, taxon_key=worms:1130 (Decapoda), life_stage='zoea'",
  "Compiled Cruises", "Carapace length", "obs_attribute", "bin_value",
    "comma-separated cell split to one row per individual, count=1",
  "Compiled Cruises", "Aliq.", "sample_measurement", "measurement_value",
    "measurement_type='prop_sorted'",
  "Compiled Cruises", "Vol (ml)", "sample_measurement", "measurement_value",
    "measurement_type='settled_volume_ml'",
  "sorting log", "SampleDate + Month", "sample", "datetime",
    "day/year text recombined with the separate Month column",
  "sorting log", "Lat / Lon", "sample", "latitude / longitude",
    "as given; one positive longitude nulled as a suspected sign error (Q07)",
  "sorting log", "Tow / MaxDepth", "sample", "tow_type / depth_max_m", "as given",
  "sorting log", "sorted vs unsorted", "obs", "(row presence)",
    "sorted -> one zero-valued M. magister obs; unsorted -> no obs at all",
  "—", "the four raw count labels", "taxon / dataset_taxon", "taxon_key",
    "metadata/measurement_taxon.csv -> WoRMS ids",
  "NOT PUBLISHED", "free text (Description, Sample Comments, specimen notes)",
    "—", "—", "no core home; stays in the archived source workbooks",
  "NOT PUBLISHED", "gear detail (Net, Mesh, Preservative), Expedition label",
    "—", "—", "beyond sample.tow_type the core does not model gear; archived source",
  "NOT PUBLISHED", "5-specimen ID verification (is_magister)", "—", "—",
    "a taxonomic determination, not an abundance observation; see Overview + Q03"
) |>
  datatable(caption = "Source -> core mapping, including what is deliberately not published",
            options = list(dom = "t", pageLength = 25), rownames = FALSE)
```

### The source shape

For reference only — it documents the wrangling above, and none of it is
published.

```{r}
#| label: schema

dcr_rels <- list(
  primary_keys = list(
    dungeness_sample      = "dungeness_sample_id",
    dungeness_sorting_log = "dungeness_sorting_log_id",
    dungeness_specimen    = "specimen_num",
    measurement_type      = "measurement_type"),
  foreign_keys = list(
    list(table = "dungeness_measurement_src", column = "dungeness_sample_id",
         ref_table = "dungeness_sample", ref_column = "dungeness_sample_id"),
    list(table = "dungeness_carapace_src", column = "dungeness_sample_id",
         ref_table = "dungeness_sample", ref_column = "dungeness_sample_id")))

cc_erd(con,
       tables = c("dungeness_sample", "dungeness_measurement_src",
                  "dungeness_carapace_src", "dungeness_sorting_log",
                  "dungeness_specimen", "dataset"),
       rels = dcr_rels,
       colors = list(
         lightblue  = c("dungeness_sample", "dungeness_measurement_src",
                        "dungeness_carapace_src"),
         lightgreen = c("dungeness_sorting_log", "dungeness_specimen"),
         white      = "dataset"))
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
# the generic checker treats every *_key as required, so the expected-NULL cases
# here (unmatched stations, a shipless sorting log, positions outside the grid)
# are reported rather than silently passed over
if (length(results$errors))
  cat("Reported as errors, all expected for this dataset:\n  ",
      paste(results$errors, collapse = "\n  "), "\n")

# referential integrity of the two FKs this ingest owns
n_orphan_meas <- dbGetQuery(con, "
  SELECT COUNT(*) FROM dungeness_measurement m
  LEFT JOIN dungeness_sample s USING (dungeness_sample_id)
  WHERE s.dungeness_sample_id IS NULL")[[1]]
n_orphan_cara <- dbGetQuery(con, "
  SELECT COUNT(*) FROM dungeness_carapace c
  LEFT JOIN dungeness_sample s USING (dungeness_sample_id)
  WHERE s.dungeness_sample_id IS NULL")[[1]]
n_orphan_type <- dbGetQuery(con, "
  SELECT COUNT(*) FROM dungeness_measurement m
  LEFT JOIN measurement_type t USING (measurement_type)
  WHERE t.measurement_type IS NULL")[[1]]
# primary keys must be unique
n_dup_pk <- dbGetQuery(con, "
  SELECT COUNT(*) FROM (SELECT dungeness_sample_id FROM dungeness_sample
                        GROUP BY 1 HAVING COUNT(*) > 1)")[[1]]
stopifnot(
  "dungeness_measurement.dungeness_sample_id must resolve" = n_orphan_meas == 0,
  "dungeness_carapace.dungeness_sample_id must resolve"    = n_orphan_cara == 0,
  "dungeness_measurement.measurement_type must resolve"    = n_orphan_type == 0,
  "dungeness_sample_id must be unique"                     = n_dup_pk == 0)
cat("FK + PK integrity: OK\n")

# the headline result of the whole dataset, asserted so a bad re-parse is loud.
# 15 positive ROWS but 14 positive station OCCUPATIONS. The Notes sheet states it
# exactly — "M. magister megalopae found in 14 samples (15 total removal vials)"
# — so the repeat at 1204OS/070.0 055.0/occ 45 (1 megalopa in each row) is a
# second removal vial from one sample, not a duplicated row. The workbook's
# "M. magister values" sheet aggregates the pair, which is why it lists 14.
# asked of the CORE now, via taxon_key — the source's 'megalopae_magister' column
# name no longer exists as a measurement_type, which is the point of the model
# reported per EFFORT, because the two are not the same survey: the 2015 time
# series counted, the 2012 log only recorded that it looked. Lumping them silently
# changes the occurrence rate (15/310 becomes 15/526), so name the scope.
mag <- dbGetQuery(con, "
  SELECT s.sample_type,
         COUNT(*) AS n_examined,
         COUNT(*) FILTER (WHERE o.measurement_value > 0) AS n_positive_rows,
         COUNT(DISTINCT CASE WHEN o.measurement_value > 0
                             THEN s.cruise_key || s.site_key || s.order_occ END)
           AS n_positive_stations,
         SUM(o.measurement_value) AS n_megalopae
  FROM obs o JOIN sample s USING (sample_key)
  WHERE o.taxon_key = 'worms:440388' AND o.life_stage = 'megalopa'
  GROUP BY 1 ORDER BY 1") |>
  (\(d) d[match(c("subsample", "tow"), d$sample_type), ])()
names(mag)[1] <- "effort"
mag$effort <- c("2015 time series (Klemmedson)", "2012 sorting log (Jones)")
mag |> datatable(caption = "M. magister megalopae by effort — the log contributes examined-but-empty samples",
                 options = list(dom = "t"), rownames = FALSE)

mag <- dbGetQuery(con, "
  SELECT COUNT(*) AS n_samples,
         COUNT(*) FILTER (WHERE o.measurement_value > 0) AS n_positive_rows,
         COUNT(DISTINCT CASE WHEN o.measurement_value > 0
                             THEN s.cruise_key || s.site_key || s.order_occ END)
           AS n_positive_stations,
         SUM(o.measurement_value) AS n_megalopae
  FROM obs o JOIN sample s USING (sample_key)
  WHERE o.taxon_key = 'worms:440388' AND o.life_stage = 'megalopa'
    AND s.sample_type = 'subsample'")
stopifnot(
  "24 M. magister megalopae in total"                 = mag$n_megalopae == 24,
  "15 positive rows / 14 positive station occupations" =
    mag$n_positive_rows == 15 && mag$n_positive_stations == 14)
n_log_examined <- sum(d_sorting$sorting_status == "sorted")
cat(glue("M. magister: {mag$n_megalopae} megalopae in {mag$n_positive_rows} of ",
         "{mag$n_samples} samples in the 2015 time series ",
         "({round(100 * mag$n_positive_rows / mag$n_samples, 1)}% occurrence), ",
         "at {mag$n_positive_stations} distinct station occupations. ",
         "Plus {n_log_examined} examined-but-empty samples from the 2012 log, ",
         "for {mag$n_positive_rows}/{mag$n_samples + n_log_examined} ",
         "({round(100 * mag$n_positive_rows / (mag$n_samples + n_log_examined), 1)}%) ",
         "across both efforts"), "\n")

# replicate subsamples of one station occupation — the reason the sample PK is a
# sequential id rather than (cruise, station, order_occ)
dbGetQuery(con, "
  SELECT cruise_orig, site_key, order_occ, COUNT(*) AS n_rows
  FROM dungeness_sample GROUP BY 1, 2, 3 HAVING COUNT(*) > 1 ORDER BY 1, 2") |>
  datatable(caption = "Station occupations examined more than once (replicate subsamples)",
            options = list(dom = "t"), rownames = FALSE)
```

## Preview

```{r}
#| label: preview-sample
dbGetQuery(con, "
  SELECT * EXCLUDE (geom) FROM dungeness_sample
  ORDER BY dungeness_sample_id LIMIT 100") |>
  datatable(caption = "dungeness_sample — first 100", rownames = FALSE, filter = "top")
```

```{r}
#| label: preview-positive
# read through the CORE, the way a consumer would: obs joined to sample and taxon,
# with the per-individual lengths from obs_attribute
dbGetQuery(con, "
  SELECT s.cruise_key, s.site_key, s.order_occ, CAST(s.datetime AS DATE) AS date,
         round(s.latitude, 3) AS lat, round(s.longitude, 3) AS lon,
         t.scientific_name, o.life_stage, o.measurement_value AS n_magister,
         (SELECT string_agg(CAST(a.bin_value AS VARCHAR), ', ' ORDER BY a.bin_value)
          FROM obs_attribute a
          WHERE a.sample_key = o.sample_key
            AND a.measurement_type = 'carapace_length') AS carapace_mm
  FROM obs o
  JOIN sample s USING (sample_key)
  LEFT JOIN taxon t USING (taxon_key)
  WHERE o.taxon_key = 'worms:440388' AND o.measurement_value > 0
  ORDER BY s.datetime") |>
  datatable(caption = "The 15 removal vials (14 samples) containing M. magister megalopae, read from the core",
            rownames = FALSE)
```

```{r}
#| label: preview-sorting
dbGetQuery(con, "
  SELECT sorting_status, COUNT(*) AS n_samples,
         COUNT(DISTINCT expedition_orig) AS n_expeditions,
         MIN(date) AS date_min, MAX(date) AS date_max,
         COUNT(grid_key) AS n_in_calcofi_grid
  FROM dungeness_sorting_log GROUP BY 1 ORDER BY 1") |>
  datatable(caption = "dungeness_sorting_log — effort by status",
            options = list(dom = "t"), rownames = FALSE)
```

```{r}
#| label: preview-specimen
dbGetQuery(con, "SELECT * FROM dungeness_specimen ORDER BY specimen_num") |>
  datatable(caption = "dungeness_specimen — independent species confirmation",
            options = list(dom = "t"), rownames = FALSE)
```

## Questions for Data Providers

```{r}
#| label: provider-questions

read_csv(here(glue("metadata/{provider}/{dataset}/questions.csv"))) |>
  arrange(factor(priority, c("blocker", "high", "normal")), id) |>
  select(priority, question, context, status) |>
  datatable(caption = "Questions (ranked)", options = list(dom = "t", pageLength = 10),
            rownames = FALSE)
```

## Write Outputs + Upload

```{r}
#| label: outputs

dir_create(dir_parquet)

# THE CORE, AND NOTHING ELSE (plus the two shared registries). No source table is
# published: the source workbooks are archived, so a per-dataset parquet copy of
# them is redundant, and consumers must read the core rather than choosing between
# two representations of the same data. The compat VIEWs are views over the core,
# so writing them would duplicate the same bytes under a second name.
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
cat(glue("publishing {length(tbls_out)} tables: {paste(tbls_out, collapse = ', ')}"), "\n")

write_parquet_outputs(
  con = con, output_dir = dir_parquet, tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE)

# prune parquet this ingest no longer publishes. Moving to the core retired
# dungeness_measurement.parquet / dungeness_carapace.parquet (now compat VIEWs),
# and a stale file left on disk is still picked up by directory scans and by
# sync_to_gcs — so the output dir must match the manifest exactly.
stale <- setdiff(
  tools::file_path_sans_ext(basename(
    list.files(dir_parquet, pattern = "[.]parquet$"))), tbls_out)
if (length(stale)) {
  file_delete(file.path(dir_parquet, paste0(stale, ".parquet")))
  cat(glue("pruned {length(stale)} stale parquet: {paste(stale, collapse = ', ')}"), "\n")
}

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)

d_tbls_rd <- read_csv(here(glue("metadata/{provider}/{dataset}/tbls_redefine.csv")))
build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  # shared core descriptions first, this dataset's overrides second — without
  # core_dictionary.csv every core table/column ships with an empty description
  metadata_derived_csv = c(
    here("metadata/core_dictionary.csv"),
    here(glue("metadata/{provider}/{dataset}/metadata_derived.csv"))),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned)

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"),
              bucket = "calcofi-db")
} else {
  cat("publish_to_gcs is FALSE — parquet kept local, NOT synced to",
      glue("gs://calcofi-db/ingest/{dir_label}"), "\n")
}
```

```{r}
#| label: metadata-gaps

# empty descriptions ship verbatim to consumers, so surface them as TODOs.
# Scoped to every table this ingest PUBLISHES (core + per-dataset) except the two
# shared registries `dataset` and `measurement_type`, which carry no column
# descriptions in any ingest and would be constant noise hiding the real gaps.
m   <- fromJSON(file.path(dir_parquet, "metadata.json"), simplifyVector = FALSE)
gap <- function(x, f) names(Filter(function(e) !nzchar(e[[f]] %||% ""), x))
checked <- setdiff(tbls_out, c("dataset", "measurement_type"))
own <- function(keys) keys[sub("[.].*$", "", keys) %in% checked]

cat(glue("metadata.json gaps in owned tables — tables missing description: ",
         "{length(own(gap(m$tables, 'description_md')))}; ",
         "columns missing description: ",
         "{length(own(gap(m$columns, 'description_md')))}"), "\n")
if (length(own(gap(m$columns, "description_md"))))
  cat("  ", paste(own(gap(m$columns, "description_md")), collapse = ", "), "\n")
```

## Not Yet Done

This is a first pass. What remains, roughly in order:

1. **Answer the one remaining blocker**, Q08 — permission to publish, a licence,
   a citation, and whether CDFW or SIO is the citable custodian. It gates both
   the release and the GCS upload. Christy Juhasz (CDFW Marine Region) supplied
   the files and is the live contact; Laura Rogers-Bennett has retired.
   (Q01, standardization, is answered — see the Notes sheet above.)
2. **Review the two modelling choices** in "Emit Core Tables" (Q11): the new
   `sample_type = 'subsample'` vocabulary value, and pointing `parent_sample_key`
   at the matched `swfsc_ichthyo` site occupation.
3. **Decide what the sorting log is for.** As absence/effort data it is the more
   scientifically useful of the two efforts — 2,011 samples over 60 years — but
   only if Q03 settles which historic identifications stand, and Q04 settles
   whether repeated rows are jars.
4. **Parse the free-text identification notes** (Q06) into `obs_attribute` stage
   and length bins.
5. Move `measurement_type_new.csv` into `metadata/measurement_type.csv` and drop
   `in_release: false`, in one change.

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
