---
title: "Ingest Underway CUFES Fish Eggs"
calcofi:
  target_name: ingest_swfsc_cufes
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/swfsc_cufes/manifest.json
  provider: swfsc
  dataset: cufes
  workflow_url: https://calcofi.io/workflows/ingest_swfsc_cufes.html
  questions_file: metadata/swfsc/cufes/questions.csv
  dataset_meta:
    dataset_name: CalCOFI Underway CUFES Fish Eggs
    description: >
      Continuous Underway Fish Egg Sampler (CUFES) egg counts (sardine, anchovy,
      jack mackerel, hake, squid, other) with underway environmental conditions,
      from CalCOFI cruises (1996-present). Source: NOAA CoastWatch ERDDAP
      erdCalCOFIcufes.
    citation_main: ""
    link_calcofi_org: ""
    link_data_source: "https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIcufes.html"
    link_others: []
    coverage_temporal: 1996 to present
    coverage_spatial: "CalCOFI region (underway)"
    license: ""
    pi_names: Noelle Bowlin
  # publishes the consolidated core: the underway sample/measurement source shape
  # is wrangled in the notebook and projected into sample / obs + the taxa refs.
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (underway grain)"}
    - {table: obs, shared: true, note: "core occurrence headline (bio, egg counts by species)"}
    - {table: taxon, shared: true, note: "shared taxa reference"}
    - {table: dataset_taxon, shared: true, note: "egg-type -> taxon_key crosswalk"}
    - {table: measurement_type, shared: true}
  erd:
    color: "#f0e8bb"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: NOAA CoastWatch ERDDAP [`erdCalCOFIcufes`](https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIcufes.html)
— Continuous Underway Fish Egg Sampler egg counts (sardine/anchovy/jack mackerel/
hake/squid/other) + underway environment. Provider `swfsc` (Noelle Bowlin), 1996-present.

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, jsonlite, knitr, lubridate, purrr,
  readr, sf, stringr, tibble, tidyr, units, quiet = T)
options(readr.show_col_types = F); options(DT.options = list(scrollX = TRUE))
source(here("libs/ingest.R"))

cc <- read_calcofi_meta(here("ingest_swfsc_cufes.qmd"))
provider <- cc$provider; dataset <- cc$dataset; tables_owned <- cc$tables_owned
dir_label <- glue("{provider}_{dataset}")
dir_parquet <- here(glue("data/parquet/{dir_label}"))
db_path <- here(glue("data/wrangling/{dir_label}.duckdb"))
if (overwrite && file_exists(db_path)) file_delete(db_path)
dir_create(dirname(db_path)); con <- get_duckdb_con(db_path); load_duckdb_extension(con, "spatial")
meas_type_csv <- here("metadata/measurement_type.csv"); d_meas_type <- read_measurement_type(meas_type_csv)
```

## Download from ERDDAP

```{r}
#| label: download

dir_dl  <- here(glue("data/cache/{dir_label}")); dir_create(dir_dl)
cufes_csv <- file.path(dir_dl, "erdCalCOFIcufes.csv")
erddap_url <- "https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIcufes.csv"
if (overwrite_all || !file_exists(cufes_csv))
  download.file(erddap_url, cufes_csv, quiet = TRUE)
# ERDDAP CSV: row 1 = column names, row 2 = units, rows 3+ = data
hdr   <- read_csv(cufes_csv, n_max = 0) |> names()
d_raw <- read_csv(cufes_csv, skip = 2, col_names = hdr)
cat(glue("Downloaded {format(nrow(d_raw), big.mark=',')} CUFES samples"), "\n")
sync_to_gcs(local_dir = dir_dl, gcs_prefix = glue("archive/{provider}/{dataset}"),
            bucket = "calcofi-files-public", exclude = c(".DS_Store"))
```

## Build Sample + Measurement Tables

```{r}
#| label: build

d <- d_raw |>
  mutate(sample_id = row_number(), .before = 1) |>
  mutate(
    datetime_start_utc = as_datetime(time),
    datetime_end_utc   = suppressWarnings(as_datetime(stop_time)),
    across(c(latitude, longitude, stop_latitude, stop_longitude,
             start_temperature, start_salinity, start_wind_speed, start_wind_direction, start_pump_speed,
             stop_temperature, stop_salinity, stop_wind_speed, stop_wind_direction, stop_pump_speed),
           ~ suppressWarnings(as.numeric(.x))))

d_sample <- d |>
  transmute(
    sample_id, cruise_orig = cruise, ship_name = ship, ship_code,
    sample_number = suppressWarnings(as.integer(sample_number)),
    datetime_start_utc, datetime_end_utc, latitude, longitude,
    latitude_stop = stop_latitude, longitude_stop = stop_longitude,
    start_temperature, start_salinity, start_wind_speed, start_wind_direction, start_pump_speed,
    stop_temperature, stop_salinity, stop_wind_speed, stop_wind_direction, stop_pump_speed)
dbWriteTable(con, "cufes_sample", d_sample, overwrite = TRUE)

# pivot egg counts -> long measurement
egg_cols <- c("sardine_eggs","anchovy_eggs","jack_mackerel_eggs","hake_eggs","squid_eggs","other_fish_eggs")
d_meas <- d |>
  select(sample_id, all_of(egg_cols)) |>
  pivot_longer(all_of(egg_cols), names_to = "measurement_type", values_to = "measurement_value") |>
  filter(!is.na(measurement_value)) |>
  mutate(measurement_value = as.double(measurement_value), measurement_qual = NA_character_) |>
  mutate(cufes_measurement_id = row_number(), .before = 1)
dbWriteTable(con, "cufes_measurement", d_meas, overwrite = TRUE)
cat(glue("cufes_sample {nrow(d_sample)}, cufes_measurement {nrow(d_meas)}"), "\n")
```

## Resolve Keys + Spatial

```{r}
#| label: keys-spatial

load_prior_tables(con, parquet_dir = here("data/parquet/swfsc_ichthyo"),
                  tables = c("ship","cruise","grid"), geom_tables = c("grid"), as_view = TRUE)
d_ship <- dbGetQuery(con, "SELECT ship_key, ship_name, ship_nodc FROM ship") |>
  mutate(ship_name_norm = ship_name |> str_to_upper() |> str_squish())
zt <- dbGetQuery(con, "SELECT sample_id, ship_name, datetime_start_utc FROM cufes_sample") |>
  mutate(ship_name_norm = ship_name |> str_replace("^R/?V\\.?\\s+","") |> str_to_upper() |> str_squish()) |>
  left_join(d_ship |> select(ship_name_norm, ship_key, ship_nodc), by = "ship_name_norm") |>
  mutate(cruise_key = if_else(is.na(ship_nodc) | is.na(datetime_start_utc), NA_character_,
                              as.character(glue("{format(datetime_start_utc,'%Y-%m')}-{ship_nodc}"))))
valid_ck <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM cruise")$cruise_key
zt <- zt |> mutate(cruise_key = if_else(cruise_key %in% valid_ck, cruise_key, NA_character_))
dbWriteTable(con, "zk", zt |> select(sample_id, ship_key, cruise_key), overwrite = TRUE)
dbExecute(con, "ALTER TABLE cufes_sample ADD COLUMN IF NOT EXISTS ship_key VARCHAR")
dbExecute(con, "ALTER TABLE cufes_sample ADD COLUMN IF NOT EXISTS cruise_key VARCHAR")
dbExecute(con, "UPDATE cufes_sample s SET ship_key=z.ship_key, cruise_key=z.cruise_key FROM zk z WHERE s.sample_id=z.sample_id")
dbExecute(con, "DROP TABLE zk")
cat(glue("ship match {sum(!is.na(zt$ship_key))}/{nrow(zt)}; cruise_key {sum(!is.na(zt$cruise_key))}/{nrow(zt)}"), "\n")

add_point_geom(con, "cufes_sample", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "cufes_sample") |> datatable(caption = "Grid assignment")
```

## Add Measurement Types

```{r}
#| label: meas-types

cufes_types <- tibble(
  measurement_type = egg_cols,
  description = c("Sardine egg count","Northern anchovy egg count","Jack mackerel egg count",
                  "Pacific hake egg count","Squid egg count","Other fish egg count"),
  units = "count", `_source_column` = egg_cols, `_source_table` = "cufes_measurement",
  `_source_datasets` = "swfsc_cufes", `_qual_column` = NA_character_, `_prec_column` = NA_character_)
new_types <- cufes_types |> filter(!measurement_type %in% d_meas_type$measurement_type)
if (nrow(new_types) > 0) { d_meas_type <- bind_rows(d_meas_type, new_types); write_csv(d_meas_type, meas_type_csv, na = "") }
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
cat(glue("added {nrow(new_types)} measurement types"), "\n")
```

## Metadata, Schema, Validate, Outputs

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`) via `calcofi4db::emit_core_tables()`. These
core tables **are** this ingest's output: `release_database.qmd` concatenates the
per-dataset shards rather than re-deriving the core, so there is exactly one
projection to keep correct.

CUFES bakes the taxon into the measurement type name (`sardine_eggs`,
`anchovy_eggs`, …). `metadata/measurement_taxon.csv` decomposes each raw type
into a real `taxon_key`, the canonical `measurement_type`, and `life_stage`, so
the egg counts land in `obs` as ordinary taxon-resolved occurrences.

```{r}
#| label: emit_core

ds_key <- "swfsc_cufes"
# filter the crosswalk to THIS dataset: emit_core_tables() did that internally, so
# an unfiltered read leaks other datasets' taxa into this shard
mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

# This projection lives here, in the notebook that owns the dataset, not in a
# switch(dataset_key, ...) arm inside calcofi4db. The reusable SHAPES stay in the
# package (sample_arm_self / compat_measurement_sql), so this is a declaration.
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- build_dataset_taxon(con,   mt_taxon, tx_over)
dbWriteTable(con, "_measurement_taxon", as.data.frame(mt_taxon), overwrite = TRUE)

append_sample(con, sample_arm_self(ds_key, "cufes_sample", "sample_id", "underway"))

append_obs(con, glue("
  SELECT 'bio', '{ds_key}', {ns_key(ds_key, 'underway', 'c.sample_id')},
         c.grid_key, c.cruise_key, c.latitude, c.longitude,
         CAST(c.datetime_start_utc AS TIMESTAMP), 0::DOUBLE, 0::DOUBLE,
         mx.taxon_key, mx.life_stage, mx.measurement_type, m.measurement_value,
         m.measurement_qual, NULL::DOUBLE
  FROM cufes_measurement m JOIN cufes_sample c USING (sample_id)
  JOIN _measurement_taxon mx ON mx.dataset_key = '{ds_key}'
                            AND mx.raw_measurement_type = m.measurement_type
                            AND mx.target = 'obs'
  WHERE c.grid_key IS NOT NULL"))

core <- list(
  sample        = dbGetQuery(con, "SELECT COUNT(*) FROM sample")[[1]],
  obs           = dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]],
  taxon         = n_taxon,
  dataset_taxon = n_ds_taxon)
cat(glue(
  "core projection — sample={core$sample %||% 0} obs={core$obs %||% 0} ",
  "taxon={core$taxon %||% 0} dataset_taxon={core$dataset_taxon %||% 0}\n"))

# every grid-matched measurement whose raw type is registered must reach obs,
# with a real taxon_key — an unregistered raw type is dropped by the INNER join,
# so assert the registry actually covers the vocabulary in the data
n_obs <- dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]]
n_exp <- dbGetQuery(con, "
  SELECT COUNT(*) FROM cufes_measurement m JOIN cufes_sample c USING (sample_id)
  JOIN _measurement_taxon mx ON mx.raw_measurement_type = m.measurement_type
                            AND mx.target = 'obs'
  WHERE c.grid_key IS NOT NULL")[[1]]
n_unreg <- dbGetQuery(con, "
  SELECT COUNT(DISTINCT m.measurement_type) FROM cufes_measurement m
  WHERE m.measurement_type NOT IN (
    SELECT raw_measurement_type FROM _measurement_taxon)")[[1]]
stopifnot(
  "obs must be one row per registered grid-matched measurement" = n_obs == n_exp,
  "every cufes measurement_type must be in measurement_taxon.csv" = n_unreg == 0,
  "cufes obs must all carry a taxon_key" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs WHERE taxon_key IS NULL")[[1]] == 0)
cat(glue("obs parity: {format(n_obs, big.mark = ',')} rows, all taxon-resolved"), "\n")
# serve the retired per-dataset table names as VIEWs over the core, so
# in-notebook consumers and ad-hoc queries keep working against the old names
# (exact for every column the core models; see ?create_compat_views)
invisible(dbExecute(con, "DROP TABLE IF EXISTS cufes_measurement_src"))
invisible(dbExecute(con, "ALTER TABLE cufes_measurement RENAME TO cufes_measurement_src"))
invisible(dbExecute(con, glue(
  "CREATE OR REPLACE VIEW cufes_measurement AS
   {compat_measurement_sql(ds_key, 'underway', 'sample_id', 'cufes_measurement_id')}")))
cat(glue("compat view cufes_measurement over obs: ",
         "{dbGetQuery(con, 'SELECT COUNT(*) FROM cufes_measurement')[[1]]} rows"), "\n")

```


```{r}
#| label: finalize

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)

cufes_rels <- list(
  primary_keys = list(cufes_sample = "sample_id", cufes_measurement = "cufes_measurement_id",
                      measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table="cufes_measurement", column="sample_id", ref_table="cufes_sample", ref_column="sample_id"),
    list(table="cufes_measurement", column="measurement_type", ref_table="measurement_type", ref_column="measurement_type")))
# the SOURCE shape, documenting the wrangling above; the published tables are
# the consolidated core, so relationships.json comes from core_relationships()
cc_erd(con, tables = c("cufes_sample","cufes_measurement","measurement_type","dataset"), rels = cufes_rels,
       colors = list(lightblue = c("cufes_sample","cufes_measurement"), lightyellow = "measurement_type", white = "dataset"))

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")

dir_create(dir_parquet)
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
write_parquet_outputs(con = con, output_dir = dir_parquet,
  tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")), strip_provenance = FALSE)
build_relationships_json(rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)
d_tbls_rd <- read_csv(here("metadata/swfsc/cufes/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/swfsc/cufes/flds_redefine.csv"))
build_metadata_json(con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           here("metadata/swfsc/cufes/metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned)
sync_to_gcs(local_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
```

## Questions for Data Providers

```{r}
#| label: provider-questions
read_csv(here(glue("metadata/{provider}/{dataset}/questions.csv"))) |>
  arrange(factor(priority, c("blocker","high","normal")), id) |>
  select(priority, question, context, status) |>
  datatable(caption = "Questions (ranked)", options = list(dom="t"), rownames = FALSE)
```

```{r}
#| label: cleanup
close_duckdb(con); cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
