---
title: "Ingest SWFSC Invertebrates"
calcofi:
  target_name: ingest_swfsc_invert
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/swfsc_invert/manifest.json
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: NOAA CoastWatch ERDDAP

- [erdCalCOFIinvcnt](https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIinvcnt.html) — invertebrate counts
- [erdCalCOFIinvsiz](https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIinvsiz.html) — invertebrate sizes

Counts and sizes of invertebrates (primarily squids) captured in CalCOFI
ichthyoplankton nets, processed by SWFSC. The two ERDDAP datasets share
the same sampling metadata columns; `invcnt` adds count/abundance fields,
`invsiz` adds a size measurement field.

This ingest creates two output tables:

- **`invert_net`** — unique sampling events (xyt: time, location, net metadata)
- **`invert`** — measurements in long format (count, standardized counts, size)
  with FK to `invert_net` and species identifiers

- **Provider**: `swfsc`
- **Coverage**: 1978–2023 (counts), 1980–2023 (sizes)
- **Spatial**: Pacific coast, 0–51°N

### Data Flow

```{mermaid}
graph LR
  A[ERDDAP] --> B[CSV on Google Drive]
  B --> C[DuckDB Wrangling]
  C --> D[Load Prior ichthyo Tables]
  D --> E[Build cruise_key + invert_net]
  E --> F[Pivot to long-format invert]
  F --> G[Parquet Export]
  G --> H[GCS Archive]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI / calcofi4db,
  CalCOFI / calcofi4r,
  DBI,
  dplyr,
  DT,
  fs,
  glue,
  here,
  janitor,
  jsonlite,
  readr,
  rerddap,
  sf,
  stringr,
  tibble,
  tidyr,
  quiet = T
)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# common ingest settings (overwrite, dir_data)
source(here("libs/ingest.R"))

# define paths
dataset_name <- "SWFSC CalCOFI Invertebrates"
provider <- "swfsc"
dataset <- "invert"
dir_label <- glue("{provider}_{dataset}")
dir_csv <- path_expand(glue("{dir_data}/{provider}/{dataset}"))
dir_parquet <- here(glue("data/parquet/{dir_label}"))
db_path <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path)) {
    file_delete(db_path)
  }
  if (dir_exists(dir_parquet)) dir_delete(dir_parquet)
}

dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Download from ERDDAP

Fetch both datasets and save as CSVs in Google Drive. Skips download if
CSV already exists (delete to re-download).

```{r}
#| label: download_erddap

erddap_url <- "https://coastwatch.pfeg.noaa.gov/erddap"
dir_create(dir_csv)

erddap_datasets <- list(
  erdCalCOFIinvcnt = "erdCalCOFIinvcnt",
  erdCalCOFIinvsiz = "erdCalCOFIinvsiz"
)

for (ds_name in names(erddap_datasets)) {
  csv_path <- path(dir_csv, glue("{ds_name}.csv"))

  if (!file_exists(csv_path)) {
    message(glue("downloading {ds_name} from ERDDAP..."))
    ds_info <- rerddap::info(erddap_datasets[[ds_name]], url = erddap_url)
    d_raw <- rerddap::tabledap(ds_info)
    write_csv(as_tibble(d_raw), csv_path)
    message(glue("  wrote {nrow(d_raw)} rows to {csv_path}"))
  } else {
    message(glue("  {ds_name}: CSV exists, skipping download"))
  }
}
```

## Read CSVs

```{r}
#| label: read_csvs

d <- read_csv_files(
  provider = provider,
  dataset = dataset,
  dir_data = dir_data,
  sync_archive = TRUE,
  metadata_dir = here("metadata")
)

message(glue("loaded {nrow(d$source_files)} tables from {d$paths$dir_csv}"))
message(glue("total rows: {sum(d$source_files$nrow)}"))
```

## Check Data Integrity

```{r}
#| label: data_integrity
#| output: asis

# type_exceptions = "all": ERDDAP type names (String, int, float) don't
# match R column classes (character, integer, numeric) — accept all
integrity <- check_data_integrity(
  d = d,
  dataset_name = dataset_name,
  halt_on_fail = TRUE,
  type_exceptions = "all"
)

render_integrity_message(integrity)
```

## Source Files

```{r}
#| label: source_files

show_source_files(d)
```

## Tables & Fields

```{r}
#| label: show_redefine

show_tables_redefine(d)
show_fields_redefine(d)
```

## Load into Database

```{r}
#| label: load_to_db

tbl_stats <- ingest_dataset(
  con = con,
  d = d,
  mode = if (overwrite) "replace" else "append",
  verbose = TRUE
)

tbl_stats |>
  datatable(rownames = FALSE, filter = "top")
```

## Load Dataset Metadata

```{r}
#| label: load_dataset_metadata

d_dataset <- read_csv(here("metadata/dataset.csv"))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
```

## Load Prior Tables

Load reference tables from ichthyo ingest for cross-referencing cruise_key.

```{r}
#| label: load_prior

prior <- load_prior_tables(
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  con         = con,
  tables      = c("ship", "cruise"),
  as_view     = TRUE
)

message(glue(
  "loaded {nrow(prior)} prior tables: {paste(prior$table, collapse=', ')}"
))

prior |>
  datatable(caption = "Prior tables loaded from ichthyo ingest")
```

## Build cruise_key

The ERDDAP `cruise` column is YYYYMM format (e.g., 197804 = April 1978)
and `ship_code` is the 2-letter ship key (e.g., "JD"). Combine with the
ship table's `ship_nodc` to create `cruise_key` in YYYY-MM-NODC format
for matching to the ichthyo cruise hierarchy.

```{r}
#| label: build_cruise_key

# add cruise_key to both source tables
for (tbl in c("invert_count", "invert_size")) {
  dbExecute(con, glue(
    "ALTER TABLE {tbl} ADD COLUMN IF NOT EXISTS cruise_key VARCHAR"))
  dbExecute(con, glue("
    UPDATE {tbl} AS ic
    SET cruise_key = CONCAT(
      LEFT(CAST(ic.cruise AS VARCHAR), 4), '-',
      SUBSTR(CAST(ic.cruise AS VARCHAR), 5, 2), '-',
      s.ship_nodc)
    FROM ship s
    WHERE s.ship_key = ic.ship_code"))
}

# report match rates
for (tbl in c("invert_count", "invert_size")) {
  n_total <- dbGetQuery(con, glue("SELECT COUNT(*) FROM {tbl}"))[[1]]
  n_match <- dbGetQuery(con, glue(
    "SELECT COUNT(*) FROM {tbl} WHERE cruise_key IS NOT NULL"))[[1]]
  n_cruise <- dbGetQuery(con, glue("
    SELECT COUNT(*) FROM {tbl}
    WHERE cruise_key IN (SELECT cruise_key FROM cruise)"))[[1]]
  cat(glue("{tbl}: {n_match}/{n_total} rows got cruise_key, ",
           "{n_cruise} matched to ichthyo cruise table"), "\n")
}
```

## Create `invert_net` — sampling events

Extract unique sampling events (xyt) from the denormalized ERDDAP data.
Each unique combination of cruise + tow + net metadata defines one
sampling event.

```{r}
#| label: create_invert_net

dbExecute(con, "
  CREATE OR REPLACE TABLE invert_net AS
  SELECT DISTINCT
    cruise_key,
    CAST(cruise AS VARCHAR)  AS cruise_erddap,
    ship_code,
    order_occupied,
    tow_type,
    tow_number,
    net_location,
    time_start,
    latitude,
    longitude,
    line,
    station,
    standard_haul_factor,
    volume_sampled,
    proportion_sorted
  FROM invert_count
  WHERE cruise_key IS NOT NULL
  ORDER BY cruise_key, order_occupied, tow_number")

# assign sequential id
assign_sequential_ids(
  con        = con,
  table_name = "invert_net",
  id_col     = "invert_net_id",
  sort_cols  = c("cruise_key", "order_occupied", "tow_number")
)

n_net <- dbGetQuery(con, "SELECT COUNT(*) FROM invert_net")[[1]]
cat(glue("invert_net: {format(n_net, big.mark = ',')} unique sampling events"), "\n")

dbGetQuery(con, "SELECT * FROM invert_net LIMIT 10") |>
  datatable(caption = "invert_net — first 10 sampling events")
```

## Create `invert` — measurements in long format

Pivot count and size measurements into a single long-format table,
similar to how ichthyo stores egg/larva measurements. Each row is one
measurement for one species at one sampling event.

```{r}
#| label: create_invert

# -- pivot invert_count: 3 measurement types per row ----
# join to invert_net to get invert_net_id
sql_count <- "
  SELECT
    n.invert_net_id,
    ic.scientific_name,
    ic.common_name,
    ic.itis_tsn,
    ic.calcofi_species_code,
    mt.measurement_type,
    mt.measurement_value
  FROM invert_count ic
  JOIN invert_net n
    ON  ic.cruise_key      = n.cruise_key
    AND ic.order_occupied   = n.order_occupied
    AND ic.tow_type         = n.tow_type
    AND ic.tow_number       = n.tow_number
    AND ic.net_location IS NOT DISTINCT FROM n.net_location
  CROSS JOIN LATERAL (VALUES
    ('count',           CAST(ic.invert_count  AS DOUBLE)),
    ('count_per_10m2',  ic.inverts_10m2),
    ('count_per_100m3', ic.inverts_100m3)
  ) AS mt(measurement_type, measurement_value)
  WHERE ic.cruise_key IS NOT NULL
    AND mt.measurement_value IS NOT NULL"

# -- pivot invert_size: size measurement ----
sql_size <- "
  SELECT
    n.invert_net_id,
    isz.scientific_name,
    isz.common_name,
    isz.itis_tsn,
    isz.calcofi_species_code,
    'size' AS measurement_type,
    isz.invert_size AS measurement_value
  FROM invert_size isz
  JOIN invert_net n
    ON  isz.cruise_key      = n.cruise_key
    AND isz.order_occupied   = n.order_occupied
    AND isz.tow_type         = n.tow_type
    AND isz.tow_number       = n.tow_number
    AND isz.net_location IS NOT DISTINCT FROM n.net_location
  WHERE isz.cruise_key IS NOT NULL
    AND isz.invert_size IS NOT NULL"

# combine and assign sequential id
dbExecute(con, glue("
  CREATE OR REPLACE TABLE invert AS
  SELECT
    ROW_NUMBER() OVER (
      ORDER BY invert_net_id, scientific_name, measurement_type, measurement_value
    ) AS invert_id,
    *
  FROM (
    {sql_count}
    UNION ALL
    {sql_size}
  ) sub"))

n_invert <- dbGetQuery(con, "SELECT COUNT(*) FROM invert")[[1]]
n_by_type <- dbGetQuery(con, "
  SELECT measurement_type, COUNT(*) AS n
  FROM invert
  GROUP BY measurement_type
  ORDER BY measurement_type")

cat(glue("invert: {format(n_invert, big.mark = ',')} total measurement rows"), "\n")
n_by_type |>
  datatable(caption = "Measurement rows by type")
```

## Data Preview

```{r}
#| label: preview_tables

dbGetQuery(con, "SELECT * FROM invert_net LIMIT 100") |>
  datatable(
    caption = "invert_net (first 100 rows)",
    rownames = FALSE, filter = "top")

dbGetQuery(con, "SELECT * FROM invert LIMIT 100") |>
  datatable(
    caption = "invert (first 100 rows)",
    rownames = FALSE, filter = "top")
```

## Schema Documentation

```{r}
#| label: erd

invert_rels <- list(
  primary_keys = list(
    invert_net = "invert_net_id",
    invert     = "invert_id",
    cruise     = "cruise_key",
    ship       = "ship_key"),
  foreign_keys = list(
    list(table = "invert",     column = "invert_net_id",
         ref_table = "invert_net", ref_column = "invert_net_id"),
    list(table = "invert_net", column = "cruise_key",
         ref_table = "cruise",     ref_column = "cruise_key"),
    list(table = "cruise",     column = "ship_key",
         ref_table = "ship",       ref_column = "ship_key")))

cc_erd(
  con,
  tables = c("invert", "invert_net", "cruise", "ship"),
  rels   = invert_rels,
  colors = list(
    lightblue   = c("invert", "invert_net"),
    lightyellow = c("cruise", "ship")))
```

## Write Parquet

```{r}
#| label: write_parquet

# export only tables this ingest creates (not dependency VIEWs)
own_tables <- c("invert_net", "invert", "dataset")

parquet_stats <- write_parquet_outputs(
  con              = con,
  output_dir       = dir_parquet,
  tables           = own_tables,
  sort_by          = list(
    invert = c("measurement_type", "scientific_name")),
  strip_provenance = FALSE
)

parquet_stats |>
  mutate(file = basename(path)) |>
  select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write_metadata

metadata_path <- build_metadata_json(
  con = con,
  d_tbls_rd = d$d_tbls_rd,
  d_flds_rd = d$d_flds_rd,
  output_dir = dir_parquet,
  tables = own_tables,
  set_comments = TRUE,
  provider = provider,
  dataset = dataset,
  workflow_url = glue(
    "https://calcofi.io/workflows/ingest_swfsc_invert.html"
  )
)

# write relationships.json
build_relationships_json(
  rels       = invert_rels,
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset
)
```

```{r}
#| label: show_metadata_json

listviewer::jsonedit(
  jsonlite::fromJSON(metadata_path, simplifyVector = FALSE),
  mode = "view")
```

```{r}
#| label: show_relationships_json

listviewer::jsonedit(
  jsonlite::fromJSON(
    file.path(dir_parquet, "relationships.json"),
    simplifyVector = FALSE),
  mode = "view")
```

## Upload to GCS

```{r}
#| label: upload_gcs

gcs_ingest_prefix <- glue("ingest/{dir_label}")
gcs_bucket <- "calcofi-db"

sync_results <- sync_to_gcs(
  local_dir = dir_parquet,
  gcs_prefix = gcs_ingest_prefix,
  bucket = gcs_bucket
)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
message("local wrangling database connection closed")
message(glue("parquet outputs written to: {dir_parquet}"))
message(glue("GCS outputs at: gs://{gcs_bucket}/{gcs_ingest_prefix}/"))
```
