---
title: "Ingest DIC"
calcofi:
  target_name: ingest_calcofi_dic
  workflow_type: ingest
  dependency:
    - ingest_calcofi_bottle
  output: data/parquet/calcofi_dic/manifest.json
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: [DIC | CalCOFI.org](https://calcofi.org/data/oceanographic-data/dic/)

Dissolved Inorganic Carbon (DIC) and Total Alkalinity (TA) from CalCOFI
Niskin bottle samples, archived at NCEI accession
[0301029](https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.nodc:0301029).

- **Provider**: `calcofi` (CalCOFI program; PI: Todd Martz, Scripps)
- **File**: `CALCOFI_DIC_20250122.csv` (~500KB, 4,391 rows)
- **Coverage**: 1983–2021, 71 stations, 131 cruises
- **Measurements**: DIC (umol/kg), TA (umol/kg), CTD temperature (degC),
  salinity (PSS-78)
- **Quality flags**: WOCE convention (2=good, 3=questionable, 4=bad, 9=missing)
- **Citation**: Keeling, C.D. et al. (2025). Discrete profile dissolved
  inorganic carbon, total alkalinity, water temperature and salinity
  measurements ... for the CalCOFI program in the North East Pacific
  Ocean Coastal area from 1983-03-19 to 2021-07-20 (NCEI Accession
  0301029). NOAA NCEI. <https://doi.org/10.25921/3w9f-jd72>

**Strategy**: Load source CSV, pivot ALL four measurement columns
(DIC, TA, CTD temperature, salinity) into tidy long-format
`dic_measurement`, then summarize replicates into
`dic_measurement_summary`. The base `dic_sample` table retains only
position/time/FK columns — no measurement values.

### Data Flow

```{mermaid}
graph LR
  A[NCEI CSV] --> B[DuckDB Wrangling]
  B --> C[Parse Station + Date]
  C --> D[Match to Casts]
  D --> E[Pivot ALL → long]
  E --> F[Summarize Replicates]
  F --> G[Parquet Export]
  G --> H[GCS Archive]
```

### Measurement Types

Four measurement types from this dataset — all **new** types, distinct
from bottle/CTD equivalents because these come from a separate
instrument chain with WOCE QC:

| measurement_type | description | units | source column | status |
|-----------------|-------------|-------|---------------|--------|
| `dic` | Dissolved inorganic carbon | umol/kg | DIC | **NEW** |
| `alkalinity` | Total alkalinity | umol/kg | TA | **NEW** |
| `ctdtemp_its90` | CTD temperature (ITS-90) | degC | CTDTEMP_ITS90 | **NEW** |
| `salinity_pss78` | Practical salinity (PSS-78) | PSS-78 | Salinity_PSS78 | **NEW** |

These are kept separate from the calcofi_bottle `temperature`/`salinity`
types because the DIC dataset values come from a different QC pipeline
(WOCE flags vs calcofi.org quality codes).

## Setup

```{r}
#| label: setup

# devtools::install_local(here::here("../calcofi4db"), force = T)
devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI / calcofi4db,
  CalCOFI / calcofi4r,
  DBI,
  dplyr,
  DT,
  fs,
  glue,
  googledrive,
  here,
  htmltools,
  janitor,
  jsonlite,
  knitr,
  lubridate,
  purrr,
  readr,
  sf,
  stringr,
  tibble,
  tidyr,
  units,
  quiet = T
)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# define paths
dataset_name <- "CalCOFI DIC"
provider     <- "calcofi"
dataset      <- "dic"
dir_label    <- glue("{provider}_{dataset}")
dir_data     <- "~/My Drive/projects/calcofi/data-public"
overwrite    <- TRUE

dir_parquet <- here(glue("data/parquet/{dir_label}"))
db_path     <- here(glue("data/wrangling/{dir_label}.duckdb"))

# clean slate if overwrite
if (overwrite) {
  if (file_exists(db_path))    file_delete(db_path)
  if (dir_exists(dir_parquet)) dir_delete(dir_parquet)
}

dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

# load unified measurement_type reference
meas_type_csv <- here("metadata/measurement_type.csv")
d_meas_type   <- read_csv(meas_type_csv)
```

## Read Source Data

The NCEI CSV has a units row (row 2) that must be skipped, and uses
`-999` as the missing value sentinel.

```{r}
#| label: read-source

dic_csv <- glue(
  "{dir_data}/calcofi/dic/0301029/1.1/data/0-data/",
  "CALCOFI_DIC_20250122.csv"
)
stopifnot("DIC CSV not found" = file_exists(dic_csv))

# read, skipping the units row (row 2)
d_raw <- read_csv(dic_csv, skip = 2, col_names = FALSE)

# get header from row 1
hdr <- read_csv(dic_csv, n_max = 0) |> names()
names(d_raw) <- hdr

cat(glue("Read {nrow(d_raw)} rows, {ncol(d_raw)} columns from {basename(dic_csv)}"), "\n")
```

## Clean and Type-Cast

Replace `-999` with NA, convert character columns to proper types, and
build a datetime column.

```{r}
#| label: clean-data

d_clean <- d_raw |>
  mutate(
    # convert numeric columns stored as character (due to units row)
    Latitude      = as.numeric(Latitude),
    Longitude     = as.numeric(Longitude),
    Depth         = as.numeric(Depth),
    CTDTEMP_ITS90 = as.numeric(CTDTEMP_ITS90),
    DIC           = as.numeric(DIC),
    TA            = as.numeric(TA),

    # replace -999 sentinels with NA across all numeric columns
    across(
      where(is.numeric),
      ~ if_else(.x == -999, NA_real_, .x)),

    # build datetime from components
    datetime_utc = make_datetime(
      Year_UTC, Month_UTC, Day_UTC,
      tz = "UTC"),

    # site_key matching calcofi convention: "090.0 062.0"
    site_key = Station_ID
  )

# summary of NA replacement
na_counts <- d_clean |>
  summarise(across(where(is.numeric), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_na") |>
  filter(n_na > 0)
na_counts |> datatable(caption = "NULL counts after -999 replacement")
```

## Load into Database

Load `dic_sample` as a **tidy position table** — only station, time,
depth, and FK columns. Measurement values (DIC, TA, temp, salinity) are
pivoted into `dic_measurement` in the next step.

```{r}
#| label: load-to-db

# tidy: dic_sample has position/time only, no measurement values
d_dic <- d_clean |>
  transmute(
    expocode     = EXPOCODE,
    ship_name    = Ship_Name,
    datetime_utc = datetime_utc,
    site_key      = site_key,
    latitude     = Latitude,
    longitude    = Longitude,
    depth_m      = Depth
  )

dbWriteTable(con, "dic_sample", d_dic, overwrite = TRUE)
cat(glue("dic_sample: {nrow(d_dic)} rows, {ncol(d_dic)} columns loaded"), "\n")

# also load the wide-format data temporarily for pivoting
d_wide <- d_clean |>
  transmute(
    expocode     = EXPOCODE,
    site_key      = Station_ID,
    datetime_utc = datetime_utc,
    depth_m      = Depth,
    latitude     = Latitude,
    longitude    = Longitude,
    ctdtemp_its90     = CTDTEMP_ITS90,
    ctdtemp_its90_flag = as.integer(CTDTEMP_flag),
    salinity_pss78     = Salinity_PSS78,
    salinity_pss78_flag = as.integer(Salinity_flag),
    dic           = DIC,
    dic_flag      = as.integer(DIC_flag),
    alkalinity    = TA,
    alkalinity_flag = as.integer(TA_flag)
  )
dbWriteTable(con, "dic_wide", d_wide, overwrite = TRUE)
```

## Match to Existing Casts

Load existing casts from bottle workflow parquet and match DIC samples
to casts using `site_key` + `datetime_utc` (±3 day window).

```{r}
#| label: load-casts

# load casts and bottle from bottle workflow
# include "casts" in geom_tables since it has GEOMETRY columns
load_prior_tables(
  con         = con,
  tables      = c("casts", "bottle"),
  parquet_dir = here("data/parquet/calcofi_bottle"),
  geom_tables = c("grid", "site", "segment", "casts"),
  as_view     = TRUE
)

# station matching
n_dic_sta <- dbGetQuery(
  con, "SELECT COUNT(DISTINCT site_key) FROM dic_sample")[[1]]
n_matched_sta <- dbGetQuery(
  con,
  "SELECT COUNT(DISTINCT d.site_key)
   FROM dic_sample d
   INNER JOIN casts c ON d.site_key = c.site_key")[[1]]
cat(glue(
  "Station matching: {n_matched_sta}/{n_dic_sta} DIC stations found in casts"), "\n")
```

```{r}
#| label: match-casts

# match DIC samples to casts by site_key + date (±3 day window)
# many dates are offset by 1 day between NCEI and calcofi.org
dbExecute(con, "ALTER TABLE dic_sample ADD COLUMN IF NOT EXISTS cast_id INTEGER")
dbExecute(
  con,
  "UPDATE dic_sample d
   SET cast_id = (
     SELECT c.cast_id
     FROM casts c
     WHERE c.site_key = d.site_key
       AND ABS(c.datetime_utc::DATE - d.datetime_utc::DATE) <= 3
     ORDER BY ABS(c.datetime_utc::DATE - d.datetime_utc::DATE)
     LIMIT 1
   )"
)

n_matched <- dbGetQuery(con, "SELECT COUNT(*) FROM dic_sample WHERE cast_id IS NOT NULL")[[1]]
n_total   <- dbGetQuery(con, "SELECT COUNT(*) FROM dic_sample")[[1]]
cat(glue(
  "Cast matching: {n_matched}/{n_total} DIC samples matched ",
  "({round(100*n_matched/n_total, 1)}%)"), "\n")

# show unmatched samples
d_unmatched <- dbGetQuery(
  con,
  "SELECT DISTINCT site_key, datetime_utc::DATE AS date, ship_name, expocode
   FROM dic_sample WHERE cast_id IS NULL ORDER BY datetime_utc")
if (nrow(d_unmatched) > 0)
  d_unmatched |> datatable(caption = "Unmatched DIC samples (no matching cast)")
```

## Match to Bottles

For matched casts, find the nearest bottle at the same depth.

```{r}
#| label: match-bottles

dbExecute(con, "ALTER TABLE dic_sample ADD COLUMN IF NOT EXISTS bottle_id INTEGER")
dbExecute(
  con,
  "UPDATE dic_sample d
   SET bottle_id = (
     SELECT b.bottle_id
     FROM bottle b
     WHERE b.cast_id = d.cast_id
       AND ABS(b.depth_m - d.depth_m) <= 1.0
     ORDER BY ABS(b.depth_m - d.depth_m)
     LIMIT 1
   )
   WHERE d.cast_id IS NOT NULL"
)

n_bottle_matched <- dbGetQuery(
  con, "SELECT COUNT(*) FROM dic_sample WHERE bottle_id IS NOT NULL")[[1]]
cat(glue(
  "Bottle matching: {n_bottle_matched}/{n_matched} cast-matched samples ",
  "({round(100*n_bottle_matched/n_matched, 1)}%)"), "\n")

# propagate FKs to dic_wide for pivoting
dbExecute(con, "ALTER TABLE dic_wide ADD COLUMN IF NOT EXISTS cast_id INTEGER")
dbExecute(con, "ALTER TABLE dic_wide ADD COLUMN IF NOT EXISTS bottle_id INTEGER")
dbExecute(
  con,
  "UPDATE dic_wide w SET
     cast_id   = s.cast_id,
     bottle_id = s.bottle_id
   FROM dic_sample s
   WHERE w.expocode     = s.expocode
     AND w.site_key      = s.site_key
     AND w.datetime_utc = s.datetime_utc
     AND w.depth_m IS NOT DISTINCT FROM s.depth_m"
)
```

## Pivot ALL Measurements to Long Format

Pivot all four measurement columns (DIC, TA, CTD temperature, salinity)
into tidy long-format `dic_measurement`. Each row = one measurement at
one position. This follows tidy data principles — no mixing of different
measured quantities on the same row.

```{r}
#| label: pivot-measurements

# pivot all 4 measurement types from wide to long
meas_map <- list(
  list(type = "dic",             col = "dic",             flag = "dic_flag"),
  list(type = "alkalinity",      col = "alkalinity",      flag = "alkalinity_flag"),
  list(type = "ctdtemp_its90",   col = "ctdtemp_its90",   flag = "ctdtemp_its90_flag"),
  list(type = "salinity_pss78",  col = "salinity_pss78",  flag = "salinity_pss78_flag")
)

sql_parts <- map_chr(meas_map, \(m) glue(
  "SELECT expocode, site_key, datetime_utc, depth_m,
          latitude, longitude, cast_id, bottle_id,
          '{m$type}' AS measurement_type,
          CAST({m$col} AS DOUBLE) AS measurement_value,
          CAST({m$flag} AS VARCHAR) AS measurement_qual
   FROM dic_wide
   WHERE {m$col} IS NOT NULL
     AND NOT isnan(CAST({m$col} AS DOUBLE))
     AND isfinite(CAST({m$col} AS DOUBLE))"
))

dbExecute(con, glue(
  "CREATE TABLE dic_measurement AS
   SELECT ROW_NUMBER() OVER (
     ORDER BY site_key, datetime_utc, depth_m, measurement_type
   ) AS dic_measurement_id, *
   FROM ({paste(sql_parts, collapse = '\nUNION ALL\n')}) sub"
))

# drop the temporary wide table
dbExecute(con, "DROP TABLE dic_wide")

n_meas <- dbGetQuery(con, "SELECT COUNT(*) FROM dic_measurement")[[1]]
cat(glue("dic_measurement: {format(n_meas, big.mark = ',')} rows"), "\n")

# summary by type
dbGetQuery(
  con,
  "SELECT measurement_type, COUNT(*) AS n,
          ROUND(AVG(measurement_value), 1) AS mean_val,
          ROUND(MIN(measurement_value), 1) AS min_val,
          ROUND(MAX(measurement_value), 1) AS max_val
   FROM dic_measurement
   GROUP BY measurement_type
   ORDER BY measurement_type"
) |> datatable(caption = "Measurement summary by type")
```

## Summarize Replicate Measurements

Aggregate replicate measurements at each unique position in time
and space (station + date + depth) into mean and standard deviation,
following `ctd_summary` in `ingest_calcofi_ctd-cast.qmd`. Filters out
`NaN`, `-Inf`, `Inf` values.

```{r}
#| label: dic-measurement-summary

dbExecute(
  con,
  "CREATE OR REPLACE TABLE dic_measurement_summary AS
   SELECT
     site_key,
     datetime_utc,
     depth_m,
     latitude,
     longitude,
     measurement_type,
     AVG(measurement_value)    AS avg,
     CASE
       WHEN COUNT(*) = 1 THEN 0
       ELSE COALESCE(STDDEV_SAMP(measurement_value), 0)
     END                       AS stddev,
     COUNT(*)                  AS n_obs,
     MAX(cast_id)              AS cast_id,
     MAX(bottle_id)            AS bottle_id
   FROM dic_measurement
   WHERE NOT isnan(measurement_value)
     AND isfinite(measurement_value)
   GROUP BY site_key, datetime_utc, depth_m,
            latitude, longitude, measurement_type"
)

n_summ <- dbGetQuery(con, "SELECT COUNT(*) FROM dic_measurement_summary")[[1]]
cat(glue(
  "dic_measurement_summary: {format(n_summ, big.mark = ',')} rows ",
  "(from {format(n_meas, big.mark = ',')} measurements)"), "\n")

dbGetQuery(
  con,
  "SELECT measurement_type,
          COUNT(*) AS n_positions,
          SUM(n_obs) AS n_raw_obs,
          ROUND(AVG(n_obs), 1) AS avg_replicates,
          SUM(CASE WHEN n_obs > 1 THEN 1 ELSE 0 END) AS n_with_replicates,
          ROUND(AVG(stddev), 3) AS avg_stddev
   FROM dic_measurement_summary
   GROUP BY measurement_type
   ORDER BY measurement_type"
) |> datatable(caption = "Replicate statistics by type")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- read_csv(here("metadata/dataset.csv"))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Add Measurement Types

Register the 4 measurement types from this dataset. All are **new** —
distinct from bottle/CTD types due to different QC pipelines.

```{r}
#| label: add-measurement-types

dic_types <- tibble(
  measurement_type   = c("dic",        "alkalinity",  "ctdtemp_its90",           "salinity_pss78"),
  description        = c("Dissolved inorganic carbon",
                          "Total alkalinity",
                          "CTD temperature (ITS-90 scale)",
                          "Practical salinity (PSS-78)"),
  units              = c("umol/kg",    "umol/kg",     "degC",                    "PSS-78"),
  `_source_column`   = c("DIC",        "TA",          "CTDTEMP_ITS90",           "Salinity_PSS78"),
  `_source_table`    = c("dic_measurement", "dic_measurement", "dic_measurement", "dic_measurement"),
  `_source_datasets` = c("calcofi_dic", "calcofi_dic", "calcofi_dic",            "calcofi_dic"),
  `_qual_column`     = c("dic_flag",   "ta_flag",     "ctdtemp_its90_flag",      "salinity_pss78_flag"),
  `_prec_column`     = c(NA_character_, NA_character_, NA_character_,             NA_character_)
)

existing <- d_meas_type |>
  filter(measurement_type %in% dic_types$measurement_type) |>
  pull(measurement_type)
new_types <- dic_types |> filter(!measurement_type %in% existing)

if (nrow(new_types) > 0) {
  cat(glue("Adding {nrow(new_types)} new measurement types: ",
           "{paste(new_types$measurement_type, collapse = ', ')}"), "\n")
  d_meas_type <- bind_rows(d_meas_type, new_types)
  write_csv(d_meas_type, meas_type_csv)
} else {
  cat("All measurement types already registered\n")
}

dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Verify Primary Keys

```{r}
#| label: verify-pks

# dic_sample composite key
n_dup_sample <- dbGetQuery(
  con,
  "SELECT COUNT(*) FROM (
     SELECT expocode, site_key, depth_m, datetime_utc, COUNT(*) AS n
     FROM dic_sample
     GROUP BY expocode, site_key, depth_m, datetime_utc
     HAVING COUNT(*) > 1)")[[1]]
cat(glue("dic_sample composite key duplicates: {n_dup_sample}"), "\n")

# dic_measurement_summary composite key
n_dup_summ <- dbGetQuery(
  con,
  "SELECT COUNT(*) FROM (
     SELECT site_key, datetime_utc, depth_m, measurement_type, COUNT(*) AS n
     FROM dic_measurement_summary
     GROUP BY site_key, datetime_utc, depth_m, measurement_type
     HAVING COUNT(*) > 1)")[[1]]
cat(glue("dic_measurement_summary composite key duplicates: {n_dup_summ}"), "\n")
```

## Schema Documentation

### Primary Keys and Foreign Key Relationships

```
dic_sample (position/time)
  ↓ (site_key + datetime_utc + depth_m)
dic_measurement (one row per measurement)
  dic_measurement.measurement_type (FK) → measurement_type.measurement_type
  ↓ (site_key + datetime_utc + depth_m + measurement_type)
dic_measurement_summary (avg/stddev per position)
  dic_measurement_summary.measurement_type (FK) → measurement_type.measurement_type
```

```{r}
#| label: schema

# define PK/FK relationships for visualization and relationships.json
dic_rels <- list(
  primary_keys = list(
    dic_measurement  = "dic_measurement_id",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "dic_measurement",         column = "measurement_type", ref_table = "measurement_type", ref_column = "measurement_type"),
    list(table = "dic_measurement_summary", column = "measurement_type", ref_table = "measurement_type", ref_column = "measurement_type")))

dic_tables <- c(
  "dic_sample", "dic_measurement", "dic_measurement_summary",
  "measurement_type", "dataset")
cc_erd(
  con,
  tables = dic_tables,
  rels   = dic_rels,
  colors = list(
    lightblue   = c("dic_sample", "dic_measurement", "dic_measurement_summary"),
    lightyellow = "measurement_type",
    white       = "dataset"))
```

### Write relationships.json

```{r}
#| label: write-relationships

build_relationships_json(
  rels       = dic_rels,
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset
)
```

## Add Spatial

```{r}
#| label: spatial

add_point_geom(con, "dic_sample", lon_col = "longitude", lat_col = "latitude")
```

## Data Preview

```{r}
#| label: preview-dic-sample

# exclude GEOMETRY columns for R driver compatibility
ds_cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = 'dic_sample' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(ds_cols, collapse=', ')} FROM dic_sample LIMIT 100")) |>
  datatable(
    caption = glue(
      "dic_sample — first 100 of ",
      "{format(dbGetQuery(con, 'SELECT COUNT(*) FROM dic_sample')[[1]], big.mark=',')} rows"),
    rownames = FALSE, filter = "top")
```

```{r}
#| label: preview-dic-measurement

dbGetQuery(con, "SELECT * FROM dic_measurement LIMIT 100") |>
  datatable(
    caption = glue(
      "dic_measurement — first 100 of ",
      "{format(dbGetQuery(con, 'SELECT COUNT(*) FROM dic_measurement')[[1]], big.mark=',')} rows"),
    rownames = FALSE, filter = "top")
```

```{r}
#| label: preview-dic-summary

dbGetQuery(con, "SELECT * FROM dic_measurement_summary LIMIT 100") |>
  datatable(
    caption = glue(
      "dic_measurement_summary — first 100 of ",
      "{format(dbGetQuery(con, 'SELECT COUNT(*) FROM dic_measurement_summary')[[1]], big.mark=',')} rows"),
    rownames = FALSE, filter = "top")
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

dir_create(dir_parquet)

# collect mismatches for manifest
mismatches <- list(
  measurement_types = collect_measurement_type_mismatches(
    con, here("metadata/measurement_type.csv")),
  cruise_keys       = collect_cruise_key_mismatches(con, "dic_sample"))

parquet_stats <- write_parquet_outputs(
  con        = con,
  output_dir = dir_parquet,
  tables     = c("dic_sample", "dic_measurement",
                 "dic_measurement_summary", "measurement_type",
                 "dataset"),
  strip_provenance = FALSE,
  mismatches = mismatches
)

parquet_stats |>
  mutate(file = basename(path)) |>
  select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata

d_tbls_rd <- read_csv(here("metadata/calcofi/dic/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/calcofi/dic/flds_redefine.csv"))

metadata_path <- build_metadata_json(
  con          = con,
  d_tbls_rd    = d_tbls_rd,
  d_flds_rd    = d_flds_rd,
  output_dir   = dir_parquet,
  tables       = c("dic_sample", "dic_measurement",
                   "dic_measurement_summary"),
  set_comments = TRUE,
  provider     = provider,
  dataset      = dataset,
  workflow_url = glue(
    "https://calcofi.io/workflows/",
    "ingest_calcofi_dic.html"
  )
)
```

## Upload to GCS Archive

```{r}
#| label: upload-gcs

gcs_ingest_prefix <- glue("ingest/{dir_label}")
gcs_bucket        <- "calcofi-db"

local_files <- list.files(dir_parquet, full.names = TRUE)

walk(local_files, function(f) {
  gcs_path <- glue("gs://{gcs_bucket}/{gcs_ingest_prefix}/{basename(f)}")
  put_gcs_file(f, gcs_path)
  cat(glue("Uploaded {basename(f)} -> {gcs_path}"), "\n")
})

cat(glue(
  "\nUploaded {length(local_files)} files to ",
  "gs://{gcs_bucket}/{gcs_ingest_prefix}/"), "\n")
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```

## TODO

- [x] Tidy data: all measurements pivoted to long format in
  `dic_measurement` — `dic_sample` has position/time only
- [x] Handle replicates: `dic_measurement_summary` aggregates with
  avg/stddev per position, following `ctd_summary` pattern
- [ ] Spatial matching: match unmatched DIC samples to `site` and `grid`
  via lat/lon projection to CalCOFI coordinate system (see issue #47)
- [ ] Validate against calcofi_bottle `dic_rep1`/`dic_rep2` values for
  overlapping bottles
- [ ] Add WOCE flag interpretation (2=good, 3=questionable, 4=bad,
  9=missing) to metadata
- [ ] Consider merging `dic_measurement` into `bottle_measurement` for
  matched samples via a VIEW that unions both tables for analysis

::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
