---
title: "Ingest CalCOFI METS (Underway TSG/Meteorology)"
calcofi:
  target_name: ingest_calcofi_mets
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/calcofi_mets/manifest.json
  modifies:
    - ship
  provider: calcofi
  dataset: mets
  workflow_url: https://calcofi.io/workflows/ingest_calcofi_mets.html
  questions_file: metadata/calcofi/mets/questions.csv
  dataset_meta:
    dataset_name: CalCOFI METS (Underway TSG/Meteorology)
    description: >
      Shipboard underway thermosalinograph (TSG) and meteorology data,
      ~1-minute resolution, recorded continuously along the cruise track.
    citation_main: "CalCOFI. Underway (METS) TSG/Meteorology Data. CalCOFI.org."
    link_calcofi_org: https://calcofi.org/data/oceanographic-data/underway-data/
    link_data_source: https://calcofi.org/data/oceanographic-data/underway-data/
    coverage_temporal: "TBD — spans at least 2006-2018 in files received so far"
    coverage_spatial: "28-37°N, -125 to -117°W"
  tables_owned:
    - {table: mets_sample}
    - {table: mets_measurement}
    - {table: measurement_type, shared: true, note: "shared registry across bottle/ctd/dic/mets"}
  erd:
    color: "#cce5ff"
---

## Overview

Ingest **METS** (shipboard underway TSG + meteorology data) from CalCOFI
cruises. Unlike `ingest_calcofi_ctd-cast.qmd`, source files arrive by manual
Drive upload per cruise rather than by scraping a download page, but the
same **discover -> prioritize -> classify -> bind -> bridge -> pivot**
shape applies, because the same problems show up: multiple files per
cruise (raw vs. final), multiple source schemas across cruise eras, and
a growing archive that should resume rather than re-process from scratch.

- **Provider**: `calcofi`
- **Dataset**: `mets`
- **Source**: Per-cruise underway files (xlsx/csv) manually placed in
  `dir_data/calcofi/mets/download/`, one row per ~1-minute sample.
  In scope: the **"Final Underway Data"** era from 1203SH (2012) onward —
  single-TSG (e.g. CC1704SH) and two dual-TSG variants (e.g. CC1802SH,
  CC1611SR) — the tab-separated PST variant found for 0903JD, the one-off
  "Raw Underway Data" for 1207OS (2012, schema confirmed via a bundled
  explanation doc, `mets_14`), and the ~2004-2005 **SCIMS** era. SCIMS files
  cannot be re-downloaded (calcofi.org 403s, and the page now 404s), but real
  copies for 8 cruises are already archived in Drive, so they are ingested
  from there with their columns positionally named and flagged unconfirmed
  (`mets_01`). Only **SCS** is out of scope — no example file exists anywhere
  in the archive (`mets_12`), as is 0810NH specifically (`mets_11`).

::: {.callout-important}
**This notebook has not been run.** Unlike the other ingests it has no
reproducible acquisition path: its source is a hand-staged per-cruise archive,
`calcofi.org/data/oceanographic-data/underway-data/` now returns 404, and no
METS files are present under `dir_data`. The code here has been corrected and
statically checked but every row count, coverage figure, and assertion below is
**unverified against real data** — expect to iterate on the first real run.
:::

### Data Flow

```{mermaid}
graph LR
  A[Per-cruise xlsx/csv in Drive, 4+ schemas] --> B[Classify + Prioritize per cruise]
  B --> C[Read + Standardize per schema]
  C --> D[Checkpoint]
  D --> E[Cross-Dataset Bridge: ship_key/cruise_key]
  E --> F[Dedup per cruise+timestamp]
  F --> G[Pivot wide -> long measurement]
  G --> H[Parquet Export]
  H --> I[GCS Archive]
  I --> J[Release Database]
```

## Setup

```{r}
#| label: setup

knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    .time_it_t0 <<- Sys.time()
  } else {
    elapsed <- round(difftime(Sys.time(), .time_it_t0, units = "secs"), 1)
    tnow <- format(Sys.time(), "%H:%M:%S")
    message(glue::glue("R chunk {options$label}: {elapsed}s ~ {tnow}"))
  }
})
knitr::opts_chunk$set(time_it = TRUE)

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db,
  CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue,
  here, janitor, jsonlite, knitr,
  lubridate, purrr, readr, readxl, sf, stringr,
  tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_calcofi_mets.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")
dir_dl       <- path_expand(glue("{dir_data}/{provider}/{dataset}/download"))
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))
db_checkpoint <- here(glue("data/wrangling/{dir_label}_checkpoint.duckdb"))
dir_tmp      <- here(glue("data/tmp/{dir_label}"))
dir_meta     <- here(glue("metadata/{provider}/{dataset}"))

if (overwrite) {
  if (file_exists(db_path)) file_delete(db_path)
  db_wal <- paste0(db_path, ".wal")
  db_tmp <- paste0(db_path, ".tmp")
  if (file_exists(db_wal)) file_delete(db_wal)
  if (dir_exists(db_tmp))  dir_delete(db_tmp)
  if (overwrite_all) {
    if (dir_exists(dir_parquet))    dir_delete(dir_parquet)
    if (file_exists(db_checkpoint)) file_delete(db_checkpoint)
    cat("Deleted parquet and checkpoint DB", "\n")
  }
}

if (file_exists(db_checkpoint) && !file_exists(db_path)) {
  file_copy(db_checkpoint, db_path, overwrite = TRUE)
  cat(glue("Restored from checkpoint: {db_checkpoint}"), "\n")
}
dir_create(c(dir_dl, dirname(db_path), dir_parquet, dir_tmp), recurse = TRUE)

con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
load_duckdb_extension(con, "icu")

mem_gb <- ps::ps_system_memory()$total / 1024^3 / 2 |> floor()
dbExecute(con, glue("SET memory_limit = '{mem_gb}GB'"))
dbExecute(con, glue("SET temp_directory = '{dir_tmp}'"))

d_meas_type <- read_csv(here("metadata/measurement_type.csv"), show_col_types = F)
d_flds_rd   <- read_csv(glue("{dir_meta}/flds_redefine.csv"), show_col_types = F)
d_tbls_rd   <- read_csv(glue("{dir_meta}/tbls_redefine.csv"), show_col_types = F)
```

## Check for Resumable State

```{r}
#| label: check_resume

parquet_complete <- FALSE
manifest_path <- file.path(dir_parquet, "manifest.json")
if (file_exists(manifest_path)) {
  mf <- jsonlite::read_json(manifest_path)
  parquet_ok <- all(vapply(mf$tables, function(tbl) {
    file_exists(file.path(dir_parquet, paste0(tbl, ".parquet")))
  }, logical(1)))
  if (parquet_ok && !overwrite) {
    parquet_complete <- TRUE
    cat(glue(
      "Parquet output already complete ({length(mf$tables)} tables, ",
      "{format(mf$total_rows, big.mark = ',')} rows) — ",
      "skipping computation, resuming at upload"))
  }
}

has_mets_raw <- FALSE
if (!parquet_complete) {
  has_mets_raw <- "mets_raw" %in% DBI::dbListTables(con)
  if (has_mets_raw) {
    n_raw <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM mets_raw")$n
    cat(glue(
      "Checkpoint: mets_raw already loaded ({format(n_raw, big.mark = ',')} rows) — ",
      "skipping read+classify+bind"))
  }
}

skip_read_bind <- parquet_complete || has_mets_raw
knitr::opts_chunk$set(eval = !skip_read_bind)
```

## Discover and Classify Source Files

Scope, per the full archive (73 cruises, `mets_file_inventory.xlsx`):

- **SCIMS** (~2004-2005, 8 cruises + 1 mislabeled orphan): header-less,
  column identity still unconfirmed (`mets_01`). calcofi.org 403s on
  fresh downloads, but these are already-archived files in Drive, so
  **back in scope** (reverses the earlier `wontfix`).
- **SCS**: still zero example files anywhere in the archive — still out
  of scope (`mets_12`).
- **1207OS Raw** (2012, one cruise): confirmed schema, see above.
- **`mets_schema_final_txt_pst`** (0903JD, 2009, found inside a nested
  zip): tab-separated, real header, explicit PST. Unknown how many other
  cruises use this (`mets_18`).
- **`mets_schema_final_csv_utc`** (37 cruises, the largest group): real
  header, explicit UTC. Ships under two filename conventions
  (`final_underway_data.csv` and `{cruise}_UnderwayFinaldt.csv`) that are
  the same schema — confirmed by identical headers, not assumed.
- **`mets_schema_single_tsg`/`dual_tsg_a`/`dual_tsg_b`** (24 cruises,
  `.xlsx` only): as before.
- **1MinData** (20 cruises): relationship to the official download still
  unconfirmed (`mets_15`).
- **20 cruises have no data file at all** — plots only.

**cruise_key**: derived from the **folder** code, not the filename. Three real
folder/filename mismatches exist in the archive (`mets_21`) and all three were
confirmed to belong to their folder: `1510OC_UnderwayFinaldt.csv` is really
1511OC, and two *separate* files both named `1202NH_UnderwayFinaldt.csv` sit in
the 1110NH and 1203SH folders — since only one real 1202NH cruise exists, both
filenames cannot be right, which is what settles folder-over-filename. Files
where the two disagree are still flagged in the table below so the mismatch
stays visible.

```{r}
#| label: d_files

schema_signatures <- list(
  mets_schema_single_tsg = c(
    "ID","Cruise","DateTime","Lat","Lon","TSGTemp","TSGSal"),
  mets_schema_dual_tsg_a = c(
    "ID","Cruise","DateTime","Lat","Lon","TSG_Temp","TSG_Temp_2","USWFlow"),
  mets_schema_dual_tsg_b = c(
    "ID","Cruise","DateTime","Lat","Lon","TSG_Temp","TSG_Temp_2",
    "BottomDepth","SW_pH"),
  mets_schema_final_csv_utc = c(
    "DATE_TIME_UTC","DATE_UTC","TIME_UTC","Latitude_N","Longitude_W",
    "SST_degC","Salinity_PSU"),
  mets_schema_final_txt_pst = c(
    "DATE_PST","TIME_PST","LA","LO","ST","SA","CAL_T","CAL_S"))

detect_schema <- function(file_name, header_cols) {
  # SCIMS and Raw_Underway files are header-less — signature matching
  # can't apply, so detect by filename first
  if (str_detect(file_name, regex("scims|10mindata", ignore_case = T)))
    return("mets_schema_scims")
  if (str_detect(file_name, regex("^Raw_Underway", ignore_case = T)))
    return("mets_schema_raw_2012")
  scores <- purrr::map_int(schema_signatures, ~ sum(.x %in% header_cols))
  best <- names(scores)[which.max(scores)]
  if (max(scores) == 0) "mets_schema_unrecognized" else best
}

d_files <- tibble(
  path = list.files(dir_dl, pattern = "\\.(xlsx|csv|txt)$",
                     recursive = TRUE, full.names = TRUE)
) |>
  mutate(
    file_name  = basename(path),
    # cruise code from the FILENAME (not folder) where the file encodes one
    fname_cruise = {
      m <- str_match(file_name, "(?i)(?:CC)?(\\d{4})([A-Za-z]{2})(?![A-Za-z0-9])")
      code <- str_to_upper(paste0(m[,2], m[,3]))
      if_else(is.na(m[,2]) | str_to_upper(m[,3]) == "UW", NA_character_, code)
    },
    folder_cruise = str_extract(path, "(\\d{4}[A-Z]{2})(?=[^/]*$)|(\\d{4}[A-Z]{2})(?=/[^/]*$)") |>
      str_to_upper(),
    # mets_21: folder wins; filename is only a fallback when the folder has no
    # code at all. Confirmed against the three known mismatches.
    cruise_key = coalesce(folder_cruise, fname_cruise),
    cruise_key_mismatch = !is.na(fname_cruise) & !is.na(folder_cruise) &
      fname_cruise != folder_cruise,
    header = map2(path, file_name, \(p, fn) {
      if (str_detect(fn, regex("scims|10mindata|^Raw_Underway", ignore_case = T)))
        return(character(0))
      if (str_detect(p, "\\.xlsx$")) {
        tryCatch(names(readxl::read_excel(p, sheet = 1, n_max = 0)),
                 error = function(e) character(0))
      } else {
        tryCatch(names(read_csv(p, n_max = 0, show_col_types = F)),
                 error = function(e) character(0))
      }
    }),
    schema_variant = map2_chr(file_name, header, detect_schema),
    source_system = case_when(
      str_detect(file_name, regex("scims|10mindata", ignore_case = T)) ~ "SCIMS",
      str_detect(file_name, regex("^Raw_Underway", ignore_case = T)) ~ "Raw (1207OS)",
      schema_variant == "mets_schema_final_txt_pst" ~ "Final (PST txt, 0903JD-style)",
      str_detect(file_name, regex("final_underway|underwayfinaldt", ignore_case = T)) ~ "Final",
      str_detect(file_name, regex("1min", ignore_case = T)) ~ "1MinData (unconfirmed provenance, see mets_15)",
      TRUE ~ "unknown — flag for review")
  ) |>
  select(-header)

stopifnot("cruise_key could not be determined for some files" =
  all(!is.na(d_files$cruise_key)))

n_mismatch <- sum(d_files$cruise_key_mismatch)
if (n_mismatch > 0) {
  cat(glue(
    "{n_mismatch} file(s) have a folder/filename cruise-code disagreement ",
    "— see questions.csv mets_21. Using the FOLDER cruise code, not the ",
    "filename, for these: ",
    "{paste(d_files$file_name[d_files$cruise_key_mismatch], collapse = ', ')}"), "\n")
}

# fail loudly rather than silently ingesting an unrecognized schema —
# safer than CTD's stopifnot(!any(zip_type == "unknown")) since here a
# false match would corrupt column meaning, not just miscategorize a file
n_unrecognized <- sum(d_files$schema_variant == "mets_schema_unrecognized")
if (n_unrecognized > 0) {
  cat(glue(
    "{n_unrecognized} file(s) didn't match a known schema signature — ",
    "add a schema_signatures entry once their real columns are known"))
}

d_files |>
  dt(caption = "Files to ingest — every file kept, tagged by source system and schema variant",
     fname   = "mets_files_to_ingest")
```

## Read and Standardize per Schema Variant

```{r}
#| label: read_files

read_mets_file <- function(path, file_name) {
  if (str_detect(file_name, regex("^Raw_Underway", ignore_case = T))) {
    # confirmed header-less via bundled explanation doc: date, time, temp_c,
    # conductivity, salinity_psu (Seabird TSG 45 only, no lat/lon)
    d <- read_csv(path, col_names = c(
      "col_1", "col_2", "col_3", "col_4", "col_5"), show_col_types = F)
  } else if (str_detect(path, "\\.xlsx$")) {
    d <- readxl::read_excel(path, sheet = 1)
  } else {
    d <- read_csv(path, show_col_types = F)
  }
  d |> clean_names(case = "none")  # preserve original case; flds_redefine maps exact names
}

d_files <- d_files |>
  mutate(
    data  = map2(path, file_name, \(p, fn) {
      cat(glue("Reading: {fn}"), "\n")
      read_mets_file(p, fn)
    }),
    nrows = map_int(data, nrow)
  )

d_files |>
  select(cruise_key, file_name, schema_variant, source_system, nrows) |>
  dt(caption = "Rows read per file", fname = "mets_rows_per_file") |>
  formatCurrency("nrows", currency = "", digits = 0, mark = ",")
```

## Bind per Schema Variant

Each schema variant has a different column set, so bind within variant
first (avoiding a single wide union of ~50 sparse columns), then stack
with an explicit `schema_variant` tag for `flds_redefine.csv` to key off.

```{r}
#| label: d_bind

d_bind <- d_files |>
  select(cruise_key, file_name, schema_variant, data) |>
  mutate(data = map(data, \(x) mutate(x, across(everything(), as.character)))) |>
  unnest(data) |>
  rename(`_source_file` = file_name)

write_rds(d_bind, glue("{dir_tmp}/d_bind.rds"), compress = "gz")

dbWriteTable(con, "mets_raw", d_bind, overwrite = TRUE)
cat(glue("Loaded {nrow(d_bind)} rows into mets_raw across ",
         "{n_distinct(d_bind$schema_variant)} schema variants"), "\n")
```

## Save Checkpoint

```{r}
#| label: save_checkpoint

if (!file_exists(db_checkpoint) || overwrite) {
  close_duckdb(con)
  file_copy(db_path, db_checkpoint, overwrite = TRUE)
  con <- get_duckdb_con(db_path)
  load_duckdb_extension(con, "spatial")
  load_duckdb_extension(con, "icu")
  cat(glue("Saved checkpoint: {db_checkpoint}"), "\n")
} else {
  cat(glue("Checkpoint already exists: {db_checkpoint}"), "\n")
}
```

## Apply Field Renames (per schema variant)

```{r}
#| label: rename_fields

mets_wide <- dbGetQuery(con, "SELECT * FROM mets_raw")

mets_wide <- mets_wide |>
  group_split(schema_variant) |>
  map(\(d) {
    sv <- d$schema_variant[1]
    renames <- d_flds_rd |>
      filter(tbl_old == sv, !is.na(fld_new), fld_new != "")
    for (i in seq_len(nrow(renames))) {
      old <- renames$fld_old[i]
      new <- renames$fld_new[i]
      # renaming onto a name the bind already carries (e.g. the file-derived
      # cruise_key) is a hard "Names must be unique" error, so catch it here
      # rather than letting one bad mapping abort the whole ingest
      if (old %in% names(d)) {
        if (new %in% setdiff(names(d), old))
          stop(glue(
            "flds_redefine maps {sv}.{old} -> {new}, but {new} already exists ",
            "on the bound frame; pick a different fld_new"))
        d <- d |> rename(!!new := !!old)
      }
    }
    d
  }) |>
  list_rbind()

# coerce numeric columns now that renaming is done (everything was read as
# character to allow a uniform bind across variant-specific column sets)
numeric_cols <- d_flds_rd |>
  filter(type_new %in% c("double", "integer", "smallint")) |>
  pull(fld_new) |> unique() |> intersect(names(mets_wide))
mets_wide <- mets_wide |>
  mutate(across(all_of(numeric_cols), as.numeric))

# --- datetime_start_utc, per schema variant --------------------------------
# every column arrives as VARCHAR (uniform bind), so each branch parses from
# character and the result is assembled as character before one final cast.
# Mixing a POSIXct branch with the character column in if_else() is a type
# error ("Can't combine <datetime> and <character>"), so don't.
chr_col <- function(d, nm)
  if (nm %in% names(d)) as.character(d[[nm]]) else NA_character_

sv  <- mets_wide$schema_variant
dt  <- chr_col(mets_wide, "datetime_start_utc")   # NA when no variant supplied it
fmt <- function(x) format(x, "%Y-%m-%d %H:%M:%S")

# final_csv_utc: a real DATE_TIME_UTC column, MM/DD/YYYY HH:MM:SS, already UTC
i <- sv == "mets_schema_final_csv_utc" & !is.na(dt)
if (any(i)) dt[i] <- fmt(mdy_hms(dt[i], tz = "UTC"))

# xlsx variants (single/dual TSG): mets_02 is RESOLVED as a team decision, not a
# provider confirmation -- the DateTime column is treated as local Pacific
# Standard Time at a fixed -8:00 (not DST-aware), matching the one schema whose
# timezone IS confirmed by column name (final_txt_pst). The previous version
# silently let these fall through as already-UTC, which applied no offset at
# all; that is what mets_02 records as "almost certainly wrong". Low confidence
# -- revisit if CalCOFI staff can confirm the true source timezone.
i <- sv %in% c("mets_schema_single_tsg", "mets_schema_dual_tsg_a",
               "mets_schema_dual_tsg_b") & !is.na(dt)
if (any(i)) {
  pac <- suppressWarnings(parse_date_time(
    dt[i], orders = c("Ymd HMS", "Ymd HM", "mdY HMS", "mdY HM"), tz = "UTC"))
  if (all(is.na(pac)))
    stop("xlsx TSG variants: could not parse the DateTime column")
  dt[i] <- fmt(pac + hours(8))   # assumed PST -> UTC (mets_02)
}

# raw_2012 (1207OS): header-less date + time; UTC-vs-local unconfirmed, left as
# UTC -- see questions.csv mets_17 (open)
i <- sv == "mets_schema_raw_2012" & !is.na(chr_col(mets_wide, "date_mdy"))
if (any(i)) dt[i] <- fmt(mdy_hms(
  paste(chr_col(mets_wide, "date_mdy")[i], chr_col(mets_wide, "time_hms_ms")[i]),
  tz = "UTC"))

# final_txt_pst (0903JD): explicit Pacific in the column name. The cruise runs
# 7-24 Mar 2009, entirely in standard time, so fixed -8:00 is right here; the
# general convention is still unconfirmed (mets_19). Date order is ambiguous
# from the format alone, so parse permissively and assert it resolved.
i <- sv == "mets_schema_final_txt_pst" & !is.na(chr_col(mets_wide, "date_pst"))
if (any(i)) {
  pst_chr <- paste(chr_col(mets_wide, "date_pst")[i], chr_col(mets_wide, "time_pst")[i])
  pst <- suppressWarnings(parse_date_time(
    pst_chr, orders = c("mdY HMS", "Ymd HMS", "dmY HMS"), tz = "UTC"))
  if (all(is.na(pst)))
    stop("final_txt_pst: could not parse DATE_PST/TIME_PST under any tried order")
  dt[i] <- fmt(pst + hours(8))   # PST -> UTC
}

# SCIMS: header-less numeric date (YYYYMMDD) + time (HHMMSS), column identity
# assumed (mets_01). Deriving these matters beyond the timestamp itself: the
# de-dup below partitions on datetime_start_utc, and DuckDB groups all NULLs
# into ONE partition, so a variant left with NULL timestamps would collapse to
# a single row per cruise.
i <- sv == "mets_schema_scims" & !is.na(chr_col(mets_wide, "date_ymd"))
if (any(i)) dt[i] <- fmt(ymd_hms(paste(
  chr_col(mets_wide, "date_ymd")[i],
  str_pad(chr_col(mets_wide, "time_hms")[i], 6, pad = "0")), tz = "UTC"))

mets_wide$datetime_start_utc <- as_datetime(dt, tz = "UTC")

# no in-scope variant may be left entirely without timestamps (see above)
dt_cover <- mets_wide |>
  summarise(n = n(), n_dt = sum(!is.na(datetime_start_utc)), .by = schema_variant) |>
  mutate(pct = round(100 * n_dt / n, 1))
dt_cover |> dt(caption = "datetime_start_utc coverage by schema variant",
               fname = "mets_datetime_coverage")
stopifnot(
  "every schema variant must derive at least some datetime_start_utc" =
    all(dt_cover$n_dt > 0))

dbWriteTable(con, "mets_wide", mets_wide, overwrite = TRUE)
```

```{r}
#| label: restore_eval
#| eval: true

# must always evaluate: check_resume may have set eval=FALSE to skip
# read+classify+bind when restoring from checkpoint. Without this the whole
# rest of the pipeline (rename -> dedup -> pivot -> write_parquet) would
# inherit eval=FALSE and be skipped, leaving the GCS sync to upload a stale or
# absent parquet directory. Same shape as ingest_calcofi_ctd-cast.qmd.
if (parquet_complete) {
  knitr::opts_chunk$set(eval = FALSE)
  cat("Parquet complete — skipping to upload\n")
} else {
  knitr::opts_chunk$set(eval = TRUE)
}
```

## Cross-Dataset Bridge

Derive `ship_key`/full `cruise_key` and validate against the `ship`/`cruise`
reference tables. Similar in spirit to `ingest_calcofi_ctd-cast.qmd` (same
orphan-ship handling via `match_ships()`/`ensure_interim_ships()`), but not
identical: `ctd-cast` derives `ship_key` directly (`RIGHT(cruise_key, 2)`)
and leaves `cruise_key` in short filename form, whereas this uses
`derive_cruise_key_on_casts()` to convert to a full key -- worth confirming
with Ben that this is the right target format for `cruise.cruise_key`.

```{r}
#| label: cross_dataset_bridge

modifies_tables <- c("ship")
load_prior_tables(con = con, tables = modifies_tables,
                   parquet_dir = here("data/parquet/swfsc_ichthyo"))
load_prior_tables(con = con, tables = c("cruise", "grid"),
                   parquet_dir = here("data/parquet/swfsc_ichthyo"), as_view = TRUE)

modifies_pks <- list()
for (tbl in modifies_tables) {
  pk_col <- dbGetQuery(con, glue(
    "SELECT column_name FROM information_schema.columns
     WHERE table_name = '{tbl}' ORDER BY ordinal_position LIMIT 1"))$column_name
  modifies_pks[[tbl]] <- list(
    pk_col = pk_col,
    keys   = dbGetQuery(con, glue("SELECT {pk_col} FROM {tbl}"))[[1]])
}

# short cruise code (e.g. "CC1704", "1704") -> full cruise_key via ship_renames
# and derive_cruise_key_on_casts() -- same helper bottle.qmd uses, but CTD
# does not call it (CTD keeps cruise_key in short form)
derive_cruise_key_on_casts(con, table_name = "mets_wide",
                            ship_renames_csv = here("metadata/ship_renames.csv"))

mets_ships <- dbGetQuery(con, "SELECT DISTINCT ship_key FROM mets_wide")
ref_ships  <- dbGetQuery(con, "SELECT ship_key FROM ship")
orphan_ships <- setdiff(mets_ships$ship_key, ref_ships$ship_key)

if (length(orphan_ships) > 0) {
  cat(glue("{length(orphan_ships)} ship_key(s) in METS not in ship table: ",
           "{paste(orphan_ships, collapse = ', ')}"), "\n")
  orphan_tbl <- dbGetQuery(con, glue(
    "SELECT DISTINCT ship_key AS ship_code, NULL AS ship_name FROM mets_wide
     WHERE ship_key IN ({paste(dbQuoteString(con, orphan_ships), collapse = ', ')})"))
  ship_result <- match_ships(
    unmatched_ships  = orphan_tbl,
    reference_ships  = dbReadTable(con, "ship"),
    ship_renames_csv = here("metadata/ship_renames.csv"),
    fetch_ices       = FALSE)
  ensure_interim_ships(con, ship_result)
} else {
  cat("All METS ship_keys found in ship reference table", "\n")
}

mets_cruises <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM mets_wide")
ref_cruises  <- dbGetQuery(con, "SELECT cruise_key FROM cruise")
orphan_cruises <- setdiff(mets_cruises$cruise_key, ref_cruises$cruise_key)
if (length(orphan_cruises) > 0) {
  cat(glue("{length(orphan_cruises)} cruise_key(s) in METS not in cruise table: ",
           "{paste(orphan_cruises, collapse = ', ')}"), "\n")
} else {
  cat("All METS cruise_keys found in cruise reference table", "\n")
}
```

## De-duplicate mets_sample

Unlike CTD, a duplicate here means a genuine re-upload artifact (the same
file appearing twice), **not** SCIMS vs. SCS — those are independent
systems and both are expected to have real, distinct rows for the same
cruise+timestamp (see `mets_13`). This step only catches literal
duplicate rows within a single system/schema_variant; it must not collapse
SCIMS against SCS.

```{r}
#| label: dedup_mets_sample

n_before <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM mets_wide")$n
dbExecute(con, "
  CREATE OR REPLACE TABLE mets_sample AS
  SELECT * FROM mets_wide
  QUALIFY ROW_NUMBER() OVER (
    -- schema_variant included so SCIMS and SCS (or any two distinct
    -- systems) at the same cruise+timestamp are never collapsed together.
    -- source_row_id disambiguates rows that share a timestamp WITHIN one
    -- file: DuckDB groups all NULLs into a single partition, so without it
    -- any variant with unparsed timestamps would collapse to one row per
    -- cruise instead of de-duplicating genuine re-uploads.
    PARTITION BY cruise_key, schema_variant, datetime_start_utc,
                 COALESCE(CAST(source_row_id AS VARCHAR), '')
    ORDER BY \"_source_file\") = 1")
n_after <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM mets_sample")$n

# a de-dup that removes most of the archive is a bug, not a clean-up
stopifnot(
  "de-dup removed >20% of rows — check datetime parsing per schema variant" =
    n_after >= 0.8 * n_before)
cat(glue(
  "mets_sample de-duplicated: removed {format(n_before - n_after, big.mark = ',')} ",
  "duplicate-timestamp rows; {format(n_after, big.mark = ',')} remain"), "\n")

assign_deterministic_uuids_md5(
  con = con, table_name = "mets_sample", id_col = "mets_sample_uuid",
  key_cols = c("cruise_key", "schema_variant", "datetime_start_utc"))

dbExecute(con, "DROP TABLE IF EXISTS mets_wide")
dbExecute(con, "DROP TABLE IF EXISTS mets_raw")
```

## Pivot Measurements

```{r}
#| label: pivot

# the measurement vocabulary is derived from flds_redefine rather than hand-
# listed, so a column added to a schema variant can't silently fail to pivot.
# Non-measurement fields (identifiers, position, navigation, timestamps) are
# excluded explicitly.
NON_MEASUREMENT <- c(
  "source_row_id", "cruise_key", "cruise_orig", "ship_key", "schema_variant",
  "_source_file", "mets_sample_uuid", "grid_key", "geom",
  "datetime_start_utc", "date_mdy", "time_hms_ms", "date_pst", "time_pst",
  "date_ymd", "time_hms", "latitude", "longitude",
  "course_over_ground_deg", "speed_over_ground_kt", "heading_deg",
  # SCIMS col_3: an unidentified integer *counter*, not a sensor reading
  # (mets_01's answer scopes the unknown measurements to col_6-9)
  "unknown_counter")

sample_cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = 'mets_sample'")$column_name

sensor_cols <- d_flds_rd |>
  filter(!is.na(fld_new), fld_new != "", !fld_new %in% NON_MEASUREMENT) |>
  pull(fld_new) |> unique() |> intersect(sample_cols) |> sort()

cat(glue("Pivoting {length(sensor_cols)} measurement columns: ",
         "{paste(sensor_cols, collapse = ', ')}"), "\n")

# -99 is a confirmed missing-value sentinel for the bottom-depth columns only
# (mets_10); applying it to every sensor would silently drop legitimate
# negative readings elsewhere (e.g. radiation, air temperature)
SENTINEL_99 <- c("bottom_depth_m", "bottom_depth_mb_m")

dbExecute(con, "DROP TABLE IF EXISTS mets_measurement")
for (i in seq_along(sensor_cols)) {
  col       <- sensor_cols[i]
  sentinel  <- if (col %in% SENTINEL_99) glue("AND {col} <> -99") else ""
  sql_select <- glue("
    SELECT mets_sample_uuid, cruise_key,
      '{col}' AS measurement_type,
      CAST({col} AS DOUBLE) AS measurement_value
    FROM mets_sample
    WHERE {col} IS NOT NULL
      {sentinel}
      AND NOT isnan(CAST({col} AS DOUBLE))")
  if (i == 1) {
    dbExecute(con, glue("CREATE OR REPLACE TABLE mets_measurement AS {sql_select}"))
  } else {
    dbExecute(con, glue("INSERT INTO mets_measurement {sql_select}"))
  }
}
n_meas <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM mets_measurement")$n
cat(glue("mets_measurement: {format(n_meas, big.mark = ',')} rows"), "\n")

# every schema variant must contribute measurements — SCIMS in particular
# carries its whole payload in the unknown_measurement_* columns (mets_01)
var_cover <- dbGetQuery(con, "
  SELECT s.schema_variant, COUNT(m.mets_sample_uuid) AS n_meas
  FROM mets_sample s LEFT JOIN mets_measurement m USING (mets_sample_uuid)
  GROUP BY 1 ORDER BY 1")
var_cover |> dt(caption = "Measurements per schema variant",
                fname = "mets_measurements_per_variant")
stopifnot(
  "every schema variant must contribute at least one measurement" =
    all(var_cover$n_meas > 0))

assign_deterministic_uuids_md5(
  con = con, table_name = "mets_measurement", id_col = "mets_measurement_uuid",
  key_cols = c("mets_sample_uuid", "measurement_type"))

# drop the wide sensor columns now they live in the long table — at ~1-minute
# resolution across the archive, keeping both forms doubles the stored volume
# and lets mets_sample and mets_measurement disagree (the sentinel filter above
# applies only to the long form). Same pattern as the picoplankton ingest.
for (col in sensor_cols)
  dbExecute(con, glue('ALTER TABLE mets_sample DROP COLUMN IF EXISTS "{col}"'))
cat(glue("Dropped {length(sensor_cols)} wide sensor columns from mets_sample"), "\n")
```

## measurement_type reference table

METS types are **registered from `flds_redefine.csv`**, which already carries a
description and units per field — the previous version asserted they were
already in `metadata/measurement_type.csv` without ever writing them, which
could only ever fail. Types are upserted (not skip-if-present) so a units answer
landing in `flds_redefine.csv` propagates on the next run, and existing rows
owned by other datasets gain `calcofi_mets` in `_source_datasets` rather than
being overwritten.

```{r}
#| label: measurement_type

mets_types_used <- dbGetQuery(con,
  "SELECT DISTINCT measurement_type FROM mets_measurement ORDER BY 1")$measurement_type

# one description/units per canonical field name (variants agree; take the first)
mets_types <- d_flds_rd |>
  filter(fld_new %in% mets_types_used) |>
  summarise(
    description = first(na.omit(fld_description)),
    units       = first(na.omit(units)),
    .by = fld_new) |>
  transmute(
    measurement_type   = fld_new,
    description        = coalesce(description, fld_new),
    units              = units,
    is_canonical       = FALSE,
    `_source_column`   = measurement_type,
    `_source_table`    = "mets_measurement",
    `_source_datasets` = "calcofi_mets",
    `_qual_column`     = NA_character_,
    `_prec_column`     = NA_character_,
    grain              = "obs")

stopifnot(
  "every METS measurement_type must be described in flds_redefine.csv" =
    setequal(mets_types$measurement_type, mets_types_used))

# types another dataset already owns: append provenance instead of clobbering
shared <- intersect(mets_types$measurement_type, d_meas_type$measurement_type)
d_meas_type <- d_meas_type |>
  mutate(`_source_datasets` = if_else(
    measurement_type %in% shared &
      !str_detect(coalesce(`_source_datasets`, ""), "calcofi_mets"),
    str_replace(str_squish(paste(coalesce(`_source_datasets`, ""),
                                 "calcofi_mets", sep = ";")), "^;", ""),
    `_source_datasets`))

d_meas_type <- d_meas_type |>
  bind_rows(mets_types |> filter(!measurement_type %in% shared)) |>
  arrange(measurement_type)
write_csv(d_meas_type, here("metadata/measurement_type.csv"))

cat(glue("measurement_type: {nrow(mets_types)} METS types registered ",
         "({length(shared)} shared with other datasets)"), "\n")

dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Schema Diagram

```{r}
#| label: schema

mets_rels <- list(
  primary_keys = list(
    mets_sample      = "mets_sample_uuid",
    mets_measurement = "mets_measurement_uuid",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "mets_measurement", column = "mets_sample_uuid",
         ref_table = "mets_sample", ref_column = "mets_sample_uuid"),
    list(table = "mets_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(con, rels = mets_rels)
```

## Add Spatial

`mets_schema_raw_2012` rows have no lat/lon at all (see questions.csv
mets_16) — they'll get a NULL `geom`/`grid_key` here unless position is
joined in from elsewhere before this step. `add_point_geom`/
`assign_grid_key` presumably no-op or leave NULL on missing coordinates
rather than error, but that's unverified against real behavior — worth
checking on first run whether this schema's rows survive downstream
validation with a NULL grid_key, or need excluding until mets_16 resolves.

```{r}
#| label: add_spatial

add_point_geom(con, "mets_sample", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "mets_sample")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Questions for Data Providers

```{r}
#| label: provider-questions

q_path <- glue("{dir_meta}/questions.csv")
if (file_exists(q_path)) {
  read_csv(q_path, show_col_types = F) |>
    arrange(
      factor(status,   c("blocker", "open", "asked", "answered", "wontfix")),
      factor(priority, c("blocker", "high", "normal"))) |>
    datatable(caption = "Follow-up questions for the data provider",
              options = list(dom = "tp", pageLength = 10))
}
```

## Validate and Enforce Types

```{r}
#| label: validate

validate_for_release(con)
enforce_column_types(con, d_flds_rd = d_flds_rd)
```

## Preview Tables

```{r}
#| label: preview
#| results: asis

preview_tables(con, tables = c("mets_sample", "mets_measurement", "measurement_type"))
```

## Write Parquet

```{r}
#| label: write_parquet

mismatches <- list(
  ships             = collect_ship_mismatches(con, "mets_sample"),
  measurement_types = collect_measurement_type_mismatches(
    con, here("metadata/measurement_type.csv")),
  cruise_keys       = collect_cruise_key_mismatches(con, "mets_sample"))

parquet_stats <- write_parquet_outputs(
  con          = con,
  output_dir   = dir_parquet,
  tables       = c("mets_sample", "mets_measurement", "measurement_type"),
  partition_by = list(mets_measurement = "cruise_key"),
  sort_by      = list(
    mets_measurement = c("measurement_type"),
    mets_sample      = "hilbert:longitude,latitude"),
  strip_provenance = FALSE,
  mismatches       = mismatches)

parquet_stats |> dt(fname = "mets_parquet_stats")
```

## Write Metadata JSON

```{r}
#| label: write_metadata

metadata_path <- build_metadata_json(
  con                  = con,
  d_tbls_rd            = d_tbls_rd,
  d_flds_rd            = d_flds_rd,
  metadata_derived_csv = glue("{dir_meta}/metadata_derived.csv"),
  output_dir           = dir_parquet,
  tables               = parquet_stats$table,
  provider             = provider,
  dataset              = dataset,
  workflow_url         = cc$workflow_url,
  tables_owned         = tables_owned)

build_relationships_json(
  rels = mets_rels, output_dir = dir_parquet, provider = provider, dataset = dataset)
```

## Export Modified Dependency Deltas

```{r}
#| label: export_new_deltas

for (tbl in modifies_tables) {
  pk_col   <- modifies_pks[[tbl]]$pk_col
  keys_old <- modifies_pks[[tbl]]$keys
  keys_new <- dbGetQuery(con, glue("SELECT {pk_col} FROM {tbl}"))[[1]]
  additions <- setdiff(keys_new, keys_old)
  if (length(additions) > 0) {
    pq_path <- file.path(dir_parquet, paste0(tbl, "_new.parquet"))
    vals <- paste(dbQuoteString(con, additions), collapse = ", ")
    export_parquet(con, glue("SELECT * FROM {tbl} WHERE {pk_col} IN ({vals})"), pq_path)
    cat(glue("{length(additions)} new {tbl} row(s) -> {tbl}_new.parquet"), "\n")
  }
}
```

## Upload to GCS

```{r}
#| label: upload_gcs
#| eval: true

knitr::opts_chunk$set(eval = TRUE)
if (parquet_complete) {
  cat("Parquet unchanged — skipping GCS sync", "\n")
} else {
  sync_to_gcs(local_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"),
              bucket = "calcofi-db", delete_stale = TRUE)
}
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
if (file_exists(db_checkpoint)) {
  file_delete(db_checkpoint)
  cat(glue("Removed checkpoint: {db_checkpoint}"), "\n")
}
```

::: {.callout-caution collapse="true"}
## Session Info
```{r session_info}
devtools::session_info()
```
:::
