---
title: "Ingest SWFSC Invertebrates"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: NOAA CoastWatch ERDDAP

- [erdCalCOFIinvcnt](https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIinvcnt.html) — invertebrate counts
- [erdCalCOFIinvsiz](https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIinvsiz.html) — invertebrate sizes

Counts and sizes of invertebrates (primarily squids) captured in CalCOFI
ichthyoplankton nets, processed by SWFSC. The two ERDDAP datasets share
the same sampling metadata columns; `invcnt` adds count/abundance fields,
`invsiz` adds a size measurement field.

- **Provider**: `swfsc`
- **Coverage**: 1978–2023 (counts), 1980–2023 (sizes)
- **Spatial**: Pacific coast, 0–51°N

### Data Flow

```{mermaid}
graph LR
  A[ERDDAP] --> B[CSV on Google Drive]
  B --> C[DuckDB Wrangling]
  C --> D[Load Prior ichthyo Tables]
  D --> E[Cross-Reference Tows/Nets]
  E --> F[Parquet Export]
  F --> G[GCS Archive]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI / calcofi4db,
  CalCOFI / calcofi4r,
  DBI,
  dm,
  dplyr,
  DT,
  fs,
  glue,
  here,
  janitor,
  jsonlite,
  readr,
  rerddap,
  sf,
  stringr,
  tibble,
  tidyr,
  quiet = T
)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# define paths
dataset_name <- "SWFSC CalCOFI Invertebrates"
provider <- "swfsc"
dataset <- "inverts"
dir_label <- glue("{provider}_{dataset}")
dir_data <- "~/My Drive/projects/calcofi/data-public"
dir_csv <- path_expand(glue("{dir_data}/{provider}/{dataset}"))
overwrite <- TRUE
dir_parquet <- here(glue("data/parquet/{dir_label}"))
db_path <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path)) {
    file_delete(db_path)
  }
  if (dir_exists(dir_parquet)) dir_delete(dir_parquet)
}

dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Download from ERDDAP

Fetch both datasets and save as CSVs in Google Drive. Skips download if
CSV already exists (delete to re-download).

```{r}
#| label: download_erddap

erddap_url <- "https://coastwatch.pfeg.noaa.gov/erddap"
dir_create(dir_csv)

erddap_datasets <- list(
  erdCalCOFIinvcnt = "erdCalCOFIinvcnt",
  erdCalCOFIinvsiz = "erdCalCOFIinvsiz"
)

for (ds_name in names(erddap_datasets)) {
  csv_path <- path(dir_csv, glue("{ds_name}.csv"))

  if (!file_exists(csv_path)) {
    message(glue("downloading {ds_name} from ERDDAP..."))
    ds_info <- rerddap::info(erddap_datasets[[ds_name]], url = erddap_url)
    d_raw <- rerddap::tabledap(ds_info)
    write_csv(as_tibble(d_raw), csv_path)
    message(glue("  wrote {nrow(d_raw)} rows to {csv_path}"))
  } else {
    message(glue("  {ds_name}: CSV exists, skipping download"))
  }
}
```

## Read CSVs

```{r}
#| label: read_csvs

d <- read_csv_files(
  provider = provider,
  dataset = dataset,
  dir_data = dir_data,
  sync_archive = TRUE,
  metadata_dir = here("metadata")
)

message(glue("loaded {nrow(d$source_files)} tables from {d$paths$dir_csv}"))
message(glue("total rows: {sum(d$source_files$nrow)}"))
```

## Check Data Integrity

```{r}
#| label: data_integrity
#| output: asis

integrity <- check_data_integrity(
  d = d,
  dataset_name = dataset_name,
  halt_on_fail = TRUE
)

render_integrity_message(integrity)
```

## Source Files

```{r}
#| label: source_files

show_source_files(d)
```

## Tables & Fields

```{r}
#| label: show_redefine

show_tables_redefine(d)
show_fields_redefine(d)
```

## Load into Database

```{r}
#| label: load_to_db

tbl_stats <- ingest_dataset(
  con = con,
  d = d,
  mode = if (overwrite) "replace" else "append",
  verbose = TRUE
)

tbl_stats |>
  datatable(rownames = FALSE, filter = "top")
```

## Load Dataset Metadata

```{r}
#| label: load_dataset_metadata

d_dataset <- read_csv(here("metadata/dataset.csv"))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
```

## Load Prior Tables

Load reference tables from ichthyo ingest for cross-referencing.

```{r}
#| label: load_prior

prior <- load_prior_tables(
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  con = con,
  tables = c("ship", "cruise", "site", "tow", "net", "grid")
)

message(glue(
  "loaded {length(prior)} prior tables: {paste(prior, collapse=', ')}"
))

# show loaded prior table row counts
for (tbl in prior) {
  n <- dbGetQuery(con, glue("SELECT COUNT(*) AS n FROM {tbl}"))$n
  message(glue("  {tbl}: {n} rows"))
}
```

## Data Preview

```{r}
#| label: preview_tables

for (tbl in c("invert_count", "invert_size")) {
  cat(glue("\n### {tbl}\n\n"))
  dbGetQuery(con, glue("SELECT * FROM {tbl} LIMIT 100")) |>
    datatable(
      caption = glue("{tbl} (first 100 rows)"),
      rownames = FALSE,
      filter = "top"
    ) |>
    print()
}
```

## Validate

```{r}
#| label: validate

validation <- validate_for_release(con)

if (validation$passed) {
  message("validation passed!")
  if (nrow(validation$checks) > 0) {
    validation$checks |>
      filter(status != "pass") |>
      datatable(caption = "Validation Warnings")
  }
} else {
  cat("validation FAILED:\n")
  cat(paste("-", validation$errors, collapse = "\n"))
}
```

## Enforce Column Types

```{r}
#| label: enforce_types

type_changes <- enforce_column_types(
  con = con,
  d_flds_rd = d$d_flds_rd
)

if (nrow(type_changes) > 0) {
  type_changes |>
    datatable(caption = "Column type changes applied")
}
```

## Write Parquet

```{r}
#| label: write_parquet

parquet_stats <- write_parquet_outputs(
  con = con,
  output_dir = dir_parquet,
  tables = dbListTables(con),
  strip_provenance = FALSE
)

parquet_stats |>
  mutate(file = basename(path)) |>
  select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write_metadata

metadata_path <- build_metadata_json(
  con = con,
  d_tbls_rd = d$d_tbls_rd,
  d_flds_rd = d$d_flds_rd,
  output_dir = dir_parquet,
  tables = dbListTables(con),
  set_comments = TRUE,
  provider = provider,
  dataset = dataset,
  workflow_url = glue(
    "https://calcofi.io/workflows/ingest_swfsc_inverts.html"
  )
)

# write relationships.json
geom_tbls <- DBI::dbGetQuery(
  con,
  "SELECT DISTINCT table_name FROM information_schema.columns
   WHERE data_type = 'GEOMETRY'"
)$table_name
dm_tbls <- setdiff(dbListTables(con), geom_tbls)

tryCatch(
  {
    dm_all <- dm_from_con(
      con,
      table_names = dm_tbls,
      learn_keys = FALSE
    ) |>
      add_cc_keys()
    build_relationships_json(
      dm = dm_all,
      output_dir = dir_parquet,
      provider = provider,
      dataset = dataset
    )
  },
  error = function(e) {
    message(glue("relationships.json skipped (dm error): {e$message}"))
  }
)
```

## Upload to GCS

```{r}
#| label: upload_gcs

gcs_ingest_prefix <- glue("ingest/{dir_label}")
gcs_bucket <- "calcofi-db"

sync_results <- sync_to_gcs(
  local_dir = dir_parquet,
  gcs_prefix = gcs_ingest_prefix,
  bucket = gcs_bucket
)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
message("local wrangling database connection closed")
message(glue("parquet outputs written to: {dir_parquet}"))
message(glue("GCS outputs at: gs://{gcs_bucket}/{gcs_ingest_prefix}/"))
```
