---
title: "Ingest SIO Cetacean Sonobuoy Detections"
calcofi:
  target_name: ingest_sio_cetacean_sonobuoy
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/sio_cetacean-sonobuoy/manifest.json
  provider: sio
  dataset: cetacean-sonobuoy
  # staged, not released, until the licence (Q01) and citation (Q02) are settled
  in_release: false
  workflow_url: https://calcofi.io/workflows/ingest_sio_cetacean-sonobuoy.html
  questions_file: metadata/sio/cetacean-sonobuoy/questions.csv
  dataset_meta:
    dataset_name: CalCOFI Sonobuoy Baleen Whale Call Detections
    dataset_name_short: "Whale calls (sonobuoy)"
    category: "Seabirds & Marine Mammals"
    color: "#3f8f8b"
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (deployment + analysed hour)"}
    - {table: obs, shared: true, note: "core occurrence headline (acoustic_presence per hour x species)"}
    - {table: obs_attribute, shared: true, note: "call types heard per hour x species"}
    - {table: taxon, shared: true, note: "shared taxa reference (blue, fin, humpback whale)"}
    - {table: dataset_taxon, shared: true, note: "species -> taxon_key crosswalk"}
    - {table: taxon_group, shared: true, note: "calcofi:marine_mammals"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#d3ebe9"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: `data/acoustic_data/acoustic-raw.csv` in
[CalCOFI/marmam-app](https://github.com/CalCOFI/marmam-app), read from a pinned
commit. It is the "Everything" sheet of the lab's workbook
`CalCOFI data all cruises, calculations.xlsx`: one row per analysed **hour** of a
sonobuoy recording (3,244 hours, 962 deployments, 35 cruises, 2004–2012). An
analyst scanned each hour as 60-second spectrogram windows and flagged eight call
types: blue whale B and D calls, four fin whale call types, humpback song and
non-song calls.

The app's `acoustic_detections_v2.csv` / `acoustic_station_v2.csv` are **not**
read: they are per-station summaries derived from this sheet.

- **Provider**: `sio` (Scripps Whale Acoustics Lab, Baumann-Pickering; Q03)
- **Grain**: an analysed hour (`scan`), the child of its sonobuoy `deployment`
- **Occurrence**: `acoustic_presence` = 1 when any of the species' call types was
  heard in the hour, 0 when none was; the call types heard go to `obs_attribute`
- **Position**: the sheet records the CalCOFI line/station, not the sonobuoy's
  position, so the nominal station position is used (`cc_calcofi_to_lonlat()`)

```{mermaid}
graph LR
  R[acoustic-raw.csv<br/>one row per hour] --> SC[sonobuoy_scan]
  SC --> DP[sonobuoy_deployment<br/>cruise x SB#]
  SC --> CL[_scan_call<br/>hour x call type 0/1]
  SC -.deployment_key.-> DP
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))
source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_sio_cetacean-sonobuoy.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
tables_owned <- cc$tables_owned
ds_key       <- glue("{provider}_{dataset}")
dir_label    <- ds_key
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
dir_stage    <- cc_stage_path("parquet", dir_label, create = TRUE)
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_meta     <- here(glue("metadata/{provider}/{dataset}"))
dir_dl       <- here(glue("data/cache/{dir_label}")); dir_create(dir_dl)
publish_to_gcs <- Sys.getenv("CALCOFI_SKIP_GCS") == ""

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

```{r}
#| label: read-source

# CalCOFI/marmam-app at a pinned commit (see ingest_sio_cetacean-sightings.qmd)
src_sha  <- "7bc0e18d7cb0ac7ba58d7bc4d47e57c50bd9bc80"
src_url  <- glue("https://raw.githubusercontent.com/CalCOFI/marmam-app/{src_sha}/data/acoustic_data/acoustic-raw.csv")
src_file <- file.path(dir_dl, "acoustic-raw.csv")
downloaded <- overwrite_all || !file_exists(src_file)
if (downloaded) download.file(src_url, src_file, quiet = TRUE, mode = "wb")
src_stamp <- stamp_source_access(files = src_file) |>
  mutate(source = as.character(src_url), method = if (downloaded) "download" else "file_mtime")

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_dl, gcs_prefix = glue("archive/{provider}/{dataset}"),
              bucket = "calcofi-files-public")
} else {
  cat("CALCOFI_SKIP_GCS set -- source NOT archived to gs://calcofi-files-public\n")
}

# UTF-8 with a BOM (readr strips it); every column as text, typed below
d_raw <- read_csv(src_file, col_types = cols(.default = "c"))
n_blank <- sum(is.na(d_raw$Cruise))
d_raw <- d_raw |> filter(!is.na(Cruise))   # one fully empty trailing row
stopifnot(
  "row count changed at this commit" = nrow(d_raw) == 3244L,
  "only the one blank row is dropped" = n_blank == 1L)
cat(glue("acoustic-raw: {nrow(d_raw)} analysed hours ({n_blank} blank row dropped)"), "\n")
```

## Build Scan + Deployment Tables

```{r}
#| label: scan

num <- function(x) suppressWarnings(as.numeric(x))

# the sheet rounds the fractional CalCOFI lines (83.3 -> 83); undo it (Q06).
# 60/70/80/90 are exact; 76, 82, 86 and 92 are kept as recorded.
line_fix <- c(`63` = 63.3, `67` = 66.7, `73` = 73.3, `77` = 76.7,
              `83` = 83.3, `87` = 86.7, `93` = 93.3)

# the eight call-type columns: which species each belongs to, and its label
d_call <- tribble(
  ~src_col,             ~call_col,            ~species_code, ~call_label,
  "Blue B",             "blue_b",             "blue",        "blue whale B call",
  "Blue D",             "blue_d",             "blue",        "blue whale D call",
  "Fin 20Hz Singular",  "fin_20hz_singular",  "fin",         "fin whale 20 Hz pulse (singular)",
  "Fin 20Hz Song",      "fin_20hz_song",      "fin",         "fin whale 20 Hz song",
  "Fin C.C.",           "fin_cc",             "fin",         "fin whale C.C. call",
  "Fin H.C.",           "fin_hc",             "fin",         "fin whale H.C. call",
  "Humpback Song",      "humpback_song",      "humpback",    "humpback whale song",
  "Humpback Non-song",  "humpback_nonsong",   "humpback",    "humpback whale non-song call")

d_scan <- d_raw |>
  mutate(scan_id = row_number()) |>
  rename_with(\(x) d_call$call_col[match(x, d_call$src_col)], all_of(d_call$src_col)) |>
  transmute(
    scan_id,
    cruise_label        = str_trim(Cruise),
    season              = str_trim(Season),
    sonobuoy_label      = str_trim(`SB#`),
    deployment_key      = paste(cruise_label, sonobuoy_label, sep = "_"),
    sonobuoy_type       = str_trim(`D/O?`),
    line_raw            = num(`CC Line`),
    line                = coalesce(unname(line_fix[as.character(line_raw)]), line_raw),
    station             = num(`CC Station`),
    date_raw            = Date,
    hour_raw            = Hour,
    # the hour that was analysed: its leading number, because Hour is written
    # "15:00:00", "22:00" and twice "22:000:00"; minutes are always 00.
    # Date + hour read as UTC (Q04)
    datetime_start_utc  = mdy(Date, tz = "UTC") +
                            hours(as.integer(str_extract(Hour, "^\\d{1,2}"))),
    # verbatim grade, whitespace-trimmed and lower-cased ("HIgh", "High  "):
    # the markers * / ** are kept, uninterpreted (Q05)
    recording_quality   = na_if(str_to_lower(str_trim(Quality)), ""),
    background_level    = str_trim(`Sarah's scale (amt of background)`),
    background_types    = str_trim(`Types of Background`),
    detections_flag     = str_to_lower(str_trim(`Detections?`)),
    species_detected_raw = str_trim(`Spp. Detected`),
    across(all_of(d_call$call_col), \(x) as.integer(num(x))),
    start_end_note      = `Start/End`,
    hour_span           = num(`Hour spans`),
    comments            = Comments,
    scan_method         = `Scan Method`,
    time_window         = `Time window`,
    freq_top            = `Top freq`,
    freq_low            = `Low Freq`,
    fft_length          = num(`FFT Length`),
    fft_overlap         = Overlap) |>
  mutate(
    site_key = if_else(!is.na(line) & !is.na(station),
                       normalize_site_key(sprintf("%.1f %.1f", line, station)), NA_character_))
# nominal station position: the sheet has no sonobuoy position
xy <- cc_calcofi_to_lonlat(d_scan$line, d_scan$station)
d_scan$longitude <- xy$longitude
d_scan$latitude  <- xy$latitude

stopifnot(
  "every hour has a time"                 = !anyNA(d_scan$datetime_start_utc),
  "every call flag is 0 or 1"             = all(unlist(d_scan[d_call$call_col]) %in% c(0L, 1L)),
  "detections_flag is yes / no / adhoc only" =
    all(d_scan$detections_flag %in% c("yes", "no", "adhoc only")),
  "an 'adhoc only' hour carries no call flag (Q08)" =
    all(rowSums(d_scan[d_scan$detections_flag == "adhoc only", d_call$call_col]) == 0),
  "a position exactly where there is a line and station" =
    identical(is.na(d_scan$latitude), is.na(d_scan$line) | is.na(d_scan$station)) &&
    identical(is.na(d_scan$site_key), is.na(d_scan$latitude)))
dbWriteTable(con, "sonobuoy_scan", d_scan, overwrite = TRUE)

# a deployment is one sonobuoy on one cruise; its station and type are those of
# its first analysed hour (one deployment changes station mid-record)
d_deployment <- d_scan |>
  arrange(deployment_key, datetime_start_utc) |>
  group_by(deployment_key) |>
  summarise(
    cruise_label   = first(cruise_label), sonobuoy_label = first(sonobuoy_label),
    sonobuoy_type  = first(sonobuoy_type),
    line = first(line), station = first(station), site_key = first(site_key),
    latitude = first(latitude), longitude = first(longitude),
    datetime_start_utc = min(datetime_start_utc),
    n_hours = n(), .groups = "drop")
dbWriteTable(con, "sonobuoy_deployment", d_deployment, overwrite = TRUE)

# long form of the call flags: one row per hour x call type
d_scan_call <- d_scan |>
  select(scan_id, all_of(d_call$call_col)) |>
  pivot_longer(-scan_id, names_to = "call_col", values_to = "heard") |>
  left_join(d_call |> select(call_col, species_code, call_label), by = "call_col")
dbWriteTable(con, "_scan_call", d_scan_call, overwrite = TRUE)

cat(glue("sonobuoy_scan: {nrow(d_scan)} hours; sonobuoy_deployment: {nrow(d_deployment)}; ",
         "{sum(is.na(d_scan$site_key))} hours without a station (Q10); ",
         "{sum(d_scan$detections_flag == 'adhoc only')} 'adhoc only' hours (Q08)"), "\n")
```

The free-text `Spp. Detected` column is kept for audit only; the call-type flags
decide presence (Q09). The hours where the two disagree:

```{r}
#| label: spp-crosscheck

d_xchk <- d_scan_call |>
  group_by(scan_id, species_code) |>
  summarise(flag = as.integer(max(heard) > 0), .groups = "drop") |>
  left_join(d_scan |> select(scan_id, cruise_label, sonobuoy_label, species_detected_raw), by = "scan_id") |>
  mutate(text = as.integer(str_detect(str_to_lower(coalesce(species_detected_raw, "")), species_code))) |>
  filter(flag != text)
cat(glue("Spp. Detected disagrees with the call flags on {nrow(d_xchk)} hour x species rows ",
         "({paste(names(table(d_xchk$species_code)), table(d_xchk$species_code), sep = ': ', collapse = '; ')})"), "\n")
d_xchk |> datatable(caption = "Spp. Detected vs call flags (flags win; Q09)", rownames = FALSE)
```

## Resolve cruise_key + Spatial

The cruise label (`CC1210NH`) is a year-month plus a ship/leg suffix, not a
cruise key. As for the sightings, each hour is matched to the nearest CalCOFI
station occupied within a day on the `swfsc_ichthyo` track, with a majority vote
per label; the vote reaches the hours without a station too.

```{r}
#| label: spatial-cruise

load_prior_tables(con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
                  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
for (tbl in c("sonobuoy_deployment", "sonobuoy_scan")) {
  add_point_geom(con, tbl, lon_col = "longitude", lat_col = "latitude")
  assign_grid_key(con, tbl)
}

dbExecute(con, glue(
  "CREATE OR REPLACE VIEW cruise_track AS
   SELECT DISTINCT cruise_key, datetime, latitude, longitude
   FROM read_parquet('{cc_stage_path('parquet','swfsc_ichthyo','sample.parquet')}')
   WHERE cruise_key IS NOT NULL AND datetime IS NOT NULL"))

cruise_match <- match_cruise_by_track(
  con, "sonobuoy_scan", "cruise_track",
  datetime_col = "datetime_start_utc", lon_col = "longitude", lat_col = "latitude",
  ref_datetime_col = "datetime", group_col = "cruise_label", window_days = 1)

dbExecute(con, "ALTER TABLE sonobuoy_deployment ADD COLUMN IF NOT EXISTS cruise_key VARCHAR")
dbExecute(con, "
  UPDATE sonobuoy_deployment d SET cruise_key = (
    SELECT ANY_VALUE(s.cruise_key) FROM sonobuoy_scan s
    WHERE s.cruise_label = d.cruise_label)")

grp  <- cruise_match$groups
dupe <- grp$cruise_key[!is.na(grp$cruise_key)]
dupe <- dupe[duplicated(dupe)]
stopifnot(
  "cruise_key must resolve on >90% of hours" = cruise_match$pct > 90,
  "every cruise label must resolve to a distinct cruise_key" = length(dupe) == 0)
cat(glue("cruise_key resolved on {cruise_match$matched}/{cruise_match$total} hours ",
         "({cruise_match$pct}%); unresolved labels: ",
         "{paste(grp$cruise_label[is.na(grp$cruise_key)], collapse = ', ')}"), "\n")
grp |>
  datatable(caption = "Cruise label -> cruise_key (vote share; NULL = unresolved)",
            rownames = FALSE, filter = "top") |>
  formatPercentage("share", 1)
```

## Load Dataset Metadata + Schema

```{r}
#| label: meta-schema

this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot(nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)

d_meas_type <- read_measurement_type(here("metadata/measurement_type.csv"))
stopifnot("every type this ingest emits is registered" =
            all(c("acoustic_presence", "call_type") %in% d_meas_type$measurement_type))
dbWriteTable(con, "measurement_type", as.data.frame(d_meas_type), overwrite = TRUE)

# every recording-quality grade must be a registered code (measurement_qual.csv)
d_qual <- read_csv(here("metadata/measurement_qual.csv"), col_types = cols(.default = "c")) |>
  filter(code_set == "cetacean-sonobuoy")
stopifnot("every recording_quality is a registered code" =
            all(na.omit(d_scan$recording_quality) %in% d_qual$qual_code))

src_rels <- list(
  primary_keys = list(sonobuoy_scan = "scan_id", sonobuoy_deployment = "deployment_key"),
  foreign_keys = list(
    list(table = "sonobuoy_scan", column = "deployment_key",
         ref_table = "sonobuoy_deployment", ref_column = "deployment_key")))
cc_erd(con, tables = c("sonobuoy_deployment", "sonobuoy_scan", "dataset"),
       rels = src_rels,
       colors = list(lightblue = c("sonobuoy_deployment", "sonobuoy_scan"), white = "dataset"))
```

## Questions for Data Providers

```{r}
#| label: provider-questions

questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the SIO sonobuoy data providers (ranked)")
```

## Emit Core Tables

| source | core | transform |
|---|---|---|
| `sonobuoy_deployment` | `sample` (`deployment`) | key `sio_cetacean-sonobuoy:deployment:<cruise_label>_<SB#>`; nominal station position; first analysed hour |
| `sonobuoy_scan` | `sample` (`scan`) | parent = its deployment |
| call flags, per species | `obs` `acoustic_presence` | 1 if any of the species' call types was heard, else 0; only hours scanned systematically (`Detections?` yes/no, not 'adhoc only', Q08); `measurement_qual` = `recording_quality` |
| call flags = 1 | `obs_attribute` `call_type` | `bin_label` = the call type, `count` = 1 |

**Not published** (kept in the archived source): background level and types,
`Spp. Detected`, start/end notes, hour spans, comments, and the spectrogram
settings (scan method, window, frequencies, FFT).

### Taxon vocabulary

```{r}
#| label: taxon-stage

mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

d_vocab <- tribble(
  ~ds_taxa_code, ~ds_scientific_name,       ~ds_common_name,
  "blue",        "Balaenoptera musculus",   "blue whale",
  "fin",         "Balaenoptera physalus",   "fin whale",
  "humpback",    "Megaptera novaeangliae",  "humpback whale")
n_staged <- append_dataset_taxon(con, ds_key, d_vocab)
ensure_taxon_xref(con, mt_taxon, tx_over, cache_csv = here("metadata/taxon_xref.csv"))
ensure_taxon_lineage(con, mt_taxon, tx_over, cache_csv = here("metadata/taxon_lineage.csv"))
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- resolve_dataset_taxon(con, mt_taxon, tx_over)
n_tx_group <- build_taxon_group(con, read_taxon_group_rules(here("metadata/taxon_group.csv")))
chk <- check_dataset_taxon(con, ds_key, codes = d_vocab$ds_taxa_code)

dt <- dbGetQuery(con, "SELECT ds_taxa_code, taxon_key FROM dataset_taxon WHERE dataset_key = ?",
                 params = list(ds_key))
stopifnot(
  "blue, fin, humpback key their WoRMS species" =
    setequal(dt$taxon_key, c("worms:137090", "worms:137091", "worms:137092")))
cat(glue("vocabulary: {n_staged} codes; taxon {n_taxon}, dataset_taxon {n_ds_taxon}"), "\n")
```

### Sample, obs, obs_attribute

```{r}
#| label: emit_core

dp_key <- function(col) ns_key(ds_key, "deployment", col)
sc_key <- function(col) ns_key(ds_key, "scan", col)

append_sample(con, sample_arm_self(
  ds_key, "sonobuoy_deployment", "deployment_key", "deployment", site_expr = "site_key"))

append_sample(con, glue("
  SELECT {sc_key('s.scan_id')} AS sample_key, 'scan' AS sample_type,
         {dp_key('s.deployment_key')} AS parent_sample_key,
         {dp_key('s.deployment_key')} AS root_sample_key,
         '{ds_key}' AS dataset_key, s.grid_key, s.site_key, s.cruise_key,
         NULL::INTEGER AS order_occ, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP) AS datetime,
         0::DOUBLE AS depth_min_m, 0::DOUBLE AS depth_max_m, NULL::VARCHAR AS tow_type
  FROM sonobuoy_scan s"))

# obs -- one row per systematically scanned hour x species: heard (1) or not (0)
append_obs(con, glue("
  SELECT 'bio', '{ds_key}', {sc_key('s.scan_id')},
         s.grid_key, s.cruise_key, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP), 0::DOUBLE, 0::DOUBLE,
         dt.taxon_key, NULL::VARCHAR, 'acoustic_presence',
         CAST(MAX(c.heard) AS DOUBLE), s.recording_quality, NULL::DOUBLE
  FROM _scan_call c JOIN sonobuoy_scan s USING (scan_id)
  JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}' AND dt.ds_taxa_code = c.species_code
  WHERE s.detections_flag IN ('yes', 'no')
  GROUP BY s.scan_id, s.grid_key, s.cruise_key, s.latitude, s.longitude,
           s.datetime_start_utc, s.recording_quality, dt.taxon_key"))

# obs_attribute -- the call types heard
append_obs_attribute(con, glue("
  SELECT '{ds_key}', {sc_key('s.scan_id')}, dt.taxon_key, NULL::VARCHAR, 'call_type',
         NULL::DOUBLE, c.call_label, 1, NULL::VARCHAR
  FROM _scan_call c JOIN sonobuoy_scan s USING (scan_id)
  JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}' AND dt.ds_taxa_code = c.species_code
  WHERE s.detections_flag IN ('yes', 'no') AND c.heard = 1"))

q1 <- function(sql) dbGetQuery(con, sql)[[1]]
n_sys <- sum(d_scan$detections_flag %in% c("yes", "no"))
n_heard <- d_scan_call |>
  filter(scan_id %in% d_scan$scan_id[d_scan$detections_flag %in% c("yes", "no")])
core <- list(
  sample_deployment = q1("SELECT COUNT(*) FROM sample WHERE sample_type = 'deployment'"),
  sample_scan       = q1("SELECT COUNT(*) FROM sample WHERE sample_type = 'scan'"),
  obs               = q1("SELECT COUNT(*) FROM obs"),
  obs_present       = q1("SELECT COUNT(*) FROM obs WHERE measurement_value = 1"),
  obs_attribute     = q1("SELECT COUNT(*) FROM obs_attribute"))
cat(glue("core projection -- ", paste(names(core), unlist(core), sep = "=", collapse = " ")), "\n")

stopifnot(
  "one deployment sample per deployment" = core$sample_deployment == nrow(d_deployment),
  "one scan sample per hour"             = core$sample_scan == nrow(d_scan),
  "three obs per systematic hour"        = core$obs == 3L * n_sys,
  "presence rows = hour x species with a call heard" =
    core$obs_present == nrow(distinct(filter(n_heard, heard == 1L), scan_id, species_code)),
  "call_type rows = call flags set on systematic hours" =
    core$obs_attribute == sum(n_heard$heard),
  "a presence of 1 has a call type, and a call type has a presence of 1" =
    q1("SELECT COUNT(*) FROM obs o
        FULL JOIN (SELECT DISTINCT sample_key, taxon_key FROM obs_attribute) a
          USING (sample_key, taxon_key)
        WHERE (o.measurement_value = 1) IS DISTINCT FROM (a.sample_key IS NOT NULL)") == 0,
  "every obs.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key) WHERE s.sample_key IS NULL") == 0,
  "every sample.parent_sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM sample c LEFT JOIN sample p ON p.sample_key = c.parent_sample_key
        WHERE c.parent_sample_key IS NOT NULL AND p.sample_key IS NULL") == 0,
  "every obs.taxon_key resolves in taxon" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN taxon t USING (taxon_key) WHERE t.taxon_key IS NULL") == 0)
dbExecute(con, "DROP TABLE _scan_call")
```

### Declared NULLs

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
lbl_sql <- paste0("'", c(grp$cruise_label[is.na(grp$cruise_key)], "_"), "'", collapse = ", ")
n_nostation_scan <- sum(is.na(d_scan$site_key))
n_nostation_dep  <- sum(is.na(d_deployment$site_key))
nullable <- tribble(
  ~table,   ~column,             ~n,  ~reason,
  "sample", "site_key",          n_nostation_scan + n_nostation_dep,
    "opportunistic deployments with no line/station (Q10)",
  "sample", "latitude",          n_nostation_scan + n_nostation_dep,
    "no station, so no nominal position (Q10)",
  "sample", "parent_sample_key", nrow(d_deployment),
    "deployments are roots",
  "sample", "cruise_key",        q1("SELECT COUNT(*) FROM sample WHERE cruise_key IS NULL"),
    glue("labels with no track match: {paste(grp$cruise_label[is.na(grp$cruise_key)], collapse = ', ')}"),
  "sample", "grid_key",          q1("SELECT COUNT(*) FROM sample WHERE grid_key IS NULL"),
    "no position, or a station outside the CalCOFI grid",
  "obs",    "measurement_qual",  q1("SELECT COUNT(*) FROM obs WHERE measurement_qual IS NULL"),
    "hours with a blank Quality grade")
stopifnot(
  "declared site_key / latitude NULLs match" =
    q1("SELECT COUNT(*) FROM sample WHERE site_key IS NULL") == nullable$n[1] &&
    q1("SELECT COUNT(*) FROM sample WHERE latitude IS NULL") == nullable$n[2],
  "the only roots are the deployments" =
    q1("SELECT COUNT(*) FROM sample WHERE parent_sample_key IS NULL") == nullable$n[3],
  "every NULL cruise_key is an unresolved label" =
    q1(glue("SELECT COUNT(*) FROM sonobuoy_scan WHERE cruise_key IS NULL
             AND cruise_label NOT IN ({lbl_sql})")) == 0,
  "a blank grade is the only NULL qual (<= 14 source rows)" =
    nullable$n[6] <= 3L * 14L)
cat("validate_for_release:", ifelse(results$passed, "PASSED", "reported findings -- reconciled against the declared NULLs below"), "\n")
nullable |> datatable(caption = "Declared NULLs (count and reason)", rownames = FALSE)
```

### Measurement bounds

```{r}
#| label: bounds

b <- check_measurement_bounds(con, "obs", mt = d_meas_type, dataset_key = ds_key)
bounds_datatable(b)
stopifnot("acoustic_presence is declared and respected" = all(b$status == "ok"))
```

## Preview

```{r}
#| label: preview

dbGetQuery(con, "
  SELECT t.scientific_name, year(o.datetime) AS year,
         COUNT(*) AS n_hours, SUM(o.measurement_value) AS n_hours_heard,
         ROUND(AVG(o.measurement_value), 3) AS frac_heard
  FROM obs o JOIN taxon t USING (taxon_key)
  GROUP BY ALL ORDER BY 1, 2") |>
  datatable(caption = "Fraction of systematically scanned hours with calls, by species and year",
            rownames = FALSE)
```

```{r}
#| label: erd-core

core_tbls <- core_output_tables(con)
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls))
```

## Write Outputs + Upload

```{r}
#| label: outputs

dir_create(dir_parquet)
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
write_parquet_outputs(
  con = con, output_dir = dir_parquet, tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)

d_tbls_rd <- read_csv(file.path(dir_meta, "tbls_redefine.csv"))
d_flds_rd <- read_csv(file.path(dir_meta, "flds_redefine.csv"))
build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           file.path(dir_meta, "metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned,
  sources = src_stamp)

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet,
              gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else {
  cat("CALCOFI_SKIP_GCS set -- parquet kept local, NOT synced to gs://calcofi-db\n")
}
```

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
