---
title: "Ingest Farallon Institute Bird & Mammal Census"
calcofi:
  target_name: ingest_farallon_bird_mammal
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/farallon_bird-mammal/manifest.json
  provider: farallon
  dataset: bird-mammal
  workflow_url: https://calcofi.io/workflows/ingest_farallon_bird-mammal.html
  questions_file: metadata/farallon/bird-mammal/questions.csv
  dataset_meta:
    dataset_name: CalCOFI Bird & Mammal Census
    # display trio, read by the release `dataset` table and the
    # consumer apps (calcofi4db >= 3.15.0) — see NEWS for why these
    # left the apps' own hardcoded maps
    dataset_name_short: "Seabirds & Marine Mammals"
    category: "Seabirds & Marine Mammals"
    color: "#ff8787"
  # this ingest publishes the consolidated core, not per-dataset tables: the
  # transect/observation/species/behavior source shape is wrangled in the
  # notebook and projected into sample / obs / obs_attribute + the taxa refs.
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (transect grain)"}
    - {table: obs, shared: true, note: "core occurrence headline (bio, per transect x species)"}
    - {table: obs_attribute, shared: true, note: "behavior breakdown per occurrence"}
    - {table: taxon, shared: true, note: "shared taxa reference (seabirds itis:, mammals worms:)"}
    - {table: dataset_taxon, shared: true, note: "species_code -> taxon_key crosswalk"}
    - {table: taxon_group, shared: true, note: "calcofi:seabirds / calcofi:marine_mammals"}
  erd:
    color: "#f0d2e8"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: transect (effort), observation (count), and species tables come from
NOAA ERDDAP (`CAC_FI_SBAS_tr` / `CAC_FI_SBAS_obs` / `CAC_FI_SBAS_sp`,
oceanview.pfeg.noaa.gov) as of 2026-08-25 — back online per Ben, superseding the
CCE-LTER DataZoo 255 local export (`whales-seabirds-turtles/bird-mammal-census/`)
for these three tables. They are **downloaded first** into `data/cache/` and
archived to GCS, so a run reads a file it can hash and diff rather than a live
query (see Read Source Data). Behavior lookup CONFIRMED not on ERDDAP (checked
directly against oceanview, 404) and stays DataZoo/local-sourced. Bird & mammal
observations along CalCOFI/NMFS/CPR cruise transects, 1987–2022.

What the switch changes in the rows (measured 2026-09-03 against the DataZoo
export): the two sources are **identical for 1987–2018** (60,715 shared
transects, every observation row equal); ERDDAP **adds 2019, 2020 and 2022**
(3,216 transects, 6,020 observation rows); ERDDAP **has no observations at all
for 2021** although it carries 956 transects for the two 2021 cruises, and
DataZoo published 625 observations for `CAC2021_7` — filed as Q11 and taken as
served, not patched from DataZoo.

- **Provider**: `calcofi` (tentative — curated via CCE-LTER; see questions)
- **Tables**: `bird_mammal_transect` (effort, **ERDDAP**) ⨝
  `bird_mammal_observation` (counts, **ERDDAP**) on `gis_key`, plus
  `bird_mammal_species` (**ERDDAP** `_sp`, with the ITIS TSN and include /
  unidentified flags joined back from the committed DataZoo list — see Build
  Lookups) and `bird_mammal_behavior` (**local**) lookups.
- **Taxonomy**: the species vocabulary is *declared* here with
  `calcofi4db::append_dataset_taxon()` and *resolved* by the package against
  WoRMS/ITIS (taxon plan D1–D3, calcofi4db ≥ 3.29.0); see Emit Core Tables.

```{mermaid}
graph LR
  T[transects] --> TR[bird_mammal_transect<br/>effort + position]
  O[observations] --> OB[bird_mammal_observation<br/>counts]
  OB -.gis_key.-> TR
  OB -.species_code.-> SP[bird_mammal_species]
  OB -.behavior_code.-> BH[bird_mammal_behavior]
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, units, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))
source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_farallon_bird-mammal.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
dir_stage  <- cc_stage_path("parquet", dir_label, create = TRUE)
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))
# the DataZoo 255 export on Drive: the behavior lookup is read from it, and its
# archived species list is what metadata/farallon/bird-mammal/
# species_itis_datazoo.csv was derived from -- transect, observation and species
# data come from ERDDAP, not this dir (see Read Source Data)
dir_src      <- path_expand(glue("{dir_data}/whales-seabirds-turtles/bird-mammal-census"))
# ERDDAP snapshots land here first (gitignored; archived to GCS below), so the
# notebook reads bytes it can name and hash rather than a live query: a re-run
# reuses them, `overwrite_all` (libs/ingest.R) refetches
dir_dl       <- here(glue("data/cache/{dir_label}")); dir_create(dir_dl)
# DataZoo's species list, committed once: the ITIS TSN per code plus the
# include / unidentified flags that ERDDAP's species table dropped (Build Lookups)
f_datazoo    <- here("metadata/farallon/bird-mammal/species_itis_datazoo.csv")
# GCS uploads (source archive + parquet mirror). Flip to FALSE to render without
# publishing -- a review render, or a machine without gcloud credentials.
publish_to_gcs <- TRUE

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

```{r}
#| label: read-source

# transects + observations + species: NOAA ERDDAP, not the DataZoo local export
# (Ben, 2026-08-25 — "the recently ascribed source dataset ... appears to be back
# online"). oceanview is the host already used by dataset_meta.link_data_source
# and by the ichthyo tow links; coastwatch serves the same datasets but was
# unreachable outright when last checked (2026-08-24). Column names/order/
# units-row shape confirmed 2026-09-02 against real rows pulled from oceanview.
#
# Download-first: each table is fetched into dir_dl once (skipped while the file
# exists; overwrite_all refetches), archived to GCS beside the DataZoo export,
# and read from disk. A live read_csv(url) leaves nothing to diff when the
# provider changes something upstream -- and they do (Q11).
erddap_host <- "https://oceanview.pfeg.noaa.gov/erddap/tabledap"
erddap_ids  <- c(tr = "CAC_FI_SBAS_tr", obs = "CAC_FI_SBAS_obs", sp = "CAC_FI_SBAS_sp")
erddap_urls <- setNames(as.character(glue("{erddap_host}/{erddap_ids}.csv")), names(erddap_ids))
erddap_csvs <- setNames(file.path(dir_dl, glue("{erddap_ids}.csv")), names(erddap_ids))
downloaded  <- character()
for (k in names(erddap_ids)) {
  if (overwrite_all || !file_exists(erddap_csvs[[k]])) {
    download.file(erddap_urls[[k]], erddap_csvs[[k]], quiet = TRUE)
    downloaded <- c(downloaded, k)
  }
}
# when the bytes came down: the cached file's mtime IS the download time, for a
# fetch in this run (`download`) as much as for one a previous run made
# (`file_mtime`); the recorded source is the URL, not a path on this machine.
# build_metadata_json(sources = ) writes it to metadata.json as sources[] and
# the release takes the newest stamp as this dataset's measured source_accessed.
src_stamp <- stamp_source_access(files = c(unname(erddap_csvs), file.path(
  dir_src, "CalCOFI_bird-mammal-census_behaviorcodes.csv"))) |>
  mutate(source = c(unname(erddap_urls),
                    "gs://calcofi-files-public/archive/farallon/bird-mammal/CalCOFI_bird-mammal-census_behaviorcodes.csv"),
         method = c(ifelse(names(erddap_ids) %in% downloaded, "download", "file_mtime"), "file_mtime"))
cat(glue("ERDDAP: {length(downloaded)} table(s) downloaded this run, ",
         "{length(erddap_ids) - length(downloaded)} reused from {dir_dl}"), "\n")

read_erddap <- function(f) {
  # ERDDAP's .csv puts a units row directly under the header; drop it. Reading
  # by name (not position) so a column reorder on the provider's end doesn't
  # silently misalign fields the way a positional/.csv0 read would.
  read_csv(f, col_types = cols(.default = "c")) |> slice(-1)
}
d_tr <- read_erddap(erddap_csvs[["tr"]])   # time, latitude, longitude, gis_key,
                                            # transect, bin, date, datedate,
                                            # numeric_time, length_km, width_km,
                                            # area_km2, year, month, cruise,
                                            # season, core
d_ob <- read_erddap(erddap_csvs[["obs"]])  # idnum, gis_key, behavior, species, count

# species lookup: CAC_FI_SBAS_sp -- species / type / common_name /
# scientific_name, and no ids. Replaces the local allspecieslist file as the
# vocabulary; what that file carried and this one does not (the ITIS TSN, the
# include / unidentified flags) is joined back in Build Lookups. The source has
# a couple of junk rows after the units row (one fully blank, one with only
# `type` populated as "b") -- filtered on `species` rather than assumed to be
# exactly two, in case the count varies on a re-pull.
d_sp <- read_erddap(erddap_csvs[["sp"]]) |> filter(!is.na(species) & species != "")

# behavior lookup: CONFIRMED not on ERDDAP (checked directly against oceanview
# 2026-09-02: CAC_FI_SBAS_beh -> 404 "Currently unknown datasetID"). Stays
# DataZoo/local-sourced -- a real, permanent gap, asked as Q10 (d).
stopifnot("source dir not found" = dir_exists(dir_src))
if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_src, gcs_prefix = glue("archive/{provider}/{dataset}"),
              bucket = "calcofi-files-public", exclude = c(".DS_Store", "*.tmp", "*.gdoc"))
  sync_to_gcs(local_dir = dir_dl, gcs_prefix = glue("archive/{provider}/{dataset}/erddap"),
              bucket = "calcofi-files-public")
} else {
  cat("publish_to_gcs is FALSE -- sources NOT archived to gs://calcofi-files-public\n")
}

d_bh <- read_csv(file.path(dir_src, "CalCOFI_bird-mammal-census_behaviorcodes.csv"),
                 col_types = cols(.default = "c")) |> clean_names()
cat(glue("transects {nrow(d_tr)} (ERDDAP), observations {nrow(d_ob)} (ERDDAP), ",
         "species {nrow(d_sp)} (ERDDAP), behaviors {nrow(d_bh)} (local)"), "\n")
cat("transect column names:", paste(names(d_tr), collapse=", "), "\n")
```

## Build Lookups (species, behavior)

```{r}
#| label: species-datazoo

# DataZoo 255's species list carried what ERDDAP's CAC_FI_SBAS_sp does not: an
# ITIS TSN per code and the include / unidentified flags. It is committed ONCE
# as metadata/farallon/bird-mammal/species_itis_datazoo.csv, derived from the
# archived CalCOFI_bird-mammal-census_allspecieslist.csv (the file the wrangling
# database's bird_mammal_species was built from until 2026-08-25; also at
# gs://calcofi-files-public/archive/farallon/bird-mammal/). Regenerated here only
# if the file is missing, so the derivation stays reproducible without the
# Drive archive being needed for a run. The TSNs are what the taxon crosswalk
# keys on (they resolve exactly, where a name may not); the include flag is the
# provider's "this is a taxon worth analysing" (land birds, fish, sharks and the
# coarsest "unidentified bird / mammal / fish" classes are not).
if (!file_exists(f_datazoo)) {
  read_csv(file.path(dir_src, "CalCOFI_bird-mammal-census_allspecieslist.csv"),
           col_types = cols(.default = "c")) |>
    transmute(
      species_code    = Species,
      common_name     = `Common Name`,
      scientific_name = `Latin Name`,
      itis_id         = suppressWarnings(as.integer(ITIS)),
      is_bird         = Bird == "1", is_mammal = Mammal == "1", is_fish = Fish == "1",
      is_unidentified = Unidentified == "1",
      include_flag    = Include == "1") |>
    arrange(species_code) |>
    write_csv(f_datazoo, na = "")
}
d_datazoo <- read_csv(f_datazoo, col_types = cols(
  itis_id = "i", is_bird = "l", is_mammal = "l", is_fish = "l",
  is_unidentified = "l", include_flag = "l", .default = "c"))
stopifnot("species_code must be unique in the DataZoo list" = !anyDuplicated(d_datazoo$species_code))
cat(glue("DataZoo species list: {nrow(d_datazoo)} codes, ",
         "{sum(!is.na(d_datazoo$itis_id))} with a TSN, ",
         "{sum(d_datazoo$include_flag)} include_flag, ",
         "{sum(d_datazoo$is_unidentified & d_datazoo$include_flag)} of those unidentified"), "\n")
```

```{r}
#| label: lookups

# ERDDAP's CAC_FI_SBAS_sp is thinner than DataZoo's allspecieslist: no itis_id,
# is_unidentified or include_flag (nor is_large_bird / nmfs_code / comment).
# `type` (bird / fish / mammal / turtle) replaces the per-species booleans. The
# three columns the taxon crosswalk needs are joined back from the committed
# DataZoo list (species-datazoo above) -- NA for a code DataZoo never listed.
#
# SBIG appears twice in the source: "Mew Gull" with no scientific name, and
# "Short-billed gull" Larus brachyrhynchus -- one code, one taxon, two rows from
# the 2021 AOS Mew Gull / Short-billed Gull split. Keep the row that carries the
# name (append_dataset_taxon() refuses a duplicate code, by design). The old
# code MEGU is the mirror image -- in the observations, absent from this table --
# and is handled where the vocabulary is staged (Emit Core Tables, Q10).
n_sbig <- sum(d_sp$species == "SBIG")
d_sp   <- d_sp |> filter(!(species == "SBIG" & is.na(scientific_name)))
stopifnot("species_code must be unique after the SBIG dedup" = !anyDuplicated(d_sp$species))
d_species <- d_sp |>
  transmute(
    species_code = species, common_name, scientific_name, type,
    is_bird = type == "bird", is_mammal = type == "mammal",
    is_fish = type == "fish", is_turtle = type == "turtle") |>
  left_join(d_datazoo |> select(species_code, itis_id, is_unidentified, include_flag,
                                 scientific_name_datazoo = scientific_name),
            by = "species_code")
dbWriteTable(con, "bird_mammal_species", d_species, overwrite = TRUE)

d_behavior <- d_bh |> transmute(behavior_code = id, behavior, description)
dbWriteTable(con, "bird_mammal_behavior", d_behavior, overwrite = TRUE)
cat(glue("species {nrow(d_species)} (SBIG {n_sbig} rows -> 1; ",
         "{sum(!is.na(d_species$itis_id))} with a DataZoo TSN, ",
         "{sum(is.na(d_species$include_flag))} not in DataZoo's list), ",
         "behaviors {nrow(d_behavior)}"), "\n")
```

## Build Transect (effort) Table

```{r}
#| label: transect

# ERDDAP's CAC_FI_SBAS_tr does NOT carry everything the DataZoo export did:
# no start/stop lat-lon (midpoint only), no bottom_depth_m, no julian_date/
# julian_day, no svy. Dropped here rather than backfilled with NA -- flag as a
# new provider question if any of these turn out to be needed downstream.
# lengths/area arrive in km/km2 (converted to m/m2 to match the prior schema);
# `core` (survey core-area flag) is new from ERDDAP and has no prior home, kept
# as core_flag in case it's useful, safe to drop if not.
d_transect <- d_tr |>
  transmute(
    gis_key,
    cruise_label    = cruise,
    transect_number = suppressWarnings(as.integer(transect)),
    bin_number      = suppressWarnings(as.integer(bin)),
    latitude        = suppressWarnings(as.numeric(latitude)),
    longitude       = suppressWarnings(as.numeric(longitude)),
    length_m        = suppressWarnings(as.numeric(length_km)) * 1000,
    width_m         = suppressWarnings(as.numeric(width_km)) * 1000,
    area_m2         = suppressWarnings(as.numeric(area_km2)) * 1e6,
    core_flag       = suppressWarnings(as.integer(core)),
    season,
    # ERDDAP's `time` is a real UTC timestamp straight from the provider -- this
    # RESOLVES Q01 (source tz previously unconfirmed under the old date +
    # time_sec reconstruction). See the cruise-match step below, now matched on
    # this column instead of date-only for the same reason.
    datetime_start_utc = suppressWarnings(ymd_hms(time, tz = "UTC")),
    # NOT parsed from `datedate` (confirmed via real sample rows to be ambiguous
    # US-style "5/5/1987 0:00" text -- as.Date() on that string silently returns
    # NA for every row under R's default ISO format). Derived from the
    # already-parsed UTC timestamp instead, which is unambiguous.
    date = as.Date(datetime_start_utc))

dbWriteTable(con, "bird_mammal_transect", d_transect, overwrite = TRUE)
cat(glue("bird_mammal_transect: {nrow(d_transect)} rows"), "\n")
```

## Build Observation (counts) Table

```{r}
#| label: observation

# idnum from ERDDAP is a real, stable source id -- used directly as
# observation_id instead of a locally generated row_number() (the old
# behavior, from when the source had no id column of its own).
#
# species_code: every code the observations use is either staged in
# dataset_taxon or listed as excluded, and check_dataset_taxon() enforces that
# (Emit Core Tables). behavior_code still matches against the local
# bird_mammal_behavior lookup exactly as before this rewrite -- unchanged, so
# not a new risk from this edit.
d_observation <- d_ob |>
  transmute(gis_key, species_code = species, behavior_code = behavior,
            count = suppressWarnings(as.integer(count)),
            observation_id = idnum)
dbWriteTable(con, "bird_mammal_observation", d_observation, overwrite = TRUE)
cat(glue("bird_mammal_observation: {nrow(d_observation)} rows"), "\n")
```

## Resolve cruise_key + Spatial

The source records a **survey label** (`CAC1987_05`), not a cruise, and carries no
ship column — so `cruise_key` has to be recovered from where the observers
actually were. Parsing year-month out of the label is not enough: it is ambiguous
whenever several ships sailed in one month (1998-10 had four), and it is wrong
outright for a survey straddling a month boundary (`CAC2014_01` ran 2014-01-29 →
02-04 and belongs to `2014-02-3322`). See Q02.

The observers ride a CalCOFI ship, so each transect sits on that day's station
track. `match_cruise_by_track()` matches every transect to the nearest station
occupied the same day, then takes the **majority vote per survey label** — one
label is one cruise — and applies the winner to all of that label's transects.

The reference track is the `swfsc_ichthyo` sample shard (already this notebook's
declared dependency). That choice matters for referential integrity: its
`cruise_key`s are exactly the `cruise` reference table's, so every key emitted
here resolves. Tracks assembled from other shards include cruises absent from
`cruise` (e.g. `2021-07-33P4`), which would ship a dangling FK.

```{r}
#| label: spatial-cruise

# geometry + grid from the transect midpoint
load_prior_tables(con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
                  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
add_point_geom(con, "bird_mammal_transect", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "bird_mammal_transect") |> datatable(caption = "Grid assignment")

# cruise track: one row per (cruise, day, station position); the match below
# is on datetime_start_utc now that ERDDAP's `time` is a UTC timestamp (Q01).
dbExecute(con, glue(
  "CREATE OR REPLACE VIEW cruise_track AS
   SELECT DISTINCT cruise_key, datetime, latitude, longitude
   FROM read_parquet('{cc_stage_path('parquet','swfsc_ichthyo','sample.parquet')}')
   WHERE cruise_key IS NOT NULL AND datetime IS NOT NULL"))

# CHANGED: matched on datetime_start_utc, not date-only -- the old date-only
# choice existed BECAUSE the prior source's tz was unconfirmed (Q01). ERDDAP's
# `time` is a real UTC timestamp, so that reason no longer holds. This is a
# genuine behavior change (day-boundary transects can now land differently) --
# verify the matched/total % below hasn't dropped before trusting it; revert
# datetime_col back to "date" if it has.
cruise_match <- match_cruise_by_track(
  con, "bird_mammal_transect", "cruise_track",
  datetime_col     = "datetime_start_utc", lon_col = "longitude", lat_col = "latitude",
  ref_datetime_col = "datetime",
  group_col        = "cruise_label")

cat(glue("cruise_key resolved on {cruise_match$matched}/{cruise_match$total} ",
         "transects ({cruise_match$pct}%)"), "\n")
```

```{r}
#| label: cruise-match-review

# surveys with no cruise sit at the top: they are the ones to eyeball, not the
# 100%-unanimous majority
cruise_match$groups |>
  datatable(caption = "Survey label -> cruise_key (vote share; NULL = unresolved)",
            rownames = FALSE, filter = "top") |>
  formatPercentage("share", 1)
```

```{r}
#| label: cruise-match-validate

grp <- cruise_match$groups
n_unresolved <- sum(is.na(grp$cruise_key))

# a survey is one cruise on one ship, so two labels claiming the same cruise
# would mean the vote collapsed two distinct surveys together
dupe <- grp$cruise_key[!is.na(grp$cruise_key)]
dupe <- dupe[duplicated(dupe)]

stopifnot(
  "cruise_key must resolve on >95% of transects" =
    cruise_match$pct > 95,
  "every survey label must resolve to a distinct cruise_key" =
    length(dupe) == 0,
  "resolved cruise_key must exist in the cruise reference (no dangling FK)" =
    dbGetQuery(con, "
      SELECT COUNT(*) FROM bird_mammal_transect t
      WHERE t.cruise_key IS NOT NULL AND t.cruise_key NOT IN (
        SELECT cruise_key FROM cruise_track)")[[1]] == 0)

cat(glue(
  "{nrow(grp) - n_unresolved}/{nrow(grp)} survey labels resolved; ",
  "{n_unresolved} left NULL: ",
  "{paste(grp$cruise_label[is.na(grp$cruise_key)], collapse=', ')}"), "\n")
```

The unresolved labels are correct as NULL, not gaps to be filled:

- **`Fronts_0711`** (2011-06-21 → 07-16) is a CCE-LTER *Fronts* process cruise;
  no CalCOFI cruise sailed in that window (2011-04 ended 04-27, the next began
  07-27), so there is no CalCOFI cruise to point at.
- **`CAC2021_7`** (2021-07-19 → 08-02) *did* ride a CalCOFI cruise —
  `2021-07-33P4`, an exact date-range match — but that cruise is missing from the
  `cruise` reference table, which is built from the ichthyo source alone.
  Assigning it would ship a dangling FK. It resolves for free once `cruise`
  covers every cruise the other datasets reference (workflows#75 — 296 of 987
  referenced keys are missing, so that gap is worth closing on its own merits
  rather than for this one survey).
- **`CAC2022_8`** (2022-08-14 → 08-29, new with the ERDDAP source) is the same
  case: the ichthyo reference carries `2022-04-3322` and `2022-10-33UD` but no
  August 2022 cruise, so its 523 transects (355 with observations) stay NULL
  until the reference does.

## Load Dataset Metadata + Schema

```{r}
#| label: meta-schema

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)

bmc_rels <- list(
  primary_keys = list(
    bird_mammal_transect = "gis_key", bird_mammal_observation = "observation_id",
    bird_mammal_species = "species_code", bird_mammal_behavior = "behavior_code"),
  foreign_keys = list(
    list(table="bird_mammal_observation", column="gis_key",       ref_table="bird_mammal_transect", ref_column="gis_key"),
    list(table="bird_mammal_observation", column="species_code",  ref_table="bird_mammal_species",  ref_column="species_code"),
    list(table="bird_mammal_observation", column="behavior_code", ref_table="bird_mammal_behavior", ref_column="behavior_code")))
# the SOURCE shape, documenting the wrangling below. The tables this ingest
# actually publishes are the consolidated core (see "Emit Core Tables"), so
# relationships.json is written there, from core_relationships().
cc_erd(con, tables = c("bird_mammal_transect","bird_mammal_observation","bird_mammal_species","bird_mammal_behavior","dataset"),
       rels = bmc_rels,
       colors = list(lightblue = c("bird_mammal_transect","bird_mammal_observation"),
                     lightyellow = c("bird_mammal_species","bird_mammal_behavior"), white = "dataset"))
```

## Validate + Preview

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
# orphan check: observations whose gis_key is not in transects
n_orphan <- dbGetQuery(con, "SELECT COUNT(*) FROM bird_mammal_observation o
  LEFT JOIN bird_mammal_transect t USING(gis_key) WHERE t.gis_key IS NULL")[[1]]
cat(glue("observation gis_key orphans (no transect): {n_orphan}"), "\n")
```

```{r}
#| label: preview

cols <- dbGetQuery(con, "SELECT column_name FROM information_schema.columns
  WHERE table_name='bird_mammal_transect' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM bird_mammal_transect LIMIT 100")) |>
  datatable(caption = "bird_mammal_transect — first 100", rownames = FALSE, filter = "top")
```

## Questions for Data Providers

```{r}
#| label: provider-questions

# one validated read + render for every ingest: the vocabulary and the column
# order live in calcofi4db, not in 16 hand-written factor() calls
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the Farallon bird/mammal data providers (ranked)")
```

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`). These core tables **are** this ingest's
output: `release_database.qmd` concatenates the per-dataset shards rather than
re-deriving the core from per-dataset tables, so there is exactly one projection
to keep correct.

The seabird/mammal source records one row per (transect, species, **behavior**).
The `obs` headline is one row per (transect, species) with `count` summed across
behaviors — otherwise the same birds are counted once per behavior code — and the
behavior breakdown becomes `obs_attribute` rows carrying the behavior label.

### Taxon vocabulary

The vocabulary is **declared** here and **resolved** by calcofi4db (taxon plan
D1–D3, ≥ 3.29.0): `append_dataset_taxon()` stages one `dataset_taxon` row per
code with `taxon_key` empty; `ensure_taxon_xref()` / `ensure_taxon_lineage()`
resolve ids and classification against WoRMS + ITIS (cached under `metadata/`);
`resolve_dataset_taxon()` mints the key from the **class** — `itis:<tsn>` exactly
when the class is Aves and an accepted TSN resolves, else `worms:<aphia>` — never
from a source flag. There is no farallon arm in the package any more; what the
arm used to hard-code is now registry rows, and `check_dataset_taxon()` halts the
render if a code the observations use is neither staged nor deliberately excluded.

What goes in, and what is left out (D3):

- every code DataZoo's `include_flag` admitted (the 156 codes released at
  v2026.08.25, minus `CSLI` and `XAMU`, which ERDDAP retired unreferenced — their
  observations carry `CASL` and `GUMU`/`SCMU`). Excluded codes keep a NULL
  `taxon_key` on their observations, as before.
- codes ERDDAP added since DataZoo (44) have no include flag. Those the
  observations reference are decided one by one in the chunk; the rest are
  neither staged nor excluded until Farallon Institute flags them (Q10). `SBIG`
  is staged although unreferenced: it is the provider's current code for the
  taxon `MEGU` records.
- `MEGU` (71 observations, absent from `CAC_FI_SBAS_sp`) is added: the pre-2021
  Mew Gull code, the same bird the source now lists as `SBIG` (Q10).
- the "Unidentified X" classes resolve to Aves `itis:174371` / Mammalia
  `worms:1837` through `metadata/taxon_override.csv` rows — the fallback that
  used to be hard-coded in the package's farallon arm.

```{r}
#| label: taxon-stage

ds_key <- "farallon_bird-mammal"
# registries. measurement_taxon.csv has no farallon rows (this dataset's taxa are
# a vocabulary, not measurement-type names) but the resolvers take it for
# uniformity, filtered to THIS dataset; taxon_override.csv is read whole and
# every helper filters to ds_key itself (a row for a dataset absent from this
# connection is that dataset's business).
mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

obs_codes <- dbGetQuery(con, "SELECT DISTINCT species_code FROM bird_mammal_observation")$species_code

# ERDDAP-only codes the observations reference, decided one by one. DataZoo's
# include flag drew the line at seabirds, marine mammals and sea turtles; the
# same line is drawn here, and the list is asserted complete so a code that
# appears on a re-pull cannot slip through undecided.
erddap_only <- tibble::tribble(
  ~species_code, ~stage, ~why,
  "GUMU", TRUE,  "Guadalupe Murrelet - seabird (the post-2012 split of Xantus's Murrelet XAMU)",
  "SCMU", TRUE,  "Scripps's Murrelet - seabird (the other half of that split)",
  "CHSP", TRUE,  "Chapman's Storm-Petrel - seabird, a Leach's subspecies",
  "TOSP", TRUE,  "Townsend's Storm-Petrel - seabird",
  "NABO", TRUE,  "Nazca Booby - seabird",
  "MABO", TRUE,  "Masked Booby - seabird",
  "LOTU", TRUE,  "Loggerhead Turtle - sea turtle, as DataZoo included GRTU / RITU",
  "UNLP", TRUE,  "Unidentified Leach's Storm-Petrel - seabird class, like the other UN** classes",
  "UNMT", TRUE,  "Unidentified Murrelet - seabird class, like the other UN** classes",
  "CRAB", FALSE, "'Crab pot' - gear, not an organism",
  "FISH", FALSE, "unidentified fish - DataZoo excluded every fish (UNFI, BLSH, MOLA, ...)",
  "TUNA", FALSE, "'Tuna' - a fish, no rank stated",
  "VEVE", FALSE, "Velella velella - a hydrozoan; the census vocabulary is birds, mammals, turtles",
  "RAPT", FALSE, "unidentified raptor - land-bird class, as DataZoo excluded UNPA / UNHU",
  "WIWA", FALSE, "Wilson's Warbler - land bird, as DataZoo excluded YRWA / BRSP / EUST")
new_ref <- d_species$species_code[is.na(d_species$include_flag) & d_species$species_code %in% obs_codes]
stopifnot(
  "every ERDDAP-only code the observations reference must be decided in erddap_only" =
    setequal(new_ref, erddap_only$species_code))
staged_codes <- d_species |>
  filter(include_flag %in% TRUE |
         species_code %in% erddap_only$species_code[erddap_only$stage] |
         species_code == "SBIG") |>
  pull(species_code)
excluded_codes <- setdiff(obs_codes, c(staged_codes, "MEGU"))

# the declaration: code, the source's name and common name, and the id the
# source supplied (DataZoo's TSN, NA for an ERDDAP-only code) -- which
# append_dataset_taxon() stores as ds_source_json, the audit value below
d_vocab <- d_species |>
  filter(species_code %in% staged_codes) |>
  transmute(ds_taxa_code = species_code, ds_scientific_name = scientific_name,
            ds_common_name = common_name, itis_id) |>
  bind_rows(tibble(
    ds_taxa_code = "MEGU", ds_scientific_name = "Larus brachyrhynchus",
    ds_common_name = "Mew Gull",
    # DataZoo's TSN (176832, Larus canus -- the pre-split name) rides along as
    # the source claim; the override row keys the code to the Short-billed
    # Gull like SBIG (Q10)
    itis_id = d_datazoo$itis_id[d_datazoo$species_code == "MEGU"]))
n_staged <- append_dataset_taxon(con, ds_key, d_vocab)

# cross-reference: resolve each taxon against BOTH authorities (cached in
# metadata/taxon_xref.csv, so a re-run costs no API calls) -- the TSN crosswalk
# where the source gave a TSN, the name where it gave nothing; fills the
# opposite authority's id as a cross-reference, replaces a deprecated id by its
# accepted form so the key is always an accepted id, and fetches the real
# taxonomic_status with the date it was checked. Must precede the lineage fetch.
ensure_taxon_xref(con, mt_taxon, tx_over,
                  cache_csv = here("metadata/taxon_xref.csv"))
# lineage: each taxon's classification (cached in metadata/taxon_lineage.csv),
# staged as the `taxon` hierarchy build_taxon_reference() reads and the source
# of the CLASS the key rule reads. Without it a taxon reaches the release with a
# key and a name and nothing else -- no rank, no parent, no classification.
ensure_taxon_lineage(con, mt_taxon, tx_over,
                     cache_csv = here("metadata/taxon_lineage.csv"))
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- resolve_dataset_taxon(con, mt_taxon, tx_over)
n_tx_group <- build_taxon_group(con, read_taxon_group_rules(here("metadata/taxon_group.csv")))

# the ingest asserts its own crosswalk (D6): every code the observations use is
# staged or deliberately excluded, every staged row keys an authority id, every
# Aves taxon keys itis:. Halts the render on any finding.
chk <- check_dataset_taxon(con, ds_key, codes = setdiff(obs_codes, excluded_codes))
cat(glue("vocabulary: {n_staged} codes staged; {length(excluded_codes)} of the ",
         "{length(obs_codes)} observed codes excluded; taxon {n_taxon}, ",
         "dataset_taxon {n_ds_taxon}, taxon_group {n_tx_group}"), "\n")
cat("excluded (their observations keep a NULL taxon_key):",
    paste(sort(excluded_codes), collapse = " "), "\n")
```

The audit: what the **source** claimed (DataZoo's TSN per code, kept in
`dataset_taxon.ds_source_json`) against what the **authority** says
(`taxon.itis_id` of the key the code resolved to). Every differing row must be
one of three things, and the table says which: a deprecated TSN re-keyed onto
its accepted form (`taxon.notes` records it), an override row
(`taxon_override.csv`, with its note), or a taxon the source gave no TSN for.
Never a silent pick.

```{r}
#| label: taxon-audit

tx_over_ds <- tx_over |> filter(dataset_key == ds_key)
audit <- dbGetQuery(con, "
  SELECT d.ds_taxa_code, d.ds_scientific_name, d.ds_common_name,
         json_extract(d.ds_source_json, '$.itis_id')::INTEGER AS source_itis_id,
         t.itis_id, t.worms_id, t.scientific_name, d.taxon_key, t.notes
  FROM dataset_taxon d JOIN taxon t USING (taxon_key)
  WHERE d.dataset_key = ?
    AND json_extract(d.ds_source_json, '$.itis_id')::INTEGER IS DISTINCT FROM t.itis_id
  ORDER BY d.ds_taxa_code", params = list(ds_key)) |>
  mutate(
    override = tx_over_ds$note[match(ds_taxa_code, tx_over_ds$match_value)],
    explained_by = case_when(
      !is.na(override) ~ "override row (taxon_override.csv)",
      !is.na(source_itis_id) &
        str_detect(coalesce(notes, ""),
                   fixed(as.character(glue("itis:{source_itis_id} deprecated in ITIS")))) ~
        "deprecated TSN re-keyed onto its accepted form (taxon.notes)",
      is.na(source_itis_id) ~ "no source TSN; resolved by crosswalk or name (taxon.notes)",
      TRUE ~ "UNEXPLAINED"))
n_agree <- dbGetQuery(con, "
  SELECT COUNT(*) FROM dataset_taxon d JOIN taxon t USING (taxon_key)
  WHERE d.dataset_key = ? AND json_extract(d.ds_source_json, '$.itis_id')::INTEGER = t.itis_id",
  params = list(ds_key))[[1]]
stopifnot("every audit row must be explained" = !any(audit$explained_by == "UNEXPLAINED"))
cat(glue("audit: {n_agree} codes where the source TSN is the key's own TSN; ",
         "{nrow(audit)} differ -- ",
         "{paste(names(table(audit$explained_by)), table(audit$explained_by), sep = ': ', collapse = '; ')}"), "\n")

# the D3 rows, asserted rather than described
dt <- dbGetQuery(con, "SELECT ds_taxa_code, taxon_key FROM dataset_taxon WHERE dataset_key = ?",
                 params = list(ds_key))
key_of <- function(code) dt$taxon_key[dt$ds_taxa_code == code]
unid_birds   <- tx_over_ds$match_value[tx_over_ds$itis_id  %in% 174371L]
unid_mammals <- tx_over_ds$match_value[tx_over_ds$worms_id %in% 1837L]
stopifnot(
  "SBIG is staged exactly once"                      = sum(dt$ds_taxa_code == "SBIG") == 1,
  "MEGU keys the same taxon as SBIG (Q10)"           = identical(key_of("MEGU"), key_of("SBIG")),
  "MEGU / SBIG key an accepted ITIS TSN"             = str_starts(key_of("SBIG"), "itis:"),
  "every unidentified bird class keys Aves"          = all(dt$taxon_key[dt$ds_taxa_code %in% unid_birds]   == "itis:174371"),
  "every unidentified mammal class keys Mammalia"    = all(dt$taxon_key[dt$ds_taxa_code %in% unid_mammals] == "worms:1837"),
  "no excluded code has a dataset_taxon row"         = !any(excluded_codes %in% dt$ds_taxa_code))
cat(glue("D3: SBIG once; MEGU = SBIG = {key_of('SBIG')}; ",
         "{sum(dt$ds_taxa_code %in% unid_birds)} unidentified bird classes -> itis:174371, ",
         "{sum(dt$ds_taxa_code %in% unid_mammals)} unidentified mammal classes -> worms:1837"), "\n")

audit |>
  select(ds_taxa_code, ds_scientific_name, source_itis_id, itis_id, worms_id,
         scientific_name, taxon_key, explained_by, override) |>
  datatable(caption = "Source TSN (DataZoo) vs the authority TSN of the resolved key -- every row explained",
            rownames = FALSE, filter = "top")
```

### Sample, obs, obs_attribute

```{r}
#| label: emit_core

# This projection lives here, in the notebook that owns the dataset, not in a
# switch(dataset_key, ...) arm inside calcofi4db. The reusable SHAPE stays in the
# package (sample_arm_self), so this is a declaration.
append_sample(con, sample_arm_self(
  ds_key, "bird_mammal_transect", "gis_key", "transect"))

# obs — the occurrence headline: one row per (transect, SPECIES CODE), count
# SUMmed across behaviors.
#
# Grouping is on the source species_code, NOT on taxon_key alone: the excluded
# codes (excluded_codes, above) have no taxon, so grouping by taxon_key alone
# would sum every excluded species into a single NULL-taxon row per transect,
# silently merging distinct species. taxon_key is functionally determined by
# species_code (dataset_taxon is unique on ds_taxa_code within a dataset), so
# carrying both does not split the grain.
#
# The behavior breakdown is sub-occurrence detail and goes to obs_attribute — it
# must NOT ride on the headline's life_stage, or the same bird is counted once
# per behavior code.
append_obs(con, glue("
  SELECT 'bio', '{ds_key}', {ns_key(ds_key, 'transect', 'tr.gis_key')},
         tr.grid_key, tr.cruise_key, tr.latitude, tr.longitude,
         CAST(tr.datetime_start_utc AS TIMESTAMP), 0::DOUBLE, 0::DOUBLE,
         dt.taxon_key, NULL::VARCHAR, 'count', CAST(SUM(o.count) AS DOUBLE),
         NULL::VARCHAR, NULL::DOUBLE
  FROM bird_mammal_observation o JOIN bird_mammal_transect tr USING (gis_key)
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = CAST(o.species_code AS VARCHAR)
  GROUP BY tr.gis_key, tr.grid_key, tr.cruise_key, tr.latitude, tr.longitude,
           tr.datetime_start_utc, o.species_code, dt.taxon_key"))

# obs_attribute — the behavior breakdown, one row per source (transect, species,
# behavior): measurement_type = 'behavior', bin_label = the behavior description,
# bin_value NULL (behavior is categorical, not a numeric bin).
append_obs_attribute(con, glue("
  SELECT '{ds_key}', {ns_key(ds_key, 'transect', 'tr.gis_key')},
         dt.taxon_key, NULL::VARCHAR, 'behavior',
         NULL::DOUBLE, bb.description, CAST(o.count AS INTEGER), NULL::VARCHAR
  FROM bird_mammal_observation o JOIN bird_mammal_transect tr USING (gis_key)
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = CAST(o.species_code AS VARCHAR)
  LEFT JOIN bird_mammal_behavior bb ON bb.behavior_code = o.behavior_code"))

core <- list(
  sample        = dbGetQuery(con, "SELECT COUNT(*) FROM sample")[[1]],
  obs           = dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]],
  obs_attribute = dbGetQuery(con, "SELECT COUNT(*) FROM obs_attribute")[[1]],
  taxon         = n_taxon,
  dataset_taxon = n_ds_taxon,
  taxon_group   = n_tx_group)
cat(glue(
  "core projection — sample={core$sample %||% 0} obs={core$obs %||% 0} ",
  "obs_attribute={core$obs_attribute %||% 0} ",
  "taxon={core$taxon %||% 0} dataset_taxon={core$dataset_taxon %||% 0} ",
  "taxon_group={core$taxon_group %||% 0}\n"))

# the headline must collapse behaviors, and the attribution must add back up
n_obs <- core$obs
# one headline row per (transect, species_code) — NOT per behavior, and NOT
# collapsed by taxon_key (the excluded codes have none; grouping on taxon_key
# alone would merge every excluded species into one row per transect)
n_exp <- dbGetQuery(con, "
  SELECT COUNT(*) FROM (
    SELECT tr.gis_key, o.species_code
    FROM bird_mammal_observation o JOIN bird_mammal_transect tr USING (gis_key)
    GROUP BY 1, 2)")[[1]]
n_mismatch <- dbGetQuery(con, "
  WITH a AS (SELECT sample_key, taxon_key, SUM(count) s FROM obs_attribute
             WHERE measurement_type = 'behavior' GROUP BY 1, 2),
       o AS (SELECT sample_key, taxon_key, SUM(measurement_value) v FROM obs
             GROUP BY 1, 2)
  SELECT COUNT(*) FROM a JOIN o USING (sample_key, taxon_key) WHERE a.s <> o.v")[[1]]
# regression guard: grouping the headline on taxon_key alone (instead of
# species_code) merged every excluded species on a transect into ONE NULL-taxon
# row. If that ever comes back, transects with several excluded species collapse
# to a single row and this max drops to 1.
n_null_max <- dbGetQuery(con, "
  SELECT COALESCE(MAX(n), 0) FROM (
    SELECT sample_key, COUNT(*) n FROM obs WHERE taxon_key IS NULL GROUP BY 1)")[[1]]
n_null <- dbGetQuery(con, "SELECT COUNT(*) FROM obs WHERE taxon_key IS NULL")[[1]]
# the NULL-taxon rows must be exactly the excluded codes' rows: a staged code
# with no key would be a resolution failure check_dataset_taxon() should have
# caught, and an excluded code with a key would be a leak
excluded_sql <- paste0("'", excluded_codes, "'", collapse = ", ")
n_null_bad <- dbGetQuery(con, glue("
  SELECT COUNT(*) FROM bird_mammal_observation o
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}' AND dt.ds_taxa_code = o.species_code
  WHERE (dt.taxon_key IS NULL) <> (o.species_code IN ({excluded_sql}))"))[[1]]
stopifnot(
  "obs must be one row per (transect, species), not per behavior" = n_obs == n_exp,
  "obs_attribute behavior counts must sum to the obs headline"    = n_mismatch == 0,
  "a NULL taxon_key is an excluded code and nothing else"          = n_null_bad == 0,
  "excluded species must stay distinct, not merge into one NULL-taxon row per transect" =
    n_null_max > 1,
  "every obs.sample_key must resolve in sample" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key)
                     WHERE s.sample_key IS NULL")[[1]] == 0,
  "every obs.taxon_key must RESOLVE in taxon (not merely be non-NULL)" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs o LEFT JOIN taxon t USING (taxon_key)
                     WHERE o.taxon_key IS NOT NULL AND t.taxon_key IS NULL")[[1]] == 0,
  "behavior is the only obs_attribute type here" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs_attribute
                     WHERE measurement_type <> 'behavior'")[[1]] == 0)
cat(glue("obs parity: {format(n_obs, big.mark = ',')} headline rows ",
         "({format(n_null, big.mark = ',')} without taxon_key = the ",
         "{length(excluded_codes)} excluded codes, up to {n_null_max} of them ",
         "on one transect); behavior attribution reconciles"), "\n")
```

## Write Outputs + Upload

```{r}
#| label: outputs

dir_create(dir_parquet)
tbls_out <- core_output_tables(con, extra = "dataset")
write_parquet_outputs(
  con = con, output_dir = dir_parquet, tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)

d_tbls_rd <- read_csv(here("metadata/farallon/bird-mammal/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/farallon/bird-mammal/flds_redefine.csv"))
build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  # shared core descriptions first, this dataset's overrides second
  metadata_derived_csv = c(
    here("metadata/core_dictionary.csv"),
    here("metadata/farallon/bird-mammal/metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned,
  # when the ERDDAP tables were fetched (Read Source Data) -> metadata.json sources[]
  sources = src_stamp)

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else {
  cat("publish_to_gcs is FALSE -- parquet kept local, NOT synced to gs://calcofi-db\n")
}
```

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
