---
title: "Ingest SIO PIC Zooplankton Net Tows"
calcofi:
  target_name: ingest_sio_pic_zooplankton
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/sio_pic-zooplankton/manifest.json
  provider: sio
  dataset: pic-zooplankton
  workflow_url: https://calcofi.io/workflows/ingest_sio_pic-zooplankton.html
  questions_file: metadata/sio/pic-zooplankton/questions.csv
  dataset_meta:
    dataset_name: SIO PIC Zooplankton Net Tows
    # display trio, read by the release `dataset` table and the
    # consumer apps (calcofi4db >= 3.15.0) — see NEWS for why these
    # left the apps' own hardcoded maps
    dataset_name_short: Zooplankton Net Tows
    category: Zooplankton
    color: "#69db7c"
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (tow grain; effort/position only)"}
  erd:
    color: "#bbe0f0"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: `SIOPIC_DB_PNTtable_allRecords_9Feb2026.csv` (~19 MB, 148,129 net
tows), the parent-net-tow (PNT) registry of the SIO Pelagic Invertebrate
Collection (PI: Rasmus Swalethorp, Ed Weber, Linsey Sala).

- **Provider**: `pic`
- **Grain**: one row per zooplankton net tow (sample metadata only)
- **Scope decision**: ingest **CalCOFI-program tows only**, selected by
  `Expedition_Type_pnt == "CalCOFI"` (Q02). This replaced a CalCOFI-region
  *bounding box*, which was a proxy for the program and was wrong in both
  directions — see below.
- **Biovolume is NOT in this file** — only the tow/sample registry is built here
  as `zooplankton_tow`; the biovolume measurement table is pending from the
  provider (blocker Q01) and will be added as `zooplankton_measurement`.

```{mermaid}
graph LR
  A[PNTtable.csv<br/>148,129 tows] --> B[filter Expedition_Type = CalCOFI]
  B --> C[zooplankton_tow<br/>position + gear + keys]
  C -.ship/cruise/grid.-> D[(shared refs)]
  C -.biovolume pending.-> E[zooplankton_measurement]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_sio_pic-zooplankton.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")

dir_parquet  <- here(glue("data/parquet/{dir_label}"))
dir_stage  <- cc_stage_path("parquet", dir_label, create = TRUE)
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))  dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

Read directly from `{dir_data}/calcofi/zooplankton/` (the file is not under a
`pic/` subdirectory). Columns renamed per
`metadata/sio/pic-zooplankton/flds_redefine.csv`.

```{r}
#| label: read-source

zoo_csv <- path_expand(glue(
  "{dir_data}/calcofi/zooplankton/SIOPIC_DB_PNTtable_allRecords_9Feb2026.csv"))
stopifnot("zooplankton CSV not found" = file_exists(zoo_csv))

sync_to_gcs(
  local_dir  = path_dir(zoo_csv),
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket     = "calcofi-files-public",
  exclude    = c(".DS_Store", "*.tmp", "*.gdoc"))

d_raw <- read_csv(zoo_csv, col_types = cols(.default = "c"))  # read all as text, cast below
cat(glue("Read {format(nrow(d_raw), big.mark=',')} tows, {ncol(d_raw)} columns"), "\n")
```

## Clean, Type-Cast, and Filter to CalCOFI Cruises

Canonical names from `metadata/field_dictionary.csv`. Datetimes are built from
date (MM/DD/YYYY) + HHMM time (treated as given pending the timezone answer,
Q03).

The scope filter is the **program**, `Expedition_Type_pnt == "CalCOFI"`, not a
CalCOFI-region bounding box. The flag is trustworthy: it agrees with `"CALCOFI"`
appearing in `EXPEDITION_pnt` on all 148,129 source rows, with zero disagreement
either way.

The bounding box it replaces was a proxy for the program, and it was wrong in
both directions:

| | tows | resolve to a CalCOFI cruise |
|---|---:|---:|
| in bbox, **not** CalCOFI — now dropped | 32,719 | 35.5% |
| CalCOFI, **outside** bbox — now restored | 15,532 | 83.2% |
| CalCOFI and in bbox — kept either way | 66,811 | 81.8% |

The 15,532 restored tows are not coordinate errors — the CalCOFI subset contains
**zero** missing, impossible or `NaN` coordinates, so Q04's antimeridian problem
lives entirely in the non-CalCOFI holdings. They are the historical grid: 1950s–80s
inshore stations, the Gulf of California, and the Baja lines down to 19.9°N. That
they resolve to a CalCOFI cruise *more* often than the average tow (83.2%) is the
evidence they belong.

The dropped tows (CWPA Squid Survey, RACER, SOSO, Huntington Beach, CCE-LTER
process cruises …) matter more than their count suggests: 11,613 of them carried a
`cruise_key`, but that key is derived from ship + year-month, so a non-CalCOFI
cruise sharing a vessel and month with a real CalCOFI cruise was being stamped with
it. Dropping them removes false attributions, not just out-of-scope rows.

```{r}
#| label: clean-filter

hhmm_to_dt <- function(date, hhmm) {
  t <- str_pad(na_if(str_trim(hhmm), ""), 4, pad = "0")
  out <- suppressWarnings(as_datetime(paste0(
    date, " ", substr(t, 1, 2), ":", substr(t, 3, 4), ":00")))
  out[is.na(t) & !is.na(date)] <- as_datetime(as.character(date[is.na(t) & !is.na(date)]))
  out
}

d_clean <- d_raw |>
  transmute(
    expedition      = EXPEDITION_pnt,
    expedition_code = EXPED_CODE_pnt,
    expedition_type = Expedition_Type_pnt,
    ship_name       = SHIP_pnt,
    order_occ       = suppressWarnings(as.integer(SWFOrder_Occ)),
    line            = suppressWarnings(as.numeric(STATION_LINE_pnt)),
    station         = suppressWarnings(as.numeric(STATION_NUMBER_pnt)),
    latitude        = suppressWarnings(as.numeric(LAT_DECIMAL_pnt)),
    longitude       = suppressWarnings(as.numeric(LONG_DECIMAL_pnt)),
    date            = suppressWarnings(as.Date(SAMPLE_DATE_pnt, format = "%m/%d/%Y")),
    datetime_start_utc = hhmm_to_dt(date, START_TIME_pnt),  # Q03: assumed UTC
    datetime_end_utc   = hhmm_to_dt(date, END_TIME_pnt),
    depth_min_m     = suppressWarnings(as.numeric(DEPTH_MIN_pnt)),
    depth_max_m     = suppressWarnings(as.numeric(DEPTH_MAX_pnt)),
    max_wire_out_m  = suppressWarnings(as.numeric(MAX_MWO_pnt)),
    net_type        = NET_TYPE_pnt,
    mesh_size_mm    = suppressWarnings(as.numeric(MESH_SIZE_pnt)),
    tow_type        = TOW_TYPE_pnt,
    fixative        = FIXATIVE_pnt,
    preservative    = PRESERVATIVE_pnt)

n_all <- nrow(d_clean)
# scope filter: the CalCOFI program (Q02), not a region. the coordinate guard is
# a validity check only -- it rejects impossible/NaN positions (Q04) rather than
# clipping the historical grid to the modern one.
d_cc <- d_clean |>
  filter(expedition_type == "CalCOFI") |>
  filter(!is.na(latitude), !is.na(longitude),
         !is.nan(latitude), !is.nan(longitude),
         abs(latitude) <= 90, abs(longitude) <= 180) |>
  mutate(
    site_key = if_else(
      is.na(line) | is.na(station), NA_character_,
      sprintf("%05.1f %05.1f", line, station)),
    tow_id = row_number())

n_cc     <- sum(d_clean$expedition_type == "CalCOFI", na.rm = TRUE)
n_badpos <- n_cc - nrow(d_cc)
cat(glue(
  "CalCOFI-program filter: {format(n_cc, big.mark=',')}/",
  "{format(n_all, big.mark=',')} tows are CalCOFI ",
  "({round(100*n_cc/n_all, 1)}%); ",
  "{n_badpos} dropped for an invalid position; ",
  "{format(nrow(d_cc), big.mark=',')} retained"), "\n")

# the program flag must agree with the expedition name, or the scope is not what
# it claims -- this is the assumption the whole filter rests on
disagree <- d_clean |>
  filter(xor(coalesce(expedition_type, "") == "CalCOFI",
             str_detect(coalesce(expedition, ""), regex("calcofi", ignore_case = TRUE))))
stopifnot(
  "Expedition_Type must agree with CALCOFI in the expedition name" = nrow(disagree) == 0,
  "CalCOFI subset must have no invalid positions (Q04 is non-CalCOFI only)" = n_badpos == 0)

dbWriteTable(con, "zooplankton_tow", d_cc, overwrite = TRUE)
```

## Resolve Ship and Cruise Keys

```{r}
#| label: resolve-keys

load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  tables = c("ship", "cruise", "grid"), geom_tables = c("grid"), as_view = TRUE)

d_ship <- dbGetQuery(con, "SELECT ship_key, ship_name, ship_nodc FROM ship") |>
  mutate(ship_name_norm = ship_name |> str_to_upper() |> str_squish())

zt <- dbGetQuery(con, "SELECT tow_id, ship_name, date FROM zooplankton_tow") |>
  mutate(ship_name_norm = ship_name |>
           str_replace("^R/?V\\.?\\s+", "") |> str_to_upper() |> str_squish()) |>
  left_join(d_ship |> select(ship_name_norm, ship_key, ship_nodc), by = "ship_name_norm")

# cruise_key: the reference cruise whose date span holds the tow, else its month,
# keys not in `cruise` left NULL (calcofi4db::resolve_cruise_key). A tow in the
# first days of a month belongs to the cruise that began the month before — the
# old month rule put it on the NEXT cruise of the same ship (30 of this dataset's
# consecutive-month cruise pairs ran into each other within 3 days).
dbWriteTable(con, "zt_keys", zt |> select(tow_id, ship_key, datetime = date), overwrite = TRUE)
resolve_cruise_key(con, "zt_keys", datetime_col = "datetime", require_in_cruise = TRUE) |>
  datatable(caption = "cruise_key resolution method")
zt <- zt |> left_join(dbGetQuery(con, "SELECT tow_id, cruise_key FROM zt_keys"), by = "tow_id")
dbExecute(con, "ALTER TABLE zooplankton_tow ADD COLUMN IF NOT EXISTS ship_key VARCHAR")
dbExecute(con, "ALTER TABLE zooplankton_tow ADD COLUMN IF NOT EXISTS cruise_key VARCHAR")
dbExecute(con,
  "UPDATE zooplankton_tow t SET ship_key = k.ship_key, cruise_key = k.cruise_key
   FROM zt_keys k WHERE t.tow_id = k.tow_id")
dbExecute(con, "DROP TABLE zt_keys")

cat(glue(
  "ship match: {sum(!is.na(zt$ship_key))}/{nrow(zt)} ",
  "({round(100*mean(!is.na(zt$ship_key)),1)}%); ",
  "cruise_key match: {sum(!is.na(zt$cruise_key))}/{nrow(zt)} ",
  "({round(100*mean(!is.na(zt$cruise_key)),1)}%)"), "\n")
```

## Add Spatial

```{r}
#| label: spatial

add_point_geom(con, "zooplankton_tow", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "zooplankton_tow") |> datatable(caption = "Grid assignment")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Schema Documentation

```{r}
#| label: schema

zoo_rels <- list(
  primary_keys = list(zooplankton_tow = "tow_id"),
  foreign_keys = list())  # cross-dataset FKs (cruise/ship/grid) in relationships_cross.csv

cc_erd(
  con, tables = c("zooplankton_tow", "dataset"), rels = zoo_rels,
  colors = list(lightblue = "zooplankton_tow", white = "dataset"))

# the SOURCE shape above documents the wrangling; relationships.json is written
# with the parquet outputs below, from core_relationships().
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
# NULL cruise_key/ship_key/site_key are the expected unmatched remainder (Q03/Q05),
# the same accepted behavior as calcofi_dic; not a hard failure.

n_dup <- dbGetQuery(con,
  "SELECT COUNT(*) FROM (SELECT tow_id, COUNT(*) n FROM zooplankton_tow GROUP BY tow_id HAVING COUNT(*)>1)")[[1]]
cat(glue("zooplankton_tow tow_id duplicates: {n_dup}"), "\n")
```

## Data Preview

```{r}
#| label: preview

cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='zooplankton_tow' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM zooplankton_tow LIMIT 100")) |>
  datatable(caption = "zooplankton_tow — first 100 rows", rownames = FALSE, filter = "top")
```

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`). These core tables **are** this ingest's
output: `release_database.qmd` concatenates the per-dataset shards rather than
re-deriving the core from per-dataset tables, so there is exactly one projection
to keep correct.

This source carries tow effort and position only — **no measurements** — so it
contributes `sample` and nothing else.

```{r}
#| label: emit_core

ds_key <- "sio_pic-zooplankton"

# This projection lives here, in the notebook that owns the dataset, not in a
# switch(dataset_key, ...) arm inside calcofi4db. The reusable SHAPE stays in the
# package (sample_arm_self), so this is a declaration.
append_sample(con, sample_arm_self(
  ds_key, "zooplankton_tow", "tow_id", "tow",
  site_expr = "site_key", ord_expr = "CAST(order_occ AS INTEGER)",
  depth_min = "depth_min_m", depth_max = "depth_max_m"))

core <- list(sample = dbGetQuery(con, "SELECT COUNT(*) FROM sample")[[1]])
cat(glue("core projection — sample={core$sample %||% 0}"), "\n")

# one sample row per source tow, and the cross-dataset columns must survive the
# projection (site_key/order_occ are how `site` is rebuilt from the core).
# Assert the no-measurement case explicitly, rather than let a silent zero-row
# obs look like a projection bug.
n_tow <- dbGetQuery(con, "SELECT COUNT(*) FROM zooplankton_tow")[[1]]
n_site <- dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE site_key IS NOT NULL")[[1]]
n_ord  <- dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE order_occ IS NOT NULL")[[1]]
stopifnot(
  "sample must be one row per source tow"          = core$sample == n_tow,
  "sio_pic-zooplankton contributes sample rows"        = core$sample > 0,
  "sio_pic-zooplankton has no measurements to project" = !"obs" %in% dbListTables(con),
  "sample must carry site_key for most tows"       = n_site > 0,
  "sample must carry order_occ where the source has it" = n_ord > 0)
cat(glue("sample parity: {format(core$sample, big.mark = ',')} tows; ",
         "{format(n_site, big.mark = ',')} with site_key, ",
         "{format(n_ord, big.mark = ',')} with order_occ"), "\n")
```


## Write Parquet Outputs

```{r}
#| label: write-parquet

dir_create(dir_parquet)
mismatches <- list(cruise_keys = collect_cruise_key_mismatches(con, "zooplankton_tow"))
tbls_out <- core_output_tables(con, extra = "dataset")
parquet_stats <- write_parquet_outputs(
  con = con, output_dir = dir_parquet,
  tables = tbls_out,
  strip_provenance = FALSE, mismatches = mismatches)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)
parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata

d_tbls_rd <- read_csv(here("metadata/sio/pic-zooplankton/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/sio/pic-zooplankton/flds_redefine.csv"))
metadata_path <- build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           here("metadata/sio/pic-zooplankton/metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned)
```

## Upload to GCS

```{r}
#| label: upload-gcs

sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
```

## Questions for Data Providers

Follow-up questions for SIO PIC (Rasmus Swalethorp, Ed Weber, Linsey Sala),
**ranked by importance**. The headline blocker (Q01) is that biovolume is absent
from the provided file, so only the tow registry is ingested here. Tracked in
`metadata/sio/pic-zooplankton/questions.csv`; aggregated by `questions_email.qmd`.

```{r}
#| label: provider-questions

# one validated read + render for every ingest: the vocabulary and the column
# order live in calcofi4db, not in 16 hand-written factor() calls
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the SIO PIC zooplankton data providers (ranked)")
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
