---
title: "Ingest NOAA SWFSC Additional CTD Casts"
calcofi:
  target_name: ingest_swfsc_ctd_noaa
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
    - ingest_calcofi_bottle
    - ingest_calcofi_ctd_cast
  output: data/parquet/swfsc_ctd-noaa/manifest.json
  provider: swfsc
  dataset: ctd-noaa
  workflow_url: https://calcofi.io/workflows/ingest_swfsc_ctd-noaa.html
  questions_file: metadata/swfsc/ctd-noaa/questions.csv
  # staged, not released: first ingest of this holding (CalCOFI/workflows#84).
  in_release: false
  dataset_meta:
    dataset_name: CalCOFI Additional CTD Hydrographic Cast Data-NOAA SWFSC
    dataset_name_short: "NOAA additional CTD casts"
    category: "Physical Oceanography"
    color: "#2f6f8f"
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (one CTD cast per row; own sample grain)"}
    - {table: obs, shared: true, note: "core observation headline (env realm; uncalibrated sensor temperature, salinity, oxygen at standard depths)"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#c8dfea"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: NOAA CoastWatch ERDDAP `erdCalCOFINOAAhydros` ("CalCOFI NOAA
Additional CTD", NOAA SWFSC, Ed Weber), the whole table: 15,151 rows, 14
columns, 1,235 casts on 28 cruise x ship combinations, 2002-02 to 2014-08
(checked 2026-10-06). Values are CTD readings at standard depths (0 to 1,000 m).
ERDDAP's own summary: "oxygen and salinity mostly uncalibrated".

- **Provider**: `swfsc` (NOAA SWFSC curates it; ERDDAP is only the portal).
- **Grain**: `sample` = one CTD cast, keyed by the source's own
  `(cruise, ship_code, order_occupied)`; `obs` = one value per cast x depth x
  measurement.
- **Measurements**: the uncorrected-sensor types the registry already holds for
  CTD casts: `temperature_1` (degC), `salinity_1` (PSU), `oxygen_ml_l_1` (ml/L).
  Not calibrated against bottle samples, which is exactly what those types
  describe.
- **Not published**: `percent_saturation` (present on 5 cruises only, and
  derivable from oxygen, temperature and salinity, Q05).
- **Overlap with CalCOFI's own casts (held out)**: on some cruises (e.g. 2004-01
  David Starr Jordan, 2005-01 New Horizon) these casts are the same casts as
  `calcofi_ctd-cast` / `calcofi_bottle`, at the same station within minutes. The
  SIO versions are calibrated, so a NOAA cast that matches a CalCOFI cast (same
  `site_key`, within 6 h) is **held out** and only the casts CalCOFI does not
  already have are published (Q01).
- **Cruise key**: `ship_code` is the ship reference's own `ship_key` (NH, JD, SH,
  ..., asserted), and `cruise` is the YYYYMM designation, so
  `resolve_cruise_key()` keys every cast (span, then designation).

```{mermaid}
graph LR
  A[ERDDAP erdCalCOFINOAAhydros<br/>15,151 rows] --> B[ctd_cast<br/>one row per cast]
  A --> C[ctd_measurement<br/>long: temperature_1, salinity_1, oxygen_ml_l_1]
  B -.site_key + time.-> D[(calcofi_ctd-cast / calcofi_bottle casts<br/>match = held out)]
  B -.ship_key + cruise.-> E[(cruise_key)]
  B --> F[[core: sample, cast]]
  C --> G[[core: obs, realm=env]]
```

## Setup

```{r}
#| label: setup
devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))
cc             <- read_calcofi_meta(here("ingest_swfsc_ctd-noaa.qmd"))
provider       <- cc$provider
dataset        <- cc$dataset
tables_owned   <- cc$tables_owned
ds_key         <- as.character(glue("{provider}_{dataset}"))
dir_label      <- ds_key
dir_parquet    <- here(glue("data/parquet/{dir_label}"))
dir_stage      <- cc_stage_path("parquet", dir_label, create = TRUE)
dir_src        <- cc_stage_path("source", dir_label, create = TRUE)
db_path        <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_meta       <- here(glue("metadata/{provider}/{dataset}"))
publish_to_gcs <- Sys.getenv("CALCOFI_SKIP_GCS") == ""

# a prior ingest's shard: the local stage copy, else its public GCS copy
shard_path <- function(ds, tbl) {
  p <- cc_stage_path("parquet", ds, glue("{tbl}.parquet"))
  if (file_exists(p)) return(p)
  glue("https://storage.googleapis.com/calcofi-db/ingest/{ds}/{tbl}.parquet")
}

meas_map <- tribble(
  ~measurement_type, ~value_col,
  "temperature_1",   "temperature",
  "salinity_1",      "salinity",
  "oxygen_ml_l_1",   "oxygen")
d_meas_type <- read_measurement_type(here("metadata/measurement_type.csv"))
stopifnot("the three sensor types are registered" =
            all(meas_map$measurement_type %in% d_meas_type$measurement_type))

if (overwrite) {
  if (file_exists(db_path)) file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp"))) dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
load_duckdb_extension(con, "httpfs")
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Read Source Data

```{r}
#| label: read-source
source(here("libs/download_ctd_noaa.R"))
downloaded <- overwrite_all || !file_exists(file.path(dir_src, "data.csv"))
src_csv <- download_ctd_noaa(dir_src, overwrite = overwrite_all)
src_stamp <- stamp_source_access(files = src_csv) |>
  mutate(source = CTD_NOAA_URL, method = if (downloaded) "download" else "file_mtime")

if (publish_to_gcs) {
  sync_to_gcs(
    local_dir  = dir_src,
    gcs_prefix = glue("archive/{provider}/{dataset}"),
    bucket     = "calcofi-files-public",
    exclude    = c(".DS_Store", "*.tmp", "*.part"))
} else {
  cat("CALCOFI_SKIP_GCS set -- source NOT archived to gs://calcofi-files-public\n")
}

hdr_expected <- c(
  "cruise", "ship", "ship_code", "order_occupied", "latitude", "longitude",
  "line", "station", "time", "ctd_depth", "temperature", "salinity", "oxygen",
  "percent_saturation")
units_expected <- c(
  "", "", "", "", "degrees_north", "degrees_east", "", "", "UTC", "meters",
  "degrees Celsius", "practical salinity units", "ml/l", "percent")
hdr <- readLines(src_csv, n = 2)
stopifnot(
  "ERDDAP header changed: re-check flds_redefine.csv" =
    identical(strsplit(hdr[1], ",")[[1]], hdr_expected),
  "ERDDAP units line changed: re-check the measurement units" =
    identical(c(strsplit(hdr[2], ",")[[1]], rep("", 14))[1:14], units_expected))

d_raw <- read_csv(src_csv, skip = 2, col_names = hdr_expected,
                  col_types = cols(.default = "c"), na = character())
cat(glue("Read {format(nrow(d_raw), big.mark=',')} rows, {ncol(d_raw)} columns"), "\n")
stopifnot("expected 15,151 rows (ERDDAP, checked 2026-10-06)" = nrow(d_raw) == 15151)
```

## Clean, Type-Cast

ERDDAP writes a missing value as `NaN`; it becomes `NA` here (R's
`as.numeric("NaN")` is `NaN`, which is not `NA` and would survive an
`IS NOT NULL`).

```{r}
#| label: clean-cast
num <- function(x) {
  v <- suppressWarnings(as.numeric(if_else(x %in% c("", "NaN"), NA_character_, x)))
  v[is.nan(v)] <- NA_real_
  v
}

d_rows <- d_raw |>
  transmute(
    cruise_ym          = cruise,
    ship_name          = ship,
    ship_code,
    order_occupied     = as.integer(order_occupied),
    latitude           = num(latitude),
    longitude          = num(longitude),
    line               = num(line),
    station            = num(station),
    datetime_start_utc = ymd_hms(time, tz = "UTC"),
    depth_m            = num(ctd_depth),
    temperature        = num(temperature),
    salinity           = num(salinity),
    oxygen             = num(oxygen),
    percent_saturation = num(percent_saturation)) |>
  mutate(sample_key_id = paste(cruise_ym, ship_code, order_occupied, sep = "-"))

stopifnot(
  "every row parses a cast id, time, position, line/station and depth" =
    !anyNA(select(d_rows, order_occupied, datetime_start_utc, latitude, longitude,
                  line, station, depth_m)),
  "cruise is a YYYYMM designation" = all(str_detect(d_rows$cruise_ym, "^\\d{4}(0[1-9]|1[0-2])$")),
  "one row per cast x depth" = !anyDuplicated(d_rows[c("sample_key_id", "depth_m")]),
  "temperature and salinity on every row" = !anyNA(d_rows$temperature) && !anyNA(d_rows$salinity))

d_cast <- d_rows |>
  group_by(sample_key_id, cruise_ym, ship_name, ship_code, order_occupied,
           datetime_start_utc, latitude, longitude, line, station) |>
  summarize(depth_min_m = min(depth_m), depth_max_m = max(depth_m),
            n_depths = n(), .groups = "drop") |>
  # a negative line or station (the extended numbering off Washington and British
  # Columbia, e.g. line -10.7) is not a CalCOFI station: no site_key
  mutate(site_key = if_else(line < 0 | station < 0, NA_character_,
                            normalize_site_key(sprintf("%.1f %.1f", line, station)))) |>
  arrange(datetime_start_utc, sample_key_id) |>
  mutate(cast_id = row_number())
stopifnot(
  "a cast has one time, position and station (1,235 casts)" =
    !anyDuplicated(d_cast$sample_key_id) && nrow(d_cast) == 1235)

d_meas <- d_rows |>
  inner_join(d_cast |> select(sample_key_id, cast_id), by = "sample_key_id") |>
  pivot_longer(all_of(meas_map$value_col), names_to = "value_col",
               values_to = "measurement_value", values_drop_na = TRUE) |>
  inner_join(meas_map, by = "value_col") |>
  arrange(cast_id, depth_m, measurement_type) |>
  transmute(measurement_id = row_number(), cast_id, depth_m,
            measurement_type, measurement_value)

d_meas |> count(measurement_type) |>
  datatable(caption = "Values per measurement_type (all casts)", rownames = FALSE)
cat(glue("percent_saturation held out (not published): ",
         "{sum(!is.na(d_rows$percent_saturation))} values on ",
         "{n_distinct(d_rows$cruise_ym[!is.na(d_rows$percent_saturation)])} cruises (Q05)"), "\n")
stopifnot("15,151 temperature + 15,151 salinity + 10,676 oxygen values" =
            nrow(d_meas) == 15151 + 15151 + 10676)

dbWriteTable(con, "ctd_cast", d_cast, overwrite = TRUE)
dbWriteTable(con, "ctd_measurement", d_meas, overwrite = TRUE)
```

## Resolve Ship and Cruise Keys

The source's `ship_code` (NH, JD, SH, ...) is the ship reference's own
`ship_key`, and the ship names agree (asserted), so no ship matching is needed.
`resolve_cruise_key()` then keys each cast by the cruise span, falling back to
the source's `cruise` designation.

```{r}
#| label: resolve-keys
load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  gcs_prefix = "ingest/swfsc_ichthyo",
  tables = c("ship", "cruise"), as_view = TRUE)

d_ship_chk <- dbGetQuery(con, "
  SELECT DISTINCT c.ship_code, c.ship_name, s.ship_key, s.ship_name AS ref_name, s.ship_nodc
  FROM ctd_cast c LEFT JOIN ship s ON s.ship_key = c.ship_code ORDER BY 1")
d_ship_chk |> datatable(caption = "Source ship_code = ship reference ship_key", rownames = FALSE)
stopifnot(
  "every ship_code is a ship_key in the ship reference" = !anyNA(d_ship_chk$ship_key),
  "and the ship names agree" = all(toupper(d_ship_chk$ship_name) == toupper(d_ship_chk$ref_name)))

dbExecute(con, "ALTER TABLE ctd_cast ADD COLUMN IF NOT EXISTS ship_key VARCHAR")
dbExecute(con, "UPDATE ctd_cast SET ship_key = ship_code")

key_stats <- resolve_cruise_key(
  con, "ctd_cast", datetime_col = "datetime_start_utc",
  ship_key_col = "ship_key", cruise_ym_col = "cruise_ym")
key_stats |> datatable(caption = "cruise_key resolution: span > source designation > month")

n_ck <- dbGetQuery(con, "SELECT COUNT(*) FROM ctd_cast WHERE cruise_key IS NOT NULL")[[1]]
cat(glue("cruise_key: {n_ck}/{nrow(d_cast)} casts"), "\n")
stopifnot("every cast resolves a cruise_key" = n_ck == nrow(d_cast))
```

## Casts CalCOFI Already Has (held out)

A NOAA cast that is the same cast as a `calcofi_ctd-cast` or `calcofi_bottle`
cast is a duplicate of a calibrated SIO record. `match_by_site_datetime()` finds,
for each NOAA cast, the CalCOFI cast at the same `site_key` on the nearest date
(within 1 day), **once against the bottle casts and once against the CTD casts**:
the two CalCOFI datasets can time the same cast hours apart (first render,
2026-10-06: on 2005-01 New Horizon the CTD casts read about 8 h earlier than the
bottle casts), and a single match picks only one candidate per date. A cast is
held out when either match is within **6 hours**. The threshold and the hold-out
are Q01.

```{r}
#| label: overlap
ref_tbls <- c(ref_bottle = "calcofi_bottle", ref_ctd_cast = "calcofi_ctd-cast")
for (tbl in names(ref_tbls)) {
  ds <- ref_tbls[[tbl]]
  dbExecute(con, glue("
    CREATE OR REPLACE TABLE {tbl} AS
    SELECT sample_key AS ref_cast_key, site_key, datetime
    FROM read_parquet('{shard_path(ds, 'sample')}')
    WHERE sample_type = 'cast' AND site_key IS NOT NULL AND datetime IS NOT NULL"))
}

match_hours <- function(ref_tbl, fk, hrs) {
  m <- match_by_site_datetime(
    con, "ctd_cast", ref_tbl, fk_col = fk, ref_pk = "ref_cast_key",
    key_col = "site_key", datetime_col = "datetime_start_utc",
    ref_datetime_col = "datetime", window_days = 1)
  dbExecute(con, glue("ALTER TABLE ctd_cast ADD COLUMN IF NOT EXISTS {hrs} DOUBLE"))
  dbExecute(con, glue("
    UPDATE ctd_cast SET {hrs} = ABS(EPOCH(ctd_cast.datetime_start_utc) - EPOCH(r.datetime)) / 3600
    FROM {ref_tbl} r WHERE r.ref_cast_key = ctd_cast.{fk}"))
  m
}
m_btl <- match_hours("ref_bottle", "bottle_cast_key", "bottle_hours")
m_ctd <- match_hours("ref_ctd_cast", "ctd_cast_key", "ctd_hours")

dbExecute(con, "ALTER TABLE ctd_cast ADD COLUMN IF NOT EXISTS is_calcofi_dup BOOLEAN")
dbExecute(con, "
  UPDATE ctd_cast SET is_calcofi_dup =
    COALESCE(bottle_hours <= 6, FALSE) OR COALESCE(ctd_hours <= 6, FALSE)")

d_overlap <- dbGetQuery(con, "
  SELECT cruise_ym, ship_code, COUNT(*) AS n_casts,
         SUM(is_calcofi_dup::INT)                 AS n_held_dup,
         COUNT(*) FILTER (WHERE bottle_hours <= 6) AS n_bottle_6h,
         COUNT(*) FILTER (WHERE ctd_hours <= 6)    AS n_ctd_6h,
         COUNT(*) FILTER (WHERE NOT is_calcofi_dup
                           AND (bottle_cast_key IS NOT NULL OR ctd_cast_key IS NOT NULL)) AS n_near_not_dup
  FROM ctd_cast GROUP BY 1, 2 ORDER BY 1, 2")
d_overlap |> datatable(caption = "Casts per cruise that CalCOFI already has (held out, Q01)", rownames = FALSE)

# the CTD-vs-bottle timing offset that motivated two matches, reported
dbGetQuery(con, "
  SELECT cruise_ym, ship_code, COUNT(*) AS n_both,
         ROUND(MEDIAN(ctd_hours - bottle_hours), 1) AS median_ctd_minus_bottle_h
  FROM ctd_cast WHERE bottle_hours IS NOT NULL AND ctd_hours IS NOT NULL
  GROUP BY 1, 2 ORDER BY 1, 2") |>
  datatable(caption = "NOAA casts matched to both CalCOFI datasets: hours to each", rownames = FALSE)

n_dup <- sum(d_overlap$n_held_dup)
n_pub <- nrow(d_cast) - n_dup
cat(glue("{n_dup} of {nrow(d_cast)} casts duplicate a CalCOFI cast and are held out; ",
         "{n_pub} casts are published"), "\n")
stopifnot("some casts remain to publish" = n_pub > 0)

dbExecute(con, "CREATE OR REPLACE TABLE ctd_cast_pub AS SELECT * FROM ctd_cast WHERE NOT is_calcofi_dup")
```

## Add Spatial

```{r}
#| label: spatial
load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  gcs_prefix = "ingest/swfsc_ichthyo",
  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
add_point_geom(con, "ctd_cast_pub", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "ctd_cast_pub") |> datatable(caption = "Grid assignment (published casts)")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata
this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot(nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)
```

## Schema Documentation

```{r}
#| label: schema
ctd_rels <- list(
  primary_keys = list(
    ctd_cast         = "cast_id",
    ctd_measurement  = "measurement_id",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "ctd_measurement", column = "cast_id",
         ref_table = "ctd_cast", ref_column = "cast_id"),
    list(table = "ctd_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))
cc_erd(
  con, tables = c("ctd_cast", "ctd_measurement", "measurement_type", "dataset"),
  rels = ctd_rels,
  colors = list(lightblue = c("ctd_cast", "ctd_measurement"),
                lightyellow = "measurement_type", white = "dataset"))
```

| source | core | transform |
|---|---|---|
| one row per `(cruise, ship_code, order_occupied)` | `sample` (`cast`) | key `swfsc_ctd-noaa:cast:{cruise}-{ship_code}-{order_occupied}`; `depth_min_m`/`depth_max_m` = the cast's shallowest/deepest standard depth; time = `time` (UTC) |
| `temperature` | `obs` `temperature_1` | verbatim, degC, at `ctd_depth` |
| `salinity` | `obs` `salinity_1` | verbatim, PSU, at `ctd_depth` |
| `oxygen` | `obs` `oxygen_ml_l_1` | verbatim, ml/L, at `ctd_depth`; `NaN` (4,475 rows) not published |
| `percent_saturation` | -- | **not published** (5 cruises only, derivable, Q05) |
| casts matching a CalCOFI cast | -- | **held out** (Q01) |

## Check Declared Bounds

All three types already carry bounds in the registry (temperature_1 -2 to 40
degC, salinity_1 0 to 45 PSU, oxygen_ml_l_1 0 to 15 ml/L). The suspicious values
the provider is asked about in Q02 (surface salinity of 22 and 24 PSU; oxygen of
12-13 ml/L at 6 degC on 2012-04) are inside those bounds, so nothing is dropped.

```{r}
#| label: bounds-check
bounds_pre <- check_measurement_bounds(con, "ctd_measurement", mt = d_meas_type)
bounds_datatable(bounds_pre)
oob_tally <- drop_out_of_bounds(con, "ctd_measurement", mt = d_meas_type)
cat(glue("out-of-range rows removed: {sum(oob_tally$n_bad %||% 0)}"), "\n")
bounds_post <- check_measurement_bounds(
  con, "ctd_measurement", mt = d_meas_type, include_undeclared = FALSE)
stopifnot(
  "every emitted type has declared bounds" = all(bounds_pre$status != "undeclared"),
  "no value outside a declared bound" = sum(bounds_post$status == "out_of_range") == 0)
```

## Enforce Column Types

```{r}
#| label: enforce-types
enforce_column_types(con)
```

## Data Preview

```{r}
#| label: preview-cast
cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='ctd_cast_pub' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM ctd_cast_pub LIMIT 200")) |>
  datatable(caption = "ctd_cast_pub -- first 200 published casts", rownames = FALSE, filter = "top")
```

## Emit Core Tables

```{r}
#| label: emit_core
append_sample(con, sample_arm_self(
  ds_key, "ctd_cast_pub", "sample_key_id", "cast",
  dt_col = "datetime_start_utc", site_expr = "site_key",
  ord_expr = "order_occupied",
  depth_min = "depth_min_m", depth_max = "depth_max_m"))

append_obs(con, glue("
  SELECT 'env', '{ds_key}', {ns_key(ds_key, 'cast', 'c.sample_key_id')},
         c.grid_key, c.cruise_key, c.latitude, c.longitude,
         CAST(c.datetime_start_utc AS TIMESTAMP), m.depth_m, m.depth_m,
         NULL::VARCHAR, NULL::VARCHAR, m.measurement_type, m.measurement_value,
         NULL::VARCHAR, NULL::DOUBLE
  FROM ctd_measurement m
  JOIN ctd_cast_pub c USING (cast_id)"))

q1 <- function(sql) dbGetQuery(con, sql)[[1]]
n_exp_obs <- q1("SELECT COUNT(*) FROM ctd_measurement m JOIN ctd_cast_pub c USING (cast_id)")
core <- list(sample = q1("SELECT COUNT(*) FROM sample"), obs = q1("SELECT COUNT(*) FROM obs"))
cat(glue("core projection -- sample={core$sample} obs={core$obs}"), "\n")
stopifnot(
  "one sample per published cast" = core$sample == n_pub,
  "obs is one row per published value" = core$obs == n_exp_obs,
  "only the three sensor types are emitted" =
    setequal(dbGetQuery(con, glue("SELECT DISTINCT measurement_type FROM obs
                                   WHERE dataset_key = '{ds_key}'"))$measurement_type,
             meas_map$measurement_type),
  "sample_key is unique" =
    q1(glue("SELECT COUNT(*) - COUNT(DISTINCT sample_key) FROM sample WHERE dataset_key = '{ds_key}'")) == 0,
  "every obs.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key) WHERE s.sample_key IS NULL") == 0,
  "obs depth lies within its cast's depth range" =
    q1("SELECT COUNT(*) FROM obs o JOIN sample s USING (sample_key)
        WHERE o.depth_max_m < s.depth_min_m OR o.depth_max_m > s.depth_max_m") == 0)

invisible(dbExecute(con, "DROP TABLE IF EXISTS ctd_measurement_src"))
invisible(dbExecute(con, "ALTER TABLE ctd_measurement RENAME TO ctd_measurement_src"))
invisible(dbExecute(con, glue(
  "CREATE OR REPLACE VIEW ctd_measurement AS
   SELECT o.obs_id AS measurement_id, o.sample_key, o.depth_max_m AS depth_m,
          o.measurement_type, o.measurement_value, o.measurement_qual
   FROM obs o WHERE o.dataset_key = '{ds_key}'")))
```

### Declared NULLs

```{r}
#| label: validate
results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("validate_for_release:", ifelse(results$passed, "PASSED",
    "reported findings -- reconciled against the declared NULLs below"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")

n_null <- function(col) q1(glue(
  "SELECT COUNT(*) FROM sample WHERE dataset_key = '{ds_key}' AND {col} IS NULL"))
nullable <- tribble(
  ~table,   ~column,             ~n,                          ~reason,
  "sample", "parent_sample_key", n_null("parent_sample_key"), "single-level dataset: every cast is a root",
  "sample", "grid_key",          n_null("grid_key"),          "cast outside the CalCOFI grid (the NOAA cruises off Oregon, Washington and British Columbia)",
  "sample", "site_key",          n_null("site_key"),          "negative line or station number (extended numbering north of the CalCOFI grid): not a CalCOFI station")
nullable |> datatable(caption = "Declared NULLs (count and reason)", rownames = FALSE)
stopifnot(
  "every cast is a root" = nullable$n[nullable$column == "parent_sample_key"] == n_pub,
  "no cast lacks a time or cruise_key" = n_null("datetime") == 0 && n_null("cruise_key") == 0,
  "a NULL site_key only where the line or station is negative" =
    n_null("site_key") == q1("SELECT COUNT(*) FROM ctd_cast_pub WHERE line < 0 OR station < 0"))
```

### Core ERD

```{r}
#| label: erd-core
core_tbls <- core_output_tables(con)
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls))
```

## Write Parquet Outputs

```{r}
#| label: write-parquet
dir_create(dir_parquet)
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
parquet_stats <- write_parquet_outputs(
  con = con, output_dir = dir_parquet,
  tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE)
build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)
parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata
d_tbls_rd <- read_csv(file.path(dir_meta, "tbls_redefine.csv"))
d_flds_rd <- read_csv(file.path(dir_meta, "flds_redefine.csv"))
metadata_path <- build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           file.path(dir_meta, "metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned,
  sources = src_stamp)
```

## Upload to GCS

```{r}
#| label: upload-gcs
if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet,
              gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else cat("CALCOFI_SKIP_GCS set -- skipping the upload (local verification run)\n")
```

## Questions for Data Providers

Follow-up questions for NOAA SWFSC (Ed Weber), ranked. Q01 (the overlap with
CalCOFI's own casts) decides how much of this dataset is new.

```{r}
#| label: provider-questions
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the swfsc ctd-noaa data providers (ranked)")
```

## Cleanup

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("parquet outputs written to: {dir_parquet}"), "\n")
```
