---
title: "Ingest CalCOFI CTD Cast Data"
calcofi:
  target_name: ingest_calcofi_ctd_cast
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/calcofi_ctd-cast/manifest.json
  modifies:
    - ship
---

## Overview

This notebook ingests CalCOFI CTD data from <https://calcofi.org/data/oceanographic-data/ctd-cast-files/>, downloads and unzips **final** and **preliminary** CTD files, normalizes into tidy tables (`ctd_cast`, `ctd_measurement`, `ctd_summary`), and exports to Parquet for the CalCOFI integrated database.

### Key Features

1.  **Web Scraping**: Scrapes all CTD .zip download links from calcofi.org

2.  **Smart Filtering**:
    -   Downloads all .zip files for archival completeness
    -   Only unzips `final` and `preliminary` files
    -   Skips raw/cast/test files

3.  **Priority-based Selection**:
    -   For each cruise, selects final if available, otherwise preliminary
    -   Excludes raw/test/prodo cast files

4.  **Tidy Normalization**:
    -   `ctd_cast`: one row per unique cast (cruise/station/direction)
    -   `ctd_measurement`: long-format sensor readings at each depth
    -   `ctd_summary`: summary stats per station/depth/measurement_type across cast directions
    -   `measurement_type`: reference table for measurement codes

5.  **Standardized Workflow**: follows patterns from `ingest_calcofi_bottle.qmd` — deterministic UUIDs, GCS parquet uploads, metadata sidecars, calcofi4db utilities

## Setup

```{r}
#| label: setup

# chunk timing hook — prints elapsed time for each chunk
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    .time_it_t0 <<- Sys.time()
  } else {
    elapsed <- round(difftime(Sys.time(), .time_it_t0, units = "secs"), 1)
    tnow <- format(Sys.time(), "%H:%M:%S")
    message(glue::glue("R chunk {options$label}: {elapsed}s ~ {tnow}"))
  }
})
knitr::opts_chunk$set(time_it = TRUE)

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI / calcofi4db,
  CalCOFI / calcofi4r,
  DBI,
  dplyr,
  DT,
  fs,
  glue,
  here,
  httr,
  janitor,
  lubridate,
  mapview,
  plotly,
  purrr,
  ps,
  readr,
  rvest,
  sf,
  stringr,
  tibble,
  tidyr,
  zip,
  quiet = T
)

dataset_name <- "CalCOFI CTD Cast Database"
provider <- "calcofi"
dataset <- "ctd-cast"
dir_label <- glue("{provider}_{dataset}")
dir_data <- "~/My Drive/projects/calcofi/data-public"
dir_dl <- path_expand(glue("{dir_data}/{provider}/{dataset}/download"))
dir_parquet <- here(glue("data/parquet/{dir_label}"))
db_path <- here(glue("data/wrangling/{dir_label}.duckdb"))
db_checkpoint <- here(glue("data/wrangling/{dir_label}_checkpoint.duckdb"))
dir_tmp <- here(glue("data/tmp/{dir_label}"))
url <- "https://calcofi.org/data/oceanographic-data/ctd-cast-files/"
dir_meta <- here(glue("metadata/{provider}/{dataset}"))

# use checkpoint DB if available (skips expensive read+bind+filter steps)
if (file_exists(db_checkpoint)) {
  if (file_exists(db_path)) file_delete(db_path)
  file_copy(db_checkpoint, db_path, overwrite = TRUE)
  message(glue("Restored from checkpoint: {db_checkpoint}"))
} else if (file_exists(db_path)) {
  file_delete(db_path)
}
dir_create(c(dir_dl, dirname(db_path), dir_parquet, dir_tmp), recurse = TRUE)

con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
load_duckdb_extension(con, "icu")

# limit memory to half of system RAM so DuckDB spills to disk instead of consuming all RAM
mem_gb <- ps::ps_system_memory()$total / 1024^3 / 2 |> floor()
dbExecute(con, glue("SET memory_limit = '{mem_gb}GB'"))
dbExecute(con, glue("SET temp_directory = '{dir_tmp}'"))

# load metadata
d_meas_type <- read_csv(
  here("metadata/measurement_type.csv"),
  show_col_types = F
)
d_flds_rd <- read_csv(glue("{dir_meta}/flds_redefine.csv"), show_col_types = F)
d_tbls_rd <- read_csv(glue("{dir_meta}/tbls_redefine.csv"), show_col_types = F)
```

## Check for Resumable State

```{r}
#| label: check_resume

# detect if parquet outputs are already complete (e.g. prior run failed
# only during GCS upload). If so, skip ALL computation and jump to upload.
parquet_complete <- FALSE
manifest_path <- file.path(dir_parquet, "manifest.json")
if (file_exists(manifest_path)) {
  mf <- jsonlite::read_json(manifest_path)
  expected <- setdiff(mf$tables, unlist(mf$supplemental))
  parquet_ok <- all(vapply(expected, function(tbl) {
    p <- file.path(dir_parquet, paste0(tbl, ".parquet"))
    d <- file.path(dir_parquet, tbl)
    file_exists(p) || dir_exists(d)
  }, logical(1)))
  if (parquet_ok) {
    parquet_complete <- TRUE
    message(glue(
      "Parquet output already complete ({length(mf$tables)} tables, ",
      "{format(mf$total_rows, big.mark = ',')} rows) — ",
      "skipping computation, resuming at upload"))
  }
}

# if parquet not complete, check for checkpoint (ctd_raw pre-computed)
has_ctd_raw <- FALSE
if (!parquet_complete) {
  has_ctd_raw <- "ctd_raw" %in% DBI::dbListTables(con)
  if (has_ctd_raw) {
    n_raw <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_raw")$n
    message(glue(
      "Checkpoint: ctd_raw already loaded ",
      "({format(n_raw, big.mark = ',')} rows) — ",
      "skipping read+bind+filter"))
  }
}

# set eval for read+bind+filter chunks
skip_read_bind <- parquet_complete || has_ctd_raw
knitr::opts_chunk$set(eval = !skip_read_bind)
```

## Scrape CTD Download Links

```{r}
#| label: d_zips

# read web page, extract all .zip download links; cache to CSV
cache_csv <- file.path(dir_meta, "ctd_zip_urls.csv")

d_zips <- tryCatch(
  {
    d <- read_html(url) |>
      html_nodes("a[href$='.zip']") |>
      html_attr("href") |>
      tibble(url = _) |>
      mutate(
        url = if_else(
          str_starts(url, "http"),
          url,
          paste0("https://calcofi.org", url)
        ),
        file_zip = basename(url),
        year = str_extract(url, "/(\\d{4})/", group = 1) |> as.integer(),
        month = str_extract(url, "-\\d{2}(\\d{2})", group = 1) |> as.integer(),
        cruise_key = str_extract(
          file_zip,
          "\\d{2}-(\\d{4}[A-Z0-9]{2,4})",
          group = 1
        ),
        zip_type = case_when(
          str_detect(file_zip, "CTDFinal") ~ "final",
          str_detect(file_zip, "CTDPrelim") ~ "preliminary",
          str_detect(file_zip, "CTDCast|CTD_Cast") ~ "raw",
          str_detect(file_zip, "CTDTest") ~ "test",
          TRUE ~ "unknown"
        )
      )
    write_csv(d, cache_csv)
    message(glue("Scraped {nrow(d)} zip URLs from {url}"))
    d
  },
  error = function(e) {
    if (file_exists(cache_csv)) {
      message(glue("Website unavailable, using cached URLs from {cache_csv}"))
      read_csv(cache_csv, show_col_types = FALSE)
    } else {
      stop(e)
    }
  }
)

stopifnot(!any(d_zips$zip_type == "unknown"))

d_zips |>
  mutate(
    file_zip = glue("<a href={url}>{file_zip}</a>")
  ) |>
  select(-url) |>
  dt(
    caption = "All zip files to download",
    fname = "ctd_zip_files",
    escape = F
  )
```

## Download and Unzip Files

```{r}
#| label: download_and_unzip

download_and_unzip <- function(url, dest_dir, zip_type, unzip = TRUE) {
  file_zip <- basename(url)
  dest_file <- file.path(dest_dir, file_zip)
  dir_unzip <- file.path(dest_dir, str_remove(file_zip, "\\.zip$"))

  if (file_exists(dest_file)) {
    message(glue("Already exists: {file_zip}"))
  } else {
    message(glue("Downloading: {file_zip}"))
    tryCatch(
      download.file(url, dest_file, mode = "wb"),
      error = function(e) {
        warning(glue("Failed to download {file_zip}: {e$message}"))
        if (file_exists(dest_file)) {
          file_delete(dest_file)
        }
        return(invisible(NULL))
      }
    )
    if (!file_exists(dest_file)) return(invisible(NULL))
  }

  if (unzip && dir_exists(dir_unzip)) {
    message(glue("Already unzipped: {file_zip}"))
  } else {
    if (zip_type %in% c("final", "preliminary")) {
      message(glue("Unzipping: {file_zip}"))
      dir_create(dir_unzip, recurse = TRUE)
      unzip(dest_file, exdir = dir_unzip)
    } else if (unzip) {
      message(glue("Skipping unzip (not final/preliminary): {file_zip}"))
    }
  }
}

d_zips |>
  pwalk(function(url, file_zip, zip_type, ...) {
    download_and_unzip(url, dir_dl, zip_type, unzip = TRUE)
  })

message("Download and extraction complete!")
```

## Find and Prioritize CTD Data Files

```{r}
#| label: d_csv

d_csv <- tibble(
  path = list.files(
    dir_dl,
    pattern = "\\.csv$",
    recursive = TRUE,
    full.names = TRUE
  )
) |>
  mutate(
    file_csv = basename(path),
    path_unzip = str_replace(path, glue("{dir_dl}/"), ""),
    dir_unzip = str_extract(path_unzip, "^[^/]+"),
    cruise_key = str_extract(
      path_unzip,
      "\\d{2}-(\\d{4}[A-Z0-9]{2,4})_.*",
      group = 1
    ),
    data_stage = case_when(
      str_detect(path_unzip, "Final.*db[_|-]csv")  ~ "final",
      str_detect(path_unzip, "Prelim.*db[_|-]csv") ~ "preliminary",
      # prelim CSVs directly in unzipped dir (no db-csv subfolder)
      str_detect(dir_unzip, "Prelim") &
        path_unzip == paste0(dir_unzip, "/", file_csv) ~ "preliminary",
      # edge case: 2111SR has CSVs only in csvs-plots subfolder
      cruise_key == "2111SR" &
        str_detect(
          path_unzip,
          "Prelim.*csvs-plots.*2111SR.*csv"
        ) ~ "preliminary",
      .default = NA_character_
    ),
    cast_dir = case_when(
      str_detect(file_csv, regex("U\\.csv$", ignore_case = T)) ~ "U",
      str_detect(file_csv, regex("D\\.csv$", ignore_case = T)) ~ "D"
    ),
    priority = case_when(
      data_stage == "final" ~ 1,
      data_stage == "preliminary" ~ 2,
      TRUE ~ 3
    )
  ) |>
  relocate(cruise_key, path_unzip) |>
  arrange(cruise_key, path_unzip) |>
  filter(data_stage %in% c("final", "preliminary"))

# for each cruise, keep only final if available, otherwise preliminary
d_priority <- d_csv |>
  group_by(cruise_key) |>
  summarize(
    best_priority = min(priority),
    .groups = "drop"
  )

d_csv <- d_csv |>
  inner_join(d_priority, by = "cruise_key") |>
  filter(priority == best_priority) |>
  select(-best_priority)

cruises_csv_notzip <- setdiff(
  unique(d_csv$cruise_key),
  unique(d_zips$cruise_key)
) |>
  sort()
stopifnot(length(cruises_csv_notzip) == 0)

cruises_zip_notcsv <- setdiff(
  unique(d_zips$cruise_key),
  unique(d_csv$cruise_key)
) |>
  sort()

d_csv |>
  select(-any_of(c("path", "data", "col_empties", "col_types"))) |>
  relocate(cruise_key, path_unzip) |>
  arrange(cruise_key, path_unzip) |>
  dt(
    caption = "Files to Ingest",
    fname = "ctd_files_to_ingest"
  )
```

## Read and Standardize CTD Files

Remove repeat header rows that can occur within files.

```{r}
#| label: read_csv

d_csv <- d_csv |>
  arrange(basename(path)) |>
  mutate(
    data = map2(path, seq_along(path), \(path, idx) {
      message(glue("Reading {idx}/{nrow(d_csv)}: {basename(path)}"))

      all_lines <- read_lines(path)
      header_line <- all_lines[1]

      # find repeat header rows (excluding row 1)
      repeat_header_rows <- which(all_lines[-1] == header_line)

      if (length(repeat_header_rows) > 0) {
        all_lines <- all_lines[-(repeat_header_rows + 1)]
        tmp_file <- tempfile(fileext = ".csv")
        write_lines(all_lines, tmp_file)
        data <- read_csv(tmp_file, guess_max = Inf, show_col_types = F) |>
          clean_names()
        file_delete(tmp_file)
      } else {
        data <- read_csv(path, guess_max = Inf, show_col_types = F) |>
          clean_names()
      }
      data
    }),
    nrows = map_int(data, ~ if (is.null(.x)) 0 else nrow(.x))
  )

d_csv |>
  arrange(cruise_key, file_csv) |>
  select(cruise_key, file_csv, nrows) |>
  dt(
    caption = "Number of rows read per file",
    fname = "ctd_rows_per_file"
  ) |>
  formatCurrency("nrows", currency = "", digits = 0, mark = ",")
```

## Detect and Correct Column Type Mismatches

```{r}
#| label: d_bind

d_csv <- d_csv |>
  mutate(
    col_empties = map(data, \(x) {
      tibble(
        col_name = names(x),
        n_empty = map_int(x, \(col) sum(is.na(col)))
      ) |>
        filter(n_empty == nrow(x)) |>
        pull(col_name)
    }),
    col_types = map2(data, col_empties, \(x, y) {
      tibble(
        col_name = names(x),
        col_type = map_chr(x, \(col) class(col)[1])
      ) |>
        filter(!col_name %in% y)
    })
  )

# find most common type for each column across all files
d_types <- d_csv |>
  select(path, col_types) |>
  unnest(col_types) |>
  count(col_name, col_type) |>
  group_by(col_name) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(col_name, expected_type = col_type)

d_mismatches <- d_csv |>
  select(cruise_key, path, col_types) |>
  unnest(col_types) |>
  left_join(d_types, by = "col_name") |>
  filter(col_type != expected_type) |>
  arrange(col_name, path)

if (nrow(d_mismatches) > 0) {
  message("Type mismatches detected - converting columns...")
}

# bind data, converting mismatched columns to expected type
d_bind <- d_csv |>
  mutate(
    data = map2(data, path, \(x, p) {
      x_mismatches <- d_mismatches |>
        filter(path == p)

      if (nrow(x_mismatches) > 0) {
        for (i in 1:nrow(x_mismatches)) {
          col <- x_mismatches$col_name[i]
          expected <- x_mismatches$expected_type[i]
          na_before <- sum(is.na(x[[col]]))

          suppressWarnings({
            x[[col]] <- switch(
              expected,
              "numeric" = as.numeric(x[[col]]),
              "integer" = as.integer(x[[col]]),
              "logical" = as.logical(x[[col]]),
              "character" = as.character(x[[col]]),
              x[[col]]
            )
          })

          na_after <- sum(is.na(x[[col]]))
          na_generated <- na_after - na_before

          if (na_generated > 0) {
            message(glue(
              "  {basename(p)}: {col} ({x_mismatches$col_type[i]} -> {expected}) generated {na_generated} NAs"
            ))
          }

          d_mismatches[
            d_mismatches$path == p & d_mismatches$col_name == col,
            "nas_generated"
          ] <<- na_generated
        }
      }
      x
    })
  ) |>
  unnest(data)

if (nrow(d_mismatches) > 0) {
  d_mismatches |>
    group_by(col_name, expected_type, col_type) |>
    summarize(
      n_files = n(),
      total_nas = sum(nas_generated, na.rm = TRUE),
      files = paste(basename(path), collapse = "; "),
      .groups = "drop"
    ) |>
    arrange(desc(total_nas)) |>
    dt(
      caption = "Type mismatches by column",
      fname = "ctd_type_mismatches_by_column"
    ) |>
    formatCurrency(
      c("n_files", "total_nas"),
      currency = "",
      digits = 0,
      mark = ","
    )
}

# reconcile cruise_key from study where mismatched
d_bind <- d_bind |>
  mutate(
    cruise_key = if_else(
      !is.na(study) &
        cruise_key != study &
        study != "Study",
      study,
      cruise_key
    ),
    `_source_file` = path_unzip
  ) |>
  relocate(`_source_file`, .after = cruise_key) |>
  select(
    -path_unzip,
    -path,
    -file_csv,
    -dir_unzip,
    -priority,
    -project,
    -study,
    -nrows,
    -col_empties,
    -col_types
  )

# fill missing ord_occ from cast_id
d_bind <- d_bind |>
  mutate(
    ord_occ = if_else(
      is.na(ord_occ) & !is.na(cast_id),
      str_extract(cast_id, "_(\\d{3})", group = 1),
      ord_occ
    )
  )

# save intermediate for resumability
write_rds(d_bind, glue("{dir_tmp}/d_bind_pre_datetime.rds"), compress = "gz")
```

## Format Date-Time Column

```{r}
#| label: format_datetime

d_bind <- read_rds(glue("{dir_tmp}/d_bind_pre_datetime.rds"))

d_bind <- d_bind |>
  mutate(
    date_time_utc = trimws(date_time_utc),
    date_time_format = case_when(
      str_detect(
        date_time_utc,
        "^\\d{1,2}-[A-z]{3}-\\d{4} \\d{1,2}:\\d{1,2}:\\d{1,2}$"
      ) ~ "dmy_hms",
      str_detect(
        date_time_utc,
        "^\\d{1,2}/\\d{1,2}/\\d{4} \\d{1,2}:\\d{1,2}$"
      ) ~ "mdy_hm",
      TRUE ~ "unknown"
    ),
    datetime_utc = NA,
    datetime_utc = if_else(
      date_time_format == "dmy_hms",
      suppressWarnings(dmy_hms(date_time_utc, tz = "UTC")),
      datetime_utc
    ),
    datetime_utc = if_else(
      date_time_format == "mdy_hm",
      suppressWarnings(mdy_hm(date_time_utc, tz = "UTC")),
      datetime_utc
    )
  ) |>
  relocate(date_time_format, datetime_utc, .after = date_time_utc) |>
  arrange(datetime_utc, depth)

stopifnot(all(!is.na(d_bind$datetime_utc)))

write_rds(d_bind, glue("{dir_tmp}/d_bind_post_datetime.rds"), compress = "gz")

d_bind |>
  group_by(cruise_key, `_source_file`, date_time_format) |>
  summarize(
    n_records = n(),
    datetime_min = min(datetime_utc),
    datetime_max = max(datetime_utc),
    .groups = "drop"
  ) |>
  arrange(cruise_key, `_source_file`, date_time_format) |>
  pivot_wider(
    names_from = date_time_format,
    values_from = n_records,
    values_fill = NA_real_
  ) |>
  relocate(dmy_hms, mdy_hm, .after = cruise_key) |>
  dt(
    caption = "Date-Time formats detected by cruise_key and source_file",
    fname = "ctd_datetime_formats"
  ) |>
  formatDate(c("datetime_min", "datetime_max"), method = "toLocaleString")
```

## Pseudo-NA Values and Fill Missing Coordinates

```{r}
#| label: fill_lonlat

d_bind <- read_rds(glue("{dir_tmp}/d_bind_post_datetime.rds"))

pseudoNA_values <- c(-9.99e-29, -99)

d_bind <- d_bind |>
  mutate(
    is_lnsta_pseudoNA = ifelse(
      as.numeric(line) == 0 & as.numeric(sta) == 0,
      TRUE,
      FALSE
    ),
    is_lon_pseudoNA = map_lgl(lon_dec, ~ some(pseudoNA_values, near, .x)),
    is_lat_pseudoNA = map_lgl(lat_dec, ~ some(pseudoNA_values, near, .x)),
    lon_dec = if_else(
      is_lon_pseudoNA | is_lat_pseudoNA,
      NA_real_,
      lon_dec
    ),
    lat_dec = if_else(
      is_lon_pseudoNA | is_lat_pseudoNA,
      NA_real_,
      lat_dec
    ),
    lon_lnst = if_else(
      !is.na(line) & !is.na(sta) & !is_lnsta_pseudoNA,
      as.numeric(sf_project(
        from = "+proj=calcofi",
        to = "+proj=longlat +datum=WGS84",
        pts = cbind(x = as.numeric(line), y = as.numeric(sta))
      )[, 1]),
      NA_real_
    ),
    lat_lnst = if_else(
      !is.na(line) & !is.na(sta) & !is_lnsta_pseudoNA,
      as.numeric(sf_project(
        from = "+proj=calcofi",
        to = "+proj=longlat +datum=WGS84",
        pts = cbind(x = as.numeric(line), y = as.numeric(sta))
      )[, 2]),
      NA_real_
    ),
    lon_dec = if_else(is.na(lon_dec), lon_lnst, lon_dec),
    lat_dec = if_else(is.na(lat_dec), lat_lnst, lat_dec)
  )

stopifnot(sum(is.na(d_bind$lon_dec)) == 0)
stopifnot(sum(is.na(d_bind$lat_dec)) == 0)

d_bind |>
  filter(is_lon_pseudoNA | is_lat_pseudoNA) |>
  group_by(cruise_key) |>
  summarize(
    n_lon_pseudoNA = sum(is_lon_pseudoNA),
    n_lat_pseudoNA = sum(is_lat_pseudoNA)
  ) |>
  dt(
    caption = glue(
      "Cruises with pseudo-NAs ({paste(pseudoNA_values, collapse = ',')}) ",
      "set to NA, then filled from line/station coordinates."
    ),
    fname = "ctd_cruises_pseudoNA_lonlat"
  ) |>
  formatCurrency(
    c("n_lon_pseudoNA", "n_lat_pseudoNA"),
    currency = "",
    digits = 0,
    mark = ","
  )
```

## Distance Filtering

```{r}
#| label: pts_distance_from_lnst

max_dist_dec_lnst_km <- 10

pts <- d_bind |>
  st_as_sf(coords = c("lon_dec", "lat_dec"), remove = F, crs = 4326) |>
  mutate(
    geom_lnst = purrr::map2(lon_lnst, lat_lnst, \(x, y) st_point(c(x, y))) |>
      st_sfc(crs = 4326),
    dist_dec_lnst_km = st_distance(geometry, geom_lnst, by_element = T) |>
      units::set_units(km) |>
      units::drop_units(),
    is_dist_dec_lnst_within_max = dist_dec_lnst_km <= max_dist_dec_lnst_km
  ) |>
  st_join(
    calcofi4r::cc_grid |>
      rename(any_of(c(site_key = "sta_key"))) |>
      select(grid_site = site_key),
    join = st_intersects
  )
st_agr(pts) <- "constant"
```

### View sample cruise with excess distance points

```{r}
#| label: view_badcr

badcr_id <- "1507OC"
pts_badcr <- pts |>
  filter(cruise_key == badcr_id)

bb_badcr <- st_bbox(pts_badcr) |> st_as_sfc()

mapView(bb_badcr) +
  mapView(
    pts_badcr |>
      filter(is_dist_dec_lnst_within_max) |>
      slice_sample(n = 1000) |>
      bind_rows(
        pts_badcr |>
          filter(!is_dist_dec_lnst_within_max)
      ) |>
      select(
        cruise_key,
        datetime_utc,
        lon_dec,
        lat_dec,
        sta_id,
        line,
        sta,
        dist_dec_lnst_km,
        is_dist_dec_lnst_within_max
      ),
    layer.name = glue(
      "Cruise {badcr_id}<br>distance (km)<br>lon/lat to line/station"
    ),
    zcol = "dist_dec_lnst_km",
    cex = 5,
    alpha = 0.5
  )
```

### Filter points

```{r}
#| label: filter_pts

d_pts_cruise_filt_smry <- pts |>
  st_drop_geometry() |>
  group_by(cruise_key) |>
  filter(any(!is_dist_dec_lnst_within_max)) |>
  summarize(
    n_all = n(),
    n_outside_grid = sum(is.na(grid_site)),
    pct_outside_grid = n_outside_grid / n_all,
    n_gt_cutoff = sum(!is_dist_dec_lnst_within_max, na.rm = T),
    avg_dist_gt_cutoff_km = if_else(
      n_gt_cutoff > 0,
      mean(dist_dec_lnst_km[!is_dist_dec_lnst_within_max], na.rm = T),
      NA_real_
    ),
    pct_gt_cutoff = n_gt_cutoff / n_all,
    n_rm = sum(!is_dist_dec_lnst_within_max | is.na(grid_site), na.rm = T),
    pct_rm = n_rm / n_all,
    .groups = "drop"
  ) |>
  filter(n_rm > 0) |>
  arrange(desc(pct_rm))

pts_filt <- pts |>
  filter(
    is_dist_dec_lnst_within_max,
    !is.na(grid_site)
  )

d_pts_cruise_filt_smry |>
  dt(
    caption = glue(
      "Cruises with rows filtered: outside CalCOFI grid or exceeded ",
      "{max_dist_dec_lnst_km} km cutoff from line/station coordinates."
    ),
    escape = F,
    fname = "ctd_cruises_distance_from_lnst"
  ) |>
  formatCurrency(
    c(
      "n_all",
      "n_gt_cutoff",
      "avg_dist_gt_cutoff_km",
      "n_outside_grid",
      "n_rm"
    ),
    currency = "",
    digits = 0,
    mark = ","
  ) |>
  formatPercentage(
    c("pct_gt_cutoff", "pct_outside_grid", "pct_rm"),
    digits = 2
  )
```

### View filtered points

```{r}
#| label: view_pts_filt

bb_cr <- st_bbox(pts_filt) |> st_as_sfc()

mapView(bb_cr) +
  mapView(
    pts_filt |>
      group_by(cruise_key) |>
      slice_sample(n = 100) |>
      ungroup() |>
      select(
        cruise_key,
        lon_dec,
        lat_dec,
        sta_id,
        line,
        sta,
        dist_dec_lnst_km,
        datetime_utc
      ),
    zcol = "cruise_key",
    cex = 5,
    alpha = 0.5
  )
```

### Check for duplicates by datetime and depth

```{r}
#| label: check_dupes

d_dupes <- pts_filt |>
  st_drop_geometry() |>
  select(datetime_utc, depth) |>
  group_by(datetime_utc, depth) |>
  summarize(n = n(), .groups = "drop") |>
  filter(n > 1) |>
  arrange(desc(n))

n_distinct_dtime_depth <- pts_filt |>
  st_drop_geometry() |>
  select(datetime_utc, depth) |>
  n_distinct()

d_dupes |>
  slice(c(1:10, (n() - 9):n())) |>
  dt(
    caption = glue(
      "First and last 10 duplicates by datetime_utc, depth ",
      "(nrows = {format(nrow(d_dupes), big.mark = ',')}; ",
      "n_distinct = {format(n_distinct_dtime_depth, big.mark = ',')})"
    ),
    fname = "ctd_duplicates_by_datetime_depth"
  )
```

## Rename Fields

Apply field renames from `flds_redefine.csv`.

```{r}
#| label: rename_fields

# apply column renames from flds_redefine
renames <- d_flds_rd |>
  filter(fld_old != fld_new) |>
  select(fld_old, fld_new)

d_filt <- pts_filt |>
  st_drop_geometry() |>
  select(-geom_lnst)

for (i in seq_len(nrow(renames))) {
  old <- renames$fld_old[i]
  new <- renames$fld_new[i]
  if (old %in% names(d_filt)) {
    d_filt <- d_filt |> rename(!!new := !!old)
  }
}

# drop intermediate columns
d_filt <- d_filt |>
  select(
    -any_of(c(
      "date_time_pst",
      "date_time_utc",
      "date_time_format",
      "is_lon_pseudoNA",
      "is_lat_pseudoNA",
      "is_lnsta_pseudoNA",
      "lon_lnst",
      "lat_lnst",
      "dist_dec_lnst_km",
      "is_dist_dec_lnst_within_max",
      "grid_site"
    ))
  )

# write raw table for pivoting
dbWriteTable(con, "ctd_raw", d_filt, overwrite = TRUE)
message(glue("Loaded {nrow(d_filt)} rows into ctd_raw"))
```

```{r}
#| label: restore_eval

# if parquet already complete, skip computation through write_parquet
# only re-enable for upload_gcs and cleanup
if (parquet_complete) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Parquet complete — skipping to upload")
} else {
  knitr::opts_chunk$set(eval = TRUE)
}
```

## Cross-Dataset Bridge

Load reference tables (ship, cruise, grid) from the ichthyo ingest, derive `ship_key`, and validate against known ships/cruises.

```{r}
#| label: cross_dataset_bridge

# tables declared in calcofi.modifies — loaded as TABLEs (writable)
modifies_tables <- c("ship")  # from calcofi.modifies in YAML frontmatter
load_prior_tables(
  con         = con,
  tables      = modifies_tables,
  parquet_dir = here("data/parquet/swfsc_ichthyo")
)
# cruise + grid as VIEWs (read-only: cruise used for JOIN, grid for assign_grid_key)
load_prior_tables(
  con         = con,
  tables      = c("cruise", "grid"),
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  as_view     = TRUE
)

# snapshot PKs of modified tables before modifications
modifies_pks <- list()
for (tbl in modifies_tables) {
  pk_col <- dbGetQuery(con, glue(
    "SELECT column_name FROM information_schema.columns
     WHERE table_name = '{tbl}' ORDER BY ordinal_position LIMIT 1"))$column_name
  modifies_pks[[tbl]] <- list(
    pk_col = pk_col,
    keys   = dbGetQuery(con, glue("SELECT {pk_col} FROM {tbl}"))[[1]])
}

# detect whether cruise_key already in YYYY-MM-NODC format (from checkpoint)
sample_ck <- dbGetQuery(
  con, "SELECT cruise_key FROM ctd_raw LIMIT 1")$cruise_key
cruise_key_already_converted <- grepl("^\\d{4}-\\d{2}-", sample_ck)

if (!cruise_key_already_converted) {
  # 0404NHJD is a combined New Horizon + Jordan Davis cruise; use primary ship NH
  dbExecute(
    con,
    "UPDATE ctd_raw SET cruise_key = LEFT(cruise_key, 6)
     WHERE LENGTH(cruise_key) > 6"
  )

  # derive ship_key from last 2 chars of cruise_key (YYMMKK format from filename)
  dbExecute(con, "ALTER TABLE ctd_raw ADD COLUMN IF NOT EXISTS ship_key VARCHAR")
  dbExecute(con, "UPDATE ctd_raw SET ship_key = RIGHT(cruise_key, 2)")

  # validate ship_key against ship reference table
  ctd_ships <- dbGetQuery(con, "SELECT DISTINCT ship_key FROM ctd_raw")
  ref_ships <- dbGetQuery(con, "SELECT ship_key FROM ship")
  orphan_ships <- setdiff(ctd_ships$ship_key, ref_ships$ship_key)

  if (length(orphan_ships) > 0) {
    message(glue(
      "{length(orphan_ships)} ship_key(s) in CTD not in ship table: ",
      "{paste(orphan_ships, collapse = ', ')}"
    ))

    # run match_ships() with centralized renames for orphan ships
    orphan_tbl <- dbGetQuery(con, glue(
      "SELECT DISTINCT ship_key AS ship_code, NULL AS ship_name
       FROM ctd_raw
       WHERE ship_key IN ({paste(shQuote(orphan_ships, type = 'sh'), collapse = ', ')})")) |>
      mutate(ship_name = as.character(ship_name))

    ship_result <- match_ships(
      unmatched_ships  = orphan_tbl,
      reference_ships  = dbReadTable(con, "ship"),
      ship_renames_csv = here("metadata/ship_renames.csv"),
      fetch_ices       = FALSE)

    # insert interim entries for still-unmatched ships (ship_nodc = "?XX?")
    ensure_interim_ships(con, ship_result)

    dbGetQuery(
      con,
      glue(
        "SELECT ship_key, COUNT(DISTINCT cruise_key) AS n_cruises,
                COUNT(*) AS n_rows
         FROM ctd_raw
         WHERE ship_key IN ({paste(shQuote(orphan_ships, type = 'sh'), collapse = ', ')})
         GROUP BY ship_key
         ORDER BY ship_key"
      )
    ) |>
      dt(
        caption = "Orphan ship_keys (in CTD but not in ship table)",
        fname = "ctd_orphan_ship_keys"
      )
  } else {
    message("All CTD ship_keys found in ship reference table")
  }

  # convert cruise_key from YYMMKK → YYYY-MM-NODC format
  convert_cruise_key_format(con, "ctd_raw", old_key_col = "cruise_key")
  dbExecute(con, "ALTER TABLE ctd_raw DROP COLUMN cruise_key")
  dbExecute(con, "ALTER TABLE ctd_raw RENAME COLUMN cruise_key_new TO cruise_key")
} else {
  message("cruise_key already in YYYY-MM-NODC format (from checkpoint)")
}

# validate cruise_key against cruise reference table
ctd_cruises <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM ctd_raw")
ref_cruises <- dbGetQuery(con, "SELECT cruise_key FROM cruise")
orphan_cruises <- setdiff(ctd_cruises$cruise_key, ref_cruises$cruise_key)

if (length(orphan_cruises) > 0) {
  message(glue(
    "{length(orphan_cruises)} cruise_key(s) in CTD not in cruise table: ",
    "{paste(head(orphan_cruises, 10), collapse = ', ')}",
    "{if (length(orphan_cruises) > 10) '...' else ''}"
  ))
  dbGetQuery(
    con,
    glue(
      "SELECT cruise_key, COUNT(*) AS n_rows
       FROM ctd_raw
       WHERE cruise_key IN ({paste(shQuote(orphan_cruises, type = 'sh'), collapse = ', ')})
       GROUP BY cruise_key
       ORDER BY cruise_key"
    )
  ) |>
    dt(
      caption = "Orphan cruise_keys (in CTD but not in cruise table)",
      fname = "ctd_orphan_cruise_keys"
    )
} else {
  message("All CTD cruise_keys found in cruise reference table")
}
```

## Save Checkpoint

Save a DuckDB checkpoint after the expensive read+bind+filter+bridge steps so subsequent runs can resume from here.

```{r}
#| label: save_checkpoint

if (!file_exists(db_checkpoint)) {
  close_duckdb(con)
  file_copy(db_path, db_checkpoint, overwrite = TRUE)
  con <- get_duckdb_con(db_path)
  load_duckdb_extension(con, "spatial")
  load_duckdb_extension(con, "icu")
  message(glue("Saved checkpoint: {db_checkpoint}"))
} else {
  message(glue("Checkpoint already exists: {db_checkpoint}"))
}
```

## Wide-Format Output (for ERDDAP)

Create a wide-format table preserving the original column layout for serving into ERDDAP. This is a supplemental output — not part of the normalized database.

```{r}
#| label: ctd_wide

# identify measurement and quality columns from measurement_type metadata
d_meas_ctd <- d_meas_type |>
  filter(str_detect(`_source_datasets`, "calcofi_ctd-cast"))
meas_cols <- d_meas_ctd$`_source_column`
qual_cols <- d_meas_ctd$`_qual_column` |> na.omit() |> as.character()
qual_cols <- qual_cols[qual_cols != ""]

# get raw columns
raw_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns WHERE table_name = 'ctd_raw'"
)$column_name

# wide-format: cast metadata + depth + all measurement/quality columns
# exclude provenance and intermediate columns
wide_keep <- intersect(
  raw_cols,
  c("cruise_key", "site_key", "cast_key", "cast_dir",
    "datetime_utc", "lon_dec", "lat_dec", "ship_key",
    "data_stage", "depth_m", meas_cols, qual_cols))

dbExecute(
  con,
  glue(
    "CREATE OR REPLACE TABLE ctd_wide AS
     SELECT {paste(wide_keep, collapse = ', ')}
     FROM ctd_raw
     ORDER BY cruise_key, site_key, cast_dir, depth_m")
)

n_wide <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_wide")$n
message(glue("ctd_wide: {format(n_wide, big.mark = ',')} rows (supplemental, for ERDDAP)"))
```

## Split into Tidy Tables

### ctd_cast — cast-level metadata

```{r}
#| label: ctd_cast

# d_meas_ctd, meas_cols, qual_cols already defined in ctd_wide chunk above

# use ctd_raw columns (includes ship_key and data_stage added via SQL)
raw_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns WHERE table_name = 'ctd_raw'"
)$column_name

cast_cols <- setdiff(
  raw_cols,
  c(meas_cols, qual_cols, "depth_m", "_source_file")
)

# extract unique casts
dbExecute(
  con,
  glue(
    "
  CREATE OR REPLACE TABLE ctd_cast AS
  SELECT DISTINCT {paste(cast_cols, collapse = ', ')}
  FROM ctd_raw
  ORDER BY datetime_utc, cruise_key"
  )
)

n_cast <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_cast")$n
message(glue("ctd_cast: {format(n_cast, big.mark = ',')} rows"))

# assign deterministic UUID from composite natural key (md5-based, DuckDB-native)
assign_deterministic_uuids_md5(
  con = con,
  table_name = "ctd_cast",
  id_col = "ctd_cast_uuid",
  key_cols = c("cruise_key", "cast_key", "cast_dir", "datetime_utc")
)
```

### ctd_measurement — long-format sensor readings

```{r}
#| label: ctd_measurement

# add ctd_cast_uuid to ctd_raw for FK linkage
dbExecute(
  con,
  "
  ALTER TABLE ctd_raw ADD COLUMN IF NOT EXISTS ctd_cast_uuid VARCHAR"
)
dbExecute(
  con,
  "
  UPDATE ctd_raw AS r
  SET ctd_cast_uuid = c.ctd_cast_uuid
  FROM ctd_cast AS c
  WHERE r.cruise_key   = c.cruise_key
    AND r.cast_key     = c.cast_key
    AND r.cast_dir     = c.cast_dir
    AND r.datetime_utc = c.datetime_utc"
)

# pivot wide-to-long one measurement type at a time to avoid OOM
dbExecute(con, "DROP TABLE IF EXISTS ctd_measurement")

for (i in seq_len(nrow(d_meas_ctd))) {
  row <- d_meas_ctd[i, ]
  src_col <- row$`_source_column`
  meas_type <- row$measurement_type
  qual_col <- row$`_qual_column`

  qual_expr <- if (!is.na(qual_col) && qual_col != "") {
    glue("CAST({qual_col} AS VARCHAR)")
  } else {
    "NULL"
  }

  sql_select <- glue(
    "
    SELECT ctd_cast_uuid, depth_m,
      '{meas_type}' AS measurement_type,
      CAST({src_col} AS DOUBLE) AS measurement_value,
      {qual_expr} AS measurement_qual
    FROM ctd_raw
    WHERE {src_col} IS NOT NULL
      AND NOT isnan(CAST({src_col} AS DOUBLE))
      AND isfinite(CAST({src_col} AS DOUBLE))
    "
  )

  if (i == 1) {
    dbExecute(con, glue("CREATE TABLE ctd_measurement AS {sql_select}"))
  } else {
    dbExecute(con, glue("INSERT INTO ctd_measurement {sql_select}"))
  }
  message(glue("  {i}/{nrow(d_meas_ctd)}: {meas_type}"))
}

n_meas <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_measurement")$n
message(glue("ctd_measurement: {format(n_meas, big.mark = ',')} rows"))

# assign deterministic UUID (md5-based, DuckDB-native — single SQL, no R data transfer)
assign_deterministic_uuids_md5(
  con = con,
  table_name = "ctd_measurement",
  id_col = "ctd_measurement_uuid",
  key_cols = c("ctd_cast_uuid", "depth_m", "measurement_type")
)

# add cruise_key for partitioned parquet output
dbExecute(
  con,
  "ALTER TABLE ctd_measurement ADD COLUMN IF NOT EXISTS cruise_key VARCHAR"
)
dbExecute(
  con,
  "UPDATE ctd_measurement AS m
   SET cruise_key = c.cruise_key
   FROM ctd_cast AS c
   WHERE m.ctd_cast_uuid = c.ctd_cast_uuid"
)
```

### ctd_summary — summary stats across cast directions

```{r}
#| label: ctd_summary

dbExecute(
  con,
  "
  CREATE OR REPLACE TABLE ctd_summary AS
  SELECT
    c.cruise_key,
    c.site_key,
    m.depth_m,
    m.measurement_type,
    AVG(m.measurement_value)    AS avg,
    CASE
      WHEN COUNT(*) = 1 THEN 0
      ELSE COALESCE(STDDEV_SAMP(m.measurement_value), 0)
    END                          AS stddev,
    COUNT(*)                     AS n_obs
  FROM ctd_measurement m
  INNER JOIN ctd_cast c ON m.ctd_cast_uuid = c.ctd_cast_uuid
  WHERE NOT isnan(m.measurement_value)
    AND isfinite(m.measurement_value)
  GROUP BY c.cruise_key, c.site_key, m.depth_m, m.measurement_type"
)

n_summ <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_summary")$n
message(glue("ctd_summary: {format(n_summ, big.mark = ',')} rows"))

assign_deterministic_uuids_md5(
  con = con,
  table_name = "ctd_summary",
  id_col = "ctd_summary_uuid",
  key_cols = c("cruise_key", "site_key", "depth_m", "measurement_type")
)
```

### measurement_type — reference table

```{r}
#| label: measurement_type

dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
message(glue("measurement_type: {nrow(d_meas_type)} rows"))

# validate: all measurement_types used by this dataset are registered
ctd_types_used <- dbGetQuery(
  con,
  "SELECT DISTINCT measurement_type FROM ctd_measurement"
)$measurement_type

registered <- d_meas_type |>
  filter(str_detect(`_source_datasets`, "calcofi_ctd-cast")) |>
  pull(measurement_type)

unregistered <- setdiff(ctd_types_used, registered)
stopifnot(
  "all measurement_types used by calcofi_ctd-cast must be registered in _source_datasets" = length(
    unregistered
  ) ==
    0
)
```

### Drop intermediate table

```{r}
#| label: drop_ctd_raw

dbExecute(con, "DROP TABLE IF EXISTS ctd_raw")
```

## Schema Diagram

```{r}
#| label: schema

# define PK/FK relationships for visualization and relationships.json
ctd_rels <- list(
  primary_keys = list(
    ctd_cast        = "ctd_cast_uuid",
    ctd_measurement = "ctd_measurement_uuid",
    ctd_summary     = "ctd_summary_uuid",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "ctd_measurement", column = "ctd_cast_uuid", ref_table = "ctd_cast", ref_column = "ctd_cast_uuid")))

cc_erd(con, rels = ctd_rels)
```

## Add Spatial

```{r}
#| label: add_spatial

add_point_geom(con, "ctd_cast", lon_col = "lon_dec", lat_col = "lat_dec")

# grid already loaded in Cross-Dataset Bridge section
assign_grid_key(con, "ctd_cast")
```

## Derived View

```{r}
#| label: derived_view

dbExecute(
  con,
  "
  CREATE OR REPLACE VIEW ctd_cast_derived AS
  SELECT *,
    datetime_utc AT TIME ZONE 'UTC' AT TIME ZONE 'US/Pacific' AS datetime_pst,
    EXTRACT(YEAR FROM datetime_utc)  AS year,
    EXTRACT(MONTH FROM datetime_utc) AS month,
    EXTRACT(DOY FROM datetime_utc)   AS julian_day
  FROM ctd_cast"
)
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- read_csv(here("metadata/dataset.csv"))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
message(glue("dataset: {nrow(d_dataset)} datasets registered"))
```

## Validate and Enforce Types

```{r}
#| label: validate

validate_for_release(con)
enforce_column_types(con, d_flds_rd = d_flds_rd)
```

## Preview Tables

```{r}
#| label: preview
#| results: asis

preview_tables(
  con,
  tables = c("ctd_cast", "ctd_measurement", "ctd_summary",
             "measurement_type", "ctd_wide")
)
```

## Write Parquet

```{r}
#| label: write_parquet

# collect mismatches for manifest
mismatches <- list(
  ships             = collect_ship_mismatches(con, "ctd_cast"),
  measurement_types = collect_measurement_type_mismatches(
    con, here("metadata/measurement_type.csv")),
  cruise_keys       = collect_cruise_key_mismatches(con, "ctd_cast"))

parquet_stats <- write_parquet_outputs(
  con          = con,
  output_dir   = dir_parquet,
  tables       = c("ctd_cast", "ctd_measurement", "ctd_summary",
                    "measurement_type", "ctd_wide"),
  partition_by = list(
    ctd_measurement = "cruise_key",
    ctd_summary     = "cruise_key"
  ),
  strip_provenance = FALSE,
  mismatches       = mismatches,
  supplemental     = "ctd_wide"
)

parquet_stats |>
  dt(fname = "ctd_parquet_stats")
```

## Write Metadata JSON

```{r}
#| label: write_metadata

build_metadata_json(
  con = con,
  d_tbls_rd = d_tbls_rd,
  d_flds_rd = d_flds_rd,
  metadata_derived_csv = glue("{dir_meta}/metadata_derived.csv"),
  output_dir = dir_parquet,
  provider = provider,
  dataset = dataset
)

# write relationships.json sidecar with PKs/FKs
build_relationships_json(
  rels       = ctd_rels,
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset
)
```

## Export Modified Dependency Deltas

```{r}
#| label: export_new_deltas

# export _new delta sidecars for modified dependency tables (from calcofi.modifies)
# these are NOT in the manifest — picked up by build_release_table_registry()
for (tbl in modifies_tables) {
  pk_col   <- modifies_pks[[tbl]]$pk_col
  keys_old <- modifies_pks[[tbl]]$keys
  keys_new <- dbGetQuery(con, glue("SELECT {pk_col} FROM {tbl}"))[[1]]
  additions <- setdiff(keys_new, keys_old)

  if (length(additions) > 0) {
    pq_path <- file.path(dir_parquet, paste0(tbl, "_new.parquet"))
    vals <- paste(shQuote(additions, "sh"), collapse = ", ")
    export_parquet(con,
      glue("SELECT * FROM {tbl} WHERE {pk_col} IN ({vals})"), pq_path)
    message(glue("{length(additions)} new {tbl} row(s) -> {tbl}_new.parquet"))
  } else {
    message(glue("No new {tbl} rows — {tbl}_new not exported"))
  }
}
```

## Upload to GCS

```{r}
#| label: resume_for_upload
#| eval: true

# always re-enable eval for upload and cleanup
knitr::opts_chunk$set(eval = TRUE)
```

```{r}
#| label: upload_gcs

# skip sync if all parquet tables were unchanged (no re-write)
if (parquet_complete) {
  message("Parquet unchanged — skipping GCS sync")
} else {
  sync_to_gcs(
    local_dir    = dir_parquet,
    gcs_prefix   = glue("ingest/{dir_label}"),
    bucket       = "calcofi-db",
    delete_stale = TRUE
  )
}
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)

# remove checkpoint after successful completion (fresh next run)
if (file_exists(db_checkpoint)) {
  file_delete(db_checkpoint)
  message(glue("Removed checkpoint: {db_checkpoint}"))
}
```

## Appendix: Interactive Gantt Chart {.appendix}

```{r}
#| label: gantt_chart

# reconnect for gantt chart
con_pq <- get_duckdb_con()
load_duckdb_extension(con_pq, "spatial")

# read ctd_cast from parquet (exclude GEOMETRY columns for R driver compat)
dbExecute(con_pq, glue("
  CREATE OR REPLACE VIEW _ctd_probe AS
  SELECT * FROM read_parquet('{dir_parquet}/ctd_cast.parquet') LIMIT 0"))
pq_geom_cols <- dbGetQuery(con_pq,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = '_ctd_probe' AND data_type LIKE 'GEOMETRY%'")$column_name
pq_all_cols <- dbGetQuery(con_pq,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = '_ctd_probe'
   ORDER BY ordinal_position")$column_name
pq_safe_cols <- setdiff(pq_all_cols, pq_geom_cols)
dbExecute(con_pq, "DROP VIEW IF EXISTS _ctd_probe")
dbExecute(con_pq, glue("
  CREATE VIEW ctd_cast AS
  SELECT {paste(pq_safe_cols, collapse = ', ')}
  FROM read_parquet('{dir_parquet}/ctd_cast.parquet')"))

cruise_spans <- tbl(con_pq, "ctd_cast") |>
  group_by(cruise_key, data_stage) |>
  summarize(
    beg_date = min(datetime_utc, na.rm = TRUE) |> as.Date(),
    end_date = max(datetime_utc, na.rm = TRUE) |> as.Date(),
    .groups = "drop"
  ) |>
  collect() |>
  filter(!is.na(beg_date), !is.na(end_date))

# split multi-year cruises into separate rows
cruise_spans <- cruise_spans |>
  rowwise() |>
  mutate(
    years_spanned = list(year(beg_date):year(end_date))
  ) |>
  unnest(years_spanned) |>
  mutate(
    year = years_spanned,
    adj_beg_date = if_else(
      years_spanned == year(beg_date),
      beg_date,
      as.Date(paste0(years_spanned, "-01-01"))
    ),
    adj_end_date = if_else(
      years_spanned == year(end_date),
      end_date,
      as.Date(paste0(years_spanned, "-12-31"))
    ),
    begin_jday = yday(adj_beg_date),
    end_jday = yday(adj_end_date),
    stage_label = data_stage,
    hover_text = glue(
      "Cruise: {cruise_key}<br>",
      "Stage: {data_stage}<br>",
      "Begin: {format(beg_date, '%Y-%m-%d')}<br>",
      "End: {format(end_date, '%Y-%m-%d')}<br>",
      "Duration: {as.numeric(difftime(end_date, beg_date, units = 'days'))} days"
    )
  ) |>
  ungroup() |>
  arrange(adj_beg_date)

colors <- c("final" = "#23d355ff", "preliminary" = "#A23B72")

p <- plot_ly()

for (i in 1:nrow(cruise_spans)) {
  row <- cruise_spans[i, ]
  p <- p |>
    add_trace(
      type = "scatter",
      mode = "lines",
      x = c(row$begin_jday, row$end_jday),
      y = c(row$year, row$year),
      line = list(
        color = colors[row$stage_label],
        width = 8
      ),
      text = row$hover_text,
      hoverinfo = "text",
      showlegend = FALSE,
      legendgroup = row$stage_label
    )
}

p <- p |>
  add_trace(
    type = "scatter",
    mode = "lines",
    x = c(NA, NA),
    y = c(NA, NA),
    line = list(color = colors["final"], width = 8),
    name = "Final",
    showlegend = TRUE
  ) |>
  add_trace(
    type = "scatter",
    mode = "lines",
    x = c(NA, NA),
    y = c(NA, NA),
    line = list(color = colors["preliminary"], width = 8),
    name = "Preliminary",
    showlegend = TRUE
  )

p <- p |>
  layout(
    title = "CalCOFI CTD Cruise Timeline",
    xaxis = list(
      title = "Day of Year",
      range = c(1, 365),
      dtick = 30,
      tickangle = 0
    ),
    yaxis = list(
      title = "Year",
      autorange = "reversed",
      dtick = 1
    ),
    hovermode = "closest",
    plot_bgcolor = "#f8f9fa",
    paper_bgcolor = "white",
    legend = list(
      x = 1.02,
      y = 1,
      xanchor = "left",
      yanchor = "top"
    )
  )

p

close_duckdb(con_pq)
```


::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
