---
title: "Ingest CCE-LTER Euphausiids"
calcofi:
  target_name: ingest_cce_lter_euphausiids
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/cce-lter_euphausiids/manifest.json
  provider: cce-lter
  dataset: euphausiids
  workflow_url: https://calcofi.io/workflows/ingest_cce-lter_euphausiids.html
  questions_file: metadata/cce-lter/euphausiids/questions.csv
  dataset_meta:
    dataset_name: CCE-LTER Euphausiid Abundance
    description: >
      Euphausiid (krill) abundance from CalCOFI / CCE-LTER net tows,
      1951-2019, one row per tow. Provisional pending provider answers on
      abundance units and taxonomic scope (see Questions for Data Providers).
    citation_main: ""
    link_calcofi_org: ""
    link_data_source: "https://portal.edirepository.org/ (EDI package TBD)"
    link_others: []
    coverage_temporal: 1951-01 to 2019-04
    coverage_spatial: "CalCOFI grid incl. Baja lines"
    license: ""
    pi_names: Rasmus Swalethorp; Linsey Sala
  tables_owned:
    - {table: euphausiids_tow}
    - {table: euphausiids_measurement}
    - {table: euphausiids_summary}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#d2f0bb"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: CCE-LTER euphausiid abundance time series (`data.csv`, ~1.2 MB,
10,150 tows, 1951–2019), curated by CCE-LTER (Rasmus Swalethorp, Linsey Sala)
and published on EDI.

- **Provider**: `cce-lter`
- **Grain**: one row per net tow, with a single `Abundance` measurement
- **Strategy**: load the source CSV into a tidy `euphausiids_tow` position
  table; resolve `ship_key`, `cruise_key`, `site_key`, and `grid_key` against
  the shared reference tables; pivot `Abundance` into long-format
  `euphausiids_measurement`; summarize replicates into `euphausiids_summary`.

This ingest **proceeds with documented best-guess decisions** for the open
provider questions (units, taxonomic scope, coordinate/cruise/ship fixes); each
decision is flagged inline and tracked, ranked by importance, in the
*Questions for Data Providers* section at the end.

```{mermaid}
graph LR
  A[data.csv<br/>10,150 tows] --> B[euphausiids_tow<br/>position + keys]
  B --> C[euphausiids_measurement<br/>long format]
  C --> D[euphausiids_summary<br/>avg/stddev]
  B -.ship/cruise/grid.-> E[(shared refs)]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# common ingest settings (overwrite, dir_data)
source(here("libs/ingest.R"))

# provider/dataset/metadata from this file's authoritative YAML block
cc           <- read_calcofi_meta(here("ingest_cce-lter_euphausiids.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")

dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path))                  file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal")))  file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))   dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

# load unified measurement_type reference
meas_type_csv <- here("metadata/measurement_type.csv")
d_meas_type   <- read_csv(meas_type_csv)
```

## Read Source Data

The source CSV currently lives at `{dir_data}/euphausiids/data.csv` (not under
a `cce-lter/` subdirectory), so it is read directly rather than via
`read_csv_files()`. Source columns are renamed to canonical names per
`metadata/cce-lter/euphausiids/flds_redefine.csv`.

```{r}
#| label: read-source

euph_csv <- path_expand(glue("{dir_data}/euphausiids/data.csv"))
stopifnot("euphausiids data.csv not found" = file_exists(euph_csv))

# archive source to GCS for provenance
sync_to_gcs(
  local_dir  = path_dir(euph_csv),
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket     = "calcofi-files-public",
  exclude    = c(".DS_Store", "*.tmp", "*.gdoc"))  # .gdoc is a Drive stub

d_raw <- read_csv(euph_csv)
cat(glue("Read {format(nrow(d_raw), big.mark=',')} rows, ",
         "{ncol(d_raw)} columns from {basename(euph_csv)}"), "\n")
```

## Clean, Type-Cast, and Correct

Apply the documented data-quality decisions (each tied to a ranked provider
question). Canonical field names follow `metadata/field_dictionary.csv`.

```{r}
#| label: clean-data

d_clean <- d_raw |>
  transmute(
    tow_id             = as.integer(RowNumber),
    cruise_orig        = Cruise,
    ship_name          = Ship,
    date               = as.Date(Date),
    line               = suppressWarnings(as.numeric(Line)),
    station            = suppressWarnings(as.numeric(Station)),
    region             = Region,
    datetime_start_utc = as_datetime(TowBegin),     # Q07: assumed already UTC
    datetime_end_utc   = as_datetime(TowEnd),
    latitude           = as.numeric(Latitude),
    longitude          = as.numeric(Longitude),
    abundance          = as.numeric(Abundance)) |>
  mutate(
    # Q05: ~15 tows have positive longitudes (sign errors) -> flip to negative
    longitude = if_else(longitude > 0, -longitude, longitude),
    # Q05: one garbage coordinate (lat 87.25, far outside survey) -> drop coords
    bad_coord = latitude > 51 | latitude < 20 | longitude < -135 | longitude > -105,
    latitude  = if_else(bad_coord, NA_real_, latitude),
    longitude = if_else(bad_coord, NA_real_, longitude),
    # Q06: one TowEnd in year 2371 (typo) -> null out-of-range tow-end times
    datetime_end_utc = if_else(
      year(datetime_end_utc) > 2025 | year(datetime_end_utc) < 1949,
      as_datetime(NA), datetime_end_utc),
    # canonical CalCOFI site_key "LLL.L SSS.S" from line + station
    site_key = if_else(
      is.na(line) | is.na(station), NA_character_,
      sprintf("%05.1f %05.1f", line, station)))

cat(glue("Cleaned {nrow(d_clean)} tows; ",
         "{sum(d_clean$bad_coord, na.rm=TRUE)} coordinate(s) nulled (Q05), ",
         "{sum(is.na(d_clean$datetime_end_utc))} tow-end null (incl. Q06)"), "\n")
```

## Resolve Ship and Cruise Keys

Load shared reference tables, normalize the messy vessel names (Q04), resolve
`ship_key`/`ship_nodc`, then derive `cruise_key` as `{YYYY}-{MM}-{ship_nodc}`
(the CalCOFI natural-key form) and validate against the `cruise` table (Q03).

```{r}
#| label: resolve-keys

# shared reference tables from the ichthyo ingest
load_prior_tables(
  con,
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables      = c("ship", "cruise", "grid"),
  geom_tables = c("grid"),
  as_view     = TRUE)

d_ship <- dbGetQuery(con, "SELECT ship_key, ship_name, ship_nodc FROM ship") |>
  mutate(ship_name_norm = ship_name |> str_to_upper() |> str_squish())

# normalize euphausiids vessel names (strip 'R/V ', collapse spaces, upper-case)
d_keys <- d_clean |>
  mutate(
    ship_name_norm = ship_name |>
      str_replace("^R/?V\\.?\\s+", "") |> str_to_upper() |> str_squish()) |>
  left_join(
    d_ship |> select(ship_name_norm, ship_key, ship_nodc),
    by = "ship_name_norm") |>
  mutate(
    # cruise_key = YYYY-MM-{ship_nodc} (NA when ship unresolved)
    cruise_key = if_else(
      is.na(ship_nodc), NA_character_,
      glue("{format(date, '%Y-%m')}-{ship_nodc}") |> as.character()))

n_ship  <- sum(!is.na(d_keys$ship_key))
# validate derived cruise_key exists in the cruise table
valid_ck <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM cruise")$cruise_key
d_keys <- d_keys |>
  mutate(cruise_key = if_else(cruise_key %in% valid_ck, cruise_key, NA_character_))
n_cruise <- sum(!is.na(d_keys$cruise_key))

cat(glue(
  "Ship match: {n_ship}/{nrow(d_keys)} ({round(100*n_ship/nrow(d_keys),1)}%); ",
  "cruise_key match: {n_cruise}/{nrow(d_keys)} ",
  "({round(100*n_cruise/nrow(d_keys),1)}%)"), "\n")

# report unresolved ship names for the provider follow-up (Q04)
d_keys |> filter(is.na(ship_key)) |> count(ship_name, sort = TRUE) |>
  datatable(caption = "Unresolved vessel names (Q04)")
```

## Load Tidy Tow Table

`euphausiids_tow` carries position, time, and FK columns only — the
`abundance` measurement is pivoted out in the next step.

```{r}
#| label: load-tow

d_tow <- d_keys |>
  transmute(
    tow_id, cruise_key, ship_key, ship_name, cruise_orig,
    site_key, line, station, region,
    datetime_start_utc, datetime_end_utc,
    latitude, longitude, abundance)

dbWriteTable(con, "euphausiids_tow_wide", d_tow, overwrite = TRUE)

# tidy tow table: drop the raw measurement column
dbExecute(con,
  "CREATE OR REPLACE TABLE euphausiids_tow AS
   SELECT tow_id, cruise_key, ship_key, ship_name, cruise_orig,
          site_key, line, station, region,
          datetime_start_utc, datetime_end_utc, latitude, longitude
   FROM euphausiids_tow_wide")

cat(glue("euphausiids_tow: {dbGetQuery(con,'SELECT COUNT(*) FROM euphausiids_tow')[[1]]} rows"), "\n")
```

## Add Spatial

```{r}
#| label: spatial

add_point_geom(con, "euphausiids_tow", lon_col = "longitude", lat_col = "latitude")
grid_stats <- assign_grid_key(con, "euphausiids_tow")
grid_stats |> datatable(caption = "Grid assignment")
```

## Pivot Measurement to Long Format

A single measurement type, `euphausiid_abundance`. **Q01 (units) and Q02
(taxonomic scope) are unresolved** — registered provisionally and flagged.

```{r}
#| label: pivot-measurement

dbExecute(con,
  "CREATE OR REPLACE TABLE euphausiids_measurement AS
   SELECT ROW_NUMBER() OVER (ORDER BY tow_id) AS euphausiids_measurement_id,
          tow_id,
          'euphausiid_abundance' AS measurement_type,
          CAST(abundance AS DOUBLE) AS measurement_value,
          NULL::VARCHAR AS measurement_qual
   FROM euphausiids_tow_wide
   WHERE abundance IS NOT NULL
     AND NOT isnan(CAST(abundance AS DOUBLE))
     AND isfinite(CAST(abundance AS DOUBLE))")

dbExecute(con, "DROP TABLE euphausiids_tow_wide")

n_meas <- dbGetQuery(con, "SELECT COUNT(*) FROM euphausiids_measurement")[[1]]
cat(glue("euphausiids_measurement: {format(n_meas, big.mark=',')} rows"), "\n")
```

## Summarize Replicate Measurements

Aggregate replicate tows at each unique position (site + datetime + type) into
mean and standard deviation, following the `dic_summary`/`ctd_summary` pattern.

```{r}
#| label: measurement-summary

dbExecute(con,
  "CREATE OR REPLACE TABLE euphausiids_summary AS
   SELECT
     t.site_key, t.datetime_start_utc, t.latitude, t.longitude,
     m.measurement_type,
     AVG(m.measurement_value) AS avg,
     CASE WHEN COUNT(*) = 1 THEN 0
          ELSE COALESCE(STDDEV_SAMP(m.measurement_value), 0) END AS stddev,
     COUNT(*) AS n_obs
   FROM euphausiids_measurement m
   JOIN euphausiids_tow t USING (tow_id)
   WHERE NOT isnan(m.measurement_value) AND isfinite(m.measurement_value)
   GROUP BY t.site_key, t.datetime_start_utc, t.latitude, t.longitude,
            m.measurement_type")

cat(glue("euphausiids_summary: {dbGetQuery(con,'SELECT COUNT(*) FROM euphausiids_summary')[[1]]} positions"), "\n")
```

## Add Measurement Type

```{r}
#| label: add-measurement-type

euph_types <- tibble(
  measurement_type   = "euphausiid_abundance",
  description        = paste(
    "Euphausiid (krill) abundance per net tow. PROVISIONAL: units and",
    "taxonomic scope unconfirmed (see questions Q01, Q02)."),
  units              = "count/1000m3",   # Q01: PROVISIONAL best guess
  `_source_column`   = "Abundance",
  `_source_table`    = "euphausiids_measurement",
  `_source_datasets` = "cce-lter_euphausiids",
  `_qual_column`     = NA_character_,
  `_prec_column`     = NA_character_)

if (!"euphausiid_abundance" %in% d_meas_type$measurement_type) {
  d_meas_type <- bind_rows(d_meas_type, euph_types)
  write_csv(d_meas_type, meas_type_csv)
  cat("Added measurement type: euphausiid_abundance (provisional units)\n")
} else {
  cat("euphausiid_abundance already registered\n")
}
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Schema Documentation

```{r}
#| label: schema

euph_rels <- list(
  primary_keys = list(
    euphausiids_tow         = "tow_id",
    euphausiids_measurement = "euphausiids_measurement_id",
    measurement_type        = "measurement_type"),
  foreign_keys = list(
    list(table = "euphausiids_measurement", column = "tow_id",
         ref_table = "euphausiids_tow", ref_column = "tow_id"),
    list(table = "euphausiids_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

euph_tables <- c(
  "euphausiids_tow", "euphausiids_measurement", "euphausiids_summary",
  "measurement_type", "dataset")
cc_erd(
  con, tables = euph_tables, rels = euph_rels,
  colors = list(
    lightblue   = c("euphausiids_tow", "euphausiids_measurement", "euphausiids_summary"),
    lightyellow = "measurement_type",
    white       = "dataset"))

build_relationships_json(
  rels = euph_rels, output_dir = dir_parquet,
  provider = provider, dataset = dataset)
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
if (length(results$warnings) > 0)
  cat("Warnings:\n", paste("-", results$warnings, collapse = "\n"), "\n")

# NOTE: NULLs in the cross-dataset FK keys (cruise_key, ship_key, site_key,
# grid_key) are the EXPECTED unmatched remainder of partial matching — the same
# accepted behavior as calcofi_dic's nullable cast_id/bottle_id (issue #47).
# They are tracked in questions Q03 (cruise) and Q04 (ship), not a hard failure.
cat(glue(
  "\nMatch coverage (non-NULL): ",
  "ship_key {round(100*dbGetQuery(con,\"SELECT AVG(CASE WHEN ship_key IS NOT NULL THEN 1 ELSE 0 END) FROM euphausiids_tow\")[[1]],1)}%, ",
  "cruise_key {round(100*dbGetQuery(con,\"SELECT AVG(CASE WHEN cruise_key IS NOT NULL THEN 1 ELSE 0 END) FROM euphausiids_tow\")[[1]],1)}%, ",
  "grid_key {round(100*dbGetQuery(con,\"SELECT AVG(CASE WHEN grid_key IS NOT NULL THEN 1 ELSE 0 END) FROM euphausiids_tow\")[[1]],1)}%"), "\n")

# composite-key duplicate check
n_dup <- dbGetQuery(con,
  "SELECT COUNT(*) FROM (
     SELECT tow_id, COUNT(*) n FROM euphausiids_tow GROUP BY tow_id HAVING COUNT(*)>1)")[[1]]
cat(glue("euphausiids_tow tow_id duplicates: {n_dup}"), "\n")
```

## Data Preview

```{r}
#| label: preview-tow

tow_cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='euphausiids_tow' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(tow_cols, collapse=', ')} FROM euphausiids_tow LIMIT 100")) |>
  datatable(caption = "euphausiids_tow — first 100 rows", rownames = FALSE, filter = "top")
```

```{r}
#| label: preview-measurement

dbGetQuery(con, "SELECT * FROM euphausiids_measurement LIMIT 100") |>
  datatable(caption = "euphausiids_measurement — first 100 rows", rownames = FALSE)
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

dir_create(dir_parquet)

mismatches <- list(
  measurement_types = collect_measurement_type_mismatches(
    con, here("metadata/measurement_type.csv")),
  cruise_keys       = collect_cruise_key_mismatches(con, "euphausiids_tow"))

parquet_stats <- write_parquet_outputs(
  con        = con,
  output_dir = dir_parquet,
  tables     = c("euphausiids_tow", "euphausiids_measurement",
                 "euphausiids_summary", "measurement_type", "dataset"),
  sort_by    = list(euphausiids_summary = "measurement_type"),
  strip_provenance = FALSE,
  mismatches = mismatches)

parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata

d_tbls_rd <- read_csv(here("metadata/cce-lter/euphausiids/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/cce-lter/euphausiids/flds_redefine.csv"))

metadata_path <- build_metadata_json(
  con        = con,
  d_tbls_rd  = d_tbls_rd,
  d_flds_rd  = d_flds_rd,
  output_dir = dir_parquet,
  tables     = c("euphausiids_tow", "euphausiids_measurement", "euphausiids_summary"),
  set_comments = TRUE,
  provider   = provider,
  dataset    = dataset,
  workflow_url = cc$workflow_url,
  tables_owned = tables_owned)
```

## Upload to GCS

```{r}
#| label: upload-gcs

sync_to_gcs(
  local_dir  = dir_parquet,
  gcs_prefix = glue("ingest/{dir_label}"),
  bucket     = "calcofi-db")
```

## Questions for Data Providers

Follow-up questions for CCE-LTER (Rasmus Swalethorp, Linsey Sala), **ranked by
importance**. This ingest proceeded with the documented provisional decisions
noted above; `blocker`/`high` answers may change registered units, taxonomy, or
matched rows on the next run. Tracked in
`metadata/cce-lter/euphausiids/questions.csv`; aggregated per provider by
`questions_email.qmd`.

```{r}
#| label: provider-questions

q_path <- here(glue("metadata/{provider}/{dataset}/questions.csv"))
read_csv(q_path) |>
  arrange(factor(priority, c("blocker", "high", "normal")), id) |>
  select(priority, question, context, related_field, status) |>
  datatable(
    caption = "Questions for data providers (ranked by importance)",
    options = list(dom = "t", pageLength = 20), rownames = FALSE)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```

::: {.callout-caution collapse="true"}
## Session Info
```{r session_info}
devtools::session_info()
```
:::
