---
title: "Ingest Bottle Database"
calcofi:
  target_name: ingest_calcofi_bottle
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/calcofi_bottle/manifest.json
  modifies:
    - ship
  provider: calcofi
  dataset: bottle
  workflow_url: https://calcofi.io/workflows/ingest_calcofi_bottle.html
  questions_file: metadata/calcofi/bottle/questions.csv
  dataset_meta:
    dataset_name: CalCOFI Bottle Database
    # display trio, read by the release `dataset` table and the
    # consumer apps (calcofi4db >= 3.15.0) — see NEWS for why these
    # left the apps' own hardcoded maps
    dataset_name_short: Hydrographic Bottle
    category: Physical Oceanography
    color: "#4dabf7"
    description: >
      Hydrographic bottle data including temperature, salinity, oxygen,
      nutrients, pigments, and C14 from CalCOFI cruises.
    citation_main: "CalCOFI. (2023). CalCOFI Bottle Database 194903-202105. CalCOFI.org."
    link_calcofi_org: https://calcofi.org/data/oceanographic-data/bottle-database/
    link_data_source: https://calcofi.org/downloads/database/CalCOFI_Database_194903-202105_csv_16October2023.zip
    # calcofi.org states no license anywhere on this page (checked 2026-09-03);
    # left empty, see metadata/calcofi/bottle/questions.csv Q10 (proposed
    # CC-BY-4.0 to Rasmus Swalethorp)
    license: ""
    # no PI stated by the source; left empty, see questions.csv Q11
    # (proposed Rasmus Swalethorp)
    pi_names: ""
  # publishes the consolidated core: the cast/bottle/measurement/condition source
  # shape is wrangled in the notebook and projected into sample / obs /
  # sample_measurement, surviving as compat VIEWs over the core.
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (cast -> bottle adjacency list)"}
    - {table: obs, shared: true, note: "core observations (env, bottle hydrography)"}
    - {table: sample_measurement, shared: true, note: "cast conditions + bottom depth"}
    - {table: measurement_type, shared: true, note: "shared registry across bottle/ctd/dic"}
  erd:
    color: "#f5cad9"
---

## Overview

**Source**: [Bottle Database | CalCOFI.org](https://calcofi.org/data/oceanographic-data/bottle-database/)

- Download: <https://calcofi.org/downloads/database/CalCOFI_Database_194903-202105_csv_16October2023.zip>

**Goal**: Ingest bottle database from source CSV files into a local
wrangling DuckDB, then export as parquet files for downstream use.

**Two source tables**:

- `194903-202105_Cast.csv` (61 fields) — station-level: cruise,
  location, meteorology, integrated measurements
- `194903-202105_Bottle.csv` (62 fields) — sample-level: depth,
  temperature, salinity, O₂, nutrients, pigments, C14, pH

**Coverage**: 1949-03 to 2021-05 (72 years)

**Join key**: `cast_id` (in both tables; source column `Cst_Cnt`)

This workflow processes CalCOFI bottle database CSV files and outputs
parquet files. The workflow:

1. Reads CSV files from source directory (with GCS archive sync)
2. Loads into local wrangling DuckDB with column renames
3. Applies ship name/code corrections
4. Consolidates temporal columns into `datetime_start_utc`, drops derivable columns, creates `casts_derived` VIEW
5. Pivots bottle measurements into long-format `bottle_measurement` table
6. Pivots cast environmental conditions into `cast_condition` table
7. Loads `measurement_type` reference table
8. Validates primary keys and foreign key integrity
9. Adds spatial geometry and grid assignments
10. Exports to parquet files (for later integration into Working DuckLake)

```{r}
#| label: setup

# devtools::install_local(here::here("../calcofi4db"), force = T)
devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI / calcofi4db,
  CalCOFI / calcofi4r,
  DBI,
  dplyr,
  DT,
  fs,
  glue,
  gargle,
  googledrive,
  here,
  htmltools,
  janitor,
  jsonlite,
  knitr,
  listviewer,
  litedown,
  lubridate,
  mapview,
  purrr,
  readr,
  rlang,
  sf,
  stringr,
  tibble,
  tidyr,
  units,
  webshot2,
  quiet = T
)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# define paths
# common ingest settings (overwrite, dir_data)
source(here("libs/ingest.R"))

# say(): status output to STDOUT so it renders in the notebook.
#
# This notebook had 37 message() calls against 5 cat() — the worst ratio in the
# repo. message() writes to stderr, which Quarto does not reliably surface under
# `code-fold: true`, so most of the diagnostics below were being computed and then
# never shown. The repo convention is cat() (see .claude/skills/ingest-new/SKILL.md),
# but cat() adds no trailing newline, so this thin wrapper keeps message()'s line
# semantics. Matches the same helper in ingest_calcofi_ctd-cast.qmd.
say <- function(...) cat(..., "\n", sep = "")

# provider/dataset/metadata read from this file's authoritative YAML block
cc           <- read_calcofi_meta(here("ingest_calcofi_bottle.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label <- glue("{provider}_{dataset}")
subdir <- "CalCOFI_Database_194903-202105_csv_16October2023"
dir_parquet <- here(glue("data/parquet/{dir_label}"))
dir_stage <- cc_stage_path("parquet", dir_label, create = TRUE)
db_path <- here(glue("data/wrangling/{dir_label}.duckdb"))
# load data using calcofi4db package
# - reads from local Google Drive mount
# - syncs to GCS archive if files changed (creates new timestamped archive)
# - tracks GCS archive path for provenance
dir_dataset <- glue("{dir_data}/{provider}/{dataset}")
# unified measurement_type reference (needed for pivot mappings)
meas_type_csv <- here("metadata/measurement_type.csv")
```

```{r}
#| label: load_db

if (overwrite) {
  if (file_exists(db_path)) {
    file_delete(db_path)
  }
  # clear stale WAL/tmp from an interrupted run (avoids "WAL checkpoint
  # iteration does not match" errors when reopening)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))  dir_delete(paste0(db_path, ".tmp"))
  # keep dir_parquet so write_parquet_outputs can content-hash dedup against
  # the prior run (only changed partitions re-written/uploaded)
}

dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

```{r}
#| label: read_csv_files

d_meas_type <- read_measurement_type(meas_type_csv)

d <- read_csv_files(
  provider = provider,
  dataset = dataset,
  subdir = subdir,
  dir_data = dir_data,
  sync_archive = TRUE,
  metadata_dir = here("metadata"),
  field_descriptions = list(
    "194903-202105_Cast" = glue(
      "{dir_dataset}/Cast Field Descriptions.csv"
    ),
    "194903-202105_Bottle" = glue(
      "{dir_dataset}/Bottle Field Descriptions - UTF-8.csv"
    )
  )
)

# show source files summary
say(glue("Loaded {nrow(d$source_files)} tables from {d$paths$dir_csv}"))
say(glue("Total rows: {sum(d$source_files$nrow)}"))
```

## Check for any mismatched tables and fields

```{r}
#| label: data_integrity_check

# check data integrity
# type_exceptions = "all": type mismatches (34) are expected since readr
# infers types differently (e.g. quality codes as numeric, Time as hms);
# these are resolved during ingestion via flds_redefine.csv type_new
integrity <- check_data_integrity(
  d = d,
  dataset_name = dataset_name,
  type_exceptions = "all"
)
```

```{r}
#| label: data_integrity_message
#| output: asis
#| echo: false

# render the pass/fail message as markdown
render_integrity_message(integrity)
```

## Show Source Files

```{r}
#| label: source-files
show_source_files(d)
```

## Show CSV Tables and Fields to Ingest

```{r}
#| label: tbls_in

d$d_csv$tables |>
  datatable(caption = "Tables to ingest.")
```

```{r}
#| label: flds_in

d$d_csv$fields |>
  datatable(caption = "Fields to ingest.")
```

## Show tables and fields redefined

```{r}
#| label: tbls_rd

show_tables_redefine(d)
```

```{r}
#| label: flds_rd

show_fields_redefine(d)
```

## Load Tables into Database

```{r}
#| label: load_tbls_to_db

# use ingest_dataset() which handles:
# - transform_data() for applying redefinitions
# - provenance tracking via gcs_path from read_csv_files()
# - ingest_to_working() for each table
tbl_stats <- ingest_dataset(
  con = con,
  d = d,
  mode = if (overwrite) "replace" else "append",
  verbose = TRUE
)

tbl_stats |>
  datatable(rownames = FALSE, filter = "top")
```

## Apply Ship Corrections

Apply known ship name/code corrections from `ship_renames.csv`.

```{r}
#| label: apply_ship_corrections

ship_renames_csv <- here("metadata/ship_renames.csv")
d_ship_renames <- read_csv(ship_renames_csv)

# apply each correction
n_updated_total <- 0
for (i in seq_len(nrow(d_ship_renames))) {
  old_name <- d_ship_renames$csv_name[i]
  old_code <- d_ship_renames$csv_code[i]
  new_name <- d_ship_renames$csv_name_new[i]
  new_code <- d_ship_renames$csv_code_new[i]

  # update where both name and code match old values
  n_updated <- dbExecute(
    con,
    glue_sql(
      "UPDATE casts
     SET ship_name = {new_name}, ship_code = {new_code}
     WHERE ship_name = {old_name} AND ship_code = {old_code}",
      .con = con
    )
  )

  if (n_updated > 0) {
    say(glue(
      "Updated {n_updated} casts: ",
      "'{old_name}' [{old_code}] -> '{new_name}' [{new_code}]"
    ))
  }
  n_updated_total <- n_updated_total + n_updated
}

say(glue("\nTotal casts updated: {n_updated_total}"))
```

## Column Naming Rationale

The column renames in `flds_redefine.csv` follow these db conventions:

- **`*_id`** for integer primary/foreign keys (e.g. `Cst_Cnt` → `cast_id`, `Btl_Cnt` → `bottle_id`)
- **`*_key`** for varchar natural/composite keys (e.g. `Cast_ID` → `cast_key`, `Sta_ID` → `site_key`)
- Note: `Cruise_ID` → `cruise_key_0` (interim `_0` suffix avoids collision with SWFSC `cruise_key` YYYY-MM-NODC; discarded after merge)
- **unit suffixes** for measurements (e.g. `Depthm` → `depth_m`)

### Ambiguous column pairs

| Column pair | Why different |
|------------|---------------|
| `Cst_Cnt` (→ `cast_id`) vs `Cast_ID` (→ `cast_key`) | integer counter (1, 2, 3...) vs varchar composite (19-4903CR-HY-060-0930-05400560) |
| `Sta_ID` (→ `site_key`) vs `DbSta_ID` (→ `db_sta_key`) | "054.0 056.0" (space-separated) vs "05400560" (compact, derivable) |
| `Depth_ID` (→ `depth_key`) vs `Depthm` (→ `depth_m`) | varchar identifier vs numeric depth in meters |

### `R_*` = "Reported" (pre-QC) values

The `R_*` prefix means original instrument values before quality control.
**Critical**: `R_Sal` (Specific Volume Anomaly, values ~233, units 10⁻⁸ m³/kg)
is a **completely different parameter/scale** than `Salnty` (Practical Salinity
PSS-78, values ~33.44). These are pivoted as distinct measurement types
(`r_salinity_sva` vs `salinity`).

## Consolidate Casts Columns

Combine `date` + `time` into `datetime_start_utc`, drop 16 derivable columns,
and create a `casts_derived` convenience VIEW.

```{r}
#| label: consolidate_casts

# -- create datetime_start_utc from date + time (idempotent) ----
casts_cols <- dbListFields(con, "casts")

if ("date" %in% casts_cols && "time" %in% casts_cols) {
  # date is DATE type (e.g. 1949-03-01), time is INTERVAL (e.g. 09:30:00)
  # adding DATE + INTERVAL yields TIMESTAMP; 324 casts have NULL time
  dbExecute(con, "ALTER TABLE casts ADD COLUMN IF NOT EXISTS datetime_start_utc TIMESTAMP")
  dbExecute(
    con,
    "
    UPDATE casts SET datetime_start_utc =
      CASE
        WHEN \"time\" IS NOT NULL
        THEN CAST(\"date\" AS TIMESTAMP) + \"time\"
        ELSE CAST(\"date\" AS TIMESTAMP)
      END"
  )
  say(glue(
    "Created datetime_start_utc: {dbGetQuery(con, 'SELECT COUNT(*) FROM casts WHERE datetime_start_utc IS NOT NULL')[[1]]} non-null values"
  ))

  # -- drop derivable columns ----
  cols_to_drop <- c(
    "year", "month", "quarter", "julian_day", "julian_date",
    "time_zone", "lat_deg", "lat_min", "lat_hem",
    "lon_deg", "lon_min", "lon_hem",
    # NOT `cruise`: the source's YYYYMM designation is the cruise's identity and is
    # not derivable from a cast's date — 5508BD ran 7 Aug - 25 Sep 1955, and
    # STRFTIME(datetime, '%Y%m') put its September casts on cruise 5509BD
    "cruz_sta", "db_sta_key", "cruz_num",
    "date", "time"
  )

  n_before <- length(casts_cols)
  for (col in cols_to_drop) {
    tryCatch(
      dbExecute(con, glue('ALTER TABLE casts DROP COLUMN "{col}"')),
      error = function(e) NULL)
  }
  n_after <- length(dbListFields(con, "casts"))
  say(glue("casts columns: {n_before} -> {n_after} (dropped {n_before - n_after})"))
} else {
  say("casts already consolidated (datetime_start_utc exists, date/time dropped)")
}
```

## Pivot Bottle Measurements

Pivot wide-format measurement columns from `bottle` into long-format
`bottle_measurement` table. Each measurement type has a value, optional
precision, and optional quality code.

```{mermaid}
%%| label: pivot_bottle_meas_diagram
%%| fig-cap: "Pivot wide bottle columns into long-format bottle_measurement table."
erDiagram
    bottle_BEFORE {
        int bottle_id PK
        int cast_id FK
        dbl depth_m
        dbl t_deg_c
        dbl t_prec
        str t_qual
        dbl salnty
        dbl s_prec
        str s_qual
        dbl o2ml_l
        str o_qual
        dbl chlor_a
        str chlqua
        dbl _etc "... 30 measurement columns"
    }
    bottle_AFTER {
        int bottle_id PK
        int cast_id FK
        dbl depth_m
        str depth_qual
    }
    bottle_measurement {
        int bottle_measurement_id PK
        int bottle_id FK
        str measurement_type
        dbl measurement_value
        dbl measurement_prec
        str measurement_qual
    }
    bottle_BEFORE ||--|{ bottle_measurement : "pivot into"
    bottle_AFTER ||--o{ bottle_measurement : "bottle_id"
```

```{r}
#| label: pivot_bottle_measurements

# build measurement mapping from unified CSV
bottle_meas_map <- d_meas_type |>
  filter(
    str_detect(`_source_datasets`, "calcofi_bottle"),
    `_source_table` == "bottle"
  ) |>
  select(
    measurement_type,
    value_col = `_source_column`,
    prec_col = `_prec_column`,
    qual_col = `_qual_column`
  ) |>
  mutate(across(c(prec_col, qual_col), ~ na_if(.x, "")))

# build UNION ALL SQL from mapping
sql_parts <- purrr::pmap_chr(
  bottle_meas_map,
  function(
    measurement_type,
    value_col,
    prec_col,
    qual_col
  ) {
    prec_expr <- if (is.na(prec_col)) "NULL" else prec_col
    qual_expr <- if (is.na(qual_col)) "NULL" else qual_col
    glue(
      "SELECT bottle_id, '{measurement_type}' AS measurement_type,
       CAST({value_col} AS DOUBLE) AS measurement_value,
       CAST({prec_expr} AS DOUBLE) AS measurement_prec,
       regexp_replace(CAST({qual_expr} AS VARCHAR), '\\.0+$', '') AS measurement_qual
     FROM bottle WHERE {value_col} IS NOT NULL"
    )
  }
)

sql_create <- glue(
  "CREATE OR REPLACE TABLE bottle_measurement AS
   SELECT ROW_NUMBER() OVER (ORDER BY bottle_id, measurement_type)
     AS bottle_measurement_id, *
   FROM (
     {paste(sql_parts, collapse = '\nUNION ALL\n')}
   ) sub"
)
dbExecute(con, sql_create)

n_meas <- dbGetQuery(con, "SELECT COUNT(*) FROM bottle_measurement")[[1]]
n_types <- dbGetQuery(
  con,
  "SELECT COUNT(DISTINCT measurement_type) FROM bottle_measurement"
)[[1]]
say(glue(
  "bottle_measurement: {format(n_meas, big.mark = ',')} rows, {n_types} measurement types"
))

# -- drop measurement columns from bottle base table ----
cols_to_drop <- unique(c(
  bottle_meas_map$value_col,
  na.omit(bottle_meas_map$prec_col),
  na.omit(bottle_meas_map$qual_col)
))

n_before <- length(dbListFields(con, "bottle"))
for (col in cols_to_drop) {
  tryCatch(
    dbExecute(con, glue('ALTER TABLE bottle DROP COLUMN "{col}"')),
    error = function(e) warning(glue("could not drop {col}: {e$message}"))
  )
}

# rename p_qual to depth_qual (pressure/depth quality code stays in base table)
dbExecute(con, 'ALTER TABLE bottle RENAME COLUMN p_qual TO depth_qual')

n_after <- length(dbListFields(con, "bottle"))
say(glue(
  "bottle columns: {n_before} -> {n_after} (dropped {n_before - n_after})"
))
```

## Pivot Cast Conditions

Pivot meteorological and environmental observation columns from `casts`
into long-format `cast_condition` table.

```{mermaid}
%%| label: pivot_cast_cond_diagram
%%| fig-cap: "Pivot wide cast condition columns into long-format cast_condition table."
erDiagram
    casts_BEFORE {
        int cast_id PK
        str site_key
        dbl latitude
        dbl longitude
        ts  datetime_start_utc
        dbl wave_dir
        dbl wave_ht
        dbl wave_prd
        dbl wind_dir
        dbl wind_spd
        dbl barometer
        dbl dry_t
        dbl wet_t
        dbl wea
        dbl cloud_typ
        dbl cloud_amt
        dbl visibility
        dbl secchi
        dbl forel_u
    }
    casts_AFTER {
        int cast_id PK
        str site_key
        dbl latitude
        dbl longitude
        ts  datetime_start_utc
    }
    cast_condition {
        int cast_condition_id PK
        int cast_id FK
        str condition_type
        dbl condition_value
    }
    casts_BEFORE ||--|{ cast_condition : "pivot into"
    casts_AFTER ||--o{ cast_condition : "cast_id"
```

```{r}
#| label: pivot_cast_conditions

# build cast condition mapping from unified CSV
cast_cond_map <- d_meas_type |>
  filter(
    str_detect(`_source_datasets`, "calcofi_bottle"),
    `_source_table` == "casts"
  ) |>
  select(
    condition_type = measurement_type,
    source_col = `_source_column`
  )

# build UNION ALL SQL
sql_parts <- purrr::pmap_chr(
  cast_cond_map,
  function(condition_type, source_col) {
    glue(
      "SELECT cast_id, '{condition_type}' AS condition_type,
       CAST({source_col} AS DOUBLE) AS condition_value
     FROM casts WHERE {source_col} IS NOT NULL"
    )
  }
)

sql_create <- glue(
  "CREATE OR REPLACE TABLE cast_condition AS
   SELECT ROW_NUMBER() OVER (ORDER BY cast_id, condition_type)
     AS cast_condition_id, *
   FROM (
     {paste(sql_parts, collapse = '\nUNION ALL\n')}
   ) sub"
)
dbExecute(con, sql_create)

n_cond <- dbGetQuery(con, "SELECT COUNT(*) FROM cast_condition")[[1]]
n_types <- dbGetQuery(
  con,
  "SELECT COUNT(DISTINCT condition_type) FROM cast_condition"
)[[1]]
say(glue(
  "cast_condition: {format(n_cond, big.mark = ',')} rows, {n_types} condition types"
))

# -- drop condition columns from casts ----
n_before <- length(dbListFields(con, "casts"))
for (col in cast_cond_map$source_col) {
  tryCatch(
    dbExecute(con, glue('ALTER TABLE casts DROP COLUMN "{col}"')),
    error = function(e) warning(glue("could not drop {col}: {e$message}"))
  )
}
n_after <- length(dbListFields(con, "casts"))
say(glue(
  "casts columns: {n_before} -> {n_after} (dropped {n_before - n_after})"
))
```

## Load Measurement Type Reference

Load the `measurement_type` lookup table that provides units and
descriptions for all measurement/condition types across datasets.

### `measurement_type` vs `lookup`

The `measurement_type` table and the swfsc `lookup` table serve fundamentally different purposes:

- **`measurement_type`** answers "what was measured?" — provides metadata (units, description, source provenance) for measurement codes used in `bottle_measurement`, `cast_condition`, and `ichthyo`
- **`lookup`** answers "what does this code mean?" — translates categorical values (egg developmental stages 1-11, larva stages, tow types) into human-readable descriptions

Merging would create a sparse table with many NULL columns (e.g., `units`/`_source_column` NULL for egg stages; `lookup_num`/`lookup_chr` NULL for temperature). Both tables coexist in the Working DuckLake, loaded from their respective upstream workflows.

```{r}
#| label: load_measurement_type

# derive provider/dataset from _source_datasets for downstream consumers
# (callers want `WHERE provider = 'calcofi'` not `_source_datasets LIKE ...`)
d_meas_type <- d_meas_type |>
  mutate(
    provider = sub("_.*$",    "", `_source_datasets`),
    dataset  = sub("^[^_]+_", "", `_source_datasets`))
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)

# validate FK: all bottle_measurement types exist in reference table
orphan_btl_types <- dbGetQuery(
  con,
  "
  SELECT DISTINCT measurement_type FROM bottle_measurement
  WHERE measurement_type NOT IN (SELECT measurement_type FROM measurement_type)"
)
stopifnot(
  "all bottle_measurement.measurement_type must exist in measurement_type" = nrow(
    orphan_btl_types
  ) ==
    0
)

# validate FK: all cast_condition types exist in reference table
orphan_cond_types <- dbGetQuery(
  con,
  "
  SELECT DISTINCT condition_type FROM cast_condition
  WHERE condition_type NOT IN (SELECT measurement_type FROM measurement_type)"
)
stopifnot(
  "all cast_condition.condition_type must exist in measurement_type" = nrow(
    orphan_cond_types
  ) ==
    0
)

say(glue(
  "measurement_type: {nrow(d_meas_type)} types loaded, FK checks passed"
))

# validate: all measurement_types used by this dataset are registered
bottle_types_used <- dbGetQuery(
  con,
  "SELECT DISTINCT measurement_type FROM bottle_measurement"
)$measurement_type
cast_types_used <- dbGetQuery(
  con,
  "SELECT DISTINCT condition_type FROM cast_condition"
)$condition_type

registered <- d_meas_type |>
  filter(str_detect(`_source_datasets`, "calcofi_bottle")) |>
  pull(measurement_type)

unregistered <- setdiff(c(bottle_types_used, cast_types_used), registered)
stopifnot(
  "all measurement_types used by calcofi_bottle must be registered in _source_datasets" = length(
    unregistered
  ) ==
    0
)
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

# dataset registry built from authoritative ingest_*.qmd YAML (was dataset.csv)
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
say(glue("dataset: {nrow(d_dataset)} datasets registered"))
```

## Questions for Data Providers

Open questions tracked in `metadata/calcofi/bottle/questions.csv`. Both blockers
concern **quality-flag semantics**, and they gate more than this notebook: with
`measurement_qual` an uninterpreted pass-through, no downstream QC rule can act on
it. `S_qual` alone holds 253 distinct values where every sibling column has 3-10.
Two further questions cover ammonium left-censoring, where the flag's usage
changes mid-record and filtering on it fabricates a trend.

```{r}
#| label: provider-questions

# one validated read + render for every ingest: the vocabulary and the column
# order live in calcofi4db, not in 16 hand-written factor() calls
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the CalCOFI bottle data providers (ranked)")
```

## Verify Primary Keys

```{r}
#| label: verify_primary_keys

# verify cast_id uniqueness in casts
cast_keys <- tbl(con, "casts") |> pull(cast_id)
n_dup_cast <- sum(duplicated(cast_keys))
stopifnot("cast_id must be unique in casts" = n_dup_cast == 0)
say(glue("casts.cast_id: {length(cast_keys)} unique values, 0 duplicates"))

# verify bottle_id uniqueness in bottle
btl_keys <- tbl(con, "bottle") |> pull(bottle_id)
n_dup_btl <- sum(duplicated(btl_keys))
stopifnot("bottle_id must be unique in bottle" = n_dup_btl == 0)
say(glue(
  "bottle.bottle_id: {length(btl_keys)} unique values, 0 duplicates"
))

# verify FK: bottle.cast_id -> casts.cast_id
orphan_bottles <- tbl(con, "bottle") |>
  anti_join(
    tbl(con, "casts"),
    by = "cast_id"
  ) |>
  collect()
stopifnot(
  "all bottle.cast_id must exist in casts.cast_id" = nrow(orphan_bottles) == 0
)
say(glue(
  "FK integrity: all {length(btl_keys)} bottle.cast_id ",
  "found in casts.cast_id"
))

# verify bottle_measurement_id uniqueness
n_dup_bm_id <- dbGetQuery(
  con,
  "
  SELECT COUNT(*) FROM (
    SELECT bottle_measurement_id, COUNT(*) AS n
    FROM bottle_measurement
    GROUP BY bottle_measurement_id
    HAVING COUNT(*) > 1)"
)[[1]]
stopifnot("bottle_measurement_id must be unique" = n_dup_bm_id == 0)
say(glue("bottle_measurement.bottle_measurement_id: unique, 0 duplicates"))

# verify composite PK: bottle_measurement (bottle_id, measurement_type)
n_dup_meas <- dbGetQuery(
  con,
  "
  SELECT COUNT(*) FROM (
    SELECT bottle_id, measurement_type, COUNT(*) AS n
    FROM bottle_measurement
    GROUP BY bottle_id, measurement_type
    HAVING COUNT(*) > 1)"
)[[1]]
stopifnot(
  "bottle_measurement (bottle_id, measurement_type) must be unique" = n_dup_meas ==
    0
)
say(glue("bottle_measurement: composite key unique, 0 duplicates"))

# verify FK: bottle_measurement.bottle_id -> bottle.bottle_id
n_orphan_meas <- dbGetQuery(
  con,
  "
  SELECT COUNT(DISTINCT bottle_id) FROM bottle_measurement
  WHERE bottle_id NOT IN (SELECT bottle_id FROM bottle)"
)[[1]]
stopifnot(
  "all bottle_measurement.bottle_id must exist in bottle" = n_orphan_meas == 0
)
say(glue("bottle_measurement FK: all bottle_ids found in bottle"))

# verify cast_condition_id uniqueness
n_dup_cc_id <- dbGetQuery(
  con,
  "
  SELECT COUNT(*) FROM (
    SELECT cast_condition_id, COUNT(*) AS n
    FROM cast_condition
    GROUP BY cast_condition_id
    HAVING COUNT(*) > 1)"
)[[1]]
stopifnot("cast_condition_id must be unique" = n_dup_cc_id == 0)
say(glue("cast_condition.cast_condition_id: unique, 0 duplicates"))

# verify composite PK: cast_condition (cast_id, condition_type)
n_dup_cond <- dbGetQuery(
  con,
  "
  SELECT COUNT(*) FROM (
    SELECT cast_id, condition_type, COUNT(*) AS n
    FROM cast_condition
    GROUP BY cast_id, condition_type
    HAVING COUNT(*) > 1)"
)[[1]]
stopifnot(
  "cast_condition (cast_id, condition_type) must be unique" = n_dup_cond == 0
)
say(glue("cast_condition: composite key unique, 0 duplicates"))

# verify FK: cast_condition.cast_id -> casts.cast_id
n_orphan_cond <- dbGetQuery(
  con,
  "
  SELECT COUNT(DISTINCT cast_id) FROM cast_condition
  WHERE cast_id NOT IN (SELECT cast_id FROM casts)"
)[[1]]
stopifnot("all cast_condition.cast_id must exist in casts" = n_orphan_cond == 0)
say(glue("cast_condition FK: all cast_ids found in casts"))
```

## Schema Documentation

```{r}
#| label: erd_tbls

# exclude views (casts_derived)
db_tables <- function(con) {
  setdiff(dbListTables(con), "casts_derived") |> sort()
}

tbls <- db_tables(con)
cc_erd(con, tables = tbls)
```

### Primary Key Strategy

**Note on UUID-first approach**: The SWFSC ichthyo tables use source UUIDs (minted
at sea) as primary identifiers. The bottle/cast tables are different — they use
source integer counters (`Cst_Cnt`, `Btl_Cnt`) as natural keys, which are stable
identifiers from the source system. The pivoted tables (`bottle_measurement`,
`cast_condition`) have no source identifier and use sequential integer IDs for
convenience, with composite natural keys enforced as unique.

| Table | Primary Key | Type |
|-------|-------------|------|
| `casts` | `cast_id` | Source integer counter (`Cst_Cnt`) — stable source key |
| `bottle` | `bottle_id` | Source integer counter (`Btl_Cnt`) — stable source key |
| `bottle_measurement` | `bottle_measurement_id` | Sequential (derived/pivoted table); composite `(bottle_id, measurement_type)` enforced unique |
| `cast_condition` | `cast_condition_id` | Sequential (derived/pivoted table); composite `(cast_id, condition_type)` enforced unique |
| `measurement_type` | `measurement_type` | Natural key (type code string) |
| `grid` | `grid_key` | Natural key |

### Foreign Key Relationships

```
casts.cast_id (PK, integer) — station-level
    ↓
bottle.bottle_id (PK, integer) — sample-level
bottle.cast_id (FK) → casts.cast_id
    ↓
bottle_measurement.bottle_measurement_id (PK, integer) — ROW_NUMBER(ORDER BY bottle_id, measurement_type)
bottle_measurement.bottle_id (FK) → bottle.bottle_id
bottle_measurement.measurement_type (FK) → measurement_type.measurement_type

cast_condition.cast_condition_id (PK, integer) — ROW_NUMBER(ORDER BY cast_id, condition_type)
cast_condition.cast_id (FK) → casts.cast_id
cast_condition.condition_type (FK) → measurement_type.measurement_type

casts.grid_key (FK) → grid.grid_key (after spatial join)
casts.ship_code (soft FK) → ship.ship_nodc (validated after integration with swfsc)
```

```{r}
#| label: erd_fk

# define PK/FK relationships for visualization and relationships.json
bottle_rels <- list(
  primary_keys = list(
    casts = "cast_id",
    bottle = "bottle_id",
    bottle_measurement = "bottle_measurement_id",
    cast_condition = "cast_condition_id",
    measurement_type = "measurement_type"
  ),
  foreign_keys = list(
    list(
      table = "bottle",
      column = "cast_id",
      ref_table = "casts",
      ref_column = "cast_id"
    ),
    list(
      table = "bottle_measurement",
      column = "bottle_id",
      ref_table = "bottle",
      ref_column = "bottle_id"
    ),
    list(
      table = "bottle_measurement",
      column = "measurement_type",
      ref_table = "measurement_type",
      ref_column = "measurement_type"
    ),
    list(
      table = "cast_condition",
      column = "cast_id",
      ref_table = "casts",
      ref_column = "cast_id"
    ),
    list(
      table = "cast_condition",
      column = "condition_type",
      ref_table = "measurement_type",
      ref_column = "measurement_type"
    )
  )
)

cc_erd(con, tables = db_tables(con), rels = bottle_rels)
```

## Add Spatial

### Add `casts.geom`

```{r}
#| label: mk_cast_pts

# add geometry column using DuckDB spatial
# note: DuckDB spatial doesn't track SRID metadata (unlike PostGIS)
# all geometries assumed WGS84 (EPSG:4326) by convention
add_point_geom(con, "casts", lon_col = "longitude", lat_col = "latitude")
```

### Load `grid` table

Load grid from the ichthyo workflow's local parquet output (single source of
truth — grid is created in `ingest_swfsc_ichthyo.qmd`).

```{r}
#| label: load_grid

load_prior_tables(
  con = con,
  tables = "grid",
  parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  as_view = TRUE
)
```

### Update `casts.grid_key`

```{r}
#| label: update_cast_from_grid

grid_stats <- assign_grid_key(con, "casts")
grid_stats |> datatable()
```

### Create `casts_derived` VIEW

Create convenience VIEW after all casts table modifications are complete
(column drops, pivots, spatial join). The VIEW adds back temporal and
coordinate columns derivable from the base table.

```{r}
#| label: create_casts_derived

dbExecute(con, "DROP VIEW IF EXISTS casts_derived")
dbExecute(
  con,
  "
  CREATE VIEW casts_derived AS
  SELECT *,
    EXTRACT(YEAR FROM datetime_start_utc)::SMALLINT    AS year,
    EXTRACT(MONTH FROM datetime_start_utc)::SMALLINT   AS month,
    EXTRACT(QUARTER FROM datetime_start_utc)::SMALLINT AS quarter,
    EXTRACT(DOY FROM datetime_start_utc)::SMALLINT     AS julian_day,
    (datetime_start_utc::DATE - DATE '1899-12-30')     AS julian_date,
    FLOOR(ABS(latitude))::SMALLINT                AS lat_deg,
    (ABS(latitude) - FLOOR(ABS(latitude))) * 60    AS lat_min,
    CASE WHEN latitude >= 0 THEN 'N' ELSE 'S' END AS lat_hem,
    FLOOR(ABS(longitude))::SMALLINT                AS lon_deg,
    (ABS(longitude) - FLOOR(ABS(longitude))) * 60    AS lon_min,
    CASE WHEN longitude >= 0 THEN 'E' ELSE 'W' END AS lon_hem,
    REPLACE(REPLACE(site_key, '.', ''), ' ', '')   AS db_sta_key
  FROM casts"
)
say("Created casts_derived VIEW with derived temporal/coordinate columns")
```

### Map Cast Locations

```{r}
#| label: map_casts

# slowish, so use cached figure
map_casts_png <- here(glue("figures/{provider}_{dataset}_casts_map.png"))

if (!file_exists(map_casts_png)) {
  # exclude native GEOMETRY column; convert via ST_AsText for sf
  cast_cols <- dbGetQuery(
    con,
    "SELECT column_name FROM information_schema.columns
     WHERE table_name = 'casts' AND data_type != 'GEOMETRY'"
  )$column_name
  cast_sql <- paste(
    "SELECT",
    paste(cast_cols, collapse = ", "),
    ", ST_AsText(geom) AS geom_wkt FROM casts"
  )
  casts_sf <- dbGetQuery(con, cast_sql) |>
    st_as_sf(wkt = "geom_wkt", crs = 4326) |>
    select(-geom_wkt) |>
    mutate(year = year(datetime_start_utc))

  m <- mapView(casts_sf, zcol = "year")
  mapshot2(m, file = map_casts_png)
}

htmltools::img(
  src = map_casts_png |> str_replace(here(), "."),
  width = "600px"
)
```

## Report

```{r}
#| label: show_latest

# cc_erd handles GEOMETRY columns natively (unlike dm_from_con)
cc_erd(
  con,
  tables = db_tables(con),
  rels = bottle_rels,
  colors = list(
    lightyellow = c("bottle_measurement", "cast_condition"),
    lightblue = "measurement_type"
  )
)
```

```{r}
#| label: summary_stats

# summary statistics
n_casts <- dbGetQuery(con, "SELECT COUNT(*) FROM casts")[[1]]
n_bottles <- tbl(con, "bottle") |> tally() |> pull(n)
n_meas <- dbGetQuery(con, "SELECT COUNT(*) FROM bottle_measurement")[[1]]
n_cond <- dbGetQuery(con, "SELECT COUNT(*) FROM cast_condition")[[1]]

year_range <- dbGetQuery(
  con,
  "
  SELECT
    EXTRACT(YEAR FROM MIN(datetime_start_utc))::INTEGER AS yr_min,
    EXTRACT(YEAR FROM MAX(datetime_start_utc))::INTEGER AS yr_max
  FROM casts"
)

fmt <- function(x, ...) format(x, big.mark = ",", ...)
say(glue("Total casts: {fmt(n_casts)}"))
say(glue("Total bottles: {fmt(n_bottles)}"))
say(glue("Total bottle measurements: {fmt(n_meas)}"))
say(glue("Total cast conditions: {fmt(n_cond)}"))
say(glue("Year range: {year_range$yr_min} - {year_range$yr_max}"))
```

## Validate Local Database

Validate data quality in the local wrangling database before exporting
to parquet. The parquet outputs from this workflow can later be used to
update the Working DuckLake.

```{r}
#| label: validate

# validate data quality
validation <- validate_for_release(con)

if (validation$passed) {
  say("Validation passed!")
  if (nrow(validation$checks) > 0) {
    validation$checks |>
      filter(status != "pass") |>
      datatable(caption = "Validation Warnings")
  }
} else {
  cat("Validation FAILED:\n")
  cat(paste("-", validation$errors, collapse = "\n"))
}
```

## Enforce Column Types

Force integer/smallint types on columns that R's `numeric` mapped to
`DOUBLE` during `dbWriteTable()`. Uses `flds_redefine.csv` (`type_new`)
as the source of truth for source-table columns, plus explicit overrides
for derived-table columns.

```{r}
#| label: enforce_types

type_changes <- enforce_column_types(
  con = con,
  d_flds_rd = d$d_flds_rd,
  type_overrides = list(),
  tables = db_tables(con),
  verbose = TRUE
)

if (nrow(type_changes) > 0) {
  type_changes |>
    datatable(caption = "Column type changes applied")
}

# recreate casts_derived VIEW after type changes
dbExecute(con, "DROP VIEW IF EXISTS casts_derived")
dbExecute(
  con,
  "
  CREATE VIEW casts_derived AS
  SELECT *,
    EXTRACT(YEAR FROM datetime_start_utc)::SMALLINT    AS year,
    EXTRACT(MONTH FROM datetime_start_utc)::SMALLINT   AS month,
    EXTRACT(QUARTER FROM datetime_start_utc)::SMALLINT AS quarter,
    EXTRACT(DOY FROM datetime_start_utc)::SMALLINT     AS julian_day,
    (datetime_start_utc::DATE - DATE '1899-12-30')     AS julian_date,
    FLOOR(ABS(latitude))::SMALLINT                AS lat_deg,
    (ABS(latitude) - FLOOR(ABS(latitude))) * 60    AS lat_min,
    CASE WHEN latitude >= 0 THEN 'N' ELSE 'S' END AS lat_hem,
    FLOOR(ABS(longitude))::SMALLINT                AS lon_deg,
    (ABS(longitude) - FLOOR(ABS(longitude))) * 60    AS lon_min,
    CASE WHEN longitude >= 0 THEN 'E' ELSE 'W' END AS lon_hem,
    REPLACE(REPLACE(site_key, '.', ''), ' ', '')   AS db_sta_key
  FROM casts"
)
say("Recreated casts_derived VIEW after column type enforcement")
```

## Data Preview

Preview first and last rows of each table before writing parquet outputs.

```{r}
#| label: preview_tables
#| results: asis

preview_tables(
  con,
  c(
    "casts",
    "bottle",
    "bottle_measurement",
    "cast_condition",
    "measurement_type"
  )
) # grid canonical from swfsc
```

```{r}
#| label: preview_casts_derived

# casts_derived is a VIEW; exclude GEOMETRY column for R driver compatibility
cd_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = 'casts_derived' AND data_type != 'GEOMETRY'"
)$column_name
dbGetQuery(
  con,
  glue("SELECT {paste(cd_cols, collapse=', ')} FROM casts_derived LIMIT 100")
) |>
  datatable(
    caption = "casts_derived — first 100 rows (VIEW)",
    rownames = FALSE,
    filter = "top"
  )
```

## Cross-Dataset Integration

Link bottle casts to SWFSC ship/cruise reference tables via ship matching
and cruise key derivation. This replaces the separate `merge_ichthyo_bottle.qmd`
workflow step.

### Load Reference Tables

```{r}
#| label: load_reference_tables

# tables declared in calcofi.modifies frontmatter — loaded as TABLEs (writable)
# all other dependency tables loaded as VIEWs (zero-copy, read-only)
modifies_tables <- c("ship") # from calcofi.modifies in YAML frontmatter
load_prior_tables(
  con = con,
  tables = modifies_tables,
  parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo")
)
load_prior_tables(
  con = con,
  tables = setdiff(c("ship", "cruise"), modifies_tables),
  parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  as_view = TRUE
)

# snapshot PKs of modified tables before modifications
modifies_pks <- list()
for (tbl in modifies_tables) {
  pk_col <- dbGetQuery(
    con,
    glue(
      "SELECT column_name FROM information_schema.columns
     WHERE table_name = '{tbl}' ORDER BY ordinal_position LIMIT 1"
    )
  )$column_name
  modifies_pks[[tbl]] <- list(
    pk_col = pk_col,
    keys = dbGetQuery(con, glue("SELECT {pk_col} FROM {tbl}"))[[1]]
  )
}
```

### Ship Matching + Cruise Bridge

```{r}
#| label: derive_cruise_key

# cruise_key = YYYY-MM-NODC where YYYY-MM is the cruise's DESIGNATED month:
# resolved by the reference cruise whose date span holds the cast, then by the
# source's own `cruise` (YYYYMM), and only last by the cast's month. Before
# calcofi4db 3.20.0 only the last rule ran, which split every cruise that
# straddled a month boundary (184 of 664) onto two keys — and onto a REAL
# neighbouring cruise of the same ship, so no FK ever objected.
bridge_result <- derive_cruise_key_on_casts(
  con = con,
  ship_renames_csv = here("metadata/ship_renames.csv"),
  fetch_ices = TRUE,
  datetime_col = "datetime_start_utc",
  cruise_ym_col = "cruise"
)
bridge_result$key_stats |>
  datatable(caption = "cruise_key resolution: span (reference date span) > source (Cruise YYYYMM) > month")

# the source designation and the resolved key disagree only where the reference
# cruise designates the cruise differently (ichthyo calls the 9 Feb - 29 Mar 1984
# Jordan cruise 8403, this database 8402) — the reference wins, every dataset
# joins to it. Report the disagreements; a large count means a ship-code problem.
dbGetQuery(con, "
  SELECT cruise, cruise_key, cruise_key_method, COUNT(*) AS n_casts
  FROM casts
  WHERE cruise_key IS NOT NULL
    AND SUBSTR(cruise_key, 1, 4) || SUBSTR(cruise_key, 6, 2) <> cruise
  GROUP BY 1, 2, 3 ORDER BY n_casts DESC") |>
  datatable(caption = "Casts whose resolved cruise month differs from the source Cruise designation")

# insert interim entries for still-unmatched ships (ship_nodc = "?XX?")
ensure_interim_ships(con, bridge_result$ship_matches)

bridge_result$ship_matches |>
  datatable(caption = "Ship matching results")
```

```{r}
#| label: ship_match_stats

bridge_result$ship_matches |>
  count(match_type) |>
  datatable(caption = "Ship match type counts")
```

```{r}
#| label: cruise_bridge_stats

bridge_result$cruise_stats |>
  datatable(caption = "Cruise bridge match statistics")
```

```{r}
#| label: unmatched_ships

if (nrow(bridge_result$unmatched_report) > 0) {
  bridge_result$unmatched_report |>
    datatable(caption = "Unmatched ship codes (no ship_key)")
} else {
  say("All ship codes matched!")
}
```

### Cross-Dataset Validation

```{r}
#| label: cross_validate

# use information_schema to check columns (avoids GEOMETRY type issues with dbListFields)
casts_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = 'casts'"
)$column_name
grid_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = 'grid'"
)$column_name

# grid_key integrity: casts.grid_key should all be in grid.grid_key
if ("grid_key" %in% casts_cols && "grid_key" %in% grid_cols) {
  grid_orphans <- dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM casts c
     WHERE c.grid_key NOT IN (SELECT grid_key FROM grid)"
  )$n
  say(glue("Grid key orphans in casts: {grid_orphans}"))
}

# ship PK uniqueness
ship_dups <- dbGetQuery(
  con,
  "SELECT ship_key, COUNT(*) AS n FROM ship
   GROUP BY ship_key HAVING COUNT(*) > 1"
)
if (nrow(ship_dups) > 0) {
  warning(glue("Duplicate ship_key values: {nrow(ship_dups)}"))
}

# cruise PK uniqueness
cruise_dups <- dbGetQuery(
  con,
  "SELECT cruise_key, COUNT(*) AS n FROM cruise
   GROUP BY cruise_key HAVING COUNT(*) > 1"
)
if (nrow(cruise_dups) > 0) {
  warning(glue("Duplicate cruise_key values: {nrow(cruise_dups)}"))
}

# cruise bridge coverage
bridge_coverage <- dbGetQuery(
  con,
  "SELECT
    COUNT(*) AS total_casts,
    SUM(CASE WHEN ship_key IS NOT NULL THEN 1 ELSE 0 END) AS with_ship_key,
    SUM(CASE WHEN cruise_key IS NOT NULL THEN 1 ELSE 0 END) AS with_cruise_key
  FROM casts"
)
bridge_coverage |> datatable(caption = "Cruise bridge coverage")

say("Cross-dataset validation complete")
```


> **Ordering note.** The cruise bridge runs *before* the core projection: the
> `sample` event dimension carries `cruise_key`, so it must be derived first.
> This also retires the old `rewrite_casts_parquet` step, which existed only to
> patch `ship_key`/`cruise_key` into a `casts.parquet` already written — parquet
> is now written once, after every column is known.

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`). These core tables **are** this ingest's
output: `release_database.qmd` concatenates the per-dataset shards rather than
re-deriving the core, so there is exactly one projection to keep correct.

Bottle is a two-level hierarchy — `cast` -> `bottle` — carried in `sample` as an
adjacency list. Cast conditions become `sample_measurement` (they describe the
event, not an observation); the long `bottle_measurement` values become `obs` at
`realm = 'env'`.

`bottom_depth` rides along in `cast_condition`: the pivot above turns the source
`Bottom_D` column into a condition row and drops it from `casts`, so the cast's
water depth reaches `sample_measurement` through the same arm as wind speed and
sea state rather than needing one of its own. An added
`UNION ALL SELECT bottom_depth_m FROM casts` would be both redundant *and* a
binder error — that column no longer exists by the time this runs.

::: {.callout-important}
## This section must run AFTER Cross-Dataset Integration

`sample` carries `cruise_key`, and `cruise_key` is only resolved on `casts` by
the ship match + `derive_cruise_key()` above. If the projection runs first,
`cruise_key` is silently NULL on all 895,371 bottles instead of 99.93% populated
— no error, just a column of NULLs in the release.
:::

```{r}
#| label: emit_core

ds_key <- "calcofi_bottle"

# This projection lives here, in the notebook that owns the dataset, not in a
# switch(dataset_key, ...) arm inside calcofi4db. Bottle is env-only: no taxa
# references, no taxon_key on obs.
#
# sample — two levels: cast (root) and bottle (leaf, parent = its cast). The
# bottle inherits the cast's position/time/grid/cruise; only depth is its own.
append_sample(con, glue("
  SELECT {ns_key(ds_key, 'cast', 'cast_id')} AS sample_key, 'cast' AS sample_type,
         NULL::VARCHAR AS parent_sample_key,
         {ns_key(ds_key, 'cast', 'cast_id')} AS root_sample_key,
         '{ds_key}' AS dataset_key, grid_key, site_key, cruise_key,
         CAST(order_occ AS INTEGER) AS order_occ,
         latitude, longitude, CAST(datetime_start_utc AS TIMESTAMP) AS datetime,
         NULL::DOUBLE AS depth_min_m, NULL::DOUBLE AS depth_max_m,
         NULL::VARCHAR AS tow_type
  FROM casts"))
append_sample(con, glue("
  SELECT {ns_key(ds_key, 'bottle', 'b.bottle_id')} AS sample_key, 'bottle' AS sample_type,
         {ns_key(ds_key, 'cast', 'b.cast_id')} AS parent_sample_key,
         {ns_key(ds_key, 'cast', 'b.cast_id')} AS root_sample_key,
         '{ds_key}' AS dataset_key, c.grid_key, b.site_key, c.cruise_key,
         CAST(c.order_occ AS INTEGER) AS order_occ, c.latitude, c.longitude,
         CAST(c.datetime_start_utc AS TIMESTAMP) AS datetime,
         b.depth_m AS depth_min_m, b.depth_m AS depth_max_m,
         NULL::VARCHAR AS tow_type
  FROM bottle b JOIN casts c USING (cast_id)"))

# obs — the long bottle measurements, at realm = 'env'
append_obs(con, glue("
  SELECT 'env', '{ds_key}', {ns_key(ds_key, 'bottle', 'b.bottle_id')},
         c.grid_key, c.cruise_key, c.latitude, c.longitude,
         CAST(c.datetime_start_utc AS TIMESTAMP), b.depth_m, b.depth_m,
         NULL::VARCHAR, NULL::VARCHAR,
         m.measurement_type, m.measurement_value, m.measurement_qual, m.measurement_prec
  FROM bottle_measurement m JOIN bottle b USING (bottle_id) JOIN casts c USING (cast_id)"))

# sample_measurement — event-level cast conditions (wind, sea state, bottom
# depth, …). `cast_id` is DOUBLE in cast_condition, so cast through BIGINT first
# or every key comes out as '…:cast:5.0' and orphans against sample.
append_sample_measurement(con, glue("
  SELECT {ns_key(ds_key, 'cast', 'CAST(cast_id AS BIGINT)')}, '{ds_key}',
         condition_type, condition_value, NULL::VARCHAR
  FROM cast_condition"))

core <- list(
  sample             = dbGetQuery(con, "SELECT COUNT(*) FROM sample")[[1]],
  obs                = dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]],
  sample_measurement = dbGetQuery(con, "SELECT COUNT(*) FROM sample_measurement")[[1]])
cat(glue(
  "core projection — sample={core$sample %||% 0} obs={core$obs %||% 0} ",
  "sample_measurement={core$sample_measurement %||% 0}\n"))

n_cast <- dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE sample_type='cast'")[[1]]
n_btl  <- dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE sample_type='bottle'")[[1]]
n_bd   <- dbGetQuery(con, "SELECT COUNT(*) FROM sample_measurement WHERE measurement_type='bottom_depth'")[[1]]
n_ck   <- dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE sample_type='bottle' AND cruise_key IS NOT NULL")[[1]]
stopifnot(
  "sample must hold one row per cast"   = n_cast == dbGetQuery(con, "SELECT COUNT(DISTINCT cast_id) FROM casts")[[1]],
  "sample must hold one row per bottle" = n_btl  == dbGetQuery(con, "SELECT COUNT(DISTINCT bottle_id) FROM bottle")[[1]],
  "sample_key must be globally unique" =
    dbGetQuery(con, "SELECT COUNT(*) FROM (SELECT sample_key FROM sample
                     GROUP BY 1 HAVING COUNT(*) > 1)")[[1]] == 0,
  "obs must be one row per bottle measurement" =
    core$obs ==
    dbGetQuery(con, "SELECT COUNT(*) FROM bottle_measurement m JOIN bottle b USING (bottle_id)
                     JOIN casts c USING (cast_id)")[[1]],
  "every bottle must resolve to its cast" =
    dbGetQuery(con, "SELECT COUNT(*) FROM sample c LEFT JOIN sample p
                     ON c.parent_sample_key = p.sample_key
                     WHERE c.parent_sample_key IS NOT NULL AND p.sample_key IS NULL")[[1]] == 0,
  "sample_measurement must be one row per cast condition, not duplicated" =
    core$sample_measurement == dbGetQuery(con, "SELECT COUNT(*) FROM cast_condition")[[1]],
  "every sample_measurement.sample_key must resolve in sample (the DOUBLE cast_id trap)" =
    dbGetQuery(con, "SELECT COUNT(*) FROM sample_measurement m
                     LEFT JOIN sample s USING (sample_key)
                     WHERE s.sample_key IS NULL")[[1]] == 0,
  "bottom_depth must reach sample_measurement" = n_bd > 0,
  # guards the chunk-order constraint above: cruise_key comes from the
  # Cross-Dataset Integration section, which must already have run
  "cruise_key must be resolved on nearly every bottle" = n_ck > 0.99 * n_btl)
cat(glue("sample: {format(n_cast, big.mark=',')} casts / {format(n_btl, big.mark=',')} bottles ",
         "({round(100 * n_ck / max(n_btl, 1), 2)}% with cruise_key); ",
         "{format(n_bd, big.mark=',')} bottom_depth rows"), "\n")

# serve the retired per-dataset table names as VIEWs over the core, so
# in-notebook consumers (casts_derived, the mismatch collectors) and ad-hoc
# queries keep working against the old names. The real tables are DROPped rather
# than kept as `_src`: bottle_measurement is 11M rows and the wrangling DB is
# file-backed, so retaining a second copy is not free.
compat <- c(
  casts = "
    SELECT CAST(split_part(s.sample_key, ':', 3) AS BIGINT) AS cast_id,
           s.site_key, s.grid_key, s.cruise_key, s.order_occ,
           s.latitude, s.longitude, s.datetime AS datetime_start_utc, s.geom
    FROM sample s
    WHERE s.dataset_key = 'calcofi_bottle' AND s.sample_type = 'cast'",
  bottle = "
    SELECT CAST(split_part(s.sample_key, ':', 3) AS BIGINT) AS bottle_id,
           CAST(split_part(s.parent_sample_key, ':', 3) AS BIGINT) AS cast_id,
           s.site_key, s.depth_min_m AS depth_m
    FROM sample s
    WHERE s.dataset_key = 'calcofi_bottle' AND s.sample_type = 'bottle'",
  cast_condition = "
    SELECT sample_measurement_id AS cast_condition_id,
           CAST(split_part(sample_key, ':', 3) AS BIGINT) AS cast_id,
           measurement_type AS condition_type, measurement_value AS condition_value
    FROM sample_measurement
    WHERE dataset_key = 'calcofi_bottle'",
  bottle_measurement = "
    SELECT obs_id AS bottle_measurement_id,
           CAST(split_part(sample_key, ':', 3) AS BIGINT) AS bottle_id,
           measurement_type, measurement_value, measurement_qual, measurement_prec
    FROM obs WHERE dataset_key = 'calcofi_bottle'")
for (nm in names(compat)) {
  # DROP has to know which it is: DuckDB refuses "DROP TABLE" on a view (and
  # vice versa), and a re-run inside one session finds the view already there
  t <- dbGetQuery(con, glue(
    "SELECT table_type FROM information_schema.tables WHERE table_name = '{nm}'"))
  if (nrow(t)) {
    kind <- if (grepl("VIEW", t$table_type[1], ignore.case = TRUE)) "VIEW" else "TABLE"
    invisible(dbExecute(con, glue('DROP {kind} IF EXISTS "{nm}"')))
  }
  invisible(dbExecute(con, glue("CREATE OR REPLACE VIEW {nm} AS {compat[[nm]]}")))
}
cat(glue("compat views over core: {paste(names(compat), collapse = ', ')}"), "\n")
```


## Write Parquet Outputs

Export tables to parquet files for downstream use.

```{r}
#| label: write_parquet

# collect mismatches for manifest
mismatches <- list(
  ships = collect_ship_mismatches(con, "casts"),
  measurement_types = collect_measurement_type_mismatches(
    con,
    here("metadata/measurement_type.csv")
  ),
  cruise_keys = collect_cruise_key_mismatches(con, "casts")
)

# write parquet files with manifest
# export only tables this ingest creates (dependency _new deltas added later)
# core shards + the shared refs this ingest owns; casts/bottle/bottle_measurement
# /cast_condition are VIEWs over the core now, not stored bytes
own_tables <- core_output_tables(con, extra = c("measurement_type", "dataset"))

parquet_stats <- write_parquet_outputs(
  con              = con,
  output_dir       = dir_parquet,
  tables           = own_tables,
  sort_by          = list(
    obs    = c("grid_key", "measurement_type"),
    sample = "hilbert:longitude,latitude"),
  strip_provenance = FALSE,
  mismatches       = mismatches
)

parquet_stats |>
  mutate(file = basename(path)) |>
  select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

Build `metadata.json` sidecar file documenting all tables and columns in
parquet outputs. DuckDB `COMMENT ON` does not propagate to parquet, so
this provides the metadata externally.

```{r}
#| label: write_metadata

metadata_path <- build_metadata_json(
  con = con,
  d_tbls_rd = d$d_tbls_rd,
  d_flds_rd = d$d_flds_rd,
  metadata_derived_csv = c(
    here("metadata/core_dictionary.csv"),
    here("metadata/calcofi/bottle/metadata_derived.csv")
  ),
  output_dir = dir_parquet,
  tables = own_tables,
  set_comments = TRUE,
  provider = provider,
  dataset = dataset,
  workflow_url = cc$workflow_url,
  tables_owned = tables_owned
)

# write relationships.json sidecar with PKs/FKs
build_relationships_json(
  rels = core_relationships(own_tables),
  output_dir = dir_parquet,
  provider = provider,
  dataset = dataset
)

# show metadata summary
metadata <- jsonlite::fromJSON(metadata_path)
tibble(
  table = names(metadata$tables),
  n_cols = map_int(
    names(metadata$tables),
    ~ sum(grepl(glue("^{.x}\\."), names(metadata$columns)))
  ),
  name_long = map_chr(metadata$tables, ~ .x$name_long)
) |>
  datatable(caption = "Table metadata summary")
```

```{r}
#| label: show_metadata_json

listviewer::jsonedit(
  jsonlite::fromJSON(metadata_path, simplifyVector = FALSE),
  mode = "view")
```

```{r}
#| label: show_relationships_json

listviewer::jsonedit(
  jsonlite::fromJSON(
    file.path(dir_parquet, "relationships.json"),
    simplifyVector = FALSE),
  mode = "view")
```

## Upload to GCS Archive

Upload parquet files, manifest, and metadata sidecar to
`gs://calcofi-db/ingest/{provider}_{dataset}/`.

```{r}
#| label: upload_gcs

sync_to_gcs(
  local_dir   = dir_stage,
  sidecar_dir = dir_parquet,
  gcs_prefix = glue("ingest/{dir_label}"),
  bucket     = "calcofi-db")
```

## Cleanup

```{r}
#| label: cleanup

# close local wrangling database connection
close_duckdb(con)
say("Local wrangling database connection closed")

# note: parquet outputs are in data/parquet/calcofi_bottle/
# these can be used to update the Working DuckLake in a separate workflow
say(glue("Parquet outputs written to: {dir_parquet}"))
say(glue("GCS outputs at: gs://calcofi-db/ingest/{dir_label}/"))
```

## TODO

- [x] Review column names and apply db naming conventions, per @docs/db.qmd and @workflows/README_PLAN.qmd: cast_cnt -> cast_id, btl_cnt -> bottle_id, sta_id -> site_key, etc.
- [x] Pivot longer casts/bottle tables for easier analysis. Created bottle_measurement (value/prec/qual) and cast_condition tables, plus measurement_type reference.
- [x] Consolidate casts temporal columns (date+time -> datetime_start_utc), drop 16 derivable columns, create casts_derived VIEW.

- [ ] Link possible fields to @workflows/ingest_swfsc_ichthyo.qmd. Flag mismatches.
- [x] Cruise bridge prep: renamed `Cruise_ID` → `cruise_key_0` (was `cruise_key`) to avoid collision with SWFSC `cruise_key` (YYMMKK). Merge workflow adds `cruise_key` as new column derived from `datetime_start_utc` + `ship_key`.
- [ ] Validate casts.ship_code -> ship.ship_nodc FK after integration with swfsc data.

- [ ] Review documentation for more comment descriptions at [Data \>
  Data Formats \| CalCOFI.org](https://calcofi.com/index.php?option=com_content&view=category&id=73&Itemid=993)

::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
