---
title: "CalCOFI NOAA Fish Larvae Sizes on ERDDAP"
editor: visual
editor_options: 
  chunk_output_type: console
---

## Data

**Goal**: Read in the "CalCOFI NOAA Fish Larvae Sizes" dataset from ERDDAP with the following:

- erddap: [coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIlrvsiz.html](https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIlrvsiz.html)\
  ERDDAP - CalCOFI NOAA Fish Larvae Sizes - Data Access Form
  - `{provider}`: `coastwatch.pfeg.noaa.gov`
  - `{dataset}`: `erdCalCOFIlrvsiz`
- workflow: `ingest_{provider}_{dataset}.qmd`
- data: Google Drive `calcofi/data/{provider}/{dataset}`
  - data: `{dataset}.csv`
  - metadata: `{dataset}_info.csv`
- database definitions: `ingest/{provider}/{dataset}`
  - tables: `tbls_redefine.csv`
  - columns: `flds_redefine.csv`

```{r}
librarian::shelf(
  dplyr, DT, glue, here, readr, rerddap, stringr, tibble)

# variables
dir_data     <- "/Users/bbest/Library/CloudStorage/GoogleDrive-ben@ecoquants.com/My Drive/projects/calcofi/data" 
dir_provider <- "coastwatch.pfeg.noaa.gov"
ds_url       <- "https://coastwatch.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIlrvsiz.html"

# extract ERDDAP url and dataset ID from the dataset URL
ed_pattern  <- "(https://.*)/tabledap/([A-Za-z0-9]+)\\.html"
ed_url      <- str_replace(ds_url, ed_pattern, "\\1") # "erdCalCOFIlrvsiz"
ed_id       <- str_replace(ds_url, ed_pattern, "\\2") # "https://coastwatch.pfeg.noaa.gov/erddap"
dir_dataset <- glue("{dir_data}/{dir_provider}")
d_csv       <- glue("{dir_dataset}/{ed_id}.csv")
m_csv       <- glue("{dir_dataset}/{ed_id}_info.csv")

if (!dir.exists(dir_dataset))
  dir.create(dir_dataset)

ed_info <- info(ed_id, url = ed_url)

if (!file.exists(d_csv)){
  d <- tabledap(ed_info)
  write_csv(d, d_csv)
} else {
  d <- read_csv(d_csv)
}
dim(d)

head(d) |> 
  datatable()
```

## Metadata

- [ERDDAP - Information about CalCOFI NOAA Fish Larvae Sizes, from NOAA SWFSC](https://coastwatch.pfeg.noaa.gov/erddap/info/erdCalCOFIlrvsiz/index.html)

```{r}
if (!file.exists(m_csv)){
  d_m <- ed_info$alldata |> 
    bind_rows() |> 
    tibble()
  write_csv(d_m, m_csv)
} else {
  d_m <- read_csv(m_csv)
}

d_m |> 
  datatable()
```

## Google Drive

Output CSVs in Google Drive: 

[calcofi](https://drive.google.com/drive/folders/13pWB5x59WSBR0mr9jJjkx7rri9hlUsMv) / [data](https://drive.google.com/drive/folders/1xxdWa4mWkmfkJUQsHxERTp9eBBXBMbV7) / [coastwatch.pfeg.noaa.gov](https://drive.google.com/drive/folders/1PDzCCpdSD4aBwpa7AZv2FNO1-pZg_WwP) / 

- data: [**erdCalCOFIlrvsiz.csv**](https://drive.google.com/file/d/1YkWJK8aUlDuP2hm5BszR4a82hzAogr0G/view?usp=drive_link)
- metadata: [**erdCalCOFIlrvsiz_info.csv**](https://drive.google.com/open?id=1GVgoMWbmC7iL0ULqDwFimCLCd3cs61N2&usp=drive_fs)

## TODO: load into database

See:

- [5.3 Ingest datasets with documentation - Database – CalCOFI.io Docs](https://calcofi.io/docs/db.html#ingest-datasets-with-documentation)
