---
title: "Ingest CCE-LTER Euphausiids"
calcofi:
  target_name: ingest_cce_lter_euphausiids
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/cce-lter_euphausiids/manifest.json
  provider: cce-lter
  dataset: euphausiids
  workflow_url: https://calcofi.io/workflows/ingest_cce-lter_euphausiids.html
  questions_file: metadata/cce-lter/euphausiids/questions.csv
  dataset_meta:
    dataset_name: CCE-LTER Euphausiid Abundance
    description: >
      Species- and life-stage-resolved euphausiid (krill) abundance from
      CalCOFI / CCE-LTER net tows (BTEDB export), 1951-present, one row per
      tow x species x life stage. Supersedes the prior single-Abundance-
      column ingest, which had no taxonomic scope (Q02, now resolved) —
      units remain provisional pending provider confirmation (Q01).
    citation_main: ""
    link_calcofi_org: ""
    link_data_source: "BTEDB (Bongo Tow Euphausiid Database) export"
    link_others: []
    coverage_temporal: 1951-01 to present
    coverage_spatial: "CalCOFI grid incl. Baja lines"
    license: ""
    pi_names: Rasmus Swalethorp; Linsey Sala
  tables_owned:
    - {table: euphausiids_tow}
    - {table: euphausiids_taxon}
    - {table: euphausiids_measurement}
    - {table: euphausiids_summary}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#d2f0bb"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: BTEDB (Bongo Tow Euphausiid Database) export, `data.csv` — one row
per net tow, 237 columns: 12 tow/position/time columns + 225
`{Genus}_{species}_{life_stage}_Abundance` columns spanning 37 species across
8 genera (*Euphausia*, *Nematobrachion*, *Nematoscelis*, *Nyctiphanes*,
*Stylocheiron*, *Tessarabrachion*, *Thysanoessa*, *Thysanopoda*) and 16
life-stage tokens (adult, juvenile, calyptopis + C1–C3, furcilia + F1–F7,
larvae, metanauplius, damaged).

- **Provider**: `cce-lter`
- **Grain**: one row per net tow x species x life stage (long format)
- **Strategy**: load the tow-level position columns into `euphausiids_tow`
  exactly as before; build a small `euphausiids_taxon` reference table from
  the distinct species named in the abundance columns; **pivot the 225 wide
  columns into long format** (species + life_stage + value per tow) into
  `euphausiids_measurement`; summarize into `euphausiids_summary`.

**What changed from the prior ingest**: the previous version read a single
pre-aggregated `Abundance` column with no species dimension at all — that
was the entire content of open question Q02 ("no species column"). This
version reads the species-resolved BTEDB export instead, which **resolves
Q02** but does not change Q01 (units) or the ship/cruise/coordinate
questions, which are unchanged from before and still open.

```{mermaid}
graph LR
  A[data.csv<br/>7,482 tows x 225 species/stage cols] --> B[euphausiids_tow<br/>position + keys]
  A --> T[euphausiids_taxon<br/>37 species]
  A --> C[euphausiids_measurement<br/>long: tow x taxon x life_stage]
  C --> D[euphausiids_summary<br/>avg/stddev]
  B -.ship/cruise/grid.-> E[(shared refs)]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# common ingest settings (overwrite, dir_data)
source(here("libs/ingest.R"))

# provider/dataset/metadata from this file's authoritative YAML block
cc           <- read_calcofi_meta(here("ingest_cce-lter_euphausiids.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")

dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path))                  file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal")))  file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))   dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

# load unified measurement_type reference
meas_type_csv <- here("metadata/measurement_type.csv")
d_meas_type   <- read_csv(meas_type_csv)
```

## Read Source Data

The source CSV lives at `{dir_data}/euphausiids/data.csv`, read directly
(not via `read_csv_files()`) same as before. Source columns are renamed to
canonical names per `metadata/cce-lter/euphausiids/flds_redefine.csv`.

```{r}
#| label: read-source

euph_csv <- path_expand(glue("{dir_data}/euphausiids/data.csv"))
stopifnot("euphausiids data.csv not found" = file_exists(euph_csv))

# archive source to GCS for provenance
sync_to_gcs(
  local_dir  = path_dir(euph_csv),
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket     = "calcofi-files-public",
  exclude    = c(".DS_Store", "*.tmp", "*.gdoc"))  # .gdoc is a Drive stub

d_raw <- read_csv(euph_csv)
abund_cols <- names(d_raw) |> str_subset("_Abundance$")

cat(glue("Read {format(nrow(d_raw), big.mark=',')} rows, ",
         "{ncol(d_raw)} columns from {basename(euph_csv)} ",
         "({length(abund_cols)} species/stage abundance columns)"), "\n")
```

## Clean, Type-Cast, and Correct

Same tow-level cleaning as the prior ingest (Q05/Q06 coordinate and
tow-end-time fixes, unchanged) — this section only touches the 12
non-abundance columns. Canonical field names follow
`metadata/field_dictionary.csv`.

```{r}
#| label: clean-data

d_clean <- d_raw |>
  transmute(
    tow_id             = as.integer(RowNumber),
    cruise_orig        = Cruise,
    ship_name          = Ship,
    date               = as.Date(Date),
    line               = suppressWarnings(as.numeric(Line)),
    station            = suppressWarnings(as.numeric(Station)),
    region             = Region,
    datetime_start_utc = as_datetime(TowBegin),     # Q07: assumed already UTC
    datetime_end_utc   = as_datetime(TowEnd),
    latitude           = as.numeric(Latitude),
    longitude          = as.numeric(Longitude)) |>
  mutate(
    # Q05: sign errors / garbage coordinates -> null out, same rule as before
    longitude = if_else(longitude > 0, -longitude, longitude),
    bad_coord = latitude > 51 | latitude < 20 | longitude < -135 | longitude > -105,
    latitude  = if_else(bad_coord, NA_real_, latitude),
    longitude = if_else(bad_coord, NA_real_, longitude),
    # Q06: out-of-range tow-end timestamps -> null
    datetime_end_utc = if_else(
      year(datetime_end_utc) > 2025 | year(datetime_end_utc) < 1949,
      as_datetime(NA), datetime_end_utc),
    site_key = if_else(
      is.na(line) | is.na(station), NA_character_,
      sprintf("%05.1f %05.1f", line, station)))

cat(glue("Cleaned {nrow(d_clean)} tows; ",
         "{sum(d_clean$bad_coord, na.rm=TRUE)} coordinate(s) nulled (Q05), ",
         "{sum(is.na(d_clean$datetime_end_utc))} tow-end null (incl. Q06)"), "\n")
```

## Build Taxon Reference Table

Parse the 225 abundance column names into genus/species/life_stage. Column
pattern is `{Genus}_{species}_{life_stage}_Abundance`; life_stage itself may
contain an underscore (e.g. `calyptopis_C1`, `furcilia_F6`), so the regex
captures everything between the species token and the trailing `_Abundance`
rather than assuming a fixed number of segments.

```{r}
#| label: parse-columns

col_parts <- tibble(col = abund_cols) |>
  mutate(
    stem = str_remove(col, "_Abundance$"),
    genus = str_extract(stem, "^[A-Za-z]+"),
    rest  = str_remove(stem, "^[A-Za-z]+_"),
    species    = str_extract(rest, "^[a-z]+"),
    life_stage = str_remove(rest, "^[a-z]+_") |> str_replace_all("_", " "))

n_species <- n_distinct(col_parts$genus, col_parts$species)
cat(glue("Parsed {nrow(col_parts)} columns -> ",
         "{n_species} distinct species x {n_distinct(col_parts$life_stage)} life-stage tokens"), "\n")

# Question 09 (NEW): two species present in the canonical DB taxon list are absent
# from this BTEDB export entirely (Bentheuphausia amblyops, Thysanopoda
# cristata) — nothing to ingest for these until the provider confirms
# whether BTEDB tracks them at all. Tracked below, not silently dropped.
#
# Question 10 (NEW): the DB's "Hansarsia" genus (atlantica, difficilis, gracilis,
# microps, tenella) is a synonym for BTEDB's "Nematoscelis" — same 5
# species, older vs. current genus name. Resolved here as a straight
# rename (not flagged as missing), but the taxon-standardization step
# downstream (calcofi4db::standardize_species()) should confirm the WoRMS
# AphiaID resolves identically under either name.
col_parts <- col_parts |>
  mutate(scientific_name = glue("{genus} {species}"))

euph_taxon <- col_parts |>
  distinct(scientific_name, genus, species) |>
  arrange(scientific_name) |>
  mutate(taxon_id = row_number(), .before = 1)

cat(glue("euphausiids_taxon: {nrow(euph_taxon)} species"), "\n")
```

## Pivot to Long Format

Reshape the 225 wide columns into one row per tow x species x life_stage,
dropping true zeros (a tow where a species/stage genuinely wasn't observed)
from the fact table — same convention as `euphausiids_measurement` not
storing NULLs before.

```{r}
#| label: pivot-long

d_wide <- d_raw |>
  transmute(tow_id = as.integer(RowNumber), across(all_of(abund_cols)))

d_long <- d_wide |>
  pivot_longer(-tow_id, names_to = "col", values_to = "abundance") |>
  left_join(col_parts, by = "col") |>
  left_join(euph_taxon |> select(taxon_id, scientific_name), by = "scientific_name") |>
  filter(!is.na(abundance), abundance > 0) |>
  select(tow_id, taxon_id, life_stage, abundance)

cat(glue("Pivoted to {format(nrow(d_long), big.mark=',')} non-zero tow x species x stage rows ",
         "(from {format(nrow(d_wide) * length(abund_cols), big.mark=',')} wide cells)"), "\n")
```

## Resolve Ship and Cruise Keys

Unchanged from the prior ingest — vessel name normalization (Q04) and
`cruise_key` derivation (Q03), against the same shared reference tables.

```{r}
#| label: resolve-keys

load_prior_tables(
  con,
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables      = c("ship", "cruise", "grid"),
  geom_tables = c("grid"),
  as_view     = TRUE)

d_ship <- dbGetQuery(con, "SELECT ship_key, ship_name, ship_nodc FROM ship") |>
  mutate(ship_name_norm = ship_name |> str_to_upper() |> str_squish())

d_keys <- d_clean |>
  mutate(
    ship_name_norm = ship_name |>
      str_replace("^R/?V\\.?\\s+", "") |> str_to_upper() |> str_squish()) |>
  left_join(
    d_ship |> select(ship_name_norm, ship_key, ship_nodc),
    by = "ship_name_norm") |>
  mutate(
    cruise_key = if_else(
      is.na(ship_nodc), NA_character_,
      glue("{format(date, '%Y-%m')}-{ship_nodc}") |> as.character()))

n_ship  <- sum(!is.na(d_keys$ship_key))
valid_ck <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM cruise")$cruise_key
d_keys <- d_keys |>
  mutate(cruise_key = if_else(cruise_key %in% valid_ck, cruise_key, NA_character_))
n_cruise <- sum(!is.na(d_keys$cruise_key))

cat(glue(
  "Ship match: {n_ship}/{nrow(d_keys)} ({round(100*n_ship/nrow(d_keys),1)}%); ",
  "cruise_key match: {n_cruise}/{nrow(d_keys)} ",
  "({round(100*n_cruise/nrow(d_keys),1)}%)"), "\n")
```

## Load Tidy Tow Table

```{r}
#| label: load-tow

d_tow <- d_keys |>
  transmute(
    tow_id, cruise_key, ship_key, ship_name, cruise_orig,
    site_key, line, station, region,
    datetime_start_utc, datetime_end_utc,
    latitude, longitude)

dbWriteTable(con, "euphausiids_tow", d_tow, overwrite = TRUE)
dbWriteTable(con, "euphausiids_taxon", euph_taxon, overwrite = TRUE)

cat(glue("euphausiids_tow: {dbGetQuery(con,'SELECT COUNT(*) FROM euphausiids_tow')[[1]]} rows"), "\n")
```

## Add Spatial

```{r}
#| label: spatial

add_point_geom(con, "euphausiids_tow", lon_col = "longitude", lat_col = "latitude")
grid_stats <- assign_grid_key(con, "euphausiids_tow")
grid_stats |> datatable(caption = "Grid assignment")
```

## Load Measurement Table

One measurement type, `euphausiid_abundance`, now dimensioned by
`taxon_id` + `life_stage` instead of being a single undifferentiated value
per tow — this is what resolves Q02.

```{r}
#| label: load-measurement

dbWriteTable(con, "euph_long_staged", d_long, overwrite = TRUE)

dbExecute(con,
  "CREATE OR REPLACE TABLE euphausiids_measurement AS
   SELECT ROW_NUMBER() OVER (ORDER BY tow_id, taxon_id, life_stage) AS euphausiids_measurement_id,
          tow_id, taxon_id, life_stage,
          'euphausiid_abundance' AS measurement_type,
          CAST(abundance AS DOUBLE) AS measurement_value,
          NULL::VARCHAR AS measurement_qual
   FROM euph_long_staged
   WHERE abundance IS NOT NULL
     AND NOT isnan(CAST(abundance AS DOUBLE))
     AND isfinite(CAST(abundance AS DOUBLE))")

dbExecute(con, "DROP TABLE euph_long_staged")

n_meas <- dbGetQuery(con, "SELECT COUNT(*) FROM euphausiids_measurement")[[1]]
cat(glue("euphausiids_measurement: {format(n_meas, big.mark=',')} rows"), "\n")
```

## Summarize Replicate Measurements

Aggregate replicate tows at each unique position x species x life_stage
into mean and standard deviation — same pattern as before, now with the
taxon/stage dimension carried through.

```{r}
#| label: measurement-summary

dbExecute(con,
  "CREATE OR REPLACE TABLE euphausiids_summary AS
   SELECT
     t.site_key, t.datetime_start_utc, t.latitude, t.longitude,
     m.taxon_id, m.life_stage, m.measurement_type,
     AVG(m.measurement_value) AS avg,
     CASE WHEN COUNT(*) = 1 THEN 0
          ELSE COALESCE(STDDEV_SAMP(m.measurement_value), 0) END AS stddev,
     COUNT(*) AS n_obs
   FROM euphausiids_measurement m
   JOIN euphausiids_tow t USING (tow_id)
   WHERE NOT isnan(m.measurement_value) AND isfinite(m.measurement_value)
   GROUP BY t.site_key, t.datetime_start_utc, t.latitude, t.longitude,
            m.taxon_id, m.life_stage, m.measurement_type")

cat(glue("euphausiids_summary: {dbGetQuery(con,'SELECT COUNT(*) FROM euphausiids_summary')[[1]]} rows"), "\n")
```

## Add Measurement Type

```{r}
#| label: add-measurement-type

euph_types <- tibble(
  measurement_type   = "euphausiid_abundance",
  description        = paste(
    "Euphausiid (krill) abundance per net tow, by species and life stage.",
    "PROVISIONAL: units unconfirmed (see question Q01)."),
  units              = "count/1000m3",   # Q01: PROVISIONAL best guess, unchanged
  `_source_column`   = "*_Abundance (pivoted; see euphausiids_taxon + life_stage)",
  `_source_table`    = "euphausiids_measurement",
  `_source_datasets` = "cce-lter_euphausiids",
  `_qual_column`     = NA_character_,
  `_prec_column`     = NA_character_)

if (!"euphausiid_abundance" %in% d_meas_type$measurement_type) {
  d_meas_type <- bind_rows(d_meas_type, euph_types)
  write_csv(d_meas_type, meas_type_csv)
  cat("Added measurement type: euphausiid_abundance (provisional units)\n")
} else {
  cat("euphausiid_abundance already registered\n")
}
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Schema Documentation

```{r}
#| label: schema

euph_rels <- list(
  primary_keys = list(
    euphausiids_tow         = "tow_id",
    euphausiids_taxon       = "taxon_id",
    euphausiids_measurement = "euphausiids_measurement_id",
    measurement_type        = "measurement_type"),
  foreign_keys = list(
    list(table = "euphausiids_measurement", column = "tow_id",
         ref_table = "euphausiids_tow", ref_column = "tow_id"),
    list(table = "euphausiids_measurement", column = "taxon_id",
         ref_table = "euphausiids_taxon", ref_column = "taxon_id"),
    list(table = "euphausiids_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

euph_tables <- c(
  "euphausiids_tow", "euphausiids_taxon", "euphausiids_measurement",
  "euphausiids_summary", "measurement_type", "dataset")
cc_erd(
  con, tables = euph_tables, rels = euph_rels,
  colors = list(
    lightblue   = c("euphausiids_tow", "euphausiids_measurement", "euphausiids_summary"),
    lightgreen  = "euphausiids_taxon",
    lightyellow = "measurement_type",
    white       = "dataset"))

build_relationships_json(
  rels = euph_rels, output_dir = dir_parquet,
  provider = provider, dataset = dataset)
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
if (length(results$warnings) > 0)
  cat("Warnings:\n", paste("-", results$warnings, collapse = "\n"), "\n")

cat(glue(
  "\nMatch coverage (non-NULL): ",
  "ship_key {round(100*dbGetQuery(con,\"SELECT AVG(CASE WHEN ship_key IS NOT NULL THEN 1 ELSE 0 END) FROM euphausiids_tow\")[[1]],1)}%, ",
  "cruise_key {round(100*dbGetQuery(con,\"SELECT AVG(CASE WHEN cruise_key IS NOT NULL THEN 1 ELSE 0 END) FROM euphausiids_tow\")[[1]],1)}%, ",
  "grid_key {round(100*dbGetQuery(con,\"SELECT AVG(CASE WHEN grid_key IS NOT NULL THEN 1 ELSE 0 END) FROM euphausiids_tow\")[[1]],1)}%"), "\n")

n_dup <- dbGetQuery(con,
  "SELECT COUNT(*) FROM (
     SELECT tow_id, COUNT(*) n FROM euphausiids_tow GROUP BY tow_id HAVING COUNT(*)>1)")[[1]]
cat(glue("euphausiids_tow tow_id duplicates: {n_dup}"), "\n")
```

## Data Preview

```{r}
#| label: preview-taxon

euph_taxon |> datatable(caption = "euphausiids_taxon — all species", rownames = FALSE)
```

```{r}
#| label: preview-measurement

dbGetQuery(con, "
  SELECT m.tow_id, t.scientific_name, m.life_stage, m.measurement_value
  FROM euphausiids_measurement m
  JOIN euphausiids_taxon t USING (taxon_id)
  LIMIT 100") |>
  datatable(caption = "euphausiids_measurement — first 100 rows (joined for readability)", rownames = FALSE)
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

dir_create(dir_parquet)

mismatches <- list(
  measurement_types = collect_measurement_type_mismatches(
    con, here("metadata/measurement_type.csv")),
  cruise_keys       = collect_cruise_key_mismatches(con, "euphausiids_tow"))

parquet_stats <- write_parquet_outputs(
  con        = con,
  output_dir = dir_parquet,
  tables     = c("euphausiids_tow", "euphausiids_taxon", "euphausiids_measurement",
                 "euphausiids_summary", "measurement_type", "dataset"),
  sort_by    = list(euphausiids_summary = "measurement_type"),
  strip_provenance = FALSE,
  mismatches = mismatches)

parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata

d_tbls_rd <- read_csv(here("metadata/cce-lter/euphausiids/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/cce-lter/euphausiids/flds_redefine.csv"))

metadata_path <- build_metadata_json(
  con        = con,
  d_tbls_rd  = d_tbls_rd,
  d_flds_rd  = d_flds_rd,
  output_dir = dir_parquet,
  tables     = c("euphausiids_tow", "euphausiids_taxon", "euphausiids_measurement", "euphausiids_summary"),
  set_comments = TRUE,
  provider   = provider,
  dataset    = dataset,
  workflow_url = cc$workflow_url,
  tables_owned = tables_owned)
```

## Upload to GCS

```{r}
#| label: upload-gcs

sync_to_gcs(
  local_dir  = dir_parquet,
  gcs_prefix = glue("ingest/{dir_label}"),
  bucket     = "calcofi-db")
```

## Questions for Data Providers

Follow-up questions for CCE-LTER (Rasmus Swalethorp, Linsey Sala), **ranked by
importance**. Q02 (taxonomic scope) is now resolved by this ingest — kept
below with status `resolved` for record-keeping, not because it's still
open. Questions 09/10 are new, arising directly from cross-checking this
export against the DB's canonical euphausiid species list.

```{r}
#| label: provider-questions

q_path <- here(glue("metadata/{provider}/{dataset}/questions.csv"))
read_csv(q_path) |>
  arrange(factor(priority, c("blocker", "high", "normal")), id) |>
  select(priority, question, context, related_field, status) |>
  datatable(
    caption = "Questions for data providers (ranked by importance)",
    options = list(dom = "t", pageLength = 20), rownames = FALSE)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```

::: {.callout-caution collapse="true"}
## Session Info
```{r session_info}
devtools::session_info()
```
:::
