---
title: "Ingest CalCOFI October 2022 Vertebrate eDNA"
calcofi:
  target_name: ingest_calcofi_2022_edna
  workflow_type: ingest
  dependency:
    # reads swfsc_ichthyo's staged ship/cruise/grid (load_prior_tables() below)
    - ingest_swfsc_ichthyo
  output: data/parquet/calcofi_2022-edna/manifest.json
  provider: calcofi
  dataset: 2022-edna
  workflow_url: https://calcofi.io/workflows/ingest_calcofi_2022-edna.html
  questions_file: metadata/calcofi/2022-edna/questions.csv
  dataset_meta:
    dataset_name: CalCOFI October 2022 Vertebrate eDNA
    # display trio, read by the release `dataset` table and consumer apps
    dataset_name_short: 2022 Vertebrate eDNA
    category: Genomics & eDNA
    color: "#2e8b57"
  # publishes the consolidated core: the source shape (a Darwin Core Archive,
  # not a wide CSV) is wrangled in this notebook and projected into the shared
  # core tables. This is the first ingest to populate the Genomics & eDNA
  # category with real observations (calpoly_whale-edna and
  # jcvi_ncog-metabarcoding remain holdings, no ingest).
  tables_owned:
    - {table: sample, shared: true, note: "one row per water filter (sample_type filter, parentEventID grain)"}
    - {table: obs, shared: true, note: "one row per filter x taxon x measurement_type: edna_presence (headline, all assays) + edna_reads_dloop / edna_reads_12s (reads summed over ASVs and PCR replicates)"}
    - {table: sample_measurement, shared: true, note: "co-collected water chemistry, DNA concentration and the per-assay raw and filtered read totals (the normalisers for edna_reads_*)"}
    - {table: taxon, shared: true}
    - {table: dataset_taxon, shared: true}
    - {table: measurement_type, shared: true, note: "adds edna_reads_dloop, edna_reads_12s, edna_reads_raw_dloop, edna_reads_raw_12s, edna_reads_filtered_dloop, edna_reads_filtered_12s, oxygen_mg_l, dna_concentration; reuses edna_presence (sio_cetacean-edna)"}
  erd:
    color: "#bbe0f0"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: GBIF/OBIS Darwin Core Archive, DOI
[10.15468/n52j6r](https://www.gbif.org/dataset/10524092-539d-4ea4-8539-2e91ad8581a9)
("CalCOFI October 2022 Vertebrate eDNA"). Linked to Patin et al. 2026
(*Methods in Ecology and Evolution*, https://doi.org/10.1111/2041-210x.70370).
Water filters collected 2022-10-13 to 2022-10-18 at CalCOFI/GEMCAP stations on
the Reuben Lasker, each run through one or both of two molecular assays:
mitochondrial D-loop (cetaceans -- *Delphinus*, *Megaptera novaeangliae*) and
12S rRNA "MiFish" (fish). Source `datasetID` is `CalCOFI_Intercal` -- an
intercalibration exercise, published as a CalCOFI program product (provider
`calcofi`, not `sio`: see `provider.csv` -- this dataset is branded and
identified as CalCOFI's own, unlike e.g. `sio_mesopelagic-fish`, which is an
SIO lab collection CalCOFI merely sampled for).

**What a read count is, and is not.** `organismQuantity` is a number of DNA
sequence reads -- PCR-amplified, primer- and library-depth-dependent. It is a
semi-quantitative signal, **not abundance**, and reads from the two assays are
not comparable. So the headline `obs` row is `edna_presence` (1 = detected; the
same type `sio_cetacean-edna` uses), and the reads are published per assay
(`edna_reads_dloop`, `edna_reads_12s`) beside the per-assay filtered-read totals
in `sample_measurement` (`edna_reads_filtered_*`), which turn them into a
relative read abundance within an assay. The archive carries positive detections
only: a filter, or an assay run, with no detection is absent, so a 0 is never
emitted and absence cannot be inferred (provider question Q08). The dataset
ships as **detections only** (Ben, 2026-10-04): the absence of an `obs` row is
not a non-detection here, unlike `sio_cetacean-edna`, where a 0 is a
non-detection in a sequenced sample.

**Not a wide CSV.** This ingest reads the DwC-A's three tables directly
(`occurrence.txt`, `dnaderiveddata.txt`, `extendedmeasurementorfact.txt`,
tab-delimited) instead of `read_csv_files()`, the same way
`ingest_calcofi_ctd-cast.qmd` bypasses it for its own non-CSV source. The
source zip (`dwca-calcofi_*.zip`, GBIF's DwC-A download) lives at
`dir_data/{provider}/{dataset}/` -- `data-public/calcofi/2022-edna/` -- the
same convention as every other dataset's folder (`bottle/`, `ctd-cast/`,
`dic/`, `mets/`), not a special `genomics/` location. A second zip in that
folder (GBIF's raw occurrence-download export) is not used here -- the DwC-A
is the complete, canonical source; the other is a redundant representation
kept for reference.

**Sample grain is `parentEventID` (`sample_type` `filter`), not `eventID`.**
Every `eventID` under a given `parentEventID` shares identical
`decimalLatitude`/`decimalLongitude`/`minimumDepthInMeters`/`eventDate`
(asserted below) -- an `eventID` is one sequencing library: an assay
(`_DL` D-loop, `_MiFish` 12S) and, for D-loop, a PCR replicate (`-1`, `-2`, `-3`).
Treating `eventID` as the sample would duplicate every filter's coordinates once
per library. An `occurrenceID` is one ASV (a distinct sequence) in one library,
so the same taxon appears many times per filter (dozens of *D. delphis* ASVs in
one library); `obs` sums them to one row per filter x taxon x assay.

### Data Flow

```{mermaid}
graph LR
  A[occurrence.txt + dnaderiveddata.txt + extendedmeasurementorfact.txt] --> B[DuckDB Wrangling]
  B --> C[Validate & Clean]
  C --> D[Parquet Export]
  D --> E[GCS Archive]
  E --> F[Release Database]
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr, lubridate, purrr,
  readr, sf, stringr, tibble, tidyr, units, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_calcofi_2022-edna.qmd"))
provider     <- cc$provider    # "calcofi"
dataset      <- cc$dataset     # "2022-edna"
ds_key       <- as.character(glue("{provider}_{dataset}"))
tables_owned <- cc$tables_owned
dir_label    <- ds_key
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
dir_stage    <- cc_stage_path("parquet", dir_label, create = TRUE)
dir_ext      <- cc_stage_path("2022-edna", "unzip", create = TRUE)
dir_meta     <- here(glue("metadata/{provider}/{dataset}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))  dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

meas_type_csv <- here("metadata/measurement_type.csv")
d_meas_type   <- read_measurement_type(meas_type_csv)

# required by build_metadata_json() even when read_csv_files() is bypassed for
# the main data read -- same requirement confirmed in ingest_calcofi_ctd-cast.qmd
d_flds_rd <- read_csv(glue("{dir_meta}/flds_redefine.csv"), show_col_types = FALSE)
d_tbls_rd <- read_csv(glue("{dir_meta}/tbls_redefine.csv"), show_col_types = FALSE)
```

## Read Source Data

Reads the three DwC-A tables directly as tab-delimited text -- `read_csv_files()`
assumes one wide CSV per dataset, which this source does not fit (Occurrence
core + 2 extensions). Source is a zip (GBIF's DwC-A download); the raw zip is
archived to GCS as received (`sync_to_gcs()`, the same non-CSV-source pattern
`ingest_calcofi_ctd-cast.qmd` uses), and unzipped into the wrangling scratch
dir for reading -- the extracted `.txt` files never go to GCS themselves.

```{r}
#| label: read-source

dir_source <- path(dir_data, provider, dataset)
dwca_zips  <- dir_ls(dir_source, glob = "*dwca*.zip")
stopifnot("expected exactly one dwca-*.zip in the dataset folder" = length(dwca_zips) == 1)
dwca_zip <- dwca_zips[[1]]
cat(glue("using source zip: {path_file(dwca_zip)}"), "\n")

sync_to_gcs(
  local_dir  = dir_source,
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket     = "calcofi-files-public",
  exclude    = c(".DS_Store"))

dir_dwca <- path(dir_ext, "dwca")
dir_create(dir_dwca)
unzip(dwca_zip, exdir = dir_dwca, overwrite = TRUE)

read_dwca_txt <- function(name) {
  read_tsv(path(dir_dwca, name), col_types = cols(.default = "c"),
           na = character())
}

d_occ  <- read_dwca_txt("occurrence.txt")
d_dna  <- read_dwca_txt("dnaderiveddata.txt")
d_emof <- read_dwca_txt("extendedmeasurementorfact.txt")

cat(glue(
  "read {nrow(d_occ)} occurrence rows, {nrow(d_dna)} dnaderiveddata rows, ",
  "{nrow(d_emof)} extendedmeasurementorfact rows"), "\n")
```

## Check Data Integrity

No `read_csv_files()` result to hand `check_data_integrity()`, so integrity is
checked directly against the shape this ingest actually depends on: every
`occurrenceID` unique, every occurrence joins 1:1 to `dnaderiveddata` (the
assay source), every `parentEventID` group internally agrees on
position/time/depth (the sample-grain decision above), every read count is a
positive integer of one quantity type, and no `occurrenceStatus` other than
`present` (no absence records to filter).

```{r}
#| label: check-integrity

n_dup_occid <- d_occ |> count(occurrenceID) |> filter(n > 1) |> nrow()
stopifnot("occurrenceID must be unique in occurrence.txt" = n_dup_occid == 0)

n_unjoined <- d_occ |> anti_join(d_dna, by = "occurrenceID") |> nrow()
stopifnot("every occurrence must join to dnaderiveddata by occurrenceID" = n_unjoined == 0)

occ_status <- unique(d_occ$occurrenceStatus)
stopifnot("only 'present' occurrenceStatus expected -- an 'absent' record needs its own handling" =
            setequal(occ_status, "present"))

grain_check <- d_occ |>
  distinct(parentEventID, decimalLatitude, decimalLongitude,
           minimumDepthInMeters, maximumDepthInMeters, eventDate) |>
  count(parentEventID) |>
  filter(n > 1)
stopifnot("every parentEventID must share one position/depth/date across its eventIDs" =
            nrow(grain_check) == 0)

stopifnot("dnaderiveddata must hold one row per occurrenceID" =
            !anyDuplicated(d_dna$occurrenceID))

# every organismQuantity is a positive whole number of sequence reads -- the
# aggregation below SUMs them, so a non-numeric or a different quantity type
# would be silently mixed in
stopifnot(
  "organismQuantityType must be 'DNA sequence reads' on every occurrence" =
    setequal(unique(d_occ$organismQuantityType), "DNA sequence reads"),
  "organismQuantity must be a positive integer on every occurrence" =
    all(grepl("^[0-9]+$", d_occ$organismQuantity)) &&
    all(as.numeric(d_occ$organismQuantity) > 0))

n_parent <- n_distinct(d_occ$parentEventID)
cat(glue(
  "integrity OK: {nrow(d_occ)} occurrences (ASV x library) / {n_parent} parentEventID ",
  "filters / {n_distinct(d_occ$eventID)} eventIDs (sequencing libraries: assay x PCR ",
  "replicate); 0 duplicate occurrenceIDs; 0 unjoined to dnaderiveddata; 0 grain conflicts"), "\n")
```

## Build Sample (Filter) Table

```{r}
#| label: build-sample

d_sample <- d_occ |>
  distinct(parentEventID, decimalLatitude, decimalLongitude,
           minimumDepthInMeters, maximumDepthInMeters, eventDate,
           locationID, locality) |>
  transmute(
    parent_event_id = parentEventID,
    latitude   = as.numeric(decimalLatitude),
    longitude  = as.numeric(decimalLongitude),
    depth_min_m = as.numeric(minimumDepthInMeters),
    depth_max_m = as.numeric(maximumDepthInMeters),
    datetime_start_utc = readr::parse_datetime(eventDate, format = "%Y-%m-%dT%H:%M%z"),
    location_id = locationID,
    locality = locality,
    # Q01: every filter is from the Reuben Lasker (ship_key 'RL', NODC 33UD).
    # Set here, before add_point_geom(): never UPDATE a table holding a geom.
    ship_key = "RL")

# regression guard (commit 9b1a0d9): as_datetime() once turned every eventDate
# (an offset timestamp without seconds) into NA while the render stayed green,
# leaving datetime and cruise_key NULL for every filter
stopifnot(
  "every eventDate must parse to a UTC datetime" = !anyNA(d_sample$datetime_start_utc),
  "one edna_sample row per parentEventID" =
    nrow(d_sample) == n_parent && !anyDuplicated(d_sample$parent_event_id))

dbWriteTable(con, "edna_sample", as.data.frame(d_sample), overwrite = TRUE)
cat(glue("edna_sample: {nrow(d_sample)} filters"), "\n")
```

## Standardize Taxonomy

Every occurrence's `scientificNameID` is already a resolved WoRMS AphiaID
(`urn:lsid:marinespecies.org:taxname:<id>`) -- there is no name-matching to do,
unlike `ingest_farallon_bird-mammal.qmd`'s ERDDAP-code crosswalk. The
declaration step below extracts that id directly rather than looking it up.

```{r}
#| label: taxon-stage

mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

dbWriteTable(con, "edna_occurrence", as.data.frame(d_occ), overwrite = TRUE)

d_vocab <- dbGetQuery(con, "
  SELECT DISTINCT scientificNameID                                AS ds_taxa_code,
         scientificName                                            AS ds_scientific_name,
         CAST(regexp_extract(scientificNameID, '([0-9]+)$', 1) AS INTEGER) AS worms_id
  FROM edna_occurrence
  ORDER BY 1")
stopifnot("every occurrence must carry a WoRMS scientificNameID" =
            all(!is.na(d_vocab$worms_id)))

n_staged <- append_dataset_taxon(con, ds_key, d_vocab)

ensure_taxon_xref(con, mt_taxon, tx_over,
                  cache_csv = here("metadata/taxon_xref.csv"))
ensure_taxon_lineage(con, mt_taxon, tx_over,
                     cache_csv = here("metadata/taxon_lineage.csv"))
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- resolve_dataset_taxon(con, mt_taxon, tx_over)

obs_codes <- dbGetQuery(con, "SELECT DISTINCT scientificNameID FROM edna_occurrence")[[1]]
chk_taxon <- check_dataset_taxon(con, ds_key, codes = obs_codes, allow = character())

cat(glue("taxonomy: {n_staged} taxa staged (all WoRMS-resolved), ",
         "taxon {n_taxon}, dataset_taxon {n_ds_taxon}"), "\n")
```

## Assay and Detection Grain

The molecular assay is a property of the sequencing library (`eventID`), not of
the filter: one filter can carry D-loop and MiFish libraries. It comes from
`dnaderiveddata.target_gene`, joined by `occurrenceID`, and is checked against
the library id's own `_DL` / `_MiFish` suffix. The assay goes into the
`measurement_type` (`edna_reads_dloop`, `edna_reads_12s`) rather than an
`obs_attribute` row: an attribute row has no key back to its `obs` row, so a
filter detecting one taxon by both assays could not say which read count was
which (decision B4, 2026-10-01).

Detections are then aggregated to **one row per filter x taxon x assay**: reads
summed over the taxon's ASVs and over the filter's PCR replicate libraries for
that assay. The per-assay filtered-read totals in `sample_measurement` are summed
over the same libraries, so `edna_reads_* / edna_reads_filtered_*` is the taxon's
share of the reads the assay produced for that filter.

```{r}
#| label: assay-lookup

dbWriteTable(con, "edna_dna", as.data.frame(d_dna), overwrite = TRUE)

# assay code -> the suffix of the measurement types it emits
assay_map <- tribble(
  ~target_gene,                    ~assay,  ~library_suffix,
  "D-loop",                        "dloop", "_DL",
  "12S rRNA (SSU mitochondria)",   "12s",   "_MiFish")
dbWriteTable(con, "edna_assay_map", assay_map, overwrite = TRUE)

stopifnot("every dnaderiveddata.target_gene must be a known assay" =
            all(unique(d_dna$target_gene) %in% assay_map$target_gene))

d_assay <- dbGetQuery(con, "
  SELECT o.occurrenceID, o.parentEventID, o.eventID, o.scientificNameID,
         CAST(o.organismQuantity AS BIGINT) AS reads,
         m.assay, m.library_suffix
  FROM edna_occurrence o
  JOIN edna_dna        d USING (occurrenceID)
  JOIN edna_assay_map  m ON m.target_gene = d.target_gene")
stopifnot(
  "every occurrence must resolve an assay" = nrow(d_assay) == nrow(d_occ),
  "a library's assay must match its eventID suffix (_DL = D-loop, _MiFish = 12S)" =
    all(str_detect(d_assay$eventID, fixed(d_assay$library_suffix))))
dbWriteTable(con, "edna_assay", d_assay, overwrite = TRUE)

# filter x taxon x assay: SUM reads over ASVs and PCR replicate libraries
dbExecute(con, "
  CREATE OR REPLACE TABLE edna_detection AS
  SELECT parentEventID, scientificNameID, assay,
         SUM(reads)               AS reads,
         COUNT(*)                 AS n_occurrence,
         COUNT(DISTINCT eventID)  AS n_library
  FROM edna_assay
  GROUP BY parentEventID, scientificNameID, assay")

d_det <- dbGetQuery(con, "SELECT * FROM edna_detection")
stopifnot(
  "edna_detection must be unique per filter x taxon x assay" =
    !anyDuplicated(d_det[, c("parentEventID", "scientificNameID", "assay")]),
  "the aggregation must conserve every read" =
    sum(d_det$reads) == sum(d_assay$reads),
  "every aggregated detection must have reads > 0" = all(d_det$reads > 0))

d_assay |>
  group_by(assay) |>
  summarize(
    occurrences = n(), libraries = n_distinct(eventID),
    filters = n_distinct(parentEventID), taxa = n_distinct(scientificNameID),
    reads = sum(reads), .groups = "drop") |>
  left_join(
    d_det |> count(assay, name = "filter_x_taxon_rows"), by = "assay") |>
  datatable(caption = "Occurrences (ASV x library) aggregated to filter x taxon x assay",
            rownames = FALSE)
```

## Add Spatial

No `enforce_column_types()` step in this notebook (unlike `ingest_calcofi_ctd-cast.qmd`/
`ingest_calcofi_mets.qmd`/`ingest_cdfw_dungeness-crab.qmd`, which all call it): once
`add_point_geom()` below adds a GEOMETRY column to `edna_sample`, DuckDB >= 1.5.1
cannot rewrite that table's column types (dungeness-crab's Data Preparation section
states this explicitly, ordering its own `enforce_column_types()` call *before*
`add_point_geom()` for the same reason). `ingest_farallon_bird-mammal.qmd`,
`ingest_sio_mesopelagic-fish.qmd`, `ingest_calcofi_phyllosoma.qmd`, and
`ingest_cce-lter_euphausiids.qmd` -- every other ingest using `add_point_geom()`
on its own staging table, the pattern this notebook follows -- omit
`enforce_column_types()` entirely rather than reordering around it.

```{r}
#| label: add-spatial

load_prior_tables(
  con,
  parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  tables      = c("ship", "cruise", "grid"),
  geom_tables = c("grid"),
  as_view     = TRUE)

# Q01 RESOLVED 2026-09-26: the source carries no ship/cruise id, but the
# provider metadata (GBIF/IPT page + eml.xml acknowledgements) names the
# vessel as NOAA Ship Reuben Lasker: ship_key 'RL' (ship_nodc '33UD'), set in
# build-sample. Keyed by the shared helper's span containment (cruise-key
# skill), before add_point_geom() so no UPDATE touches the geom table.
resolve_cruise_key(con, "edna_sample", datetime_col = "datetime_start_utc",
                   require_in_cruise = TRUE) |>
  datatable(caption = "cruise_key resolution method")

add_point_geom(con, "edna_sample", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "edna_sample") |> datatable(caption = "Grid assignment")

n_sk  <- dbGetQuery(con, "SELECT COUNT(*) FROM edna_sample WHERE grid_key IS NOT NULL")[[1]]
n_ck  <- dbGetQuery(con, "SELECT COUNT(*) FROM edna_sample WHERE cruise_key IS NOT NULL")[[1]]
n_all <- dbGetQuery(con, "SELECT COUNT(*) FROM edna_sample")[[1]]
stopifnot(
  "every filter must resolve a cruise_key (one Reuben Lasker cruise, Q01)" = n_ck == n_all,
  "every filter must resolve a grid_key" = n_sk == n_all)
cat(glue("grid_key resolved: {n_sk}/{n_all}; cruise_key resolved: {n_ck}/{n_all}"), "\n")
```

## Water Chemistry + Genomics QC (sample_measurement)

`extendedmeasurementorfact.txt` repeats every value on every occurrence. Most of
its distinct `measurementType` values are FAIRe-checklist protocol text (kit
names, SOP links, extraction dates) with no place in a value/unit table. What is
kept, at two grains (both asserted below, not assumed):

- **per filter** -- the co-collected nutrients, `chl_fluor`, dissolved oxygen and
  the extracted-DNA concentration: one value per `parentEventID`. Five reuse
  canonical types with the same unit (`ammonia`, `nitrate`, `nitrite`,
  `phosphate`, `silicate`) and `chl_fluor`; oxygen is reported in mg/L, which no
  existing oxygen type uses, so it is `oxygen_mg_l` rather than a converted value
  (Q03).
- **per library** -- the raw and the filtered read totals: one value per
  `eventID`. Summed over the filter's libraries of each assay into
  `edna_reads_raw_<assay>` and `edna_reads_filtered_<assay>`; the filtered total
  is the denominator of `edna_reads_<assay>`.

One library-level value is **not** published, pending the provider (Q04): the
"Total number of OTUs or ASVs assigned to taxa", which is not an OTU count -- it
equals the library's summed `organismQuantity` (a read count) in nearly every
library. The evidence is computed below.

```{r}
#| label: emof-context

dbWriteTable(con, "edna_emof", as.data.frame(d_emof), overwrite = TRUE)

src_raw  <- "The total number of raw sequence reads for each sample. Unit = reads"
src_filt <- paste(
  "The number of filtered sequence reads for each sample, that made it through",
  "bioinformatic filtering and used for final/subsequent analyses. Unit = reads")
src_otu  <- paste(
  "Total number of OTUs or ASVs assigned to taxa based on the study-specific",
  "thresholds for each sample")

# source_unit is asserted against measurementUnit, so a mapped type cannot
# silently change unit under us; level is the grain the value repeats at
emof_type_map <- tribble(
  ~source_type,                                  ~source_unit,               ~measurement_type,     ~level,
  "Concentration of ammonia in the sample",      "umol/L",                   "ammonia",             "filter",
  "Concentration of nitrate in the sample",      "umol/L",                   "nitrate",             "filter",
  "Concentration of nitrite in the sample",      "umol/L",                   "nitrite",             "filter",
  "Concentration of phosphate in the sample",    "umol/L",                   "phosphate",           "filter",
  "Concentration of silicate in the sample",     "umol/L",                   "silicate",            "filter",
  "Chlorophyll fluorescence",                    "micrograms per litre",     "chl_fluor",           "filter",
  "Concentration of dissolved oxygen",           "mg/L",                     "oxygen_mg_l",         "filter",
  "Concentration of total DNA after extraction", "nanograms per microlitre", "dna_concentration",   "filter",
  src_raw,                                       "reads",                    "edna_reads_raw",      "library",
  src_filt,                                      "reads",                    "edna_reads_filtered", "library")
dbWriteTable(con, "emof_type_map", emof_type_map, overwrite = TRUE)

d_emof_m <- dbGetQuery(con, "
  SELECT a.parentEventID, a.eventID, a.assay,
         m.measurement_type, m.level, m.source_unit,
         e.measurementUnit                       AS unit,
         e.measurementValue                      AS value_raw,
         TRY_CAST(e.measurementValue AS DOUBLE)  AS value
  FROM edna_emof e
  JOIN edna_assay    a USING (occurrenceID)
  JOIN emof_type_map m ON m.source_type = e.measurementType")

# the FAIRe checklist's own missing-value terms are the only non-numbers allowed;
# they are absences of a value, so they emit no row (S1)
faire_missing <- c("missing", "missing: not provided")
d_nonnum <- d_emof_m |> filter(is.na(value)) |> distinct(measurement_type, parentEventID, value_raw)
stopifnot(
  "every mapped eMoF measurementType must be present in the source" =
    setequal(unique(d_emof_m$measurement_type), emof_type_map$measurement_type),
  "every mapped eMoF value must carry its expected measurementUnit" =
    all(d_emof_m$unit == d_emof_m$source_unit),
  "a non-numeric eMoF value must be a FAIRe missing-value term" =
    all(d_nonnum$value_raw %in% faire_missing))

# the grain each value repeats at is asserted, never assumed: a filter value is
# identical on every occurrence of its parentEventID, a library value on every
# occurrence of its eventID. This replaces a MIN() that hid any disagreement.
n_filter_conflict <- d_emof_m |> filter(level == "filter") |>
  group_by(parentEventID, measurement_type) |>
  summarize(n = n_distinct(value_raw), .groups = "drop") |> filter(n > 1) |> nrow()
n_library_conflict <- d_emof_m |> filter(level == "library") |>
  group_by(eventID, measurement_type) |>
  summarize(n = n_distinct(value_raw), .groups = "drop") |> filter(n > 1) |> nrow()
stopifnot(
  "a per-filter eMoF value must be identical across the filter's occurrences" =
    n_filter_conflict == 0,
  "a per-library eMoF value must be identical across the library's occurrences" =
    n_library_conflict == 0)

d_ctx_filter <- d_emof_m |>
  filter(level == "filter", !is.na(value)) |>
  distinct(parentEventID, measurement_type, measurement_value = value)

# per assay: SUM over the filter's libraries of that assay, the same libraries
# the obs read counts are summed over
d_ctx_library <- d_emof_m |>
  filter(level == "library", !is.na(value)) |>
  distinct(parentEventID, eventID, assay, measurement_type, value) |>
  group_by(parentEventID, measurement_type = paste0(measurement_type, "_", assay)) |>
  summarize(measurement_value = sum(value), .groups = "drop")

d_context <- bind_rows(d_ctx_filter, d_ctx_library)
stopifnot("edna_context must be unique per filter x measurement_type" =
            !anyDuplicated(d_context[, c("parentEventID", "measurement_type")]))
dbWriteTable(con, "edna_context", as.data.frame(d_context), overwrite = TRUE)

n_dropped_types <- dbGetQuery(con, "
  SELECT COUNT(DISTINCT measurementType) FROM edna_emof
  WHERE measurementType NOT IN (SELECT source_type FROM emof_type_map)")[[1]]
cat(glue(
  "sample_measurement staged: {nrow(d_context)} rows across ",
  "{n_distinct(d_context$measurement_type)} types and ",
  "{n_distinct(d_context$parentEventID)} filters; ",
  "{nrow(d_nonnum)} FAIRe 'missing' values emit no row ",
  "({paste(unique(d_nonnum$measurement_type), collapse = ', ')}); ",
  "{n_dropped_types} other measurementTypes not published"), "\n")
```

```{r}
#| label: emof-held-evidence

# raw >= filtered in every library, and the evidence for the value held back (Q04)
d_lib_qc <- dbGetQuery(con, glue("
  SELECT a.eventID, a.assay,
         MAX(CASE WHEN e.measurementType = '{src_raw}'  THEN TRY_CAST(e.measurementValue AS BIGINT) END) AS raw,
         MAX(CASE WHEN e.measurementType = '{src_filt}' THEN TRY_CAST(e.measurementValue AS BIGINT) END) AS filtered,
         MAX(CASE WHEN e.measurementType = '{src_otu}'  THEN TRY_CAST(e.measurementValue AS BIGINT) END) AS otu_or_asv
  FROM edna_emof e JOIN edna_assay a USING (occurrenceID)
  GROUP BY a.eventID, a.assay")) |>
  left_join(
    d_assay |> group_by(eventID) |> summarize(reads_published = sum(reads)),
    by = "eventID")

n_lib          <- nrow(d_lib_qc)
n_otu_eq_reads <- sum(d_lib_qc$otu_or_asv == d_lib_qc$reads_published)
n_otu_ge_reads <- sum(d_lib_qc$otu_or_asv >= d_lib_qc$reads_published)
stopifnot(
  "every library must carry a raw and a filtered read total" =
    !anyNA(d_lib_qc$raw) && !anyNA(d_lib_qc$filtered),
  "filtered reads can never exceed raw reads" = all(d_lib_qc$filtered <= d_lib_qc$raw),
  "published reads can never exceed the library's filtered reads" =
    all(d_lib_qc$reads_published <= d_lib_qc$filtered))
cat(glue(
  "raw >= filtered >= published reads in all {n_lib} libraries"), "\n")
cat(glue(
  "'OTUs or ASVs assigned to taxa' equals the library's summed organismQuantity in ",
  "{n_otu_eq_reads} of {n_lib} libraries (>= it in {n_otu_ge_reads}); range ",
  "{min(d_lib_qc$otu_or_asv)}-{max(d_lib_qc$otu_or_asv)} -- a read count, not ",
  "richness; not published (Q04)"), "\n")

d_lib_qc |>
  arrange(assay, eventID) |>
  datatable(caption = "Per-library read totals (otu_or_asv held back, Q04)",
            rownames = FALSE)
```

## Declare and Enforce Physical Bounds

Every ingest that emits measurements checks them against
`metadata/measurement_type.csv`'s declared bounds (measurement-bounds skill):
here on `edna_context` before it becomes `sample_measurement`, and again on the
emitted `obs` after Emit Core Tables. Every type this dataset emits declares at
least `valid_min = 0`, so an `undeclared` finding is a hard stop rather than a
note; an `out_of_range` one is enforced by `drop_out_of_bounds()` and reported.

```{r}
#| label: bounds-check

d_bounds <- check_measurement_bounds(con, "edna_context", mt = d_meas_type)
bounds_datatable(d_bounds)
stopifnot("every sample_measurement type must declare a bound" =
            !any(d_bounds$status == "undeclared"))
```

```{r}
#| label: bounds-enforce

oob_tally <- drop_out_of_bounds(con, "edna_context", mt = d_meas_type)
cat(glue("out-of-bounds rows dropped from edna_context: {sum(oob_tally$n_bad)}"), "\n")
```

## Schema Documentation

Color scheme: lightblue = new dataset tables, lightgreen = taxon staging,
lightyellow = amended reference tables, white = shared metadata.

```{r}
#| label: schema

edna_rels <- list(
  primary_keys = list(
    edna_sample    = "parent_event_id",
    edna_assay     = "occurrenceID",
    edna_detection = c("parentEventID", "scientificNameID", "assay"),
    edna_context   = c("parentEventID", "measurement_type"),
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "edna_assay", column = "parentEventID",
         ref_table = "edna_sample", ref_column = "parent_event_id"),
    list(table = "edna_detection", column = "parentEventID",
         ref_table = "edna_sample", ref_column = "parent_event_id"),
    list(table = "edna_context", column = "parentEventID",
         ref_table = "edna_sample", ref_column = "parent_event_id"),
    list(table = "edna_context", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(
  con,
  tables = c("edna_sample", "edna_assay", "edna_detection", "edna_context",
             "measurement_type", "dataset"),
  rels   = edna_rels,
  colors = list(
    lightblue   = c("edna_sample", "edna_assay", "edna_detection", "edna_context"),
    lightyellow = "measurement_type",
    white       = "dataset"))
```

## Emit Core Tables

Projection lives here, in the notebook, per the no-`switch(dataset_key, ...)`
rule -- no `calcofi4db` change needed for this dataset. `sample_key` is
`calcofi_2022-edna:filter:<parentEventID>` (`dataset_key:sample_type:id`). There
is no `obs_attribute`: the assay lives in the measurement type.

```{r}
#| label: emit-core

# 1. sample -- one row per physical water filter (parentEventID grain)
append_sample(con, glue("
  SELECT {ns_key(ds_key, 'filter', 'parent_event_id')} AS sample_key,
         'filter' AS sample_type, NULL::VARCHAR AS parent_sample_key,
         {ns_key(ds_key, 'filter', 'parent_event_id')} AS root_sample_key,
         '{ds_key}' AS dataset_key, grid_key, NULL::VARCHAR AS site_key, cruise_key,
         NULL::INTEGER AS order_occ,
         latitude, longitude, datetime_start_utc AS datetime,
         depth_min_m, depth_max_m, NULL::VARCHAR AS tow_type
  FROM edna_sample"))

# 2. obs -- bound positionally: realm, dataset_key, sample_key, grid_key,
# cruise_key, latitude, longitude, datetime, depth_min_m, depth_max_m, taxon_key,
# life_stage, measurement_type, measurement_value, measurement_qual,
# measurement_prec. life_stage is NULL: a water-filter detection has none.
#   (a) edna_presence -- the cross-dataset headline, the same type and meaning as
#       sio_cetacean-edna: 1 = the taxon's DNA was detected (reads > 0 after the
#       provider's own filtering) by any assay in this filter. Positive
#       detections only, so never 0 here: no row is NOT a non-detection (Q08).
#   (b) edna_reads_<assay> -- reads summed over ASVs and PCR replicate libraries
#       (edna_detection), one row per filter x taxon x assay.
obs_cols <- glue("
  '{ds_key}', {ns_key(ds_key, 'filter', 'd.parentEventID')},
  s.grid_key, s.cruise_key, s.latitude, s.longitude, s.datetime_start_utc,
  s.depth_min_m, s.depth_max_m, dt.taxon_key, NULL::VARCHAR")
obs_from <- glue("
  JOIN edna_sample s ON s.parent_event_id = d.parentEventID
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = d.scientificNameID")

append_obs(con, glue("
  SELECT 'bio', {obs_cols},
         'edna_presence', 1::DOUBLE, NULL::VARCHAR, NULL::DOUBLE
  FROM (SELECT DISTINCT parentEventID, scientificNameID
        FROM edna_detection WHERE reads > 0) d
  {obs_from}"))

append_obs(con, glue("
  SELECT 'bio', {obs_cols},
         'edna_reads_' || d.assay, CAST(d.reads AS DOUBLE), NULL::VARCHAR, NULL::DOUBLE
  FROM edna_detection d
  {obs_from}"))

# 3. sample_measurement -- co-collected water chemistry, DNA concentration and
# the per-assay filtered-read totals; a FAIRe 'missing' emitted no row upstream
append_sample_measurement(con, glue("
  SELECT {ns_key(ds_key, 'filter', 'parentEventID')}, '{ds_key}',
         measurement_type, measurement_value, NULL::VARCHAR
  FROM edna_context
  WHERE measurement_value IS NOT NULL"))

core <- list(
  sample             = dbGetQuery(con, glue("SELECT COUNT(*) FROM sample WHERE dataset_key = '{ds_key}'"))[[1]],
  obs                = dbGetQuery(con, glue("SELECT COUNT(*) FROM obs WHERE dataset_key = '{ds_key}'"))[[1]],
  sample_measurement = dbGetQuery(con, glue("SELECT COUNT(*) FROM sample_measurement WHERE dataset_key = '{ds_key}'"))[[1]])

# the grain contract, asserted -------------------------------------------------
d_obs_chk <- dbGetQuery(con, glue("
  SELECT sample_key, taxon_key, measurement_type, measurement_value
  FROM obs WHERE dataset_key = '{ds_key}'"))
n_presence <- sum(d_obs_chk$measurement_type == "edna_presence")
stopifnot(
  "one sample row per filter" = core$sample == n_parent,
  "every sample_key must be dataset_key:filter:id" =
    dbGetQuery(con, glue("
      SELECT COUNT(*) FROM sample
      WHERE dataset_key = '{ds_key}'
        AND (sample_type <> 'filter' OR sample_key NOT LIKE '{ds_key}:filter:%')"))[[1]] == 0,
  "every eDNA observation must resolve a taxon_key" = !anyNA(d_obs_chk$taxon_key),
  "obs must be unique per (sample_key, taxon_key, measurement_type)" =
    !anyDuplicated(d_obs_chk[, c("sample_key", "taxon_key", "measurement_type")]),
  "one edna_presence row per filter x taxon detected" =
    n_presence == nrow(distinct(d_det, parentEventID, scientificNameID)),
  "edna_presence is 1 on every row (positive detections only)" =
    all(d_obs_chk$measurement_value[d_obs_chk$measurement_type == "edna_presence"] == 1),
  "one edna_reads_* row per filter x taxon x assay" =
    sum(startsWith(d_obs_chk$measurement_type, "edna_reads_")) == nrow(d_det),
  "obs must carry every published read" =
    sum(d_obs_chk$measurement_value[startsWith(d_obs_chk$measurement_type, "edna_reads_")]) ==
      sum(d_assay$reads),
  "no NULL measurement_value in obs" = !anyNA(d_obs_chk$measurement_value),
  "no NULL measurement_value in sample_measurement" =
    dbGetQuery(con, glue("
      SELECT COUNT(*) FROM sample_measurement
      WHERE dataset_key = '{ds_key}' AND measurement_value IS NULL"))[[1]] == 0)

d_core_types <- dbGetQuery(con, glue("
  SELECT 'obs' AS tbl, measurement_type, COUNT(*) AS n
  FROM obs WHERE dataset_key = '{ds_key}' GROUP BY ALL
  UNION ALL
  SELECT 'sample_measurement', measurement_type, COUNT(*)
  FROM sample_measurement WHERE dataset_key = '{ds_key}' GROUP BY ALL
  ORDER BY 1, 2"))
d_core_types |> datatable(caption = "Rows per measurement_type emitted", rownames = FALSE)

cat(glue(
  "core projection -- sample={core$sample} obs={core$obs} ",
  "({n_presence} edna_presence + {core$obs - n_presence} edna_reads_*) ",
  "sample_measurement={core$sample_measurement}; no obs_attribute"), "\n")
```

```{r}
#| label: bounds-obs

# S5: bounds on what is actually published, not only on the staging table
d_bounds_obs <- check_measurement_bounds(con, "obs", mt = d_meas_type, dataset_key = ds_key)
bounds_datatable(d_bounds_obs)
stopifnot(
  "every obs type must declare a bound" = !any(d_bounds_obs$status == "undeclared"),
  "no obs value may lie outside its declared bound" =
    !any(d_bounds_obs$status == "out_of_range"))
```

## Measurement Type Registry

The new types were added to `metadata/measurement_type.csv` once, with
`register_measurement_types()`, not by this
notebook: rewriting the shared registry on every render is a side effect on a
file other ingests edit. The notebook asserts the contract in both directions --
every type it emits is registered, and every registered type that names this
dataset in `_source_datasets` has at least one row.

```{r}
#| label: measurement-types

dbWriteTable(con, "measurement_type", as.data.frame(d_meas_type), overwrite = TRUE)

emitted    <- unique(d_core_types$measurement_type)
registered <- d_meas_type |>
  filter(map_lgl(str_split(coalesce(`_source_datasets`, ""), ";\\s*"), ~ ds_key %in% .x)) |>
  pull(measurement_type)

unregistered <- setdiff(emitted, d_meas_type$measurement_type)
unclaimed    <- setdiff(emitted, registered)
empty        <- setdiff(registered, emitted)
stopifnot(
  "every emitted measurement_type must be in metadata/measurement_type.csv" =
    length(unregistered) == 0,
  "every emitted measurement_type must list this dataset in _source_datasets" =
    length(unclaimed) == 0,
  "every registered type naming this dataset must have >= 1 row (an empty type would ship)" =
    length(empty) == 0)
cat(glue("measurement types: {length(emitted)} emitted, all registered and claimed; ",
         "0 registered-but-empty"), "\n")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot("this dataset must appear in the ingest YAML registry" = nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)
```

## Questions for Data Providers

```{r}
#| label: provider-questions

questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the calcofi 2022-edna data providers (ranked)")
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)

# validate_for_release()'s null check treats every *_id/_key column as required,
# a heuristic rather than this dataset's contract. Every nullable case is
# declared here with its count and reason (the cdfw_dungeness-crab idiom);
# anything not on the list, or whose count moved, is a hard failure.
nullable <- tribble(
  ~table,   ~column,             ~n,  ~reason,
  "sample", "parent_sample_key", 47L, "a filter is a root event: no link yet to the bottle/cast it came from (provider question Q05)",
  "sample", "site_key",          47L, "no station match yet; locationID names the CalCOFI station but is not parsed (Q05)",
  "taxon",  "itis_id",           12L, "WoRMS links no ITIS TSN for these lineage nodes",
  "taxon",  "gbif_id",           79L, "nothing asks GBIF -- no GBIF crosswalk in the xref",
  "taxon",  "ncbi_id",           79L, "never populated by ANY dataset",
  "taxon",  "inat_id",           79L, "never populated by ANY dataset",
  "taxon",  "parent_taxon_key",  1L,  "the root of the lineage (Biota) has no parent")

nulls <- results$checks |>
  filter(check == "nulls", status == "error") |>
  mutate(column = str_match(message, "column '([^']+)'")[, 2],
         n      = as.integer(str_match(message, "has (\\d+) NULL")[, 2])) |>
  select(table, column, n)

reconciled  <- full_join(nulls, nullable |> select(table, column, n_expected = n),
                         by = c("table", "column"))
unexplained <- reconciled |> filter(is.na(n_expected))
moved       <- reconciled |> filter(!is.na(n_expected), !is.na(n), n != n_expected)
resolved    <- reconciled |> filter(!is.na(n_expected), is.na(n))
other_fail  <- results$checks |> filter(check != "nulls", status == "error")
if (nrow(unexplained)) print(unexplained)
if (nrow(moved))       print(moved)
if (nrow(other_fail))  print(other_fail)
stopifnot(
  "a NULL appeared in a column with no declared reason" = nrow(unexplained) == 0,
  "a declared NULL count has changed -- the data moved, re-derive the reason" = nrow(moved) == 0,
  "a validate_for_release() check other than nulls failed" = nrow(other_fail) == 0)
if (nrow(resolved))
  cat(glue("{nrow(resolved)} declared nullable case(s) no longer occur -- prune them: ",
           "{paste(resolved$table, resolved$column, sep = '.', collapse = ', ')}"), "\n")
if (length(results$warnings) > 0)
  cat("Warnings:\n", paste("-", results$warnings, collapse = "\n"), "\n")

cat(glue("validate_for_release: {ifelse(results$passed, 'PASSED', 'FAILED')} on its null ",
         "heuristic; {nrow(nulls)} null case(s) reconciled against {nrow(nullable)} declared, ",
         "0 unexplained, 0 other failures"), "\n")
```

## Data Preview

```{r}
#| label: preview-sample
dbGetQuery(con, "SELECT * FROM sample WHERE dataset_key = 'calcofi_2022-edna'") |>
  datatable(caption = "sample", rownames = FALSE)
```

```{r}
#| label: preview-obs
dbGetQuery(con, "SELECT * FROM obs WHERE dataset_key = 'calcofi_2022-edna' LIMIT 50") |>
  datatable(caption = "obs (first 50)", rownames = FALSE)
```

```{r}
#| label: preview-sample-measurement
dbGetQuery(con, "SELECT * FROM sample_measurement WHERE dataset_key = 'calcofi_2022-edna'") |>
  datatable(caption = "sample_measurement", rownames = FALSE)
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
write_parquet_outputs(con, dir_parquet, tables = tbls_out)
```

## Write Metadata

```{r}
#| label: write-metadata

metadata_path <- build_metadata_json(
  con                  = con,
  d_tbls_rd            = d_tbls_rd,
  d_flds_rd            = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           glue("{dir_meta}/metadata_derived.csv")),
  output_dir           = dir_parquet,
  tables               = tbls_out,
  provider             = provider,
  dataset              = dataset,
  workflow_url         = cc$workflow_url,
  tables_owned         = tables_owned)

build_relationships_json(
  rels       = core_relationships(tbls_out),
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset)
```

## Upload to GCS

```{r}
#| label: upload-gcs

# local_dir/sidecar_dir split matches ingest_sio_mesopelagic-fish.qmd and
# ingest_calcofi_ctd-cast.qmd exactly -- both pass dir_stage as local_dir and
# dir_parquet as sidecar_dir here, even though every write above (parquet
# tables, metadata.json, relationships.json) targets dir_parquet.
# delete_stale: the 2026-10-01 grain change dropped obs_attribute from this
# shard, and a stale obs_attribute.parquet must not linger under the prefix
# (sidecars are protected from the delete by sync_to_gcs() itself)
sync_to_gcs(
  local_dir    = dir_stage,
  sidecar_dir  = dir_parquet,
  gcs_prefix   = glue("ingest/{dir_label}"),
  bucket       = "calcofi-db",
  delete_stale = TRUE)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("parquet outputs written to: {dir_parquet}"), "\n")
```
