---
title: "Ingest CCE-LTER Particulate Organic Carbon and Nitrogen"
calcofi:
  target_name: ingest_cce_lter_poc_pon
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
    - ingest_calcofi_bottle
  output: data/parquet/cce-lter_poc-pon/manifest.json
  provider: cce-lter
  dataset: poc-pon
  workflow_url: https://calcofi.io/workflows/ingest_cce-lter_poc-pon.html
  questions_file: metadata/cce-lter/poc-pon/questions.csv
  # staged, not released: first ingest of this holding (CalCOFI/workflows#84).
  # Remove after review, the same way the cetacean ingests were un-held.
  in_release: false
  dataset_meta:
    dataset_name: CCE-LTER Particulate Organic Carbon and Nitrogen (CalCOFI Cruises)
    dataset_name_short: "POC / PON"
    category: "Nutrients & Chemistry"
    color: "#6b4226"
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (one Niskin bottle water sample per row; own sample grain)"}
    - {table: obs, shared: true, note: "core observation headline (env realm; poc + pon in umol/L)"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#ead8c8"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: EDI package `knb-lter-cce.54.10` (Goericke), entity
*ParticulateOrganicCarbonandNitrogen* (`table_54.csv`): 17,413 rows, 18 columns,
73 cruises from 2004-11 to 2023-04 (the package title still says "November
2022", Q05). Fetched by `libs/download_poc_pon.R`, revision and md5 pinned.
Licence: CC0 1.0 (the EML's `intellectualRights`).

- **Provider**: `cce-lter` (same as `iron`, `euphausiids`, `picoplankton-bacteria`).
- **Grain**: one row per Niskin bottle water sample, filtered for POC/PON. The
  source's own `(studyName, Cast Number, Bottle Number)` is the natural key,
  asserted unique after the clean-up below.
- **Realm**: `env`. Two measurements are published per sample: `poc`
  (`umol/L_C`) and `pon` (`umol/L_N`), both in µmol/L.
- **Not published** (derivable or held, kept in the archived source):
  `ug_CperL` and `ug_NperL` (the same values in µg/L; the notebook asserts the
  conversion), `C/N ratio` and `C/N molar ratio` (ratios of the published values,
  also asserted), `Associated Bottle Number` (meaning unconfirmed, Q03) and
  `Notes` (free text on 12 rows, Q06).
- **Cleaned before publishing** (each counted and asserted below):
  - 4 rows that repeat another row in every column are dropped (2019-11 cast 25,
    Q02).
  - 2 rows share a sample key but give different depths (2005-11 cast 25 bottle
    20, 1 m vs 2 m, identical values) and are held out until Q02 is answered.
  - 10 rows with no depth are held out (Q04): depth is a coordinate in the core.
- **Cruise key**: from the ship code in `studyName` plus the source's own
  `Cruise` designation (YYYYMM), resolved by `derive_cruise_key_on_casts()`
  (span, then source designation, then month).
- **Bottle match** (evidence only): each sample is matched to its
  `calcofi_bottle` cast by `site_key` within 1 day, then to the Niskin bottle
  nearest its depth within 2 m. The match is reported, not published, and not set
  as `parent_sample_key`, following the cetacean eDNA decision (Ben,
  2026-10-04): every sample is a root.

```{mermaid}
graph LR
  A[EDI knb-lter-cce.54.10<br/>table_54.csv, 17,413 rows] --> B[drop exact duplicates<br/>hold out conflicting key + no depth]
  B --> C[poc_sample<br/>one row per bottle sample]
  B --> D[poc_measurement<br/>long: poc, pon]
  C -.ship code + Cruise.-> E[(cruise_key)]
  C -.site_key + date + depth.-> F[(calcofi_bottle<br/>evidence only)]
  C --> G[[core: sample]]
  D --> H[[core: obs, realm=env]]
```

## Setup

```{r}
#| label: setup
devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))
cc             <- read_calcofi_meta(here("ingest_cce-lter_poc-pon.qmd"))
provider       <- cc$provider
dataset        <- cc$dataset
tables_owned   <- cc$tables_owned
ds_key         <- as.character(glue("{provider}_{dataset}"))
dir_label      <- ds_key
dir_parquet    <- here(glue("data/parquet/{dir_label}"))
dir_stage      <- cc_stage_path("parquet", dir_label, create = TRUE)
dir_src        <- cc_stage_path("source", dir_label, create = TRUE)
db_path        <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_meta       <- here(glue("metadata/{provider}/{dataset}"))
publish_to_gcs <- Sys.getenv("CALCOFI_SKIP_GCS") == ""

# poc and pon are registered in metadata/measurement_type.csv by this change
# (valid_min = 0; valid_max proposed in Q01, not declared)
d_meas_type <- read_measurement_type(here("metadata/measurement_type.csv"))
stopifnot(
  "poc and pon must be pre-registered" =
    all(c("poc", "pon") %in% d_meas_type$measurement_type))

if (overwrite) {
  if (file_exists(db_path)) file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp"))) dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Read Source Data

Pulled from EDI by `libs/download_poc_pon.R` (package `knb-lter-cce.54.10`,
entity *ParticulateOrganicCarbonandNitrogen*), revision 10 with the EML's md5
asserted. Bulk input lives under the stage dir (`cc_stage_path("source", ...)`),
never in Drive or the git tree.

The real headers carry unit suffixes with non-ASCII characters (`Latitude (º)`,
`ug_CperL (µg/L)`). They are checked against the expected header (written with
`\u` escapes so this file stays ASCII) and then renamed **by position**, so the
notebook does not depend on the session's text encoding.

```{r}
#| label: read-source
source(here("libs/download_poc_pon.R"))
downloaded <- overwrite_all || !file_exists(file.path(dir_src, "data.csv"))
src_csv <- download_poc_pon(dir_src, overwrite = overwrite_all)
stopifnot("POC/PON data.csv not found" = file_exists(src_csv))
src_stamp <- stamp_source_access(files = src_csv) |>
  mutate(source = "https://doi.org/10.6073/pasta/830ff22e75a9de5b493af48330e5fc4c",
         method = if (downloaded) "download" else "file_mtime")

if (publish_to_gcs) {
  sync_to_gcs(
    local_dir  = dir_src,
    gcs_prefix = glue("archive/{provider}/{dataset}"),
    bucket     = "calcofi-files-public",
    exclude    = c(".DS_Store", "*.tmp", "*.gdoc"))
} else {
  cat("CALCOFI_SKIP_GCS set -- source NOT archived to gs://calcofi-files-public\n")
}

d_raw <- read_csv(src_csv, col_types = cols(.default = "c"), na = character(),
                  locale = locale(encoding = "UTF-8"))

hdr_expected <- c(
  "studyName", "Cruise", "Datetime GMT", "Latitude (º)", "Longitude (º)",
  "Line", "Station", "Cast Number", "Bottle Number", "Associated Bottle Number",
  "Depth (m)", "ug_CperL (µg/L)", "ug_NperL (µg/L)",
  "umol/L_N (µmol/L)", "umol/L_C (µmol/L)", "C/N ratio",
  "C/N molar ratio", "Notes")
hdr_new <- c(
  "study_name", "cruise_ym", "datetime_gmt", "latitude", "longitude",
  "line", "station", "cast_number", "bottle_number", "associated_bottle_number",
  "depth_m", "poc_ug_l", "pon_ug_l", "pon_umol_l", "poc_umol_l", "cn_ratio",
  "cn_molar_ratio", "notes")

cat(glue("Read {format(nrow(d_raw), big.mark=',')} rows, {ncol(d_raw)} columns"), "\n")
stopifnot(
  "expected 17,413 rows x 18 columns (EDI revision 10, checked 2026-10-06)" =
    nrow(d_raw) == 17413 && ncol(d_raw) == 18,
  "source header changed: re-check flds_redefine.csv before renaming by position" =
    identical(names(d_raw), hdr_expected))
names(d_raw) <- hdr_new
```

## Clean, Type-Cast

Three clean-up steps, each counted and asserted against what the source held on
2026-10-06, so a new revision that changes any of them stops the render:

1. **Exact duplicates.** 4 rows repeat another row in every column (2019-11
   cast 25, bottles 16, 17, 20, 22). The copy is dropped (Q02).
2. **Conflicting key.** After that, one key is still listed twice: 2005-11 cast
   25 bottle 20, at 2 m and at 1 m with identical values. Both rows are held out
   (not published) until the provider says which depth is right (Q02).
3. **No depth.** 10 rows have an empty `Depth (m)`. They are held out (Q04):
   depth is a coordinate in the core and is never guessed.

Every measurement is blank-or-numeric with no sentinel: the EML declares `-999`
for Depth, and the file uses none (asserted).

```{r}
#| label: clean-cast
num <- function(x) suppressWarnings(as.numeric(if_else(x == "", NA_character_, x)))

stopifnot("no -999 sentinel in this revision (the EML declares it for Depth)" =
            !any(as.matrix(d_raw) == "-999"))

# 1. exact duplicates
n_raw   <- nrow(d_raw)
d_dedup <- distinct(d_raw)
n_exact <- n_raw - nrow(d_dedup)
stopifnot("4 exact duplicate rows expected (2019-11 cast 25, Q02)" = n_exact == 4)

d_clean <- d_dedup |>
  transmute(
    study_name,
    cruise_ym,
    datetime_start_utc = suppressWarnings(ymd_hms(datetime_gmt, tz = "UTC")),
    latitude           = num(latitude),
    longitude          = num(longitude),
    line               = num(line),
    station            = num(station),
    cast_number        = suppressWarnings(as.integer(cast_number)),
    bottle_number      = suppressWarnings(as.integer(bottle_number)),
    associated_bottle_number = suppressWarnings(as.integer(associated_bottle_number)),
    depth_m            = num(depth_m),
    poc_umol_l         = num(poc_umol_l),
    pon_umol_l         = num(pon_umol_l),
    poc_ug_l           = num(poc_ug_l),
    pon_ug_l           = num(pon_ug_l),
    cn_ratio           = num(cn_ratio),
    cn_molar_ratio     = num(cn_molar_ratio),
    notes              = if_else(notes == "", NA_character_, notes)) |>
  mutate(
    site_key      = normalize_site_key(sprintf("%.1f %.1f", line, station)),
    # the source's own natural key; '-' not ':' so the id stays one token
    sample_key_id = paste(study_name, cast_number, bottle_number, sep = "-"))

stopifnot(
  "every row parses a datetime, position, line/station, cast and bottle" =
    !anyNA(d_clean$datetime_start_utc) && !anyNA(d_clean$latitude) &&
    !anyNA(d_clean$longitude) && !anyNA(d_clean$site_key) &&
    !anyNA(d_clean$cast_number) && !anyNA(d_clean$bottle_number),
  "Cruise is a YYYYMM designation on every row" =
    all(str_detect(d_clean$cruise_ym, "^\\d{4}(0[1-9]|1[0-2])$")))

# 2. a key still listed twice after dropping exact duplicates -> hold out
dup_keys <- d_clean |> count(sample_key_id) |> filter(n > 1) |> pull(sample_key_id)
stopifnot("one conflicting key expected (2005-11 cast 25 bottle 20, Q02)" =
            identical(dup_keys, "2005-11-04-C-32NM-25-20"))

# 3. no depth -> hold out
d_clean <- d_clean |>
  mutate(hold_reason = case_when(
    sample_key_id %in% dup_keys ~ "conflicting depth for one key (Q02)",
    is.na(depth_m)              ~ "no depth (Q04)",
    .default                    = NA_character_))
d_hold <- d_clean |> filter(!is.na(hold_reason))
stopifnot(
  "12 rows held out: 2 conflicting-key + 10 no-depth" =
    nrow(d_hold) == 12 &&
    sum(d_hold$hold_reason == "no depth (Q04)") == 10)

d_sample <- d_clean |>
  filter(is.na(hold_reason)) |>
  select(-hold_reason) |>
  arrange(study_name, cast_number, bottle_number) |>
  mutate(sample_id = row_number())
stopifnot("sample_key_id is unique" = !anyDuplicated(d_sample$sample_key_id))

cat(glue(
  "{format(n_raw, big.mark=',')} rows read; {n_exact} exact duplicates dropped; ",
  "{nrow(d_hold)} held out; {format(nrow(d_sample), big.mark=',')} samples on ",
  "{n_distinct(d_sample$study_name)} cruises"), "\n")
d_hold |>
  select(study_name, cast_number, bottle_number, depth_m, poc_umol_l, pon_umol_l, hold_reason) |>
  datatable(caption = "Rows held out of the core (counted, asserted)", rownames = FALSE)

dbWriteTable(con, "poc_sample", d_sample, overwrite = TRUE)
```

### Derived columns agree with the published ones

The µg/L columns and both ratios are not published because they follow from
`poc` and `pon`. That is only safe if they really do, so it is checked here: µg/L
is µmol/L times the atomic weight (C 12.0107, N 14.0067), and the ratios are
`C/N` in µg and in µmol. The EML's definition of both ratio columns reads
"Nitrogen:Carbon ratio"; the values are C:N (Q07).

```{r}
#| label: derived-check
d_chk <- d_sample |>
  mutate(
    rel_c  = abs(poc_ug_l / 12.0107 - poc_umol_l) / poc_umol_l,
    rel_n  = abs(pon_ug_l / 14.0067 - pon_umol_l) / pon_umol_l,
    rel_cn  = abs(poc_ug_l / pon_ug_l - cn_ratio) / cn_ratio,
    rel_cnm = abs(poc_umol_l / pon_umol_l - cn_molar_ratio) / cn_molar_ratio)
chk <- d_chk |> summarize(across(starts_with("rel_"), ~ max(.x, na.rm = TRUE)))
chk |> datatable(caption = "Largest relative difference, derived vs published", rownames = FALSE)
stopifnot(
  "ug/L = umol/L x atomic weight, within 0.1%" = chk$rel_c < 1e-3 && chk$rel_n < 1e-3,
  "C/N ratios follow from the published values, within 0.1%" =
    chk$rel_cn < 1e-3 && chk$rel_cnm < 1e-3)
```

## Pivot Measurements to Long Format

```{r}
#| label: pivot-long
poc_meas_map <- tribble(
  ~measurement_type, ~value_col,
  "poc",             "poc_umol_l",
  "pon",             "pon_umol_l")

sql_parts <- purrr::pmap_chr(poc_meas_map, function(measurement_type, value_col) {
  glue(
    "SELECT sample_id, '{measurement_type}' AS measurement_type,
     CAST({value_col} AS DOUBLE) AS measurement_value
     FROM poc_sample WHERE {value_col} IS NOT NULL")
})
dbExecute(con, glue(
  "CREATE OR REPLACE TABLE poc_measurement AS
   SELECT ROW_NUMBER() OVER (ORDER BY sample_id, measurement_type) AS measurement_id, *
   FROM (
     {paste(sql_parts, collapse = '\nUNION ALL\n')}
   ) sub"))

d_n <- dbGetQuery(con, "
  SELECT measurement_type, COUNT(*) AS n FROM poc_measurement GROUP BY 1 ORDER BY 1")
d_n |> datatable(caption = "Rows per measurement_type", rownames = FALSE)
# one sample has no carbon value, two have no nitrogen value (blank in the source)
stopifnot(
  "poc: one value per sample with a carbon value" =
    d_n$n[d_n$measurement_type == "poc"] == sum(!is.na(d_sample$poc_umol_l)),
  "pon: one value per sample with a nitrogen value" =
    d_n$n[d_n$measurement_type == "pon"] == sum(!is.na(d_sample$pon_umol_l)))
```

## Resolve Ship and Cruise Keys

Same approach as the iron and picoplankton-bacteria notebooks: the trailing
token of `studyName` is the ship's NODC code (`2004-11-02-C-33RR` -> `33RR`).
Unlike iron, this source also ships the cruise's own `Cruise` designation
(YYYYMM), which differs from the start month in `studyName` on 3,023 rows (a
cruise that starts late in the month before its designated one), so it is passed
as `cruise_ym_col`.

```{r}
#| label: resolve-keys
load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  tables = c("ship", "cruise"), as_view = TRUE)

dbExecute(con, "ALTER TABLE poc_sample ADD COLUMN IF NOT EXISTS ship_code VARCHAR")
dbExecute(con, "UPDATE poc_sample SET ship_code = regexp_extract(study_name, '[0-9A-Z]+$')")

cruise_key_result <- derive_cruise_key_on_casts(
  con, table_name = "poc_sample", datetime_col = "datetime_start_utc",
  cruise_ym_col = "cruise_ym")
cruise_key_result$key_stats |>
  datatable(caption = "cruise_key resolution: span > source designation > month")

n_ck  <- dbGetQuery(con, "SELECT COUNT(*) FROM poc_sample WHERE cruise_key IS NOT NULL")[[1]]
n_all <- dbGetQuery(con, "SELECT COUNT(*) FROM poc_sample")[[1]]
cat(glue("cruise_key match: {n_ck}/{n_all} ({round(100*n_ck/n_all,1)}%)"), "\n")
```

## Match to calcofi_bottle (evidence only)

Each sample is matched to the `calcofi_bottle` cast at the same `site_key`
within 1 day (`match_by_site_datetime()`), then to the Niskin bottle on that
cast nearest its depth within 2 m (`match_nearest_by_depth()`). The match rate is
reported, not asserted: the bottle database may not yet hold the most recent
cruises. The matched bottle is kept on `poc_sample` (`bottle_sample_key`) as
evidence only. It is not published and not set as `parent_sample_key`.

```{r}
#| label: match-bottle
bottle_parquet <- cc_stage_path("parquet", "calcofi_bottle", "sample.parquet")
stopifnot("calcofi_bottle's sample shard is staged (run ingest_calcofi_bottle first)" =
            file_exists(bottle_parquet))

dbExecute(con, glue(
  "CREATE OR REPLACE TABLE bottle_cast AS
   SELECT sample_key AS cast_sample_key, site_key, cruise_key, datetime
   FROM read_parquet('{bottle_parquet}')
   WHERE sample_type = 'cast' AND site_key IS NOT NULL AND datetime IS NOT NULL"))
dbExecute(con, glue(
  "CREATE OR REPLACE TABLE bottle_niskin AS
   SELECT sample_key AS bottle_sample_key, parent_sample_key AS cast_sample_key,
          depth_max_m AS depth_m
   FROM read_parquet('{bottle_parquet}')
   WHERE sample_type = 'bottle' AND depth_max_m IS NOT NULL"))

cast_match <- match_by_site_datetime(
  con, "poc_sample", "bottle_cast",
  fk_col = "cast_sample_key", ref_pk = "cast_sample_key",
  key_col = "site_key", datetime_col = "datetime_start_utc",
  ref_datetime_col = "datetime", window_days = 1)
niskin_match <- match_nearest_by_depth(
  con, "poc_sample", "bottle_niskin",
  fk_col = "bottle_sample_key", ref_pk = "bottle_sample_key",
  parent_fk = "cast_sample_key", axis_col = "depth_m", tolerance = 2)

cat(glue(
  "bottle cast matched: {cast_match$matched}/{cast_match$total} ({cast_match$pct}%); ",
  "Niskin bottle within 2 m: {niskin_match$matched}/{niskin_match$eligible} ",
  "({niskin_match$pct}%)"), "\n")

dbGetQuery(con, "
  SELECT cruise_ym, COUNT(*) AS n_samples,
         COUNT(cast_sample_key) AS n_cast_matched,
         COUNT(bottle_sample_key) AS n_bottle_matched
  FROM poc_sample GROUP BY 1 ORDER BY 1") |>
  datatable(caption = "Bottle match by cruise (evidence only; not published)", rownames = FALSE)
```

## Add Spatial

```{r}
#| label: spatial
# spatial last: nothing above has to UPDATE a table that carries geometry
load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
add_point_geom(con, "poc_sample", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "poc_sample") |> datatable(caption = "Grid assignment")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata
this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot(nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)
```

## Schema Documentation

Source shape (what the notebook wrangles):

```{r}
#| label: schema
poc_rels <- list(
  primary_keys = list(
    poc_sample      = "sample_id",
    poc_measurement = "measurement_id",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "poc_measurement", column = "sample_id",
         ref_table = "poc_sample", ref_column = "sample_id"),
    list(table = "poc_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(
  con, tables = c("poc_sample", "poc_measurement", "measurement_type", "dataset"),
  rels = poc_rels,
  colors = list(lightblue = c("poc_sample", "poc_measurement"),
                lightyellow = "measurement_type", white = "dataset"))
```

Source to core (what a consumer sees):

| source | core | transform |
|---|---|---|
| `poc_sample` | `sample` (`bottle`) | one row per bottle sample; key `cce-lter_poc-pon:bottle:{studyName}-{Cast Number}-{Bottle Number}`; `depth_min_m = depth_max_m = Depth (m)`; time = `Datetime GMT`; `cruise_key` from ship code + `Cruise` |
| `umol/L_C (µmol/L)` | `obs` `poc` | verbatim, µmol/L |
| `umol/L_N (µmol/L)` | `obs` `pon` | verbatim, µmol/L |
| `ug_CperL`, `ug_NperL`, `C/N ratio`, `C/N molar ratio` | -- | **not published**: follow from `poc`/`pon` (asserted in "Derived columns agree") |
| `Associated Bottle Number` | -- | **not published**: meaning unconfirmed (Q03) |
| `Notes` | -- | **not published**: free text on 12 rows (Q06) |
| matched `calcofi_bottle` cast/bottle | -- | **not published**: evidence only, not `parent_sample_key` |
| 4 exact-duplicate rows, 2 conflicting-key rows, 10 no-depth rows | -- | dropped / held out (Q02, Q04) |

## Check Declared Bounds

`poc` and `pon` are registered with `valid_min = 0` (a concentration cannot be
negative). Their `valid_max` is **not declared**: Q01 proposes 1,000 µmol/L
(poc) and 150 µmol/L (pon), generous ceilings about twice the observed maxima
(527 and 67.5 µmol/L, both at inshore stations), and they stay `proposed` until
the provider agrees.

```{r}
#| label: bounds-check
bounds_pre <- check_measurement_bounds(con, "poc_measurement", mt = d_meas_type)
bounds_datatable(bounds_pre)

oob_tally <- drop_out_of_bounds(con, "poc_measurement", mt = d_meas_type)
cat(glue(
  "out-of-range rows removed: {format(sum(oob_tally$n_bad %||% 0), big.mark=',')} ",
  "(only valid_min = 0 is declared; the upper bounds await Q01)"), "\n")

bounds_post <- check_measurement_bounds(
  con, "poc_measurement", mt = d_meas_type, include_undeclared = FALSE)
stopifnot(
  "poc_measurement must hold no value outside a DECLARED bound" =
    sum(bounds_post$status == "out_of_range") == 0)
```

## Enforce Column Types

```{r}
#| label: enforce-types
enforce_column_types(con)
```

## Data Preview

```{r}
#| label: preview-sample
cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='poc_sample' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM poc_sample LIMIT 200")) |>
  datatable(caption = "poc_sample -- first 200 rows", rownames = FALSE, filter = "top")
```

```{r}
#| label: preview-measurement
dbGetQuery(con, "SELECT * FROM poc_measurement LIMIT 200") |>
  datatable(caption = "poc_measurement -- first 200 rows", rownames = FALSE, filter = "top")
```

## Emit Core Tables

Self-contained sample arm via `sample_arm_self()`, like iron: the bottle match
above is evidence, not a parent.

```{r}
#| label: emit_core
append_sample(con, sample_arm_self(
  ds_key, "poc_sample", "sample_key_id", "bottle",
  dt_col = "datetime_start_utc", site_expr = "site_key",
  depth_min = "depth_m", depth_max = "depth_m"))

append_obs(con, glue("
  SELECT 'env', '{ds_key}', {ns_key(ds_key, 'bottle', 's.sample_key_id')},
         s.grid_key, s.cruise_key, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP), s.depth_m, s.depth_m,
         NULL::VARCHAR, NULL::VARCHAR, m.measurement_type, m.measurement_value,
         NULL::VARCHAR, NULL::DOUBLE
  FROM poc_measurement m
  JOIN poc_sample s USING (sample_id)"))

q1 <- function(sql) dbGetQuery(con, sql)[[1]]
core <- list(
  sample = q1("SELECT COUNT(*) FROM sample"),
  obs    = q1("SELECT COUNT(*) FROM obs"))
cat(glue("core projection -- sample={core$sample} obs={core$obs}"), "\n")

n_exp_obs <- q1("SELECT COUNT(*) FROM poc_measurement")
src_total <- q1("SELECT SUM(measurement_value) FROM poc_measurement")
obs_total <- q1(glue("SELECT SUM(measurement_value) FROM obs WHERE dataset_key = '{ds_key}'"))
stopifnot(
  "one sample per source sample" = core$sample == nrow(d_sample),
  "obs is one row per measurement" = core$obs == n_exp_obs,
  "obs value total equals the source" = isTRUE(all.equal(obs_total, src_total)),
  "only poc and pon are emitted" =
    setequal(dbGetQuery(con, glue("SELECT DISTINCT measurement_type FROM obs
                                   WHERE dataset_key = '{ds_key}'"))$measurement_type,
             c("poc", "pon")),
  "sample_key is unique" =
    q1(glue("SELECT COUNT(*) - COUNT(DISTINCT sample_key) FROM sample
             WHERE dataset_key = '{ds_key}'")) == 0,
  "every obs.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key)
        WHERE s.sample_key IS NULL") == 0,
  "every obs.measurement_type resolves in measurement_type" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN measurement_type t USING (measurement_type)
        WHERE t.measurement_type IS NULL") == 0)

# serve the per-dataset name as a VIEW over the core
invisible(dbExecute(con, "DROP TABLE IF EXISTS poc_measurement_src"))
invisible(dbExecute(con, "ALTER TABLE poc_measurement RENAME TO poc_measurement_src"))
invisible(dbExecute(con, glue(
  "CREATE OR REPLACE VIEW poc_measurement AS
   SELECT o.obs_id AS measurement_id,
          o.sample_key,
          substr(o.sample_key, length('{ds_key}:bottle:') + 1) AS sample_key_id,
          o.measurement_type, o.measurement_value, o.measurement_qual
   FROM obs o WHERE o.dataset_key = '{ds_key}'")))
```

### Declared NULLs

`validate_for_release()` treats every `_key` column as required, which is a
heuristic. Each NULL this dataset legitimately carries is declared with its count
and reason; anything undeclared, or a count that moves, stops the render.

```{r}
#| label: validate
results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("validate_for_release:", ifelse(results$passed, "PASSED",
    "reported findings -- reconciled against the declared NULLs below"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")

n_null <- function(col) q1(glue(
  "SELECT COUNT(*) FROM sample WHERE dataset_key = '{ds_key}' AND {col} IS NULL"))
nullable <- tribble(
  ~table,   ~column,             ~n,                          ~reason,
  "sample", "parent_sample_key", n_null("parent_sample_key"), "single-level dataset: every sample is a root (the matched calcofi_bottle bottle is evidence, not a parent)",
  "sample", "grid_key",          n_null("grid_key"),          "position outside the CalCOFI grid cells",
  "sample", "cruise_key",        n_null("cruise_key"),        "ship code in studyName not in the ship reference")
nullable |> datatable(caption = "Declared NULLs (count and reason)", rownames = FALSE)

stopifnot(
  "every sample is a root" = nullable$n[nullable$column == "parent_sample_key"] == nrow(d_sample),
  "no sample lacks a time or a site_key" =
    n_null("datetime") == 0 && n_null("site_key") == 0,
  "no sample lacks a depth (no-depth rows are held out, Q04)" =
    n_null("depth_max_m") == 0)
# grid_key / cruise_key NULLs are reported here on the first render; once seen,
# pin each count in an assertion so a change is noticed (a ratchet).
```

### Core ERD

```{r}
#| label: erd-core
core_tbls <- core_output_tables(con)
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls))
```

## Write Parquet Outputs

```{r}
#| label: write-parquet
dir_create(dir_parquet)
mismatches <- list(cruise_keys = collect_cruise_key_mismatches(con, "poc_sample"))
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
parquet_stats <- write_parquet_outputs(
  con = con, output_dir = dir_parquet,
  tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE, mismatches = mismatches)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)
parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata
d_tbls_rd <- read_csv(file.path(dir_meta, "tbls_redefine.csv"))
d_flds_rd <- read_csv(file.path(dir_meta, "flds_redefine.csv"))
metadata_path <- build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           file.path(dir_meta, "metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned,
  sources = src_stamp)
```

## Upload to GCS

```{r}
#| label: upload-gcs
if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet,
              gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else cat("CALCOFI_SKIP_GCS set -- skipping the upload (local verification run)\n")
```

## Questions for Data Providers

Follow-up questions for CCE-LTER (Ralf Goericke), ranked. Q01 (bounds) and Q02
(duplicates) are `proposed`: each carries an answer for the provider to confirm.
None is a blocker: what is published is the source's own µmol/L values, and the
rows the questions are about are held out.

```{r}
#| label: provider-questions
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the cce-lter poc-pon data providers (ranked)")
```

## Cleanup

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("parquet outputs written to: {dir_parquet}"), "\n")
```
