---
title: "Ingest Spatial Boundaries"
editor_options:
  chunk_output_type: console
---

## Overview

Process spatial boundary datasets for:

1. **Map overlay** — PMTiles served via `file.calcofi.io/_spatial/`
2. **Spatial queries** — parquet tables loaded into CalCOFI DuckDB
3. **Future summarization** — aggregate CalCOFI data into boundary features

**Source**: `~/My Drive/projects/calcofi/data-public/_spatial/`
([US Maritime Limits](https://nauticalcharts.noaa.gov/data/us-maritime-limits-and-boundaries.html),
CA CDFW, BOEM, NOAA ONMS, MEOW, and others)

**Registry**: `metadata/spatial_layers.csv` — one row per visual layer, with
dataset grouping, styling, filter expressions, and source file paths.

### Data Flow

```{mermaid}
graph LR
  A[Source Shapefiles] --> B[Read + Reproject 4326]
  B --> C[clean_names + id column]
  C --> D[Parquet Export]
  C --> E[GeoJSON → tippecanoe → PMTiles]
  D --> F[DuckDB Tables]
  E --> G[GCS Upload]
  G --> H[file.calcofi.io/_spatial/]
```

### Pre-processing

Each dataset is standardized:

- **CRS**: reprojected to `EPSG:4326` (WGS 84 unprojected)
- **Columns**: `janitor::clean_names()` for snake_case
- **ID**: integer `id` column as first column (from source FID or `row_number()`)

## Setup

```{r}
#| label: setup
librarian::shelf(
  arcgislayers,
  calcofi / calcofi4db,
  DBI,
  dplyr,
  DT,
  duckdb,
  fs,
  glue,
  here,
  janitor,
  jsonlite,
  purrr,
  readr,
  sf,
  stringr,
  tibble,
  quiet = T
)

redo_if_exists <- FALSE

dir_spatial <- path_expand("~/My Drive/projects/calcofi/data-public/_spatial")
dir_pmtiles <- here("data/pmtiles")
dir_parquet <- here("data/parquet/spatial")
gcs_bucket <- "calcofi-files-public"
gcs_prefix <- "_spatial"

dir_create(dir_pmtiles)
dir_create(dir_parquet)

# load registry
d_layers <- read_csv(
  here("metadata/spatial_layers.csv"),
  show_col_types = FALSE
)

datatable(
  d_layers,
  caption = "Spatial layers registry from `metadata/spatial_layers.csv`"
)

# compare with folders/files in dir_spatial
spatial_dirs <- dir_ls(dir_spatial, type = "directory")
spatial_dirs_not_layers_csv <- setdiff(
  basename(spatial_dirs),
  unique(d_layers$source_folder)
)
if (length(spatial_dirs_not_layers_csv) > 0) {
  cat(
    glue(
      "⚠ WARNING: The following folders exist in `{dir_spatial}` but are not ",
      "referenced in `metadata/spatial_layers.csv`:\n",
      paste0("- ", spatial_dirs_not_layers_csv, collapse = "\n"),
      "\n\n"
    )
  )
}

# unique dataset groups (one pmtiles + parquet per group)
d_groups <- d_layers |>
  distinct(dataset_group, source_folder, source_file, source_format) |>
  mutate(
    source_path = path(dir_spatial, source_folder, source_file),
    file_exists = file_exists(source_path)
  )

cat(
  glue("{nrow(d_layers)} layers across {nrow(d_groups)} dataset groups"),
  "\n"
)
```

## ArcGIS Rest layer(s) to gpkg: `CA_watershed-boundaries`

```{r}
#| label: arcgis-rest

fs_to_sf <- function(url, where = NULL) {
  fs <- arcgislayers::arc_open(url)
  lyr <- arcgislayers::get_layer(fs, 0)

  sf <- arcgislayers::arc_select(
    lyr,
    where = where
  ) |>
    tibble::tibble() |>
    sf::st_as_sf() |>
    janitor::clean_names()

  return(sf)
}

# [USGS Watershed Boundary Dataset (WBD) HUC8, California Vicinity - Overview](https://www.arcgis.com/home/item.html?id=a98ef9ead82a4bd6aa09d46869f75a49&sublayer=0)
url_rest <- "https://services2.arcgis.com/Uq9r85Potqm3MfRV/arcgis/rest/services/WBD_HU8_wm/FeatureServer"
out_gpkg <- glue("{dir_spatial}/CA_watershed-boundaries/wbd-hu8_ca.gpkg")

if (!file_exists(out_gpkg) || redo_if_exists) {
  fs_to_sf(url_rest) |>
    write_sf(out_gpkg, delete_dsn = TRUE)
}
```


## Clip CA Current LME to US EEZ

```{r}
#| label: clip-layers

cacurrent_geo <- glue(
  "{dir_spatial}/NOAA_IEA-CA-Current/lme-ca-current.geojson"
)
usaeez_geo <- glue(
  "{dir_spatial}/NOAA_IEA-CA-Current/usa-eez.geojson"
)
cacurrent_usaeez_geo <- glue(
  "{dir_spatial}/NOAA_IEA-CA-Current/lme-ca-current_usa-eez.geojson"
)

if (!file_exists(cacurrent_usaeez_geo) || redo_if_exists) {
  cacurrent <- st_read(cacurrent_geo, quiet = TRUE)
  usaeez <- st_read(usaeez_geo, quiet = TRUE)

  cacurrent_usaeez <- st_intersection(cacurrent, usaeez)

  st_write(
    cacurrent_usaeez,
    cacurrent_usaeez_geo,
    delete_dsn = TRUE,
    quiet = TRUE
  )
}
```

## Process datasets

```{r}
#| label: process-datasets
#| output: asis

process_spatial_dataset <- function(
  dataset_group,
  source_folder,
  source_file,
  source_format,
  dir_spatial,
  dir_pmtiles,
  dir_parquet,
  redo_if_exists = FALSE
) {
  pmtiles_path <- path(dir_pmtiles, glue("{dataset_group}.pmtiles"))
  parquet_path <- path(dir_parquet, glue("{dataset_group}.parquet"))

  # skip if already processed
  if (
    !redo_if_exists &&
      file_exists(pmtiles_path) &&
      file_exists(parquet_path)
  ) {
    cat(glue("- **{dataset_group}**: skipped (already exists)\n\n"))
    return(invisible(TRUE))
  }

  # read source data ----
  sf_data <- tryCatch(
    {
      if (source_format == "rda") {
        # .rda files: load() into temp env, extract sf object
        e <- new.env()
        load(path(dir_spatial, source_folder, source_file), envir = e)
        obj_names <- ls(e)
        sf_obj <- NULL
        for (nm in obj_names) {
          if (inherits(get(nm, envir = e), "sf")) {
            sf_obj <- get(nm, envir = e)
            break
          }
        }
        sf_obj
        # drop any list columns (e.g. from complex attributes that can't be written to parquet)
        sf_obj |> select(where(~ !is.list(.)))
      } else if (source_format == "gdb") {
        # geodatabase: source_file = "file.gdb|layer_name"
        parts <- str_split_1(source_file, "\\|")
        st_read(
          path(dir_spatial, source_folder, parts[1]),
          layer = parts[2],
          quiet = TRUE
        )
      } else {
        # shp, gpkg, geojson
        st_read(
          path(dir_spatial, source_folder, source_file),
          quiet = TRUE
        )
      }
    },
    error = function(e) {
      cat(glue(
        "- **{dataset_group}**: ⚠ SKIPPED — failed to read ",
        "`{source_folder}/{source_file}`: {e$message}\n\n"
      ))
      return(NULL)
    }
  )

  if (is.null(sf_data) || nrow(sf_data) == 0) {
    cat(glue("- **{dataset_group}**: ⚠ SKIPPED — empty dataset\n\n"))
    return(invisible(FALSE))
  }

  # reproject to EPSG:4326 ----
  src_crs <- st_crs(sf_data)$epsg
  if (is.na(src_crs) || src_crs != 4326) {
    sf_data <- st_transform(sf_data, 4326)
  }

  # clean column names ----
  sf_data <- clean_names(sf_data)

  # ensure integer id column first ----
  geom_col <- attr(sf_data, "sf_column")
  if (!"id" %in% names(sf_data)) {
    sf_data <- sf_data |> mutate(id = row_number())
  }
  sf_data <- sf_data |>
    mutate(id = as.integer(id)) |>
    relocate(id)

  # write parquet ----
  # use sf + arrow for geoparquet
  tryCatch(
    arrow::write_parquet(sf_data, parquet_path),
    error = function(e) {
      # fallback: write via DuckDB spatial
      tmp_geojson <- tempfile(fileext = ".geojson")
      st_write(sf_data, tmp_geojson, delete_dsn = TRUE, quiet = TRUE)
      con_tmp <- dbConnect(duckdb())
      dbExecute(con_tmp, "INSTALL spatial; LOAD spatial;")
      dbExecute(
        con_tmp,
        glue(
          "COPY (SELECT * FROM ST_Read('{tmp_geojson}'))
         TO '{parquet_path}' (FORMAT PARQUET)"
        )
      )
      dbDisconnect(con_tmp, shutdown = TRUE)
      unlink(tmp_geojson)
    }
  )

  # write pmtiles ----
  geojson_tmp <- tempfile(fileext = ".geojson")
  st_write(sf_data, geojson_tmp, delete_dsn = TRUE, quiet = TRUE)

  tippecanoe_args <- c(
    "-o",
    pmtiles_path,
    "-z10",
    "-Z0",
    "--simplification=10",
    "--no-tile-size-limit",
    "-l",
    dataset_group,
    "--force",
    geojson_tmp
  )

  result <- system2(
    "tippecanoe",
    args = tippecanoe_args,
    stdout = TRUE,
    stderr = TRUE
  )
  unlink(geojson_tmp)

  if (!file_exists(pmtiles_path)) {
    cat(glue(
      "- **{dataset_group}**: ⚠ SKIPPED — tippecanoe failed\n\n"
    ))
    return(invisible(FALSE))
  }

  pmtiles_size <- file_size(pmtiles_path)
  parquet_size <- file_size(parquet_path)

  cat(glue(
    "- **{dataset_group}**: {nrow(sf_data)} features ",
    "(CRS {src_crs} → 4326), ",
    "PMTiles {pmtiles_size}, Parquet {parquet_size}\n\n"
  ))

  return(invisible(TRUE))
}

# process each dataset group
results <- d_groups |>
  # slice(4:nrow(d_groups)) |> # DEBUG: skip first 3 unproblematic groups
  select(dataset_group, source_folder, source_file, source_format) |>
  pmap_lgl(
    process_spatial_dataset,
    dir_spatial = dir_spatial,
    dir_pmtiles = dir_pmtiles,
    dir_parquet = dir_parquet,
    redo_if_exists = redo_if_exists
  )

n_ok <- sum(results)
n_skip <- sum(!results)
cat(glue("\n**Summary**: {n_ok} succeeded, {n_skip} skipped/failed\n"))
```

## Upload to GCS

```{r}
#| label: upload-gcs

pmtiles_files <- dir_ls(dir_pmtiles, glob = "*.pmtiles")
cat(glue(
  "Uploading {length(pmtiles_files)} PMTiles files to ",
  "gs://{gcs_bucket}/{gcs_prefix}/\n\n"
))

sync_to_gcs(
  local_dir = dir_pmtiles,
  gcs_prefix = gcs_prefix,
  bucket = gcs_bucket,
  pattern = "\\.pmtiles$"
)
```

## Load into DuckDB

Load parquet files into the wrangling database for spatial queries.

```{r}
#| label: load-duckdb
#| output: asis

db_path <- here("data/calcofi_wrangling.duckdb")
con <- dbConnect(duckdb(dbdir = db_path))
dbExecute(con, "INSTALL spatial; LOAD spatial;")

parquet_files <- dir_ls(dir_parquet, glob = "*.parquet")

for (pq in parquet_files) {
  tbl_name <- paste0("spatial_", path_ext_remove(path_file(pq)))

  # read parquet into a temp view first to discover geometry column name
  dbExecute(
    con,
    glue(
      "CREATE OR REPLACE TEMP VIEW v_tmp AS
     SELECT * FROM read_parquet('{pq}')"
    )
  )

  # find geometry column (type GEOMETRY or contains 'geom' in name)
  cols_info <- dbGetQuery(con, "DESCRIBE v_tmp")
  geom_col <- cols_info$column_name[
    grepl("GEOMETRY|WKB_GEOMETRY", cols_info$column_type, ignore.case = TRUE) |
      grepl("^geom", cols_info$column_name, ignore.case = TRUE)
  ]

  if (length(geom_col) > 0) {
    # strip CRS from geometry to avoid storage version error
    geom_col <- geom_col[1]
    other_cols <- setdiff(cols_info$column_name, geom_col)
    select_sql <- paste(
      c(
        other_cols,
        glue("ST_GeomFromWKB(ST_AsWKB({geom_col})) AS {geom_col}")
      ),
      collapse = ", "
    )
    dbExecute(
      con,
      glue(
        "CREATE OR REPLACE TABLE {tbl_name} AS
       SELECT {select_sql} FROM v_tmp"
      )
    )
  } else {
    # no geometry column, load as-is
    dbExecute(
      con,
      glue(
        "CREATE OR REPLACE TABLE {tbl_name} AS
       SELECT * FROM v_tmp"
      )
    )
  }

  n_rows <- dbGetQuery(con, glue("SELECT COUNT(*) AS n FROM {tbl_name}"))$n
  cat(glue("- {tbl_name}: {n_rows} rows\n\n"))
}

dbDisconnect(con, shutdown = TRUE)
```

## Output summary

```{r}
#| label: summary
#| output: asis

cat("### PMTiles\n\n")
dir_ls(dir_pmtiles, glob = "*.pmtiles") |>
  tibble(path = _) |>
  mutate(
    file = path_file(path),
    size = file_size(path)
  ) |>
  select(file, size) |>
  knitr::kable()

cat("\n### Parquet\n\n")
dir_ls(dir_parquet, glob = "*.parquet") |>
  tibble(path = _) |>
  mutate(
    file = path_file(path),
    size = file_size(path)
  ) |>
  select(file, size) |>
  knitr::kable()
```
