---
title: "Ingest CalCOFI Bird & Mammal Census"
calcofi:
  target_name: ingest_calcofi_bird_mammal_census
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/calcofi_bird_mammal_census/manifest.json
  provider: calcofi
  dataset: bird_mammal_census
  workflow_url: https://calcofi.io/workflows/ingest_calcofi_bird_mammal_census.html
  questions_file: metadata/calcofi/bird_mammal_census/questions.csv
  dataset_meta:
    dataset_name: CalCOFI Bird & Mammal Census
    description: >
      Bird and mammal observations along CalCOFI (and NMFS, CPR) cruise transects,
      1987-2021. Transect effort + counts + species/behavior lookups. Curated via
      CCE-LTER DataZoo 255 (PI Bill Sydeman).
    citation_main: ""
    link_calcofi_org: ""
    link_data_source: "https://oceaninformatics.ucsd.edu/datazoo/catalogs/ccelter/datasets/255"
    link_others:
      - http://dx.doi.org/10.6073/pasta/4ee1bd702acb11786277192a41626800
    coverage_temporal: 1987 to 2021
    coverage_spatial: "CalCOFI region (California Current)"
    license: ""
    pi_names: Bill Sydeman
  tables_owned:
    - {table: bird_mammal_transect}
    - {table: bird_mammal_observation}
    - {table: bird_mammal_species}
    - {table: bird_mammal_behavior}
  erd:
    color: "#f0d2e8"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: `whales-seabirds-turtles/bird-mammal-census/` (CCE-LTER DataZoo 255,
PI Bill Sydeman). Bird & mammal observations along CalCOFI/NMFS/CPR cruise
transects, 1987-2021.

- **Provider**: `calcofi` (tentative — curated via CCE-LTER; see questions)
- **Tables**: `bird_mammal_transect` (effort) ⨝ `bird_mammal_observation` (counts)
  on `gis_key`, plus `bird_mammal_species` (ITIS taxonomy) and
  `bird_mammal_behavior` lookups.

```{mermaid}
graph LR
  T[transects] --> TR[bird_mammal_transect<br/>effort + position]
  O[observations] --> OB[bird_mammal_observation<br/>counts]
  OB -.gis_key.-> TR
  OB -.species_code.-> SP[bird_mammal_species]
  OB -.behavior_code.-> BH[bird_mammal_behavior]
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, units, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))
source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_calcofi_bird_mammal_census.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_src      <- path_expand(glue("{dir_data}/whales-seabirds-turtles/bird-mammal-census"))

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

```{r}
#| label: read-source

stopifnot("source dir not found" = dir_exists(dir_src))
sync_to_gcs(local_dir = dir_src, gcs_prefix = glue("archive/{provider}/{dataset}"),
            bucket = "calcofi-files-public", exclude = c(".DS_Store", "*.tmp", "*.gdoc"))

rd <- function(f) read_csv(file.path(dir_src, glue("CalCOFI_bird-mammal-census_{f}.csv")),
                           col_types = cols(.default = "c")) |> clean_names()
d_tr <- rd("transects"); d_ob <- rd("observations")
d_sp <- rd("allspecieslist"); d_bh <- rd("behaviorcodes")
cat(glue("transects {nrow(d_tr)}, observations {nrow(d_ob)}, species {nrow(d_sp)}, behaviors {nrow(d_bh)}"), "\n")
cat("transect clean names:", paste(names(d_tr), collapse=", "), "\n")
```

## Build Lookups (species, behavior)

```{r}
#| label: lookups

b01 <- function(x) !is.na(x) & x %in% c("1","TRUE","true","Y","y")

d_species <- d_sp |>
  transmute(
    species_code = species, common_name, scientific_name = latin_name,
    itis_id = suppressWarnings(as.integer(itis)),
    is_bird = b01(bird), is_mammal = b01(mammal), is_fish = b01(fish),
    is_large_bird = b01(large_bird), is_unidentified = b01(unidentified),
    include_flag = b01(include), nmfs_code = nmfs, comment)
dbWriteTable(con, "bird_mammal_species", d_species, overwrite = TRUE)

d_behavior <- d_bh |> transmute(behavior_code = id, behavior, description)
dbWriteTable(con, "bird_mammal_behavior", d_behavior, overwrite = TRUE)
cat(glue("species {nrow(d_species)}, behaviors {nrow(d_behavior)}"), "\n")
```

## Build Transect (effort) Table

```{r}
#| label: transect

d_transect <- d_tr |>
  transmute(
    gis_key,
    cruise_label    = cruise,
    transect_number = suppressWarnings(as.integer(transect_number)),
    bin_number      = suppressWarnings(as.integer(bin_number)),
    date            = suppressWarnings(as.Date(date)),
    time_sec        = suppressWarnings(as.numeric(time_sec)),
    latitude        = suppressWarnings(as.numeric(latitude_mid_o)),
    longitude       = suppressWarnings(as.numeric(longitude_mid_o)),
    latitude_start  = suppressWarnings(as.numeric(latitude_start_o)),
    longitude_start = suppressWarnings(as.numeric(longitude_start_o)),
    latitude_stop   = suppressWarnings(as.numeric(latitude_stop_o)),
    longitude_stop  = suppressWarnings(as.numeric(longitude_stop_o)),
    length_m        = suppressWarnings(as.numeric(length_m)),
    width_m         = suppressWarnings(as.numeric(width_m)),
    area_m2         = suppressWarnings(as.numeric(area_m2)),
    bottom_depth_m  = suppressWarnings(as.numeric(depth_m)),
    julian_date     = suppressWarnings(as.numeric(julian_date)),
    julian_day      = suppressWarnings(as.integer(julian_day)),
    svy, season,
    # datetime from date + time_sec (Q01: tz unconfirmed, treated as given)
    datetime_start_utc = as_datetime(date) + dseconds(coalesce(time_sec, 0)))

dbWriteTable(con, "bird_mammal_transect", d_transect, overwrite = TRUE)
cat(glue("bird_mammal_transect: {nrow(d_transect)} rows"), "\n")
```

## Build Observation (counts) Table

```{r}
#| label: observation

d_observation <- d_ob |>
  transmute(gis_key, species_code = species, behavior_code = behavior,
            count = suppressWarnings(as.integer(count))) |>
  mutate(observation_id = row_number(), .before = 1)
dbWriteTable(con, "bird_mammal_observation", d_observation, overwrite = TRUE)
cat(glue("bird_mammal_observation: {nrow(d_observation)} rows"), "\n")
```

## Resolve cruise_key + Spatial

```{r}
#| label: spatial-cruise

# cruise_key: best-effort YYYY-MM from cruise_label/date (no ship column -> Q02).
# leave NULL for now; record format for the provider question.
dbExecute(con, "ALTER TABLE bird_mammal_transect ADD COLUMN IF NOT EXISTS cruise_key VARCHAR")

# geometry + grid from the transect midpoint
load_prior_tables(con, parquet_dir = here("data/parquet/swfsc_ichthyo"),
                  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
add_point_geom(con, "bird_mammal_transect", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "bird_mammal_transect") |> datatable(caption = "Grid assignment")
```

## Load Dataset Metadata + Schema

```{r}
#| label: meta-schema

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)

bmc_rels <- list(
  primary_keys = list(
    bird_mammal_transect = "gis_key", bird_mammal_observation = "observation_id",
    bird_mammal_species = "species_code", bird_mammal_behavior = "behavior_code"),
  foreign_keys = list(
    list(table="bird_mammal_observation", column="gis_key",       ref_table="bird_mammal_transect", ref_column="gis_key"),
    list(table="bird_mammal_observation", column="species_code",  ref_table="bird_mammal_species",  ref_column="species_code"),
    list(table="bird_mammal_observation", column="behavior_code", ref_table="bird_mammal_behavior", ref_column="behavior_code")))
cc_erd(con, tables = c("bird_mammal_transect","bird_mammal_observation","bird_mammal_species","bird_mammal_behavior","dataset"),
       rels = bmc_rels,
       colors = list(lightblue = c("bird_mammal_transect","bird_mammal_observation"),
                     lightyellow = c("bird_mammal_species","bird_mammal_behavior"), white = "dataset"))
build_relationships_json(rels = bmc_rels, output_dir = dir_parquet, provider = provider, dataset = dataset)
```

## Validate + Preview

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
# orphan check: observations whose gis_key is not in transects
n_orphan <- dbGetQuery(con, "SELECT COUNT(*) FROM bird_mammal_observation o
  LEFT JOIN bird_mammal_transect t USING(gis_key) WHERE t.gis_key IS NULL")[[1]]
cat(glue("observation gis_key orphans (no transect): {n_orphan}"), "\n")
```

```{r}
#| label: preview

cols <- dbGetQuery(con, "SELECT column_name FROM information_schema.columns
  WHERE table_name='bird_mammal_transect' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM bird_mammal_transect LIMIT 100")) |>
  datatable(caption = "bird_mammal_transect — first 100", rownames = FALSE, filter = "top")
```

## Questions for Data Providers

```{r}
#| label: provider-questions

read_csv(here(glue("metadata/{provider}/{dataset}/questions.csv"))) |>
  arrange(factor(priority, c("blocker","high","normal")), id) |>
  select(priority, question, context, status) |>
  datatable(caption = "Questions (ranked)", options = list(dom="t", pageLength=10), rownames = FALSE)
```

## Write Outputs + Upload

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`) via `calcofi4db::emit_core_tables()` — the
single source of truth for the per-dataset projection into `sample` / `obs` /
`obs_freq` / `sample_measurement`, also used by `release_database.qmd` to
assemble the authoritative cross-dataset release.

```{r}
#| label: emit_core

core <- emit_core_tables(con, "calcofi_bird_mammal_census")
cat(glue(
  "core projection — sample={core$sample %||% 0} obs={core$obs %||% 0} ",
  "obs_freq={core$obs_freq %||% 0} sample_measurement={core$sample_measurement %||% 0}\n"))

# the frozen release assembles the authoritative cross-dataset core; keep this
# ingest's parquet output per-dataset by dropping the local core projection here.
for (tbl_core in c("obs", "obs_freq", "sample_measurement", "sample"))
  dbExecute(con, glue("DROP TABLE IF EXISTS {tbl_core}"))
```


```{r}
#| label: outputs

dir_create(dir_parquet)
write_parquet_outputs(
  con = con, output_dir = dir_parquet,
  tables = c("bird_mammal_transect","bird_mammal_observation","bird_mammal_species","bird_mammal_behavior","dataset"),
  strip_provenance = FALSE)

d_tbls_rd <- read_csv(here("metadata/calcofi/bird_mammal_census/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/calcofi/bird_mammal_census/flds_redefine.csv"))
build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = here("metadata/calcofi/bird_mammal_census/metadata_derived.csv"),
  output_dir = dir_parquet, tables = c("bird_mammal_transect","bird_mammal_observation","bird_mammal_species","bird_mammal_behavior"),
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned)

sync_to_gcs(local_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
```

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
