---
title: "Ingest CCE-LTER Picoplankton and Bacteria Abundance"
calcofi:
  target_name: ingest_cce_lter_picoplankton_bacteria
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/cce-lter_picoplankton-bacteria/manifest.json
  provider: cce-lter
  dataset: picoplankton-bacteria
  workflow_url: https://calcofi.io/workflows/ingest_cce-lter_picoplankton-bacteria.html
  questions_file: metadata/cce-lter/picoplankton-bacteria/questions.csv
  dataset_meta:
    dataset_name: Picoplankton and Bacteria Abundance (CalCOFI Cruise)
    description: >
      Picophytoplankton (Prochlorococcus, Synechococcus, picoeukaryotes) and
      heterotrophic bacteria abundances analyzed by flow cytometry (FCM) from
      CCE-CalCOFI Augmented cruises in the California Current System,
      2004-2023 (ongoing). Seawater collected from Niskin bottles at 3-8
      depths per station; cells fixed shipboard with paraformaldehyde,
      stained with a DNA-specific dye, and enumerated on an Altra flow
      cytometer with dual argon-ion lasers.
    citation_main: "Landry, M. (2004-2023). Picoplankton and Bacteria Abundance (CalCOFI Cruise). CCE LTER."
    link_calcofi_org: ""
    link_data_source: "https://oceaninformatics.ucsd.edu/datazoo/catalogs/ccelter/datasets/159"
    link_others:
      - "http://dx.doi.org/10.6073/pasta/bc2915c8448214d2841b064a7414064b"
    coverage_temporal: 2004-11-02 to 2023-07-18 (ongoing)
    coverage_spatial: "CalCOFI grid (California Current Ecosystem)"
    license: ""
    pi_names: Michael Landry
  # publishes the consolidated core: the source shape is wrangled in the notebook
  # and projected into the shared core tables.
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (bottle grain)"}
    - {table: obs, shared: true, note: "core observations (env, flow-cytometry cell counts)"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#bbe0f0"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: Datazoo table download (`https://oceaninformatics.ucsd.edu/datazoo/catalogs/ccelter/datasets/159/datatables/159/download`),
16,017 rows, 16 columns (73 cruises, 2004-11-02 to 2023-07-18).

- **Provider**: `cce-lter` (matches the existing provider group used by
  `ingest_cce-lter_euphausiids`/`_zoodb`/`_zooscan` -- an earlier draft of
  this notebook used the unhyphenated `ccelter`, which would have created a
  duplicate, inconsistent provider group)
- **Grain**: one row per bottle/depth per cast (station-level, multiple depths per cast)
- **Scope decision**: no bounding-box filter applied -- source is already
  scoped to CalCOFI/CCE cruises by the provider. `studyName` (e.g.
  `2004-11-02-C-33RR`) is the cruise identifier used for key resolution, NOT
  the `Cruise` column (which is a separate YYYYMM short code, e.g. `200411`
  -- kept as a raw field but not used for joins).
- **Completeness**: Depth, Heterotrophic Bacteria, Synechococcus, and
  Picoeukaryotes are >99.9% populated. Prochlorococcus is ~80% populated
  (3,228/16,017 blank) -- plausible non-detection at higher latitudes/depths,
  not yet confirmed with the provider (see Questions, Q01). `Notes` is 100%
  blank in the current export; it is read and normalized to `NA` rather than
  dropped, so a future export that starts populating it flows through without a
  schema change. No literal-zero values appear
  in any measurement column in this export (confirmed directly against the
  raw file) -- unlike ZooDB, there is no analyzed-but-absent convention to
  encode; blank simply means not measured.

```{mermaid}
graph LR
  A[Datazoo table 159<br/>16,017 rows] --> B[rename + type-cast]
  B --> C[picoplankton_bacteria_bottle<br/>position + depth, one row per bottle]
  B --> D[picoplankton_bacteria_measurement<br/>long format: 4 measurement types]
  C -.studyName -> cruise_key.-> E[(shared refs)]
  D -.bottle_id.-> C
```

## Setup

```{r}
#| label: setup
devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))
cc <- read_calcofi_meta(here("ingest_cce-lter_picoplankton-bacteria.qmd"))
provider <- cc$provider
dataset <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label <- glue("{provider}_{dataset}")
dir_parquet <- here(glue("data/parquet/{dir_label}"))
db_path <- here(glue("data/wrangling/{dir_label}.duckdb"))
meas_type_csv <- here("metadata/measurement_type.csv")
d_meas_type <- read_csv(meas_type_csv)

if (overwrite) {
  if (file_exists(db_path)) file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp"))) dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

Fetched reproducibly by `libs/download_picoplankton_bacteria.R` into
`{dir_data}/cce-lter/picoplankton-bacteria/`. The dataset page links to the
Datazoo table download, but that URL redirects to a *Login / Accept Data
Agreement* form and so cannot be fetched unattended; EDI carries the same table
(the Datazoo DOI resolves to `knb-lter-cce.159`), so we pull from there. The
revision is deliberately unpinned — this series is ongoing — and the revision
actually used is reported below. Columns are renamed per
`metadata/cce-lter/picoplankton-bacteria/flds_redefine.csv`.

```{r}
#| label: read-source
source(here("libs/download_picoplankton_bacteria.R"))
pico_csv <- download_picoplankton_bacteria(
  path_expand(glue("{dir_data}/{provider}/{dataset}")),
  overwrite = overwrite_all)
stopifnot("picoplankton/bacteria CSV not found" = file_exists(pico_csv))

sync_to_gcs(
  local_dir = path_dir(pico_csv),
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket = "calcofi-files-public",
  exclude = c(".DS_Store", "*.tmp", "*.gdoc"))

d_raw <- read_csv(pico_csv, col_types = cols(.default = "c"))
cat(glue("Read {format(nrow(d_raw), big.mark=',')} rows, {ncol(d_raw)} columns"), "\n")
```

## Clean, Type-Cast

Canonical names from `metadata/cce-lter/picoplankton-bacteria/flds_redefine.csv`.
`studyName` (the cruise identifier string, e.g. `2004-11-02-C-33RR`) is
retained separately from `Cruise` (the YYYYMM short code) -- only the former
is used for cruise-key resolution. The four measurement columns
(Heterotrophic Bacteria, Prochlorococcus, Synechococcus, Picoeukaryotes) are
NOT kept as wide columns -- they are pivoted to long format below, following
the `bottle_measurement`/`zoodb_measurement` pattern used elsewhere in the
integrated DB.

```{r}
#| label: clean-cast
d_clean <- d_raw |>
  transmute(
    study_name       = studyName,
    cruise_code      = Cruise,
    datetime_utc     = as_datetime(`Datetime GMT`),
    latitude         = suppressWarnings(as.numeric(`Latitude (º)`)),
    longitude        = suppressWarnings(as.numeric(`Longitude (º)`)),
    line             = suppressWarnings(as.numeric(Line)),
    station          = suppressWarnings(as.numeric(Station)),
    cast_number      = suppressWarnings(as.integer(`Cast Number`)),
    bottle_number    = suppressWarnings(as.integer(`Bottle Number`)),
    assoc_bottle_number = suppressWarnings(as.integer(`Associated Bottle Number`)),
    depth_m          = suppressWarnings(as.numeric(`Depth (m)`)),
    het_bacteria_n_ml    = suppressWarnings(as.numeric(`Heterotrophic Bacteria (number/ml)`)),
    prochlorococcus_n_ml = suppressWarnings(as.numeric(`Prochlorococcus (number/ml)`)),
    synechococcus_n_ml   = suppressWarnings(as.numeric(`Synechococcus (number/ml)`)),
    picoeukaryotes_n_ml  = suppressWarnings(as.numeric(`Picoeukaryotes (number/ml)`)),
    notes            = na_if(Notes, "")) |>
  mutate(
    site_key = if_else(
      is.na(line) | is.na(station), NA_character_,
      sprintf("%05.1f %05.1f", line, station)),
    bottle_id = row_number())

n_all <- nrow(d_clean)
cat(glue("{format(n_all, big.mark=',')} rows read, no bounding-box filter applied ",
         "(source pre-scoped to CalCOFI/CCE cruises)"), "\n")

dbWriteTable(con, "picoplankton_bacteria_bottle", d_clean, overwrite = TRUE)
```

## Pivot Measurements to Long Format

Unlike ZooDB, this source has no analyzed-but-absent convention -- every
blank cell in the raw export is genuinely not measured (confirmed: zero
literal `0` values appear in Heterotrophic Bacteria or Prochlorococcus,
only blanks). So this pivot simply drops NA, with no explicit-zero
retention logic needed.

`measurement_type` codes are unit-less (`het_bacteria`, not
`het_bacteria_n_ml`) -- units live in `measurement_type.csv`'s `units`
column, matching the convention used by `bottle_measurement`/
`zoodb_measurement` elsewhere in the DB. An earlier draft baked the unit
into the type code itself; fixed here.

```{r}
#| label: pivot-long
pb_meas_map <- tribble(
  ~measurement_type,     ~value_col,
  "het_bacteria",         "het_bacteria_n_ml",
  "prochlorococcus",      "prochlorococcus_n_ml",
  "synechococcus",        "synechococcus_n_ml",
  "picoeukaryotes",       "picoeukaryotes_n_ml")

sql_parts <- purrr::pmap_chr(pb_meas_map, function(measurement_type, value_col) {
  glue(
    "SELECT bottle_id, '{measurement_type}' AS measurement_type,
     CAST({value_col} AS DOUBLE) AS measurement_value
     FROM picoplankton_bacteria_bottle WHERE {value_col} IS NOT NULL")
})

sql_create <- glue(
  "CREATE OR REPLACE TABLE picoplankton_bacteria_measurement AS
   SELECT ROW_NUMBER() OVER (ORDER BY bottle_id, measurement_type) AS measurement_id, *
   FROM (
     {paste(sql_parts, collapse = '\nUNION ALL\n')}
   ) sub")
dbExecute(con, sql_create)

n_meas <- dbGetQuery(con, "SELECT COUNT(*) FROM picoplankton_bacteria_measurement")[[1]]
cat(glue("picoplankton_bacteria_measurement: {format(n_meas, big.mark=',')} rows"), "\n")

for (col in pb_meas_map$value_col) {
  tryCatch(
    dbExecute(con, glue('ALTER TABLE picoplankton_bacteria_bottle DROP COLUMN "{col}"')),
    error = function(e) NULL)
}
```

## Register Measurement Types

Registers the four new `measurement_type` codes into the shared
`metadata/measurement_type.csv` and loads the `measurement_type` table into
this connection -- an earlier draft referenced `measurement_type` as a FK
target in Schema Documentation / Validate without ever creating it, which
would fail. Pattern matches `ingest_calcofi_phyllosoma.qmd`'s `finalize` step.

```{r}
#| label: register-measurement-types
pb_types <- tribble(
  ~measurement_type, ~description,                              ~units,
  "het_bacteria",     "Heterotrophic bacteria abundance (FCM)",  "number/ml",
  "prochlorococcus",  "Prochlorococcus abundance (FCM); ~80% populated, see Q01", "number/ml",
  "synechococcus",    "Synechococcus abundance (FCM)",           "number/ml",
  "picoeukaryotes",   "Picoeukaryote abundance (FCM)",           "number/ml") |>
  mutate(is_canonical = TRUE,
         grain        = "obs",
         `_source_column` = case_when(
           measurement_type == "het_bacteria"    ~ "Heterotrophic Bacteria (number/ml)",
           measurement_type == "prochlorococcus" ~ "Prochlorococcus (number/ml)",
           measurement_type == "synechococcus"   ~ "Synechococcus (number/ml)",
           measurement_type == "picoeukaryotes"  ~ "Picoeukaryotes (number/ml)"),
         `_source_table` = "picoplankton_bacteria_measurement",
         `_source_datasets` = "cce-lter_picoplankton-bacteria",
         `_qual_column` = NA_character_, `_prec_column` = NA_character_)

# upsert so a units/description correction propagates on re-run, and keep the
# registry sorted so its on-disk order is deterministic across ingests
d_meas_type <- d_meas_type |>
  filter(!measurement_type %in% pb_types$measurement_type) |>
  bind_rows(pb_types) |>
  arrange(measurement_type)
new_types <- pb_types
write_csv(d_meas_type, meas_type_csv)
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
cat(glue("measurement_type: {nrow(d_meas_type)} types registered ({nrow(new_types)} new)"), "\n")
```

## Resolve Ship and Cruise Keys

`cruise_key` in the shared `cruise` table is a natural key in format
`YYYY-MM-NODC`, built by the real `calcofi4db::derive_cruise_key_on_casts()`
function (confirmed from `ship.R`). That function needs a `ship_code` column
(matched against `ship.ship_nodc`) and a `datetime_utc` column, and accepts
the target table directly via `table_name=` (confirmed via
`ingest_calcofi_mets.qmd`'s usage) -- no need to rename anything to `casts`.
This source has no `ship_code` natively -- only `studyName` (e.g.
`2004-11-02-C-33RR`), whose trailing code (`33RR`) is the ship code embedded
in the string -- so extract that into a `ship_code` column, then call the
real function against `picoplankton_bacteria_bottle` directly rather than
reimplementing its SQL by hand (an earlier draft did this incorrectly).

```{r}
#| label: resolve-keys
load_prior_tables(
  con, parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables = c("ship", "cruise", "grid"), geom_tables = c("grid"), as_view = TRUE)

# extract trailing ship code from studyName (e.g. "2004-11-02-C-33RR" -> "33RR")
# NOTE (Q02, resolved): verified against the live release DB for 3 sample
# codes (33RR, 32NM, 31JD); all matched real casts.parquet rows. Still not
# verified across all 73 distinct studyName formats in this dataset.
dbExecute(con, "
  ALTER TABLE picoplankton_bacteria_bottle ADD COLUMN IF NOT EXISTS ship_code VARCHAR")
dbExecute(con, "
  UPDATE picoplankton_bacteria_bottle
  SET ship_code = regexp_extract(study_name, '[0-9A-Z]+$')")

cruise_key_result <- derive_cruise_key_on_casts(
  con, table_name = "picoplankton_bacteria_bottle", datetime_col = "datetime_utc")

n_ck <- dbGetQuery(con, "SELECT COUNT(*) FROM picoplankton_bacteria_bottle WHERE cruise_key IS NOT NULL")[[1]]
n_all <- dbGetQuery(con, "SELECT COUNT(*) FROM picoplankton_bacteria_bottle")[[1]]
cat(glue("cruise_key match: {n_ck}/{n_all} ({round(100*n_ck/n_all,1)}%)"), "\n")
```

## Add Spatial

```{r}
#| label: spatial
add_point_geom(con, "picoplankton_bacteria_bottle", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "picoplankton_bacteria_bottle") |> datatable(caption = "Grid assignment")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Schema Documentation

```{r}
#| label: schema
pb_rels <- list(
  primary_keys = list(
    picoplankton_bacteria_bottle = "bottle_id",
    picoplankton_bacteria_measurement = "measurement_id",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "picoplankton_bacteria_measurement", column = "bottle_id",
         ref_table = "picoplankton_bacteria_bottle", ref_column = "bottle_id"),
    list(table = "picoplankton_bacteria_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(
  con, tables = c("picoplankton_bacteria_bottle", "picoplankton_bacteria_measurement",
                  "measurement_type", "dataset"),
  rels = pb_rels,
  colors = list(lightblue = c("picoplankton_bacteria_bottle", "picoplankton_bacteria_measurement"),
                lightyellow = "measurement_type", white = "dataset"))

build_relationships_json(
  rels = pb_rels, output_dir = dir_parquet, provider = provider, dataset = dataset)
```

## Validate

```{r}
#| label: validate
results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")

# NULL cruise_key is the expected unmatched remainder pending confirmation of
# study_name -> cruise_key format (see Questions, Q02); not a hard failure.
n_dup <- dbGetQuery(con,
  "SELECT COUNT(*) FROM (SELECT bottle_id, COUNT(*) n FROM picoplankton_bacteria_bottle GROUP BY bottle_id HAVING COUNT(*)>1)")[[1]]
cat(glue("picoplankton_bacteria_bottle bottle_id duplicates: {n_dup}"), "\n")

n_orphan <- dbGetQuery(con,
  "SELECT COUNT(*) FROM picoplankton_bacteria_measurement m
   LEFT JOIN picoplankton_bacteria_bottle b ON m.bottle_id = b.bottle_id
   WHERE b.bottle_id IS NULL")[[1]]
cat(glue("Orphan measurements (no matching bottle): {n_orphan}"), "\n")

n_orphan_type <- dbGetQuery(con,
  "SELECT COUNT(*) FROM picoplankton_bacteria_measurement m
   WHERE m.measurement_type NOT IN (SELECT measurement_type FROM measurement_type)")[[1]]
cat(glue("Orphan measurement_types (not registered): {n_orphan_type}"), "\n")
```

## Data Preview

```{r}
#| label: preview
cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='picoplankton_bacteria_bottle' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM picoplankton_bacteria_bottle LIMIT 100")) |>
  datatable(caption = "picoplankton_bacteria_bottle — first 100 rows", rownames = FALSE, filter = "top")

dbGetQuery(con, "SELECT * FROM picoplankton_bacteria_measurement LIMIT 100") |>
  datatable(caption = "picoplankton_bacteria_measurement — first 100 rows", rownames = FALSE, filter = "top")
```

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`) via `calcofi4db::emit_core_tables()`, the
same projection `release_database.qmd` uses to assemble the cross-dataset
release. This is a bottle-shaped sample — leaf grain `bottle_id`, and since the
export carries no cast-level event table the bottle is its own root — with an
`obs` headline of the four flow-cytometry counts. Those counts are an
environmental measurement vocabulary rather than taxa, so the arm emits
`realm = 'env'` with a NULL `taxon_key`.

```{r}
#| label: emit_core

core <- emit_core_tables(con, "cce-lter_picoplankton-bacteria")
cat(glue(
  "core projection — sample={core$sample %||% 0} obs={core$obs %||% 0} ",
  "sample_measurement={core$sample_measurement %||% 0}\n"))

# every measurement on a grid-matched bottle must reach obs
n_obs <- dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]]
n_exp <- dbGetQuery(con,
  "SELECT COUNT(*) FROM picoplankton_bacteria_measurement m
   JOIN picoplankton_bacteria_bottle b USING (bottle_id)
   WHERE b.grid_key IS NOT NULL")[[1]]
stopifnot("obs must be one row per grid-matched measurement" = n_obs == n_exp)
cat(glue("obs parity: {format(n_obs, big.mark=',')} rows"), "\n")

stopifnot(
  "every obs.sample_key must resolve in sample" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key)
                     WHERE s.sample_key IS NULL")[[1]] == 0)
# serve the retired per-dataset table names as VIEWs over the core, so
# in-notebook consumers and ad-hoc queries keep working against the old names
# (exact for every column the core models; see ?create_compat_views)
compat <- create_compat_views(con, "cce-lter_picoplankton-bacteria")
cat(glue("compat views over core: {paste(compat, collapse = ', ')}"), "\n")

```

## Write Parquet Outputs

```{r}
#| label: write-parquet
dir_create(dir_parquet)
mismatches <- list(cruise_keys = collect_cruise_key_mismatches(con, "picoplankton_bacteria_bottle"))
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
parquet_stats <- write_parquet_outputs(
  con = con, output_dir = dir_parquet,
  tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE, mismatches = mismatches)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)
parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata
d_tbls_rd <- read_csv(here("metadata/cce-lter/picoplankton-bacteria/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/cce-lter/picoplankton-bacteria/flds_redefine.csv"))
metadata_path <- build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           here("metadata/cce-lter/picoplankton-bacteria/metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned)
```

## Upload to GCS

```{r}
#| label: upload-gcs
sync_to_gcs(local_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
```

## Questions for Data Providers

Follow-up questions for CCE-LTER (Michael Landry), ranked by importance.
Tracked in `metadata/cce-lter/picoplankton-bacteria/questions.csv`.

```{r}
#| label: provider-questions
read_csv(here(glue("metadata/{provider}/{dataset}/questions.csv"))) |>
  arrange(factor(priority, c("blocker", "high", "normal")), id) |>
  select(priority, question, context, related_field, status) |>
  datatable(caption = "Questions for data providers (ranked)",
            options = list(dom = "t", pageLength = 20), rownames = FALSE)
```

## Cleanup

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
