---
title: "Ingest CCE-LTER ZooDB Holoplankton"
calcofi:
  target_name: ingest_cce_lter_zoodb
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/cce-lter_zoodb/manifest.json
  provider: cce-lter
  dataset: zoodb
  workflow_url: https://calcofi.io/workflows/ingest_cce-lter_zoodb.html
  questions_file: metadata/cce-lter/zoodb/questions.csv
  dataset_meta:
    dataset_name: ZooDB Holoplankton Community
    description: >
      Holoplankton abundance and carbon biomass by 33 higher taxa (copepods
      to family, plus amphipods, chaetognaths, salps, pyrosomes, pteropods,
      ostracods, foraminifera, radiolaria and more) from CalCOFI net tows,
      1951-2021, enumerated by microscopy (Ohman Lab / SIO Pelagic Invertebrate
      Collection). One row per (sample, taxon, measurement_type); unpooled tows
      are station-resolved, pooled samples are per-cruise regional composites.
      Raw (uncorrected) abundance per 1000 m3 and per m2, plus biomass mg C/m2.
    citation_main: ""
    citation_others: "Plankton sample analysis supported by NSF grants to M.D. Ohman and the CCE-LTER site, and by the SIO Pelagic Invertebrate Collection."
    link_calcofi_org: "https://calcofi.org/data/marine-ecosystem-data/zooplankton/"
    link_data_source: "https://oceaninformatics.ucsd.edu/zoodb/"
    link_others: []
    coverage_temporal: 1951-03 to 2021-05
    coverage_spatial: "CalCOFI lines 60-93 (Southern + Central California)"
    license: ""
    pi_names: Mark D. Ohman; Bertha Lavaniegos; Linsey Sala
  tables_owned:
    - {table: zoodb_sample}
    - {table: zoodb_measurement}
    - {table: zoodb_taxon}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#d9c2f0"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: [ZooDB](https://oceaninformatics.ucsd.edu/zoodb/) — the SIO Ocean
Informatics zooplankton database (Mark Ohman Lab, SIO Pelagic Invertebrate
Collection). Per-taxon CSVs are scripted from the portal's `save.php` endpoint
(public access) by `libs/download_zoodb.R` and consolidated to
`zoodb_holoplankton.csv` (see the Acquire step below + `by_taxon/_PROVENANCE.md`).

- **Provider**: `cce-lter` (curated by CCE-LTER / Ohman Lab; sibling of
  `cce-lter_euphausiids`, which the ZooDB methods note as a *separate* database)
- **Grain**: one row per **(sample × taxon × measurement_type)**. A measurement
  row means the taxon was *analyzed* for that sample; `measurement_value = 0`
  means analyzed-but-absent. Coverage varies by taxon/cruise — a **missing**
  (sample, taxon) is *not analyzed*, so we never impute zeros.
- **Samples** (506): 351 `unpooled` net tows (station-resolved) + 155 `pooled`
  per-cruise regional composites (Lavaniegos & Ohman 2007; no per-station
  position; a cruise/region may hold >1 composite, e.g. shallow vs deep strata).
- **Taxa** (33): higher taxa — all genera/species and phases/stages combined —
  standardized to WoRMS (`taxon_worms.csv`).
- **Measurements**: `zooplankton_abundance` (count/1000 m³),
  `zooplankton_abundance_areal` (count/m²), `zooplankton_biomass_carbon`
  (mg C/m²). Raw (uncorrected for net type).

This ingest **proceeds with documented provisional decisions** (cruise/ship
resolution, timezone, raw vs corrected, pooled positioning); each is tracked in
the *Questions for Data Providers* section.

```{mermaid}
graph LR
  A[zoodb_holoplankton.csv<br/>10,316 sample x taxon rows] --> B[zoodb_sample<br/>503 samples + keys]
  A --> C[zoodb_measurement<br/>long: abundance x2 + biomass]
  A --> D[zoodb_taxon<br/>33 higher taxa + WoRMS]
  C -.sample_id.-> B
  C -.taxon_id.-> D
  B -.ship/cruise/grid.-> E[(shared refs)]
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr, lubridate, purrr,
  readr, sf, stringr, tibble, tidyr, units, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))   # overwrite, dir_data

cc           <- read_calcofi_meta(here("ingest_cce-lter_zoodb.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")

dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))  dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

meas_type_csv <- here("metadata/measurement_type.csv")
d_meas_type   <- read_csv(meas_type_csv)
```

## Acquire Source Data

ZooDB is an interactive portal, not a file download. `libs/download_zoodb.R`
scripts its public login + `save.php` CSV export — the 33 higher taxa × 2
abundance units (`pooled=combined`, raw/uncorrected, all years/months/times/
regions) — and merges the two unit files per taxon (on the source `RowNumber`)
into `zoodb_holoplankton.csv`. It runs once then caches; set `overwrite_all =
TRUE` in `libs/ingest.R` to re-scrape. Exact query parameters are recorded in
`by_taxon/_PROVENANCE.md`.

```{r}
#| label: acquire

source(here("libs/download_zoodb.R"))
zoo_dir <- path_expand(glue("{dir_data}/cce-lter/ZooDB"))
zoo_csv <- download_zoodb(zoo_dir, overwrite = overwrite_all)
```

## Read Source Data

The consolidated long table (one row per taxon × sample) carries `N/A` for the
per-station fields on pooled rows, read as `NA`.

```{r}
#| label: read-source

stopifnot("zoodb_holoplankton.csv not found" = file_exists(zoo_csv))

# archive source (consolidated + per-taxon exports + portal method PDFs) to GCS
sync_to_gcs(
  local_dir  = zoo_dir,
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket     = "calcofi-files-public",
  exclude    = c(".DS_Store", "*.tmp", "*.gdoc"))

d_raw <- read_csv(zoo_csv, na = c("", "NA", "N/A"))
cat(glue("Read {format(nrow(d_raw), big.mark=',')} rows, {ncol(d_raw)} columns ",
         "from {basename(zoo_csv)}"), "\n")
```

## Clean and Type-Cast

Canonical field names follow `metadata/field_dictionary.csv`. A `sample_key`
uniquely identifies each sample: `U|cruise|towbegin|line|station` (unpooled) or
`P|cruise|date|region` (pooled).

```{r}
#| label: clean

d <- d_raw |>
  transmute(
    taxon_slug           = taxon,
    source               = str_to_lower(source),
    cruise_orig          = cruise,
    ship_name            = ship,
    date                 = as.Date(date),
    line                 = suppressWarnings(as.numeric(line)),
    station              = suppressWarnings(as.numeric(station)),
    region               = region,
    datetime_start_utc   = as_datetime(tow_begin),   # Q01: tz assumed local/ship
    datetime_end_utc     = as_datetime(tow_end),
    latitude             = suppressWarnings(as.numeric(latitude)),
    longitude            = suppressWarnings(as.numeric(longitude)),
    max_depth_m          = suppressWarnings(as.numeric(max_depth_m)),
    min_depth_m          = suppressWarnings(as.numeric(min_depth_m)),
    net_type             = net_type,
    n_tows_pooled        = suppressWarnings(as.integer(n_tows_pooled)),
    abundance_per_1000m3 = suppressWarnings(as.numeric(abundance_per_1000m3)),
    abundance_per_m2     = suppressWarnings(as.numeric(abundance_per_m2)),
    biomass_mgC_per_m2   = suppressWarnings(as.numeric(biomass_mgC_per_m2))) |>
  mutate(
    site_key = if_else(
      is.na(line) | is.na(station), NA_character_,
      sprintf("%05.1f %05.1f", line, station)),
    # pooled composites can co-occur per cruise/region (e.g. shallow vs deep
    # strata) — n_tows_pooled disambiguates the otherwise-identical key
    sample_key = if_else(
      source == "unpooled",
      paste("U", cruise_orig, datetime_start_utc, line, station, sep = "|"),
      paste("P", cruise_orig, date, region, n_tows_pooled, sep = "|")))

cat(glue(
  "Cleaned {format(nrow(d), big.mark=',')} (sample x taxon) rows; ",
  "{n_distinct(d$sample_key)} samples ",
  "({n_distinct(d$sample_key[d$source=='unpooled'])} unpooled, ",
  "{n_distinct(d$sample_key[d$source=='pooled'])} pooled); ",
  "{n_distinct(d$taxon_slug)} taxa"), "\n")
```

## Build Taxon Table (WoRMS)

`zoodb_taxon` is the 33 higher taxa with WoRMS AphiaIDs + classification,
resolved offline and cached in `metadata/cce-lter/zoodb/taxon_worms.csv`
(33/33 matched; 2 WoRMS-unaccepted concepts flagged, Q07).

```{r}
#| label: taxon

d_worms <- read_csv(here("metadata/cce-lter/zoodb/taxon_worms.csv"))

zoodb_taxon <- d_worms |>
  arrange(taxon_zoodb) |>
  transmute(
    taxon_id        = row_number(),
    taxon_zoodb, taxon_slug,
    aphia_id        = as.integer(aphia_id),
    scientific_name = scientific_name_accepted,
    rank, taxon_status = status,
    kingdom, phylum, class, order_taxon = order, family)   # 'order' is a SQL reserved word

stopifnot("taxon_slug mismatch source vs WoRMS map" =
            all(unique(d$taxon_slug) %in% zoodb_taxon$taxon_slug))
dbWriteTable(con, "zoodb_taxon", zoodb_taxon, overwrite = TRUE)
cat(glue("zoodb_taxon: {nrow(zoodb_taxon)} taxa ",
         "({sum(!is.na(zoodb_taxon$aphia_id))} WoRMS-matched, ",
         "{sum(zoodb_taxon$taxon_status=='unaccepted')} unaccepted [Q07])"), "\n")
```

## Build Sample Table + Resolve Keys

One row per distinct sample. Resolve `ship_key`/`ship_nodc` against the shared
ship registry, derive `cruise_key` as `YYYY-MM-{ship_nodc}` (validated against
`cruise`), and fall back to a unique year-month match for pooled samples (which
have no ship). Q01 (timezone) and Q02 (cruise mapping) are tracked.

```{r}
#| label: sample-keys

# distinct sample-level attributes (constant across taxa within a sample)
d_sample0 <- d |>
  distinct(
    sample_key, source, cruise_orig, ship_name, date,
    datetime_start_utc, datetime_end_utc, line, station, site_key, region,
    latitude, longitude, max_depth_m, min_depth_m, net_type, n_tows_pooled)
stopifnot("sample_key not unique over sample attributes" =
            n_distinct(d_sample0$sample_key) == nrow(d_sample0))

# shared reference tables from the ichthyo ingest
load_prior_tables(
  con, parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables = c("ship", "cruise", "grid"), geom_tables = c("grid"), as_view = TRUE)

d_ship <- dbGetQuery(con, "SELECT ship_key, ship_name, ship_nodc FROM ship") |>
  mutate(ship_name_norm = str_squish(str_to_upper(ship_name))) |>
  distinct(ship_name_norm, .keep_all = TRUE)

# year-month -> unique cruise_key (pooled samples + fallback)
cr_ym <- dbGetQuery(con, "SELECT cruise_key, date_ym FROM cruise") |>
  mutate(ym = format(as.Date(date_ym), "%Y%m")) |>
  group_by(ym) |>
  summarize(cruise_key_ym = if (n() == 1) cruise_key[1] else NA_character_,
            .groups = "drop")
valid_ck <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM cruise")$cruise_key

d_sample <- d_sample0 |>
  mutate(ship_name_norm = ship_name |>
           str_replace("^R/?V\\.?\\s+", "") |> str_to_upper() |> str_squish()) |>
  left_join(d_ship |> select(ship_name_norm, ship_key, ship_nodc),
            by = "ship_name_norm") |>
  mutate(
    ym = format(date, "%Y%m"),
    cruise_key_ship = if_else(
      is.na(ship_nodc), NA_character_,
      as.character(glue("{format(date, '%Y-%m')}-{ship_nodc}"))),
    cruise_key_ship = if_else(cruise_key_ship %in% valid_ck,
                              cruise_key_ship, NA_character_)) |>
  left_join(cr_ym, by = "ym") |>
  mutate(cruise_key = coalesce(cruise_key_ship, cruise_key_ym)) |>
  arrange(source, date, cruise_orig, region, line, station, sample_key) |>
  mutate(sample_id = row_number(), .before = 1)

# id <-> key map for the measurement join, then drop the key from the table
sample_map   <- d_sample |> select(sample_id, sample_key)
zoodb_sample <- d_sample |>
  select(sample_id, source, cruise_orig, cruise_key, ship_name, ship_key,
         date, datetime_start_utc, datetime_end_utc, line, station, site_key,
         region, latitude, longitude, max_depth_m, min_depth_m, net_type,
         n_tows_pooled)
dbWriteTable(con, "zoodb_sample", zoodb_sample, overwrite = TRUE)

n_ship   <- sum(!is.na(d_sample$ship_key))
n_cruise <- sum(!is.na(d_sample$cruise_key))
cat(glue(
  "zoodb_sample: {nrow(zoodb_sample)} samples; ",
  "ship_key {n_ship}/{nrow(d_sample)} ({round(100*n_ship/nrow(d_sample),1)}%), ",
  "cruise_key {n_cruise}/{nrow(d_sample)} ({round(100*n_cruise/nrow(d_sample),1)}%)"), "\n")
```

## Add Spatial

Unpooled tows get a point geometry and a CalCOFI `grid_key`; pooled samples have
no position (NULL geometry / grid, Q05).

```{r}
#| label: spatial

add_point_geom(con, "zoodb_sample", lon_col = "longitude", lat_col = "latitude")
grid_stats <- assign_grid_key(con, "zoodb_sample")
grid_stats |> datatable(caption = "Grid assignment (unpooled tows)")
```

## Pivot Measurements to Long Format

The three measured quantities pivot into long form, joined to `sample_id`
(via `sample_key`) and `taxon_id` (via `taxon_slug`). Explicit zeros (analyzed-
but-absent) are retained; only NA / non-finite values are dropped.

```{r}
#| label: measurement

meas_recode <- c(
  abundance_per_1000m3 = "zooplankton_abundance",
  abundance_per_m2     = "zooplankton_abundance_areal",
  biomass_mgC_per_m2   = "zooplankton_biomass_carbon")

zoodb_measurement <- d |>
  left_join(sample_map, by = "sample_key") |>
  left_join(zoodb_taxon |> select(taxon_id, taxon_slug), by = "taxon_slug") |>
  select(sample_id, taxon_id,
         abundance_per_1000m3, abundance_per_m2, biomass_mgC_per_m2) |>
  pivot_longer(
    cols = c(abundance_per_1000m3, abundance_per_m2, biomass_mgC_per_m2),
    names_to = "src_col", values_to = "measurement_value") |>
  filter(!is.na(measurement_value), is.finite(measurement_value)) |>
  mutate(measurement_type = unname(meas_recode[src_col])) |>
  arrange(sample_id, taxon_id, measurement_type) |>
  transmute(measurement_id = row_number(),
            sample_id, taxon_id, measurement_type, measurement_value)

dbWriteTable(con, "zoodb_measurement", zoodb_measurement, overwrite = TRUE)
cat(glue(
  "zoodb_measurement: {format(nrow(zoodb_measurement), big.mark=',')} rows ",
  "({format(sum(zoodb_measurement$measurement_value==0), big.mark=',')} explicit zeros = ",
  "analyzed-but-absent, Q04)"), "\n")
```

## Add Measurement Types

```{r}
#| label: add-measurement-type

zoo_types <- tibble(
  measurement_type = c("zooplankton_abundance", "zooplankton_abundance_areal",
                       "zooplankton_biomass_carbon"),
  description = c(
    "Zooplankton numerical abundance (volumetric density) per net tow, by higher taxon. Raw (uncorrected for net type).",
    "Zooplankton areal abundance (depth-integrated density) per net tow, by higher taxon. Raw (uncorrected).",
    "Zooplankton carbon biomass per net tow, by higher taxon. Raw (uncorrected for net type)."),
  units = c("count/1000m3", "count/m2", "mgC/m2"),
  is_canonical = c(TRUE, NA, TRUE),
  `_source_column` = c("abundance_per_1000m3", "abundance_per_m2", "biomass_mgC_per_m2"),
  `_source_table` = "zoodb_measurement",
  `_source_datasets` = "cce-lter_zoodb",
  `_qual_column` = NA_character_, `_prec_column` = NA_character_)

new_types <- zoo_types |> filter(!measurement_type %in% d_meas_type$measurement_type)
if (nrow(new_types) > 0) {
  d_meas_type <- bind_rows(d_meas_type, new_types)
  write_csv(d_meas_type, meas_type_csv)
  cat(glue("Added measurement type(s): {paste(new_types$measurement_type, collapse=', ')}"), "\n")
} else cat("zooplankton measurement types already registered\n")
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} dataset(s) registered"), "\n")
```

## Schema Documentation

```{r}
#| label: schema

zoodb_rels <- list(
  primary_keys = list(
    zoodb_sample      = "sample_id",
    zoodb_measurement = "measurement_id",
    zoodb_taxon       = "taxon_id",
    measurement_type  = "measurement_type"),
  foreign_keys = list(
    list(table = "zoodb_measurement", column = "sample_id",
         ref_table = "zoodb_sample", ref_column = "sample_id"),
    list(table = "zoodb_measurement", column = "taxon_id",
         ref_table = "zoodb_taxon", ref_column = "taxon_id"),
    list(table = "zoodb_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(
  con,
  tables = c("zoodb_sample", "zoodb_measurement", "zoodb_taxon",
             "measurement_type", "dataset"),
  rels = zoodb_rels,
  colors = list(
    lightblue   = c("zoodb_sample", "zoodb_measurement"),
    lightgreen  = "zoodb_taxon",
    lightyellow = "measurement_type",
    white       = "dataset"))

build_relationships_json(
  rels = zoodb_rels, output_dir = dir_parquet,
  provider = provider, dataset = dataset)
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
if (length(results$warnings) > 0)
  cat("Warnings:\n", paste("-", results$warnings, collapse = "\n"), "\n")

# NULLs in cross-dataset keys (cruise_key, ship_key, site_key, grid_key) are the
# EXPECTED unmatched / not-applicable remainder (pooled samples have no station;
# cruise/ship mapping is partial — Q02). Not a hard failure.
cov <- dbGetQuery(con,
  "SELECT
     AVG(CASE WHEN ship_key   IS NOT NULL THEN 1 ELSE 0 END) ship,
     AVG(CASE WHEN cruise_key IS NOT NULL THEN 1 ELSE 0 END) cruise,
     AVG(CASE WHEN grid_key   IS NOT NULL THEN 1 ELSE 0 END) grid
   FROM zoodb_sample")
cat(glue("Match coverage: ship_key {round(100*cov$ship,1)}%, ",
         "cruise_key {round(100*cov$cruise,1)}%, grid_key {round(100*cov$grid,1)}%"), "\n")

# FK integrity: every measurement resolves to a sample and a taxon
orphan <- dbGetQuery(con,
  "SELECT
     SUM(CASE WHEN s.sample_id IS NULL THEN 1 ELSE 0 END) AS no_sample,
     SUM(CASE WHEN t.taxon_id  IS NULL THEN 1 ELSE 0 END) AS no_taxon
   FROM zoodb_measurement m
   LEFT JOIN zoodb_sample s USING (sample_id)
   LEFT JOIN zoodb_taxon  t USING (taxon_id)")
cat(glue("Orphan measurements: {orphan$no_sample} w/o sample, {orphan$no_taxon} w/o taxon"), "\n")
```

## Data Preview

```{r}
#| label: preview-sample

samp_cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='zoodb_sample' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(samp_cols, collapse=', ')} FROM zoodb_sample LIMIT 100")) |>
  datatable(caption = "zoodb_sample — first 100 rows", rownames = FALSE, filter = "top")
```

```{r}
#| label: preview-measurement

dbGetQuery(con,
  "SELECT m.measurement_id, m.sample_id, x.taxon_zoodb, m.measurement_type, m.measurement_value
   FROM zoodb_measurement m JOIN zoodb_taxon x USING (taxon_id)
   ORDER BY m.measurement_id LIMIT 100") |>
  datatable(caption = "zoodb_measurement — first 100 rows", rownames = FALSE)
```

```{r}
#| label: preview-taxon

dbGetQuery(con,
  "SELECT t.taxon_id, t.taxon_zoodb, t.scientific_name, t.aphia_id, t.rank,
          t.phylum, t.class, COUNT(DISTINCT m.sample_id) AS n_samples
   FROM zoodb_taxon t LEFT JOIN zoodb_measurement m USING (taxon_id)
   GROUP BY ALL ORDER BY t.taxon_id") |>
  datatable(caption = "zoodb_taxon — 33 higher taxa (samples analyzed)", rownames = FALSE)
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

dir_create(dir_parquet)

mismatches <- list(
  measurement_types = collect_measurement_type_mismatches(
    con, here("metadata/measurement_type.csv")),
  cruise_keys = collect_cruise_key_mismatches(con, "zoodb_sample"))

parquet_stats <- write_parquet_outputs(
  con        = con,
  output_dir = dir_parquet,
  tables     = c("zoodb_sample", "zoodb_measurement", "zoodb_taxon",
                 "measurement_type", "dataset"),
  sort_by    = list(zoodb_measurement = "sample_id"),
  strip_provenance = FALSE,
  mismatches = mismatches)

parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata

d_tbls_rd <- read_csv(here("metadata/cce-lter/zoodb/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/cce-lter/zoodb/flds_redefine.csv"))

metadata_path <- build_metadata_json(
  con        = con,
  d_tbls_rd  = d_tbls_rd,
  d_flds_rd  = d_flds_rd,
  metadata_derived_csv = here("metadata/cce-lter/zoodb/metadata_derived.csv"),
  output_dir = dir_parquet,
  tables     = c("zoodb_sample", "zoodb_measurement", "zoodb_taxon"),
  set_comments = TRUE,
  provider   = provider,
  dataset    = dataset,
  workflow_url = cc$workflow_url,
  tables_owned = tables_owned)
```

## Upload to GCS

```{r}
#| label: upload-gcs

sync_to_gcs(
  local_dir  = dir_parquet,
  gcs_prefix = glue("ingest/{dir_label}"),
  bucket     = "calcofi-db")
```

## Questions for Data Providers

Follow-up questions for CCE-LTER (Mark Ohman, Bertha Lavaniegos, Linsey Sala),
**ranked by importance**. This ingest proceeded with the documented provisional
decisions noted above; `high` answers may change cruise/ship matching or the
timezone assumption on the next run. Tracked in
`metadata/cce-lter/zoodb/questions.csv`.

```{r}
#| label: provider-questions

read_csv(here(glue("metadata/{provider}/{dataset}/questions.csv"))) |>
  arrange(factor(priority, c("blocker", "high", "normal", "low")), id) |>
  select(priority, question, context, related_field, status) |>
  datatable(
    caption = "Questions for data providers (ranked by importance)",
    options = list(dom = "t", pageLength = 20), rownames = FALSE)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```

::: {.callout-caution collapse="true"}
## Session Info
```{r session_info}
devtools::session_info()
```
:::
