---
title: "Ingest CalCOFI CTD Cast Data"
calcofi:
  target_name: ingest_calcofi_ctd_cast
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/calcofi_ctd-cast/manifest.json
  modifies:
    - ship
  provider: calcofi
  dataset: ctd-cast
  workflow_url: https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html
  questions_file: metadata/calcofi/ctd-cast/questions.csv
  dataset_meta:
    dataset_name: CalCOFI CTD Cast Files
    description: >
      CTD downcast profiles with temperature, salinity, oxygen, fluorescence,
      and other sensors at standard depths.
    citation_main: "CalCOFI. (2023). CalCOFI CTD Cast Files. CalCOFI.org."
    link_calcofi_org: https://calcofi.org/data/oceanographic-data/ctd-cast-files/
    link_data_source: https://calcofi.org/data/oceanographic-data/ctd-cast-files/
    # measured from the published `sample` shard, not asserted: the previous values
    # (1993-01 to 2021-05; 28-37°N, -125 to -117°W) predated several years of
    # cruises and understated the western extent, and they ship to consumers via
    # the release's `dataset` table and the schema site.
    coverage_temporal: 1998-01 to 2026-04
    coverage_spatial: "29.8-37.8°N, -126.5 to -117.3°W"
  # publishes the consolidated core. obs carries the THINNED series (ctd_thin);
  # the full scan set ships as the supplemental obs_ctd_full. ctd_wide is retired:
  # ERDDAP now serves CTD through EDDTableFromDatabase over DuckDB views plus a
  # separate netCDF build, so the 1 GB wide file has no consumer (its whole-file
  # heap read is what OOM'd ERDDAP in the first place — see bench_erddap_ctd.qmd).
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (one row per physical cast)"}
    - {table: obs, shared: true, note: "core observations (env, thinned CTD profile)"}
    - {table: obs_ctd_full, supplemental: true, note: "full-resolution scans (~216M rows), opt-in"}
    - {table: measurement_type, shared: true, note: "shared registry across bottle/ctd/dic"}
  erd:
    color: "#e0ccf0"
---

## Overview

This notebook ingests CalCOFI CTD data from <https://calcofi.org/data/oceanographic-data/ctd-cast-files/>, downloads and unzips **final** and **preliminary** CTD files, normalizes into tidy tables (`ctd_cast`, `ctd_measurement`, `ctd_thin`, `ctd_summary`), and exports to Parquet for the CalCOFI integrated database.

### Key Features

1.  **Web Scraping**: Scrapes all CTD .zip download links from calcofi.org

2.  **Smart Filtering**:
    -   Downloads all .zip files for archival completeness
    -   Only unzips `final` and `preliminary` files
    -   Skips raw/cast/test files

3.  **Priority-based Selection**:
    -   For each cruise, selects final if available, otherwise preliminary
    -   Excludes raw/test/prodo cast files

4.  **Tidy Normalization**:
    -   `ctd_cast`: one row per unique cast (cruise/station/direction)
    -   `ctd_measurement`: long-format sensor readings at each depth (supplemental)
    -   `ctd_thin`: adaptively-thinned `ctd_measurement` — single direction, canonical types, ~10 m grid with inflections preserved; the headline CTD table
    -   `ctd_summary`: summary stats per station/depth/measurement_type across cast directions
    -   `measurement_type`: reference table for measurement codes

5.  **Standardized Workflow**: follows patterns from `ingest_calcofi_bottle.qmd` — deterministic UUIDs, GCS parquet uploads, metadata sidecars, calcofi4db utilities

6.  **[Findings](#findings)**: the report at the end — what is known to be wrong with this data, measured from the parquet that was just written and checked by the same QC rule registry `apps/ctd-qaqc` runs. It renders even when the heavy path is skipped (see [Check for Resumable State]), so it can be revised without an hour of recomputation.

## Setup

```{r}
#| label: setup

# chunk timing hook — prints elapsed time for each chunk
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    .time_it_t0 <<- Sys.time()
  } else {
    elapsed <- round(difftime(Sys.time(), .time_it_t0, units = "secs"), 1)
    tnow <- format(Sys.time(), "%H:%M:%S")
    message(glue::glue("R chunk {options$label}: {elapsed}s ~ {tnow}"))
  }
})
knitr::opts_chunk$set(time_it = TRUE)

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI / calcofi4db,
  CalCOFI / calcofi4r,
  DBI,
  dplyr,
  DT,
  fs,
  glue,
  here,
  httr,
  janitor,
  lubridate,
  mapview,
  plotly,
  purrr,
  ps,
  readr,
  rvest,
  sf,
  stringr,
  tibble,
  tidyr,
  zip,
  quiet = T
)

# common ingest settings (overwrite, dir_data)
source(here("libs/ingest.R"))

# say(): status output to STDOUT so it renders in the notebook.
#
# This notebook previously used message() for ~37 status lines. message() writes
# to stderr, which Quarto does not reliably surface under `code-fold: true` — so
# most of the diagnostics below were being computed and then never shown. The
# repo convention is cat() (see .claude/skills/ingest-new.md), but cat() adds no
# trailing newline, so this thin wrapper keeps message()'s line semantics.
say <- function(...) cat(..., "\n", sep = "")

# provider/dataset/metadata read from this file's authoritative YAML block
cc           <- read_calcofi_meta(here("ingest_calcofi_ctd-cast.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label <- glue("{provider}_{dataset}")
dir_dl <- path_expand(glue("{dir_data}/{provider}/{dataset}/download"))
dir_parquet <- here(glue("data/parquet/{dir_label}"))
db_path <- here(glue("data/wrangling/{dir_label}.duckdb"))
db_checkpoint <- here(glue("data/wrangling/{dir_label}_checkpoint.duckdb"))
dir_tmp <- here(glue("data/tmp/{dir_label}"))
url <- "https://calcofi.org/data/oceanographic-data/ctd-cast-files/"
dir_meta <- here(glue("metadata/{provider}/{dataset}"))

dir_create(c(dir_dl, dirname(db_path), dir_parquet, dir_tmp), recurse = TRUE)

# NOTE: nothing destructive happens here, and no connection is opened. Both wait
# for the resume decision below, which needs the scraped zip list to fingerprint
# the inputs — deleting the wrangling DB before knowing whether the heavy path
# will run would throw away the one table the Gantt appendix reads.

# source archival handled by sync_gd_to_gcs.qmd (rclone)
# the ctd-cast download/ dir is 62GB+ — too large for inline
# sync_to_gcs() through the GD FUSE mount

# load metadata
# read_measurement_type(), NOT read_csv(): the registry stores empty cells that a
# default read turns back into the literal string "NA", which then survives the
# na.omit() in the measurement-column registry below and becomes a phantom
# quality-flag column name. It also validates the file on the way in, so a
# corrupted write fails here rather than reaching the release. See CLAUDE.md.
d_meas_type <- calcofi4db::read_measurement_type(
  here("metadata/measurement_type.csv")
)
d_flds_rd <- read_csv(glue("{dir_meta}/flds_redefine.csv"), show_col_types = F)
d_tbls_rd <- read_csv(glue("{dir_meta}/tbls_redefine.csv"), show_col_types = F)
d_cruise_corrections <- read_csv(
  glue("{dir_meta}/cruise_key_corrections.csv"), show_col_types = F)
```

## Scrape CTD Download Links

This runs on **every** render, ahead of the resume decision below, because the set
of published `.zip` URLs is one of the two things that decide whether there is
anything to recompute. It costs one HTTP GET and falls back to the cached list
when calcofi.org is unreachable.

```{r}
#| label: d_zips

# read web page, extract all .zip download links; cache to CSV
cache_csv <- file.path(dir_meta, "ctd_zip_urls.csv")

d_zips <- tryCatch(
  {
    d <- read_html(url) |>
      html_nodes("a[href$='.zip']") |>
      html_attr("href") |>
      tibble(url = _) |>
      mutate(
        url = if_else(
          str_starts(url, "http"),
          url,
          paste0("https://calcofi.org", url)
        ),
        file_zip = basename(url),
        year = str_extract(url, "/(\\d{4})/", group = 1) |> as.integer(),
        month = str_extract(url, "-\\d{2}(\\d{2})", group = 1) |> as.integer(),
        cruise_key = str_extract(
          file_zip,
          "\\d{2}-(\\d{4}[A-Z0-9]{2,4})",
          group = 1
        ),
        zip_type = case_when(
          str_detect(file_zip, "CTDFinal") ~ "final",
          str_detect(file_zip, "CTDPrelim") ~ "preliminary",
          str_detect(file_zip, "CTDCast|CTD_Cast") ~ "raw",
          str_detect(file_zip, "CTDTest") ~ "test",
          TRUE ~ "unknown"
        )
      )
    write_csv(d, cache_csv)
    say(glue("Scraped {nrow(d)} zip URLs from {url}"))
    d
  },
  error = function(e) {
    if (file_exists(cache_csv)) {
      say(glue("Website unavailable, using cached URLs from {cache_csv}"))
      read_csv(cache_csv, show_col_types = FALSE)
    } else {
      stop(e)
    }
  }
)

stopifnot(!any(d_zips$zip_type == "unknown"))

d_zips |>
  mutate(
    file_zip = glue("<a href={url}>{file_zip}</a>")
  ) |>
  select(-url) |>
  dt(
    caption = "All zip files to download",
    fname = "ctd_zip_files",
    escape = F
  )
```

## Check for Resumable State

The heavy path here — download, parse, pivot — is about an hour, and this notebook
is re-rendered mostly for reasons that have nothing to do with its inputs: a
narrative edit, a new diagnostic, a corrected sentence. Paying an hour to prove
that nothing changed is what stops a pipeline notebook being usable as a living
document, so the decision is made on evidence rather than on a flag: **skip the
heavy path when the parquet outputs are complete and the inputs they were built
from still fingerprint the same.**

The inputs are the published zip URL list scraped above, plus the metadata
registries that steer the pivot. `calcofi4db::input_fingerprint()` hashes them and
records the result beside the wrangling database; a *missing* input hashes as
`<missing>` rather than being skipped, so deleting a corrections file invalidates
the outputs instead of reading as "no change".

When the heavy path is skipped, **every chunk from here to the GCS upload is
displayed but not evaluated** — their code is still the record of how the tables
were built, but the tables themselves are not rebuilt, so those sections show no
output. The [Findings](#findings) section at the end reads the **parquet** rather
than the wrangling database and therefore always runs; that is what makes this
worth doing, and what keeps a skipped render honest rather than merely quiet.

```{r}
#| label: check_resume

# `overwrite` in libs/ingest.R is TRUE for every ingest and cannot be flipped
# globally — the comment there records that trying it broke a different notebook,
# because the skip path is not uniformly safe. So this notebook decides for itself.
# CTD_FORCE_REBUILD=TRUE (or overwrite_all) runs the heavy path regardless.
fingerprint_path <- here(glue("data/wrangling/{dir_label}_inputs.json"))

# kept OUT of dir_parquet on purpose: sync_to_gcs() mirrors that directory to a
# world-readable bucket, and this is local state about one machine's outputs.
fp <- input_fingerprint(
  # ONLY inputs that can change the OUTPUTS belong here. measurement_qual.csv is
  # deliberately absent: it is a vocabulary the diagnostics decode against, never
  # something the pivot reads, so hashing it would force an hour of rebuild for a
  # documentation edit that cannot alter a single row.
  files = c(
    here("metadata/measurement_type.csv"),
    here("metadata/core_dictionary.csv"),
    file.path(dir_meta, "flds_redefine.csv"),
    file.path(dir_meta, "tbls_redefine.csv"),
    file.path(dir_meta, "cruise_key_corrections.csv"),
    file.path(dir_meta, "metadata_derived.csv")),
  # the engine version is an input too: a change to calcofi4db's projection or
  # parsing would otherwise leave the fingerprint identical and keep outputs that
  # the current code would no longer produce
  values = c(sort(d_zips$url),
             paste0("calcofi4db ", as.character(packageVersion("calcofi4db")))))
fp_prior <- read_input_fingerprint(fingerprint_path)
fp_same  <- !is.null(fp_prior) && identical(fp_prior$hash, fp$hash)

# parquet is "complete" when every non-supplemental manifest table is on disk
parquet_ok    <- FALSE
manifest_path <- file.path(dir_parquet, "manifest.json")
if (file_exists(manifest_path)) {
  mf       <- jsonlite::read_json(manifest_path)
  expected <- setdiff(mf$tables, unlist(mf$supplemental))
  parquet_ok <- all(vapply(expected, function(tbl) {
    file_exists(file.path(dir_parquet, paste0(tbl, ".parquet"))) ||
      dir_exists(file.path(dir_parquet, tbl))
  }, logical(1)))
}

force_rebuild <- overwrite_all ||
  isTRUE(as.logical(Sys.getenv("CTD_FORCE_REBUILD", "FALSE")))
parquet_complete <- parquet_ok && fp_same && !force_rebuild

if (parquet_complete) {
  say(glue(
    "Inputs unchanged and parquet complete ({length(mf$tables)} tables, ",
    "{format(mf$total_rows, big.mark = ',')} rows) — ",
    "skipping download/parse/pivot"))
  say(glue("  fingerprint {substr(fp$hash, 1, 12)}, recorded {fp_prior$recorded_at}"))
} else if (force_rebuild) {
  say("Rebuilding: forced (CTD_FORCE_REBUILD / overwrite_all)")
} else if (parquet_ok && !fp_same) {
  say("Rebuilding: inputs changed since the recorded fingerprint —")
  for (f in changed_inputs(fp, fp_prior)) say(glue("  {f}"))
} else {
  say("Rebuilding: no complete parquet output found")
}

# --- only now touch anything -------------------------------------------------
# overwrite keeps the checkpoint DB, the downloads, and (importantly) dir_parquet,
# so write_parquet_outputs() can content-hash dedup against the prior run and only
# changed partitions are re-written and re-uploaded.
if (overwrite && !parquet_complete) {
  if (file_exists(db_path)) file_delete(db_path)
  # clear any stale WAL/tmp from an interrupted run so the restored checkpoint
  # DB opens cleanly (avoids "WAL checkpoint iteration does not match" errors)
  db_wal <- paste0(db_path, ".wal")
  db_tmp <- paste0(db_path, ".tmp")
  if (file_exists(db_wal)) file_delete(db_wal)
  if (dir_exists(db_tmp))  dir_delete(db_tmp)
  if (overwrite_all) {
    if (dir_exists(dir_parquet))    dir_delete(dir_parquet)
    if (file_exists(db_checkpoint)) file_delete(db_checkpoint)
    rds_files <- list.files(dir_tmp, pattern = "\\.rds$", full.names = TRUE)
    if (length(rds_files) > 0) file_delete(rds_files)
    say("Deleted parquet, checkpoint DB, and RDS intermediates")
  }
}

# restore from checkpoint DB if available (skips expensive read+bind+filter)
if (!parquet_complete && file_exists(db_checkpoint) && !file_exists(db_path)) {
  file_copy(db_checkpoint, db_path, overwrite = TRUE)
  say(glue("Restored from checkpoint: {db_checkpoint}"))
}

con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
load_duckdb_extension(con, "icu")

# limit memory to half of system RAM so DuckDB spills to disk instead of consuming all RAM
mem_gb <- ps::ps_system_memory()$total / 1024^3 / 2 |> floor()
dbExecute(con, glue("SET memory_limit = '{mem_gb}GB'"))
dbExecute(con, glue("SET temp_directory = '{dir_tmp}'"))

# a second, cheaper resume: ctd_raw already pivoted from a prior interrupted run
has_ctd_raw <- FALSE
if (!parquet_complete) {
  has_ctd_raw <- "ctd_raw" %in% DBI::dbListTables(con)
  if (has_ctd_raw) {
    n_raw <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_raw")$n
    say(glue(
      "Checkpoint: ctd_raw already loaded ",
      "({format(n_raw, big.mark = ',')} rows) — ",
      "skipping read+bind+filter"))
  }
}

# set eval for read+bind+filter chunks
skip_read_bind <- parquet_complete || has_ctd_raw
knitr::opts_chunk$set(eval = !skip_read_bind)
```

## Prime Downloads from GCS Source (optional)

The authoritative source for the CTD `.zip` files is the organization Shared Drive
("CalCOFI Data Folder"), mirrored to
`gs://calcofi-files-public/_sync/calcofi/ctd-cast/download/` by
`scripts/sync_gdrive_to_gcs.sh`. If that source exposes zips, copy them into
`dir_dl` so the next chunk simply unzips them rather than re-scraping calcofi.org.
This is a no-op (falls back to calcofi.org) until the GCS source is populated; set
`CTD_ZIP_SOURCE=""` to disable, or override it to point elsewhere.

```{r}
#| label: prime_zips_from_gcs

# authoritative zip source (gdrive→gcs); empty string disables priming
ctd_zip_source <- Sys.getenv(
  "CTD_ZIP_SOURCE",
  "gcs-calcofi:calcofi-files-public/_sync/calcofi/ctd-cast/download")

prime_zips_from_gcs <- function(src, dest_dir) {
  # skip if disabled or rclone unavailable
  if (src == "" || Sys.which("rclone") == "")
    return(invisible(0L))
  # system2() does NOT quote its args — it pastes them into a shell command — and
  # dest_dir is under "~/My Drive/…", so an unquoted path split on the space and
  # rclone saw three positional arguments instead of two:
  #   Command copy needs 2 arguments maximum: you provided 3 non flag arguments:
  #   ["gcs-calcofi:…/download" "/Users/bbest/My" "Drive/projects/…/download"]
  # It then failed SILENTLY, because the exit status was never checked: the run
  # printed "Priming N zip(s)", primed nothing, and fell through to scraping
  # calcofi.org. Harmless where the zips are already local; a slow surprise on a
  # fresh machine.
  n_src <- tryCatch(
    length(system2(
      "rclone", c("lsf", shQuote(src), "--include", shQuote("*.zip"),
                  "--max-depth", "1"),
      stdout = TRUE, stderr = FALSE)),
    error = function(e) 0L)
  if (n_src == 0) {
    say(glue("No zips at {src} — skipping GCS prime (will use calcofi.org)"))
    return(invisible(0L))
  }
  say(glue("Priming {n_src} zip(s) from {src} → {dest_dir}"))
  status <- system2("rclone", c(
    "copy", shQuote(src), shQuote(dest_dir),
    "--include", shQuote("*.zip"), "--max-depth", "1",
    "--transfers", "8", "--checkers", "16"))
  if (!identical(status, 0L)) {
    say(glue("GCS prime FAILED (rclone exit {status}) — falling back to ",
             "scraping calcofi.org. This is slower but not fatal."))
    return(invisible(0L))
  }
  invisible(n_src)
}

prime_zips_from_gcs(ctd_zip_source, dir_dl)
```

## Download and Unzip Files

```{r}
#| label: download_and_unzip

download_and_unzip <- function(url, dest_dir, zip_type, unzip = TRUE) {
  file_zip <- basename(url)
  dest_file <- file.path(dest_dir, file_zip)
  dir_unzip <- file.path(dest_dir, str_remove(file_zip, "\\.zip$"))

  if (file_exists(dest_file)) {
    say(glue("Already exists: {file_zip}"))
  } else {
    say(glue("Downloading: {file_zip}"))
    tryCatch(
      download.file(url, dest_file, mode = "wb"),
      error = function(e) {
        warning(glue("Failed to download {file_zip}: {e$message}"))
        if (file_exists(dest_file)) {
          file_delete(dest_file)
        }
        return(invisible(NULL))
      }
    )
    if (!file_exists(dest_file)) return(invisible(NULL))
  }

  if (unzip && dir_exists(dir_unzip)) {
    say(glue("Already unzipped: {file_zip}"))
  } else {
    if (zip_type %in% c("final", "preliminary")) {
      say(glue("Unzipping: {file_zip}"))
      dir_create(dir_unzip, recurse = TRUE)
      unzip(dest_file, exdir = dir_unzip)
    } else if (unzip) {
      say(glue("Skipping unzip (not final/preliminary): {file_zip}"))
    }
  }
}

d_zips |>
  pwalk(function(url, file_zip, zip_type, ...) {
    download_and_unzip(url, dir_dl, zip_type, unzip = TRUE)
  })

say("Download and extraction complete!")
```

## Find and Prioritize CTD Data Files

```{r}
#| label: d_csv

d_csv <- tibble(
  path = list.files(
    dir_dl,
    pattern = "\\.csv$",
    recursive = TRUE,
    full.names = TRUE
  )
) |>
  mutate(
    file_csv = basename(path),
    path_unzip = str_replace(path, glue("{dir_dl}/"), ""),
    dir_unzip = str_extract(path_unzip, "^[^/]+"),
    cruise_key = str_extract(
      path_unzip,
      "\\d{2}-(\\d{4}[A-Z0-9]{2,4})_.*",
      group = 1
    ),
    data_stage = case_when(
      str_detect(path_unzip, "Final.*db[_|-]csv")  ~ "final",
      str_detect(path_unzip, "Prelim.*db[_|-]csv") ~ "preliminary",
      # prelim CSVs directly in unzipped dir (no db-csv subfolder)
      str_detect(dir_unzip, "Prelim") &
        path_unzip == paste0(dir_unzip, "/", file_csv) ~ "preliminary",
      # edge case: 2111SR has CSVs only in csvs-plots subfolder
      cruise_key == "2111SR" &
        str_detect(
          path_unzip,
          "Prelim.*csvs-plots.*2111SR.*csv"
        ) ~ "preliminary",
      .default = NA_character_
    ),
    cast_dir = case_when(
      str_detect(file_csv, regex("U\\.csv$", ignore_case = T)) ~ "U",
      str_detect(file_csv, regex("D\\.csv$", ignore_case = T)) ~ "D"
    ),
    priority = case_when(
      data_stage == "final" ~ 1,
      data_stage == "preliminary" ~ 2,
      TRUE ~ 3
    )
  ) |>
  relocate(cruise_key, path_unzip) |>
  arrange(cruise_key, path_unzip) |>
  filter(data_stage %in% c("final", "preliminary"))

# for each cruise, keep only final if available, otherwise preliminary
d_priority <- d_csv |>
  group_by(cruise_key) |>
  summarize(
    best_priority = min(priority),
    .groups = "drop"
  )

d_csv <- d_csv |>
  inner_join(d_priority, by = "cruise_key") |>
  filter(priority == best_priority) |>
  select(-best_priority)

cruises_csv_notzip <- setdiff(
  unique(d_csv$cruise_key),
  unique(d_zips$cruise_key)
) |>
  sort()
stopifnot(length(cruises_csv_notzip) == 0)

cruises_zip_notcsv <- setdiff(
  unique(d_zips$cruise_key),
  unique(d_csv$cruise_key)
) |>
  sort()

d_csv |>
  select(-any_of(c("path", "data", "col_empties", "col_types"))) |>
  relocate(cruise_key, path_unzip) |>
  arrange(cruise_key, path_unzip) |>
  dt(
    caption = "Files to Ingest",
    fname = "ctd_files_to_ingest"
  )
```

## Read and Standardize CTD Files

Remove repeat header rows that can occur within files.

```{r}
#| label: read_csv

d_csv <- d_csv |>
  arrange(basename(path)) |>
  mutate(
    data = map2(path, seq_along(path), \(path, idx) {
      say(glue("Reading {idx}/{nrow(d_csv)}: {basename(path)}"))

      all_lines <- read_lines(path)
      header_line <- all_lines[1]

      # find repeat header rows (excluding row 1)
      repeat_header_rows <- which(all_lines[-1] == header_line)

      if (length(repeat_header_rows) > 0) {
        all_lines <- all_lines[-(repeat_header_rows + 1)]
        tmp_file <- tempfile(fileext = ".csv")
        write_lines(all_lines, tmp_file)
        data <- read_csv(tmp_file, guess_max = Inf, show_col_types = F) |>
          clean_names()
        file_delete(tmp_file)
      } else {
        data <- read_csv(path, guess_max = Inf, show_col_types = F) |>
          clean_names()
      }
      data
    }),
    nrows = map_int(data, ~ if (is.null(.x)) 0 else nrow(.x))
  )

d_csv |>
  arrange(cruise_key, file_csv) |>
  select(cruise_key, file_csv, nrows) |>
  dt(
    caption = "Number of rows read per file",
    fname = "ctd_rows_per_file"
  ) |>
  formatCurrency("nrows", currency = "", digits = 0, mark = ",")
```

## Detect and Correct Column Type Mismatches

```{r}
#| label: d_bind

d_csv <- d_csv |>
  mutate(
    col_empties = map(data, \(x) {
      tibble(
        col_name = names(x),
        n_empty = map_int(x, \(col) sum(is.na(col)))
      ) |>
        filter(n_empty == nrow(x)) |>
        pull(col_name)
    }),
    col_types = map2(data, col_empties, \(x, y) {
      tibble(
        col_name = names(x),
        col_type = map_chr(x, \(col) class(col)[1])
      ) |>
        filter(!col_name %in% y)
    })
  )

# find most common type for each column across all files
d_types <- d_csv |>
  select(path, col_types) |>
  unnest(col_types) |>
  count(col_name, col_type) |>
  group_by(col_name) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(col_name, expected_type = col_type)

d_mismatches <- d_csv |>
  select(cruise_key, path, col_types) |>
  unnest(col_types) |>
  left_join(d_types, by = "col_name") |>
  filter(col_type != expected_type) |>
  arrange(col_name, path)

if (nrow(d_mismatches) > 0) {
  say("Type mismatches detected - converting columns...")
}

# bind data, converting mismatched columns to expected type
d_bind <- d_csv |>
  mutate(
    data = map2(data, path, \(x, p) {
      x_mismatches <- d_mismatches |>
        filter(path == p)

      if (nrow(x_mismatches) > 0) {
        for (i in 1:nrow(x_mismatches)) {
          col <- x_mismatches$col_name[i]
          expected <- x_mismatches$expected_type[i]
          na_before <- sum(is.na(x[[col]]))

          suppressWarnings({
            x[[col]] <- switch(
              expected,
              "numeric" = as.numeric(x[[col]]),
              "integer" = as.integer(x[[col]]),
              "logical" = as.logical(x[[col]]),
              "character" = as.character(x[[col]]),
              x[[col]]
            )
          })

          na_after <- sum(is.na(x[[col]]))
          na_generated <- na_after - na_before

          if (na_generated > 0) {
            say(glue(
              "  {basename(p)}: {col} ({x_mismatches$col_type[i]} -> {expected}) generated {na_generated} NAs"
            ))
          }

          d_mismatches[
            d_mismatches$path == p & d_mismatches$col_name == col,
            "nas_generated"
          ] <<- na_generated
        }
      }
      x
    })
  ) |>
  unnest(data)

if (nrow(d_mismatches) > 0) {
  d_mismatches |>
    group_by(col_name, expected_type, col_type) |>
    summarize(
      n_files = n(),
      total_nas = sum(nas_generated, na.rm = TRUE),
      files = paste(basename(path), collapse = "; "),
      .groups = "drop"
    ) |>
    arrange(desc(total_nas)) |>
    dt(
      caption = "Type mismatches by column",
      fname = "ctd_type_mismatches_by_column"
    ) |>
    formatCurrency(
      c("n_files", "total_nas"),
      currency = "",
      digits = 0,
      mark = ","
    )
}

# reconcile cruise_key from study where mismatched
d_bind <- d_bind |>
  mutate(
    cruise_key = if_else(
      !is.na(study) &
        cruise_key != study &
        study != "Study",
      study,
      cruise_key
    ),
    `_source_file` = path_unzip
  ) |>
  relocate(`_source_file`, .after = cruise_key) |>
  select(
    -path_unzip,
    -path,
    -file_csv,
    -dir_unzip,
    -priority,
    -project,
    -study,
    -nrows,
    -col_empties,
    -col_types
  )

# fill missing ord_occ from cast_id
d_bind <- d_bind |>
  mutate(
    ord_occ = if_else(
      is.na(ord_occ) & !is.na(cast_id),
      str_extract(cast_id, "_(\\d{3})", group = 1),
      ord_occ
    )
  )

# normalize raw coordinate column names (CalCOFI CTD CSV ships Lon_Dec/Lat_Dec ->
# clean_names lon_dec/lat_dec) to the canonical dictionary names used downstream
d_bind <- d_bind |>
  rename(longitude = lon_dec, latitude = lat_dec)

# save intermediate for resumability
write_rds(d_bind, glue("{dir_tmp}/d_bind_pre_datetime.rds"), compress = "gz")
```

## Format Date-Time Column

```{r}
#| label: format_datetime

d_bind <- read_rds(glue("{dir_tmp}/d_bind_pre_datetime.rds"))

d_bind <- d_bind |>
  mutate(
    date_time_utc = trimws(date_time_utc),
    date_time_format = case_when(
      str_detect(
        date_time_utc,
        "^\\d{1,2}-[A-z]{3}-\\d{4} \\d{1,2}:\\d{1,2}:\\d{1,2}$"
      ) ~ "dmy_hms",
      str_detect(
        date_time_utc,
        "^\\d{1,2}/\\d{1,2}/\\d{4} \\d{1,2}:\\d{1,2}$"
      ) ~ "mdy_hm",
      TRUE ~ "unknown"
    ),
    datetime_start_utc = NA,
    datetime_start_utc = if_else(
      date_time_format == "dmy_hms",
      suppressWarnings(dmy_hms(date_time_utc, tz = "UTC")),
      datetime_start_utc
    ),
    datetime_start_utc = if_else(
      date_time_format == "mdy_hm",
      suppressWarnings(mdy_hm(date_time_utc, tz = "UTC")),
      datetime_start_utc
    )
  ) |>
  relocate(date_time_format, datetime_start_utc, .after = date_time_utc) |>
  arrange(datetime_start_utc, depth)

stopifnot(all(!is.na(d_bind$datetime_start_utc)))

write_rds(d_bind, glue("{dir_tmp}/d_bind_post_datetime.rds"), compress = "gz")

d_bind |>
  group_by(cruise_key, `_source_file`, date_time_format) |>
  summarize(
    n_records = n(),
    datetime_min = min(datetime_start_utc),
    datetime_max = max(datetime_start_utc),
    .groups = "drop"
  ) |>
  arrange(cruise_key, `_source_file`, date_time_format) |>
  pivot_wider(
    names_from = date_time_format,
    values_from = n_records,
    values_fill = NA_real_
  ) |>
  relocate(dmy_hms, mdy_hm, .after = cruise_key) |>
  dt(
    caption = "Date-Time formats detected by cruise_key and source_file",
    fname = "ctd_datetime_formats"
  ) |>
  formatDate(c("datetime_min", "datetime_max"), method = "toLocaleString")
```

## Pseudo-NA Values and Fill Missing Coordinates

```{r}
#| label: fill_lonlat

d_bind <- read_rds(glue("{dir_tmp}/d_bind_post_datetime.rds"))

pseudoNA_values <- c(-9.99e-29, -99)

d_bind <- d_bind |>
  mutate(
    is_lnsta_pseudoNA = ifelse(
      as.numeric(line) == 0 & as.numeric(sta) == 0,
      TRUE,
      FALSE
    ),
    is_lon_pseudoNA = map_lgl(longitude, ~ some(pseudoNA_values, near, .x)),
    is_lat_pseudoNA = map_lgl(latitude, ~ some(pseudoNA_values, near, .x)),
    longitude = if_else(
      is_lon_pseudoNA | is_lat_pseudoNA,
      NA_real_,
      longitude
    ),
    latitude = if_else(
      is_lon_pseudoNA | is_lat_pseudoNA,
      NA_real_,
      latitude
    ),
    lon_lnst = if_else(
      !is.na(line) & !is.na(sta) & !is_lnsta_pseudoNA,
      as.numeric(sf_project(
        from = "+proj=calcofi",
        to = "+proj=longlat +datum=WGS84",
        pts = cbind(x = as.numeric(line), y = as.numeric(sta))
      )[, 1]),
      NA_real_
    ),
    lat_lnst = if_else(
      !is.na(line) & !is.na(sta) & !is_lnsta_pseudoNA,
      as.numeric(sf_project(
        from = "+proj=calcofi",
        to = "+proj=longlat +datum=WGS84",
        pts = cbind(x = as.numeric(line), y = as.numeric(sta))
      )[, 2]),
      NA_real_
    ),
    longitude = if_else(is.na(longitude), lon_lnst, longitude),
    latitude = if_else(is.na(latitude), lat_lnst, latitude)
  )

stopifnot(sum(is.na(d_bind$longitude)) == 0)
stopifnot(sum(is.na(d_bind$latitude)) == 0)

d_bind |>
  filter(is_lon_pseudoNA | is_lat_pseudoNA) |>
  group_by(cruise_key) |>
  summarize(
    n_lon_pseudoNA = sum(is_lon_pseudoNA),
    n_lat_pseudoNA = sum(is_lat_pseudoNA)
  ) |>
  dt(
    caption = glue(
      "Cruises with pseudo-NAs ({paste(pseudoNA_values, collapse = ',')}) ",
      "set to NA, then filled from line/station coordinates."
    ),
    fname = "ctd_cruises_pseudoNA_lonlat"
  ) |>
  formatCurrency(
    c("n_lon_pseudoNA", "n_lat_pseudoNA"),
    currency = "",
    digits = 0,
    mark = ","
  )
```

## Distance Filtering

```{r}
#| label: pts_distance_from_lnst

max_dist_dec_lnst_km <- 10

pts <- d_bind |>
  st_as_sf(coords = c("longitude", "latitude"), remove = F, crs = 4326) |>
  mutate(
    geom_lnst = purrr::map2(lon_lnst, lat_lnst, \(x, y) st_point(c(x, y))) |>
      st_sfc(crs = 4326),
    dist_dec_lnst_km = st_distance(geometry, geom_lnst, by_element = T) |>
      units::set_units(km) |>
      units::drop_units(),
    is_dist_dec_lnst_within_max = dist_dec_lnst_km <= max_dist_dec_lnst_km
  ) |>
  st_join(
    calcofi4r::cc_grid |>
      rename(any_of(c(site_key = "sta_key"))) |>
      select(grid_site = site_key),
    join = st_intersects
  )
st_agr(pts) <- "constant"
```

### View sample cruise with excess distance points

```{r}
#| label: view_badcr

badcr_id <- "1507OC"
pts_badcr <- pts |>
  filter(cruise_key == badcr_id)

bb_badcr <- st_bbox(pts_badcr) |> st_as_sfc()

mapView(bb_badcr) +
  mapView(
    pts_badcr |>
      filter(is_dist_dec_lnst_within_max) |>
      slice_sample(n = 1000) |>
      bind_rows(
        pts_badcr |>
          filter(!is_dist_dec_lnst_within_max)
      ) |>
      select(
        cruise_key,
        datetime_start_utc,
        longitude,
        latitude,
        sta_id,
        line,
        sta,
        dist_dec_lnst_km,
        is_dist_dec_lnst_within_max
      ),
    layer.name = glue(
      "Cruise {badcr_id}<br>distance (km)<br>lon/lat to line/station"
    ),
    zcol = "dist_dec_lnst_km",
    cex = 5,
    alpha = 0.5
  )
```

### Filter points

```{r}
#| label: filter_pts

d_pts_cruise_filt_smry <- pts |>
  st_drop_geometry() |>
  group_by(cruise_key) |>
  filter(any(!is_dist_dec_lnst_within_max)) |>
  summarize(
    n_all = n(),
    n_outside_grid = sum(is.na(grid_site)),
    pct_outside_grid = n_outside_grid / n_all,
    n_gt_cutoff = sum(!is_dist_dec_lnst_within_max, na.rm = T),
    avg_dist_gt_cutoff_km = if_else(
      n_gt_cutoff > 0,
      mean(dist_dec_lnst_km[!is_dist_dec_lnst_within_max], na.rm = T),
      NA_real_
    ),
    pct_gt_cutoff = n_gt_cutoff / n_all,
    n_rm = sum(!is_dist_dec_lnst_within_max | is.na(grid_site), na.rm = T),
    pct_rm = n_rm / n_all,
    .groups = "drop"
  ) |>
  filter(n_rm > 0) |>
  arrange(desc(pct_rm))

pts_filt <- pts |>
  filter(
    is_dist_dec_lnst_within_max,
    !is.na(grid_site)
  )

d_pts_cruise_filt_smry |>
  dt(
    caption = glue(
      "Cruises with rows filtered: outside CalCOFI grid or exceeded ",
      "{max_dist_dec_lnst_km} km cutoff from line/station coordinates."
    ),
    escape = F,
    fname = "ctd_cruises_distance_from_lnst"
  ) |>
  formatCurrency(
    c(
      "n_all",
      "n_gt_cutoff",
      "avg_dist_gt_cutoff_km",
      "n_outside_grid",
      "n_rm"
    ),
    currency = "",
    digits = 0,
    mark = ","
  ) |>
  formatPercentage(
    c("pct_gt_cutoff", "pct_outside_grid", "pct_rm"),
    digits = 2
  )
```

### View filtered points

```{r}
#| label: view_pts_filt

bb_cr <- st_bbox(pts_filt) |> st_as_sfc()

mapView(bb_cr) +
  mapView(
    pts_filt |>
      group_by(cruise_key) |>
      slice_sample(n = 100) |>
      ungroup() |>
      select(
        cruise_key,
        longitude,
        latitude,
        sta_id,
        line,
        sta,
        dist_dec_lnst_km,
        datetime_start_utc
      ),
    zcol = "cruise_key",
    cex = 5,
    alpha = 0.5
  )
```

### Check for duplicates by datetime and depth

```{r}
#| label: check_dupes

d_dupes <- pts_filt |>
  st_drop_geometry() |>
  select(datetime_start_utc, depth) |>
  group_by(datetime_start_utc, depth) |>
  summarize(n = n(), .groups = "drop") |>
  filter(n > 1) |>
  arrange(desc(n))

n_distinct_dtime_depth <- pts_filt |>
  st_drop_geometry() |>
  select(datetime_start_utc, depth) |>
  n_distinct()

d_dupes |>
  slice(c(1:10, (n() - 9):n())) |>
  dt(
    caption = glue(
      "First and last 10 duplicates by datetime_start_utc, depth ",
      "(nrows = {format(nrow(d_dupes), big.mark = ',')}; ",
      "n_distinct = {format(n_distinct_dtime_depth, big.mark = ',')})"
    ),
    fname = "ctd_duplicates_by_datetime_depth"
  )
```

## Rename Fields

Apply field renames from `flds_redefine.csv`.

```{r}
#| label: rename_fields

# apply column renames from flds_redefine
renames <- d_flds_rd |>
  filter(fld_old != fld_new) |>
  select(fld_old, fld_new)

d_filt <- pts_filt |>
  st_drop_geometry() |>
  select(-geom_lnst)

for (i in seq_len(nrow(renames))) {
  old <- renames$fld_old[i]
  new <- renames$fld_new[i]
  if (old %in% names(d_filt)) {
    d_filt <- d_filt |> rename(!!new := !!old)
  }
}

# drop intermediate columns
d_filt <- d_filt |>
  select(
    -any_of(c(
      "date_time_pst",
      "date_time_utc",
      "date_time_format",
      "is_lon_pseudoNA",
      "is_lat_pseudoNA",
      "is_lnsta_pseudoNA",
      "lon_lnst",
      "lat_lnst",
      "dist_dec_lnst_km",
      "is_dist_dec_lnst_within_max",
      "grid_site"
    ))
  )

# write raw table for pivoting
dbWriteTable(con, "ctd_raw", d_filt, overwrite = TRUE)
say(glue("Loaded {nrow(d_filt)} rows into ctd_raw"))
```

## De-duplicate ctd_raw

Some cruises ship the same casts twice in the download archive — a primary copy plus a reorganized copy in a sub-folder (e.g. `db_csvs/` alongside `db_csvs/south-north/`), both classified `final` by the path regex, so both pass the priority filter (see the duplicates reported above). The copies are identical except for re-rounded coordinates (and occasionally a re-processed measurement value). Collapse to one row per measurement position so the deterministic UUIDs downstream (`ctd_cast_uuid`, `ctd_measurement_uuid`, …) are unique.

```{r}
#| label: dedup_ctd_raw

raw_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns WHERE table_name = 'ctd_raw'"
)$column_name

# a measurement position is identified by these five columns; duplicate copies
# share all five and differ only in re-rounded coordinates / re-processed values
dedup_key  <- c("cruise_key", "cast_key", "cast_dir", "datetime_start_utc", "depth_m")
stopifnot(
  "ctd_raw missing expected columns for de-duplication" =
    all(c(dedup_key, "_source_file", "latitude", "longitude") %in% raw_cols))

# keep the primary copy: reorganized duplicates sit in deeper source paths, so
# prefer the shortest `_source_file`; coordinates tie-break for full determinism
n_before <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_raw")$n
dbExecute(con, glue("
  CREATE OR REPLACE TABLE ctd_raw AS
  SELECT * FROM ctd_raw
  QUALIFY ROW_NUMBER() OVER (
    PARTITION BY {paste(dedup_key, collapse = ', ')}
    ORDER BY length(\"_source_file\"), \"_source_file\", latitude, longitude) = 1"))
n_after <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_raw")$n

cat(glue(
  "ctd_raw de-duplicated: removed {format(n_before - n_after, big.mark = ',')} ",
  "duplicate-position rows ",
  "({round(100 * (n_before - n_after) / max(n_before, 1), 1)}%); ",
  "{format(n_after, big.mark = ',')} rows remain"), "\n")
```

```{r}
#| label: restore_eval
#| eval: true

# must always evaluate: line ~217 may have set eval=FALSE to skip read+bind
# when restoring from checkpoint; this chunk re-enables computation through
# write_parquet. without `eval: true` it would inherit eval=FALSE and be
# skipped, so write_parquet never runs yet downstream chunks expect its output.
if (parquet_complete) {
  knitr::opts_chunk$set(eval = FALSE)
  say("Parquet complete — skipping to upload")
} else {
  knitr::opts_chunk$set(eval = TRUE)
}
```

## Cross-Dataset Bridge

Load reference tables (ship, cruise, grid) from the ichthyo ingest, derive `ship_key`, and validate against known ships/cruises.

```{r}
#| label: cross_dataset_bridge

# tables declared in calcofi.modifies — loaded as TABLEs (writable)
modifies_tables <- c("ship")  # from calcofi.modifies in YAML frontmatter
load_prior_tables(
  con         = con,
  tables      = modifies_tables,
  parquet_dir = here("data/parquet/swfsc_ichthyo")
)
# cruise + grid as VIEWs (read-only: cruise used for JOIN, grid for assign_grid_key)
load_prior_tables(
  con         = con,
  tables      = c("cruise", "grid"),
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  as_view     = TRUE
)

# snapshot PKs of modified tables before modifications
modifies_pks <- list()
for (tbl in modifies_tables) {
  pk_col <- dbGetQuery(con, glue(
    "SELECT column_name FROM information_schema.columns
     WHERE table_name = '{tbl}' ORDER BY ordinal_position LIMIT 1"))$column_name
  modifies_pks[[tbl]] <- list(
    pk_col = pk_col,
    keys   = dbGetQuery(con, glue("SELECT {pk_col} FROM {tbl}"))[[1]])
}

# apply cruise_key corrections for known study-column data quality errors.
# runs unconditionally so it fires for both fresh runs (YYMMKK keys) and
# checkpoint-restored runs (already-converted YYYY-MM-NODC keys).
for (i in seq_len(nrow(d_cruise_corrections))) {
  n_wrong <- dbGetQuery(con, glue(
    "SELECT COUNT(*) AS n FROM ctd_raw
     WHERE cruise_key = '{d_cruise_corrections$cruise_key_raw[i]}'"))$n
  if (n_wrong > 0) {
    dbExecute(con, glue(
      "UPDATE ctd_raw SET cruise_key = '{d_cruise_corrections$cruise_key_correct[i]}'
       WHERE cruise_key = '{d_cruise_corrections$cruise_key_raw[i]}'"))
    say(glue(
      "cruise_key correction: {d_cruise_corrections$cruise_key_raw[i]} → ",
      "{d_cruise_corrections$cruise_key_correct[i]} ({format(n_wrong, big.mark=',')} rows)"))
  }
}

# detect whether cruise_key already in YYYY-MM-NODC format (from checkpoint)
sample_ck <- dbGetQuery(
  con, "SELECT cruise_key FROM ctd_raw LIMIT 1")$cruise_key
cruise_key_already_converted <- grepl("^\\d{4}-\\d{2}-", sample_ck)

if (!cruise_key_already_converted) {
  # 0404NHJD is a combined New Horizon + Jordan Davis cruise; use primary ship NH
  dbExecute(
    con,
    "UPDATE ctd_raw SET cruise_key = LEFT(cruise_key, 6)
     WHERE LENGTH(cruise_key) > 6"
  )

  # derive ship_key from last 2 chars of cruise_key (YYMMKK format from filename)
  dbExecute(con, "ALTER TABLE ctd_raw ADD COLUMN IF NOT EXISTS ship_key VARCHAR")
  dbExecute(con, "UPDATE ctd_raw SET ship_key = RIGHT(cruise_key, 2)")

  # validate ship_key against ship reference table
  ctd_ships <- dbGetQuery(con, "SELECT DISTINCT ship_key FROM ctd_raw")
  ref_ships <- dbGetQuery(con, "SELECT ship_key FROM ship")
  orphan_ships <- setdiff(ctd_ships$ship_key, ref_ships$ship_key)

  if (length(orphan_ships) > 0) {
    say(glue(
      "{length(orphan_ships)} ship_key(s) in CTD not in ship table: ",
      "{paste(orphan_ships, collapse = ', ')}"
    ))

    # run match_ships() with centralized renames for orphan ships
    orphan_tbl <- dbGetQuery(con, glue(
      "SELECT DISTINCT ship_key AS ship_code, NULL AS ship_name
       FROM ctd_raw
       WHERE ship_key IN ({paste(shQuote(orphan_ships, type = 'sh'), collapse = ', ')})")) |>
      mutate(ship_name = as.character(ship_name))

    ship_result <- match_ships(
      unmatched_ships  = orphan_tbl,
      reference_ships  = dbReadTable(con, "ship"),
      ship_renames_csv = here("metadata/ship_renames.csv"),
      fetch_ices       = FALSE)

    # insert interim entries for still-unmatched ships (ship_nodc = "?XX?")
    ensure_interim_ships(con, ship_result)

    dbGetQuery(
      con,
      glue(
        "SELECT ship_key, COUNT(DISTINCT cruise_key) AS n_cruises,
                COUNT(*) AS n_rows
         FROM ctd_raw
         WHERE ship_key IN ({paste(shQuote(orphan_ships, type = 'sh'), collapse = ', ')})
         GROUP BY ship_key
         ORDER BY ship_key"
      )
    ) |>
      dt(
        caption = "Orphan ship_keys (in CTD but not in ship table)",
        fname = "ctd_orphan_ship_keys"
      )
  } else {
    say("All CTD ship_keys found in ship reference table")
  }

  # convert cruise_key from YYMMKK → YYYY-MM-NODC format
  convert_cruise_key_format(con, "ctd_raw", old_key_col = "cruise_key")
  dbExecute(con, "ALTER TABLE ctd_raw DROP COLUMN cruise_key")
  dbExecute(con, "ALTER TABLE ctd_raw RENAME COLUMN cruise_key_new TO cruise_key")
} else {
  say("cruise_key already in YYYY-MM-NODC format (from checkpoint)")
}

# validate cruise_key against cruise reference table
ctd_cruises <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM ctd_raw")
ref_cruises <- dbGetQuery(con, "SELECT cruise_key FROM cruise")
orphan_cruises <- setdiff(ctd_cruises$cruise_key, ref_cruises$cruise_key)

if (length(orphan_cruises) > 0) {
  say(glue(
    "{length(orphan_cruises)} cruise_key(s) in CTD not in cruise table: ",
    "{paste(head(orphan_cruises, 10), collapse = ', ')}",
    "{if (length(orphan_cruises) > 10) '...' else ''}"
  ))
  dbGetQuery(
    con,
    glue(
      "SELECT cruise_key, COUNT(*) AS n_rows
       FROM ctd_raw
       WHERE cruise_key IN ({paste(shQuote(orphan_cruises, type = 'sh'), collapse = ', ')})
       GROUP BY cruise_key
       ORDER BY cruise_key"
    )
  ) |>
    dt(
      caption = "Orphan cruise_keys (in CTD but not in cruise table)",
      fname = "ctd_orphan_cruise_keys"
    )
} else {
  say("All CTD cruise_keys found in cruise reference table")
}
```

## Save Checkpoint

Save a DuckDB checkpoint after the expensive read+bind+filter+bridge steps so subsequent runs can resume from here.

```{r}
#| label: save_checkpoint

if (!file_exists(db_checkpoint) || overwrite) {
  close_duckdb(con)
  file_copy(db_path, db_checkpoint, overwrite = TRUE)
  con <- get_duckdb_con(db_path)
  load_duckdb_extension(con, "spatial")
  load_duckdb_extension(con, "icu")
  say(glue("Saved checkpoint: {db_checkpoint}"))
} else {
  say(glue("Checkpoint already exists: {db_checkpoint}"))
}
```

## Measurement Column Registry

Resolve which `ctd_raw` columns are **measurements** and which are their
**quality flags**, from the shared `metadata/measurement_type.csv` registry. This
split drives everything downstream: measurement columns pivot into
`ctd_measurement`, and whatever is left over is by definition cast-level and
becomes `ctd_cast`. Getting it wrong is how a per-scan column once leaked into
`ctd_cast` and broke its primary key (workflows#53).

::: {.callout-note title="`ctd_wide` was retired here"}
This section used to also build `ctd_wide`, a ~1 GB wide-format copy of
`ctd_raw` for ERDDAP. It has no consumer: ERDDAP now serves CTD through
`EDDTableFromDatabase` over DuckDB views plus the netCDF files built below, and
the wide file's whole-file heap read is what OOM'd ERDDAP in the first place
(see [the serving benchmark](https://calcofi.io/workflows/bench_erddap_ctd.html)).
The column-registry logic it carried is load-bearing, so it stays; the table
itself is gone.
:::

```{r}
#| label: meas_col_registry

# identify measurement and quality columns from measurement_type metadata
d_meas_ctd <- d_meas_type |>
  filter(str_detect(`_source_datasets`, "calcofi_ctd-cast"))
meas_cols <- d_meas_ctd$`_source_column`
qual_cols <- d_meas_ctd$`_qual_column` |> na.omit() |> as.character()
qual_cols <- qual_cols[qual_cols != ""]

raw_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns WHERE table_name = 'ctd_raw'"
)$column_name

# DIAGNOSTIC: a measurement column in the registry that is absent from ctd_raw is
# a silent no-op in the pivot; a numeric ctd_raw column that is NOT registered
# falls through to cast_cols and can break the ctd_cast PK. Report both.
meas_missing <- setdiff(meas_cols, raw_cols)
cast_leftover <- setdiff(raw_cols, c(meas_cols, qual_cols, "depth_m", "_source_file"))

cat(glue("registered measurement types : {nrow(d_meas_ctd)}\n"))
cat(glue("  canonical (-> ctd_thin)    : {sum(d_meas_ctd$is_canonical, na.rm = TRUE)}\n"))
cat(glue("  quality-flag columns       : {length(qual_cols)}\n"))
cat(glue("registry columns absent from ctd_raw: {length(meas_missing)}",
         if (length(meas_missing)) glue(" -> {paste(meas_missing, collapse = ', ')}") else "", "\n"))
cat(glue("cast-level columns (-> ctd_cast)    : {length(cast_leftover)}\n"))
```

```{r}
#| label: meas_col_registry_tbl
# the canonical set is what ctd_thin and the thinned netCDF carry, so show it
# explicitly rather than leaving it implicit in a flag column
d_meas_ctd |>
  select(measurement_type, units, is_canonical, source_column = `_source_column`) |>
  arrange(desc(is_canonical), measurement_type) |>
  datatable(
    rownames  = FALSE,
    caption   = "CTD measurement registry — is_canonical drives ctd_thin membership",
    options   = list(pageLength = 10, scrollX = TRUE, dom = "ftip"))
```

## Split into Tidy Tables

### ctd_cast — cast-level metadata

```{r}
#| label: ctd_cast

# d_meas_ctd, meas_cols, qual_cols come from the Measurement Column Registry above

# use ctd_raw columns (includes ship_key and data_stage added via SQL)
raw_cols <- dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns WHERE table_name = 'ctd_raw'"
)$column_name

cast_cols <- setdiff(
  raw_cols,
  c(meas_cols, qual_cols, "depth_m", "_source_file")
)

# extract one row per cast, keyed (cruise_key, cast_key, cast_dir,
# datetime_start_utc). within that key the only cast-level column that varies is
# lat/lon (sub-meter GPS jitter between same-second scans), so order by it for a
# deterministic pick. measurement-like columns (e.g. ox_aveu_m_sta_corr) must be
# registered in measurement_type.csv so they pivot to ctd_measurement and never
# reach cast_cols — otherwise SELECT DISTINCT would emit multiple rows per cast.
dbExecute(
  con,
  glue(
    "
  CREATE OR REPLACE TABLE ctd_cast AS
  SELECT {paste(cast_cols, collapse = ', ')}
  FROM ctd_raw
  QUALIFY ROW_NUMBER() OVER (
    PARTITION BY cruise_key, cast_key, cast_dir, datetime_start_utc
    ORDER BY latitude, longitude) = 1
  ORDER BY datetime_start_utc, cruise_key"
  )
)

n_cast <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_cast")$n
say(glue("ctd_cast: {format(n_cast, big.mark = ',')} rows"))

# assign deterministic UUID from composite natural key (md5-based, DuckDB-native)
assign_deterministic_uuids_md5(
  con = con,
  table_name = "ctd_cast",
  id_col = "ctd_cast_uuid",
  key_cols = c("cruise_key", "cast_key", "cast_dir", "datetime_start_utc")
)

# guard: ctd_cast_uuid must be a valid unique PK — fail loudly if a
# measurement-like column ever leaks into cast_cols again (see #53)
stopifnot(
  "ctd_cast_uuid is not unique" =
    dbGetQuery(
      con,
      "SELECT COUNT(*) - COUNT(DISTINCT ctd_cast_uuid) AS n FROM ctd_cast"
    )$n == 0)
```

### ctd_measurement — long-format sensor readings

```{r}
#| label: ctd_measurement

# add ctd_cast_uuid to ctd_raw for FK linkage
dbExecute(
  con,
  "
  ALTER TABLE ctd_raw ADD COLUMN IF NOT EXISTS ctd_cast_uuid VARCHAR"
)
dbExecute(
  con,
  "
  UPDATE ctd_raw AS r
  SET ctd_cast_uuid = c.ctd_cast_uuid
  FROM ctd_cast AS c
  WHERE r.cruise_key   = c.cruise_key
    AND r.cast_key     = c.cast_key
    AND r.cast_dir     = c.cast_dir
    AND r.datetime_start_utc = c.datetime_start_utc"
)

# pivot wide-to-long one measurement type at a time to avoid OOM
dbExecute(con, "DROP TABLE IF EXISTS ctd_measurement")

for (i in seq_len(nrow(d_meas_ctd))) {
  row <- d_meas_ctd[i, ]
  src_col <- row$`_source_column`
  meas_type <- row$measurement_type
  qual_col <- row$`_qual_column`

  # a bare NULL is typed INT32 by DuckDB, so if the FIRST measurement type has no
  # qual column the CREATE TABLE below makes measurement_qual an integer column —
  # and the first later type that does have one fails inserting 'nan'. Type it.
  #
  # The source stores these codes in DOUBLE columns, so a plain cast yields "9.0"
  # rather than "9" and no longer matches the controlled vocabulary in
  # metadata/measurement_qual.csv (the CTD set: 0/1/2/8/9). Strip a trailing ".0" TEXTUALLY
  # rather than casting through INTEGER: a numeric cast would silently round an
  # unexpected "9.5" to 9, whereas this leaves anything non-integral verbatim so it
  # shows up in the audit below instead of being quietly normalized away.
  qual_expr <- if (!is.na(qual_col) && qual_col != "") {
    glue("regexp_replace(CAST({qual_col} AS VARCHAR), '\\.0+$', '')")
  } else {
    "CAST(NULL AS VARCHAR)"
  }

  sql_select <- glue(
    "
    SELECT ctd_cast_uuid, depth_m,
      '{meas_type}' AS measurement_type,
      CAST({src_col} AS DOUBLE) AS measurement_value,
      {qual_expr} AS measurement_qual
    FROM ctd_raw
    WHERE {src_col} IS NOT NULL
      AND NOT isnan(CAST({src_col} AS DOUBLE))
      AND isfinite(CAST({src_col} AS DOUBLE))
    "
  )

  if (i == 1) {
    dbExecute(con, glue("CREATE OR REPLACE TABLE ctd_measurement AS {sql_select}"))
  } else {
    dbExecute(con, glue("INSERT INTO ctd_measurement {sql_select}"))
  }
  say(glue("  {i}/{nrow(d_meas_ctd)}: {meas_type}"))
}

n_meas_raw <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_measurement")$n
say(glue("ctd_measurement (pre-sentinel): {format(n_meas_raw, big.mark = ',')} rows"))

# --- strip the -99 missing-value sentinel from MEASUREMENTS -------------------
# -99 is this source's documented missing marker. The notebook already knows that:
# `pseudoNA_values <- c(-9.99e-29, -99)` in the fill_lonlat chunk strips it from
# longitude/latitude. It was never applied to the measurement columns, so -99 has
# been flowing through ctd_measurement -> ctd_thin -> obs as though it were a real
# reading — 84,302 rows in release v2026.07.17, including CANONICAL oxygen
# (oxygen_ml_l_ave_sta_corr 953, oxygen_umol_kg_ave_sta_corr 4,294) alongside
# isus_v 40,479, ph 31,493 and spar 6,189.
#
# The NOT isnan / isfinite guard in the pivot above cannot catch it: -99 is a
# perfectly finite double. A -99 mL/L oxygen silently corrupts any mean, minimum
# or anomaly a consumer computes, which is why this is a DELETE and not merely a
# report — a missing value has no business existing as a row in long format.
#
# Matched exactly (= -99, not a tolerance): every observed case is exactly -99.00,
# and an exact test cannot swallow a real reading that merely rounds near it. The
# two genuinely signed types (dynamic_height, specific_volume_anomaly) are
# reported separately below so a real -99 there would be visible rather than
# assumed away.
SENTINEL <- -99

sentinel_tally <- dbGetQuery(con, glue("
  SELECT measurement_type, COUNT(*) AS n,
         COUNT(DISTINCT ctd_cast_uuid) AS n_casts
  FROM ctd_measurement WHERE measurement_value = {SENTINEL}
  GROUP BY 1 ORDER BY n DESC"))

n_sentinel <- sum(sentinel_tally$n)
dbExecute(con, glue(
  "DELETE FROM ctd_measurement WHERE measurement_value = {SENTINEL}"))

n_meas <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_measurement")$n
say(glue("-99 sentinel rows removed: {format(n_sentinel, big.mark = ',')} ",
         "across {nrow(sentinel_tally)} measurement type(s)"))
say(glue("ctd_measurement: {format(n_meas, big.mark = ',')} rows"))
stopifnot("sentinel delete accounting" = n_meas_raw - n_sentinel == n_meas)

# assign deterministic UUID (md5-based, DuckDB-native — single SQL, no R data transfer)
assign_deterministic_uuids_md5(
  con = con,
  table_name = "ctd_measurement",
  id_col = "ctd_measurement_uuid",
  key_cols = c("ctd_cast_uuid", "depth_m", "measurement_type")
)

# add cruise_key for partitioned parquet output
dbExecute(
  con,
  "ALTER TABLE ctd_measurement ADD COLUMN IF NOT EXISTS cruise_key VARCHAR"
)
dbExecute(
  con,
  "UPDATE ctd_measurement AS m
   SET cruise_key = c.cruise_key
   FROM ctd_cast AS c
   WHERE m.ctd_cast_uuid = c.ctd_cast_uuid"
)

# remove orphaned measurements with no parent cast
n_orphan <- dbGetQuery(
  con,
  "SELECT COUNT(*) FROM ctd_measurement WHERE ctd_cast_uuid IS NULL"
)[[1]]
if (n_orphan > 0) {
  dbExecute(con, "DELETE FROM ctd_measurement WHERE ctd_cast_uuid IS NULL")
  cat(glue("{format(n_orphan, big.mark = ',')} orphaned ctd_measurement ",
           "rows removed (NULL ctd_cast_uuid)"), "\n")
}
```

### ctd_summary — summary stats across cast directions

```{r}
#| label: ctd_summary

dbExecute(
  con,
  "
  CREATE OR REPLACE TABLE ctd_summary AS
  SELECT
    c.cruise_key,
    c.site_key,
    m.depth_m,
    m.measurement_type,
    AVG(m.measurement_value)    AS avg,
    CASE
      WHEN COUNT(*) = 1 THEN 0
      ELSE COALESCE(STDDEV_SAMP(m.measurement_value), 0)
    END                          AS stddev,
    COUNT(*)                     AS n_obs
  FROM ctd_measurement m
  INNER JOIN ctd_cast c ON m.ctd_cast_uuid = c.ctd_cast_uuid
  WHERE NOT isnan(m.measurement_value)
    AND isfinite(m.measurement_value)
  GROUP BY c.cruise_key, c.site_key, m.depth_m, m.measurement_type"
)

n_summ <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_summary")$n
say(glue("ctd_summary: {format(n_summ, big.mark = ',')} rows"))

assign_deterministic_uuids_md5(
  con = con,
  table_name = "ctd_summary",
  id_col = "ctd_summary_uuid",
  key_cols = c("cruise_key", "site_key", "depth_m", "measurement_type")
)
```

### measurement_type — reference table

```{r}
#| label: measurement_type

dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
say(glue("measurement_type: {nrow(d_meas_type)} rows"))

# validate: all measurement_types used by this dataset are registered
ctd_types_used <- dbGetQuery(
  con,
  "SELECT DISTINCT measurement_type FROM ctd_measurement"
)$measurement_type

registered <- d_meas_type |>
  filter(str_detect(`_source_datasets`, "calcofi_ctd-cast")) |>
  pull(measurement_type)

unregistered <- setdiff(ctd_types_used, registered)
stopifnot(
  "all measurement_types used by calcofi_ctd-cast must be registered in _source_datasets" = length(
    unregistered
  ) ==
    0
)
```

### Measurement provenance — instrument and accuracy eras

Which instrument produced a value, to what stated accuracy, and over which years.
Recovered from the CalCOFI hydrographic master's `0-Measurements` table into
`metadata/calcofi/hydro-master/measurement_method.csv`; the pipeline had no
equivalent. It matters most where a method changed mid-record — water temperature
switches from reversing thermometer (through 1993-04-15) to CTD thermistor (from
1993-08-11), which is the boundary this dataset begins at, and phosphate crosses
four instruments from 1949.

The registry is **one-to-many** by design (Temperature has 6 eras), so it is joined
by *property* rather than by exact type: CTD types carry sensor and correction
suffixes (`temperature_1`, `salinity_ave_corr`, `oxygen_ml_l_ave_sta_corr`) that
the era table knows nothing about, so those are stripped back to the base property
before matching.

```{r}
#| label: meas_provenance

path_method <- here("metadata/calcofi/hydro-master/measurement_method.csv")

d_method <- read_csv(path_method, show_col_types = F) |>
  filter(!is.na(measurement_type)) |>
  select(property = measurement_type, method, accuracy, year_started, year_ended)

# strip sensor/correction suffixes and the bottle marker to reach the base property:
#   temperature_ave  -> temperature ; oxygen_ml_l_ave_sta_corr -> oxygen_ml_l
#   btl_temperature  -> temperature ; salinity_btl            -> salinity
#   oxygen_btl_ml_l  -> oxygen_ml_l  (the marker is INFIX here, not a prefix/suffix —
#                                     the three positions are why this is not one regex)
base_property <- function(x) {
  x |>
    str_remove("^btl_") |>
    str_remove("_btl$") |>
    str_replace("_btl_", "_") |>
    str_remove("_(cruise|sta)_corr$") |>
    str_remove("_corr$") |>
    str_remove("_ave$") |>
    str_remove("_[12]$")
}

d_prov <- d_meas_ctd |>
  transmute(measurement_type, units, is_canonical,
            property = base_property(measurement_type)) |>
  inner_join(d_method, by = "property", relationship = "many-to-many") |>
  arrange(desc(is_canonical), measurement_type, year_started)

n_typ <- n_distinct(d_prov$measurement_type)
say(glue("CTD types with instrument provenance: {n_typ} of {nrow(d_meas_ctd)} ",
         "({n_distinct(d_prov$property)} distinct properties, ",
         "{nrow(d_prov)} type x era rows)"))
say(glue("  unmatched (no era recorded): ",
         "{paste(setdiff(d_meas_ctd$measurement_type, d_prov$measurement_type) |> head(8), collapse = ', ')}",
         if (nrow(d_meas_ctd) - n_typ > 8) " ..." else ""))
```

```{r}
#| label: meas_provenance_tbl
d_prov |>
  transmute(measurement_type, canonical = is_canonical, units,
            method, accuracy, from = year_started, to = year_ended) |>
  datatable(
    caption  = paste("CTD measurement provenance — instrument, stated accuracy and",
                     "era, from the CalCOFI hydrographic master's 0-Measurements"),
    rownames = FALSE,
    options  = list(pageLength = 10, scrollX = TRUE, dom = "ftip"))
```

::: {.callout-note title="Documentation only, for now"}
This table is rendered, not written into `metadata.json`. Promoting it into the
sidecar so consumers get instrument provenance alongside units is a follow-up
that belongs in `calcofi4db::build_metadata_json()`, and 18 of the 35 era rows
still need a data manager to confirm their mapping to canonical types
(`hydro_master_06`).
:::

### Drop intermediate table

```{r}
#| label: drop_ctd_raw

dbExecute(con, "DROP TABLE IF EXISTS ctd_raw")
```

## Measurement Frequency

`ctd_measurement` is by far the largest table in this dataset — in the prior release the `calcofi_ctd-cast` parquet totaled 22.16 GB, with `ctd_measurement` alone at 15.8 GB (233 M rows). Three stacked multipliers drive that size. The queries below quantify each, motivating the adaptively-thinned `ctd_thin` table built in the next section.

```{r}
#| label: freq_table_sizes

freq_tbl_sizes <- dbGetQuery(con, "
  SELECT 'ctd_cast'         AS table_name, COUNT(*) AS n_rows FROM ctd_cast
  UNION ALL
  SELECT 'ctd_measurement' AS table_name, COUNT(*) AS n_rows FROM ctd_measurement
  UNION ALL
  SELECT 'ctd_summary'     AS table_name, COUNT(*) AS n_rows FROM ctd_summary")

n_meas_rows <- freq_tbl_sizes$n_rows[freq_tbl_sizes$table_name == "ctd_measurement"]
cat(glue("ctd_measurement: {format(n_meas_rows, big.mark = ',')} rows"), "\n")

freq_tbl_sizes |>
  dt(caption = "Normalized CTD table row counts (this run)",
     fname   = "ctd_freq_table_sizes")
```

### Multiplier 1 — sub-meter depth resolution

CTD profiles record a sample roughly every metre from surface to seafloor (~3,600 m max). Thinning to a 10 m grid is therefore an order-of-magnitude reduction on the depth axis.

```{r}
#| label: freq_depth_resolution

# depth interval between consecutive samples within a cast, using `pressure`
# as a representative measurement_type (all types share a cast's depth grid)
freq_depth_int <- dbGetQuery(con, "
  WITH gaps AS (
    SELECT depth_m - LAG(depth_m) OVER (
             PARTITION BY ctd_cast_uuid ORDER BY depth_m) AS d_gap
    FROM ctd_measurement
    WHERE measurement_type = 'pressure')
  SELECT
    ROUND(MIN(d_gap), 2)                 AS min_m,
    ROUND(QUANTILE_CONT(d_gap, 0.25), 2) AS q25_m,
    ROUND(MEDIAN(d_gap), 2)              AS median_m,
    ROUND(QUANTILE_CONT(d_gap, 0.75), 2) AS q75_m,
    ROUND(MAX(d_gap), 2)                 AS max_m
  FROM gaps
  WHERE d_gap IS NOT NULL")

cat(glue(
  "median within-cast depth interval: {freq_depth_int$median_m} m ",
  "(~{round(10 / freq_depth_int$median_m)}x reduction at a 10 m grid)"), "\n")

freq_depth_int |>
  dt(caption = "Within-cast depth interval (m), representative type `pressure`",
     fname   = "ctd_freq_depth_resolution")
```

### Multiplier 2 — up- and down-casts both stored

Nearly every physical cast is recorded as a separate down-cast and up-cast. Keeping a single direction roughly halves the rows.

```{r}
#| label: freq_cast_direction

# physical cast = (cruise_key, site_key, cast_base); direction resolved from
# cast_dir, falling back to the cast_key suffix (trailing d/u)
freq_cast_dir <- dbGetQuery(con, "
  WITH phys AS (
    SELECT DISTINCT
      cruise_key, site_key,
      regexp_replace(cast_key, '[duDU]$', '') AS cast_base,
      COALESCE(
        cast_dir,
        CASE
          WHEN RIGHT(cast_key, 1) IN ('d', 'D') THEN 'D'
          WHEN RIGHT(cast_key, 1) IN ('u', 'U') THEN 'U'
        END) AS dir
    FROM ctd_cast)
  SELECT
    COUNT(DISTINCT cruise_key || '|' || site_key || '|' || cast_base)
                                       AS n_physical_casts,
    COUNT(*) FILTER (WHERE dir = 'D')   AS n_with_downcast,
    COUNT(*) FILTER (WHERE dir = 'U')   AS n_with_upcast,
    COUNT(*) FILTER (WHERE dir IS NULL) AS n_unknown_dir
  FROM phys")

cat(glue(
  "{format(freq_cast_dir$n_physical_casts, big.mark = ',')} physical casts; ",
  "{format(freq_cast_dir$n_with_downcast, big.mark = ',')} have a downcast, ",
  "{format(freq_cast_dir$n_with_upcast, big.mark = ',')} an upcast"), "\n")

freq_cast_dir |>
  dt(caption = "Cast direction coverage per physical cast",
     fname   = "ctd_freq_cast_direction")
```

### Multiplier 3 — redundant measurement-type variants

Many of the measurement types below are variants of the same property (multiple oxygen units and sensor/correction combinations, dual temperature and salinity sensors, derived estimates). `ctd_thin` keeps one canonical type per property — see the `is_canonical` flag in `metadata/measurement_type.csv`.

```{r}
#| label: freq_measurement_types

freq_meas_types <- dbGetQuery(con, "
  SELECT
    measurement_type,
    COUNT(*)                      AS n_rows,
    COUNT(DISTINCT ctd_cast_uuid) AS n_casts
  FROM ctd_measurement
  GROUP BY measurement_type
  ORDER BY n_rows DESC")

cat(glue(
  "{nrow(freq_meas_types)} distinct measurement_type values in ctd_measurement"),
  "\n")

freq_meas_types |>
  dt(caption = "Rows per measurement_type",
     fname   = "ctd_freq_measurement_types")
```

## CTD Thin

`ctd_thin` is an adaptively-thinned `ctd_measurement`: one cast direction per physical cast, canonical measurement types only, and a ~10 m depth grid with thermocline / halocline inflections preserved via Ramer–Douglas–Peucker line simplification. It is the headline CTD table; full `ctd_measurement` is retained as a supplemental output. See the Measurement Frequency section above for the size rationale.

### Verify cast pairing

```{r}
#| label: ctd_thin_verify_pairing

# physical cast = (cruise_key, site_key, cast_base); direction comes from the
# cast_key suffix (trailing d/u), which is part of the deterministic
# ctd_cast_uuid key and fully consistent per cast. confirm casts are paired
# before thinning by direction.
ctd_pairing <- dbGetQuery(con, "
  WITH phys AS (
    SELECT DISTINCT
      cruise_key, site_key,
      regexp_replace(cast_key, '[duDU]$', '') AS cast_base,
      CASE WHEN RIGHT(cast_key, 1) IN ('d', 'D') THEN 'D'
           WHEN RIGHT(cast_key, 1) IN ('u', 'U') THEN 'U' END AS dir
    FROM ctd_cast)
  SELECT n_dirs, COUNT(*) AS n_physical_casts
  FROM (
    SELECT cruise_key, site_key, cast_base, COUNT(DISTINCT dir) AS n_dirs
    FROM phys GROUP BY 1, 2, 3)
  GROUP BY n_dirs ORDER BY n_dirs")

n_paired <- sum(ctd_pairing$n_physical_casts[ctd_pairing$n_dirs == 2])
n_single <- sum(ctd_pairing$n_physical_casts[ctd_pairing$n_dirs == 1])
cat(glue(
  "{format(n_paired, big.mark = ',')} physical casts have both directions, ",
  "{format(n_single, big.mark = ',')} have one"), "\n")

# sanity: most casts should be paired — a near-zero n_paired means the pairing
# key is wrong and the single-direction lever would silently keep everything
stopifnot(
  "ctd_thin: cast pairing key looks wrong (no two-direction casts)" =
    n_paired > n_single)

ctd_pairing |>
  dt(caption = "Cast directions per physical cast", fname = "ctd_thin_pairing")
```

### Choose a single cast direction

```{r}
#| label: ctd_thin_chosen_dir

# canonical measurement types for ctd_thin (one per property; see is_canonical
# in metadata/measurement_type.csv)
canon_types <- d_meas_type |>
  filter(str_detect(`_source_datasets`, "calcofi_ctd-cast"), is_canonical) |>
  pull(measurement_type)
canon_in <- paste0("'", canon_types, "'", collapse = ", ")
cat(glue("{length(canon_types)} canonical measurement types: ",
         "{paste(canon_types, collapse = ', ')}"), "\n")

# choose one direction per physical cast: prefer downcast, fall back to upcast.
# _chosen_dir maps every chosen-direction ctd_cast_uuid to its physical cast
# (phys_cast_id) and direction.
dbExecute(con, "
  CREATE OR REPLACE TEMP TABLE _chosen_dir AS
  WITH resolved AS (
    SELECT DISTINCT
      ctd_cast_uuid, cruise_key, site_key, cast_key,
      regexp_replace(cast_key, '[duDU]$', '') AS cast_base,
      CASE WHEN RIGHT(cast_key, 1) IN ('d', 'D') THEN 'D'
           WHEN RIGHT(cast_key, 1) IN ('u', 'U') THEN 'U' END AS dir
    FROM ctd_cast),
  pick AS (
    SELECT cruise_key, site_key, cast_base, cast_key, dir
    FROM (SELECT DISTINCT cruise_key, site_key, cast_base, cast_key, dir
          FROM resolved)
    QUALIFY ROW_NUMBER() OVER (
      PARTITION BY cruise_key, site_key, cast_base
      ORDER BY CASE dir WHEN 'D' THEN 1 WHEN 'U' THEN 2 ELSE 3 END,
               cast_key) = 1)
  SELECT
    r.ctd_cast_uuid,
    r.cruise_key,
    r.cruise_key || '|' || r.site_key || '|' || r.cast_base AS phys_cast_id,
    p.dir AS cast_dir
  FROM resolved r
  JOIN pick p USING (cruise_key, site_key, cast_base, cast_key)")

n_chosen <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM _chosen_dir")$n
cat(glue("_chosen_dir: {format(n_chosen, big.mark = ',')} cast records ",
         "(single direction per physical cast)"), "\n")
```

### 10 m depth grid

```{r}
#| label: ctd_thin_grid

# depth backbone: every distinct depth a canonical measurement is recorded at,
# per chosen-direction physical cast
dbExecute(con, glue("
  CREATE OR REPLACE TEMP TABLE _profile_depths AS
  SELECT DISTINCT cd.phys_cast_id, m.depth_m
  FROM ctd_measurement m
  JOIN _chosen_dir cd ON m.ctd_cast_uuid = cd.ctd_cast_uuid
  WHERE m.measurement_type IN ({canon_in})"))

# 10 m grid: the sample nearest each 10 m node per profile, plus the shallowest
# and deepest sample (RDP needs fixed profile endpoints)
dbExecute(con, "
  CREATE OR REPLACE TEMP TABLE _grid_depths AS
  WITH ranked AS (
    SELECT phys_cast_id, depth_m,
      ROW_NUMBER() OVER (
        PARTITION BY phys_cast_id, ROUND(depth_m / 10.0) * 10.0
        ORDER BY ABS(depth_m - ROUND(depth_m / 10.0) * 10.0), depth_m) AS rn
    FROM _profile_depths),
  endpoints AS (
    SELECT phys_cast_id, MIN(depth_m) AS depth_m FROM _profile_depths GROUP BY 1
    UNION
    SELECT phys_cast_id, MAX(depth_m) AS depth_m FROM _profile_depths GROUP BY 1)
  SELECT phys_cast_id, depth_m FROM ranked WHERE rn = 1
  UNION
  SELECT phys_cast_id, depth_m FROM endpoints")

n_prof <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM _profile_depths")$n
n_grid <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM _grid_depths")$n
cat(glue("_grid_depths: {format(n_grid, big.mark = ',')} depths on the 10 m grid ",
         "(from {format(n_prof, big.mark = ',')} profile depths)"), "\n")
```

### Inflection points (Ramer–Douglas–Peucker)

```{r}
#| label: ctd_thin_rdp

# Ramer-Douglas-Peucker line simplification: keep depths where a profile deviates
# from the local straight line by more than `eps` (captures thermocline /
# halocline inflections). pure DP — the ~10 m backbone is provided separately by
# _grid_depths, so RDP only adds genuine inflections on top of it.
rdp_keep <- function(x, y, eps) {
  n    <- length(x)
  keep <- rep(FALSE, n)
  if (n == 0L) return(keep)
  keep[c(1L, n)] <- TRUE
  if (n <= 2L) return(keep)
  stack <- list(c(1L, n))
  while (length(stack) > 0L) {
    seg   <- stack[[length(stack)]]
    stack[[length(stack)]] <- NULL
    i <- seg[1L]; j <- seg[2L]
    if (j - i < 2L) next
    idx <- (i + 1L):(j - 1L)
    dx  <- x[j] - x[i]
    dy  <- y[j] - y[i]
    den <- sqrt(dx * dx + dy * dy)
    d <- if (den == 0) {
      abs(y[idx] - y[i])
    } else {
      abs(dy * x[idx] - dx * y[idx] + x[j] * y[i] - y[j] * x[i]) / den
    }
    if (max(d) > eps) {
      k <- idx[which.max(d)]
      keep[k] <- TRUE
      stack[[length(stack) + 1L]] <- c(i, k)
      stack[[length(stack) + 1L]] <- c(k, j)
    }
  }
  keep
}

# per-variable tolerance in measurement units — the primary tuning knob, set so
# the median retained-sample gap lands near the 10 m grid spacing (a looser eps
# keeps fewer inflections; revisit with the spot-check in the verification section)
rdp_eps <- c(temperature_ave = 0.2, salinity_ave_corr = 0.04)

# run RDP per cruise to keep memory flat: pull the two key variables for
# chosen-direction casts, simplify each profile, collect retained depths
cruise_keys <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM _chosen_dir")$cruise_key

rdp_retained <- purrr::map(cruise_keys, function(ck) {
  d_prof <- dbGetQuery(con, glue("
    SELECT cd.phys_cast_id, m.measurement_type, m.depth_m,
           AVG(m.measurement_value) AS measurement_value
    FROM ctd_measurement m
    JOIN _chosen_dir cd
      ON m.ctd_cast_uuid = cd.ctd_cast_uuid AND cd.cruise_key = '{ck}'
    WHERE m.cruise_key = '{ck}'
      AND m.measurement_type IN ('temperature_ave', 'salinity_ave_corr')
      AND m.measurement_value IS NOT NULL
    GROUP BY cd.phys_cast_id, m.measurement_type, m.depth_m"))
  if (nrow(d_prof) == 0) return(NULL)
  d_prof |>
    arrange(phys_cast_id, measurement_type, depth_m) |>
    group_by(phys_cast_id, measurement_type) |>
    filter(rdp_keep(depth_m, measurement_value,
                    eps = rdp_eps[[measurement_type[1]]])) |>
    ungroup() |>
    distinct(phys_cast_id, depth_m)
}) |>
  purrr::list_rbind() |>
  distinct(phys_cast_id, depth_m)

dbWriteTable(con, "_rdp_retained", rdp_retained,
             temporary = TRUE, overwrite = TRUE)
cat(glue("_rdp_retained: {format(nrow(rdp_retained), big.mark = ',')} ",
         "(profile, depth) pairs flagged as inflections"), "\n")
```

### Assemble ctd_thin

```{r}
#| label: ctd_thin_assemble

# Union three retention reasons, in precedence order: 10 m grid, RDP inflections,
# then bottle-trip depths.
#
# The third is why this section changed. Thinning is designed for DENSE SENSOR
# SCANS: the 10 m backbone and the RDP inflections are both derived from
# temperature/salinity profiles, so a depth is kept because the *sensor* profile
# bends there. Bottle values are not a profile — they are a handful of discrete,
# lab-analysed samples per cast at depths chosen by the watch, and nothing about
# the sensor profile knows where they are. Selecting depths by sensor geometry and
# then subsetting bottle values to those depths silently discards most of them.
#
# Measured against release v2026.07.17: of 133,206 distinct (cast, depth) bottle
# ammonium measurements, only 35,549 (26.7%) sit at a depth this logic retained.
# Roughly half the loss is the cast-direction choice (by design); the rest was
# depth thinning throwing away real lab samples. So flagging btl_ammonium as
# canonical is necessary but NOT sufficient — without this clause it would have
# landed about a quarter of the ammonium record in `obs` and looked like success.
#
# A bottle depth costs nothing to keep (a few rows per cast) and cannot be
# reconstructed by interpolation, so retain every one carrying a canonical value.
canon_btl <- d_meas_ctd |>
  filter(is_canonical, str_starts(measurement_type, "btl_") |
           measurement_type %in% c("salinity_btl", "oxygen_btl_ml_l",
                                   "oxygen_btl_umol_kg")) |>
  pull(measurement_type)
say(glue("bottle-grain canonical types retained whole: ",
         "{if (length(canon_btl)) paste(canon_btl, collapse = ', ') else '(none)'}"))

btl_clause <- if (length(canon_btl)) glue("
  UNION
  SELECT cd.phys_cast_id, m.depth_m, 'bottle' AS retained_reason
  FROM ctd_measurement m
  JOIN _chosen_dir cd ON m.ctd_cast_uuid = cd.ctd_cast_uuid
  WHERE m.measurement_type IN ({paste(sprintf(\"'%s'\", canon_btl), collapse = ', ')})
    AND NOT EXISTS (
      SELECT 1 FROM _grid_depths g
      WHERE g.phys_cast_id = cd.phys_cast_id AND g.depth_m = m.depth_m)
    AND NOT EXISTS (
      SELECT 1 FROM _rdp_retained r
      WHERE r.phys_cast_id = cd.phys_cast_id AND r.depth_m = m.depth_m)") else ""

dbExecute(con, glue("
  CREATE OR REPLACE TEMP TABLE _retained_depths AS
  SELECT phys_cast_id, depth_m, 'grid' AS retained_reason FROM _grid_depths
  UNION
  SELECT r.phys_cast_id, r.depth_m, 'inflection' AS retained_reason
  FROM _rdp_retained r
  WHERE NOT EXISTS (
    SELECT 1 FROM _grid_depths g
    WHERE g.phys_cast_id = r.phys_cast_id AND g.depth_m = r.depth_m)
  {btl_clause}"))

# ctd_thin: ctd_measurement rows that are a canonical type, on the chosen
# direction, at a retained depth — a pure row subset (values never interpolated).
# the QUALIFY de-dups defensively on (ctd_cast_uuid, depth_m, measurement_type):
# the dedup_ctd_raw chunk above already makes ctd_measurement unique, so this is
# a no-op safeguard that keeps ctd_thin's PK guaranteed-unique on its own.
dbExecute(con, glue("
  CREATE OR REPLACE TABLE ctd_thin AS
  WITH candidates AS (
    SELECT
      m.ctd_cast_uuid,
      m.depth_m,
      m.measurement_type,
      m.measurement_value,
      m.measurement_qual,
      cd.cast_dir,
      rd.retained_reason,
      m.cruise_key
    FROM ctd_measurement m
    JOIN _chosen_dir cd      ON m.ctd_cast_uuid = cd.ctd_cast_uuid
    JOIN _retained_depths rd ON rd.phys_cast_id = cd.phys_cast_id
                            AND rd.depth_m      = m.depth_m
    WHERE m.measurement_type IN ({canon_in}))
  SELECT *
  FROM candidates
  QUALIFY ROW_NUMBER() OVER (
    PARTITION BY ctd_cast_uuid, depth_m, measurement_type
    ORDER BY measurement_qual NULLS LAST, measurement_value) = 1
  ORDER BY cruise_key, ctd_cast_uuid, measurement_type, depth_m"))

n_thin <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_thin")$n
n_meas <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_measurement")$n
cat(glue(
  "ctd_thin: {format(n_thin, big.mark = ',')} rows ",
  "({round(100 * n_thin / n_meas, 1)}% of ctd_measurement's ",
  "{format(n_meas, big.mark = ',')})"), "\n")

# deterministic PK from the natural key, mirroring ctd_measurement
assign_deterministic_uuids_md5(
  con        = con,
  table_name = "ctd_thin",
  id_col     = "ctd_thin_uuid",
  key_cols   = c("ctd_cast_uuid", "depth_m", "measurement_type"))

# drop scratch tables; _chosen_dir is kept for the verification chunk that
# follows and dropped there (downstream metadata helpers enumerate all DB tables)
dbExecute(con, "DROP TABLE IF EXISTS _profile_depths")
dbExecute(con, "DROP TABLE IF EXISTS _grid_depths")
dbExecute(con, "DROP TABLE IF EXISTS _rdp_retained")
dbExecute(con, "DROP TABLE IF EXISTS _retained_depths")
```

### Verify ctd_thin

```{r}
#| label: ctd_thin_verify

# 1. row count + thinning ratio
n_thin <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_thin")$n
n_meas <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM ctd_measurement")$n
cat(glue("ctd_thin: {format(n_thin, big.mark = ',')} rows = ",
         "{round(100 * n_thin / n_meas, 1)}% of ctd_measurement"), "\n")
stopifnot(
  "ctd_thin is empty"                          = n_thin > 0,
  "ctd_thin not smaller than ctd_measurement"  = n_thin < n_meas)

# 2. measurement_type is a subset of the canonical list
thin_types <- dbGetQuery(con,
  "SELECT DISTINCT measurement_type FROM ctd_thin")$measurement_type
stopifnot("ctd_thin has non-canonical measurement_type" =
  length(setdiff(thin_types, canon_types)) == 0)

# 3. ctd_thin_uuid is unique
stopifnot("ctd_thin_uuid not unique" = dbGetQuery(con,
  "SELECT COUNT(*) - COUNT(DISTINCT ctd_thin_uuid) AS n FROM ctd_thin")$n == 0)

# 4. faithful subset of ctd_measurement — every ctd_thin row corresponds to a
#    real ctd_measurement row with the same value (thinning selects rows, never
#    interpolates). a NOT EXISTS check, since ctd_measurement has duplicate keys.
stopifnot("ctd_thin rows not a faithful subset of ctd_measurement" =
  dbGetQuery(con, "
    SELECT COUNT(*) AS n
    FROM ctd_thin t
    WHERE NOT EXISTS (
      SELECT 1 FROM ctd_measurement m
      WHERE m.ctd_cast_uuid     = t.ctd_cast_uuid
        AND m.depth_m           = t.depth_m
        AND m.measurement_type  = t.measurement_type
        AND m.measurement_value IS NOT DISTINCT FROM t.measurement_value)")$n == 0)

# 5. retained_reason populated with the expected categories. 'bottle' joins
#    'grid'/'inflection' only when a canonical bottle-grain type exists (see the
#    assemble chunk), so it is required conditionally rather than always.
reason_breakdown <- dbGetQuery(con,
  "SELECT retained_reason, COUNT(*) AS n, COUNT(DISTINCT depth_m) AS n_depths
   FROM ctd_thin GROUP BY 1 ORDER BY n DESC")
expected_reasons <- c("grid", "inflection", if (length(canon_btl)) "bottle")
stopifnot(
  "ctd_thin retained_reason has unexpected values" =
    all(reason_breakdown$retained_reason %in% expected_reasons),
  "ctd_thin retained_reason missing a category" =
    all(expected_reasons %in% reason_breakdown$retained_reason))
# previously computed only for the assertion and never shown
say("retained_reason breakdown:")
print(reason_breakdown)

# 5b. the point of the 'bottle' reason: every canonical bottle-grain measurement on
#     a chosen-direction cast must survive thinning, not just those that happened to
#     land on a sensor-derived depth. Assert 100% retention rather than trusting it.
if (length(canon_btl)) {
  btl_in <- paste(sprintf("'%s'", canon_btl), collapse = ", ")
  btl_cov <- dbGetQuery(con, glue("
    WITH avail AS (
      SELECT m.ctd_cast_uuid, m.depth_m, m.measurement_type
      FROM ctd_measurement m JOIN _chosen_dir cd USING (ctd_cast_uuid)
      WHERE m.measurement_type IN ({btl_in}))
    SELECT (SELECT COUNT(*) FROM avail) AS available,
           (SELECT COUNT(*) FROM ctd_thin WHERE measurement_type IN ({btl_in})) AS retained"))
  say(glue("bottle-grain canonical: {format(btl_cov$retained, big.mark = ',')} of ",
           "{format(btl_cov$available, big.mark = ',')} retained ",
           "({round(100 * btl_cov$retained / btl_cov$available, 1)}%)"))
  stopifnot("every canonical bottle measurement on the chosen direction must be retained" =
              btl_cov$retained == btl_cov$available)
}

# 6. one cast direction per physical cast; no physical cast dropped entirely
coverage <- dbGetQuery(con, glue("
  WITH meas_casts AS (
    SELECT DISTINCT cd.phys_cast_id
    FROM ctd_measurement m
    JOIN _chosen_dir cd ON m.ctd_cast_uuid = cd.ctd_cast_uuid
    WHERE m.measurement_type IN ({canon_in})),
  thin AS (
    SELECT cd.phys_cast_id, COUNT(DISTINCT t.cast_dir) AS n_dir
    FROM ctd_thin t
    JOIN _chosen_dir cd ON t.ctd_cast_uuid = cd.ctd_cast_uuid
    GROUP BY cd.phys_cast_id)
  SELECT
    (SELECT COUNT(*) FROM thin)                  AS n_thin_casts,
    (SELECT COUNT(*) FROM thin WHERE n_dir > 1)  AS n_multi_dir,
    (SELECT COUNT(*) FROM (
       SELECT phys_cast_id FROM meas_casts
       EXCEPT
       SELECT phys_cast_id FROM thin))           AS n_dropped"))
stopifnot(
  "ctd_thin: physical cast with >1 direction" = coverage$n_multi_dir == 0,
  "ctd_thin: physical cast dropped entirely"  = coverage$n_dropped   == 0)
cat(glue("{format(coverage$n_thin_casts, big.mark = ',')} physical casts ",
         "retained (single direction each)"), "\n")

# 7. depth-gap distribution between retained samples
gap_stats <- dbGetQuery(con, "
  WITH g AS (
    SELECT t.depth_m - LAG(t.depth_m) OVER (
             PARTITION BY cd.phys_cast_id, t.measurement_type
             ORDER BY t.depth_m) AS gap_m
    FROM ctd_thin t
    JOIN _chosen_dir cd ON t.ctd_cast_uuid = cd.ctd_cast_uuid)
  SELECT
    ROUND(MEDIAN(gap_m), 2)              AS median_gap_m,
    ROUND(QUANTILE_CONT(gap_m, 0.95), 2) AS p95_gap_m,
    ROUND(MAX(gap_m), 2)                 AS max_gap_m
  FROM g WHERE gap_m IS NOT NULL")
cat(glue("depth gaps between retained samples: median {gap_stats$median_gap_m} m, ",
         "p95 {gap_stats$p95_gap_m} m, max {gap_stats$max_gap_m} m"), "\n")

# 8. every ctd_thin type is registered as canonical in measurement_type
stopifnot("ctd_thin types not registered as canonical in measurement_type" =
  length(dbGetQuery(con, "
    SELECT DISTINCT t.measurement_type
    FROM ctd_thin t
    LEFT JOIN measurement_type mt ON t.measurement_type = mt.measurement_type
    WHERE mt.measurement_type IS NULL OR NOT mt.is_canonical")$measurement_type) == 0)

# done with _chosen_dir — drop before downstream metadata helpers enumerate tables
dbExecute(con, "DROP TABLE IF EXISTS _chosen_dir")

reason_breakdown |>
  dt(caption = "ctd_thin rows by retained_reason", fname = "ctd_thin_reason")
```

### Spot-check thinned profiles

Overlay the full `ctd_measurement` profile (line) against the `ctd_thin` samples (points) for the six temperature profiles with the most retained inflection points — confirming thermocline curvature is tracked while flat segments stay sparse.

```{r}
#| label: ctd_thin_spotcheck
#| fig-width: 9
#| fig-height: 6

# a ctd_cast_uuid is a single depth sample, not a profile — group by the
# physical cast (cruise_key + cast_key). pick six temperature profiles with the
# most retained inflection points (the hardest cases to thin faithfully).
cast_map <- "(SELECT DISTINCT ctd_cast_uuid, cast_key FROM ctd_cast)"

spot <- dbGetQuery(con, glue("
  WITH thin AS (
    SELECT t.cruise_key, cm.cast_key, t.retained_reason
    FROM ctd_thin t
    JOIN {cast_map} cm ON t.ctd_cast_uuid = cm.ctd_cast_uuid
    WHERE t.measurement_type = 'temperature_ave')
  SELECT cruise_key, cast_key
  FROM thin
  GROUP BY cruise_key, cast_key
  HAVING COUNT(*) >= 20
  ORDER BY COUNT(*) FILTER (WHERE retained_reason = 'inflection') DESC,
           cruise_key, cast_key
  LIMIT 6"))
spot$cast <- paste("cast", seq_len(nrow(spot)))
spot_in <- if (nrow(spot) > 0) {
  paste(sprintf("('%s', '%s')", spot$cruise_key, spot$cast_key), collapse = ", ")
} else "('', '')"

# full ctd_measurement profile (de-duplicated) and the thinned samples
d_full <- dbGetQuery(con, glue("
  SELECT t.cruise_key, cm.cast_key, t.depth_m,
         AVG(t.measurement_value) AS measurement_value
  FROM ctd_measurement t
  JOIN {cast_map} cm ON t.ctd_cast_uuid = cm.ctd_cast_uuid
  WHERE t.measurement_type = 'temperature_ave'
    AND (t.cruise_key, cm.cast_key) IN ({spot_in})
  GROUP BY t.cruise_key, cm.cast_key, t.depth_m
  ORDER BY t.cruise_key, cm.cast_key, t.depth_m"))
d_thin <- dbGetQuery(con, glue("
  SELECT t.cruise_key, cm.cast_key, t.depth_m, t.measurement_value,
         t.retained_reason
  FROM ctd_thin t
  JOIN {cast_map} cm ON t.ctd_cast_uuid = cm.ctd_cast_uuid
  WHERE t.measurement_type = 'temperature_ave'
    AND (t.cruise_key, cm.cast_key) IN ({spot_in})
  ORDER BY t.cruise_key, cm.cast_key, t.depth_m"))

# short facet labels keyed on cruise_key + cast_key
prof_key  <- function(d) paste(d$cruise_key, d$cast_key)
cast_lab  <- setNames(spot$cast, prof_key(spot))
d_full$cast <- cast_lab[prof_key(d_full)]
d_thin$cast <- cast_lab[prof_key(d_thin)]

ggplot2::ggplot() +
  ggplot2::geom_path(
    data = d_full, ggplot2::aes(measurement_value, depth_m),
    color = "grey70", linewidth = 0.3) +
  ggplot2::geom_point(
    data = d_thin,
    ggplot2::aes(measurement_value, depth_m, color = retained_reason),
    size = 1.1) +
  ggplot2::scale_y_reverse() +
  ggplot2::scale_color_manual(
    values = c(grid = "#0077cc", inflection = "#ff6600")) +
  ggplot2::facet_wrap(~ cast, scales = "free", nrow = 2) +
  ggplot2::labs(
    x = "temperature_ave (degC)", y = "depth (m)", color = "retained",
    title = "Full ctd_measurement profile (line) vs ctd_thin samples (points)") +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::theme(legend.position = "bottom")
```

## Schema Diagram

```{r}
#| label: schema

# define PK/FK relationships for visualization and relationships.json
ctd_rels <- list(
  primary_keys = list(
    ctd_cast        = "ctd_cast_uuid",
    ctd_measurement = "ctd_measurement_uuid",
    ctd_thin        = "ctd_thin_uuid",
    ctd_summary     = "ctd_summary_uuid",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "ctd_measurement", column = "ctd_cast_uuid", ref_table = "ctd_cast", ref_column = "ctd_cast_uuid"),
    list(table = "ctd_thin", column = "ctd_cast_uuid", ref_table = "ctd_cast", ref_column = "ctd_cast_uuid"),
    list(table = "ctd_measurement", column = "measurement_type", ref_table = "measurement_type", ref_column = "measurement_type"),
    list(table = "ctd_thin", column = "measurement_type", ref_table = "measurement_type", ref_column = "measurement_type"),
    list(table = "ctd_summary", column = "measurement_type", ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(con, rels = ctd_rels)
```

## Add Spatial

```{r}
#| label: add_spatial

add_point_geom(con, "ctd_cast", lon_col = "longitude", lat_col = "latitude")

# grid already loaded in Cross-Dataset Bridge section
assign_grid_key(con, "ctd_cast")
```

## Derived View

```{r}
#| label: derived_view

dbExecute(
  con,
  "
  CREATE OR REPLACE VIEW ctd_cast_derived AS
  SELECT *,
    datetime_start_utc AT TIME ZONE 'UTC' AT TIME ZONE 'US/Pacific' AS datetime_pst,
    EXTRACT(YEAR FROM datetime_start_utc)  AS year,
    EXTRACT(MONTH FROM datetime_start_utc) AS month,
    EXTRACT(DOY FROM datetime_start_utc)   AS julian_day
  FROM ctd_cast"
)
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

# dataset registry built from authoritative ingest_*.qmd YAML (was dataset.csv)
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
say(glue("dataset: {nrow(d_dataset)} datasets registered"))
```

## Data Quality Diagnostics

Range and sentinel audit over `ctd_measurement`. This exists because the
`v2026.07.17` release shipped values that cannot be real — `salinity_ave_corr`
spanning −45 to 1016 PSU, oxygen maxima around 1e9–1e10 — alongside `-99`
sentinels that were ingested as genuine measurements. Nothing in the pipeline
flagged either, so this section makes both visible at ingest time.

::: {.callout-note title="Empty on a fast re-render — see Findings instead"}
This section audits the **intermediate** `ctd_measurement` table, which only exists
when the heavy path ran. On a re-render with unchanged inputs its chunks are shown
but not evaluated, so they have no output (see [Check for Resumable State]).

That is not a gap in coverage: the equivalent checks over the **published** `obs`
and `obs_ctd_full` are in [Findings](#findings), which reads the parquet and
therefore always runs. This section is the *pre-write* audit; that one is the
*as-shipped* report.
:::

::: {.callout-important title="`-99` is fixed at the source; the extremes are not"}
**`-99` is handled.** The pivot above now deletes it, because the notebook already
treated `-99` as this source's missing marker for longitude/latitude — the rule had
simply never been extended to the measurement columns. The audit below should
therefore report **zero** `-99` rows; if it does not, that fix has regressed.

**The extremes are still only reported.** They do not `stopifnot()`: until the
providers confirm whether values like 2.1e9 mL/L oxygen are bad scans or a units
error (question 02), dropping them would be guessing.
The plausible ranges below are a first pass and belong in
`metadata/measurement_type.csv` as `valid_min`/`valid_max` once agreed — which
would also let the CF netCDF writer emit them as real `valid_min`/`valid_max`
variable attributes.
:::

```{r}
#| label: diag_sentinel
# audit trail for the -99 deletion in the pivot above
if (nrow(sentinel_tally)) {
  sentinel_tally |>
    mutate(pct_of_removed = round(100 * n / sum(n), 1)) |>
    datatable(
      caption  = glue("-99 sentinel rows removed from ctd_measurement ",
                      "({format(n_sentinel, big.mark = ',')} total)"),
      rownames = FALSE,
      options  = list(pageLength = 10, scrollX = TRUE, dom = "tip"))
} else {
  say("no -99 sentinel values found in ctd_measurement")
}
```

```{r}
#| label: diag_ranges

# Plausible physical ranges now come from the REGISTRY, not a tribble local to this
# chunk — `metadata/measurement_type.csv` columns valid_min / valid_max, populated by
# libs/build_ctd_measurement_registry.R. Same values as the inline table they replace;
# declaring them once means they are reviewable in a diff, emittable by the CF netCDF
# writer as real valid_min/valid_max variable attributes, and available to any QC rule
# rather than only to this audit.
#
# Still deliberately generous, and still report-only: they catch impossible values,
# they are not the agreed oceanographic ranges, and nothing below drops a row.
plaus <- d_meas_type |>
  filter(!is.na(valid_min) | !is.na(valid_max)) |>
  select(measurement_type, lo = valid_min, hi = valid_max)

diag_sql <- glue("
  SELECT measurement_type,
         COUNT(*)                                           AS n,
         COUNT(*) FILTER (WHERE measurement_value = -99)     AS n_neg99,
         COUNT(*) FILTER (WHERE isnan(measurement_value)
                             OR NOT isfinite(measurement_value)) AS n_nonfinite,
         MIN(measurement_value)                             AS v_min,
         MAX(measurement_value)                             AS v_max
  FROM ctd_measurement GROUP BY 1")
d_diag <- dbGetQuery(con, diag_sql) |>
  left_join(plaus, by = "measurement_type") |>
  mutate(
    pct_neg99   = round(100 * n_neg99 / n, 2),
    # count out-of-range only where a plausible range is declared, and exclude the
    # -99 sentinel so it is reported once (as a sentinel) rather than twice
    has_range   = !is.na(lo),
    out_of_range = NA_integer_)

for (i in which(d_diag$has_range)) {
  r <- d_diag[i, ]
  d_diag$out_of_range[i] <- dbGetQuery(con, glue("
    SELECT COUNT(*) AS n FROM ctd_measurement
    WHERE measurement_type = '{r$measurement_type}'
      AND measurement_value <> -99
      AND (measurement_value < {r$lo} OR measurement_value > {r$hi})"))$n
}

say(glue("types audited                     : {nrow(d_diag)}"))
say(glue("  with a declared plausible range : {sum(d_diag$has_range)}"))
say(glue("  WITHOUT a range (unchecked)     : {sum(!d_diag$has_range)}"))
say(glue("-99 sentinel values (all types)   : ",
         "{format(sum(d_diag$n_neg99), big.mark = ',')}"))
say(glue("out-of-range values (ranged types): ",
         "{format(sum(d_diag$out_of_range, na.rm = TRUE), big.mark = ',')}"))
```

```{r}
#| label: diag_ranges_tbl
d_diag |>
  transmute(
    measurement_type, n,
    `-99` = n_neg99, `%-99` = pct_neg99, `non-finite` = n_nonfinite,
    min = signif(v_min, 4), max = signif(v_max, 4),
    range = if_else(has_range, glue("{lo}..{hi}"), glue("—")),
    `out of range` = out_of_range) |>
  arrange(desc(`-99` + coalesce(`out of range`, 0L)), measurement_type) |>
  datatable(
    caption  = paste("ctd_measurement range + sentinel audit —",
                     "rows with -99 or out-of-range values sort first"),
    rownames = FALSE,
    options  = list(pageLength = 12, scrollX = TRUE, dom = "ftip")) |>
  formatStyle("-99",          color = styleInterval(0, c("inherit", "#d03b3b"))) |>
  formatStyle("out of range", color = styleInterval(0, c("inherit", "#d03b3b")))
```

### Quality flags — decoded against the controlled vocabulary

`measurement_qual` was a verbatim pass-through for years: carried through every
table, never interpreted, with no vocabulary anywhere in the repo. There is one
now — `metadata/measurement_qual.csv`, recovered from the CalCOFI hydrographic
master's own field documentation (`6` = data OK but taken from CTD, `8` = value
is suspect, `9` = missing data) and confirmed to apply to CTD, whose flags use
the same codes.

::: {.callout-warning title="Decoded and reported — deliberately not acted on"}
Nothing here drops or rewrites a flagged value. Three questions are open with the
data providers (`metadata/calcofi/hydro-master/questions.csv`, two of them
blockers): what the undocumented single-digit codes mean, whether a NULL flag
means "good" or "never assessed", and whether `S_qual`'s 253 distinct values in
the bottle master indicate a bitmask or a corrupted column. Acting on a flag
whose semantics are unconfirmed would be guessing with released data.

**The headline variables carry no flags at all**, and that is the more important
gap: source flags attach to the *component sensors* (`Temp1Q`, `Salt1Q`, `Ox1Q`,
`Ox2Q`), while the canonical types are the *averages* — so quality information is
lost in the mean. Deciding that propagation rule (worst-of? both-must-pass?) is
`ctd-qaqc` work, not something to invent here.
:::

```{r}
#| label: diag_qual

# qual_code is forced to character: every value in the vocabulary is numeric, so a
# type-guessing read makes it a double, and joining that to the VARCHAR
# measurement_qual is a hard error rather than a silent miss.
d_qual_vocab <- read_csv(
  here("metadata/measurement_qual.csv"),
  col_types = cols(qual_code = col_character(), .default = col_guess())) |>
  select(measurement_qual = qual_code, label, description, is_documented)

d_qual_obs <- dbGetQuery(con, "
  SELECT measurement_qual, COUNT(*) AS n,
         COUNT(DISTINCT measurement_type) AS n_types
  FROM ctd_measurement
  WHERE measurement_qual IS NOT NULL
  GROUP BY 1 ORDER BY n DESC") |>
  left_join(d_qual_vocab, by = "measurement_qual") |>
  mutate(
    label         = coalesce(label, "UNRECOGNIZED"),
    is_documented = coalesce(is_documented, FALSE))

n_flagged   <- sum(d_qual_obs$n)
n_unrecog   <- sum(d_qual_obs$n[d_qual_obs$label == "UNRECOGNIZED"])
n_canon_flg <- dbGetQuery(con, glue("
  SELECT COUNT(DISTINCT measurement_type) AS n FROM ctd_measurement
  WHERE measurement_qual IS NOT NULL
    AND measurement_type IN ({canon_in})"))$n

say(glue("flagged measurements        : {format(n_flagged, big.mark = ',')}"))
say(glue("  codes seen                : {nrow(d_qual_obs)}"))
say(glue("  NOT in the vocabulary     : {format(n_unrecog, big.mark = ',')}"))
say(glue("canonical types carrying any flag: {n_canon_flg} of {length(canon_types)}"))

# a code outside the vocabulary is a finding, not noise — surface it loudly rather
# than letting a left_join quietly render it as NA in the table below
if (n_unrecog > 0) {
  say(glue("  unrecognized code(s): ",
           "{paste(d_qual_obs$measurement_qual[d_qual_obs$label == 'UNRECOGNIZED'], collapse = ', ')}"))
}
```

```{r}
#| label: diag_qual_tbl
d_qual_obs |>
  transmute(code = measurement_qual, label, description,
            documented = is_documented, n, `types` = n_types) |>
  datatable(
    caption  = paste("ctd_measurement quality flags decoded against",
                     "metadata/measurement_qual.csv"),
    rownames = FALSE,
    options  = list(pageLength = 10, scrollX = TRUE, dom = "tip")) |>
  formatStyle("documented",
              color = styleEqual(c(TRUE, FALSE), c("inherit", "#d03b3b")))
```

## Validate and Enforce Types

```{r}
#| label: validate

validate_for_release(con)
enforce_column_types(con, d_flds_rd = d_flds_rd)
```

## Preview Tables

One `datatable()` per table, each in its own chunk — `preview_tables()` in a loop
renders only the first DT widget (see `.claude/skills/ingest-new.md`), which is
why several of these previews were previously invisible. Every preview is
horizontally scrollable (`scrollX`) and column-capped: `ctd_cast` is ~40 columns
wide and overflowed the page as a static table.

```{r}
#| label: preview_helper
# shared preview: LIMIT in SQL (never collect a 216M-row table), scrollX so wide
# tables scroll inside their own box instead of overflowing the article, and
# geometry dropped (the R DuckDB driver cannot marshal GEOMETRY).
preview_tbl <- function(tbl, n = 100, cols = NULL) {
  all_cols <- dbGetQuery(con, glue(
    "SELECT column_name, data_type FROM information_schema.columns
      WHERE table_name = '{tbl}'"))
  keep <- all_cols$column_name[all_cols$data_type != "GEOMETRY"]
  if (!is.null(cols)) keep <- intersect(keep, cols)
  n_row <- dbGetQuery(con, glue("SELECT COUNT(*) AS n FROM {tbl}"))$n
  d <- dbGetQuery(con, glue(
    "SELECT {paste(keep, collapse = ', ')} FROM {tbl} LIMIT {n}"))
  datatable(
    d, rownames = FALSE,
    caption = glue("{tbl} — {format(n_row, big.mark = ',')} rows x ",
                   "{nrow(all_cols)} cols (showing first {nrow(d)} rows",
                   if (!is.null(cols)) ", selected cols" else "", ")"),
    options = list(pageLength = 5, scrollX = TRUE, dom = "tip"))
}
```

```{r}
#| label: preview_ctd_cast
# ~40 columns; show the identity + position core and let scrollX handle the rest
preview_tbl("ctd_cast", cols = c(
  "ctd_cast_uuid", "cruise_key", "site_key", "cast_key", "cast_dir",
  "datetime_start_utc", "latitude", "longitude", "ship_key", "data_stage",
  "ord_occ"))
```

```{r}
#| label: preview_ctd_measurement
preview_tbl("ctd_measurement")
```

```{r}
#| label: preview_ctd_thin
preview_tbl("ctd_thin")
```

```{r}
#| label: preview_ctd_summary
preview_tbl("ctd_summary")
```

```{r}
#| label: preview_measurement_type
preview_tbl("measurement_type")
```

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`). These core tables **are** this ingest's output: `release_database.qmd` concatenates the
per-dataset shards rather than re-deriving the core, so there is exactly one
projection to keep correct.

Two grains ship from here. The default `obs` carries the **thinned** `ctd_thin`
series (~5.5M rows) at `realm = 'env'`, which is what every consumer reads. The
full-resolution scan set becomes **`obs_ctd_full`** (~216M rows), a supplemental
output that is hosted and catalogued but excluded from the default table list —
`cc_get_db(supplemental = TRUE)` opts into it. Building it was previously the
release's job, reaching back into this ingest's `ctd_measurement` parquet; it
belongs here, with the dataset that produces it.

`sample` is one row per **physical cast** (`cast_key`), not per scan: `ctd_cast`
is per-scan, so the arm deduplicates to the cast before minting the event.

```{r}
#| label: emit_core

ds_key <- "calcofi_ctd-cast"

# This projection lives here, in the notebook that owns the dataset, rather than in
# a switch(dataset_key, …) arm inside calcofi4db. No taxa: CTD is the env realm.
#
# 1. sample — one row per PHYSICAL cast. ctd_cast is per-scan, so the QUALIFY is
#    load-bearing: without it `sample` would get one row per scan (see the
#    scan-grain gotcha in CLAUDE.md).
#    16 columns, not 15: `data_stage` is append_sample()'s optional trailing
#    column, so the final/preliminary distinction the source insists on ("for
#    non-publication use") reaches consumers instead of dying here (question
#    calcofi_ctd-cast_14).
append_sample(con, glue("
  SELECT * FROM (
    SELECT {ns_key(ds_key, 'cast', 'cast_key')} AS sample_key, 'cast' AS sample_type,
           NULL::VARCHAR AS parent_sample_key,
           {ns_key(ds_key, 'cast', 'cast_key')} AS root_sample_key,
           '{ds_key}' AS dataset_key, grid_key, site_key, cruise_key,
           TRY_CAST(ord_occ AS INTEGER) AS order_occ, latitude, longitude,
           CAST(datetime_start_utc AS TIMESTAMP) AS datetime,
           NULL::DOUBLE AS depth_min_m, NULL::DOUBLE AS depth_max_m,
           NULL::VARCHAR AS tow_type, data_stage
    FROM ctd_cast
  ) q QUALIFY row_number() OVER (PARTITION BY sample_key ORDER BY datetime) = 1"))

# 2. obs — the THINNED scan set is the headline (ctd_thin, not ctd_measurement);
#    full resolution goes to the supplemental obs_ctd_full below.
append_obs(con, glue("
  SELECT 'env', '{ds_key}',
         {ns_key(ds_key, 'cast', 'cc.cast_key')},
         cc.grid_key, cc.cruise_key, cc.latitude, cc.longitude,
         CAST(cc.datetime_start_utc AS TIMESTAMP), t.depth_m, t.depth_m,
         NULL::VARCHAR, NULL::VARCHAR, t.measurement_type, t.measurement_value,
         t.measurement_qual, NULL::DOUBLE
  FROM ctd_thin t JOIN ctd_cast cc ON t.ctd_cast_uuid = cc.ctd_cast_uuid
  WHERE cc.grid_key IS NOT NULL"))

core <- list(
  sample = dbGetQuery(con, "SELECT COUNT(*) FROM sample")[[1]],
  obs    = dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]])
cat(glue(
  "core projection — sample={core$sample %||% 0} obs={core$obs %||% 0}\n"))

n_obs <- dbGetQuery(con, "SELECT COUNT(*) FROM obs")[[1]]
n_exp <- dbGetQuery(con,
  "SELECT COUNT(*) FROM ctd_thin t JOIN ctd_cast cc ON t.ctd_cast_uuid = cc.ctd_cast_uuid
   WHERE cc.grid_key IS NOT NULL")[[1]]
n_cast <- dbGetQuery(con, "SELECT COUNT(DISTINCT cast_key) FROM ctd_cast")[[1]]
stopifnot(
  "obs must be one row per grid-matched thinned scan" = n_obs == n_exp,
  "sample must be one row per PHYSICAL cast, not per scan" =
    dbGetQuery(con, "SELECT COUNT(*) FROM sample")[[1]] == n_cast,
  "every obs.sample_key must resolve in sample" =
    dbGetQuery(con, "SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key)
                     WHERE s.sample_key IS NULL")[[1]] == 0,
  # ctd_cast is filtered to final|preliminary upstream, so a NULL here means the
  # 16th positional column silently fell off the arm rather than that a cast has
  # no stage
  "every sample row must carry data_stage (final|preliminary)" =
    dbGetQuery(con, "SELECT COUNT(*) FROM sample
                     WHERE data_stage NOT IN ('final','preliminary')
                        OR data_stage IS NULL")[[1]] == 0,
  # the whole point of ctd_thin: obs must not silently become the full scan set
  "obs must be fed by ctd_thin, not ctd_measurement" =
    n_obs < dbGetQuery(con, "SELECT COUNT(*) FROM ctd_measurement")[[1]])
cat(glue("obs parity: {format(n_obs, big.mark=',')} thinned rows across ",
         "{format(n_cast, big.mark=',')} casts"), "\n")

# --- obs_ctd_full: supplemental full-resolution scans -----------------------
# heavy (~216M rows); set BUILD_OBS_CTD_FULL=FALSE for a fast structural render.
build_obs_ctd_full <- as.logical(Sys.getenv("BUILD_OBS_CTD_FULL", "TRUE"))
if (build_obs_ctd_full) {
  n_full <- append_obs(con, obs_tbl = "obs_ctd_full", select_sql = "
    SELECT 'env' realm, 'calcofi_ctd-cast' dataset_key,
           'calcofi_ctd-cast:cast:' || CAST(cc.cast_key AS VARCHAR) sample_key,
           cc.grid_key, cc.cruise_key, cc.latitude, cc.longitude,
           CAST(cc.datetime_start_utc AS TIMESTAMP) datetime,
           m.depth_m depth_min_m, m.depth_m depth_max_m,
           NULL::VARCHAR taxon_key, NULL::VARCHAR life_stage,
           m.measurement_type, m.measurement_value, m.measurement_qual,
           NULL::DOUBLE measurement_prec
    FROM ctd_measurement m
    JOIN ctd_cast cc ON m.ctd_cast_uuid = cc.ctd_cast_uuid
    WHERE cc.grid_key IS NOT NULL")
  stopifnot(
    "obs_ctd_full must cover every grid-matched scan" = n_full ==
      dbGetQuery(con, "SELECT COUNT(*) FROM ctd_measurement m
                       JOIN ctd_cast cc ON m.ctd_cast_uuid = cc.ctd_cast_uuid
                       WHERE cc.grid_key IS NOT NULL")[[1]])
  cat(glue("obs_ctd_full: {format(n_full, big.mark=',')} rows"), "\n")
} else {
  cat("obs_ctd_full skipped (BUILD_OBS_CTD_FULL=FALSE)\n")
}
```


## Write Parquet

```{r}
#| label: write_parquet

# collect mismatches for manifest
mismatches <- list(
  ships             = collect_ship_mismatches(con, "ctd_cast"),
  measurement_types = collect_measurement_type_mismatches(
    con, here("metadata/measurement_type.csv")),
  cruise_keys       = collect_cruise_key_mismatches(con, "ctd_cast"))

# core shards + the supplemental outputs this dataset owns. ctd_cast/ctd_thin/
# ctd_measurement/ctd_summary are superseded by sample/obs/obs_ctd_full, and
# ctd_wide is retired (ERDDAP reads DuckDB views + netCDF now).
tbls_out <- core_output_tables(
  con, extra = c("obs_ctd_full", "measurement_type"))
parquet_stats <- write_parquet_outputs(
  con          = con,
  output_dir   = dir_parquet,
  tables       = tbls_out,
  partition_by = list(
    obs          = "cruise_key",
    obs_ctd_full = "cruise_key"),
  sort_by = list(
    obs          = c("measurement_type", "depth_min_m"),
    obs_ctd_full = c("measurement_type", "depth_min_m"),
    sample       = "hilbert:longitude,latitude"),
  strip_provenance = FALSE,
  mismatches       = mismatches,
  supplemental     = c("obs_ctd_full")
)

parquet_stats |>
  dt(fname = "ctd_parquet_stats")
```

## Write Metadata JSON

```{r}
#| label: write_metadata

metadata_path <- build_metadata_json(
  con                  = con,
  d_tbls_rd            = d_tbls_rd,
  d_flds_rd            = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           glue("{dir_meta}/metadata_derived.csv")),
  output_dir           = dir_parquet,
  tables               = tbls_out,
  provider             = provider,
  dataset              = dataset,
  workflow_url         = cc$workflow_url,
  tables_owned         = tables_owned
)

# write relationships.json sidecar with PKs/FKs
build_relationships_json(
  rels       = core_relationships(tbls_out),
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset
)
```

Both sidecars are read back **from disk** rather than from the objects the chunk
above returned, and both viewers always evaluate. They describe files that exist in
either mode, so gating them on the heavy path would have silently dropped two of
this notebook's published outputs from a fast render — which is exactly the kind of
quiet degradation a skip path must not introduce.

```{r}
#| label: show_metadata_json
#| eval: true

path_meta_json <- file.path(dir_parquet, "metadata.json")
if (file_exists(path_meta_json)) {
  listviewer::jsonedit(
    jsonlite::fromJSON(path_meta_json, simplifyVector = FALSE), mode = "view")
} else {
  say(glue("metadata.json not found at {path_meta_json}"))
}
```

```{r}
#| label: show_relationships_json
#| eval: true

path_rels_json <- file.path(dir_parquet, "relationships.json")
if (file_exists(path_rels_json)) {
  listviewer::jsonedit(
    jsonlite::fromJSON(path_rels_json, simplifyVector = FALSE), mode = "view")
} else {
  say(glue("relationships.json not found at {path_rels_json}"))
}
```

## Export Modified Dependency Deltas

```{r}
#| label: export_new_deltas

# export _new delta sidecars for modified dependency tables (from calcofi.modifies)
# these are NOT in the manifest — picked up by build_release_table_registry()
for (tbl in modifies_tables) {
  pk_col   <- modifies_pks[[tbl]]$pk_col
  keys_old <- modifies_pks[[tbl]]$keys
  keys_new <- dbGetQuery(con, glue("SELECT {pk_col} FROM {tbl}"))[[1]]
  additions <- setdiff(keys_new, keys_old)

  if (length(additions) > 0) {
    pq_path <- file.path(dir_parquet, paste0(tbl, "_new.parquet"))
    vals <- paste(shQuote(additions, "sh"), collapse = ", ")
    export_parquet(con,
      glue("SELECT * FROM {tbl} WHERE {pk_col} IN ({vals})"), pq_path)
    say(glue("{length(additions)} new {tbl} row(s) -> {tbl}_new.parquet"))
  } else {
    say(glue("No new {tbl} rows — {tbl}_new not exported"))
  }
}
```

## Upload to GCS

```{r}
#| label: resume_for_upload
#| eval: true

# always re-enable eval for upload and cleanup
knitr::opts_chunk$set(eval = TRUE)
```

```{r}
#| label: upload_gcs

# skip sync if all parquet tables were unchanged (no re-write)
if (parquet_complete) {
  say("Parquet unchanged — skipping GCS sync")
} else {
  sync_to_gcs(
    local_dir    = dir_parquet,
    gcs_prefix   = glue("ingest/{dir_label}"),
    bucket       = "calcofi-db",
    delete_stale = TRUE
  )
}
```

## Findings {#findings}

What this ingest just published, and the condition it is in. The sections above
are the *build*; this is the *report* — the place a reader should be able to learn
what is known to be wrong with CalCOFI CTD data without opening an app, a rule
file or a release.

Three things make it worth having here rather than only in `ctd-qaqc`:

- it reads the **parquet that was just written**, not the wrangling database, so it
  is checking what shipped rather than what was in memory before shipping;
- it runs the **same rule registry** (`metadata/qc_rules/`) through the same engine
  (`calcofi4db::qc_run_all()`) that the app runs, so a number here and a number
  there cannot disagree;
- it renders on a **fast re-run** (see [Check for Resumable State]), so the
  narrative can be revised without an hour of recomputation.

It sits after the GCS upload deliberately. Everything above is the ingest's actual
product and is already delivered by the time this runs; a broken diagnostic must
not be able to discard it — the lesson the Gantt appendix records.

```{r}
#| label: findings_setup

# in-memory DuckDB with VIEWS over this run's parquet. Views cost nothing until
# queried, and reading the outputs rather than the wrangling tables is what makes
# this section render identically on a full run and on a fast re-run.
fcon <- get_duckdb_con(":memory:")
dbExecute(fcon, "SET threads TO 4")

pq_view <- function(tbl, path) {
  if (!file_exists(path) && !dir_exists(path)) return(FALSE)
  src <- if (dir_exists(path))
    glue("read_parquet('{path}/**/*.parquet', hive_partitioning = true)") else
      glue("read_parquet('{path}')")
  dbExecute(fcon, glue("CREATE OR REPLACE VIEW {tbl} AS SELECT * FROM {src}"))
  TRUE
}
has_obs  <- pq_view("obs",          file.path(dir_parquet, "obs"))
has_smp  <- pq_view("sample",       file.path(dir_parquet, "sample.parquet"))
has_full <- pq_view("obs_ctd_full", file.path(dir_parquet, "obs_ctd_full"))
stopifnot("no obs/sample parquet to report on" = has_obs && has_smp)

# Reference inputs the rules join against — the quality vocabulary, the harmonic
# climatology and station bottom depths mined from the Access master, and a
# seafloor depth per cast from the GEBCO crop that apps/ctd-viz already commits.
# None of it is part of the release. A missing input is left as a MISSING TABLE so
# its rules ERROR rather than returning zero rows and reading as clean.
gebco_tif <- normalizePath(
  file.path(here(), "..", "apps/ctd-viz/data/gebco_calcofi.tif"), mustWork = FALSE)
say("QC reference data staged:")
qc_ref <- qc_stage_reference(fcon, here(), gebco_tif = gebco_tif)
```

```{r}
#| label: findings_headline

f_head <- dbGetQuery(fcon, "
  SELECT COUNT(*)                                       AS n_obs,
         COUNT(DISTINCT sample_key)                     AS n_casts,
         COUNT(DISTINCT cruise_key)                     AS n_cruises,
         COUNT(DISTINCT measurement_type)               AS n_types,
         CAST(MIN(datetime) AS DATE)                    AS d0,
         CAST(MAX(datetime) AS DATE)                    AS d1,
         COUNT(*) FILTER (WHERE measurement_value = -99) AS n_neg99
  FROM obs")
n_smp  <- dbGetQuery(fcon, "SELECT COUNT(*) AS n FROM sample")$n
f_full <- if (has_full) dbGetQuery(fcon,
  "SELECT COUNT(*) AS n, COUNT(*) FILTER (WHERE measurement_value = -99) AS n_neg99
   FROM obs_ctd_full") else NULL

say(glue("obs           : {format(f_head$n_obs, big.mark = ',')} rows, ",
         "{f_head$n_types} measurement types, ",
         "{format(f_head$n_casts, big.mark = ',')} casts, ",
         "{f_head$n_cruises} cruises, {f_head$d0} to {f_head$d1}"))
say(glue("sample        : {format(n_smp, big.mark = ',')} cast events"))
if (!is.null(f_full))
  say(glue("obs_ctd_full  : {format(f_full$n, big.mark = ',')} rows (supplemental)"))
```

### The `-99` sentinel — fixed at the source, and now guarded

`-99` is this source's documented missing marker. The notebook had always known
that — `pseudoNA_values <- c(-9.99e-29, -99)` strips it from longitude and latitude
— but the rule was never extended to the measurement columns, so `-99` flowed
through `ctd_measurement` → `ctd_thin` → `obs` as though it were a reading.
Release `v2026.07.17` shipped **84,302** of them, including canonical oxygen
(`oxygen_ml_l_ave_sta_corr` 953, `oxygen_umol_kg_ave_sta_corr` 4,294) alongside
`isus_v` 40,479, `ph` 31,493 and `spar` 6,189.

The `NOT isnan / isfinite` guard in the pivot could never catch it: `-99` is a
perfectly finite double. A `-99` mL/L oxygen silently corrupts any mean, minimum
or anomaly a consumer computes, which is why the pivot now **deletes** the rows
rather than reporting them — a missing value has no business existing as a row in
long format.

The count below is an assertion, not a statistic: it is the regression guard for
that fix, and the `ctd_sentinel_neg99` rule is the same test run from the registry.

```{r}
#| label: findings_sentinel

say(glue("-99 in obs          : {f_head$n_neg99}   (was 84,302 in v2026.07.17)"))
if (!is.null(f_full))
  say(glue("-99 in obs_ctd_full : {f_full$n_neg99}"))

stopifnot(
  "the -99 sentinel is back in obs — the pivot fix has regressed" =
    f_head$n_neg99 == 0,
  "the -99 sentinel is back in obs_ctd_full" =
    is.null(f_full) || f_full$n_neg99 == 0)
```

### `btl_*` promoted to canonical — and the depth-thinning trap

The CTD cast files ship the bottle values taken on the same cast: nutrients,
chlorophyll, and the lab references (`BTL_Temp`, `SaltB`, `OxB`) that sensor
calibration is judged against. Eleven of those twelve types were registered but
flagged `is_canonical = FALSE`, so they were excluded from `ctd_thin` and therefore
from `obs` — only `btl_ammonium` was canonical, which read like a stray edit rather
than a decision. Making the rest canonical is what enables the bottle-vs-sensor
calibration rules below; without the reference side, those rules have nothing to
compare and (correctly) report `skip`, not `pass`.

::: {.callout-important title="Flagging them canonical was necessary but not sufficient"}
Thinning is designed for **dense sensor scans**: the ~10 m backbone and the RDP
inflections are both derived from the temperature/salinity profile, so a depth is
kept because *the sensor profile bends there*. Bottle values are not a profile —
they are a handful of discrete lab samples at depths chosen by the watch, and
nothing about the sensor profile knows where they are.

Selecting depths by sensor geometry and then subsetting bottle values to those
depths silently discards most of them. Measured against `v2026.07.17`: of 133,206
distinct (cast, depth) bottle ammonium measurements, only **35,549 (26.7%)** sat at
a retained depth. So promoting the types alone would have landed about a quarter of
the bottle record in `obs` and looked like a success.

`ctd_thin` therefore keeps a third retention reason — `bottle` — that retains every
depth carrying a canonical bottle-grain value. The table below verifies it: within
the cast direction `obs` covers, retention must be **100%**.
:::

```{r}
#| label: findings_bottle

# bottle-grain types carry the marker as a prefix (btl_ammonium), a suffix
# (salinity_btl) or an infix (oxygen_btl_ml_l) — three positions, hence two LIKEs
btl_where <- "(measurement_type LIKE 'btl\\_%' ESCAPE '\\'
               OR measurement_type LIKE '%\\_btl%' ESCAPE '\\')"

f_btl <- dbGetQuery(fcon, glue("
  SELECT COUNT(*) FILTER (WHERE {btl_where}) AS n_btl, COUNT(*) AS n FROM obs"))
say(glue("bottle-grain rows in obs: {format(f_btl$n_btl, big.mark = ',')} of ",
         "{format(f_btl$n, big.mark = ',')} ",
         "({round(100 * f_btl$n_btl / f_btl$n, 1)}%)"))

# retention WITHIN the direction obs keeps. The other direction is not loss: the
# source duplicates every bottle value onto both cast records (question 05), so
# comparing against the full scan set would report a 50% "loss" that is really the
# duplicate copy being dropped.
if (has_full) {
  f_ret <- dbGetQuery(fcon, glue("
    WITH chosen AS (SELECT DISTINCT sample_key FROM obs),
    f AS (SELECT DISTINCT sample_key, depth_min_m, measurement_type
          FROM obs_ctd_full
          WHERE {btl_where} AND sample_key IN (SELECT sample_key FROM chosen)),
    t AS (SELECT DISTINCT sample_key, depth_min_m, measurement_type
          FROM obs WHERE {btl_where})
    SELECT COUNT(*) AS n_avail,
           COUNT(*) FILTER (WHERE t.sample_key IS NOT NULL) AS n_kept
    FROM f LEFT JOIN t USING (sample_key, depth_min_m, measurement_type)"))
  say(glue("bottle values retained by thinning: ",
           "{format(f_ret$n_kept, big.mark = ',')} of ",
           "{format(f_ret$n_avail, big.mark = ',')} on the chosen direction ",
           "({round(100 * f_ret$n_kept / f_ret$n_avail, 1)}%)"))
  stopifnot(
    "depth thinning is dropping bottle values again (retained_reason = 'bottle')" =
      f_ret$n_kept == f_ret$n_avail)
}
```

```{r}
#| label: findings_bottle_tbl

dbGetQuery(fcon, glue("
  SELECT measurement_type, COUNT(*) AS n,
         COUNT(DISTINCT sample_key) AS n_casts,
         ROUND(MIN(measurement_value), 3) AS v_min,
         ROUND(MAX(measurement_value), 3) AS v_max
  FROM obs WHERE {btl_where} GROUP BY 1 ORDER BY n DESC")) |>
  datatable(
    caption  = paste("Bottle-grain measurement types now carried in the default",
                     "obs — the reference side of every calibration check"),
    rownames = FALSE,
    options  = list(pageLength = 12, scrollX = TRUE, dom = "tip"))
```

### One direction per physical cast: half of `sample` carries no `obs`

`sample` is one row per `cast_key`, and `cast_key` carries the direction suffix
(`…d` / `…u`) — so a station occupation that recorded both a down- and an upcast
contributes **two** `sample` rows. `obs` carries `ctd_thin`, which by design keeps
one direction per physical cast. The consequence is arithmetic, but it is not
stated anywhere a consumer would look: roughly half of the CTD rows in the release's
`sample` have no observations attached to them in the default tier.

That is not data loss — the other direction is in the supplemental
`obs_ctd_full` — but a consumer counting casts from `sample` and a consumer
counting casts from `obs` will get answers that differ by about a factor of two,
and neither is wrong. Which direction should be authoritative is question
`calcofi_ctd-cast_06`.

```{r}
#| label: findings_direction

f_dir <- dbGetQuery(fcon, "
  SELECT COUNT(*) AS n_sample,
         COUNT(*) FILTER (
           WHERE sample_key IN (SELECT DISTINCT sample_key FROM obs)) AS n_with_obs
  FROM sample")
say(glue("sample rows                 : {format(f_dir$n_sample, big.mark = ',')}"))
say(glue("  with any obs (default tier): {format(f_dir$n_with_obs, big.mark = ',')} ",
         "({round(100 * f_dir$n_with_obs / f_dir$n_sample, 1)}%)"))
say(glue("  no obs (other direction)   : ",
         "{format(f_dir$n_sample - f_dir$n_with_obs, big.mark = ',')}"))
```

### Quality codes, as published

`measurement_qual` was a verbatim pass-through for years — carried through every
table, never interpreted, with no vocabulary anywhere in the repo. There is one
now: `metadata/measurement_qual.csv`, recovered from the CalCOFI hydrographic
master's own field documentation (`6` = data OK but taken from CTD, `8` = value is
suspect, `9` = missing data) and confirmed to apply to CTD, whose flags use the
same codes. The codes were also being stored as `"9.0"` by a double→string cast, so
none of them matched the vocabulary; stripping that textually (not via an `INTEGER`
cast, which would round an unexpected `"9.5"`) turned 3,860 spurious
"unrecognized code" hits into zero.

Two things below are worth a reader's attention, and both are open with the
providers rather than acted on here:

1. **The headline variables carry no flags at all.** Source flags attach to the
   *component sensors* (`Temp1Q`, `Salt1Q`, `Ox1Q`, `Ox2Q`) while the canonical
   types are the *averages*, so the quality information is lost in the mean. A
   consumer cannot currently tell whether a headline CTD value is trustworthy
   (question `calcofi_ctd-cast_09`).
2. **Every row flagged `9` = "missing data" carries a value** — including
   physically impossible ones. Either the flag or the value is wrong, and nothing
   in the source says which (question `calcofi_ctd-cast_13`).

```{r}
#| label: findings_qual

f_qual <- dbGetQuery(fcon, "
  SELECT o.measurement_qual AS code, q.label, q.description,
         COUNT(*) AS n,
         COUNT(o.measurement_value)              AS n_with_value,
         COUNT(DISTINCT o.measurement_type)      AS n_types,
         ROUND(MIN(o.measurement_value), 3)      AS v_min,
         ROUND(MAX(o.measurement_value), 3)      AS v_max
  FROM obs o LEFT JOIN measurement_qual q ON o.measurement_qual = q.qual_code
  WHERE o.measurement_qual IS NOT NULL
  GROUP BY 1, 2, 3 ORDER BY n DESC")

n_typed <- dbGetQuery(fcon, "
  SELECT COUNT(DISTINCT measurement_type) AS n_all,
         COUNT(DISTINCT measurement_type) FILTER (
           WHERE measurement_qual IS NOT NULL) AS n_flagged FROM obs")

say(glue("flagged obs rows       : {format(sum(f_qual$n), big.mark = ',')} of ",
         "{format(f_head$n_obs, big.mark = ',')} ",
         "({round(100 * sum(f_qual$n) / f_head$n_obs, 3)}%)"))
say(glue("measurement types with any flag: {n_typed$n_flagged} of {n_typed$n_all}"))
n_missing_flag <- sum(f_qual$n_with_value[f_qual$code == "9"])
if (length(n_missing_flag) && n_missing_flag > 0)
  say(glue("rows flagged 9 = 'missing data' that nonetheless carry a value: ",
           "{format(n_missing_flag, big.mark = ',')}"))
```

```{r}
#| label: findings_qual_tbl

f_qual |>
  mutate(label = coalesce(label, "UNRECOGNIZED")) |>
  datatable(
    caption  = paste("Quality codes in the published obs, decoded against",
                     "metadata/measurement_qual.csv"),
    rownames = FALSE,
    options  = list(pageLength = 10, scrollX = TRUE, dom = "tip")) |>
  formatStyle("label",
              color = styleEqual("UNRECOGNIZED", "#d03b3b", default = "inherit"))
```

### What the source documentation says about these numbers

Each source zip ships its own documentation (`CTD-CSV-Format.pdf`,
`CTD Data Files.pdf`, `FilenamingAndAscHeaderNotes.txt`, and a
`PreliminaryDataReadme.txt` on preliminary cruises). Combed 2026-08-01; the full
account is in the
[QA/QC protocol](https://calcofi.io/workflows/ctd-cast_qa-qc-protocol.html).
Four points change how a consumer should read what this ingest publishes.

**The canonical types are the providers' own recommendation, not our preference.**
Every property is published three times — SBE-processed, `_CruiseCorr` (one
regression per cruise, from 4-second averages against ~1400 bottles) and
`_StaCorr` (a regression per cast, from 1 m bin averages against that cast's ~20
bottles). The source states station-corrected data are the best, particularly for
nitrate, which is why `*_sta_corr` are the types flagged canonical here. Salinity
is corrected separately, from **bottle salinities below 350 m**, applied at all
depths.

**A missing `*_sta_corr` is often correct.** Station correction needs a 500 m cast
with roughly ten or more bottles. On winter and spring cruises, lines north of
76.7 may be occupied with as few as 0–12 bottles and receive cruise-corrected
values only. Counting nulls as defects will mis-read a whole class of casts.

**Values are 1 m bin averages, not raw scans.** `obs_ctd_full` is full *published*
resolution. That is also why the loop-edit rule's 1 m threshold is a noise floor
rather than a tolerance.

::: {.callout-note title="`data_stage` now reaches the release"}
The source is explicit that preliminary data are **for non-publication use**, and
specific about what moves: *"Temperature and salinities may change very little but
oxygen, nitrate and chlorophyll data may change significantly after post-cruise
calibrations."*

Until v2026.08 the released `sample` carried no such column, so a consumer could
not tell a final cruise from a preliminary one and the caveat above could not
reach them — question `calcofi_ctd-cast_14`. Core `sample` now has
**`data_stage`**, supplied by this ingest as the optional trailing column of
`append_sample()` (calcofi4db 3.4.0). It is `final` or `preliminary` here; the
other 15 ingests, none of which draw the distinction, leave it `NULL` rather than
having to change.
:::

```{r}
#| label: findings_stage

# how much of what we publish is preliminary? Read from the wrangling DB when the
# heavy path ran; otherwise say so rather than guessing.
if ("ctd_cast" %in% DBI::dbListTables(con)) {
  dbGetQuery(con, "
    SELECT data_stage, COUNT(DISTINCT cruise_key) AS n_cruises,
           COUNT(DISTINCT cast_key) AS n_casts
    FROM ctd_cast GROUP BY 1 ORDER BY n_casts DESC") |>
    datatable(
      caption = paste("Final vs preliminary in this ingest — carried into the",
                      "release on `sample.data_stage`"),
      rownames = FALSE, options = list(dom = "t"))
} else {
  say(paste("data_stage not available on a fast re-render (ctd_cast is not",
            "materialized); it is charted in the Gantt appendix on a full run."))
}
```

### QC rules over this ingest's output

The rules are **data**, not code: `metadata/qc_rules/rules.csv` plus one
`sql/*.sql` file per rule, versioned with the pipeline that produces the data they
check, so a data manager can review one in a diff. They are executed here by
`calcofi4db::qc_run_all()` — the same engine `apps/ctd-qaqc` uses, moved into the
package once it had two callers so the app and this notebook cannot report
different numbers for the same rule.

Every rule targets `obs` / `sample`, never a per-dataset table. That is what lets
the identical registry run against a release, against this parquet, and (next)
against an uploaded file.

::: {.callout-note title="`skip` is not `pass`"}
A rule whose input is absent returns zero rows, which is indistinguishable from
"the data is clean". The engine therefore reports `skip` with a reason, never a
silent pass — this is not hypothetical: the three bottle-vs-sensor rules did
exactly that against `v2026.07.30`, which predates the `btl_*` promotion above.
Rules scoped to a single cruise (they read the full-resolution `obs_ctd_full`) skip
here too, and are run separately below.
:::

```{r}
#| label: findings_rules

dir_rules <- here("metadata/qc_rules")
rules_all <- qc_read_rules(dir_rules, active_only = FALSE)
rules_act <- filter(rules_all, active)
rules_now <- filter(rules_act, coalesce(scope, "all") == "all")

f_types <- qc_present_types(fcon, "calcofi_ctd-cast")
qc_res  <- qc_run_all(fcon, rules_now, limit = 500L, present_types = f_types)
qc_sum  <- qc_summarize(qc_res, rules_now)

say(glue("rules in registry : {nrow(rules_all)} ",
         "({nrow(rules_act)} active, {nrow(rules_all) - nrow(rules_act)} parked)"))
say(glue("run here          : {nrow(rules_now)} release-wide ",
         "(+{nrow(rules_act) - nrow(rules_now)} cruise-scoped, below)"))
for (s in c("FAIL", "ERROR", "flag", "skip", "pass")) {
  n <- sum(qc_sum$status == s)
  if (n > 0) say(glue("  {format(s, width = 6)}: {n}"))
}
say(glue("total findings    : ",
         "{format(sum(qc_sum$n, na.rm = TRUE), big.mark = ',')}"))
```

```{r}
#| label: findings_rules_tbl

qc_sum |>
  transmute(rule_key, status, findings = n, severity, type = rule_type, target,
            description, `sec` = elapsed_s, note) |>
  datatable(
    caption  = paste("QC rule results over this ingest's parquet —",
                     "registry metadata/qc_rules/, engine calcofi4db::qc_run_all()"),
    rownames = FALSE,
    options  = list(pageLength = 15, scrollX = TRUE, dom = "tip")) |>
  formatStyle(
    "status",
    color = styleEqual(c("pass", "flag", "FAIL", "ERROR", "skip"),
                       c("#1a7f37", "#b06000", "#d03b3b", "#d03b3b", "#6e7781")),
    fontWeight = styleEqual(c("FAIL", "ERROR"), c("bold", "bold"),
                            default = "normal"))
```

#### Values outside their declared range

Ranges live in `metadata/measurement_type.csv` (`valid_min` / `valid_max`), moved
there from an inline tribble so they are reviewable in a diff, emittable by the CF
netCDF writer as real variable attributes, and usable by any rule rather than only
by this notebook. They are deliberately generous: they catch values that cannot be
real, they do not police oceanography. Nothing below is dropped.

```{r}
#| label: findings_range_tbl

f_range <- dbGetQuery(fcon, "
  SELECT o.measurement_type,
         COUNT(*)                                    AS n_bad,
         COUNT(DISTINCT o.cruise_key)                AS n_cruises,
         ROUND(MIN(o.measurement_value), 3)          AS v_min,
         ROUND(MAX(o.measurement_value), 3)          AS v_max,
         m.valid_min, m.valid_max
  FROM obs o JOIN measurement_type m USING (measurement_type)
  WHERE m.valid_min IS NOT NULL
    AND (o.measurement_value < m.valid_min OR o.measurement_value > m.valid_max)
  GROUP BY 1, 6, 7 ORDER BY n_bad DESC")

f_range |>
  left_join(
    dbGetQuery(fcon, "SELECT measurement_type, COUNT(*) AS n_total FROM obs GROUP BY 1"),
    by = "measurement_type") |>
  transmute(measurement_type, n_bad, n_total,
            `% bad` = round(100 * n_bad / n_total, 2),
            n_cruises, v_min, v_max,
            range = glue("{valid_min}..{valid_max}")) |>
  datatable(
    caption  = "Published values outside their declared plausible range",
    rownames = FALSE,
    options  = list(pageLength = 15, scrollX = TRUE, dom = "tip"))
```

#### Impossible temperatures

The single finding most worth a provider's time. `temperature_ave` reaches
**60.4 °C** in the published product, and this is not a sensor warming on deck:
most of the values above the 40 °C ceiling are at depth, they span more than two
decades and more than one ship, and where a bottle thermometer sampled the same
cast at the same depth it disagrees by 30–40 °C. Three independent lines — the
range rule, the depth distribution, and the bottle reference — say the same thing,
which is why this is stated as a finding rather than a suspicion.

```{r}
#| label: findings_temp

f_t <- dbGetQuery(fcon, "
  SELECT COUNT(*) FILTER (WHERE measurement_value > 40)                       AS n_hot,
         COUNT(*) FILTER (WHERE measurement_value > 40 AND depth_min_m > 2)   AS n_hot_deep,
         COUNT(*) FILTER (WHERE measurement_value < -2)                       AS n_cold,
         COUNT(DISTINCT cruise_key) FILTER (
           WHERE measurement_value > 40 OR measurement_value < -2)            AS n_cruises
  FROM obs WHERE measurement_type = 'temperature_ave'")
say(glue("temperature_ave > 40 degC : {f_t$n_hot} ",
         "({f_t$n_hot_deep} of them deeper than 2 m — not a deck warm-up)"))
say(glue("temperature_ave < -2 degC : {f_t$n_cold}"))
say(glue("cruises affected          : {f_t$n_cruises}"))
```

```{r}
#| label: findings_temp_tbl

dbGetQuery(fcon, "
  SELECT cruise_key,
         COUNT(*)                             AS n,
         COUNT(DISTINCT sample_key)           AS n_casts,
         ROUND(MIN(depth_min_m), 1)           AS depth_min,
         ROUND(MAX(depth_min_m), 1)           AS depth_max,
         ROUND(MIN(measurement_value), 2)     AS t_min,
         ROUND(MAX(measurement_value), 2)     AS t_max
  FROM obs
  WHERE measurement_type = 'temperature_ave'
    AND (measurement_value > 40 OR measurement_value < -2)
  GROUP BY 1 ORDER BY n DESC") |>
  datatable(
    caption  = paste("Physically impossible temperature_ave by cruise —",
                     "the table to send with question calcofi_ctd-cast_02"),
    rownames = FALSE,
    options  = list(pageLength = 15, scrollX = TRUE, dom = "tip"))
```

```{r}
#| label: findings_temp_btl_tbl

# the corroboration: where a bottle thermometer sampled the same cast at the same
# depth, what did IT read? This is only possible because btl_temperature is now
# canonical (above) — before that the reference side was not in obs at all.
f_tbtl <- dbGetQuery(fcon, "
  WITH hot AS (
    SELECT sample_key, cruise_key, depth_min_m, measurement_value AS v_sensor
    FROM obs
    WHERE measurement_type = 'temperature_ave' AND measurement_value > 40)
  SELECT h.cruise_key,
         COUNT(*)                              AS n_pairs,
         ROUND(MIN(h.depth_min_m), 1)          AS depth_min,
         ROUND(MAX(h.depth_min_m), 1)          AS depth_max,
         ROUND(MIN(b.measurement_value), 2)    AS bottle_min,
         ROUND(MAX(b.measurement_value), 2)    AS bottle_max,
         ROUND(MIN(h.v_sensor), 2)             AS sensor_min,
         ROUND(MAX(h.v_sensor), 2)             AS sensor_max
  FROM hot h
  JOIN obs b ON b.sample_key = h.sample_key AND b.depth_min_m = h.depth_min_m
            AND b.measurement_type = 'btl_temperature'
  GROUP BY 1 ORDER BY n_pairs DESC")

if (nrow(f_tbtl)) {
  datatable(
    f_tbtl,
    caption  = paste("Bottle thermometer vs sensor on the same cast and depth,",
                     "where the sensor reads above 40 degC"),
    rownames = FALSE,
    options  = list(pageLength = 10, scrollX = TRUE, dom = "tip"))
} else {
  say("no bottle temperature at the same cast+depth as any >40 degC sensor value")
}
```

#### Negative pH, clustered by cruise

Not previously reported, and the largest single range violation in the release:
`ph` accounts for the great majority of out-of-range values, and the bad values are
not scattered — they cluster tightly *within* a cruise, often thousands of rows
sitting in a narrow band around −2.5 to −3. Scattered impossible values look like
bad scans; a whole cruise pinned at −2.7 looks like a column that was never
converted from sensor voltage to pH units. The dataset has two explicit voltage
columns (`isus_v`, `fluorescence_v`) but no `ph_v` twin, so a raw-voltage `ph`
would have nowhere else to go. That is a hypothesis for the providers, not a
conclusion (question `calcofi_ctd-cast_12`).

```{r}
#| label: findings_ph_tbl

dbGetQuery(fcon, "
  SELECT cruise_key,
         COUNT(*)                                                        AS n_bad,
         ROUND(MIN(measurement_value), 2)                                AS v_min,
         ROUND(MAX(measurement_value), 2)                                AS v_max,
         COUNT(*) FILTER (WHERE measurement_value < 0)                   AS n_negative,
         ROUND(MAX(measurement_value) - MIN(measurement_value), 2)       AS spread
  FROM obs
  WHERE measurement_type = 'ph' AND (measurement_value < 6 OR measurement_value > 9)
  GROUP BY 1 ORDER BY n_bad DESC") |>
  datatable(
    caption  = paste("Out-of-range pH by cruise — a narrow `spread` on thousands",
                     "of rows is the signature of an unconverted column"),
    rownames = FALSE,
    options  = list(pageLength = 10, scrollX = TRUE, dom = "tip"))
```

#### Full-resolution profile rules, one cruise at a time

Spikes, loop edits and up/down disagreement are invisible in `obs`: thinning keeps
one direction and roughly one sample per 10 m, which is exactly the structure these
rules examine. They read the supplemental `obs_ctd_full` instead, which is
hive-partitioned by `cruise_key` so a scoped query prunes to one cruise rather than
scanning 212M rows.

The cruise below is chosen by evidence, not by recency: it is the one with the most
out-of-range canonical values, so the profile rules run where there is most reason
to look. This is a **sample, not a sweep** — the remaining cruises are the app's job.

```{r}
#| label: findings_cruise_rules

rules_cruise <- filter(rules_act, coalesce(scope, "all") == "cruise")

if (has_full && nrow(rules_cruise)) {
  worst_cruise <- dbGetQuery(fcon, "
    SELECT o.cruise_key, COUNT(*) AS n
    FROM obs o JOIN measurement_type m USING (measurement_type)
    WHERE m.valid_min IS NOT NULL
      AND (o.measurement_value < m.valid_min OR o.measurement_value > m.valid_max)
    GROUP BY 1 ORDER BY n DESC LIMIT 1")$cruise_key
  say(glue("cruise scoped to: {worst_cruise} (most out-of-range values)"))

  qc_res_cr <- qc_run_all(
    fcon, rules_cruise, limit = 500L, present_types = f_types,
    scope_values = list(cruise_key = worst_cruise))
  qc_sum_cr <- qc_summarize(qc_res_cr, rules_cruise)
} else {
  worst_cruise <- NA_character_
  qc_sum_cr    <- NULL
  say("obs_ctd_full not present — the profile rules are SKIPPED, not passed")
}
```

```{r}
#| label: findings_cruise_rules_tbl

if (!is.null(qc_sum_cr)) {
  qc_sum_cr |>
    transmute(rule_key, status, findings = n, severity, type = rule_type,
              description, `sec` = elapsed_s, note) |>
    datatable(
      caption  = glue("Profile rules over obs_ctd_full for cruise {worst_cruise}"),
      rownames = FALSE,
      options  = list(pageLength = 10, scrollX = TRUE, dom = "t")) |>
    formatStyle(
      "status",
      color = styleEqual(c("pass", "flag", "FAIL", "ERROR", "skip"),
                         c("#1a7f37", "#b06000", "#d03b3b", "#d03b3b", "#6e7781")))
} else {
  say("no cruise-scoped rule results this render")
}
```

### Questions for Data Providers

Open questions on this dataset, tracked in
`metadata/calcofi/ctd-cast/questions.csv` and surfaced here so they travel with the
workflow rather than living in someone's inbox. Everything above that could not be
resolved from the data alone is written down here, with the measured evidence that
raised it, so that asking is a matter of sending a link rather than reconstructing
an investigation.

```{r}
#| label: provider-questions

# one validated read + render for every ingest: the vocabulary and the column
# order live in calcofi4db, not in 16 hand-written factor() calls
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the CalCOFI CTD data providers (ranked)")
```

```{r}
#| label: findings_close

close_duckdb(fcon)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)

# remove checkpoint after successful completion (fresh next run)
if (file_exists(db_checkpoint)) {
  file_delete(db_checkpoint)
  say(glue("Removed checkpoint: {db_checkpoint}"))
}

# record the input fingerprint LAST, and only when this run actually rebuilt.
# Written any earlier, a run that died mid-pivot would leave a fingerprint saying
# "these inputs produced those outputs" for outputs that were never finished — and
# the next render would skip the heavy path on the strength of it.
if (!parquet_complete) {
  write_input_fingerprint(fingerprint_path, fp)
  say(glue("Recorded input fingerprint {substr(fp$hash, 1, 12)} → {fingerprint_path}"))
}
```

## Appendix: Interactive Gantt Chart {.appendix}

::: {.callout-note title="This appendix cannot fail the ingest"}
`error: true` is deliberate. Everything above this point — the parquet outputs and
the GCS sync — is the ingest's actual product, and it is all complete by the time
this renders. On 2026-07-30 a stale path in this cosmetic chart aborted the whole
target *after* a successful 15-minute run, so `targets` recorded a failure and
`release_database` never started, over a chart. A decorative appendix should
degrade to a visible error message, not discard finished work.
:::

```{r}
#| label: gantt_guard

# On a fast re-render (inputs unchanged) the heavy path never ran, so ctd_cast was
# never built in the wrangling DB. Gate the chart on its presence rather than
# letting a cosmetic appendix throw — the ingest's real product is already done.
con_pq <- get_duckdb_con(db_path)
load_duckdb_extension(con_pq, "spatial")
has_ctd_cast <- "ctd_cast" %in% DBI::dbListTables(con_pq)
if (!has_ctd_cast) {
  close_duckdb(con_pq)
  say(paste0(
    "ctd_cast not materialized (inputs unchanged, heavy path skipped) — no Gantt ",
    "chart this render. Re-run with CTD_FORCE_REBUILD=TRUE to rebuild it."))
}
```

```{r}
#| label: gantt_chart
#| error: true
#| eval: !expr has_ctd_cast

# Read ctd_cast from the WRANGLING DB, not from parquet.
#
# This chunk used to read '{dir_parquet}/ctd_cast.parquet'. That file stopped
# being written at the consolidated-core cut-over — Write Parquet now exports
# sample / obs / obs_ctd_full / measurement_type, with ctd_cast, ctd_thin,
# ctd_measurement and ctd_summary superseded. The breakage went unnoticed only
# because this target was excluded from `_targets.R` for the whole period since,
# so re-enabling it surfaced a latent failure: the ingest completed all of its
# real work (parquet written, GCS synced) and then died in a cosmetic appendix.
#
# sample.parquet is the natural modern substitute but has no `data_stage`, and
# the final-vs-preliminary split is the point of the chart. ctd_cast is still a
# real table in the wrangling DB at render time, so read it there — the connection
# is opened by the guard chunk above.
cruise_spans <- tbl(con_pq, "ctd_cast") |>
  group_by(cruise_key, data_stage) |>
  summarize(
    beg_date = min(datetime_start_utc, na.rm = TRUE) |> as.Date(),
    end_date = max(datetime_start_utc, na.rm = TRUE) |> as.Date(),
    .groups = "drop"
  ) |>
  collect() |>
  filter(!is.na(beg_date), !is.na(end_date))

# split multi-year cruises into separate rows
cruise_spans <- cruise_spans |>
  rowwise() |>
  mutate(
    years_spanned = list(year(beg_date):year(end_date))
  ) |>
  unnest(years_spanned) |>
  mutate(
    year = years_spanned,
    adj_beg_date = if_else(
      years_spanned == year(beg_date),
      beg_date,
      as.Date(paste0(years_spanned, "-01-01"))
    ),
    adj_end_date = if_else(
      years_spanned == year(end_date),
      end_date,
      as.Date(paste0(years_spanned, "-12-31"))
    ),
    begin_jday = yday(adj_beg_date),
    end_jday = yday(adj_end_date),
    stage_label = data_stage,
    hover_text = glue(
      "Cruise: {cruise_key}<br>",
      "Stage: {data_stage}<br>",
      "Begin: {format(beg_date, '%Y-%m-%d')}<br>",
      "End: {format(end_date, '%Y-%m-%d')}<br>",
      "Duration: {as.numeric(difftime(end_date, beg_date, units = 'days'))} days"
    )
  ) |>
  ungroup() |>
  arrange(adj_beg_date)

colors <- c("final" = "#23d355ff", "preliminary" = "#A23B72")

p <- plot_ly()

for (i in 1:nrow(cruise_spans)) {
  row <- cruise_spans[i, ]
  p <- p |>
    add_trace(
      type = "scatter",
      mode = "lines",
      x = c(row$begin_jday, row$end_jday),
      y = c(row$year, row$year),
      line = list(
        color = colors[row$stage_label],
        width = 8
      ),
      text = row$hover_text,
      hoverinfo = "text",
      showlegend = FALSE,
      legendgroup = row$stage_label
    )
}

p <- p |>
  add_trace(
    type = "scatter",
    mode = "lines",
    x = c(NA, NA),
    y = c(NA, NA),
    line = list(color = colors["final"], width = 8),
    name = "Final",
    showlegend = TRUE
  ) |>
  add_trace(
    type = "scatter",
    mode = "lines",
    x = c(NA, NA),
    y = c(NA, NA),
    line = list(color = colors["preliminary"], width = 8),
    name = "Preliminary",
    showlegend = TRUE
  )

p <- p |>
  layout(
    title = "CalCOFI CTD Cruise Timeline",
    xaxis = list(
      title = "Day of Year",
      range = c(1, 365),
      dtick = 30,
      tickangle = 0
    ),
    yaxis = list(
      title = "Year",
      autorange = "reversed",
      dtick = 1
    ),
    hovermode = "closest",
    plot_bgcolor = "#f8f9fa",
    paper_bgcolor = "white",
    legend = list(
      x = 1.02,
      y = 1,
      xanchor = "left",
      yanchor = "top"
    )
  )

p

close_duckdb(con_pq)
```


::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
