---
title: "Ingest CCE-LTER HPLC Pigments (CalCOFI Cruises)"
calcofi:
  target_name: ingest_cce_lter_hplc_pigments_calcofi
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/cce-lter_hplc-pigments-calcofi/manifest.json
  provider: cce-lter
  dataset: hplc-pigments-calcofi
  workflow_url: https://calcofi.io/workflows/ingest_cce-lter_hplc-pigments-calcofi.html
  questions_file: metadata/cce-lter/hplc-pigments-calcofi/questions.csv
  # staged, not released: first ingest of this holding (CalCOFI/workflows#84).
  # Remove after review, the same way the cetacean ingests were un-held.
  in_release: false
  dataset_meta:
    dataset_name: "High Performance Liquid Chromatography (HPLC) pigment analysis from rosette bottle samples at various depths from CalCOFI-CCE Augmented cruises in the California Current System, 2002 to 2023 (ongoing)"
    dataset_name_short: "HPLC pigments (CalCOFI cruises)"
    category: "Productivity & Pigments"
    color: "#3f7f3a"
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (one Niskin bottle per row; own sample grain)"}
    - {table: obs, shared: true, note: "core observation headline (env realm; 44 HPLC pigment types in ug/L, one set per analysing lab; QC flag in measurement_qual)"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#d5ead2"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: EDI package `knb-lter-cce.316.1` (Goericke, Stukel, Dovel), entity
*ChromatographyPigmentsCalCOFI* (`table_324.csv`; DataZoo dataset 316): 9,153
rows, 60 columns, 81 CalCOFI cruises from 2002-03 to 2023-04. Fetched by
`libs/download_hplc_pigments_calcofi.R`, revision and md5 pinned. Licence: CC0 1.0
(the EML's `intellectualRights`).

This is the **CalCOFI-cruise** HPLC table. The CCE Process-cruise HPLC table
(EDI `knb-lter-cce.72`) is a separate holding (`cce-lter_hplc-pigments`), not
ingested here.

- **Provider**: `cce-lter`.
- **Grain**: one row per HPLC analysis of a Niskin bottle; the sample is the
  bottle, keyed `{studyName}-{Cast Number}-{Bottle Number}`. Seven bottles carry
  two analyses with different values (replicates, Q03); both are published as
  separate obs of the one sample, as in the TOC ingest.
- **Two labs, two sets of measurement types.** Samples from 2002-2014 were run by
  the Goericke Lab (SIO), from 2014 on by Horn Point Laboratory (UMCES); 31
  bottles in 2014 were run by both. The provider's own comparison (EML methods)
  finds some pigments comparable between labs and others not (neoxanthin and
  prasinoxanthin run higher at UMCES; carotene is defined differently). So each
  lab's columns get their own type, `hplc_{pigment}_sio` and
  `hplc_{pigment}_umces` (44 types, `metadata/cce-lter/hplc-pigments-calcofi/pigment_columns.csv`),
  the same way CTD sensor pairs are kept apart. Joining the comparable ones is a
  `variable` crosswalk decision, not made here.
- **Realm**: `env`, µg/L. The provider's `Quality Control flag` (0-3,
  `metadata/measurement_qual.csv` code set `hplc-pigments-calcofi`) is carried on
  every obs of the row as `measurement_qual`.
- **Not published**: `Total chlorophyll b` and `Total Chlorophyll c` for both
  labs (sums of published pigments; asserted), `Gyroxanthin diester (UMCES)`
  (empty), `Associated Bottle Number` (Q05). `Total chlorophyll a` **is**
  published: it is the provider's headline biomass value and the basis of its QC
  flags, and the UMCES total differs from the sum of its parts by rounding.
- **Held out** (each counted and asserted below):
  - 2 rows: one bottle listed at two depths (2005-11 cast 25 bottle 20, 1 m vs
    2 m), the same conflict as `cce-lter_poc-pon` Q02 (Q01).
  - 8 rows: two bottles each listed four times, two sets of values crossed with QC
    flags 0 and 3, so which flag belongs to which value is unknown (Q02).
  - 1 value: `Carotene (Goericke_SIO)` = 10,575 µg/L on 2013-11 cast 41 bottle
    19 (QC flag 2; the next-highest is 0.64 µg/L). The rest of that row is
    published (Q04).
- **Cruise key**: from the ship code in `studyName` (translated to NODC through
  `metadata/ship_renames.csv`, including CCE-LTER's `33SH` for the 2021-05 Bell
  M. Shimada cruise), the sample time and each cruise's designated month, by
  `derive_cruise_key_on_casts()`. This table has no `Cruise` column, so the
  designation comes from `cruise_designation.csv` (the same cruises' `Cruise`
  column in the POC/PON and TOC tables, or the SWFSC cruise for the 11 those
  tables lack). Without it, 701 samples on cruises with no date span fell back
  to the sample's own month, and 20 on the 2023-03-25 Shimada cruise were keyed
  to a cruise that does not exist (2023-03-3322 rather than 2023-04-3322).

```{mermaid}
graph LR
  A[EDI knb-lter-cce.316.1<br/>table_324.csv, 9,153 rows] --> B[hold out conflicting depth<br/>and flag-crossed rows]
  B --> C[hplc_sample<br/>one row per bottle]
  B --> D[hplc_measurement<br/>long: 44 pigment x lab types<br/>+ QC flag]
  C -.ship code + time.-> E[(cruise_key)]
  C --> G[[core: sample]]
  D --> H[[core: obs, realm=env]]
```

## Setup

```{r}
#| label: setup
devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))
cc             <- read_calcofi_meta(here("ingest_cce-lter_hplc-pigments-calcofi.qmd"))
provider       <- cc$provider
dataset        <- cc$dataset
tables_owned   <- cc$tables_owned
ds_key         <- as.character(glue("{provider}_{dataset}"))
dir_label      <- ds_key
dir_parquet    <- here(glue("data/parquet/{dir_label}"))
dir_stage      <- cc_stage_path("parquet", dir_label, create = TRUE)
dir_src        <- cc_stage_path("source", dir_label, create = TRUE)
db_path        <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_meta       <- here(glue("metadata/{provider}/{dataset}"))
publish_to_gcs <- Sys.getenv("CALCOFI_SKIP_GCS") == ""

# source column -> measurement_type, one row per pigment column
d_pig <- read_csv(file.path(dir_meta, "pigment_columns.csv"),
                  col_types = cols(.default = "c", position = "i", published = "l"))
d_pub <- d_pig |> filter(published)

d_meas_type <- read_measurement_type(here("metadata/measurement_type.csv"))
stopifnot(
  "44 published pigment columns" = nrow(d_pub) == 44,
  "every published pigment type is pre-registered" =
    all(d_pub$measurement_type %in% d_meas_type$measurement_type))

d_qual <- read_csv(here("metadata/measurement_qual.csv"), col_types = cols(.default = "c")) |>
  filter(code_set == "hplc-pigments-calcofi")
stopifnot("QC flag codes 0-3 are registered" = setequal(d_qual$qual_code, as.character(0:3)))

if (overwrite) {
  if (file_exists(db_path)) file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp"))) dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Read Source Data

Pulled from EDI by `libs/download_hplc_pigments_calcofi.R` (package
`knb-lter-cce.316.1`, entity *ChromatographyPigmentsCalCOFI*), revision 1 with the
EML's md5 asserted. Bulk input lives under the stage dir.

The real headers carry unit suffixes with non-ASCII characters (`Latitude (º)`,
`Fucoxanthin (UMCES) (µg/L)`). They are checked against the expected header
(written with `\u` escapes, the pigment part built from `pigment_columns.csv`)
and then renamed **by position**.

```{r}
#| label: read-source
source(here("libs/download_hplc_pigments_calcofi.R"))
downloaded <- overwrite_all || !file_exists(file.path(dir_src, "data.csv"))
src_csv <- download_hplc_pigments_calcofi(dir_src, overwrite = overwrite_all)
stopifnot("HPLC data.csv not found" = file_exists(src_csv))
src_stamp <- stamp_source_access(files = src_csv) |>
  mutate(source = "https://doi.org/10.6073/pasta/99e2bad73df5fd4890237d31415820b8",
         method = if (downloaded) "download" else "file_mtime")

if (publish_to_gcs) {
  sync_to_gcs(
    local_dir  = dir_src,
    gcs_prefix = glue("archive/{provider}/{dataset}"),
    bucket     = "calcofi-files-public",
    exclude    = c(".DS_Store", "*.tmp", "*.gdoc"))
} else {
  cat("CALCOFI_SKIP_GCS set -- source NOT archived to gs://calcofi-files-public\n")
}

d_raw <- read_csv(src_csv, col_types = cols(.default = "c"), na = character(),
                  locale = locale(encoding = "UTF-8"))

hdr_meta_expected <- c(
  "studyName", "Datetime GMT", "Latitude (º)", "Longitude (º)",
  "Line", "Station", "Cast Number", "Bottle Number", "Associated Bottle Number",
  "Depth (m)", "Quality Control flag")
hdr_meta_new <- c(
  "study_name", "datetime_gmt", "latitude", "longitude",
  "line", "station", "cast_number", "bottle_number", "associated_bottle_number",
  "depth_m", "qc_flag")
pig_col <- paste0("pig_", d_pig$position)   # positional names for the 49 pigment columns

cat(glue("Read {format(nrow(d_raw), big.mark=',')} rows, {ncol(d_raw)} columns"), "\n")
stopifnot(
  "expected 9,153 rows x 60 columns (EDI revision 1, checked 2026-10-06)" =
    nrow(d_raw) == 9153 && ncol(d_raw) == 60,
  "source header changed: re-check pigment_columns.csv before renaming by position" =
    identical(names(d_raw),
              c(hdr_meta_expected, paste0(d_pig$source_column, " (µg/L)"))),
  "pigment_columns.csv positions follow the header" =
    identical(d_pig$position, 12:60))
names(d_raw) <- c(hdr_meta_new, pig_col)
```

## Clean, Type-Cast

```{r}
#| label: clean-cast
num <- function(x) suppressWarnings(as.numeric(if_else(x == "", NA_character_, x)))

stopifnot(
  "no exact duplicate rows" = !anyDuplicated(d_raw),
  "no -999 sentinel" = !any(as.matrix(d_raw) == "-999"),
  "every pigment cell is blank or a positive number" =
    all(vapply(pig_col, function(p) {
      v <- num(d_raw[[p]]); all(is.na(v) == (d_raw[[p]] == "")) && all(v[!is.na(v)] > 0)
    }, logical(1))),
  "Gyroxanthin diester (UMCES) is empty in this revision" =
    all(d_raw[[pig_col[d_pig$source_column == "Gyroxanthin diester (UMCES)"]]] == ""))

d_row <- d_raw |>
  mutate(
    row_id             = row_number(),
    datetime_start_utc = suppressWarnings(ymd_hms(datetime_gmt, tz = "UTC")),
    across(c(latitude, longitude, line, station, depth_m), num),
    across(c(cast_number, bottle_number, associated_bottle_number), ~ suppressWarnings(as.integer(.x))),
    across(all_of(pig_col), num),
    site_key      = normalize_site_key(sprintf("%.1f %.1f", line, station)),
    # the source's own natural key; '-' not ':' so the id stays one token
    sample_key_id = paste(study_name, cast_number, bottle_number, sep = "-")) |>
  select(-datetime_gmt)

stopifnot(
  "every row parses a datetime, position, line/station, cast, bottle and depth" =
    !anyNA(d_row$datetime_start_utc) && !anyNA(d_row$latitude) &&
    !anyNA(d_row$longitude) && !anyNA(d_row$site_key) &&
    !anyNA(d_row$cast_number) && !anyNA(d_row$bottle_number) && !anyNA(d_row$depth_m),
  "QC flag is a registered code on every row" = all(d_row$qc_flag %in% d_qual$qual_code))
```

### Keys listed more than once

Ten bottles appear on more than one row. They fall in three groups, each counted
and asserted:

1. **Conflicting depth** (1 bottle, 2 rows): 2005-11-04-C-32NM cast 25 bottle 20,
   at 1 m and 2 m with identical values. Held out (Q01).
2. **Flag-crossed** (2 bottles, 8 rows): each listed four times, two sets of
   values each paired with QC flag 0 and with flag 3. Held out (Q02).
3. **Replicates** (7 bottles, 14 rows): two rows that agree on time, position,
   depth, associated bottle and QC flag but differ in pigment values. Both kept
   as obs of the one sample (Q03).

```{r}
#| label: dup-keys
key_meta <- d_row |>
  group_by(sample_key_id) |>
  summarize(
    n_rows  = n(),
    n_where = n_distinct(datetime_start_utc, latitude, longitude, depth_m, associated_bottle_number),
    n_flag  = n_distinct(qc_flag),
    .groups = "drop") |>
  filter(n_rows > 1) |>
  mutate(group = case_when(
    n_where > 1 ~ "conflicting depth (Q01)",
    n_flag  > 1 ~ "flag-crossed (Q02)",
    .default    = "replicate (Q03)"))
key_meta |> datatable(caption = "Bottles listed on more than one row", rownames = FALSE)

stopifnot(
  "1 conflicting-depth bottle: 2005-11-04-C-32NM-25-20" =
    identical(key_meta$sample_key_id[key_meta$group == "conflicting depth (Q01)"],
              "2005-11-04-C-32NM-25-20"),
  "2 flag-crossed bottles, 4 rows each" =
    sum(key_meta$group == "flag-crossed (Q02)") == 2 &&
    all(key_meta$n_rows[key_meta$group == "flag-crossed (Q02)"] == 4),
  "7 replicate bottles, 2 rows each" =
    sum(key_meta$group == "replicate (Q03)") == 7 &&
    all(key_meta$n_rows[key_meta$group == "replicate (Q03)"] == 2))

hold_keys <- key_meta$sample_key_id[key_meta$group != "replicate (Q03)"]
d_hold    <- d_row |> filter(sample_key_id %in% hold_keys)
d_row     <- d_row |> filter(!sample_key_id %in% hold_keys)
stopifnot("10 rows held out" = nrow(d_hold) == 10)

d_sample <- d_row |>
  distinct(sample_key_id, study_name, datetime_start_utc, latitude, longitude,
           line, station, site_key, cast_number, bottle_number,
           associated_bottle_number, depth_m) |>
  arrange(study_name, cast_number, bottle_number) |>
  mutate(sample_id = row_number())
stopifnot(
  "one sample per bottle (replicate rows agree on time, position and depth)" =
    !anyDuplicated(d_sample$sample_key_id),
  "9,136 samples from 9,143 rows" = nrow(d_sample) == 9136 && nrow(d_row) == 9143)
d_row <- d_row |> left_join(d_sample |> select(sample_key_id, sample_id), by = "sample_key_id")

cat(glue(
  "9,153 rows read; {nrow(d_hold)} held out; {format(nrow(d_row), big.mark=',')} analyses ",
  "of {format(nrow(d_sample), big.mark=',')} bottles on {n_distinct(d_sample$study_name)} cruises"), "\n")

dbWriteTable(con, "hplc_sample", d_sample, overwrite = TRUE)
```

### Totals agree with their parts

The provider defines `TChl a = Chl a + DVChl a + Chlide a` for both labs, and
the total chl b and chl c columns as sums of their parts. Totals b and c are not
published because they follow from published pigments; that is checked here. A
blank part counts as 0 (the provider's own sums treat it so). UMCES values are
rounded to 0.001 µg/L, so their sums agree to within rounding.

```{r}
#| label: totals-check
pc <- function(src) d_row[[pig_col[d_pig$source_column == src]]]
z  <- function(x) coalesce(x, 0)
max_diff <- function(tot, parts) max(abs(tot - parts), na.rm = TRUE)
d_tot <- tribble(
  ~total, ~max_abs_diff,
  "TChl a (SIO)",   max_diff(pc("Total chlorophyll a (Goericke_SIO)"),
                             z(pc("Chlorophyll a (Goericke_SIO)")) + z(pc("Divinyl chlorophyll a (Goericke_SIO)")) + z(pc("Chlorophyllide a (Goericke_SIO)"))),
  "TChl a (UMCES)", max_diff(pc("Total chlorophyll a (UMCES)"),
                             z(pc("Chlorophyll a (UMCES)")) + z(pc("Divinyl chlorophyll a (UMCES)")) + z(pc("Chlorophyllide a (UMCES)"))),
  "TChl b (SIO)",   max_diff(pc("Total chlorophyll b (Goericke_SIO)"), z(pc("Chlorophyll b (Goericke_SIO)"))),
  "TChl b (UMCES)", max_diff(pc("Total Chlorophyll b (UMCES)"),
                             z(pc("Chlorophyll b (UMCES)")) + z(pc("Divinyl Chlorophlyll b (UMCES)"))),
  "TChl c (SIO)",   max_diff(pc("Total Chlorophyll c (Goericke_SIO)"),
                             z(pc("Chlorophyll c (Goericke_SIO)")) + z(pc("Chlorophyll c 1 (Goericke_SIO)")) +
                             z(pc("Chlorophyll c 2 (Goericke_SIO)")) + z(pc("Chlorophyll c 3 (Goericke_SIO)"))),
  "TChl c (UMCES)", max_diff(pc("Total Chlorophyll c (UMCES)"),
                             z(pc("Chlorophyll c 12 (UMCES)")) + z(pc("Chlorophyll c 3 (UMCES)"))))
d_tot |> datatable(caption = "Largest |total - sum of parts| (ug/L)", rownames = FALSE)
stopifnot(
  "SIO totals equal the sum of their parts" =
    all(d_tot$max_abs_diff[str_detect(d_tot$total, "SIO")] < 1e-9),
  "UMCES totals equal the sum of their parts within rounding (0.0015 ug/L)" =
    all(d_tot$max_abs_diff[str_detect(d_tot$total, "UMCES")] < 0.0015))
```

## Pivot Measurements to Long Format

One row per (analysis, published pigment) with a value. The QC flag of the row
becomes `measurement_qual`. The carotene value of 10,575 µg/L is held out here
(Q04).

```{r}
#| label: pivot-long
d_meas <- d_row |>
  select(row_id, sample_id, qc_flag, all_of(paste0("pig_", d_pub$position))) |>
  pivot_longer(starts_with("pig_"), names_to = "pig", values_to = "measurement_value",
               values_drop_na = TRUE) |>
  left_join(d_pub |> transmute(pig = paste0("pig_", position), measurement_type), by = "pig") |>
  transmute(row_id, sample_id, measurement_type, measurement_value,
            measurement_qual = qc_flag)

carotene_hold <- d_meas |>
  filter(measurement_type == "hplc_carotene_sio", measurement_value > 100)
stopifnot("one carotene (SIO) value above 100 ug/L: 10,575 on 2013-11 cast 41 bottle 19 (Q04)" =
            nrow(carotene_hold) == 1 &&
            d_sample$sample_key_id[d_sample$sample_id == carotene_hold$sample_id] ==
              "2013-11-08-C-32NM-41-19")
d_meas <- d_meas |> anti_join(carotene_hold, by = c("row_id", "measurement_type")) |>
  arrange(sample_id, measurement_type, row_id) |>
  mutate(measurement_id = row_number(), .before = 1)

dbWriteTable(con, "hplc_measurement", d_meas, overwrite = TRUE)

d_n <- d_meas |> count(measurement_type, name = "n") |>
  left_join(d_pub |> select(measurement_type, source_column), by = "measurement_type")
d_n |> datatable(caption = "Rows per measurement_type", rownames = FALSE)
stopifnot(
  "every published pigment type has values" = nrow(d_n) == 44,
  "154,459 pigment values (one held out)" = nrow(d_meas) == 154459)
```

## Resolve Ship and Cruise Keys

Same as `cce-lter_poc-pon`: the trailing token of `studyName` is the ship code,
translated to NODC by code through `metadata/ship_renames.csv` (CCE-LTER uses
CalCOFI's own codes 325S, 33SR, 33RL, 31OS, 32BH, 33SH). This table has no
`Cruise` designation column, so the cruise's designated month comes from
`metadata/cce-lter/hplc-pigments-calcofi/cruise_designation.csv`: the `Cruise`
column the POC/PON (EDI 54) and TOC (EDI 253) tables give the same `studyName`,
or, for the 11 cruises neither has, the SWFSC cruise whose dates hold it. A
sample is keyed by the cruise whose dates hold it (span), else by that
designation; never by its own month, which split the 2023-03-25 Shimada cruise
across two months on the first render.

```{r}
#| label: resolve-keys
load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  gcs_prefix = "ingest/swfsc_ichthyo",  # public GCS copy when not staged locally
  tables = c("ship", "cruise"), as_view = TRUE)

dbExecute(con, "ALTER TABLE hplc_sample ADD COLUMN IF NOT EXISTS ship_code VARCHAR")
dbExecute(con, "UPDATE hplc_sample SET ship_code = regexp_extract(study_name, '[0-9A-Z]+$')")

d_ship_xw <- read_csv(here("metadata/ship_renames.csv"), col_types = cols(.default = "c")) |>
  distinct(csv_code, csv_code_new) |>
  filter(csv_code != csv_code_new)
stopifnot("each ship code in ship_renames.csv maps to one NODC code" =
            !anyDuplicated(d_ship_xw$csv_code))
dbWriteTable(con, "_ship_code_xw", as.data.frame(d_ship_xw), overwrite = TRUE)
dbGetQuery(con, "
  SELECT s.ship_code AS source_code, x.csv_code_new AS nodc_code, COUNT(*) AS n
  FROM hplc_sample s JOIN _ship_code_xw x ON s.ship_code = x.csv_code
  GROUP BY 1, 2 ORDER BY 1") |>
  datatable(caption = "Ship codes translated to NODC (metadata/ship_renames.csv)", rownames = FALSE)
dbExecute(con, "
  UPDATE hplc_sample SET ship_code = x.csv_code_new
  FROM _ship_code_xw x WHERE hplc_sample.ship_code = x.csv_code")

d_cruise_des <- read_csv(file.path(dir_meta, "cruise_designation.csv"),
                         col_types = cols(.default = "c"))
stopifnot(
  "every studyName has a designated cruise month" =
    setequal(d_cruise_des$study_name, unique(d_sample$study_name)),
  "cruise_ym is YYYYMM" = all(str_detect(d_cruise_des$cruise_ym, "^\\d{4}(0[1-9]|1[0-2])$")))
dbWriteTable(con, "_cruise_des", as.data.frame(d_cruise_des), overwrite = TRUE)
dbExecute(con, "ALTER TABLE hplc_sample ADD COLUMN IF NOT EXISTS cruise_ym VARCHAR")
dbExecute(con, "
  UPDATE hplc_sample SET cruise_ym = d.cruise_ym
  FROM _cruise_des d WHERE hplc_sample.study_name = d.study_name")

cruise_key_result <- derive_cruise_key_on_casts(
  con, table_name = "hplc_sample", datetime_col = "datetime_start_utc",
  cruise_ym_col = "cruise_ym")
cruise_key_result$key_stats |>
  datatable(caption = "cruise_key resolution: span > source designation > month")
stopifnot("no sample is keyed by its own month (every cruise has a designation)" =
            !"month" %in% cruise_key_result$key_stats$method)

n_ck  <- dbGetQuery(con, "SELECT COUNT(*) FROM hplc_sample WHERE cruise_key IS NOT NULL")[[1]]
n_all <- dbGetQuery(con, "SELECT COUNT(*) FROM hplc_sample")[[1]]
cat(glue("cruise_key match: {n_ck}/{n_all} ({round(100*n_ck/n_all,1)}%)"), "\n")
stopifnot("every sample resolves a cruise_key (an unmatched ship code needs a row in metadata/ship_renames.csv)" =
            n_ck == n_all)
dbGetQuery(con, "
  SELECT study_name, cruise_key, COUNT(*) AS n FROM hplc_sample
  GROUP BY 1, 2 ORDER BY 1") |>
  datatable(caption = "studyName -> cruise_key", rownames = FALSE)
```

## Add Spatial

```{r}
#| label: spatial
# spatial last: nothing above has to UPDATE a table that carries geometry
load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  gcs_prefix = "ingest/swfsc_ichthyo",
  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
add_point_geom(con, "hplc_sample", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "hplc_sample") |> datatable(caption = "Grid assignment")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata
this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot(nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)
```

## Schema Documentation

Source shape (what the notebook wrangles):

```{r}
#| label: schema
hplc_rels <- list(
  primary_keys = list(
    hplc_sample      = "sample_id",
    hplc_measurement = "measurement_id",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "hplc_measurement", column = "sample_id",
         ref_table = "hplc_sample", ref_column = "sample_id"),
    list(table = "hplc_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(
  con, tables = c("hplc_sample", "hplc_measurement", "measurement_type", "dataset"),
  rels = hplc_rels,
  colors = list(lightblue = c("hplc_sample", "hplc_measurement"),
                lightyellow = "measurement_type", white = "dataset"))
```

Source to core (what a consumer sees):

| source | core | transform |
|---|---|---|
| `hplc_sample` | `sample` (`bottle`) | one row per bottle; key `cce-lter_hplc-pigments-calcofi:bottle:{studyName}-{Cast Number}-{Bottle Number}`; `depth_min_m = depth_max_m = Depth (m)`; time = `Datetime GMT`; `cruise_key` from ship code + time + designated month (`cruise_designation.csv`) |
| 44 pigment columns (`pigment_columns.csv`) | `obs` `hplc_{pigment}_{sio,umces}` | verbatim, µg/L; one obs per analysis, so a replicate bottle has two of each |
| `Quality Control flag` | `obs.measurement_qual` | verbatim code 0-3 (`measurement_qual.csv`, code set `hplc-pigments-calcofi`) |
| `Total chlorophyll b`, `Total Chlorophyll c` (both labs) | -- | **not published**: sums of published pigments (asserted) |
| `Gyroxanthin diester (UMCES)` | -- | **not published**: empty |
| `Associated Bottle Number` | -- | **not published** (Q05) |
| 10 rows (Q01, Q02); 1 carotene value (Q04) | -- | held out |

## Check Declared Bounds

Every pigment type carries `valid_min = 0` (every value is positive, asserted).
No upper bound is declared.

```{r}
#| label: bounds-check
bounds_pre <- check_measurement_bounds(con, "hplc_measurement", mt = d_meas_type)
bounds_datatable(bounds_pre)

bounds_post <- check_measurement_bounds(
  con, "hplc_measurement", mt = d_meas_type, include_undeclared = FALSE)
stopifnot(
  "hplc_measurement must hold no value outside a DECLARED bound" =
    sum(bounds_post$status == "out_of_range") == 0)
```

## Enforce Column Types

```{r}
#| label: enforce-types
enforce_column_types(con)
```

## Data Preview

```{r}
#| label: preview-sample
cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='hplc_sample' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM hplc_sample LIMIT 200")) |>
  datatable(caption = "hplc_sample -- first 200 rows", rownames = FALSE, filter = "top")
```

```{r}
#| label: preview-measurement
dbGetQuery(con, "SELECT * FROM hplc_measurement LIMIT 200") |>
  datatable(caption = "hplc_measurement -- first 200 rows", rownames = FALSE, filter = "top")
```

## Emit Core Tables

```{r}
#| label: emit_core
append_sample(con, sample_arm_self(
  ds_key, "hplc_sample", "sample_key_id", "bottle",
  dt_col = "datetime_start_utc", site_expr = "site_key",
  depth_min = "depth_m", depth_max = "depth_m"))

append_obs(con, glue("
  SELECT 'env', '{ds_key}', {ns_key(ds_key, 'bottle', 's.sample_key_id')},
         s.grid_key, s.cruise_key, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP), s.depth_m, s.depth_m,
         NULL::VARCHAR, NULL::VARCHAR, m.measurement_type, m.measurement_value,
         CAST(m.measurement_qual AS VARCHAR), NULL::DOUBLE
  FROM hplc_measurement m
  JOIN hplc_sample s USING (sample_id)"))

q1 <- function(sql) dbGetQuery(con, sql)[[1]]
core <- list(
  sample = q1("SELECT COUNT(*) FROM sample"),
  obs    = q1("SELECT COUNT(*) FROM obs"))
cat(glue("core projection -- sample={core$sample} obs={core$obs}"), "\n")

src_total <- q1("SELECT SUM(measurement_value) FROM hplc_measurement")
obs_total <- q1(glue("SELECT SUM(measurement_value) FROM obs WHERE dataset_key = '{ds_key}'"))
d_qual_n  <- dbGetQuery(con, glue("
  SELECT measurement_qual, COUNT(*) AS n FROM obs WHERE dataset_key = '{ds_key}' GROUP BY 1 ORDER BY 1"))
d_qual_n |> datatable(caption = "obs per QC flag (measurement_qual)", rownames = FALSE)
stopifnot(
  "one sample per bottle" = core$sample == nrow(d_sample),
  "obs is one row per measurement" = core$obs == nrow(d_meas),
  "obs value total equals the source" = isTRUE(all.equal(obs_total, src_total)),
  "only the 44 pigment types are emitted" =
    setequal(dbGetQuery(con, glue("SELECT DISTINCT measurement_type FROM obs
                                   WHERE dataset_key = '{ds_key}'"))$measurement_type,
             d_pub$measurement_type),
  "every obs carries a registered QC flag" =
    all(d_qual_n$measurement_qual %in% d_qual$qual_code),
  "sample_key is unique" =
    q1(glue("SELECT COUNT(*) - COUNT(DISTINCT sample_key) FROM sample
             WHERE dataset_key = '{ds_key}'")) == 0,
  "every obs.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key)
        WHERE s.sample_key IS NULL") == 0,
  "every obs.measurement_type resolves in measurement_type" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN measurement_type t USING (measurement_type)
        WHERE t.measurement_type IS NULL") == 0)

# serve the per-dataset name as a VIEW over the core
invisible(dbExecute(con, "DROP TABLE IF EXISTS hplc_measurement_src"))
invisible(dbExecute(con, "ALTER TABLE hplc_measurement RENAME TO hplc_measurement_src"))
invisible(dbExecute(con, glue(
  "CREATE OR REPLACE VIEW hplc_measurement AS
   SELECT o.obs_id AS measurement_id,
          o.sample_key,
          substr(o.sample_key, length('{ds_key}:bottle:') + 1) AS sample_key_id,
          o.measurement_type, o.measurement_value, o.measurement_qual
   FROM obs o WHERE o.dataset_key = '{ds_key}'")))
```

### Declared NULLs

```{r}
#| label: validate
results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("validate_for_release:", ifelse(results$passed, "PASSED",
    "reported findings -- reconciled against the declared NULLs below"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")

n_null <- function(col) q1(glue(
  "SELECT COUNT(*) FROM sample WHERE dataset_key = '{ds_key}' AND {col} IS NULL"))
nullable <- tribble(
  ~table,   ~column,             ~n,                          ~reason,
  "sample", "parent_sample_key", n_null("parent_sample_key"), "single-level dataset: every sample is a root",
  "sample", "grid_key",          n_null("grid_key"),          "position outside the CalCOFI grid cells (none on the first render, 2026-10-06; pinned below)")
nullable |> datatable(caption = "Declared NULLs (count and reason)", rownames = FALSE)

stopifnot(
  "every sample is a root" = nullable$n[nullable$column == "parent_sample_key"] == nrow(d_sample),
  "no sample lacks a time, site_key, cruise_key or depth" =
    n_null("datetime") == 0 && n_null("site_key") == 0 &&
    n_null("cruise_key") == 0 && n_null("depth_max_m") == 0)
# ratchet: the first render (2026-10-06) put every sample in a grid cell
stopifnot("every sample falls in a CalCOFI grid cell (first render: 0 NULL)" =
            n_null("grid_key") == 0)
```

### Core ERD

```{r}
#| label: erd-core
core_tbls <- core_output_tables(con)
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls))
```

## Write Parquet Outputs

```{r}
#| label: write-parquet
dir_create(dir_parquet)
mismatches <- list(cruise_keys = collect_cruise_key_mismatches(con, "hplc_sample"))
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
parquet_stats <- write_parquet_outputs(
  con = con, output_dir = dir_parquet,
  tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE, mismatches = mismatches)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)
parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata
d_tbls_rd <- read_csv(file.path(dir_meta, "tbls_redefine.csv"))
d_flds_rd <- read_csv(file.path(dir_meta, "flds_redefine.csv"))
metadata_path <- build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           file.path(dir_meta, "metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned,
  sources = src_stamp)
```

## Upload to GCS

```{r}
#| label: upload-gcs
if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet,
              gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else cat("CALCOFI_SKIP_GCS set -- skipping the upload (local verification run)\n")
```

## Questions for Data Providers

Follow-up questions for CCE-LTER (Ralf Goericke, Mike Stukel), ranked. None is a
blocker: what is published is the source's own µg/L values with its own QC flag,
and the rows the questions are about are held out.

```{r}
#| label: provider-questions
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the cce-lter hplc-pigments-calcofi data providers (ranked)")
```

## Cleanup

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("parquet outputs written to: {dir_parquet}"), "\n")
```
