---
title: "Ingest Size-Fractionated Chlorophyll a (MLRG, 1994-1996)"
calcofi:
  target_name: ingest_cce_lter_chl_size_fractionated
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
    - ingest_calcofi_bottle
  output: data/parquet/cce-lter_chl-size-fractionated/manifest.json
  provider: cce-lter
  dataset: chl-size-fractionated
  workflow_url: https://calcofi.io/workflows/ingest_cce-lter_chl-size-fractionated.html
  questions_file: metadata/cce-lter/chl-size-fractionated/questions.csv
  # staged, not released: first ingest of this holding (CalCOFI/workflows#84).
  in_release: false
  dataset_meta:
    dataset_name: Size fractionation of total Chl a larger and smaller than 8 μm data
      generated by Mike Mullin of the Marine Life Research Group, aboard CalCOFI (California
      Cooperative Oceanic Fisheries Investigations) cruises of the coast of California,
      January 1994 - October 1996
    dataset_name_short: "Size-fractionated chl a"
    category: "Productivity & Pigments"
    color: "#5a8f29"
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (one station occupation per row, 0-100 m composite; time and position from its calcofi_bottle cast)"}
    - {table: obs, shared: true, note: "core observation headline (env realm; chl_a_int_100m + chl_a_gt8um_int_100m in mg/m2)"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#dcebc8"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: EDI package `knb-lter-cce.249.3` (Marine Life Research Group, Michael
Mullin), entity *SizeFractionatedChl* (`table_249.csv`): 422 rows, 8 columns, 12
quarterly CalCOFI cruises, 1994-01 to 1996-10, completed. Fetched by
`libs/download_chl_size_fractionated.R`, revision and md5 pinned.

- **Grain**: one row per station occupation (`studyName` x `Station ID`), each a
  value integrated over 0-100 m from the bottle cast: total chlorophyll a (ten
  GF/F samples, integrated algebraically) and chlorophyll a >= 8 um (one
  physically combined sample through an 8 um membrane filter).
- **No time, position or depth in the source.** Each row carries only its
  cruise and station, and the station's line is rounded to a whole number for
  lines 76.7, 81.8, 83.3 and 86.7 (`077.0 060.0` is station 76.7 60.0);
  `station_crosswalk.csv` maps all 48 source stations to official ones (Q03).
  Time and position come from the matching `calcofi_bottle` cast: same
  `cruise_key` and `site_key`
  (`match_by_site_datetime()`), and the cast's own `cruise_key` must equal the
  one minted from the source's `Cruise` designation and ship code (asserted).
- **Published**: `chl_a_int_100m` and `chl_a_gt8um_int_100m` (mg/m2), as `obs` on
  a `cast` sample spanning 0-100 m.
- **Not published** (kept in the archived source):
  - `Chl a < 8um`: total minus >= 8 um, per the EML (asserted to the rounding).
  - `Nitracline Depth`: a property of the cast's nutrient profile, which
    `calcofi_bottle` already holds.
  - `Total Zooplankton Displacement Volume`: from a net tow, not the bottle
    cast; whether it is the same measurement `swfsc_ichthyo` already publishes
    is Q02.

```{mermaid}
graph LR
  A[EDI knb-lter-cce.249.3<br/>422 rows] --> B[chl_size_station<br/>cruise x station]
  B -.cruise_key + site_key.-> C[(calcofi_bottle cast<br/>time, position)]
  B --> D[chl_size_measurement<br/>long: chl_a_int_100m, chl_a_gt8um_int_100m]
  B --> E[[core: sample, cast 0-100 m]]
  D --> F[[core: obs, realm=env]]
```

## Setup

```{r}
#| label: setup
devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))
cc             <- read_calcofi_meta(here("ingest_cce-lter_chl-size-fractionated.qmd"))
provider       <- cc$provider
dataset        <- cc$dataset
tables_owned   <- cc$tables_owned
ds_key         <- as.character(glue("{provider}_{dataset}"))
dir_label      <- ds_key
dir_parquet    <- here(glue("data/parquet/{dir_label}"))
dir_stage      <- cc_stage_path("parquet", dir_label, create = TRUE)
dir_src        <- cc_stage_path("source", dir_label, create = TRUE)
db_path        <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_meta       <- here(glue("metadata/{provider}/{dataset}"))
publish_to_gcs <- Sys.getenv("CALCOFI_SKIP_GCS") == ""

shard_path <- function(ds, tbl) {
  p <- cc_stage_path("parquet", ds, glue("{tbl}.parquet"))
  if (file_exists(p)) return(p)
  glue("https://storage.googleapis.com/calcofi-db/ingest/{ds}/{tbl}.parquet")
}

meas_map <- tribble(
  ~measurement_type,       ~value_col,
  "chl_a_int_100m",        "chl_a_total",
  "chl_a_gt8um_int_100m",  "chl_a_gt8um")
d_meas_type <- read_measurement_type(here("metadata/measurement_type.csv"))
stopifnot("both types are registered" =
            all(meas_map$measurement_type %in% d_meas_type$measurement_type))

if (overwrite) {
  if (file_exists(db_path)) file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp"))) dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
load_duckdb_extension(con, "httpfs")
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Read Source Data

```{r}
#| label: read-source
source(here("libs/download_chl_size_fractionated.R"))
downloaded <- overwrite_all || !file_exists(file.path(dir_src, "data.csv"))
src_csv <- download_chl_size_fractionated(dir_src, overwrite = overwrite_all)
src_stamp <- stamp_source_access(files = src_csv) |>
  mutate(source = "https://doi.org/10.6073/pasta/02183db3ad041992304d26757f1b25f9",
         method = if (downloaded) "download" else "file_mtime")

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_src, gcs_prefix = glue("archive/{provider}/{dataset}"),
              bucket = "calcofi-files-public", exclude = c(".DS_Store", "*.tmp"))
} else {
  cat("CALCOFI_SKIP_GCS set -- source NOT archived to gs://calcofi-files-public\n")
}

d_raw <- read_csv(src_csv, col_types = cols(.default = "c"), na = character(),
                  locale = locale(encoding = "UTF-8"))
hdr_expected <- c(
  "studyName", "Cruise", "Station ID", "Total Chl a (mg/m²)",
  "Chl a > 8µm (mg/m²)", "Chl a < 8µm (mg/m²)",
  "Total Zooplankton Displacement Volume (ml/1000m3)", "Nitracline Depth (m)")
hdr_new <- c("study_name", "cruise_ym", "site_key", "chl_a_total", "chl_a_gt8um",
             "chl_a_lt8um", "zooplankton_dv_ml_1000m3", "nitracline_depth_m")
stopifnot(
  "expected 422 rows x 8 columns (EDI revision 3, checked 2026-10-06)" =
    nrow(d_raw) == 422 && ncol(d_raw) == 8,
  "source header changed: re-check flds_redefine.csv before renaming by position" =
    identical(names(d_raw), hdr_expected))
names(d_raw) <- hdr_new
```

## Clean, Type-Cast

```{r}
#| label: clean-cast
num <- function(x) suppressWarnings(as.numeric(if_else(x == "", NA_character_, x)))

# the source writes lines 76.7 / 81.8 / 83.3 / 86.7 rounded to whole numbers
# (077.0, 082.0, 083.0, 087.0); metadata/.../station_crosswalk.csv maps each
# source Station ID to the official station (Q03)
d_xwalk <- read_csv(file.path(dir_meta, "station_crosswalk.csv"), col_types = cols(.default = "c"))
stopifnot("every source Station ID is in station_crosswalk.csv" =
            all(d_raw$site_key %in% d_xwalk$station_id_src))

d_station <- d_raw |>
  rename(station_id_src = site_key) |>
  left_join(d_xwalk |> select(station_id_src, site_key), by = "station_id_src") |>
  mutate(
    across(c(chl_a_total, chl_a_gt8um, chl_a_lt8um, zooplankton_dv_ml_1000m3,
             nitracline_depth_m), num),
    site_key   = normalize_site_key(site_key),
    ship_code  = str_extract(study_name, "[0-9A-Z]+$"),
    # the cruise key the source designates: YYYY-MM (its Cruise) + the ship's NODC code
    cruise_key_src = paste0(substr(cruise_ym, 1, 4), "-", substr(cruise_ym, 5, 6), "-", ship_code),
    sample_key_id  = paste(study_name, str_replace(site_key, " ", "_"), sep = "-")) |>
  arrange(study_name, site_key) |>
  mutate(station_id = row_number())

stopifnot(
  "every row has a station, a YYYYMM Cruise and both published values" =
    !anyNA(d_station$site_key) && !anyNA(d_station$chl_a_total) && !anyNA(d_station$chl_a_gt8um) &&
    all(str_detect(d_station$cruise_ym, "^\\d{4}(0[1-9]|1[0-2])$")),
  "one row per cruise x station" = !anyDuplicated(d_station$sample_key_id),
  "< 8 um = total - >= 8 um (not published, derivable)" =
    max(abs(d_station$chl_a_total - d_station$chl_a_gt8um - d_station$chl_a_lt8um)) < 1e-6,
  ">= 8 um never exceeds total (the EML caps it at 100%)" =
    all(d_station$chl_a_gt8um <= d_station$chl_a_total))
cat(glue("{nrow(d_station)} station occupations on {n_distinct(d_station$study_name)} cruises"), "\n")
dbWriteTable(con, "chl_size_station", d_station, overwrite = TRUE)
```

## Time, Position and Cruise from the Bottle Cast

The source names the cruise and station only. `match_by_site_datetime()` finds
the `calcofi_bottle` cast with the same `cruise_key` and `site_key` (key
`cruise_site`), taking the one nearest the 15th of the designated month if a
station was occupied twice. The cast gives the row its time, position and
`cruise_key`.

```{r}
#| label: match-cast
dbExecute(con, glue("
  CREATE OR REPLACE TABLE bottle_cast AS
  SELECT sample_key AS cast_sample_key, cruise_key, site_key, datetime, latitude, longitude,
         cruise_key || ' ' || site_key AS cruise_site
  FROM read_parquet('{shard_path('calcofi_bottle', 'sample')}')
  WHERE sample_type = 'cast' AND site_key IS NOT NULL AND cruise_key IS NOT NULL
    AND datetime IS NOT NULL
    AND cruise_key IN (SELECT DISTINCT cruise_key_src FROM chl_size_station)"))

dbExecute(con, "ALTER TABLE chl_size_station ADD COLUMN IF NOT EXISTS cruise_site VARCHAR")
dbExecute(con, "ALTER TABLE chl_size_station ADD COLUMN IF NOT EXISTS nominal_date TIMESTAMP")
dbExecute(con, "
  UPDATE chl_size_station SET
    cruise_site  = cruise_key_src || ' ' || site_key,
    nominal_date = make_timestamp(CAST(substr(cruise_ym, 1, 4) AS BIGINT),
                                  CAST(substr(cruise_ym, 5, 2) AS BIGINT), 15, 0, 0, 0)")

n_multi <- dbGetQuery(con, "
  SELECT COUNT(*) FROM (SELECT cruise_site FROM bottle_cast
    WHERE cruise_site IN (SELECT cruise_site FROM chl_size_station)
    GROUP BY 1 HAVING COUNT(*) > 1)")[[1]]
cat(glue("station occupations with more than one bottle cast on the cruise: {n_multi}"), "\n")

m_cast <- match_by_site_datetime(
  con, "chl_size_station", "bottle_cast",
  fk_col = "cast_sample_key", ref_pk = "cast_sample_key",
  key_col = "cruise_site", datetime_col = "nominal_date",
  ref_datetime_col = "datetime", window_days = 45)

dbExecute(con, "ALTER TABLE chl_size_station ADD COLUMN IF NOT EXISTS cruise_key VARCHAR")
dbExecute(con, "ALTER TABLE chl_size_station ADD COLUMN IF NOT EXISTS datetime_start_utc TIMESTAMP")
dbExecute(con, "ALTER TABLE chl_size_station ADD COLUMN IF NOT EXISTS latitude DOUBLE")
dbExecute(con, "ALTER TABLE chl_size_station ADD COLUMN IF NOT EXISTS longitude DOUBLE")
dbExecute(con, "
  UPDATE chl_size_station s SET cruise_key = c.cruise_key, datetime_start_utc = c.datetime,
         latitude = c.latitude, longitude = c.longitude
  FROM bottle_cast c WHERE c.cast_sample_key = s.cast_sample_key")

d_unmatched <- dbGetQuery(con, "
  SELECT study_name, cruise_key_src, site_key FROM chl_size_station
  WHERE cast_sample_key IS NULL ORDER BY 1, 3")
d_unmatched |> datatable(caption = "Station occupations with no calcofi_bottle cast (held out, Q01)", rownames = FALSE)
cat(glue("bottle cast matched: {m_cast$matched}/{m_cast$total}"), "\n")
stopifnot(
  "a matched cast's cruise_key is the one the source designates" =
    dbGetQuery(con, "SELECT COUNT(*) FROM chl_size_station
                     WHERE cast_sample_key IS NOT NULL AND cruise_key <> cruise_key_src")[[1]] == 0,
  "most station occupations find their cast" = m_cast$matched >= 0.9 * m_cast$total)

dbExecute(con, "CREATE OR REPLACE TABLE chl_size_pub AS
                SELECT * FROM chl_size_station WHERE cast_sample_key IS NOT NULL")
```

## Pivot Measurements

```{r}
#| label: pivot-long
d_meas <- dbGetQuery(con, "SELECT station_id, chl_a_total, chl_a_gt8um FROM chl_size_pub") |>
  pivot_longer(-station_id, names_to = "value_col", values_to = "measurement_value") |>
  inner_join(meas_map, by = "value_col") |>
  arrange(station_id, measurement_type) |>
  transmute(measurement_id = row_number(), station_id, measurement_type, measurement_value)
dbWriteTable(con, "chl_size_measurement", d_meas, overwrite = TRUE)
d_meas |> count(measurement_type) |> datatable(caption = "Values per measurement_type", rownames = FALSE)
```

## Add Spatial

```{r}
#| label: spatial
load_prior_tables(
  con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
  gcs_prefix = "ingest/swfsc_ichthyo",
  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
add_point_geom(con, "chl_size_pub", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "chl_size_pub") |> datatable(caption = "Grid assignment")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata
this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot(nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)
```

## Schema Documentation

```{r}
#| label: schema
rels <- list(
  primary_keys = list(chl_size_station = "station_id", chl_size_measurement = "measurement_id",
                      measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "chl_size_measurement", column = "station_id",
         ref_table = "chl_size_station", ref_column = "station_id"),
    list(table = "chl_size_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))
cc_erd(con, tables = c("chl_size_station", "chl_size_measurement", "measurement_type", "dataset"),
       rels = rels,
       colors = list(lightblue = c("chl_size_station", "chl_size_measurement"),
                     lightyellow = "measurement_type", white = "dataset"))
```

| source | core | transform |
|---|---|---|
| one row per `studyName` x `Station ID` | `sample` (`cast`) | key `cce-lter_chl-size-fractionated:cast:{studyName}-{LLL.L_SSS.S}`; depth 0-100 m; time, position and `cruise_key` from the matched `calcofi_bottle` cast |
| `Total Chl a` | `obs` `chl_a_int_100m` | verbatim, mg/m2 |
| `Chl a > 8µm` | `obs` `chl_a_gt8um_int_100m` | verbatim, mg/m2 |
| `Chl a < 8µm` | -- | **not published**: total minus >= 8 um (asserted) |
| `Nitracline Depth` | -- | **not published**: derivable from `calcofi_bottle` nutrients |
| `Total Zooplankton Displacement Volume` | -- | **not published** (Q02) |
| rows with no bottle cast | -- | held out (Q01) |

## Check Declared Bounds

Both types carry `valid_min = 0` only. No upper bound is declared.

```{r}
#| label: bounds-check
bounds_pre <- check_measurement_bounds(con, "chl_size_measurement", mt = d_meas_type)
bounds_datatable(bounds_pre)
bounds_post <- check_measurement_bounds(
  con, "chl_size_measurement", mt = d_meas_type, include_undeclared = FALSE)
stopifnot("no value outside a declared bound" = sum(bounds_post$status == "out_of_range") == 0)
```

## Enforce Column Types

```{r}
#| label: enforce-types
enforce_column_types(con)
```

## Emit Core Tables

```{r}
#| label: emit_core
append_sample(con, sample_arm_self(
  ds_key, "chl_size_pub", "sample_key_id", "cast",
  dt_col = "datetime_start_utc", site_expr = "site_key",
  depth_min = "0::DOUBLE", depth_max = "100::DOUBLE"))

append_obs(con, glue("
  SELECT 'env', '{ds_key}', {ns_key(ds_key, 'cast', 's.sample_key_id')},
         s.grid_key, s.cruise_key, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP), 0::DOUBLE, 100::DOUBLE,
         NULL::VARCHAR, NULL::VARCHAR, m.measurement_type, m.measurement_value,
         NULL::VARCHAR, NULL::DOUBLE
  FROM chl_size_measurement m JOIN chl_size_pub s USING (station_id)"))

q1 <- function(sql) dbGetQuery(con, sql)[[1]]
core <- list(sample = q1("SELECT COUNT(*) FROM sample"), obs = q1("SELECT COUNT(*) FROM obs"))
cat(glue("core projection -- sample={core$sample} obs={core$obs}"), "\n")
stopifnot(
  "one sample per published station occupation" = core$sample == q1("SELECT COUNT(*) FROM chl_size_pub"),
  "two obs per sample" = core$obs == 2 * core$sample,
  "obs value total equals the source" =
    isTRUE(all.equal(q1("SELECT SUM(measurement_value) FROM obs"),
                     q1("SELECT SUM(measurement_value) FROM chl_size_measurement"))),
  "sample_key is unique" = q1("SELECT COUNT(*) - COUNT(DISTINCT sample_key) FROM sample") == 0,
  "every obs.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key) WHERE s.sample_key IS NULL") == 0)
```

### Declared NULLs

```{r}
#| label: validate
results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("validate_for_release:", ifelse(results$passed, "PASSED",
    "reported findings -- reconciled against the declared NULLs below"), "\n")
n_null <- function(col) q1(glue("SELECT COUNT(*) FROM sample WHERE {col} IS NULL"))
nullable <- tribble(
  ~table,   ~column,             ~n,                          ~reason,
  "sample", "parent_sample_key", n_null("parent_sample_key"), "single-level dataset: the matched bottle cast is the source of time and position, not a parent",
  "sample", "grid_key",          n_null("grid_key"),          "position outside the CalCOFI grid cells")
nullable |> datatable(caption = "Declared NULLs (count and reason)", rownames = FALSE)
stopifnot(
  "no sample lacks a time, position, site_key or cruise_key" =
    n_null("datetime") == 0 && n_null("latitude") == 0 && n_null("site_key") == 0 &&
    n_null("cruise_key") == 0)
```

### Core ERD

```{r}
#| label: erd-core
core_tbls <- core_output_tables(con)
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls))
```

## Write Parquet Outputs

```{r}
#| label: write-parquet
dir_create(dir_parquet)
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
parquet_stats <- write_parquet_outputs(
  con = con, output_dir = dir_parquet, tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")), strip_provenance = FALSE)
build_relationships_json(rels = core_relationships(tbls_out), output_dir = dir_parquet,
                         provider = provider, dataset = dataset)
parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata
d_tbls_rd <- read_csv(file.path(dir_meta, "tbls_redefine.csv"))
d_flds_rd <- read_csv(file.path(dir_meta, "flds_redefine.csv"))
metadata_path <- build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           file.path(dir_meta, "metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned, sources = src_stamp)
```

## Upload to GCS

```{r}
#| label: upload-gcs
if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet,
              gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else cat("CALCOFI_SKIP_GCS set -- skipping the upload (local verification run)\n")
```

## Questions for Data Providers

The dataset is completed (1994-1996); questions go to the CCE-LTER Information
Manager, the package's contact.

```{r}
#| label: provider-questions
questions_datatable(here(cc$questions_file),
                    caption = "Questions for the cce-lter chl-size-fractionated providers (ranked)")
```

## Cleanup

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("parquet outputs written to: {dir_parquet}"), "\n")
```
