---
title: "Ingest SIO Cetacean Sightings (CalCOFI visual surveys)"
calcofi:
  target_name: ingest_sio_cetacean_sightings
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/sio_cetacean-sightings/manifest.json
  provider: sio
  dataset: cetacean-sightings
  workflow_url: https://calcofi.io/workflows/ingest_sio_cetacean-sightings.html
  questions_file: metadata/sio/cetacean-sightings/questions.csv
  # HELD (Ben, 2026-10-08): staged, never released or uploaded, until the
  # Scripps Whale Acoustics Lab reviews and signs off (Simone Baumann-Pickering,
  # 2026-10-07). The source is Michaela Alksne's QA/QC'd on-effort product, the
  # basis of a density-estimation manuscript still pending; the earlier
  # marmam-app copy was exploratory and must not be used. `in_release: false`
  # keeps it out of every release; the notebook also skips every GCS upload
  # while it is set (publish_to_gcs below).
  in_release: false
  dataset_meta:
    dataset_name: CalCOFI Cetacean Visual Line-Transect Sightings
    dataset_name_short: "Cetacean sightings"
    category: "Seabirds & Marine Mammals"
    color: "#5b7fa6"
  # publishes the consolidated core: the segment / sighting / species source shape
  # is wrangled here and projected into sample / obs / obs_attribute /
  # sample_measurement + the taxa refs.
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (effort segment + sighting)"}
    - {table: obs, shared: true, note: "core occurrence headline (group_size per counted sighting)"}
    - {table: obs_attribute, shared: true, note: "transect status, group size min/max, calves per counted sighting (behaviour held, Q03)"}
    - {table: sample_measurement, shared: true, note: "transect_length per on-transect effort segment"}
    - {table: taxon, shared: true, note: "shared taxa reference (cetaceans, worms:)"}
    - {table: dataset_taxon, shared: true, note: "species_code -> taxon_key crosswalk"}
    - {table: taxon_group, shared: true, note: "calcofi:marine_mammals"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#d6e2f0"
editor_options:
  chunk_output_type: console
---

## Overview

> **Held, not released.** This dataset is staged only (`in_release: false`) and
> nothing from it is uploaded, until the Scripps Whale Acoustics Lab reviews and
> signs off (Simone Baumann-Pickering, 2026-10-07).

**Source**: the QA/QC'd on-effort visual-survey product of the Scripps Whale
Acoustics Lab (Hildebrand / Baumann-Pickering), worked up by Michaela Alksne for
a density-estimation manuscript (pending), shared by Simone Baumann-Pickering on
2026-10-07 and saved to the CalCOFI Data Folder
(`data-private/sio/cetacean-sightings/`, held while the lab reviews it):

- `CalCOFI_sightings_on_effort.xlsx` — one row per on-effort sighting, with the
  sighting conditions (Beaufort, swell, visibility, glare, cloud), the observer,
  the reticle / distance measurements and the computed perpendicular distance.
- `CalCOFI_segments_on_effort.xlsx` — the on-effort survey segments: start / end
  position and time, the on/off-transect flag and the segment length.
- `CalCOFI_sighting_codes.csv` — the lab's species-code list (Latin-1, with
  non-breaking spaces).

The row counts are asserted in [Read Source Data]. The two workbooks are
**on-effort only**: every sighting carries a segment id, and the segment ids are
unique across cruises, so a sighting joins its segment on the id alone.

They replace the CSVs in the CalCOFI marine mammal Shiny app repository
(CalCOFI/marmam-app) this ingest read until 2026-10-07: the lab calls the app's
copy exploratory, and it is no longer read. The marmam-app copy also carried
off-effort sightings, which this product does not; it adds 2022-2025 effort and
2023-2025 sightings the app never had.

**Behaviour is held back.** The two behaviour columns are bare numeric codes
(`1`, `34`, `2, 5, 8`) with no codebook, and the same `behavior` measurement type
carries words (Flying, Feeding) for `farallon_bird-mammal`, so a consumer
grouping by `behavior` would get codes beside labels. The code path stays below,
gated by `publish_behavior <- FALSE`; it is switched on when the lab supplies
the codebook (Q03; Ben, 2026-10-04).

- **Provider**: `sio` (Scripps Whale Acoustics Lab, Hildebrand / Baumann-Pickering)
- **Grain**: a sighting inside an on-effort segment (`segment_id`)
- **Taxonomy**: species codes declared with `append_dataset_taxon()` from
  `metadata/sio/cetacean-sightings/species_codes.csv` (reconciled against the
  lab's code list, whose common names it carries) and resolved against WoRMS;
  the unidentified classes and the two synonyms are keyed by
  `metadata/taxon_override.csv`.

```{mermaid}
graph LR
  E[Segments workbook] --> SG[cetacean_segment<br/>on-effort segment]
  S[Sightings workbook] --> SI[cetacean_sighting<br/>one row per sighting]
  C[lab code list +<br/>species_codes.csv] --> SP[cetacean_species]
  SI -.segment_id.-> SG
  SI -.species_code.-> SP
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr,
  lubridate, purrr, readr, readxl, sf, stringr, tibble, tidyr, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))
source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_sio_cetacean-sightings.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
tables_owned <- cc$tables_owned
ds_key       <- as.character(glue("{provider}_{dataset}"))
dir_label    <- ds_key
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
dir_stage    <- cc_stage_path("parquet", dir_label, create = TRUE)
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_meta     <- here(glue("metadata/{provider}/{dataset}"))
# the lab's files sit in data-private, beside data-public in the CalCOFI Data
# Folder, while the dataset is held: a sync of data-public to GCS publishes
# whatever it holds (Ben, 2026-10-08). CALCOFI_DATA_PRIVATE_DIR overrides
dir_data_private <- Sys.getenv(
  "CALCOFI_DATA_PRIVATE_DIR", file.path(dirname(path_expand(dir_data)), "data-private"))
dir_src      <- path_expand(glue("{dir_data_private}/{provider}/{dataset}"))
# GCS uploads (source archive + parquet mirror) both target world-readable
# buckets, so they run only for a released dataset: `in_release: false` (held
# until the lab signs off) keeps everything local. CALCOFI_SKIP_GCS=1 also skips
# them for a review render of a released dataset.
in_release     <- isTRUE(cc$in_release %||% TRUE)
publish_to_gcs <- in_release && Sys.getenv("CALCOFI_SKIP_GCS") == ""

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

```{r}
#| label: read-source

stopifnot("source dir not found" = dir_exists(dir_src))
src_files <- c(
  sightings = file.path(dir_src, "CalCOFI_sightings_on_effort.xlsx"),
  segments  = file.path(dir_src, "CalCOFI_segments_on_effort.xlsx"),
  species   = file.path(dir_src, "CalCOFI_sighting_codes.csv"))
stopifnot("a source file is missing" = all(file_exists(src_files)))
# when the bytes were saved (file mtime); recorded by their place in the Data
# Folder, not by this machine's path to it
src_stamp <- stamp_source_access(files = unname(src_files)) |>
  mutate(source = as.character(glue("data-private/{provider}/{dataset}/{basename(src_files)}")))

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_src, gcs_prefix = glue("archive/{provider}/{dataset}"),
              bucket = "calcofi-files-public", exclude = c(".DS_Store", "*.tmp"))
} else {
  cat("held (in_release: false) or CALCOFI_SKIP_GCS set -- sources NOT archived to gs://calcofi-files-public\n")
}

# every cell as text, so nothing is coerced before it is looked at; the
# workbooks write the literal "NA" for a blank in several columns
xlsx_na  <- c("", "NA")
d_si_raw <- read_excel(src_files[["sightings"]], sheet = "Sheet1", col_types = "text", na = xlsx_na)
d_sg_raw <- read_excel(src_files[["segments"]],  sheet = "Sheet1", col_types = "text", na = xlsx_na)
# Latin-1, CR CR LF line ends, and a non-breaking space (0xA0) around one name
d_sp_raw <- read_csv(src_files[["species"]], col_types = cols(.default = "c"),
                     locale = locale(encoding = "latin1")) |>
  setNames(c("lab_code", "lab_common_name")) |>
  mutate(across(everything(), \(x) str_squish(str_replace_all(x, " ", " "))))

# the counts profiled on 2026-10-08; a changed file must be looked at, not absorbed
stopifnot(
  "sightings row count changed" = nrow(d_si_raw) == 4961L && ncol(d_si_raw) == 53L,
  "segments row count changed"  = nrow(d_sg_raw) == 3165L && ncol(d_sg_raw) == 12L,
  "species code count changed"  = nrow(d_sp_raw) == 34L)
cat(glue("sightings {nrow(d_si_raw)}, segments {nrow(d_sg_raw)}, lab species codes {nrow(d_sp_raw)}"), "\n")
```

## Build Species Vocabulary

```{r}
#| label: species

# species_codes.csv is the vocabulary: its `species_code` is the key the
# taxon_override.csv rows and ingest_sio_cetacean-edna.qmd use (mixed case, as
# the first delivery wrote it); `lab_code` is the code in the lab's list. The
# 2026-10 sightings write the codes upper-case, so a sighting is matched to the
# vocabulary case-insensitively (UBW, which the lab's list spells UWB, included).
# The lab's list is authoritative for the common name.
d_species_meta <- read_csv(file.path(dir_meta, "species_codes.csv"), col_types = cols(.default = "c"))
d_species <- d_species_meta |>
  select(species_code, lab_code, suborder, scientific_name, common_name_meta = common_name) |>
  left_join(d_sp_raw, by = "lab_code") |>
  mutate(common_name = coalesce(lab_common_name, common_name_meta)) |>
  select(species_code, lab_code, common_name, suborder, scientific_name)
stopifnot(
  "species_code must be unique, case-insensitively" = !anyDuplicated(toupper(d_species$species_code)),
  "every code in the lab's list is in species_codes.csv" =
    setequal(na.omit(d_species$lab_code), d_sp_raw$lab_code),
  "species_codes.csv carries the lab's common names" =
    with(d_species_meta, all(is.na(lab_code) | lab_common_name ==
                               d_sp_raw$lab_common_name[match(lab_code, d_sp_raw$lab_code)])))
dbWriteTable(con, "cetacean_species", d_species, overwrite = TRUE)
cat(glue("cetacean_species: {nrow(d_species)} codes ({sum(!is.na(d_species$lab_code))} in the lab's list), ",
         "{sum(is.na(d_species$scientific_name))} without a taxon"), "\n")
```

## Build Effort Segment Table

```{r}
#| label: segment

# Excel stores a date-time as days since 1899-12-30; the workbooks hold local
# Pacific clock time (Q04), read here as a clock reading (tagged UTC, not yet
# converted)
xl_clock <- function(x) as.POSIXct(round(as.numeric(x) * 86400), origin = "1899-12-30", tz = "UTC")
# clock time -> UTC (Q04). EDI knb-lter-cce.262.2 records each 2005-2015
# cruise's own GMT offset: most kept Pacific standard time (UTC-8) all year, even
# in summer; a few used PDT (UTC-7). Those cruises take their recorded offset
# (cruise_utc_offset.csv, keyed by this file's cruise label); the rest are read
# as America/Los_Angeles, daylight saving applied.
d_utc_offset <- read_csv(file.path(dir_meta, "cruise_utc_offset.csv"),
                         col_types = cols(utc_offset_h = "i", .default = "c"))
stopifnot("an offset is 7 or 8 h" = all(d_utc_offset$utc_offset_h %in% 7:8),
          "one offset per label"  = !anyDuplicated(d_utc_offset$cruise_label))
local_to_utc <- function(clock, label) {
  off <- d_utc_offset$utc_offset_h[match(label, d_utc_offset$cruise_label)]
  la  <- with_tz(force_tz(clock, "America/Los_Angeles"), "UTC")
  if_else(is.na(off), la, clock + hours(off))
}
num <- function(x) suppressWarnings(as.numeric(x))
fmt_clock <- function(x) format(x, "%Y-%m-%d %H:%M:%S")

d_segment <- d_sg_raw |>
  mutate(start_clock = xl_clock(StartDateTimeLocal), end_clock = xl_clock(EndDateTimeLocal)) |>
  transmute(
    segment_id           = as.integer(num(SegmentID)),
    # the effort file's cruise label (CCYYYY-MM) is the canonical one: the
    # sightings take it from their segment
    cruise_label         = Cruise,
    season               = str_to_lower(str_trim(Season)),
    year                 = as.integer(Year),
    datetime_start_local = fmt_clock(start_clock),
    datetime_end_local   = fmt_clock(end_clock),
    datetime_start_utc   = local_to_utc(start_clock, cruise_label),
    datetime_end_utc     = local_to_utc(end_clock, cruise_label),
    latitude_start       = num(StartLat),  longitude_start = num(StartLon),
    latitude_end         = num(EndLat),    longitude_end   = num(EndLon),
    # the core carries one point per event: the segment midpoint
    latitude             = (latitude_start  + latitude_end)  / 2,
    longitude            = (longitude_start + longitude_end) / 2,
    # ON = on effort and on transect; OFF = on effort, off transect
    transect_flag        = str_to_upper(str_trim(TransectFlag)),
    segment_length_km    = num(SegmentLength_km))

# a segment whose start is more than two months from its cruise label's month:
# segment 2886 is labelled CC2022-04 and dated 2022-11-05 (the first day of
# CC2022-11). Kept, but given no cruise_key below (it would take April's).
label_month <- ymd(paste0(str_sub(d_segment$cruise_label, 3, 9), "-01"))
d_segment <- d_segment |>
  mutate(misdated = abs(as.numeric(difftime(as.Date(datetime_start_local), label_month, units = "days"))) > 62)

stopifnot(
  "segment_id must be unique"                = !anyDuplicated(d_segment$segment_id),
  "every segment needs a start and end time" = !anyNA(d_segment$datetime_start_utc) && !anyNA(d_segment$datetime_end_utc),
  "no segment ends before it starts"         = all(d_segment$datetime_end_utc >= d_segment$datetime_start_utc),
  "every segment needs a midpoint"           = !anyNA(d_segment$latitude) && !anyNA(d_segment$longitude),
  "cruise_label must be CCYYYY-MM"           = all(str_detect(d_segment$cruise_label, "^CC\\d{4}-\\d{2}$")),
  "Year is the start's year"                 = all(d_segment$year == year(ymd_hms(d_segment$datetime_start_local))),
  "transect_flag is ON or OFF"               = all(d_segment$transect_flag %in% c("ON", "OFF")),
  "every segment has a length"               = !anyNA(d_segment$segment_length_km),
  "only segment 2886 is dated away from its label (CC2022-04 on 2022-11-05)" =
    identical(d_segment$segment_id[d_segment$misdated], 2886L))
dbWriteTable(con, "cetacean_segment", d_segment, overwrite = TRUE)
cat(glue("cetacean_segment: {nrow(d_segment)} segments over ",
         "{n_distinct(d_segment$cruise_label)} cruises, {min(d_segment$year)}-{max(d_segment$year)}; ",
         "{sum(d_segment$transect_flag == 'ON')} on transect, {sum(d_segment$transect_flag == 'OFF')} off; ",
         "{round(sum(d_segment$segment_length_km))} km"), "\n")
```

## Build Sighting Table

```{r}
#| label: sighting

d_sighting <- d_si_raw |>
  mutate(sighting_id = row_number(),
         clock       = xl_clock(Date_Time_Local_),
         segment_id  = as.integer(num(segID_ON_Effort_MNA))) |>
  left_join(d_segment |> select(segment_id, segment_label = cruise_label,
                                effort_status = transect_flag), by = "segment_id") |>
  transmute(
    sighting_id,
    # the sighting's own Cruise disagrees with its segment's on four cruises
    # (202504SH / 202507SR / 202511SR for CC2025-03 / -07 / -11, and CC2007-04
    # for CC2007-03): the segment's label wins, the sighting's is kept
    cruise_label          = coalesce(segment_label, Cruise),
    cruise_label_source   = Cruise,
    unique_id             = str_squish(UniqueID),
    # the survey code the UniqueID starts with (CC0407, CC2204SH, CC202504SH)
    # and its CCYYMM form, the key of cruise_label_crosswalk.csv
    survey_code           = str_extract(unique_id, "^[^_ ]+"),
    survey_code6          = paste0("CC", str_sub(str_extract(survey_code, "\\d+"), -4)),
    sighting_no           = SightingNo,
    season                = str_to_lower(str_trim(Season)),
    datetime_local        = fmt_clock(clock),
    datetime_start_utc    = local_to_utc(clock, cruise_label),
    date_local_check      = as.Date(xl_clock(Date_Local_)),
    latitude              = num(DecLat),
    longitude             = num(DecLong),
    segment_id,
    observer_port         = PortObs,
    observer_starboard    = StbdObs,
    observer_initial      = InitObserver,
    sighting_quality      = Quality,
    visibility            = num(Visibility),
    precipitation         = Precipitation,
    cloud_pct             = num(Cloud_),
    glare_bearing         = GlareBearing,
    glare_bearing_1       = GlareBearing_1,
    glare_quality         = GlareQuality_SL_M_SE_,
    wind_dir_true         = num(WindDir_T_),
    wind_speed            = num(WindSpd_T_),
    beaufort              = num(Bft),
    swell_ft              = num(Swell_ft_),
    cue                   = Cue,
    ship_heading_true     = num(ShipHeadingTru),
    sighting_bearing_true = num(SightingBearingTru),
    bino_reticle          = num(BinoReticle),
    distance_m            = num(Distance_m_),
    obs_height_m          = num(ObsHeight_m_),
    port_or_starboard     = PortOrStarbord,
    angle_formula         = num(AngleFormula),
    dist_calc_m           = num(dist_calc_m_),
    perp_dist_m           = num(perp_dist_m),
    group_size_best       = num(Best),
    group_size_min        = num(Min),
    group_size_max        = num(Max),
    calf_raw              = Calf,
    # plain integers only ("001" included); ND / nd / 0NA / UD / O / DN / N /
    # 1? / Juv stay NULL (Q09)
    calf_count            = if_else(str_detect(str_trim(Calf), "^\\d+$"),
                                    suppressWarnings(as.integer(str_trim(Calf))), NA_integer_),
    species_code_source   = str_trim(Species1),
    # case-insensitive: the vocabulary's DC is the source's DC, its Ubw the source's UBW
    species_code          = d_species$species_code[match(toupper(species_code_source),
                                                         toupper(d_species$species_code))],
    species2_raw          = Species2,
    sp1_percent           = num(sp1Percent),
    sp2_percent           = num(sp2Percent),
    effort_flag_original  = OriginalEffortFlag_ON_OFF_,
    transect_flag_original = OriginalTransect_ON_OFF_,
    # the segment's on/off-transect flag (every sighting here is on effort)
    effort_status,
    behavior_primary      = PrimaryBehaviour,
    behavior_other1       = OtherBehaviour1,
    comments              = Comments,
    time_diff_min         = num(Time_Diff_minutes))
n_seg_orphan <- sum(!d_sighting$segment_id %in% d_segment$segment_id)
d_lbl_src <- d_sighting |> distinct(cruise_label_source, cruise_label)
d_label   <- d_sighting |> distinct(cruise_label, survey_code6)
d_in_seg  <- d_sighting |>
  inner_join(d_segment |> select(segment_id, datetime_start_utc_seg = datetime_start_utc,
                                 datetime_end_utc_seg = datetime_end_utc), by = "segment_id")
stopifnot(
  "every sighting has a time"                       = !anyNA(d_sighting$datetime_start_utc),
  "Date_Local_ is the day of Date_Time_Local_ (blank on 7 CC0911 / CC1008 rows)" =
    all(is.na(d_sighting$date_local_check) | as.Date(ymd_hms(d_sighting$datetime_local)) == d_sighting$date_local_check) &&
    sum(is.na(d_sighting$date_local_check)) == 7L,
  "every sighting has a position"                   = !anyNA(d_sighting$latitude) && !anyNA(d_sighting$longitude),
  "every position is off California (west of -117)" = all(d_sighting$longitude < -117),
  "every sighting carries a segment id (on effort only)" = !anyNA(d_sighting$segment_id),
  "only the 2 sightings on segment 419 (CC2006-07) lack their segment" =
    n_seg_orphan == 2L && all(d_sighting$segment_id[!d_sighting$segment_id %in% d_segment$segment_id] == 419L),
  "a sighting falls inside its segment's time span (+- 1 min)" =
    all(d_in_seg$datetime_start_utc >= d_in_seg$datetime_start_utc_seg - 60 &
        d_in_seg$datetime_start_utc <= d_in_seg$datetime_end_utc_seg + 60),
  "each source cruise label maps to one effort label" = !anyDuplicated(d_lbl_src$cruise_label_source),
  "each effort label has one survey code"           = !anyDuplicated(d_label$cruise_label),
  "every effort label has sightings"                = all(d_segment$cruise_label %in% d_label$cruise_label),
  "every species code is in the vocabulary"         = !anyNA(d_sighting$species_code),
  "Species2 is always blank"                        = all(is.na(d_sighting$species2_raw)),
  "every sighting has a Best group size"            = !anyNA(d_sighting$group_size_best),
  "effort_status is ON or OFF wherever the segment exists" =
    all(d_sighting$effort_status %in% c("ON", "OFF") | !d_sighting$segment_id %in% d_segment$segment_id))
d_sighting <- d_sighting |> select(-date_local_check)
dbWriteTable(con, "cetacean_sighting", d_sighting, overwrite = TRUE)

n_dup_uid <- sum(duplicated(d_sighting$unique_id) | duplicated(d_sighting$unique_id, fromLast = TRUE))
cat(glue("cetacean_sighting: {nrow(d_sighting)} rows over {n_distinct(d_sighting$cruise_label)} cruises, ",
         "{min(year(d_sighting$datetime_start_utc))}-{max(year(d_sighting$datetime_start_utc))}; ",
         "{n_seg_orphan} on a segment_id not in the segments file; ",
         "{n_dup_uid} rows share a UniqueID; {sum(!is.na(d_sighting$calf_count))} with a calf count; ",
         "{sum(!is.na(d_sighting$perp_dist_m))} with a perpendicular distance"), "\n")
d_lbl_src |> filter(cruise_label_source != cruise_label) |>
  datatable(caption = "Sighting cruise labels the segments spell differently", rownames = FALSE)
```

Two sightings (CC2006-07) reference segment 419, which the segments file lacks.
They keep their `segment_id` in the source table but get no parent `sample`, so
the core never points at a sample that does not exist.

## Resolve cruise_key + Spatial

The workbooks carry a **survey label** (`CC2004-07`), not a cruise, and no
ship. As for `farallon_bird-mammal`, `match_cruise_by_track()` matches each
sighting to the nearest CalCOFI station occupied within a day (the
`swfsc_ichthyo` sample track, whose `cruise_key`s are exactly the `cruise`
reference's) and takes the **majority vote per label**. The effort segments (the
ship's own track) vote the same way, and the two votes must agree wherever both
exist.

The surveys on cruises `swfsc_ichthyo` has no tows for (Sally Ride, Oceanus,
Bold Horizon, and every cruise after 2023-01) take their cruise from
`cruise_label_crosswalk.csv`, keyed by the survey code (`CC1611`): the one
cruise the release's CTD casts (and bottles) place on the same dates (Q14),
checked here against the ship reference. That file is shared with
`ingest_sio_cetacean-edna.qmd`, which re-checks that each key is its label's
year-month plus the ship's NODC code, so two surveys whose designated cruise
month differs from the survey code's are left unresolved (CC2023-03 on
2023-04-3322 and CC2024-07 on 2024-08-33P4, by the CTD casts). The chunk below
asserts that the crosswalk is used only where the ichthyo track has no
occupation at all during a survey's dates, never to overrule a track match.

```{r}
#| label: spatial-cruise

dbExecute(con, glue(
  "CREATE OR REPLACE VIEW cruise_track AS
   SELECT DISTINCT cruise_key, datetime, latitude, longitude
   FROM read_parquet('{cc_stage_path('parquet','swfsc_ichthyo','sample.parquet')}')
   WHERE cruise_key IS NOT NULL AND datetime IS NOT NULL"))

# window_days = 1: the sighting time is converted from Pacific clock time, so an
# evening sighting falls on the next UTC day
cruise_match <- match_cruise_by_track(
  con, "cetacean_sighting", "cruise_track",
  datetime_col = "datetime_start_utc", lon_col = "longitude", lat_col = "latitude",
  ref_datetime_col = "datetime", group_col = "cruise_label", window_days = 1)

# the effort segments vote on the same labels from the ship's own positions (the
# misdated segment 2886 is one vote of 25 on CC2022-04; no ichthyo station was
# occupied within a day of it, so it casts none)
seg_match <- match_cruise_by_track(
  con, "cetacean_segment", "cruise_track",
  datetime_col = "datetime_start_utc", lon_col = "longitude", lat_col = "latitude",
  ref_datetime_col = "datetime", group_col = "cruise_label", window_days = 1)

# surveys on cruises with no ichthyo tows: the cruise the release's CTD casts and
# bottles place on the same dates (Q14); each key must be the survey code's
# year-month plus its ship's NODC code, so a typo in the crosswalk cannot slip
# through (the same rule ingest_sio_cetacean-edna.qmd applies to this file)
d_ship  <- dbGetQuery(con, glue(
  "SELECT ship_key, ship_nodc FROM read_parquet('{cc_stage_path('parquet','swfsc_ichthyo','ship.parquet')}')"))
d_xwalk <- read_csv(file.path(dir_meta, "cruise_label_crosswalk.csv"), col_types = cols(.default = "c")) |>
  rename(survey_code6 = cruise_label) |>
  left_join(d_ship, by = "ship_key") |>
  mutate(minted = paste0("20", substr(survey_code6, 3, 4), "-", substr(survey_code6, 5, 6), "-", ship_nodc))
stopifnot(
  "every crosswalk ship is in the ship reference" = !anyNA(d_xwalk$ship_nodc),
  "each crosswalk cruise_key is the survey code's year-month + its ship's NODC code" =
    all(d_xwalk$minted == d_xwalk$cruise_key))

d_labels <- dbGetQuery(con, "
  SELECT cruise_label FROM cetacean_sighting UNION SELECT cruise_label FROM cetacean_segment")
grp <- d_labels |>
  left_join(d_label, by = "cruise_label") |>
  left_join(cruise_match$groups |> select(cruise_label, sighting_key = cruise_key,
                                          sighting_votes = votes, sighting_share = share),
            by = "cruise_label") |>
  left_join(seg_match$groups |> select(cruise_label, segment_key = cruise_key,
                                       segment_votes = votes, segment_share = share),
            by = "cruise_label") |>
  left_join(d_xwalk |> select(survey_code6, crosswalk_key = cruise_key), by = "survey_code6") |>
  mutate(
    cruise_key = coalesce(sighting_key, segment_key, crosswalk_key),
    cruise_key_method = case_when(
      !is.na(sighting_key)  ~ "track: sightings",
      !is.na(segment_key)   ~ "track: effort",
      !is.na(crosswalk_key) ~ "crosswalk: release CTD/bottle (Q14)")) |>
  arrange(cruise_label)
stopifnot(
  "the sighting and segment votes agree wherever both resolve a label" =
    with(grp, all(is.na(sighting_key) | is.na(segment_key) | sighting_key == segment_key)),
  "the crosswalk only fills labels the track cannot resolve" =
    with(grp, all(is.na(crosswalk_key) | (is.na(sighting_key) & is.na(segment_key)))))

# each fills the other's gaps; the misdated segment keeps a NULL cruise_key
dbWriteTable(con, "_label_ck", grp |> select(cruise_label, cruise_key), overwrite = TRUE)
for (tbl in c("cetacean_sighting", "cetacean_segment"))
  dbExecute(con, glue("
    UPDATE {tbl} t SET cruise_key = l.cruise_key FROM _label_ck l
    WHERE t.cruise_key IS NULL AND l.cruise_key IS NOT NULL AND l.cruise_label = t.cruise_label"))
dbExecute(con, "UPDATE cetacean_segment SET cruise_key = NULL WHERE misdated")
dbExecute(con, "DROP TABLE _label_ck")

# spatial last: nothing above has to UPDATE a table that carries geometry
load_prior_tables(con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
                  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
for (tbl in c("cetacean_segment", "cetacean_sighting")) {
  add_point_geom(con, tbl, lon_col = "longitude", lat_col = "latitude")
  assign_grid_key(con, tbl)
}

n_sight_ck <- dbGetQuery(con, "SELECT COUNT(cruise_key) FROM cetacean_sighting")[[1]]
cat(glue("cruise_key resolved on {n_sight_ck}/{cruise_match$total} sightings ",
         "({round(100 * n_sight_ck / cruise_match$total, 1)}%; ",
         "{cruise_match$matched} by their own label vote)"), "\n")
grp |>
  datatable(caption = "Cruise label -> cruise_key: sighting vote, effort-segment vote, crosswalk (NULL = unresolved)",
            rownames = FALSE, filter = "top") |>
  formatPercentage(c("sighting_share", "segment_share"), 1)
```

```{r}
#| label: cruise-match-validate

dupe <- grp$cruise_key[!is.na(grp$cruise_key)]
dupe <- dupe[duplicated(dupe)]
n_seg_null <- dbGetQuery(con, "SELECT COUNT(*) FROM cetacean_segment WHERE cruise_key IS NULL")[[1]]
unresolved_labels <- grp$cruise_label[is.na(grp$cruise_key)]
xwalk_labels      <- grp$cruise_label[grp$cruise_key_method %in% "crosswalk: release CTD/bottle (Q14)"]
xw_sql            <- paste0("'", c(xwalk_labels, "_"), "'", collapse = ", ")
# a label keyed by the crosswalk (or left unresolved) must be one whose cruise the
# ichthyo track lacks entirely (no occupation within a day of its dates), never a
# failed match
dbWriteTable(con, "_off_track", tibble(cruise_label = c(unresolved_labels, xwalk_labels)),
             overwrite = TRUE)
d_unres_track <- dbGetQuery(con, "
  WITH lab AS (
    SELECT cruise_label,
           MIN(datetime_start_utc) :: DATE AS dmin,
           MAX(datetime_start_utc) :: DATE AS dmax
    FROM (SELECT cruise_label, datetime_start_utc FROM cetacean_sighting
          UNION ALL SELECT cruise_label, datetime_start_utc FROM cetacean_segment WHERE NOT misdated)
    GROUP BY 1)
  SELECT l.cruise_label, COUNT(*) AS n_track
  FROM lab l JOIN cruise_track t ON t.datetime :: DATE BETWEEN l.dmin - 1 AND l.dmax + 1
  WHERE l.cruise_label IN (SELECT cruise_label FROM _off_track)
  GROUP BY 1")
dbExecute(con, "DROP TABLE _off_track")
stopifnot(
  "a crosswalk or unresolved label has no ichthyo occupation during its dates (Q14)" =
    nrow(d_unres_track) == 0,
  "every cruise label must resolve to a distinct cruise_key" = length(dupe) == 0,
  "only CC2023-03 and CC2024-07 are unresolved (their cruise month is not the survey code's)" =
    setequal(unresolved_labels, c("CC2023-03", "CC2024-07")),
  "a track-resolved cruise_key exists in the track (no dangling FK)" =
    dbGetQuery(con, glue("
      SELECT COUNT(*) FROM cetacean_sighting WHERE cruise_key IS NOT NULL
        AND cruise_label NOT IN ({xw_sql})
        AND cruise_key NOT IN (SELECT cruise_key FROM cruise_track)"))[[1]] == 0)
cat(glue(
  "{sum(!is.na(grp$cruise_key))}/{nrow(grp)} labels resolved ({length(xwalk_labels)} by the crosswalk); unresolved: ",
  "{paste(unresolved_labels, collapse = ', ')}; ",
  "{n_seg_null} segment(s) without a cruise_key"), "\n")
```

## Load Dataset Metadata + Schema

```{r}
#| label: meta-schema

this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot(nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)

# the registry this dataset's types were added to (metadata/measurement_type.csv,
# 2026-09-27): read, never re-derived, and asserted complete
d_meas_type <- read_measurement_type(here("metadata/measurement_type.csv"))
my_types <- c("group_size", "group_size_min", "group_size_max", "calf_count",
              "behavior", "effort_status", "transect_length")
stopifnot("every type this ingest emits is registered" =
            all(my_types %in% d_meas_type$measurement_type))
dbWriteTable(con, "measurement_type", as.data.frame(d_meas_type), overwrite = TRUE)

# the SOURCE shape; the published shape is the core (Emit Core Tables)
src_rels <- list(
  primary_keys = list(cetacean_segment = "segment_id", cetacean_sighting = "sighting_id",
                      cetacean_species = "species_code"),
  foreign_keys = list(
    list(table = "cetacean_sighting", column = "segment_id",
         ref_table = "cetacean_segment", ref_column = "segment_id"),
    list(table = "cetacean_sighting", column = "species_code",
         ref_table = "cetacean_species", ref_column = "species_code")))
cc_erd(con, tables = c("cetacean_segment", "cetacean_sighting", "cetacean_species", "dataset"),
       rels = src_rels,
       colors = list(lightblue = c("cetacean_segment", "cetacean_sighting"),
                     lightyellow = "cetacean_species", white = "dataset"))
```

## Questions for Data Providers

```{r}
#| label: provider-questions

questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the SIO cetacean sightings data providers (ranked)")
```

## Emit Core Tables

What the source becomes, and what it does not:

| source | core | transform |
|---|---|---|
| `cetacean_segment` row | `sample` (`transect`) | every on-effort segment, on or off transect; key `sio_cetacean-sightings:transect:<segment_id>`; midpoint position; start time |
| `cetacean_segment.segment_length_km` | `sample_measurement` `transect_length` | km; on-transect segments only (`transect_flag` ON), as the type is defined |
| `cetacean_sighting` row | `sample` (`sighting`) | every sighting; parent = its segment when `segment_id` resolves, else none |
| `cetacean_sighting.group_size_best` | `obs` `group_size` | only where Best > 0: a Best of 0 keeps its `sample` row and publishes no obs or attribute (Q13); `measurement_qual` NULL |
| `effort_status` (the segment's `transect_flag`) | `obs_attribute` `effort_status` | `bin_label` ON / OFF, `bin_value` 1 / 0; none where the segment is missing |
| `group_size_min` / `_max` | `obs_attribute` `group_size_min` / `_max` | value in `count` |
| `calf_count` | `obs_attribute` `calf_count`, `life_stage = calf` | plain integers only (Q09) |
| `behavior_primary` / `_other1` | `obs_attribute` `behavior` | **held** (Q03): no rows until the lab supplies the codebook (`publish_behavior <- FALSE`); when enabled, one row per code token, `bin_label` = the code, `bin_value` = 1 / 2 for primary / other1 |
| `species_code` | `obs.taxon_key` via `dataset_taxon` | `Oth`, `EL` and `UNID_OBJCT` keep a NULL taxon |

**Not published** (they stay in the source table and the lab's files): the
behaviour codes (Q03), the sighting conditions (Beaufort, swell, visibility,
precipitation, cloud, glare, wind), observer initials and positions, cue, the
reticle / bearing / distance measurements and the computed perpendicular
distance, the original effort/transect flags, the mixed-group percentages,
comments, `time_diff_min`, `calf_raw`, and the segment start/end coordinates and
times. Perpendicular distance, Beaufort and visibility are what a
line-transect density estimate needs; they are candidates for `obs_attribute`
types once the lab has signed off on publishing this product.

### Taxon vocabulary

```{r}
#| label: taxon-stage

mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

obs_codes      <- unique(d_sighting$species_code)
# not a taxon: "Other", an unidentified object, and EL (not in the lab's list,
# meaning unknown); their sightings keep a NULL taxon_key
excluded_codes <- intersect(obs_codes, d_species$species_code[is.na(d_species$scientific_name)])
stopifnot("Oth, EL and UNID_OBJCT are the only codes without a taxon" =
            setequal(excluded_codes, c("Oth", "EL", "UNID_OBJCT")))

# the declaration: code, the name the code is resolved from, and the common name.
# The source supplied no ids, so none ride along as ds_source_json.
# only the codes the sightings use: an unused code (Be, Beba, CU) would be resolved for nothing
d_vocab <- d_species |>
  filter(!is.na(scientific_name), species_code %in% obs_codes) |>
  transmute(ds_taxa_code = species_code, ds_scientific_name = scientific_name,
            ds_common_name = common_name)
n_staged <- append_dataset_taxon(con, ds_key, d_vocab)

ensure_taxon_xref(con, mt_taxon, tx_over, cache_csv = here("metadata/taxon_xref.csv"))
ensure_taxon_lineage(con, mt_taxon, tx_over, cache_csv = here("metadata/taxon_lineage.csv"))
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- resolve_dataset_taxon(con, mt_taxon, tx_over)
n_tx_group <- build_taxon_group(con, read_taxon_group_rules(here("metadata/taxon_group.csv")))

chk <- check_dataset_taxon(con, ds_key, codes = setdiff(obs_codes, excluded_codes))

# every staged code keys a Cetacea taxon (the vocabulary is cetaceans only)
dt <- dbGetQuery(con, "
  SELECT d.ds_taxa_code, d.taxon_key, t.scientific_name, t.rank
  FROM dataset_taxon d LEFT JOIN taxon t USING (taxon_key)
  WHERE d.dataset_key = ?", params = list(ds_key))
key_of <- function(code) dt$taxon_key[dt$ds_taxa_code == code]
stopifnot(
  "every staged code has a taxon_key"   = !anyNA(dt$taxon_key),
  "Bbo keys Berardius bairdii"          = identical(dt$scientific_name[dt$ds_taxa_code == "Bbo"], "Berardius bairdii"),
  "Mesoplodon spp keys the genus"       = identical(dt$scientific_name[dt$ds_taxa_code == "Mesoplodon spp"], "Mesoplodon") &&
                                          identical(dt$rank[dt$ds_taxa_code == "Mesoplodon spp"], "Genus"),
  "Dc keys Delphinus capensis (Q10)"    = identical(key_of("Dc"), "worms:383816"),
  "Dc and Dd are distinct taxa (Q10)"   = !identical(key_of("Dc"), key_of("Dd")),
  "Lo keys Sagmatias obliquidens (Q10)" = identical(key_of("Lo"), "worms:1571909"))
cat(glue("vocabulary: {n_staged} codes staged, {length(excluded_codes)} excluded (Oth, EL, UNID_OBJCT); ",
         "taxon {n_taxon}, dataset_taxon {n_ds_taxon}, taxon_group {n_tx_group}"), "\n")
dt |> datatable(caption = "species_code -> taxon", rownames = FALSE, filter = "top")
```

### Sample, obs, obs_attribute, sample_measurement

```{r}
#| label: emit_core

tr_key <- function(col) ns_key(ds_key, "transect", col)
si_key <- function(col) ns_key(ds_key, "sighting", col)

# transects: one sample per on-effort segment, on or off transect
append_sample(con, sample_arm_self(ds_key, "cetacean_segment", "segment_id", "transect"))

# sightings: one sample per sighting, the child of its transect when on one. The
# parent is looked up, not assumed: a segment_id missing from the effort file
# gets no parent rather than a dangling one.
append_sample(con, glue("
  SELECT {si_key('s.sighting_id')} AS sample_key, 'sighting' AS sample_type,
         CASE WHEN g.segment_id IS NOT NULL THEN {tr_key('g.segment_id')} END AS parent_sample_key,
         COALESCE(CASE WHEN g.segment_id IS NOT NULL THEN {tr_key('g.segment_id')} END,
                  {si_key('s.sighting_id')}) AS root_sample_key,
         '{ds_key}' AS dataset_key, s.grid_key, NULL::VARCHAR AS site_key, s.cruise_key,
         NULL::INTEGER AS order_occ, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP) AS datetime,
         0::DOUBLE AS depth_min_m, 0::DOUBLE AS depth_max_m, NULL::VARCHAR AS tow_type
  FROM cetacean_sighting s
  LEFT JOIN cetacean_segment g USING (segment_id)"))

# effort: segment length on the transect sample, for the on-transect segments
# only (`transect_length` is the length of an on-effort, ON-transect segment);
# an off-transect segment keeps its sample row and its length in the source
append_sample_measurement(con, glue("
  SELECT {tr_key('segment_id')}, '{ds_key}', 'transect_length',
         segment_length_km, NULL::VARCHAR
  FROM cetacean_segment WHERE segment_length_km IS NOT NULL AND transect_flag = 'ON'"))

# obs -- the headline: one row per COUNTED sighting, the group size (best
# estimate). One species per sighting (Q05), so the sighting IS the occurrence
# grain (a mixed group is split by the lab into one row per species, a/b). A
# Best of 0 is animals seen but not counted (a blow, a splash):
# a row here would assert a measured group of 0, which reads as an absence, so
# those sightings keep their sample row and publish no obs (Q13; Ben, 2026-10-01).
# measurement_qual stays NULL: the on/off-effort flag is a sampling regime, not a
# quality flag, and cc_qual_ok_sql() would keep OFF rows anyway (obs_attribute
# effort_status below).
counted <- "s.group_size_best > 0"
append_obs(con, glue("
  SELECT 'bio', '{ds_key}', {si_key('s.sighting_id')},
         s.grid_key, s.cruise_key, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP), 0::DOUBLE, 0::DOUBLE,
         dt.taxon_key, NULL::VARCHAR, 'group_size', s.group_size_best,
         NULL::VARCHAR, NULL::DOUBLE
  FROM cetacean_sighting s
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = s.species_code
  WHERE {counted}"))

# obs_attribute -- effort status, group size range, calves, behaviour, each on a
# counted sighting only (an attribute of an occurrence that is not published
# would describe nothing). Not a breakdown of the headline (none of these sums
# to group_size), so there is no sum-parity to assert; each is asserted against
# its source column instead.
attr_sel <- function(type, life_stage, bin_value, bin_label, count, where) glue("
  SELECT '{ds_key}', {si_key('s.sighting_id')}, dt.taxon_key, {life_stage}, '{type}',
         {bin_value}, {bin_label}, {count}, NULL::VARCHAR
  FROM cetacean_sighting s
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = s.species_code
  WHERE {counted} AND {where}")
# the segment's on-transect flag (every sighting is on effort): line-transect
# density uses ON only; OFF is a different sampling regime, not bad data
append_obs_attribute(con, attr_sel("effort_status", "NULL::VARCHAR",
                                   "CAST(CASE s.effort_status WHEN 'ON' THEN 1 ELSE 0 END AS DOUBLE)",
                                   "s.effort_status", "NULL::INTEGER", "s.effort_status IS NOT NULL"))
append_obs_attribute(con, attr_sel("group_size_min", "NULL::VARCHAR", "NULL::DOUBLE", "NULL::VARCHAR",
                                   "CAST(s.group_size_min AS INTEGER)", "s.group_size_min IS NOT NULL"))
append_obs_attribute(con, attr_sel("group_size_max", "NULL::VARCHAR", "NULL::DOUBLE", "NULL::VARCHAR",
                                   "CAST(s.group_size_max AS INTEGER)", "s.group_size_max IS NOT NULL"))
append_obs_attribute(con, attr_sel("calf_count", "'calf'", "NULL::DOUBLE", "NULL::VARCHAR",
                                   "s.calf_count", "s.calf_count IS NOT NULL"))

# behaviour: HELD (Q03, Ben 2026-10-04). The two columns hold bare numeric codes
# ("1", "1, 34", "2, 5, 8") with no codebook, in the same `behavior` type that
# farallon_bird-mammal fills with words, so no rows are emitted until the lab
# supplies the codebook. Flip `publish_behavior` then, map code -> label here (or
# emit under a dataset-specific type), and add the behavior assertions back.
# When on, each token is its own row, labelled with the code as recorded.
publish_behavior <- FALSE
if (publish_behavior) {
  d_behavior <- d_sighting |>
    select(sighting_id, behavior_primary, behavior_other1) |>
    pivot_longer(-sighting_id, names_to = "slot", values_to = "code") |>
    filter(!is.na(code)) |>
    mutate(slot = match(slot, c("behavior_primary", "behavior_other1")),
           # split lists of numeric codes; free text ("Slow Travel") stays whole
           code = map(code, \(x) if (str_detect(x, "^[0-9?,/&\\s]+$"))
                                     str_split_1(x, "[,/&\\s]+") else x)) |>
    unnest(code) |>
    mutate(code = str_trim(code)) |>
    filter(code != "")
  dbWriteTable(con, "_behavior", d_behavior, overwrite = TRUE)
  append_obs_attribute(con, glue("
    SELECT '{ds_key}', {si_key('b.sighting_id')}, dt.taxon_key, NULL::VARCHAR, 'behavior',
           CAST(b.slot AS DOUBLE), b.code, NULL::INTEGER, NULL::VARCHAR
    FROM _behavior b JOIN cetacean_sighting s USING (sighting_id)
    LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                              AND dt.ds_taxa_code = s.species_code
    WHERE {counted}"))
  dbExecute(con, "DROP TABLE _behavior")
}

q1 <- function(sql) dbGetQuery(con, sql)[[1]]
d_counted  <- d_sighting |> filter(group_size_best > 0)
n_uncounted <- nrow(d_sighting) - nrow(d_counted)
core <- list(
  sample             = q1("SELECT COUNT(*) FROM sample"),
  sample_transect    = q1("SELECT COUNT(*) FROM sample WHERE sample_type = 'transect'"),
  sample_sighting    = q1("SELECT COUNT(*) FROM sample WHERE sample_type = 'sighting'"),
  sample_measurement = q1("SELECT COUNT(*) FROM sample_measurement"),
  sighting_uncounted = n_uncounted,
  obs                = q1("SELECT COUNT(*) FROM obs"),
  obs_attribute      = q1("SELECT COUNT(*) FROM obs_attribute"),
  effort_on          = q1("SELECT COUNT(*) FROM obs_attribute WHERE measurement_type = 'effort_status' AND bin_label = 'ON'"),
  effort_off         = q1("SELECT COUNT(*) FROM obs_attribute WHERE measurement_type = 'effort_status' AND bin_label = 'OFF'"))
cat(glue("core projection -- ", paste(names(core), unlist(core), sep = "=", collapse = " ")), "\n")

stopifnot(
  "one transect sample per segment"  = core$sample_transect == nrow(d_segment),
  "one sighting sample per sighting, counted or not" = core$sample_sighting == nrow(d_sighting),
  "the uncounted sightings are the 2 zeros (Q13: MN on 202504SH, UD on 202511SR)" =
    n_uncounted == 2L && sum(d_sighting$group_size_best %in% 0) == 2L,
  "one obs per counted sighting"     = core$obs == nrow(d_counted),
  "no obs has a group size of 0 or NULL (Q13)" =
    q1("SELECT COUNT(*) FROM obs WHERE measurement_value IS NULL OR measurement_value <= 0") == 0,
  "obs value total = source Best total" =
    isTRUE(all.equal(q1("SELECT SUM(measurement_value) FROM obs"),
                     sum(d_sighting$group_size_best, na.rm = TRUE))),
  "measurement_qual is not used (effort status is an obs_attribute)" =
    q1("SELECT COUNT(*) FROM obs WHERE measurement_qual IS NOT NULL") == 0,
  "effort_status rows = counted sightings with the flag, ON and OFF as in the source" =
    core$effort_on  == sum(d_counted$effort_status %in% "ON") &&
    core$effort_off == sum(d_counted$effort_status %in% "OFF"),
  "group_size_min rows = non-NULL Min on counted sightings" =
    q1("SELECT COUNT(*) FROM obs_attribute WHERE measurement_type = 'group_size_min'") ==
    sum(!is.na(d_counted$group_size_min)),
  "calf_count total = source total on counted sightings" =
    q1("SELECT COALESCE(SUM(count), 0) FROM obs_attribute WHERE measurement_type = 'calf_count'") ==
    sum(d_counted$calf_count, na.rm = TRUE),
  "no behavior rows while the codes are held (Q03)" =
    publish_behavior || q1("SELECT COUNT(*) FROM obs_attribute WHERE measurement_type = 'behavior'") == 0,
  "every obs_attribute row sits on a sighting that has an obs" =
    q1("SELECT COUNT(*) FROM obs_attribute a WHERE a.sample_key NOT IN (SELECT sample_key FROM obs)") == 0,
  "transect_length rows and total = the on-transect segments'" =
    core$sample_measurement == sum(d_segment$transect_flag == "ON") &&
    isTRUE(all.equal(q1("SELECT SUM(measurement_value) FROM sample_measurement"),
                     sum(d_segment$segment_length_km[d_segment$transect_flag == "ON"]))),
  "every obs.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key) WHERE s.sample_key IS NULL") == 0,
  "every obs_attribute.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs_attribute a LEFT JOIN sample s USING (sample_key) WHERE s.sample_key IS NULL") == 0,
  "every sample.parent_sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM sample c LEFT JOIN sample p ON p.sample_key = c.parent_sample_key
       WHERE c.parent_sample_key IS NOT NULL AND p.sample_key IS NULL") == 0,
  "every obs.taxon_key RESOLVES in taxon" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN taxon t USING (taxon_key)
       WHERE o.taxon_key IS NOT NULL AND t.taxon_key IS NULL") == 0,
  "a NULL obs.taxon_key is an Oth, EL or UNID_OBJCT sighting and nothing else" =
    q1("SELECT COUNT(*) FROM obs WHERE taxon_key IS NULL") == sum(d_counted$species_code %in% excluded_codes),
  "every obs / obs_attribute / sample_measurement type is registered" =
    q1("SELECT COUNT(*) FROM (SELECT measurement_type FROM obs UNION
                             SELECT measurement_type FROM obs_attribute UNION
                             SELECT measurement_type FROM sample_measurement) x
       WHERE measurement_type NOT IN (SELECT measurement_type FROM measurement_type)") == 0)
```

### Declared NULLs

`validate_for_release()` treats every `_key` column as required. These NULLs are
the dataset's contract, each with its count and reason; anything else fails.

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)

lbl_sql <- paste0("'", c(unresolved_labels, "_"), "'", collapse = ", ")
nullable <- tribble(
  ~table,   ~column,             ~n,                                              ~reason,
  "sample", "cruise_key",        q1("SELECT COUNT(*) FROM sample WHERE cruise_key IS NULL"),
    glue("labels resolved by neither the ichthyo track nor the crosswalk ({paste(unresolved_labels, collapse = ', ')}), their segments, and the misdated segment 2886"),
  "sample", "parent_sample_key", n_seg_orphan + nrow(d_segment),
    "transects are roots; the sightings whose segment_id is not in the segments file (419) have no parent",
  "sample", "site_key",          nrow(d_segment) + nrow(d_sighting),
    "underway survey: no event sits on a CalCOFI station",
  "sample", "grid_key",          q1("SELECT COUNT(*) FROM sample WHERE grid_key IS NULL"),
    "positions outside the CalCOFI grid (e.g. the CC2015-10 transit north of 39 N)",
  "sample", "datetime",          0L,
    "none: every sighting and segment has a time",
  "obs",    "taxon_key",         sum(d_counted$species_code %in% excluded_codes),
    "'Oth' (Other) and 'UNID_OBJCT' are not taxa; 'EL' is not in the lab's list (meaning unknown)")
# the counts above that are measured rather than derived from the source must
# still be explained: every NULL cruise_key belongs to an unresolved label
stopifnot(
  "every NULL sighting cruise_key belongs to an unresolved label" =
    q1(glue("SELECT COUNT(*) FROM cetacean_sighting WHERE cruise_key IS NULL
            AND cruise_label NOT IN ({lbl_sql})")) == 0,
  "every NULL segment cruise_key belongs to an unresolved label or is misdated" =
    q1(glue("SELECT COUNT(*) FROM cetacean_segment WHERE cruise_key IS NULL
            AND cruise_label NOT IN ({lbl_sql}) AND NOT misdated")) == 0,
  "the declared NULL counts match the tables" =
    q1("SELECT COUNT(*) FROM sample WHERE parent_sample_key IS NULL") ==
      nullable$n[nullable$column == "parent_sample_key"] &&
    q1("SELECT COUNT(*) FROM sample WHERE datetime IS NULL") ==
      nullable$n[nullable$column == "datetime"])
cat("validate_for_release:", ifelse(results$passed, "PASSED", "reported findings -- reconciled against the declared NULLs below"), "\n")
nullable |> datatable(caption = "Declared NULLs (count and reason)", rownames = FALSE)
```

### Measurement bounds

```{r}
#| label: bounds

b <- bind_rows(
  check_measurement_bounds(con, "obs",                mt = d_meas_type, dataset_key = ds_key),
  check_measurement_bounds(con, "sample_measurement", mt = d_meas_type, dataset_key = ds_key))
bounds_datatable(b)
stopifnot("every emitted bound is declared and respected" = all(b$status == "ok"))
```

## Preview

```{r}
#| label: preview-obs

dbGetQuery(con, "
  SELECT o.taxon_key, t.scientific_name, COUNT(*) AS n_sightings,
         SUM(o.measurement_value) AS n_animals,
         COUNT(*) FILTER (WHERE a.bin_label = 'ON') AS n_on_effort,
         MIN(year(o.datetime)) AS year_min, MAX(year(o.datetime)) AS year_max
  FROM obs o LEFT JOIN taxon t USING (taxon_key)
  LEFT JOIN obs_attribute a ON a.sample_key = o.sample_key AND a.measurement_type = 'effort_status'
  GROUP BY ALL ORDER BY n_sightings DESC") |>
  datatable(caption = "obs by taxon", rownames = FALSE)
```

```{r}
#| label: preview-sample

dbGetQuery(con, "
  SELECT sample_key, sample_type, parent_sample_key, cruise_key, grid_key,
         latitude, longitude, datetime
  FROM sample ORDER BY datetime LIMIT 200") |>
  datatable(caption = "sample -- first 200 by time", rownames = FALSE, filter = "top")
```

```{r}
#| label: erd-core

core_tbls <- core_output_tables(con)
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls))
```

## Write Outputs + Upload

```{r}
#| label: outputs

dir_create(dir_parquet)
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
write_parquet_outputs(
  con = con, output_dir = dir_parquet, tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)

d_tbls_rd <- read_csv(file.path(dir_meta, "tbls_redefine.csv"))
d_flds_rd <- read_csv(file.path(dir_meta, "flds_redefine.csv"))
build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           file.path(dir_meta, "metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned,
  sources = src_stamp)

# gs://calcofi-db is world-readable: a held dataset (in_release: false, until the
# lab signs off) is never mirrored there
if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet,
              gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else {
  cat("held (in_release: false) or CALCOFI_SKIP_GCS set -- parquet kept local, NOT synced to gs://calcofi-db\n")
}
```

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
