---
title: "Ingest CalCOFI October 2022 Vertebrate eDNA"
calcofi:
  target_name: ingest_calcofi_2022_edna
  workflow_type: ingest
  dependency:
    - ingest_calcofi_bottle
  output: data/parquet/calcofi_2022-edna/manifest.json
  provider: calcofi
  dataset: 2022-edna
  workflow_url: https://calcofi.io/workflows/ingest_calcofi_2022-edna.html
  questions_file: metadata/calcofi/2022-edna/questions.csv
  dataset_meta:
    dataset_name: CalCOFI October 2022 Vertebrate eDNA
    # display trio, read by the release `dataset` table and consumer apps
    dataset_name_short: 2022 Vertebrate eDNA
    category: Genomics & eDNA
    color: "#2e8b57"
  # publishes the consolidated core: the source shape (a Darwin Core Archive,
  # not a wide CSV) is wrangled in this notebook and projected into the shared
  # core tables. This is the first ingest to populate the Genomics & eDNA
  # category with real observations (calpoly_whale-edna and
  # jcvi_ncog-metabarcoding remain holdings, no ingest).
  tables_owned:
    - {table: sample, shared: true, note: "one row per water-filter sample (parentEventID grain, not eventID -- see Overview)"}
    - {table: obs, shared: true, note: "species detections; measurement_type = sequence_reads"}
    - {table: obs_attribute, shared: true, note: "assay dimension (D-loop vs 12S/MiFish) per detection"}
    - {table: sample_measurement, shared: true, note: "co-collected water chemistry + genomics QC, from extendedmeasurementorfact.txt"}
    - {table: taxon, shared: true}
    - {table: dataset_taxon, shared: true}
    - {table: measurement_type, shared: true, note: "adds sequence_reads, oxygen_mg_l, dna_concentration, reads_raw_total, reads_filtered_total, otu_richness"}
  erd:
    color: "#bbe0f0"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: GBIF/OBIS Darwin Core Archive, DOI
[10.15468/n52j6r](https://www.gbif.org/dataset/10524092-539d-4ea4-8539-2e91ad8581a9)
("CalCOFI October 2022 Vertebrate eDNA"). Linked to Patin et al. 2026
(*Methods in Ecology and Evolution*, https://doi.org/10.1111/2041-210x.70370).
201 occurrence records from 47 water-filter samples collected 2022-10-13 to
2022-10-18 at CalCOFI/GEMCAP stations, each filter run through up to two
molecular assays: mitochondrial D-loop (cetacean detections -- *Delphinus*,
*Megaptera novaeangliae*) and 12S rRNA "MiFish" (fish detections). Source
`datasetID` is `CalCOFI_Intercal` -- an intercalibration exercise, published
as a CalCOFI program product (provider `calcofi`, not `sio`: see
`provider.csv` -- this dataset is branded and identified as CalCOFI's own,
unlike e.g. `sio_mesopelagic-fish`, which is an SIO lab collection CalCOFI
merely sampled for).

**Not a wide CSV.** This ingest reads the DwC-A's three tables directly
(`occurrence.txt`, `dnaderiveddata.txt`, `extendedmeasurementorfact.txt`,
tab-delimited) instead of `read_csv_files()`, the same way
`ingest_calcofi_ctd-cast.qmd` bypasses it for its own non-CSV source. The
source zip (`dwca-calcofi_*.zip`, GBIF's DwC-A download) lives at
`dir_data/{provider}/{dataset}/` -- `data-public/calcofi/2022-edna/` -- the
same convention as every other dataset's folder (`bottle/`, `ctd-cast/`,
`dic/`, `mets/`), not a special `genomics/` location. A second zip in that
folder (GBIF's raw occurrence-download export) is not used here -- the DwC-A
is the complete, canonical source; the other is a redundant representation
kept for reference.

**Sample grain is `parentEventID`, not `eventID`.** Every `eventID` under a
given `parentEventID` shares identical `decimalLatitude`/`decimalLongitude`/
`minimumDepthInMeters`/`eventDate` (checked against all 201 rows: 0/47 groups
disagree) -- the `eventID` split is the assay dimension (D-loop vs MiFish), not
a distinct physical event. Treating `eventID` as the sample grain would
duplicate every filter's coordinates once per assay it ran (up to 2x) and
double-count "samples per station" in any consumer query. This mirrors
`calcofi_bottle`, where one physical bottle (not each `bottle_measurement`
row) is the sample.

### Data Flow

```{mermaid}
graph LR
  A[occurrence.txt + dnaderiveddata.txt + extendedmeasurementorfact.txt] --> B[DuckDB Wrangling]
  B --> C[Validate & Clean]
  C --> D[Parquet Export]
  D --> E[GCS Archive]
  E --> F[Release Database]
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr, lubridate, purrr,
  readr, sf, stringr, tibble, tidyr, units, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_calcofi_2022-edna.qmd"))
provider     <- cc$provider    # "calcofi"
dataset      <- cc$dataset     # "2022-edna"
ds_key       <- glue("{provider}_{dataset}")
tables_owned <- cc$tables_owned
dir_label    <- ds_key
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
dir_stage    <- cc_stage_path("parquet", dir_label, create = TRUE)
dir_ext      <- cc_stage_path("2022-edna", "unzip", create = TRUE)
dir_meta     <- here(glue("metadata/{provider}/{dataset}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))  dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

meas_type_csv <- here("metadata/measurement_type.csv")
d_meas_type   <- read_measurement_type(meas_type_csv)

# required by build_metadata_json() even when read_csv_files() is bypassed for
# the main data read -- same requirement confirmed in ingest_calcofi_ctd-cast.qmd
d_flds_rd <- read_csv(glue("{dir_meta}/flds_redefine.csv"), show_col_types = FALSE)
d_tbls_rd <- read_csv(glue("{dir_meta}/tbls_redefine.csv"), show_col_types = FALSE)
```

## Read Source Data

Reads the three DwC-A tables directly as tab-delimited text -- `read_csv_files()`
assumes one wide CSV per dataset, which this source does not fit (Occurrence
core + 2 extensions). Source is a zip (GBIF's DwC-A download); the raw zip is
archived to GCS as received (`sync_to_gcs()`, the same non-CSV-source pattern
`ingest_calcofi_ctd-cast.qmd` uses), and unzipped into the wrangling scratch
dir for reading -- the extracted `.txt` files never go to GCS themselves.

```{r}
#| label: read-source

dir_source <- path(dir_data, provider, dataset)
dwca_zips  <- dir_ls(dir_source, glob = "*dwca*.zip")
stopifnot("expected exactly one dwca-*.zip in the dataset folder" = length(dwca_zips) == 1)
dwca_zip <- dwca_zips[[1]]
cat(glue("using source zip: {path_file(dwca_zip)}"), "\n")

sync_to_gcs(
  local_dir  = dir_source,
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket     = "calcofi-files-public",
  exclude    = c(".DS_Store"))

dir_dwca <- path(dir_ext, "dwca")
dir_create(dir_dwca)
unzip(dwca_zip, exdir = dir_dwca, overwrite = TRUE)

read_dwca_txt <- function(name) {
  read_tsv(path(dir_dwca, name), col_types = cols(.default = "c"),
           na = character())
}

d_occ  <- read_dwca_txt("occurrence.txt")
d_dna  <- read_dwca_txt("dnaderiveddata.txt")
d_emof <- read_dwca_txt("extendedmeasurementorfact.txt")

cat(glue(
  "read {nrow(d_occ)} occurrence rows, {nrow(d_dna)} dnaderiveddata rows, ",
  "{nrow(d_emof)} extendedmeasurementorfact rows"), "\n")
```

## Check Data Integrity

No `read_csv_files()` result to hand `check_data_integrity()`, so integrity is
checked directly against the shape this ingest actually depends on: every
`occurrenceID` unique, every occurrence joins 1:1 to `dnaderiveddata` (the
assay source), every `parentEventID` group internally agrees on
position/time/depth (the sample-grain decision above), and no `occurrenceStatus`
other than `present` (no absence records to filter).

```{r}
#| label: check-integrity

n_dup_occid <- d_occ |> count(occurrenceID) |> filter(n > 1) |> nrow()
stopifnot("occurrenceID must be unique in occurrence.txt" = n_dup_occid == 0)

n_unjoined <- d_occ |> anti_join(d_dna, by = "occurrenceID") |> nrow()
stopifnot("every occurrence must join to dnaderiveddata by occurrenceID" = n_unjoined == 0)

occ_status <- unique(d_occ$occurrenceStatus)
stopifnot("only 'present' occurrenceStatus expected -- an 'absent' record needs its own handling" =
            setequal(occ_status, "present"))

grain_check <- d_occ |>
  distinct(parentEventID, decimalLatitude, decimalLongitude,
           minimumDepthInMeters, maximumDepthInMeters, eventDate) |>
  count(parentEventID) |>
  filter(n > 1)
stopifnot("every parentEventID must share one position/depth/date across its eventIDs" =
            nrow(grain_check) == 0)

n_parent <- n_distinct(d_occ$parentEventID)
cat(glue(
  "integrity OK: {nrow(d_occ)} occurrences / {n_parent} parentEventID samples / ",
  "{n_distinct(d_occ$eventID)} eventIDs (assay runs); 0 duplicate occurrenceIDs; ",
  "0 unjoined to dnaderiveddata; 0 grain conflicts"), "\n")
```

## Build Sample (Filter) Table

```{r}
#| label: build-sample

d_sample <- d_occ |>
  distinct(parentEventID, decimalLatitude, decimalLongitude,
           minimumDepthInMeters, maximumDepthInMeters, eventDate,
           locationID, locality) |>
  transmute(
    parent_event_id = parentEventID,
    latitude   = as.numeric(decimalLatitude),
    longitude  = as.numeric(decimalLongitude),
    depth_min_m = as.numeric(minimumDepthInMeters),
    depth_max_m = as.numeric(maximumDepthInMeters),
    datetime_start_utc = as_datetime(eventDate),
    location_id = locationID,
    locality = locality)

dbWriteTable(con, "edna_sample", as.data.frame(d_sample), overwrite = TRUE)
cat(glue("edna_sample: {nrow(d_sample)} rows (expect 47)"), "\n")
```

## Standardize Taxonomy

Every occurrence's `scientificNameID` is already a resolved WoRMS AphiaID
(`urn:lsid:marinespecies.org:taxname:<id>`) -- there is no name-matching to do,
unlike `ingest_farallon_bird-mammal.qmd`'s ERDDAP-code crosswalk. The
declaration step below extracts that id directly rather than looking it up.

```{r}
#| label: taxon-stage

mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

dbWriteTable(con, "edna_occurrence", as.data.frame(d_occ), overwrite = TRUE)

d_vocab <- dbGetQuery(con, "
  SELECT DISTINCT scientificNameID                                AS ds_taxa_code,
         scientificName                                            AS ds_scientific_name,
         CAST(regexp_extract(scientificNameID, '([0-9]+)$', 1) AS INTEGER) AS worms_id
  FROM edna_occurrence
  ORDER BY 1")
stopifnot("every occurrence must carry a WoRMS scientificNameID" =
            all(!is.na(d_vocab$worms_id)))

n_staged <- append_dataset_taxon(con, ds_key, d_vocab)

ensure_taxon_xref(con, mt_taxon, tx_over,
                  cache_csv = here("metadata/taxon_xref.csv"))
ensure_taxon_lineage(con, mt_taxon, tx_over,
                     cache_csv = here("metadata/taxon_lineage.csv"))
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- resolve_dataset_taxon(con, mt_taxon, tx_over)

obs_codes <- dbGetQuery(con, "SELECT DISTINCT scientificNameID FROM edna_occurrence")[[1]]
chk_taxon <- check_dataset_taxon(con, ds_key, codes = obs_codes, allow = character())

cat(glue("taxonomy: {n_staged} taxa staged (all WoRMS-resolved), ",
         "taxon {n_taxon}, dataset_taxon {n_ds_taxon}"), "\n")
```

## Assay (D-loop vs MiFish) Lookup

The molecular assay is a property of the *detection*, not the sample (one
filter carries detections from both assays) -- it comes from
`dnaderiveddata.target_gene`, joined by `occurrenceID`. Carried as an
`obs_attribute` row (measurement_type = 'assay', bin_label = the assay name),
matching how `ingest_farallon_bird-mammal.qmd` records `behavior` -- Q02,
RESOLVED against the real obs_attribute schema.

```{r}
#| label: assay-lookup

dbWriteTable(con, "edna_dna", as.data.frame(d_dna), overwrite = TRUE)

d_assay <- dbGetQuery(con, "
  SELECT o.occurrenceID, o.parentEventID, o.eventID, o.scientificNameID,
         o.organismQuantity,
         CASE d.target_gene
           WHEN 'D-loop' THEN 'D-loop'
           WHEN '12S rRNA (SSU mitochondria)' THEN '12S (MiFish)'
           ELSE d.target_gene
         END AS assay
  FROM edna_occurrence o JOIN edna_dna d USING (occurrenceID)")
dbWriteTable(con, "edna_assay", d_assay, overwrite = TRUE)

d_assay |> count(assay) |>
  datatable(caption = "Detections by assay", rownames = FALSE)
```

## Add Spatial

No `enforce_column_types()` step in this notebook (unlike `ingest_calcofi_ctd-cast.qmd`/
`ingest_calcofi_mets.qmd`/`ingest_cdfw_dungeness-crab.qmd`, which all call it): once
`add_point_geom()` below adds a GEOMETRY column to `edna_sample`, DuckDB >= 1.5.1
cannot rewrite that table's column types (dungeness-crab's Data Preparation section
states this explicitly, ordering its own `enforce_column_types()` call *before*
`add_point_geom()` for the same reason). `ingest_farallon_bird-mammal.qmd`,
`ingest_sio_mesopelagic-fish.qmd`, `ingest_calcofi_phyllosoma.qmd`, and
`ingest_cce-lter_euphausiids.qmd` -- every other ingest using `add_point_geom()`
on its own staging table, the pattern this notebook follows -- omit
`enforce_column_types()` entirely rather than reordering around it.

```{r}
#| label: add-spatial

add_point_geom(con, "edna_sample", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "edna_sample") |> datatable(caption = "Grid assignment")

n_sk  <- dbGetQuery(con, "SELECT COUNT(*) FROM edna_sample WHERE grid_key IS NOT NULL")[[1]]
n_all <- dbGetQuery(con, "SELECT COUNT(*) FROM edna_sample")[[1]]
cat(glue("grid_key resolved: {n_sk}/{n_all}"), "\n")

# Q01 RESOLVED 2026-09-26: the source carries no ship/cruise id, but the
# provider metadata (GBIF/IPT page + eml.xml acknowledgements) names the
# vessel as NOAA Ship Reuben Lasker. Confirmed against calcofi.io/db-query:
# `ship` has ship_key 'RL' = REUBEN LASKER (ship_nodc '33UD'), and `cruise`
# has exactly one row overlapping this dataset's 2022-10-13..18 window --
# cruise_key '2022-10-33UD' (its own '33UD' suffix is that same NODC code).
# Resolved via the shared helper (cce-lter_euphausiids.qmd precedent), not
# hand-written SQL (issue #47); ship_key is a constant here since all 47
# samples come from this one cruise.
dbExecute(con, "ALTER TABLE edna_sample ADD COLUMN IF NOT EXISTS ship_key VARCHAR")
dbExecute(con, "UPDATE edna_sample SET ship_key = 'RL'")
resolve_cruise_key(con, "edna_sample", datetime_col = "datetime_start_utc",
                   require_in_cruise = TRUE) |>
  datatable(caption = "cruise_key resolution method")

n_ck <- dbGetQuery(con, "SELECT COUNT(*) FROM edna_sample WHERE cruise_key IS NOT NULL")[[1]]
cat(glue("cruise_key resolved: {n_ck}/{n_all}"), "\n")
```

## Water Chemistry + Genomics QC (sample_measurement)

`extendedmeasurementorfact.txt` is 12,632 rows, but only 7 fields are real
per-sample measurements (identical across every occurrence sharing a
`parentEventID`, confirmed for all 47 groups) plus 4 genomics QC values; the
other ~50 distinct `measurementType` values are FAIRe-checklist protocol text
(kit names, SOP links, extraction dates) with no place in a value/unit table.
Six of the seven measurements reuse existing canonical `measurement_type`
rows by exact unit match (`ammonia`, `nitrate`, `nitrite`, `phosphate`,
`silicate`, `chl_fluor`, all confirmed against the live v2026.07.30 release
schema); dissolved oxygen is reported in mg/L, which no existing oxygen type
uses, so it gets a new `oxygen_mg_l` type rather than a guessed conversion
(Q03).

```{r}
#| label: emof-context

dbWriteTable(con, "edna_emof", as.data.frame(d_emof), overwrite = TRUE)

emof_type_map <- tribble(
  ~source_type,                              ~measurement_type,
  "Concentration of ammonia in the sample",   "ammonia",
  "Concentration of nitrate in the sample",   "nitrate",
  "Concentration of nitrite in the sample",   "nitrite",
  "Concentration of phosphate in the sample", "phosphate",
  "Concentration of silicate in the sample",  "silicate",
  "Chlorophyll fluorescence",                 "chl_fluor",
  "Concentration of dissolved oxygen",        "oxygen_mg_l",
  "Concentration of total DNA after extraction", "dna_concentration",
  "The total number of raw sequence reads for each sample.", "reads_raw_total",
  "The number of filtered sequence reads for each sample, that made it through bioinformatic filtering and used for final/subsequent analyses.", "reads_filtered_total",
  "Total number of OTUs or ASVs assigned to taxa based on the study-specific thresholds for each sample", "otu_richness")

dbWriteTable(con, "emof_type_map", emof_type_map, overwrite = TRUE)

# dedupe to one value per (parentEventID, measurement_type): every eventID
# child of a parentEventID carries an identical context value (checked above
# for ammonia; same join logic covers all seven), so MIN() here is a
# tie-break over identical values, not an aggregation.
d_context <- dbGetQuery(con, "
  SELECT o.parentEventID, m.measurement_type,
         MIN(TRY_CAST(e.measurementValue AS DOUBLE)) AS measurement_value
  FROM edna_emof e
  JOIN edna_occurrence o USING (occurrenceID)
  JOIN emof_type_map m ON m.source_type = e.measurementType
  GROUP BY o.parentEventID, m.measurement_type")
dbWriteTable(con, "edna_context", d_context, overwrite = TRUE)

n_dropped_types <- dbGetQuery(con, "
  SELECT COUNT(DISTINCT measurementType) FROM edna_emof
  WHERE measurementType NOT IN (SELECT source_type FROM emof_type_map)")[[1]]
cat(glue(
  "sample_measurement staged: {nrow(d_context)} rows across {nrow(emof_type_map)} types, ",
  "{n_parent <- n_distinct(d_context$parentEventID)} samples; ",
  "{n_dropped_types} protocol-text measurementTypes excluded (not measurements)"), "\n")
```

## Declare and Enforce Physical Bounds

Per RUNBOOK.md 3b: every ingest with numeric `sample_measurement` values checks
them against `metadata/measurement_type.csv`'s declared bounds before leaving
the notebook (`ingest_calcofi_bottle.qmd`/`ingest_calcofi_mets.qmd` do this on
their own measurement staging table; this is the same step, run on
`edna_context`). All 11 types here (6 reused, 5 new) already carry a
`valid_min = 0`; `oxygen_mg_l` also has `valid_max = 20`. A finding that isn't
`ok` gets a `proposed` row in `questions.csv`, not a silent drop.

```{r}
#| label: bounds-check

d_bounds <- check_measurement_bounds(con, "edna_context", mt = d_meas_type)
bounds_datatable(d_bounds)
```

```{r}
#| label: bounds-enforce

oob_tally <- drop_out_of_bounds(con, "edna_context", mt = d_meas_type)
cat(glue("out-of-bounds rows dropped from edna_context: {oob_tally}"), "\n")
```

## Schema Documentation

Color scheme: lightblue = new dataset tables, lightgreen = taxon staging,
lightyellow = amended reference tables, white = shared metadata.

```{r}
#| label: schema

edna_rels <- list(
  primary_keys = list(
    edna_sample = "parent_event_id",
    measurement_type = "measurement_type"),
  foreign_keys = list(
    list(table = "edna_assay", column = "parentEventID",
         ref_table = "edna_sample", ref_column = "parent_event_id"),
    list(table = "edna_context", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))

cc_erd(
  con,
  tables = c("edna_sample", "edna_assay", "edna_context",
             "measurement_type", "dataset"),
  rels   = edna_rels,
  colors = list(
    lightblue   = c("edna_sample", "edna_assay", "edna_context"),
    lightyellow = "measurement_type",
    white       = "dataset"))
```

## Emit Core Tables

Projection lives here, in the notebook, per the no-`switch(dataset_key, ...)`
rule -- no `calcofi4db` change needed for this dataset.

```{r}
#| label: emit-core

build_grid_reference(con)

# 1. sample -- one row per physical water filter (parentEventID grain)
append_sample(con, glue("
  SELECT {ns_key(ds_key, 'sample', 'parent_event_id')} AS sample_key,
         'filter' AS sample_type, NULL::VARCHAR AS parent_sample_key,
         {ns_key(ds_key, 'sample', 'parent_event_id')} AS root_sample_key,
         '{ds_key}' AS dataset_key, grid_key, cruise_key,
         latitude, longitude, datetime_start_utc AS datetime,
         depth_min_m, depth_max_m
  FROM edna_sample"))

# 2. obs -- species detections, headline measurement = sequence_reads
# Positional order confirmed against calcofi_bottle/calcofi_dic/cce-lter_euphausiids
# (identical arg order in all three): realm, dataset_key, sample_key, grid_key,
# cruise_key, latitude, longitude, datetime, depth_min_m, depth_max_m, taxon_key,
# life_stage, measurement_type, measurement_value, measurement_qual,
# measurement_prec. life_stage is NULL for eDNA reads -- there is no life-stage
# field on a water-filter detection (occurrence.txt has no lifeStage column
# either); the source's identificationRemarks was a guess with no home in this
# schema and has been dropped.
append_obs(con, glue("
  SELECT 'bio', '{ds_key}', {ns_key(ds_key, 'sample', 'a.parentEventID')},
         s.grid_key, s.cruise_key, s.latitude, s.longitude, s.datetime_start_utc,
         s.depth_min_m, s.depth_max_m,
         dt.taxon_key, NULL::VARCHAR,
         'sequence_reads', CAST(a.organismQuantity AS DOUBLE),
         NULL::VARCHAR, NULL::DOUBLE
  FROM edna_assay a
  JOIN edna_sample s ON s.parent_event_id = a.parentEventID
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = a.scientificNameID"))

# 3. obs_attribute -- assay dimension per detection. Positional order confirmed
# against swfsc_ichthyo/calcofi_phyllosoma/cdfw_dungeness-crab/farallon_bird-mammal:
# dataset_key, sample_key, taxon_key, life_stage, measurement_type, bin_value,
# bin_label, count, measurement_qual. The assay label ('D-loop'/'12S (MiFish)')
# is categorical text, so it belongs in bin_label (position 7) -- confirmed
# against the real obs_attribute schema at calcofi.io/db-schema. Resolves Q02.
append_obs_attribute(con, glue("
  SELECT '{ds_key}', {ns_key(ds_key, 'sample', 'a.parentEventID')},
         dt.taxon_key, NULL::VARCHAR, 'assay',
         NULL::DOUBLE, a.assay, NULL::INTEGER, NULL::VARCHAR
  FROM edna_assay a
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = a.scientificNameID"))

# 4. sample_measurement -- co-collected water chemistry + genomics QC
append_sample_measurement(con, glue("
  SELECT {ns_key(ds_key, 'sample', 'parentEventID')}, '{ds_key}',
         measurement_type, measurement_value, NULL::VARCHAR
  FROM edna_context"))

core <- list(
  sample             = dbGetQuery(con, "SELECT COUNT(*) FROM sample WHERE dataset_key = 'calcofi_2022-edna'")[[1]],
  obs                = dbGetQuery(con, "SELECT COUNT(*) FROM obs WHERE dataset_key = 'calcofi_2022-edna'")[[1]],
  obs_attribute      = dbGetQuery(con, "SELECT COUNT(*) FROM obs_attribute WHERE dataset_key = 'calcofi_2022-edna'")[[1]],
  sample_measurement = dbGetQuery(con, "SELECT COUNT(*) FROM sample_measurement WHERE dataset_key = 'calcofi_2022-edna'")[[1]])
cat(glue(
  "core projection -- sample={core$sample} (expect 47) obs={core$obs} (expect 201) ",
  "obs_attribute={core$obs_attribute} (expect 201) ",
  "sample_measurement={core$sample_measurement} (expect up to 47*11)"), "\n")

n_unresolved_taxon <- dbGetQuery(con,
  "SELECT COUNT(*) FROM obs WHERE dataset_key = 'calcofi_2022-edna' AND taxon_key IS NULL")[[1]]
stopifnot("every eDNA detection must resolve a taxon_key" = n_unresolved_taxon == 0)
```

## Register New Measurement Types

```{r}
#| label: measurement-types

new_types <- c("sequence_reads", "oxygen_mg_l", "dna_concentration",
               "reads_raw_total", "reads_filtered_total", "otu_richness")
stopifnot("new measurement_type rows must already be in metadata/measurement_type.csv" =
            all(new_types %in% d_meas_type$measurement_type))

# extend _source_datasets on the 6 reused canonical types, without duplicating rows
reused_types <- c("ammonia", "nitrate", "nitrite", "phosphate", "silicate", "chl_fluor")
d_meas_type <- d_meas_type |>
  mutate(`_source_datasets` = if_else(
    measurement_type %in% reused_types &
      !str_detect(coalesce(`_source_datasets`, ""), ds_key),
    str_replace(str_squish(paste(coalesce(`_source_datasets`, ""), ds_key, sep = ";")), "^;", ""),
    `_source_datasets`))
write_csv(d_meas_type, meas_type_csv, na = "")
dbWriteTable(con, "measurement_type", d_meas_type, overwrite = TRUE)
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot("this dataset must appear in the ingest YAML registry" = nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)
```

## Questions for Data Providers

```{r}
#| label: provider-questions

questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the calcofi 2022-edna data providers (ranked)")
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
if (length(results$warnings) > 0)
  cat("Warnings:\n", paste("-", results$warnings, collapse = "\n"), "\n")
```

## Data Preview

```{r}
#| label: preview-sample
dbGetQuery(con, "SELECT * FROM sample WHERE dataset_key = 'calcofi_2022-edna'") |>
  datatable(caption = "sample", rownames = FALSE)
```

```{r}
#| label: preview-obs
dbGetQuery(con, "SELECT * FROM obs WHERE dataset_key = 'calcofi_2022-edna' LIMIT 50") |>
  datatable(caption = "obs (first 50)", rownames = FALSE)
```

```{r}
#| label: preview-sample-measurement
dbGetQuery(con, "SELECT * FROM sample_measurement WHERE dataset_key = 'calcofi_2022-edna'") |>
  datatable(caption = "sample_measurement", rownames = FALSE)
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
write_parquet_outputs(con, dir_parquet, tables = tbls_out)
```

## Write Metadata

```{r}
#| label: write-metadata

metadata_path <- build_metadata_json(
  con                  = con,
  d_tbls_rd            = d_tbls_rd,
  d_flds_rd            = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           glue("{dir_meta}/metadata_derived.csv")),
  output_dir           = dir_parquet,
  tables               = tbls_out,
  provider             = provider,
  dataset              = dataset,
  workflow_url         = cc$workflow_url,
  tables_owned         = tables_owned)

build_relationships_json(
  rels       = core_relationships(tbls_out),
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset)
```

## Upload to GCS

```{r}
#| label: upload-gcs

# local_dir/sidecar_dir split matches ingest_sio_mesopelagic-fish.qmd and
# ingest_calcofi_ctd-cast.qmd exactly -- both pass dir_stage as local_dir and
# dir_parquet as sidecar_dir here, even though every write above (parquet
# tables, metadata.json, relationships.json) targets dir_parquet.
sync_to_gcs(
  local_dir   = dir_stage,
  sidecar_dir = dir_parquet,
  gcs_prefix  = glue("ingest/{dir_label}"),
  bucket      = "calcofi-db")
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("parquet outputs written to: {dir_parquet}"), "\n")
```
