---
title: "Ingest SIO PIC Zooplankton Net Tows"
calcofi:
  target_name: ingest_pic_zooplankton
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/pic_zooplankton/manifest.json
  provider: pic
  dataset: zooplankton
  workflow_url: https://calcofi.io/workflows/ingest_pic_zooplankton.html
  questions_file: metadata/pic/zooplankton/questions.csv
  dataset_meta:
    dataset_name: SIO PIC Zooplankton Net Tows
    description: >
      Zooplankton net-tow sample registry from the SIO Pelagic Invertebrate
      Collection, filtered to the CalCOFI region. Position, time, depth, and gear
      metadata per tow. Biovolume measurements are not in the source file and are
      pending from the provider (see Questions for Data Providers).
    citation_main: ""
    link_calcofi_org: ""
    link_data_source: "SIO Pelagic Invertebrate Collection DB (CSV export)"
    link_others: []
    coverage_temporal: 1926 to present (CalCOFI-region subset)
    coverage_spatial: "CalCOFI grid (~23-51N, -135 to -117W incl. Baja)"
    license: ""
    pi_names: Rasmus Swalethorp; Ed Weber; Linsey Sala
  tables_owned:
    - {table: zooplankton_tow}
  erd:
    color: "#bbe0f0"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: `SIOPIC_DB_PNTtable_allRecords_9Feb2026.csv` (~19 MB, 148,129 net
tows), the parent-net-tow (PNT) registry of the SIO Pelagic Invertebrate
Collection (PI: Rasmus Swalethorp, Ed Weber, Linsey Sala).

- **Provider**: `pic`
- **Grain**: one row per zooplankton net tow (sample metadata only)
- **Scope decision**: ingest tows within the **CalCOFI region** bounding box
  (~23–51°N, −135→−117°W incl. Baja); this also drops the antimeridian
  coordinate errors (Q04).
- **Biovolume is NOT in this file** — only the tow/sample registry is built here
  as `zooplankton_tow`; the biovolume measurement table is pending from the
  provider (blocker Q01) and will be added as `zooplankton_measurement`.

```{mermaid}
graph LR
  A[PNTtable.csv<br/>148,129 tows] --> B[filter CalCOFI bbox]
  B --> C[zooplankton_tow<br/>position + gear + keys]
  C -.ship/cruise/grid.-> D[(shared refs)]
  C -.biovolume pending.-> E[zooplankton_measurement]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))

librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_pic_zooplankton.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
dataset_name <- cc$dataset_meta$dataset_name
tables_owned <- cc$tables_owned
dir_label    <- glue("{provider}_{dataset}")

dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
  if (dir_exists(paste0(db_path, ".tmp")))  dir_delete(paste0(db_path, ".tmp"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

Read directly from `{dir_data}/calcofi/zooplankton/` (the file is not under a
`pic/` subdirectory). Columns renamed per
`metadata/pic/zooplankton/flds_redefine.csv`.

```{r}
#| label: read-source

zoo_csv <- path_expand(glue(
  "{dir_data}/calcofi/zooplankton/SIOPIC_DB_PNTtable_allRecords_9Feb2026.csv"))
stopifnot("zooplankton CSV not found" = file_exists(zoo_csv))

sync_to_gcs(
  local_dir  = path_dir(zoo_csv),
  gcs_prefix = glue("archive/{provider}/{dataset}"),
  bucket     = "calcofi-files-public",
  exclude    = c(".DS_Store", "*.tmp", "*.gdoc"))

d_raw <- read_csv(zoo_csv, col_types = cols(.default = "c"))  # read all as text, cast below
cat(glue("Read {format(nrow(d_raw), big.mark=',')} tows, {ncol(d_raw)} columns"), "\n")
```

## Clean, Type-Cast, and Filter to CalCOFI Region

Canonical names from `metadata/field_dictionary.csv`. Tows are filtered to the
CalCOFI bounding box (the scope decision); this also excludes the |lon|>180
coordinate errors (Q04). Datetimes are built from date (MM/DD/YYYY) + HHMM time
(treated as given pending the timezone answer, Q03).

```{r}
#| label: clean-filter

hhmm_to_dt <- function(date, hhmm) {
  t <- str_pad(na_if(str_trim(hhmm), ""), 4, pad = "0")
  out <- suppressWarnings(as_datetime(paste0(
    date, " ", substr(t, 1, 2), ":", substr(t, 3, 4), ":00")))
  out[is.na(t) & !is.na(date)] <- as_datetime(as.character(date[is.na(t) & !is.na(date)]))
  out
}

d_clean <- d_raw |>
  transmute(
    expedition      = EXPEDITION_pnt,
    expedition_code = EXPED_CODE_pnt,
    expedition_type = Expedition_Type_pnt,
    ship_name       = SHIP_pnt,
    order_occ       = suppressWarnings(as.integer(SWFOrder_Occ)),
    line            = suppressWarnings(as.numeric(STATION_LINE_pnt)),
    station         = suppressWarnings(as.numeric(STATION_NUMBER_pnt)),
    latitude        = suppressWarnings(as.numeric(LAT_DECIMAL_pnt)),
    longitude       = suppressWarnings(as.numeric(LONG_DECIMAL_pnt)),
    date            = suppressWarnings(as.Date(SAMPLE_DATE_pnt, format = "%m/%d/%Y")),
    datetime_start_utc = hhmm_to_dt(date, START_TIME_pnt),  # Q03: assumed UTC
    datetime_end_utc   = hhmm_to_dt(date, END_TIME_pnt),
    depth_min_m     = suppressWarnings(as.numeric(DEPTH_MIN_pnt)),
    depth_max_m     = suppressWarnings(as.numeric(DEPTH_MAX_pnt)),
    max_wire_out_m  = suppressWarnings(as.numeric(MAX_MWO_pnt)),
    net_type        = NET_TYPE_pnt,
    mesh_size_mm    = suppressWarnings(as.numeric(MESH_SIZE_pnt)),
    tow_type        = TOW_TYPE_pnt,
    fixative        = FIXATIVE_pnt,
    preservative    = PRESERVATIVE_pnt)

n_all <- nrow(d_clean)
# scope filter: CalCOFI region bounding box (incl. Baja); drops global + bad coords
d_cc <- d_clean |>
  filter(!is.na(latitude), !is.na(longitude),
         latitude  >= 23, latitude  <= 51,
         longitude >= -135, longitude <= -117) |>
  mutate(
    site_key = if_else(
      is.na(line) | is.na(station), NA_character_,
      sprintf("%05.1f %05.1f", line, station)),
    tow_id = row_number())

cat(glue(
  "CalCOFI-region filter: {format(nrow(d_cc), big.mark=',')}/",
  "{format(n_all, big.mark=',')} tows retained ",
  "({round(100*nrow(d_cc)/n_all, 1)}%)"), "\n")

dbWriteTable(con, "zooplankton_tow", d_cc, overwrite = TRUE)
```

## Resolve Ship and Cruise Keys

```{r}
#| label: resolve-keys

load_prior_tables(
  con, parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables = c("ship", "cruise", "grid"), geom_tables = c("grid"), as_view = TRUE)

d_ship <- dbGetQuery(con, "SELECT ship_key, ship_name, ship_nodc FROM ship") |>
  mutate(ship_name_norm = ship_name |> str_to_upper() |> str_squish())

zt <- dbGetQuery(con, "SELECT tow_id, ship_name, date FROM zooplankton_tow") |>
  mutate(ship_name_norm = ship_name |>
           str_replace("^R/?V\\.?\\s+", "") |> str_to_upper() |> str_squish()) |>
  left_join(d_ship |> select(ship_name_norm, ship_key, ship_nodc), by = "ship_name_norm") |>
  mutate(cruise_key = if_else(
    is.na(ship_nodc) | is.na(date), NA_character_,
    as.character(glue("{format(date, '%Y-%m')}-{ship_nodc}"))))

valid_ck <- dbGetQuery(con, "SELECT DISTINCT cruise_key FROM cruise")$cruise_key
zt <- zt |> mutate(cruise_key = if_else(cruise_key %in% valid_ck, cruise_key, NA_character_))

# write resolved keys back
dbWriteTable(con, "zt_keys", zt |> select(tow_id, ship_key, cruise_key), overwrite = TRUE)
dbExecute(con, "ALTER TABLE zooplankton_tow ADD COLUMN IF NOT EXISTS ship_key VARCHAR")
dbExecute(con, "ALTER TABLE zooplankton_tow ADD COLUMN IF NOT EXISTS cruise_key VARCHAR")
dbExecute(con,
  "UPDATE zooplankton_tow t SET ship_key = k.ship_key, cruise_key = k.cruise_key
   FROM zt_keys k WHERE t.tow_id = k.tow_id")
dbExecute(con, "DROP TABLE zt_keys")

cat(glue(
  "ship match: {sum(!is.na(zt$ship_key))}/{nrow(zt)} ",
  "({round(100*mean(!is.na(zt$ship_key)),1)}%); ",
  "cruise_key match: {sum(!is.na(zt$cruise_key))}/{nrow(zt)} ",
  "({round(100*mean(!is.na(zt$cruise_key)),1)}%)"), "\n")
```

## Add Spatial

```{r}
#| label: spatial

add_point_geom(con, "zooplankton_tow", lon_col = "longitude", lat_col = "latitude")
assign_grid_key(con, "zooplankton_tow") |> datatable(caption = "Grid assignment")
```

## Load Dataset Metadata

```{r}
#| label: load-dataset-metadata

d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here()))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Schema Documentation

```{r}
#| label: schema

zoo_rels <- list(
  primary_keys = list(zooplankton_tow = "tow_id"),
  foreign_keys = list())  # cross-dataset FKs (cruise/ship/grid) in relationships_cross.csv

cc_erd(
  con, tables = c("zooplankton_tow", "dataset"), rels = zoo_rels,
  colors = list(lightblue = "zooplankton_tow", white = "dataset"))

build_relationships_json(
  rels = zoo_rels, output_dir = dir_parquet, provider = provider, dataset = dataset)
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
# NULL cruise_key/ship_key/site_key are the expected unmatched remainder (Q03/Q05),
# the same accepted behavior as calcofi_dic; not a hard failure.

n_dup <- dbGetQuery(con,
  "SELECT COUNT(*) FROM (SELECT tow_id, COUNT(*) n FROM zooplankton_tow GROUP BY tow_id HAVING COUNT(*)>1)")[[1]]
cat(glue("zooplankton_tow tow_id duplicates: {n_dup}"), "\n")
```

## Data Preview

```{r}
#| label: preview

cols <- dbGetQuery(con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name='zooplankton_tow' AND data_type NOT LIKE 'GEOMETRY%'")$column_name
dbGetQuery(con, glue("SELECT {paste(cols, collapse=', ')} FROM zooplankton_tow LIMIT 100")) |>
  datatable(caption = "zooplankton_tow — first 100 rows", rownames = FALSE, filter = "top")
```

## Emit Core Tables

Project this dataset into the shared consolidated core model
(`design_env-bio-consolidation.md`) via `calcofi4db::emit_core_tables()` — the
single source of truth for the per-dataset projection into `sample` / `obs` /
`obs_freq` / `sample_measurement`, also used by `release_database.qmd` to
assemble the authoritative cross-dataset release.

```{r}
#| label: emit_core

core <- emit_core_tables(con, "pic_zooplankton")
cat(glue(
  "core projection — sample={core$sample %||% 0} obs={core$obs %||% 0} ",
  "obs_freq={core$obs_freq %||% 0} sample_measurement={core$sample_measurement %||% 0}\n"))

# the frozen release assembles the authoritative cross-dataset core; keep this
# ingest's parquet output per-dataset by dropping the local core projection here.
for (tbl_core in c("obs", "obs_freq", "sample_measurement", "sample"))
  dbExecute(con, glue("DROP TABLE IF EXISTS {tbl_core}"))
```


## Write Parquet Outputs

```{r}
#| label: write-parquet

dir_create(dir_parquet)
mismatches <- list(cruise_keys = collect_cruise_key_mismatches(con, "zooplankton_tow"))
parquet_stats <- write_parquet_outputs(
  con = con, output_dir = dir_parquet,
  tables = c("zooplankton_tow", "dataset"),
  strip_provenance = FALSE, mismatches = mismatches)
parquet_stats |> mutate(file = basename(path)) |> select(-path) |>
  datatable(caption = "Parquet export statistics")
```

## Write Metadata

```{r}
#| label: write-metadata

d_tbls_rd <- read_csv(here("metadata/pic/zooplankton/tbls_redefine.csv"))
d_flds_rd <- read_csv(here("metadata/pic/zooplankton/flds_redefine.csv"))
metadata_path <- build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = here("metadata/pic/zooplankton/metadata_derived.csv"),
  output_dir = dir_parquet, tables = c("zooplankton_tow"),
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned)
```

## Upload to GCS

```{r}
#| label: upload-gcs

sync_to_gcs(local_dir = dir_parquet, gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
```

## Questions for Data Providers

Follow-up questions for SIO PIC (Rasmus Swalethorp, Ed Weber, Linsey Sala),
**ranked by importance**. The headline blocker (Q01) is that biovolume is absent
from the provided file, so only the tow registry is ingested here. Tracked in
`metadata/pic/zooplankton/questions.csv`; aggregated by `questions_email.qmd`.

```{r}
#| label: provider-questions

read_csv(here(glue("metadata/{provider}/{dataset}/questions.csv"))) |>
  arrange(factor(priority, c("blocker", "high", "normal")), id) |>
  select(priority, question, context, related_field, status) |>
  datatable(caption = "Questions for data providers (ranked)",
            options = list(dom = "t", pageLength = 20), rownames = FALSE)
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
