---
title: "Ingest SIO Cetacean Sightings (CalCOFI visual surveys)"
calcofi:
  target_name: ingest_sio_cetacean_sightings
  workflow_type: ingest
  dependency:
    - ingest_swfsc_ichthyo
  output: data/parquet/sio_cetacean-sightings/manifest.json
  provider: sio
  dataset: cetacean-sightings
  # staged, not released, until the licence (Q01) and citation (Q02) are settled
  in_release: false
  workflow_url: https://calcofi.io/workflows/ingest_sio_cetacean-sightings.html
  questions_file: metadata/sio/cetacean-sightings/questions.csv
  dataset_meta:
    dataset_name: CalCOFI Cetacean Visual Line-Transect Sightings
    dataset_name_short: "Cetacean sightings"
    category: "Seabirds & Marine Mammals"
    color: "#5b7fa6"
  # publishes the consolidated core: the effort / sighting / species source shape
  # is wrangled here and projected into sample / obs / obs_attribute /
  # sample_measurement + the taxa refs.
  tables_owned:
    - {table: sample, shared: true, note: "core event dimension (transect segment + sighting)"}
    - {table: obs, shared: true, note: "core occurrence headline (group_size per sighting)"}
    - {table: obs_attribute, shared: true, note: "group size min/max, calves, behaviour per sighting"}
    - {table: sample_measurement, shared: true, note: "transect_length per effort segment"}
    - {table: taxon, shared: true, note: "shared taxa reference (cetaceans, worms:)"}
    - {table: dataset_taxon, shared: true, note: "species_code -> taxon_key crosswalk"}
    - {table: taxon_group, shared: true, note: "calcofi:marine_mammals"}
    - {table: measurement_type, shared: true, note: "shared registry across datasets"}
  erd:
    color: "#d6e2f0"
editor_options:
  chunk_output_type: console
---

## Overview

**Source**: the curated cetacean visual-survey files in
[CalCOFI/marmam-app](https://github.com/CalCOFI/marmam-app) (`data/whale_visual_data/`),
read from a **pinned commit** so a run reads bytes it can name:

- `CalCOFI_2004-2022_CombinedSightings.csv` — one row per sighting, 67 cruises
  2004-07 → 2022-11 (6,107 rows).
- `CalCOFI_2004-2021_Effort_OnTransectOnEffortONLY_MNA.csv` — on-effort,
  on-transect survey segments, 2004 → 2021 (1,988 rows, 8 of them placeholder
  rows for cruises with no effort).
- `data/CalCOFI_species_codes.csv` — the 34 species codes.

Both CSVs are **Latin-1**, not UTF-8 (a `°` in a comment breaks a UTF-8 read).
The app's derived `whale.csv` / `transect.csv` are **not** read: they drop rows
and rename cruises for display. The raw CountCOFI event log for 2023-11
(`new-marine-mammal-data/CC-202311.txt`) is not ingested either (Q08).

This is the dataset the catalog held as the `sio_cetacean-sightings` holding
(EDI `knb-lter-cce.262`, 2005–2015); the marmam-app files are the lab's wider,
curated 2004–2022 version of it (Q02).

- **Provider**: `sio` (Scripps Whale Acoustics Lab, Hildebrand / Baumann-Pickering)
- **Grain**: a sighting, optionally inside an effort segment (`segment_id`, NULL
  when off effort / off transect)
- **Taxonomy**: species codes declared with `append_dataset_taxon()` from
  `metadata/sio/cetacean-sightings/species_codes.csv` and resolved against WoRMS;
  the unidentified classes and the two synonyms are keyed by
  `metadata/taxon_override.csv`.

```{mermaid}
graph LR
  E[Effort CSV] --> SG[cetacean_segment<br/>effort segment]
  S[Sightings CSV] --> SI[cetacean_sighting<br/>one row per sighting]
  C[species_codes] --> SP[cetacean_species]
  SI -.segment_id.-> SG
  SI -.species_code.-> SP
```

## Setup

```{r}
#| label: setup
#| message: false

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db, CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, here, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr, tibble, tidyr, quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))
source(here("libs/ingest.R"))

cc           <- read_calcofi_meta(here("ingest_sio_cetacean-sightings.qmd"))
provider     <- cc$provider
dataset      <- cc$dataset
tables_owned <- cc$tables_owned
ds_key       <- glue("{provider}_{dataset}")
dir_label    <- ds_key
dir_parquet  <- here(glue("data/parquet/{dir_label}"))
dir_stage    <- cc_stage_path("parquet", dir_label, create = TRUE)
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))
dir_meta     <- here(glue("metadata/{provider}/{dataset}"))
# downloads land here first (gitignored; archived to GCS below); a re-run reuses
# them, `overwrite_all` (libs/ingest.R) refetches
dir_dl       <- here(glue("data/cache/{dir_label}")); dir_create(dir_dl)
# GCS uploads (source archive + parquet mirror). Set CALCOFI_SKIP_GCS=1 to render
# without publishing -- a review render, or an account without write access to
# gs://calcofi-files-public / gs://calcofi-db.
publish_to_gcs <- Sys.getenv("CALCOFI_SKIP_GCS") == ""

if (overwrite) {
  if (file_exists(db_path))                 file_delete(db_path)
  if (file_exists(paste0(db_path, ".wal"))) file_delete(paste0(db_path, ".wal"))
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

```{r}
#| label: read-source

# CalCOFI/marmam-app at a pinned commit: the repository is a Shiny app whose
# data/ folder is edited by hand, so `main` is a moving target. Bump the SHA
# deliberately (and re-check the counts asserted below) to take new data.
src_sha  <- "7bc0e18d7cb0ac7ba58d7bc4d47e57c50bd9bc80"
src_base <- glue("https://raw.githubusercontent.com/CalCOFI/marmam-app/{src_sha}/data")
src_paths <- c(
  sightings = "whale_visual_data/CalCOFI_2004-2022_CombinedSightings.csv",
  effort    = "whale_visual_data/CalCOFI_2004-2021_Effort_OnTransectOnEffortONLY_MNA.csv",
  species   = "CalCOFI_species_codes.csv")
src_urls  <- setNames(as.character(glue("{src_base}/{src_paths}")), names(src_paths))
src_files <- setNames(file.path(dir_dl, basename(src_paths)), names(src_paths))
downloaded <- character()
for (k in names(src_paths)) {
  if (overwrite_all || !file_exists(src_files[[k]])) {
    download.file(src_urls[[k]], src_files[[k]], quiet = TRUE, mode = "wb")
    downloaded <- c(downloaded, k)
  }
}
# when the bytes came down (the cached file's mtime), recorded against the URL
src_stamp <- stamp_source_access(files = unname(src_files)) |>
  mutate(source = unname(src_urls),
         method = ifelse(names(src_files) %in% downloaded, "download", "file_mtime"))
cat(glue("marmam-app@{substr(src_sha, 1, 7)}: {length(downloaded)} file(s) downloaded, ",
         "{length(src_paths) - length(downloaded)} reused from {dir_dl}"), "\n")

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_dl, gcs_prefix = glue("archive/{provider}/{dataset}"),
              bucket = "calcofi-files-public")
} else {
  cat("CALCOFI_SKIP_GCS set -- sources NOT archived to gs://calcofi-files-public\n")
}

# Latin-1: a degree sign in a comment is byte 0xB0, which a UTF-8 read rejects
latin1 <- locale(encoding = "latin1")
d_si_raw <- read_csv(src_files[["sightings"]], col_types = cols(.default = "c"), locale = latin1)
d_ef_raw <- read_csv(src_files[["effort"]],    col_types = cols(.default = "c"), locale = latin1)
d_sp_raw <- read_csv(src_files[["species"]],   col_types = cols(.default = "c"))

# the counts profiled at this commit (2026-09-27); a changed file must be looked
# at, not absorbed
stopifnot(
  "sightings row count changed at this commit" = nrow(d_si_raw) == 6107L,
  "effort row count changed at this commit"    = nrow(d_ef_raw) == 1988L,
  "species row count changed at this commit"   = nrow(d_sp_raw) == 34L)
cat(glue("sightings {nrow(d_si_raw)}, effort rows {nrow(d_ef_raw)}, species codes {nrow(d_sp_raw)}"), "\n")
```

## Build Species Vocabulary

```{r}
#| label: species

# the source list gives code, common name and suborder; the scientific name each
# code resolves from is authored once in species_codes.csv (unidentified classes
# name their rank's taxon; Oth names none and is excluded below)
d_species_meta <- read_csv(file.path(dir_meta, "species_codes.csv"), col_types = cols(.default = "c"))
d_species <- d_sp_raw |>
  transmute(species_code = speciesCode, common_name = str_trim(speciesName),
            suborder = taxonomy) |>
  left_join(d_species_meta |> select(species_code, scientific_name), by = "species_code")
stopifnot(
  "species_code must be unique" = !anyDuplicated(d_species$species_code),
  "every source code must be in species_codes.csv" =
    setequal(d_species$species_code, d_species_meta$species_code))
dbWriteTable(con, "cetacean_species", d_species, overwrite = TRUE)
cat(glue("cetacean_species: {nrow(d_species)} codes, ",
         "{sum(is.na(d_species$scientific_name))} without a taxon (Oth)"), "\n")
```

## Build Effort Segment Table

```{r}
#| label: segment

# local clock time -> UTC (Q04): America/Los_Angeles, so daylight saving applies
local_to_utc <- function(x)
  with_tz(suppressWarnings(mdy_hm(x, tz = "America/Los_Angeles")), "UTC")
num <- function(x) suppressWarnings(as.numeric(x))

# the eight rows with no Segment.ID are per-cruise placeholders (length 0, no
# position): a cruise listed with no on-effort survey. Kept as a list, not a
# segment (Q07).
d_no_effort <- d_ef_raw |> filter(is.na(Segment.ID)) |> distinct(Cruise)
d_segment <- d_ef_raw |>
  filter(!is.na(Segment.ID)) |>
  transmute(
    segment_id           = as.integer(num(Segment.ID)),
    cruise_label_effort  = Cruise,
    # CC2004-07 -> CC0407, the label the sightings carry
    cruise_label         = str_replace(Cruise, "^CC\\d{2}(\\d{2})-(\\d{2})$", "CC\\1\\2"),
    season               = str_trim(Season),
    datetime_start_local = StartDateTimeLocal,
    datetime_end_local   = EndDateTimeLocal,
    datetime_start_utc   = local_to_utc(StartDateTimeLocal),
    datetime_end_utc     = local_to_utc(EndDateTimeLocal),
    latitude_start       = num(StartLat),  longitude_start = num(StartLon),
    latitude_end         = num(EndLat),    longitude_end   = num(EndLon),
    # the core carries one point per event: the segment midpoint
    latitude             = (latitude_start  + latitude_end)  / 2,
    longitude            = (longitude_start + longitude_end) / 2,
    comment              = Comment.ST,
    analyst_comment      = AG.Comment,
    station_from         = From.Station,
    station_to           = To.Station,
    segment_length_km    = num(Segment.length..km.),
    transect_number      = as.integer(num(Transect)),
    transect_length_km   = num(Transect.Length),
    line                 = num(ccLine),
    line_length_km       = num(ccLineLength))

stopifnot(
  "segment_id must be unique"            = !anyDuplicated(d_segment$segment_id),
  "every segment needs a start time"     = !anyNA(d_segment$datetime_start_utc),
  "every segment needs a midpoint"       = !anyNA(d_segment$latitude) && !anyNA(d_segment$longitude),
  "cruise_label must parse to CCYYMM"    = all(str_detect(d_segment$cruise_label, "^CC\\d{4}$")))
dbWriteTable(con, "cetacean_segment", d_segment, overwrite = TRUE)
cat(glue("cetacean_segment: {nrow(d_segment)} segments over ",
         "{n_distinct(d_segment$cruise_label)} cruises; ",
         "{nrow(d_no_effort)} cruises listed with no effort: ",
         "{paste(d_no_effort$Cruise, collapse = ', ')}"), "\n")
```

## Build Sighting Table

```{r}
#| label: sighting

d_sighting <- d_si_raw |>
  mutate(sighting_id = row_number()) |>
  transmute(
    sighting_id,
    cruise_label          = Cruise,
    season                = str_trim(Season),
    datetime_local        = DateTimeLocal,
    # '#REF!' on 36 CC0501 rows parses to NA (Q06)
    datetime_start_utc    = local_to_utc(DateTimeLocal),
    latitude              = num(DecLat),
    longitude             = num(DecLong),
    segment_id            = as.integer(num(`segID..NaN..Off.effort.and.or.off.transect..AG.`)),
    observer_port         = PortObs,
    observer_starboard    = StbdObs,
    group_size_best       = num(Best),
    group_size_min        = num(Min),
    group_size_max        = num(Max),
    calf_raw              = Calf,
    # plain integers only; ND / nd / 0NA / UD / O / 1? / Juv stay NULL (Q09)
    calf_count            = if_else(str_detect(str_trim(Calf), "^\\d+$"),
                                    as.integer(str_trim(Calf)), NA_integer_),
    species_code          = str_trim(Species1),
    # only ever "100", on CC2111: not a second species (Q05)
    species2_raw          = Species2,
    effort_flag_original  = Original.Effort.Flag..ON.OFF.,
    off_effort_code       = OffEffortCode,
    transect_flag_original = OriginalTransect.ON.OFF.,
    off_transect_code     = OffTransectCode,
    effort_status         = str_to_upper(str_trim(Adjusted.both.On.Effort.and.On.Transect)),
    behavior_primary      = PrimaryBehaviour,
    behavior_other1       = OtherBehaviour1,
    behavior_other2       = OtherBehaviour2,
    comments              = Comments,
    speed_over_ground_kn  = num(SOG..knots.),
    analyst_comment       = AGComments)

# one position is on land (Sonora, 32.31 N -112.70 E; the app drops the row):
# keep the sighting, drop the position (Q11)
on_land <- with(d_sighting, !is.na(longitude) & longitude > -114)
d_sighting <- d_sighting |>
  mutate(latitude  = if_else(on_land, NA_real_, latitude),
         longitude = if_else(on_land, NA_real_, longitude))

n_ref <- sum(d_si_raw$DateTimeLocal == "#REF!", na.rm = TRUE)
stopifnot(
  "only the '#REF!' rows may lack a UTC time (Q06)" =
    sum(is.na(d_sighting$datetime_start_utc)) == n_ref,
  "exactly one position is on land (Q11)" = sum(on_land) == 1L,
  "effort_status is ON, OFF or NULL" =
    all(d_sighting$effort_status %in% c("ON", "OFF", NA)),
  "every species_code is in the vocabulary" =
    all(d_sighting$species_code %in% d_species$species_code),
  "Species2 is never a species code (Q05)" =
    all(d_sighting$species2_raw %in% c("100", NA)))
dbWriteTable(con, "cetacean_sighting", d_sighting, overwrite = TRUE)

# every segment_id a sighting carries must exist in the effort table
n_seg_orphan <- dbGetQuery(con, "
  SELECT COUNT(*) FROM cetacean_sighting s
  LEFT JOIN cetacean_segment g USING (segment_id)
  WHERE s.segment_id IS NOT NULL AND g.segment_id IS NULL")[[1]]
cat(glue("cetacean_sighting: {nrow(d_sighting)} rows; {sum(!is.na(d_sighting$segment_id))} on a segment ",
         "({n_seg_orphan} segment_id(s) not in the effort file); ",
         "{n_ref} '#REF!' times; {sum(!is.na(d_sighting$calf_count))} with a calf count"), "\n")
```

One sighting references a segment id (236) that is absent from the effort file
(99.96% resolve). Those sightings keep their `segment_id` in the source table
but get no parent `sample`, so the core never points at a sample that does not
exist.

## Resolve cruise_key + Spatial

The sightings carry a **survey label** (`CC0407`), not a cruise, and no ship. As
for `farallon_bird-mammal`, `match_cruise_by_track()` matches each sighting to
the nearest CalCOFI station occupied within a day (the `swfsc_ichthyo` sample
track, whose `cruise_key`s are exactly the `cruise` reference's) and takes the
**majority vote per label**. The effort segments take their label's winner.

```{r}
#| label: spatial-cruise

load_prior_tables(con, parquet_dir = cc_stage_path("parquet", "swfsc_ichthyo"),
                  tables = c("grid"), geom_tables = c("grid"), as_view = TRUE)
for (tbl in c("cetacean_segment", "cetacean_sighting")) {
  add_point_geom(con, tbl, lon_col = "longitude", lat_col = "latitude")
  assign_grid_key(con, tbl)
}

dbExecute(con, glue(
  "CREATE OR REPLACE VIEW cruise_track AS
   SELECT DISTINCT cruise_key, datetime, latitude, longitude
   FROM read_parquet('{cc_stage_path('parquet','swfsc_ichthyo','sample.parquet')}')
   WHERE cruise_key IS NOT NULL AND datetime IS NOT NULL"))

# window_days = 1: the sighting time is converted from Pacific clock time, so an
# evening sighting falls on the next UTC day
cruise_match <- match_cruise_by_track(
  con, "cetacean_sighting", "cruise_track",
  datetime_col = "datetime_start_utc", lon_col = "longitude", lat_col = "latitude",
  ref_datetime_col = "datetime", group_col = "cruise_label", window_days = 1)

# the segments take the winner of their label; a label with no sightings (a
# cruise with effort but nothing seen) stays NULL and is counted below
dbExecute(con, "ALTER TABLE cetacean_segment ADD COLUMN IF NOT EXISTS cruise_key VARCHAR")
dbExecute(con, "
  UPDATE cetacean_segment g SET cruise_key = (
    SELECT ANY_VALUE(s.cruise_key) FROM cetacean_sighting s
    WHERE s.cruise_label = g.cruise_label AND s.cruise_key IS NOT NULL)")

cat(glue("cruise_key resolved on {cruise_match$matched}/{cruise_match$total} ",
         "sightings ({cruise_match$pct}%)"), "\n")
cruise_match$groups |>
  datatable(caption = "Cruise label -> cruise_key (vote share; NULL = unresolved)",
            rownames = FALSE, filter = "top") |>
  formatPercentage("share", 1)
```

```{r}
#| label: cruise-match-validate

grp  <- cruise_match$groups
dupe <- grp$cruise_key[!is.na(grp$cruise_key)]
dupe <- dupe[duplicated(dupe)]
n_seg_null <- dbGetQuery(con, "SELECT COUNT(*) FROM cetacean_segment WHERE cruise_key IS NULL")[[1]]
stopifnot(
  "cruise_key must resolve on >90% of sightings" = cruise_match$pct > 90,
  "every cruise label must resolve to a distinct cruise_key" = length(dupe) == 0,
  "a resolved cruise_key must exist in the track (no dangling FK)" =
    dbGetQuery(con, "
      SELECT COUNT(*) FROM cetacean_sighting WHERE cruise_key IS NOT NULL
        AND cruise_key NOT IN (SELECT cruise_key FROM cruise_track)")[[1]] == 0)
cat(glue(
  "{sum(!is.na(grp$cruise_key))}/{nrow(grp)} labels resolved; unresolved: ",
  "{paste(grp$cruise_label[is.na(grp$cruise_key)], collapse = ', ')}; ",
  "{n_seg_null} segment(s) without a cruise_key"), "\n")
```

## Load Dataset Metadata + Schema

```{r}
#| label: meta-schema

this_provider <- provider; this_dataset <- dataset
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot(nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)

# the registry this dataset's types were added to (metadata/measurement_type.csv,
# 2026-09-27): read, never re-derived, and asserted complete
d_meas_type <- read_measurement_type(here("metadata/measurement_type.csv"))
my_types <- c("group_size", "group_size_min", "group_size_max", "calf_count",
              "behavior", "transect_length")
stopifnot("every type this ingest emits is registered" =
            all(my_types %in% d_meas_type$measurement_type))
dbWriteTable(con, "measurement_type", as.data.frame(d_meas_type), overwrite = TRUE)

# the SOURCE shape; the published shape is the core (Emit Core Tables)
src_rels <- list(
  primary_keys = list(cetacean_segment = "segment_id", cetacean_sighting = "sighting_id",
                      cetacean_species = "species_code"),
  foreign_keys = list(
    list(table = "cetacean_sighting", column = "segment_id",
         ref_table = "cetacean_segment", ref_column = "segment_id"),
    list(table = "cetacean_sighting", column = "species_code",
         ref_table = "cetacean_species", ref_column = "species_code")))
cc_erd(con, tables = c("cetacean_segment", "cetacean_sighting", "cetacean_species", "dataset"),
       rels = src_rels,
       colors = list(lightblue = c("cetacean_segment", "cetacean_sighting"),
                     lightyellow = "cetacean_species", white = "dataset"))
```

## Questions for Data Providers

```{r}
#| label: provider-questions

questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the SIO cetacean sightings data providers (ranked)")
```

## Emit Core Tables

What the source becomes, and what it does not:

| source | core | transform |
|---|---|---|
| `cetacean_segment` row | `sample` (`transect`) | key `sio_cetacean-sightings:transect:<segment_id>`; midpoint position; start time |
| `cetacean_segment.segment_length_km` | `sample_measurement` `transect_length` | km |
| `cetacean_sighting` row | `sample` (`sighting`) | parent = its transect when `segment_id` resolves, else none |
| `cetacean_sighting.group_size_best` | `obs` `group_size` | `measurement_qual` = `effort_status` (ON/OFF) |
| `group_size_min` / `_max` | `obs_attribute` `group_size_min` / `_max` | value in `count` |
| `calf_count` | `obs_attribute` `calf_count`, `life_stage = calf` | plain integers only (Q09) |
| `behavior_primary` / `_other1` / `_other2` | `obs_attribute` `behavior` | one row per code token, `bin_label` = the code (codebook Q03), `bin_value` = 1 / 2 / 3 for primary / other1 / other2 |
| `species_code` | `obs.taxon_key` via `dataset_taxon` | `Oth` keeps a NULL taxon |

**Not published** (kept in the archived source): observer initials, original
effort/transect flags and their off-codes, comments, speed over ground,
`species2_raw`, `calf_raw`, segment start/end coordinates and times, from/to
stations, transect and line lengths.

### Taxon vocabulary

```{r}
#| label: taxon-stage

mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

obs_codes      <- unique(d_sighting$species_code)
# not a taxon: "Other"; its sightings keep a NULL taxon_key
excluded_codes <- intersect(obs_codes, d_species$species_code[is.na(d_species$scientific_name)])
stopifnot("Oth is the only code without a taxon" = identical(excluded_codes, "Oth"))

# the declaration: code, the name the code is resolved from, and the common name.
# The source supplied no ids, so none ride along as ds_source_json.
# only the codes the sightings use: an unused code (Be) would be resolved for nothing
d_vocab <- d_species |>
  filter(!is.na(scientific_name), species_code %in% obs_codes) |>
  transmute(ds_taxa_code = species_code, ds_scientific_name = scientific_name,
            ds_common_name = common_name)
n_staged <- append_dataset_taxon(con, ds_key, d_vocab)

ensure_taxon_xref(con, mt_taxon, tx_over, cache_csv = here("metadata/taxon_xref.csv"))
ensure_taxon_lineage(con, mt_taxon, tx_over, cache_csv = here("metadata/taxon_lineage.csv"))
n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
n_ds_taxon <- resolve_dataset_taxon(con, mt_taxon, tx_over)
n_tx_group <- build_taxon_group(con, read_taxon_group_rules(here("metadata/taxon_group.csv")))

chk <- check_dataset_taxon(con, ds_key, codes = setdiff(obs_codes, excluded_codes))

# every staged code keys a Cetacea taxon (the vocabulary is cetaceans only)
dt <- dbGetQuery(con, "
  SELECT d.ds_taxa_code, d.taxon_key, t.scientific_name, t.rank
  FROM dataset_taxon d LEFT JOIN taxon t USING (taxon_key)
  WHERE d.dataset_key = ?", params = list(ds_key))
key_of <- function(code) dt$taxon_key[dt$ds_taxa_code == code]
stopifnot(
  "every staged code has a taxon_key"   = !anyNA(dt$taxon_key),
  "Beba and Bbo are one taxon"          = identical(key_of("Beba"), key_of("Bbo")),
  "Dc keys Delphinus delphis (Q10)"     = identical(key_of("Dc"), "worms:137094"),
  "Lo keys Sagmatias obliquidens (Q10)" = identical(key_of("Lo"), "worms:1571909"))
cat(glue("vocabulary: {n_staged} codes staged, {length(excluded_codes)} excluded (Oth); ",
         "taxon {n_taxon}, dataset_taxon {n_ds_taxon}, taxon_group {n_tx_group}"), "\n")
dt |> datatable(caption = "species_code -> taxon", rownames = FALSE, filter = "top")
```

### Sample, obs, obs_attribute, sample_measurement

```{r}
#| label: emit_core

tr_key <- function(col) ns_key(ds_key, "transect", col)
si_key <- function(col) ns_key(ds_key, "sighting", col)

# transects: one sample per effort segment
append_sample(con, sample_arm_self(ds_key, "cetacean_segment", "segment_id", "transect"))

# sightings: one sample per sighting, the child of its transect when on one. The
# parent is looked up, not assumed: a segment_id missing from the effort file
# gets no parent rather than a dangling one.
append_sample(con, glue("
  SELECT {si_key('s.sighting_id')} AS sample_key, 'sighting' AS sample_type,
         CASE WHEN g.segment_id IS NOT NULL THEN {tr_key('g.segment_id')} END AS parent_sample_key,
         COALESCE(CASE WHEN g.segment_id IS NOT NULL THEN {tr_key('g.segment_id')} END,
                  {si_key('s.sighting_id')}) AS root_sample_key,
         '{ds_key}' AS dataset_key, s.grid_key, NULL::VARCHAR AS site_key, s.cruise_key,
         NULL::INTEGER AS order_occ, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP) AS datetime,
         0::DOUBLE AS depth_min_m, 0::DOUBLE AS depth_max_m, NULL::VARCHAR AS tow_type
  FROM cetacean_sighting s
  LEFT JOIN cetacean_segment g USING (segment_id)"))

# effort: segment length on the transect sample
append_sample_measurement(con, glue("
  SELECT {tr_key('segment_id')}, '{ds_key}', 'transect_length',
         segment_length_km, NULL::VARCHAR
  FROM cetacean_segment WHERE segment_length_km IS NOT NULL"))

# obs -- the headline: one row per sighting, the group size (best estimate).
# One species per sighting (Q05), so the sighting IS the occurrence grain.
append_obs(con, glue("
  SELECT 'bio', '{ds_key}', {si_key('s.sighting_id')},
         s.grid_key, s.cruise_key, s.latitude, s.longitude,
         CAST(s.datetime_start_utc AS TIMESTAMP), 0::DOUBLE, 0::DOUBLE,
         dt.taxon_key, NULL::VARCHAR, 'group_size', s.group_size_best,
         s.effort_status, NULL::DOUBLE
  FROM cetacean_sighting s
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = s.species_code"))

# obs_attribute -- group size range, calves, behaviour. Not a breakdown of the
# headline (none of these sums to group_size), so there is no sum-parity to
# assert; each is asserted against its source column instead.
attr_sel <- function(type, life_stage, bin_value, bin_label, count, where) glue("
  SELECT '{ds_key}', {si_key('s.sighting_id')}, dt.taxon_key, {life_stage}, '{type}',
         {bin_value}, {bin_label}, {count}, NULL::VARCHAR
  FROM cetacean_sighting s
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = s.species_code
  WHERE {where}")
append_obs_attribute(con, attr_sel("group_size_min", "NULL::VARCHAR", "NULL::DOUBLE", "NULL::VARCHAR",
                                   "CAST(s.group_size_min AS INTEGER)", "s.group_size_min IS NOT NULL"))
append_obs_attribute(con, attr_sel("group_size_max", "NULL::VARCHAR", "NULL::DOUBLE", "NULL::VARCHAR",
                                   "CAST(s.group_size_max AS INTEGER)", "s.group_size_max IS NOT NULL"))
append_obs_attribute(con, attr_sel("calf_count", "'calf'", "NULL::DOUBLE", "NULL::VARCHAR",
                                   "s.calf_count", "s.calf_count IS NOT NULL"))

# behaviour: the three columns hold one code or a list ("1, 34", "2 8"); each
# token is its own row, labelled with the code as recorded (codebook Q03)
d_behavior <- d_sighting |>
  select(sighting_id, behavior_primary, behavior_other1, behavior_other2) |>
  pivot_longer(-sighting_id, names_to = "slot", values_to = "code") |>
  filter(!is.na(code)) |>
  mutate(slot = match(slot, c("behavior_primary", "behavior_other1", "behavior_other2")),
         # split lists of numeric codes; free text ("Slow Travel") stays whole
         code = map(code, \(x) if (str_detect(x, "^[0-9?,/&\\s]+$"))
                                   str_split_1(x, "[,/&\\s]+") else x)) |>
  unnest(code) |>
  mutate(code = str_trim(code)) |>
  filter(code != "")
dbWriteTable(con, "_behavior", d_behavior, overwrite = TRUE)
append_obs_attribute(con, glue("
  SELECT '{ds_key}', {si_key('b.sighting_id')}, dt.taxon_key, NULL::VARCHAR, 'behavior',
         CAST(b.slot AS DOUBLE), b.code, NULL::INTEGER, NULL::VARCHAR
  FROM _behavior b JOIN cetacean_sighting s USING (sighting_id)
  LEFT JOIN dataset_taxon dt ON dt.dataset_key = '{ds_key}'
                            AND dt.ds_taxa_code = s.species_code"))
dbExecute(con, "DROP TABLE _behavior")

q1 <- function(sql) dbGetQuery(con, sql)[[1]]
core <- list(
  sample             = q1("SELECT COUNT(*) FROM sample"),
  sample_transect    = q1("SELECT COUNT(*) FROM sample WHERE sample_type = 'transect'"),
  sample_sighting    = q1("SELECT COUNT(*) FROM sample WHERE sample_type = 'sighting'"),
  sample_measurement = q1("SELECT COUNT(*) FROM sample_measurement"),
  obs                = q1("SELECT COUNT(*) FROM obs"),
  obs_attribute      = q1("SELECT COUNT(*) FROM obs_attribute"))
cat(glue("core projection -- ", paste(names(core), unlist(core), sep = "=", collapse = " ")), "\n")

stopifnot(
  "one transect sample per segment"  = core$sample_transect == nrow(d_segment),
  "one sighting sample per sighting" = core$sample_sighting == nrow(d_sighting),
  "one obs per sighting"             = core$obs == nrow(d_sighting),
  "obs value total = source Best total" =
    isTRUE(all.equal(q1("SELECT SUM(measurement_value) FROM obs"),
                     sum(d_sighting$group_size_best, na.rm = TRUE))),
  "group_size_min rows = non-NULL Min" =
    q1("SELECT COUNT(*) FROM obs_attribute WHERE measurement_type = 'group_size_min'") ==
    sum(!is.na(d_sighting$group_size_min)),
  "calf_count total = source total" =
    q1("SELECT COALESCE(SUM(count), 0) FROM obs_attribute WHERE measurement_type = 'calf_count'") ==
    sum(d_sighting$calf_count, na.rm = TRUE),
  "transect_length total = source total" =
    isTRUE(all.equal(q1("SELECT SUM(measurement_value) FROM sample_measurement"),
                     sum(d_segment$segment_length_km, na.rm = TRUE))),
  "every obs.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN sample s USING (sample_key) WHERE s.sample_key IS NULL") == 0,
  "every obs_attribute.sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM obs_attribute a LEFT JOIN sample s USING (sample_key) WHERE s.sample_key IS NULL") == 0,
  "every sample.parent_sample_key resolves in sample" =
    q1("SELECT COUNT(*) FROM sample c LEFT JOIN sample p ON p.sample_key = c.parent_sample_key
       WHERE c.parent_sample_key IS NOT NULL AND p.sample_key IS NULL") == 0,
  "every obs.taxon_key RESOLVES in taxon" =
    q1("SELECT COUNT(*) FROM obs o LEFT JOIN taxon t USING (taxon_key)
       WHERE o.taxon_key IS NOT NULL AND t.taxon_key IS NULL") == 0,
  "a NULL obs.taxon_key is an Oth sighting and nothing else" =
    q1("SELECT COUNT(*) FROM obs WHERE taxon_key IS NULL") == sum(d_sighting$species_code == "Oth"),
  "every obs / obs_attribute / sample_measurement type is registered" =
    q1("SELECT COUNT(*) FROM (SELECT measurement_type FROM obs UNION
                             SELECT measurement_type FROM obs_attribute UNION
                             SELECT measurement_type FROM sample_measurement) x
       WHERE measurement_type NOT IN (SELECT measurement_type FROM measurement_type)") == 0)
```

### Declared NULLs

`validate_for_release()` treats every `_key` column as required. These NULLs are
the dataset's contract, each with its count and reason; anything else fails.

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)

unresolved_labels <- grp$cruise_label[is.na(grp$cruise_key)]
lbl_sql <- paste0("'", c(unresolved_labels, "_"), "'", collapse = ", ")
nullable <- tribble(
  ~table,   ~column,             ~n,                                              ~reason,
  "sample", "cruise_key",        q1("SELECT COUNT(*) FROM sample WHERE cruise_key IS NULL"),
    glue("labels with no track match ({paste(unresolved_labels, collapse = ', ')}), and their segments"),
  "sample", "parent_sample_key", sum(is.na(d_sighting$segment_id)) + n_seg_orphan + nrow(d_segment),
    "transects are roots; off-effort sightings, and those whose segment_id is not in the effort file, have no parent",
  "sample", "site_key",          nrow(d_segment) + nrow(d_sighting),
    "underway survey: no event sits on a CalCOFI station",
  "sample", "grid_key",          q1("SELECT COUNT(*) FROM sample WHERE grid_key IS NULL"),
    "positions outside the CalCOFI grid (e.g. the CC1511 transit north of 39 N) and the one nulled position (Q11)",
  "sample", "datetime",          n_ref,
    "'#REF!' sighting times on CC0501 (Q06)",
  "obs",    "taxon_key",         sum(d_sighting$species_code == "Oth"),
    "'Oth' (Other) is not a taxon")
# the counts above that are measured rather than derived from the source must
# still be explained: every NULL cruise_key belongs to an unresolved label
stopifnot(
  "every NULL sighting cruise_key belongs to an unresolved label" =
    q1(glue("SELECT COUNT(*) FROM cetacean_sighting WHERE cruise_key IS NULL
            AND cruise_label NOT IN ({lbl_sql})")) == 0,
  "the declared NULL counts match the tables" =
    q1("SELECT COUNT(*) FROM sample WHERE parent_sample_key IS NULL") ==
      nullable$n[nullable$column == "parent_sample_key"] &&
    q1("SELECT COUNT(*) FROM sample WHERE datetime IS NULL") ==
      nullable$n[nullable$column == "datetime"])
cat("validate_for_release:", ifelse(results$passed, "PASSED", "reported findings -- reconciled against the declared NULLs below"), "\n")
nullable |> datatable(caption = "Declared NULLs (count and reason)", rownames = FALSE)
```

### Measurement bounds

```{r}
#| label: bounds

b <- bind_rows(
  check_measurement_bounds(con, "obs",                mt = d_meas_type, dataset_key = ds_key),
  check_measurement_bounds(con, "sample_measurement", mt = d_meas_type, dataset_key = ds_key))
bounds_datatable(b)
stopifnot("every emitted bound is declared and respected" = all(b$status == "ok"))
```

## Preview

```{r}
#| label: preview-obs

dbGetQuery(con, "
  SELECT o.taxon_key, t.scientific_name, COUNT(*) AS n_sightings,
         SUM(o.measurement_value) AS n_animals,
         COUNT(*) FILTER (WHERE o.measurement_qual = 'ON') AS n_on_effort,
         MIN(year(o.datetime)) AS year_min, MAX(year(o.datetime)) AS year_max
  FROM obs o LEFT JOIN taxon t USING (taxon_key)
  GROUP BY ALL ORDER BY n_sightings DESC") |>
  datatable(caption = "obs by taxon", rownames = FALSE)
```

```{r}
#| label: preview-sample

dbGetQuery(con, "
  SELECT sample_key, sample_type, parent_sample_key, cruise_key, grid_key,
         latitude, longitude, datetime
  FROM sample ORDER BY datetime LIMIT 200") |>
  datatable(caption = "sample -- first 200 by time", rownames = FALSE, filter = "top")
```

```{r}
#| label: erd-core

core_tbls <- core_output_tables(con)
cc_erd(con, tables = core_tbls, rels = core_relationships(core_tbls))
```

## Write Outputs + Upload

```{r}
#| label: outputs

dir_create(dir_parquet)
tbls_out <- core_output_tables(con, extra = c("measurement_type", "dataset"))
write_parquet_outputs(
  con = con, output_dir = dir_parquet, tables = tbls_out,
  sort_by = list(obs = c("grid_key", "measurement_type")),
  strip_provenance = FALSE)

build_relationships_json(
  rels = core_relationships(tbls_out), output_dir = dir_parquet,
  provider = provider, dataset = dataset)

d_tbls_rd <- read_csv(file.path(dir_meta, "tbls_redefine.csv"))
d_flds_rd <- read_csv(file.path(dir_meta, "flds_redefine.csv"))
build_metadata_json(
  con = con, d_tbls_rd = d_tbls_rd, d_flds_rd = d_flds_rd,
  metadata_derived_csv = c(here("metadata/core_dictionary.csv"),
                           file.path(dir_meta, "metadata_derived.csv")),
  output_dir = dir_parquet, tables = tbls_out,
  set_comments = TRUE, provider = provider, dataset = dataset,
  workflow_url = cc$workflow_url, tables_owned = tables_owned,
  sources = src_stamp)

if (publish_to_gcs) {
  sync_to_gcs(local_dir = dir_stage, sidecar_dir = dir_parquet,
              gcs_prefix = glue("ingest/{dir_label}"), bucket = "calcofi-db")
} else {
  cat("CALCOFI_SKIP_GCS set -- parquet kept local, NOT synced to gs://calcofi-db\n")
}
```

```{r}
#| label: cleanup
close_duckdb(con)
cat(glue("Parquet outputs written to: {dir_parquet}"), "\n")
```
