---
title: "Load Sanctuaries"
author: "Ben Best"
date: "`r Sys.time()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source(here::here("libs/rmd.R")) # defines function: time_it
knitr::opts_chunk$set(
  echo = TRUE, warning = F, message = F, time_chunk = TRUE)
```

## TODO: load into single `aois` using `hstore`

Use PostgreSQL [`hstore`](https://www.postgresql.org/docs/current/hstore.html) to store the many varying fields of any given AOI feature into a single column, say `keys`.

See issue [#13: load areas of interest into single `aois` table using `hstore`, starting with `sanctuaries`](https://github.com/CalCOFI/scripts/issues/13).

## Sanctuaries

* [Sanctuaries with spatial information — sanctuaries • onmsR](https://noaa-onms.github.io/onmsR/reference/sanctuaries.html)

```{r}
source(here::here("libs/db.R")) # defines variables: con, dir_gdrive
librarian::shelf(
  glue, here, mapview, sf)
mapviewOptions(fgb = FALSE)

# source: https://github.com/noaa-onms/onmsR/blob/master/data/sanctuaries.rda
# sanctuaries_rda <- "/share/data/ONMS-Sanctuaries/sanctuaries.rda"
sanctuaries_rda <- "~/github/noaa-onms/onmsR/data/sanctuaries.rda"

load(file=sanctuaries_rda)
sanctuaries <- sanctuaries %>% select(-spatial)
mapview(sanctuaries)
```

## Sanctuaries intersecting CalCOFI

```{r}
stations <- st_read(                    # n =  6,343 
  con, 
  query = "
    SELECT DISTINCT ON (sta_id)
      line || ' ' || station AS sta_id, line, station, longitude, latitude, geom
    FROM stations")
mapview(stations)
```

```{r}
hull <- st_convex_hull(st_union(stations))
st_write(hull, con, "stations_hull", delete_layer=T)
mapview(hull)
```

```{r}
sanctuaries <- st_set_crs(sanctuaries, 4326)
hull        <- st_set_crs(hull, 4326)
x <- st_intersects(sanctuaries, hull, sparse = F)[,1]
sanctuaries <- sanctuaries[x,]
mapview(sanctuaries)
```

## Load into Database

```{r}
st_write(sanctuaries, con, "aoi_fed_sanctuaries", delete_layer=T)
# add spatial index
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS aoi_fed_sanctuaries_geom_idx ON aoi_fed_sanctuaries USING GIST (geom);")
# set geographic projection for showing up at tile.calcofi.io
dbSendQuery(con, "SELECT UpdateGeometrySRID('aoi_fed_sanctuaries','geom',4326);")

# check read from db to spatial features
sanctuaries <- sf::st_read(con, "aoi_fed_sanctuaries")
sanctuaries
```

