---
title: "Load Oceanographic Data"
author: "Ben Best"
date: "`r Sys.time()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source(here::here("libs/rmd.R")) # defines function: time_it
knitr::opts_chunk$set(
  echo = TRUE, warning = F, message = F, time_chunk = TRUE)
```

## Load libraries

```{r packages}
source(here::here("libs/db.R")) # defines variables: con, dir_gdrive
librarian::shelf(
  DT, glue, janitor, lubridate, mapview, purrr, readr, readxl, 
  sf, skimr, stringr, tidyr)
options(readr.show_col_types = F)
mapviewOptions(fgb = FALSE)

# helper functions ----

# convert station ID to lon, lat using the proj library
proj <- "/Users/bbest/homebrew/bin/proj" # installed on Ben's MacBookPro from https://proj.org/apps/proj.html
lonlat_to_stationid <- function(lon, lat){
  system(glue("echo {lon} {lat} | {proj} +proj=calcofi +epsg=4326 -f '%05.1f'"), intern=T) %>%
    stringr::str_replace("\t", " ")
}
stationid_to_lonlat <- function(stationid){
  # using 5th decimal place, a la CCE_Stations.txt
  system(glue("echo {stationid} | {proj} +proj=calcofi +epsg=4326  -I -d 5"), intern=T) %>%
    stringr::str_replace("\t", " ")
}
# (a_staid  <- stations$Sta_ID[1]) # "001.0 168.0"
# (a_lonlat <- stationid_to_lonlat(a_staid)) # 93.3	26.7
# a_lon <- str_split(a_lonlat, " ", simplify=T)[1] %>% as.double()
# a_lat <- str_split(a_lonlat, " ", simplify=T)[2] %>% as.double()
# lonlat_to_stationid(a_lon, a_lat) # "1.00 168.00" -> "001.0 168.0"
# "1.0 168.0"
```

## Read input tables

Initially copied from `CalCOFI/calcofi4r:data-raw/`[`bottle.R`](https://github.com/CalCOFI/calcofi4r/blob/bc8427670fb474390326a054bda9c10cf6de1831/data-raw/bottle.R).

```{r paths}
# paths ----

# oceanographic source paths
bottle_csv <- file.path(dir_gdrive, "data/oceanographic-data/bottle-database/CalCOFI_Database_194903-202001_csv_22Sep2021/194903-202001_Bottle.csv")
cast_csv   <- file.path(dir_gdrive, "data/oceanographic-data/bottle-database/CalCOFI_Database_194903-202001_csv_22Sep2021/194903-202001_Cast.csv")
# bottle_cast_rds <- file.path(dir_gdrive, "data/oceanographic-data/bottle-database/bottle_cast.rds")
DIC_csv    <- file.path(dir_gdrive, "data/DIC/CalCOFI_DICs_200901-201507_28June2018.csv")

# stations source paths
# source: Shonna Dovel <sdovel@ucsd.edu> 2022-03-17
stations_ccelter_xls <- file.path(dir_gdrive, "data/CalCOFI-stations/CalCOFI station LatLong.xlsx")
# sourc: http://cce.lternet.edu/data/gmt-mapper
stations_cce_txt     <- file.path(dir_gdrive, "data/CalCOFI-stations/CCE_Stations.txt")
stations_sccoos_txt  <- file.path(dir_gdrive, "data/CalCOFI-stations/SCCOOS_Stations.txt")

# study area destination paths
calcofi_geo           <- here("data/calcofi_oceano-bottle-stations_convex-hull.geojson")
calcofi_offshore_geo  <- here("data/calcofi_oceano-bottle-stations_convex-hull_offshore.geojson")
calcofi_nearshore_geo <- here("data/calcofi_oceano-bottle-stations_convex-hull_nearshore.geojson")

# check paths
stopifnot(all(file.exists(
  bottle_csv, cast_csv, DIC_csv,
  stations_ccelter_xls, stations_cce_txt, stations_sccoos_txt)))

# read csv sources ----

d_bottle <- read_csv(bottle_csv, skip=1, col_names = F, guess_max = 1000000)
#d_bottle_problems() <- problems()
names(d_bottle) <- str_split(
  readLines(bottle_csv, n=1), ",")[[1]] %>%
  str_replace("\xb5", "µ")

d_cast  <- read_csv(cast_csv)

d_DIC <- read_csv(DIC_csv, skip=1, col_names = F, guess_max = 1000000)
names(d_DIC) <- str_split(
  readLines(DIC_csv, n=1), ",")[[1]] %>%
  str_replace("\xb5", "µ")
# d_DIC %>% head() %>% View()
d_DIC <- d_DIC %>%
  rename("Sta_ID"="Line Sta_ID")
```

## `stations`

```{r stations}
# get unique station IDs from the casts
stations <- d_cast %>%
  select(Sta_ID) %>%
  group_by(Sta_ID) %>%
  summarize() %>%
  mutate(
    is_cast     = TRUE,
    lonlat_proj = map_chr(Sta_ID, stationid_to_lonlat)) %>%
  separate(lonlat_proj, c("lon", "lat"), sep=" ", convert = T) %>%
  separate(
    Sta_ID, c("Sta_ID_line", "Sta_ID_station"), sep=" ", remove=F, convert=T) %>%
  mutate(
    is_offshore = ifelse(Sta_ID_station > 60, T, F)) %>%
  st_as_sf(
    coords = c("lon", "lat"), crs=4326, remove = F)
# stations$Sta_ID
# mapview(stations)

# + CCE field ----
stations_cce <- read_tsv(stations_cce_txt, skip = 2) %>%
  mutate(
    is_cce = TRUE,
    Sta_ID = map2_chr(LonDec, LatDec, lonlat_to_stationid))
stopifnot(sum(!stations_cce$Sta_ID %in% stations$Sta_ID) == 0)
stations <- stations %>%
  left_join(
    stations_cce %>%
      select(Sta_ID, is_cce),
    by = "Sta_ID")

# + CCE-LTER field ----
stations_ccelter <- read_excel(stations_ccelter_xls) %>%
  mutate(
    is_ccelter = TRUE,
    Sta_ID       = glue("{sprintf('%05.1f', Line)} {sprintf('%05.1f', Sta)}"),
    lonlat_proj  = map_chr(Sta_ID, stationid_to_lonlat)) %>%
  separate(
    lonlat_proj, c("lon", "lat"), sep=" ", convert = T) %>%
  select(Sta_ID, is_ccelter, lon, lon_0 = `Lon Dec`, lat, lat_0 = `Lat Dec`)
# stations_ccelter %>% View()

stopifnot(sum(!stations_ccelter$Sta_ID %in% stations$Sta_ID) == 0)
stations <- stations %>%
  left_join(
    stations_ccelter %>%
      select(Sta_ID, is_ccelter),
    by = "Sta_ID")

# + SCCOOS field ----
stations_sccoos <- read_tsv(stations_sccoos_txt) %>%
  mutate(
    is_sccoos = TRUE,
    Sta_ID = map2_chr(LonDec, LatDec, lonlat_to_stationid))
stopifnot(sum(!stations_sccoos$Sta_ID %in% stations$Sta_ID) == 0)
stations <- stations %>%
  left_join(
    stations_sccoos %>%
      select(Sta_ID, is_sccoos),
    by = "Sta_ID")

# stations_0 <- stations
stations <- stations %>% 
  mutate(across(where(is.logical), ~replace_na(.,FALSE)))

stations %>%
  select(is_cce, is_ccelter) %>%
  st_drop_geometry() %>%
  table(useNA = "ifany")
#           is_cce
# is_ccelter FALSE TRUE
#      FALSE  2521   47
#      TRUE      0   66
# So: 47 extra stations in CCE that are missing in  CCE-LTER
stations %>%
  select(is_cce, is_sccoos) %>%
  st_drop_geometry() %>%
  table(useNA = "ifany")
#        is_sccoos
# is_cce  FALSE TRUE
#   FALSE  2521    0
#   TRUE    104    9
# So: only 9 stations in SCCOS, none of which are in CCE

# field order ----
stations <- stations %>%
  select(
    sta_id = Sta_ID, 
    sta_id_line = Sta_ID_line, sta_id_station = Sta_ID_station,
    lon, lat,
    is_offshore,
    is_cce, is_ccelter, is_sccoos) %>% 
  rename(geom = geometry)

# write to database
st_write(stations, con, "stations")
# to get to show as layer in tile.calcofi.io
dbSendQuery(con, "SELECT UpdateGeometrySRID('stations', 'geom', 4326);")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS stations_geom_idx ON stations USING GIST (geom);")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS stations_staid_idx ON stations (sta_id)")

# show map of stations
mapview(stations)

# show table summary
stations %>% 
  st_drop_geometry() %>% 
  skim()
```

## `ctdcast`

Note that [`cast`](https://www.postgresql.org/docs/14/sql-expressions.html#SQL-SYNTAX-TYPE-CASTS) is a reserved word in SQL for a function to convert data types, so we're calling this table `ctdcast` instead of `cast`.

Writing queries to the database is also much easier if we don't have to wrap field names in quotes. This means that all field names have to be lower case and not use any spaces, dashes or special characters. Thankfully `janitor::`[`make_clean_names()`](https://www.rdocumentation.org/packages/janitor/versions/2.1.0/topics/make_clean_names) can handle this for us. We'll just also keep track of the original field names in a separate table `fields_renamed`.

```{r ctdcast}
stopifnot(length(setdiff(unique(d_cast$Sta_ID), stations$sta_id)) == 0)

d_cast <- d_cast %>% 
  mutate(
    datetime = map2_chr(Date, Time, function(f_date, f_time){
      s_time <- ifelse(
        is.na(f_time), 
        "00:00:00",
        str_split(f_time, " ")[[1]][2])
      glue("{f_date} {s_time}") }) %>% 
      as_datetime(
        format = "%m/%d/%Y %H:%M:%S",
        tz = "America/Los_Angeles"), # TODO: confirm PST and not UTC timezone
    Date = as_date(Date, format = "%m/%d/%Y")) %>% 
  st_as_sf(
    coords = c("Lon_Dec", "Lat_Dec"), crs=4326, remove = F) %>%
  rename(geom = geometry)

# rename fields
flds_orig       <- names(d_cast)
flds            <- make_clean_names(flds_orig)
names(d_cast) <- flds
d_flds <- tibble(
  table      = "ctdcast",
  field_orig = flds_orig,
  field      = flds)

# write spatial points to database
st_write(d_cast, con, "ctdcast")
# to get to show as layer in tile.calcofi.io
dbSendQuery(con, "SELECT UpdateGeometrySRID('ctdcast', 'geom', 4326);")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcast_geom_idx ON ctdcast USING GIST (geom);")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcast_staid_idx ON ctdcast (sta_id)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcast_cstcnt_idx ON ctdcast (cst_cnt)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcast_date_idx ON ctdcast (date)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcast_datetime_idx ON ctdcast (datetime)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcast_datetime_idx ON ctdcast (datetime)")

# show table summary
d_cast %>% 
  st_drop_geometry() %>% 
  skim()
```

## `ctdcast_bottle`

```{r ctdcast_bottle}
# rename fields
flds_orig       <- names(d_bottle)
flds            <- make_clean_names(flds_orig)
flds[which(flds_orig == "Depthm")] <- "depth_m"
names(d_bottle) <- flds
d_flds <- bind_rows(
  d_flds,
  tibble(
    table      = "ctdcast_bottle",
    field_orig = flds_orig,
    field      = flds))

# write table to database
dbWriteTable(con, "ctdcast_bottle", d_bottle, overwrite=T)
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcastbottle_staid_idx  ON ctdcast_bottle (sta_id)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcastbottle_cstcnt_idx ON ctdcast_bottle (cst_cnt)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcastbottle_btlcnt_idx ON ctdcast_bottle (btl_cnt)")

# show table summary
skim(d_bottle)
```

## `ctdcast_bottle_dic`

```{r ctdcast_bottle_dic}
# ensure bottle.btl_cnt is unique in 
stopifnot(d_bottle$btl_cnt %>% duplicated() %>% sum() == 0)

# rename fields
flds_orig       <- names(d_DIC)
flds            <- make_clean_names(flds_orig)
flds[which(flds_orig == "Cast_Index")]   <- "cst_cnt"
flds[which(flds_orig == "Bottle_Index")] <- "btl_cnt"
names(d_DIC) <- flds
d_flds <- bind_rows(
  d_flds,
  tibble(
    tbl        = "ctdcast_bottle_dic",
    field_orig = flds_orig,
    field      = flds))

# ensure uniquely matchable to cast_bottle
x <- anti_join(
  d_DIC,
  d_bottle, 
  by="btl_cnt")
stopifnot(nrow(x) == 0)

# write table to database
dbWriteTable(con, "ctdcast_bottle_dic", d_DIC, overwrite=T)
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcastbottledic_staid_idx  ON ctdcast_bottle_dic (sta_id)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcastbottledic_cstcnt_idx ON ctdcast_bottle_dic (cst_cnt)")
dbSendQuery(con, "CREATE INDEX IF NOT EXISTS ctdcastbottledic_btlcnt_idx ON ctdcast_bottle_dic (btl_cnt)")

# show table summary
skim(d_DIC)
```

## `fields_renamed`

```{r fields_renamed}
# write table to database
dbWriteTable(con, "fields_renamed", d_flds, overwrite=T)

# show table
datatable(d_flds)
```

