---
title: "Merge Ichthyo & Bottle to Working DuckLake"
execute:
  echo: true
  warning: false
editor_options:
  chunk_output_type: console
format:
  html:
    code-fold: true
editor:
  markdown:
    wrap: 72
---

## Overview {.unnumbered}

**Goal**: Merge parquet outputs from both upstream ingest workflows,
perform cross-dataset reconciliation (ship matching, cruise bridge,
taxonomy standardization), then load into Working DuckLake and create a
frozen release.

**Steps**:

1.  Load parquet files from both upstream workflows into local temp
    DuckDB
2.  Fuzzy match ships between datasets
3.  Build cruise bridge (casts → cruise_key)
4.  Standardize taxonomy (species → WoRMS/ITIS/GBIF)
5.  Validate cross-dataset integrity
6.  Write modified parquet outputs + manifest with GCS references
7.  Ingest all tables into Working DuckLake with provenance
8.  Freeze release (strip provenance)
9.  Upload to GCS

**Upstream workflows** (must complete first):

-   `ingest_swfsc.noaa.gov_calcofi-db.qmd` → ichthyo tables (10)
-   `ingest_calcofi.org_bottle-database.qmd` → bottle/cast tables (5)

```{mermaid}
%%| label: fig-workflow
%%| fig-cap: "Integration workflow: upstream parquet → merge → Working DuckLake → Frozen Release"

flowchart LR
    subgraph upstream["Upstream Parquet"]
        s["icthyo<br/>10 tables"]
        b["bottle<br/>5 tables"]
    end
    subgraph merge["Merge DuckDB"]
        m1["Ship matching"]
        m2["Cruise bridge"]
        m3["Taxonomy"]
    end
    subgraph output["Output"]
        p["Modified parquet<br/>+ manifest"]
        w["Working DuckLake"]
        f["Frozen Release"]
    end
    s --> merge
    b --> merge
    merge --> p --> w --> f

    style upstream fill:#e3f2fd,stroke:#1565c0
    style merge fill:#fff3e0,stroke:#ef6c00
    style output fill:#e8f4e8,stroke:#2e7d32
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4R"))
librarian::shelf(
  CalCOFI / calcofi4db,
  DBI,
  dm,
  dplyr,
  DT,
  fs,
  glue,
  here,
  jsonlite,
  purrr,
  readr,
  stringr,
  tibble,
  tidyr,
  worrms,
  taxize,
  quiet = T
)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# parquet input directories (from upstream workflows)
dir_pq_ichthyo <- here("data/parquet/swfsc.noaa.gov_calcofi-db")
dir_pq_bottle <- here("data/parquet/calcofi.org_bottle-database")

# MarineSensitivity species DB for local taxonomy lookups
spp_db_path <- Sys.getenv(
  "SPP_DB_PATH",
  unset = "/Users/bbest/_big/msens/derived/spp.duckdb"
)

# release version
release_version <- format(Sys.Date(), "v%Y.%m")

# local merge database
db_path <- here("data/wrangling/merge_ichthyo_bottle.duckdb")
overwrite <- TRUE # set TRUE to rebuild from scratch

if (overwrite && file_exists(db_path)) {
  file_delete(db_path)
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")

# check if database already has tables (from previous run)
existing_tables <- DBI::dbListTables(con)
skip_load <- length(existing_tables) > 0
if (skip_load) {
  message(glue("Using existing database with {length(existing_tables)} tables"))
}
```

## Check Upstream Workflows

Verify that parquet outputs and manifests exist from both upstream
workflows.

```{r}
#| label: check_upstream

# check ichthyo manifest
ichthyo_manifest_path <- file.path(dir_pq_ichthyo, "manifest.json")
stopifnot(
  "Missing ichthyo parquet outputs. Run ingest_swfsc.noaa.gov_calcofi-db.qmd first." = file.exists(
    ichthyo_manifest_path
  )
)

ichthyo_manifest <- read_json(ichthyo_manifest_path)
message(glue(
  "Ichthyo parquet: {ichthyo_manifest$total_rows} total rows, ",
  "created {ichthyo_manifest$created_at}"
))

# check bottle manifest
bottle_manifest_path <- file.path(dir_pq_bottle, "manifest.json")
stopifnot(
  "Missing bottle parquet outputs. Run ingest_calcofi.org_bottle-database.qmd first." = file.exists(
    bottle_manifest_path
  )
)

bottle_manifest <- read_json(bottle_manifest_path)
message(glue(
  "Bottle parquet: {bottle_manifest$total_rows} total rows, ",
  "created {bottle_manifest$created_at}"
))

# show manifest stats
tibble(
  source = c("ichthyo", "bottle"),
  tables = c(
    length(ichthyo_manifest$tables),
    length(bottle_manifest$tables)
  ),
  rows = c(
    ichthyo_manifest$total_rows,
    bottle_manifest$total_rows
  ),
  created = c(
    ichthyo_manifest$created_at,
    bottle_manifest$created_at
  )
) |>
  datatable(caption = "Upstream workflow manifest stats")
```

## Load SWFSC Tables

Load all ichthyo parquet files into the local merge DuckDB. Tables with
geometry columns (grid, site, segment) need WKB→GEOMETRY conversion.

```{r}
#| label: load_ichthyo

ichthyo_files <- list.files(
  dir_pq_ichthyo,
  pattern = "\\.parquet$",
  full.names = TRUE
)

# tables with geometry columns that need WKB conversion
geom_tables <- c("grid", "site", "segment")

if (skip_load) {
  message("Using existing ichthyo tables from database")
  ichthyo_stats <- map_dfr(ichthyo_files, function(pqt_path) {
    tbl_name <- tools::file_path_sans_ext(basename(pqt_path))
    n <- DBI::dbGetQuery(con, glue("SELECT COUNT(*) AS n FROM {tbl_name}"))$n
    tibble(table = tbl_name, rows = n, source = "ichthyo")
  })
} else {
  ichthyo_stats <- map_dfr(ichthyo_files, function(pqt_path) {
    tbl_name <- tools::file_path_sans_ext(basename(pqt_path))

    if (tbl_name %in% geom_tables) {
      # use load_gcs_parquet_to_duckdb for WKB→GEOMETRY conversion
      DBI::dbExecute(
        con,
        glue(
          "CREATE OR REPLACE TABLE {tbl_name} AS
         SELECT * FROM read_parquet('{pqt_path}')"
        )
      )
      # detect and convert WKB BLOB columns
      blob_cols <- DBI::dbGetQuery(
        con,
        glue(
          "SELECT column_name FROM information_schema.columns
         WHERE table_name = '{tbl_name}'
           AND data_type = 'BLOB'
           AND column_name LIKE '%geom%'"
        )
      )$column_name
      for (gc in blob_cols) {
        tmp_col <- paste0(gc, "_tmp")
        DBI::dbExecute(
          con,
          glue(
            'ALTER TABLE {tbl_name} ADD COLUMN {tmp_col} GEOMETRY'
          )
        )
        DBI::dbExecute(
          con,
          glue(
            'UPDATE {tbl_name} SET {tmp_col} = ST_GeomFromWKB({gc})'
          )
        )
        DBI::dbExecute(
          con,
          glue(
            'ALTER TABLE {tbl_name} DROP COLUMN {gc}'
          )
        )
        DBI::dbExecute(
          con,
          glue(
            'ALTER TABLE {tbl_name} RENAME COLUMN {tmp_col} TO {gc}'
          )
        )
      }
    } else {
      DBI::dbExecute(
        con,
        glue(
          "CREATE OR REPLACE TABLE {tbl_name} AS
         SELECT * FROM read_parquet('{pqt_path}')"
        )
      )
    }

    n <- DBI::dbGetQuery(con, glue("SELECT COUNT(*) AS n FROM {tbl_name}"))$n
    tibble(table = tbl_name, rows = n, source = "ichthyo")
  })
}

ichthyo_stats |> datatable(caption = "SWFSC ichthyo tables loaded")
```

## Load Bottle Tables

Load bottle parquet files (grid excluded from bottle export — canonical
from swfsc).

```{r}
#| label: load_bottle

bottle_files <- list.files(
  dir_pq_bottle,
  pattern = "\\.parquet$",
  full.names = TRUE
)

# belt-and-suspenders: skip grid.parquet if it still exists
bottle_files <- bottle_files[!grepl("grid\\.parquet$", bottle_files)]

if (skip_load) {
  message("Using existing bottle tables from database")
  bottle_stats <- map_dfr(bottle_files, function(pqt_path) {
    tbl_name <- tools::file_path_sans_ext(basename(pqt_path))
    n <- DBI::dbGetQuery(con, glue("SELECT COUNT(*) AS n FROM {tbl_name}"))$n
    tibble(table = tbl_name, rows = n, source = "bottle")
  })
} else {
  bottle_stats <- map_dfr(bottle_files, function(pqt_path) {
    tbl_name <- tools::file_path_sans_ext(basename(pqt_path))

    DBI::dbExecute(
      con,
      glue(
        "CREATE OR REPLACE TABLE {tbl_name} AS
       SELECT * FROM read_parquet('{pqt_path}')"
      )
    )

    n <- DBI::dbGetQuery(con, glue("SELECT COUNT(*) AS n FROM {tbl_name}"))$n
    tibble(table = tbl_name, rows = n, source = "bottle")
  })
}

bottle_stats |> datatable(caption = "Bottle tables loaded")
```

## Show Loaded Tables

```{r}
#| label: show_tables

all_stats <- bind_rows(ichthyo_stats, bottle_stats)

message(glue(
  "Loaded {nrow(all_stats)} tables: ",
  "{sum(all_stats$source == 'ichthyo')} ichthyo + ",
  "{sum(all_stats$source == 'bottle')} bottle"
))

all_stats |>
  arrange(source, table) |>
  datatable(caption = "All loaded tables")
```

## Fuzzy Ship Matching

Reconcile ship codes between bottle casts and swfsc ship reference
table. Uses `calcofi4db::match_ships()`.

```{r}
#| label: ship_matching

# get unmatched ships from casts
unmatched <- dbGetQuery(
  con,
  "
  SELECT DISTINCT c.ship_code, c.ship_name
  FROM casts c
  LEFT JOIN ship s ON c.ship_code = s.ship_nodc
  WHERE s.ship_key IS NULL"
)

message(glue("{nrow(unmatched)} unmatched ship codes in casts"))

# ship renames CSV (manual overrides from previous reconciliation)
ship_renames_csv <- here(
  "metadata/calcofi.org/bottle-database/ship_renames.csv"
)

# run fuzzy matching
ship_matches <- match_ships(
  unmatched_ships = unmatched,
  reference_ships = dbReadTable(con, "ship"),
  ship_renames_csv = ship_renames_csv,
  fetch_ices = TRUE
)

ship_matches |>
  datatable(caption = "Ship matching results")
```

### Match Statistics

```{r}
#| label: ship_match_stats

ship_matches |>
  count(match_type) |>
  datatable(caption = "Ship match type counts")
```

## Build Cruise Bridge

Link bottle casts to swfsc cruise table via ship_key → cruise_key.

### Step 1: Add ship_key to casts

```{r}
#| label: cruise_bridge_ship_key

# add ship_key column
dbExecute(con, "ALTER TABLE casts ADD COLUMN IF NOT EXISTS ship_key TEXT")

# exact match: casts.ship_code = ship.ship_nodc
dbExecute(
  con,
  "
  UPDATE casts SET ship_key = (
    SELECT s.ship_key FROM ship s
    WHERE s.ship_nodc = casts.ship_code
    LIMIT 1)"
)

n_exact <- dbGetQuery(
  con,
  "SELECT COUNT(*) AS n FROM casts WHERE ship_key IS NOT NULL"
)$n
message(glue("Exact ship_nodc match: {n_exact} casts"))

# apply fuzzy match results for previously unmatched ships
matched_ships <- ship_matches |>
  filter(match_type != "unmatched", !is.na(matched_ship_key))

if (nrow(matched_ships) > 0) {
  for (i in seq_len(nrow(matched_ships))) {
    m <- matched_ships[i, ]
    dbExecute(
      con,
      glue(
        "
      UPDATE casts SET ship_key = '{m$matched_ship_key}'
      WHERE ship_code = '{m$ship_code}'
        AND ship_key IS NULL"
      )
    )
  }
  n_fuzzy <- dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM casts WHERE ship_key IS NOT NULL"
  )$n -
    n_exact
  message(glue("Fuzzy/manual match: {n_fuzzy} additional casts"))
}

# summary
ship_key_stats <- dbGetQuery(
  con,
  "
  SELECT
    CASE WHEN ship_key IS NULL THEN 'no_ship_match'
         ELSE 'matched' END AS status,
    COUNT(*) AS n_casts
  FROM casts
  GROUP BY status"
)
ship_key_stats |> datatable(caption = "ship_key assignment stats")
```

### Step 2: Derive cruise_key

Cruise key format: YYMMKK (2-digit year, 2-digit month, 2-letter ship
key). The bottle ingest stores the original `Cruise_ID` as `cruise_key_0`
(interim), so `cruise_key` is a new column bridging bottle casts to the
SWFSC `cruise` table.

```{r}
#| label: cruise_bridge_cruise_key

# add cruise_key column
dbExecute(con, "ALTER TABLE casts ADD COLUMN IF NOT EXISTS cruise_key TEXT")

# derive YYMMKK from datetime_utc + ship_key
dbExecute(
  con,
  "
  UPDATE casts SET cruise_key = CONCAT(
    LPAD(CAST(EXTRACT(YEAR FROM datetime_utc) % 100 AS VARCHAR), 2, '0'),
    LPAD(CAST(EXTRACT(MONTH FROM datetime_utc) AS VARCHAR), 2, '0'),
    ship_key)
  WHERE ship_key IS NOT NULL"
)

n_cruise <- dbGetQuery(
  con,
  "SELECT COUNT(*) AS n FROM casts WHERE cruise_key IS NOT NULL"
)$n
message(glue("Derived cruise_key for {n_cruise} casts"))
```

### Step 3: Validate against swfsc cruise table

```{r}
#| label: cruise_bridge_validate

cruise_stats <- dbGetQuery(
  con,
  "
  SELECT
    CASE
      WHEN c.ship_key IS NULL THEN 'no_ship_match'
      WHEN c.cruise_key IS NULL THEN 'no_cruise_key'
      WHEN cr.cruise_key IS NULL THEN 'no_cruise_match'
      ELSE 'matched'
    END AS status,
    COUNT(*) AS n_casts
  FROM casts c
  LEFT JOIN cruise cr ON c.cruise_key = cr.cruise_key
  GROUP BY status
  ORDER BY status"
)

cruise_stats |> datatable(caption = "Cruise bridge match statistics")
```

### Step 4: Report unmatched ship codes

```{r}
#| label: unmatched_ships

unmatched_report <- dbGetQuery(
  con,
  "
  SELECT DISTINCT
    c.ship_code, c.ship_name,
    COUNT(*) AS n_casts,
    MIN(c.datetime_utc) AS first_cast,
    MAX(c.datetime_utc) AS last_cast
  FROM casts c
  WHERE c.ship_key IS NULL
  GROUP BY c.ship_code, c.ship_name
  ORDER BY n_casts DESC"
)

if (nrow(unmatched_report) > 0) {
  unmatched_report |>
    datatable(caption = "Unmatched ship codes (no ship_key)")
} else {
  message("All ship codes matched!")
}
```

## Standardize Taxonomy

Update species table with WoRMS/ITIS/GBIF identifiers using local
lookups against `spp.duckdb` (MarineSensitivity species DB with full
WoRMS, ITIS, GBIF tables). Falls back to WoRMS API only for species not
found locally. Build taxonomy hierarchy via recursive CTEs.

```{r}
#| label: standardize_species

# check if species already standardized (all taxonomy columns exist with data)
sp_cols <- DBI::dbGetQuery(
  con,
  "SELECT column_name FROM information_schema.columns
   WHERE table_name = 'species'"
)$column_name
sp_standardized <- all(c("worms_id", "itis_id", "gbif_id") %in% sp_cols) &&
  DBI::dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM species WHERE gbif_id IS NOT NULL"
  )$n >
    0

if (sp_standardized && !overwrite) {
  message("Species already standardized, skipping")
  sp_results <- DBI::dbGetQuery(
    con,
    "SELECT species_id, scientific_name, worms_id, itis_id, gbif_id
     FROM species"
  )
} else {
  # add missing columns
  for (col in c("worms_id", "itis_id", "gbif_id")) {
    if (!col %in% sp_cols) {
      DBI::dbExecute(
        con,
        glue(
          "ALTER TABLE species ADD COLUMN {col} INTEGER"
        )
      )
    }
  }

  # attach MarineSensitivity species DB for local lookups
  DBI::dbExecute(
    con,
    glue(
      "ATTACH '{spp_db_path}' AS spp (READ_ONLY)"
    )
  )

  # worms: validate existing worms_id and resolve synonyms
  n_worms_before <- DBI::dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM species WHERE worms_id IS NOT NULL"
  )$n

  DBI::dbExecute(
    con,
    "
    UPDATE species SET worms_id = (
      SELECT CASE
        WHEN w.taxonomicStatus = 'accepted' THEN w.taxonID
        ELSE COALESCE(w.acceptedNameUsageID, w.taxonID)
      END
      FROM spp.worms w
      WHERE w.taxonID = species.worms_id
      LIMIT 1)
    WHERE worms_id IS NOT NULL"
  )

  n_worms <- DBI::dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM species WHERE worms_id IS NOT NULL"
  )$n
  message(glue(
    "WoRMS: {n_worms} species with validated worms_id ",
    "(was {n_worms_before})"
  ))

  # worms: name match for any still missing
  DBI::dbExecute(
    con,
    "
    UPDATE species SET worms_id = (
      SELECT CASE
        WHEN w.taxonomicStatus = 'accepted' THEN w.taxonID
        ELSE COALESCE(w.acceptedNameUsageID, w.taxonID)
      END
      FROM spp.worms w
      WHERE w.scientificName = species.scientific_name
        AND w.taxonRank = 'Species'
      ORDER BY CASE
        WHEN w.taxonomicStatus = 'accepted' THEN 0 ELSE 1 END
      LIMIT 1)
    WHERE worms_id IS NULL"
  )

  n_worms2 <- DBI::dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM species WHERE worms_id IS NOT NULL"
  )$n
  message(glue("WoRMS name match: {n_worms2 - n_worms} additional species"))

  # itis: match by scientific_name
  DBI::dbExecute(
    con,
    "
    UPDATE species SET itis_id = (
      SELECT COALESCE(i.acceptedNameUsageID, i.taxonID)
      FROM spp.itis i
      WHERE i.scientificName = species.scientific_name
      ORDER BY CASE
        WHEN i.taxonomicStatus = 'valid' THEN 0 ELSE 1 END
      LIMIT 1)
    WHERE itis_id IS NULL"
  )

  n_itis <- DBI::dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM species WHERE itis_id IS NOT NULL"
  )$n
  message(glue("ITIS: {n_itis} species matched"))

  # gbif: match by canonicalName
  DBI::dbExecute(
    con,
    "
    UPDATE species SET gbif_id = (
      SELECT COALESCE(
        TRY_CAST(NULLIF(g.acceptedNameUsageID, '') AS INTEGER),
        g.taxonID)
      FROM spp.gbif g
      WHERE g.canonicalName = species.scientific_name
      ORDER BY CASE
        WHEN g.taxonomicStatus = 'accepted' THEN 0 ELSE 1 END
      LIMIT 1)
    WHERE gbif_id IS NULL"
  )

  n_gbif <- DBI::dbGetQuery(
    con,
    "SELECT COUNT(*) AS n FROM species WHERE gbif_id IS NOT NULL"
  )$n
  message(glue("GBIF: {n_gbif} species matched"))

  # API fallback for any species still missing worms_id
  missing_worms <- DBI::dbGetQuery(
    con,
    "SELECT species_id, scientific_name
     FROM species WHERE worms_id IS NULL"
  )

  if (nrow(missing_worms) > 0) {
    message(glue(
      "API fallback for {nrow(missing_worms)} species missing worms_id"
    ))
    for (i in seq_len(nrow(missing_worms))) {
      sp_name <- missing_worms$scientific_name[i]
      sp_id <- missing_worms$species_id[i]

      worms_id <- tryCatch(
        {
          records <- worrms::wm_records_name(sp_name, fuzzy = TRUE)
          if (!is.null(records) && nrow(records) > 0) {
            rec <- records[1, ]
            if (!is.na(rec$valid_AphiaID)) {
              as.integer(rec$valid_AphiaID)
            } else {
              as.integer(rec$AphiaID)
            }
          } else {
            NA_integer_
          }
        },
        error = function(e) NA_integer_
      )

      if (!is.na(worms_id)) {
        DBI::dbExecute(
          con,
          glue(
            "UPDATE species SET worms_id = {worms_id}
           WHERE species_id = {sp_id}"
          )
        )
      }
      Sys.sleep(0.5)
    }
    message("API fallback complete")
  } else {
    message("All species matched locally, no API fallback needed")
  }

  sp_results <- DBI::dbGetQuery(
    con,
    "SELECT species_id, scientific_name, worms_id, itis_id, gbif_id
     FROM species"
  )
}

sp_results |>
  datatable(caption = "Species standardization results")
```

```{r}
#| label: build_taxon

# check if taxon table already exists with data
taxon_exists <- "taxon" %in%
  existing_tables &&
  DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM taxon")$n > 0

if (taxon_exists && !overwrite) {
  message("Taxon table already exists, skipping")
  taxon_rows <- DBI::dbReadTable(con, "taxon")
} else {
  # ensure spp.duckdb is attached
  spp_attached <- tryCatch(
    {
      DBI::dbGetQuery(con, "SELECT 1 FROM spp.worms LIMIT 1")
      TRUE
    },
    error = function(e) FALSE
  )
  if (!spp_attached) {
    DBI::dbExecute(
      con,
      glue(
        "ATTACH '{spp_db_path}' AS spp (READ_ONLY)"
      )
    )
  }

  # build WoRMS hierarchy via recursive CTE
  worms_taxon <- DBI::dbGetQuery(
    con,
    "
    WITH RECURSIVE hierarchy AS (
      SELECT
        w.taxonID,
        w.parentNameUsageID,
        w.scientificName,
        w.taxonRank,
        w.taxonID AS leaf_taxonID
      FROM spp.worms w
      WHERE w.taxonID IN (
        SELECT DISTINCT worms_id FROM species
        WHERE worms_id IS NOT NULL)

      UNION ALL

      SELECT
        p.taxonID,
        p.parentNameUsageID,
        p.scientificName,
        p.taxonRank,
        h.leaf_taxonID
      FROM spp.worms p
      JOIN hierarchy h ON p.taxonID = h.parentNameUsageID
      WHERE h.parentNameUsageID IS NOT NULL
        AND h.parentNameUsageID != h.taxonID
    )
    SELECT DISTINCT
      'WoRMS'                        AS authority,
      taxonID                        AS taxonID,
      leaf_taxonID                   AS acceptedNameUsageID,
      parentNameUsageID              AS parentNameUsageID,
      scientificName                 AS scientificName,
      taxonRank                      AS taxonRank,
      'accepted'                     AS taxonomicStatus,
      CAST(NULL AS VARCHAR)          AS scientificNameAuthorship
    FROM hierarchy
    WHERE taxonRank IS NOT NULL
  "
  )

  # deduplicate: keep one acceptedNameUsageID per (authority, taxonID)
  worms_taxon <- worms_taxon |>
    distinct(authority, taxonID, .keep_all = TRUE)
  message(glue("WoRMS hierarchy: {nrow(worms_taxon)} rows"))

  # build ITIS hierarchy via recursive CTE
  itis_taxon <- DBI::dbGetQuery(
    con,
    "
    WITH RECURSIVE hierarchy AS (
      SELECT
        i.taxonID,
        i.parentNameUsageID,
        i.scientificName,
        i.taxonRank,
        i.taxonID AS leaf_taxonID
      FROM spp.itis i
      WHERE i.taxonID IN (
        SELECT DISTINCT itis_id FROM species
        WHERE itis_id IS NOT NULL)

      UNION ALL

      SELECT
        p.taxonID,
        p.parentNameUsageID,
        p.scientificName,
        p.taxonRank,
        h.leaf_taxonID
      FROM spp.itis p
      JOIN hierarchy h ON p.taxonID = h.parentNameUsageID
      WHERE h.parentNameUsageID IS NOT NULL
        AND h.parentNameUsageID != h.taxonID
    )
    SELECT DISTINCT
      'ITIS'                         AS authority,
      taxonID                        AS taxonID,
      leaf_taxonID                   AS acceptedNameUsageID,
      parentNameUsageID              AS parentNameUsageID,
      scientificName                 AS scientificName,
      taxonRank                      AS taxonRank,
      'accepted'                     AS taxonomicStatus,
      CAST(NULL AS VARCHAR)          AS scientificNameAuthorship
    FROM hierarchy
    WHERE taxonRank IS NOT NULL
  "
  )

  itis_taxon <- itis_taxon |>
    distinct(authority, taxonID, .keep_all = TRUE)
  message(glue("ITIS hierarchy: {nrow(itis_taxon)} rows"))

  # combine and write
  taxon_rows <- bind_rows(worms_taxon, itis_taxon)

  DBI::dbExecute(con, "DROP TABLE IF EXISTS taxon")
  DBI::dbWriteTable(con, "taxon", taxon_rows)
  message(glue("Created taxon table: {nrow(taxon_rows)} rows"))

  # create taxa_rank lookup table (from calcofi4db::build_taxon_table)
  taxa_ranks_chr <- c(
    "Kingdom",
    "Subkingdom",
    "Phylum",
    "Subphylum",
    "Infraphylum",
    "Superclass",
    "Class",
    "Subclass",
    "Infraclass",
    "Megacohort",
    "Supercohort",
    "Cohort",
    "Subcohort",
    "Infracohort",
    "Superorder",
    "Order",
    "Suborder",
    "Infraorder",
    "Parvorder",
    "Superfamily",
    "Family",
    "Subfamily",
    "Supertribe",
    "Tribe",
    "Subtribe",
    "Genus",
    "Subgenus",
    "Series",
    "Subseries",
    "Species",
    "Subspecies",
    "Natio",
    "Mutatio",
    "Form",
    "Forma",
    "Subform",
    "Subforma",
    "Variety",
    "Subvariety",
    "Coll. sp.",
    "Aggr."
  )

  d_taxa_rank <- tibble(
    taxonRank = taxa_ranks_chr,
    rank_order = seq_along(taxa_ranks_chr)
  )

  DBI::dbExecute(con, "DROP TABLE IF EXISTS taxa_rank")
  DBI::dbWriteTable(con, "taxa_rank", d_taxa_rank)
  message(glue("Created taxa_rank table: {nrow(d_taxa_rank)} rank levels"))

  # detach spp.duckdb
  tryCatch(
    DBI::dbExecute(con, "DETACH spp"),
    error = function(e) NULL
  )
}

# show taxon stats
if (nrow(taxon_rows) > 0) {
  taxon_rows |>
    count(authority, taxonRank) |>
    arrange(authority, taxonRank) |>
    datatable(caption = "Taxon hierarchy by authority and rank")
}

# show taxa_rank lookup
dbReadTable(con, "taxa_rank") |>
  datatable(caption = "Taxa rank ordering")
```

### Taxonomy Statistics

```{r}
#| label: taxonomy_stats

n_species <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM species")$n
n_worms <- dbGetQuery(
  con,
  "SELECT COUNT(*) AS n FROM species WHERE worms_id IS NOT NULL"
)$n
n_itis <- dbGetQuery(
  con,
  "SELECT COUNT(*) AS n FROM species WHERE itis_id IS NOT NULL"
)$n
n_gbif <- dbGetQuery(
  con,
  "SELECT COUNT(*) AS n FROM species WHERE gbif_id IS NOT NULL"
)$n
n_taxon <- dbGetQuery(con, "SELECT COUNT(*) AS n FROM taxon")$n

tibble(
  metric = c(
    "total species",
    "with worms_id",
    "with itis_id",
    "with gbif_id",
    "taxon hierarchy rows"
  ),
  count = c(n_species, n_worms, n_itis, n_gbif, n_taxon)
) |>
  datatable(caption = "Taxonomy standardization summary")
```

## Validate Cross-Dataset Integrity

```{r}
#| label: validate

# grid_key integrity: casts.grid_key should all be in grid.grid_key
if (
  "grid_key" %in%
    dbListFields(con, "casts") &&
    "grid_key" %in% dbListFields(con, "grid")
) {
  grid_orphans <- dbGetQuery(
    con,
    "
    SELECT COUNT(*) AS n FROM casts c
    WHERE c.grid_key IS NOT NULL
      AND c.grid_key NOT IN (SELECT grid_key FROM grid)"
  )$n
  message(glue("Grid key orphans in casts: {grid_orphans}"))
}

# ship PK uniqueness
ship_dups <- dbGetQuery(
  con,
  "
  SELECT ship_key, COUNT(*) AS n FROM ship
  GROUP BY ship_key HAVING COUNT(*) > 1"
)
if (nrow(ship_dups) > 0) {
  warning(glue("Duplicate ship_key values: {nrow(ship_dups)}"))
}

# cruise PK uniqueness
cruise_dups <- dbGetQuery(
  con,
  "
  SELECT cruise_key, COUNT(*) AS n FROM cruise
  GROUP BY cruise_key HAVING COUNT(*) > 1"
)
if (nrow(cruise_dups) > 0) {
  warning(glue("Duplicate cruise_key values: {nrow(cruise_dups)}"))
}

# cruise bridge match statistics
bridge_stats <- dbGetQuery(
  con,
  "
  SELECT
    COUNT(*) AS total_casts,
    SUM(CASE WHEN ship_key IS NOT NULL THEN 1 ELSE 0 END) AS with_ship_key,
    SUM(CASE WHEN cruise_key IS NOT NULL THEN 1 ELSE 0 END) AS with_cruise_key
  FROM casts"
)
bridge_stats |> datatable(caption = "Cruise bridge coverage")

message("Cross-dataset validation complete")
```

## Show Combined Schema

```{r}
#| label: combined_schema
#| fig-width: 12
#| fig-height: 10

# build dm from connection (exclude internal tables)
tbls <- setdiff(dbListTables(con), c("_sp_update"))

d <- dm::dm_from_con(con, table_names = tbls, learn_keys = FALSE)

# set table colors by group
d <- d |>
  dm::dm_set_colors(
    lightblue = c(cruise, site, tow, net), # ichthyo chain
    lightyellow = c(ichthyo, species, lookup), # species/lookup
    lightgreen = c(grid, segment), # spatial
    pink = c(
      casts,
      bottle,
      bottle_measurement,
      cast_condition,
      measurement_type
    )
  ) # bottle chain

# add colors for new taxonomy tables if they exist
if ("taxon" %in% tbls) {
  d <- d |> dm::dm_set_colors(lightyellow = taxon)
}
if ("taxa_rank" %in% tbls) {
  d <- d |> dm::dm_set_colors(lightyellow = taxa_rank)
}
if ("ship" %in% tbls) {
  d <- d |> dm::dm_set_colors(lightblue = ship)
}

dm::dm_draw(d, rankdir = "LR", view_type = "all")
```

## Write Modified Parquet Outputs

Only export tables that differ from upstream (modified or new). Unchanged
tables reference upstream GCS URIs in the manifest.

```{r}
#| label: write_parquet

# tables modified by this workflow
modified_tables <- c(
  "casts", # added ship_key, cruise_key columns
  "species", # updated worms_id, itis_id, gbif_id
  "taxon", # new table (taxonomy hierarchy)
  "taxa_rank"
) # new table (rank ordering)

dir_pq_merged <- here("data/parquet/merge_ichthyo_bottle")

merge_stats <- write_parquet_outputs(
  con = con,
  output_dir = dir_pq_merged,
  tables = modified_tables,
  strip_provenance = FALSE
)

merge_stats |>
  mutate(file = basename(path)) |>
  select(-path) |>
  datatable(caption = "Modified parquet export statistics")
```

### Build Combined Manifest

Reference GCS URIs for unchanged upstream tables + local parquet for
modified tables.

```{r}
#| label: build_manifest

# read upstream manifests
ichthyo_manifest <- read_json(ichthyo_manifest_path)
bottle_manifest <- read_json(bottle_manifest_path)

# build combined manifest
manifest <- list(
  created_at = as.character(Sys.time()),
  release_version = release_version,
  modified_tables = merge_stats |>
    select(table, rows, file_size) |>
    as.list(),
  upstream_refs = list(
    ichthyo = list(
      manifest = ichthyo_manifest$created_at,
      tables = setdiff(
        ichthyo_manifest$tables,
        modified_tables
      )
    ),
    bottle = list(
      manifest = bottle_manifest$created_at,
      tables = setdiff(
        bottle_manifest$tables,
        c(modified_tables, "grid")
      )
    )
  )
) # grid canonical from swfsc

write_json(
  manifest,
  file.path(dir_pq_merged, "manifest.json"),
  auto_unbox = TRUE,
  pretty = TRUE
)

message(glue(
  "Manifest written: {length(modified_tables)} modified tables, ",
  "upstream refs for {length(manifest$upstream_refs$ichthyo$tables)} ",
  "ichthyo + {length(manifest$upstream_refs$bottle$tables)} bottle"
))
```

## Ingest to Working DuckLake

Load ALL tables into Working DuckLake with provenance. Unchanged tables
read directly from upstream parquet. Modified tables read from merge
output dir.

```{r}
#| label: ingest_working

con_wdl <- get_working_ducklake()
load_duckdb_extension(con_wdl, "spatial")

# helper to ingest a parquet file with provenance
ingest_pqt <- function(pqt_path, tbl_name, source_label) {
  data <- arrow::read_parquet(pqt_path)
  ingest_to_working(
    con = con_wdl,
    data = data,
    table = tbl_name,
    source_file = source_label,
    mode = "replace"
  )
}

# ingest ichthyo upstream tables (skip modified ones)
ichthyo_upstream <- setdiff(ichthyo_manifest$tables, modified_tables)
ichthyo_wdl <- map_dfr(ichthyo_upstream, function(tbl) {
  pqt <- file.path(dir_pq_ichthyo, paste0(tbl, ".parquet"))
  if (file.exists(pqt)) {
    ingest_pqt(
      pqt,
      tbl,
      glue("parquet/swfsc.noaa.gov_calcofi-db/{tbl}.parquet")
    )
  }
})

# ingest bottle upstream tables (skip modified + grid)
bottle_upstream <- setdiff(
  bottle_manifest$tables,
  c(modified_tables, "grid")
)
bottle_wdl <- map_dfr(bottle_upstream, function(tbl) {
  pqt <- file.path(dir_pq_bottle, paste0(tbl, ".parquet"))
  if (file.exists(pqt)) {
    ingest_pqt(
      pqt,
      tbl,
      glue("parquet/calcofi.org_bottle-database/{tbl}.parquet")
    )
  }
})

# ingest modified tables from merge output
merge_wdl <- map_dfr(modified_tables, function(tbl) {
  pqt <- file.path(dir_pq_merged, paste0(tbl, ".parquet"))
  if (file.exists(pqt)) {
    ingest_pqt(pqt, tbl, glue("parquet/merge_ichthyo_bottle/{tbl}.parquet"))
  }
})

# combine stats
wdl_stats <- bind_rows(
  ichthyo_wdl |> mutate(source = "ichthyo_upstream"),
  bottle_wdl |> mutate(source = "bottle_upstream"),
  merge_wdl |> mutate(source = "merge_modified")
)

wdl_stats |> datatable(caption = "Working DuckLake ingestion stats")
```

## List Working Tables

```{r}
#| label: list_working

working_tables <- list_working_tables(con_wdl)
working_tables |>
  datatable(caption = "Working DuckLake tables with provenance")
```

## Save Working DuckLake

```{r}
#| label: save_working

save_working_ducklake(con_wdl)
message("Working DuckLake saved to GCS")
```

## Create Frozen Release

Create a frozen (immutable) release for public access. Strips provenance
columns and exports clean parquet files. See
[Frozen DuckLake](https://ducklake.select/2025/10/24/frozen-ducklake/)
pattern.

```{r}
#| label: freeze_release

dir_frozen <- here(glue("data/releases/{release_version}"))
message(glue("Creating frozen release: {release_version}"))

# get tables to freeze (exclude internal tables)
tables_to_freeze <- DBI::dbListTables(con_wdl) |>
  setdiff(c("_meta"))

# create frozen parquet (strips provenance)
freeze_stats <- write_parquet_outputs(
  con = con_wdl,
  output_dir = file.path(dir_frozen, "parquet"),
  tables = tables_to_freeze,
  strip_provenance = TRUE,
  compression = "zstd"
)

freeze_stats |>
  datatable(caption = glue("Frozen release {release_version} statistics"))
```

### Release Notes

```{r}
#| label: release_notes
#| results: asis

# build release notes using paste() to avoid glue() parsing issues with code blocks
tables_list <- paste0(
  "- ",
  freeze_stats$table,
  " (",
  format(freeze_stats$rows, big.mark = ","),
  " rows)"
)

release_notes <- paste0(
  "# CalCOFI Database Release ",
  release_version,
  "\n\n",
  "**Release Date**: ",
  Sys.Date(),
  "\n\n",
  "## Tables Included\n\n",
  paste(tables_list, collapse = "\n"),
  "\n\n",
  "## Total\n\n",
  "- **Tables**: ",
  nrow(freeze_stats),
  "\n",
  "- **Total Rows**: ",
  format(sum(freeze_stats$rows), big.mark = ","),
  "\n",
  "- **Total Size**: ",
  round(sum(freeze_stats$file_size) / 1024 / 1024, 1),
  " MB\n\n",
  "## Data Sources\n\n",
  "- `ingest_swfsc.noaa.gov_calcofi-db.qmd` - Ichthyo tables (cruise, ship, site, tow, net, species, ichthyo, grid, segment, lookup)\n",
  "- `ingest_calcofi.org_bottle-database.qmd` - Bottle/cast tables (casts, bottle, bottle_measurement, cast_condition, measurement_type)\n\n",
  "## Cross-Dataset Integration\n\n",
  "- **Ship matching**: Reconciled ship codes between bottle casts and swfsc ship reference\n",
  "- **Cruise bridge**: Derived cruise_key (YYMMKK) for bottle casts via ship_key + datetime\n",
  "- **Taxonomy**: Standardized species with WoRMS AphiaID, ITIS TSN, GBIF backbone key\n",
  "- **Taxon hierarchy**: Built taxon + taxa_rank tables from WoRMS/ITIS classification\n\n",
  "## Access\n\n",
  "Parquet files can be queried directly from GCS:\n\n",
  "```r\n",
  "library(duckdb)\n",
  "con <- dbConnect(duckdb())\n",
  "dbExecute(con, 'INSTALL httpfs; LOAD httpfs;')\n",
  "dbGetQuery(con, \"\n",
  "  SELECT * FROM read_parquet(\n",
  "    'https://storage.googleapis.com/calcofi-db/ducklake/releases/",
  release_version,
  "/parquet/ichthyo.parquet')\n",
  "  LIMIT 10\")\n",
  "```\n\n",
  "Or use calcofi4r:\n\n",
  "```r\n",
  "library(calcofi4r)\n",
  "con <- cc_get_db(version = '",
  release_version,
  "')\n",
  "```\n"
)

writeLines(release_notes, file.path(dir_frozen, "RELEASE_NOTES.md"))
message(glue(
  "Release notes written to {file.path(dir_frozen, 'RELEASE_NOTES.md')}"
))
cat(release_notes)
```

### Upload Frozen Release to GCS

```{r}
#| label: upload_frozen

# upload frozen release to GCS (creates catalog.json, updates versions.json and latest.txt)
upload_frozen_release(
  release_dir = dir_frozen,
  version = release_version,
  set_latest = TRUE
)
```

## Cleanup

```{r}
#| label: cleanup

# close working ducklake connection
close_duckdb(con_wdl)
message("Working DuckLake connection closed")

# close local merge database
close_duckdb(con)
message("Local merge database connection closed")

# summary
message(glue("\n=== Summary ==="))
message(glue("Working DuckLake: saved to GCS"))
message(glue("Frozen release: {release_version} created at {dir_frozen}"))
message(glue("Tables: {nrow(freeze_stats)}"))
message(glue("Total rows: {format(sum(freeze_stats$rows), big.mark = ',')}"))
```




::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
