---
title: "Oceanographic Bottle Database"
author: "Ben Best, Cristina Dobbelaere"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

## Packages & setup
```{r}
# packages
if (!require("librarian")){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  dplyr, DT, dygraphs, glue,googlesheets4, gstat, here, lubridate, mapview, purrr, readr, 
  raster, rmapshaper, sf, skimr, stars, stringr, tidyr)
select <- dplyr::select
mapviewOptions(fgb = FALSE)
# paths
dir_data <- switch(
  Sys.info()["nodename"],
  `ben-mbpro` = "/Users/bbest/My Drive (ben@ecoquants.com)/projects/calcofi/data",
  `Cristinas-MacBook-Pro.local` = "/Volumes/GoogleDrive/.shortcut-targets-by-id/13pWB5x59WSBR0mr9jJjkx7rri9hlUsMv/calcofi/data",
  `Bens-MacBook-Air.local` = "/Volumes/GoogleDrive/My Drive/projects/calcofi/data")
  # TODO: get Erin's Google Drive path and "nodename")
```

## Read dataset keys from gdrive
```{r}
# bottle_key_csv <- file.path(dir_data, "/dataset_keys/bottle_field_descriptions.csv")
cast_key_csv   <- file.path(dir_data, "/_dataset-keys/cast_field_descriptions.csv")

key_cast   <- read_csv(cast_key_csv)

key_bottle <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/18c6eSGRf0bSdraDocjn3-j1rUIxN2WKqxsEnPQCR6rA/edit#gid=2046976359") %>% 
  na_if("n.a.") %>% 
  mutate(
    dataset    = "bottle_cast",
    source_url = source_url[1])


key_dic <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1SGfGMJUhiYZKIh746p5pGn7cD0cOcTA3g_imAnvyz88/edit#gid=0") %>% 
  na_if("n.a.") %>% 
  mutate(
    dataset    = "dic_cast",
    source_url = source_url[1])

data_key <- bind_rows(key_bottle, key_dic) %>% 
  select(dataset, field_name, title, description_abbv, everything())


# convert to var_lookup
var_lookup_key_tbl <- data_key %>% 
  filter(!is.na(description_abbv)) 
var_lookup <- var_lookup_key_tbl %>% 
  split(seq(nrow(.))) %>% 
  lapply(as.list)
names(var_lookup) <- var_lookup_key_tbl$field_name
```

## Read Oceano Bottle data from gdrive
```{r}
# get data file paths from gdrive
bottle_csv <- file.path(dir_data, "/oceanographic-data/bottle-database/CalCOFI_Database_194903-202001_csv_22Sep2021/194903-202001_Bottle.csv")
cast_csv <- file.path(dir_data, "/oceanographic-data/bottle-database/CalCOFI_Database_194903-202001_csv_22Sep2021/194903-202001_Cast.csv")
bottle_cast_rds <- file.path(dir_data, "/oceanographic-data/bottle-database/bottle_cast.rds")
dic_csv  <- file.path(dir_data, "/dic/CalCOFI_dics_200901-201507_28June2018.csv")

calcofi_geo           <- here("data/calcofi_oceano-bottle-stations_convex-hull.geojson")
calcofi_offshore_geo  <- here("data/calcofi_oceano-bottle-stations_convex-hull_offshore.geojson")
calcofi_nearshore_geo <- here("data/calcofi_oceano-bottle-stations_convex-hull_nearshore.geojson")

# check paths
stopifnot(dir.exists(dir_data))
stopifnot(any(file.exists(bottle_csv,cast_csv)))

# read data
d_cast   <- read_csv(cast_csv) %>% 
  separate(
    Sta_ID, c("Sta_ID_Line", "Sta_ID_Station"), 
    sep=" ", remove=F) %>% 
  mutate(
    Sta_ID_Line    = as.double(Sta_ID_Line),
    Sta_ID_Station = as.double(Sta_ID_Station))

d_bottle <- read_csv(bottle_csv, skip=1, col_names = F, guess_max = 1000000)
names(d_bottle) <- str_split(
  readLines(bottle_csv, n=1), ",")[[1]] %>% 
  str_replace("\xb5", "µ")

table(is.na(d_bottle$pH1))
# FALSE   TRUE 
#    84 889416 
table(is.na(d_bottle$pH2))
# FALSE   TRUE 
#    10 889490
table(is.na(d_bottle$O2ml_L))
#  FALSE   TRUE 
# 719993 169507
table(is.na(d_bottle$O2Sat))
# FALSE   TRUE 
# 685072 204428 
table(is.na(d_bottle[,'Oxy_µmol/Kg']))
#  FALSE   TRUE 
# 685061 204439

d_dic <- read_csv(dic_csv, skip=1, col_names = F, guess_max = 1000000)
names(d_dic) <- str_split(
  readLines(dic_csv, n=1), ",")[[1]] %>% 
  str_replace("\xb5", "µ")


bottle_cast <- d_cast %>% 
  right_join(
    d_bottle %>% select(-Sta_ID),
    by = "Cst_Cnt") %>% 
  mutate(Date = lubridate::as_date(Date, format = "%m/%d/%Y")) 

dic_cast <- d_cast %>% 
  right_join(
    d_dic %>% 
      select(-`Line Sta_ID`) %>% 
      rename(Depthm = `Depth(m)`),
    by = c("Cst_Cnt" = "ID")) %>% 
  mutate(Date = lubridate::as_date(Date, format = "%m/%d/%Y"))

# check if keys' field names are equal to names(d_bottle) and names(d_dic)
setequal(
  data_key %>% filter(dataset == "bottle_cast") %>% pull(field_name),
  names(d_bottle)) # TRUE
setequal(
  data_key %>% filter(dataset == "dic_cast") %>% pull(field_name),
  names(d_dic)) # TRUE

saveRDS(bottle_cast, bottle_cast_rds)
```

## Create `stations` dataset

```{r}
# filter out stations that are present in the data
stations <- bind_rows(bottle_cast, dic_cast) %>% 
  filter(
    !is.na(Lat_Dec), 
    !is.na(Lon_Dec)) %>% 
  select(Sta_ID, Sta_ID_Line, Sta_ID_Station, Sta_Code, Lat_Dec, Lon_Dec) %>% 
  distinct() %>% 
  group_by(Sta_ID) %>% 
  summarize(
    Lon            = mean(Lon_Dec),
    Lat            = mean(Lat_Dec),
    Sta_ID_Line    = mean(Sta_ID_Line),
    Sta_ID_Station = mean(Sta_ID_Station)) %>% 
  st_as_sf(
    coords = c("Lon", "Lat"), 
    crs    = 4326, 
    remove = F) %>% 
  mutate(
    offshore = ifelse(Sta_ID_Station > 60, T, F))
```

## Exploratory summary of bottle data

```{r, eval=F}
# d_cast
skim(d_cast)
```

```{r, eval=F}
# d_bottle
skim(d_bottle)
```

```{r}
# d_dic
skim(d_dic)
```

## Get CINMS AOI

```{r}
# get example AOI (Channel Islands NMS)
sanctuaries_geo <- "https://github.com/noaa-onms/onmsR/raw/12a87dfd4b90f2e3009ccb4913315fb2df7afddc/data-raw/sanctuaries.geojson"

cinms_ply <- sf::st_read(sanctuaries_geo) %>%
  dplyr::filter(nms == "CINMS")

# get AOI geom points as WKT for later use with API
cinms_txt <- sf::st_as_text(cinms_ply$geometry)
#cinms_txt

mapview(cinms_ply) 

# get aoi
aoi = cinms_ply
```

# Functions

### Get Station IDs as points (see `stations` dataset above)

```{r}
# for summary, want to group by Sta_Code because each data point has a diff Sta_ID
# get_pts <- function(data) {
#   data %>% 
#     filter(
#     !is.na(Lat_Dec),
#     !is.na(Lon_Dec)) %>% 
#   group_by(
#     Sta_ID) %>%
#   summarize(
#     lon            = mean(Lon_Dec),
#     lat            = mean(Lat_Dec),
#     Sta_ID_Line    = mean(Sta_ID_Line),
#     Sta_ID_Station = mean(Sta_ID_Station)) %>%
#   st_as_sf(
#     coords = c("lon", "lat"), crs=4326, remove = F) %>% 
#   mutate(
#     offshore = ifelse(Sta_ID_Station > 60, T, F))
# }


stations %>% mapview(zcol = "offshore")

# get_pts(bottle_cast) %>% mapview(zcol="offshore")
# get_pts(dic_cast) %>% mapview(zcol="offshore")

# pts %>% mapview(zcol="offshore")
# table(pts$offshore)
```

### Make CalCOFI total study areas

Per guidance from Erin:

>> CalCOFI Regions-- Station # > 60 = oceanic/offshore; <60 = neritic/nearshore/continental shelf (other regions include those from [Venrick et al. 2012]( http://calcofi.org/publications/calcofireports/v56/Vol56-Venrick.web.168-184.pdf) or [Stephens et al. 2018](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2018GB005930)

```{r}
make_hull <- function(x){
  st_union(x) %>% 
  st_convex_hull() %>% 
  st_buffer(0.1) %>%  # ~10 km
  ms_simplify(keep = 0.05) }

hull_geos <- c(calcofi_geo, calcofi_nearshore_geo, calcofi_offshore_geo)

if (any(!file.exists(hull_geos))){
  hull <- pts %>% 
    make_hull()
  st_write(hull, calcofi_geo)
  mapview(hull)
  
  hull_nearshore <- pts %>% 
    filter(!offshore) %>% 
    make_hull()
  st_write(hull_nearshore, calcofi_nearshore_geo)
  mapview(hull_nearshore)
  
  hull_offshore <- pts %>% 
    filter(offshore) %>% 
    make_hull()
  st_write(hull_offshore, calcofi_offshore_geo)
  mapview(hull_offshore)
}
```

## Get latest variable for AOI

```{r}
# # find stations in aoi
# pts_aoi <- pts %>% 
#   mutate(
#     x = st_intersects(pts, aoi) %>% as.logical()) %>% 
#   filter(x)

# mapview(aoi) +
#   mapview(pts_aoi)
```

### [Update dates](https://github.com/tidyverse/lubridate/issues/239#issuecomment-73107327) for summarizing by `date_step`

```{r}
update_date <- function(x, unit = c("second", "minute", "hour", "day", "week", "month", "quarter", "year", "decade")) {
  unit <- match.arg(unit)
  switch(unit,
    second  = update(x, seconds = floor(second(x))),
    minute  = update(x, seconds = 0),
    hour    = update(x, minutes = 0, seconds = 0),
    day     = update(x, hours   = 0, minutes = 0, seconds = 0),
    week    = update(x, wdays   = 1, hours = 0,   minutes = 0, seconds = 0),
    month   = update(x, mdays   = 1, hours = 0,   minutes = 0, seconds = 0),
    quarter = update(x, months  = ceiling(month(x)/3) * 3 - 2, mdays   = 1),
    year    = update(x, ydays   = 1,  hours = 0,  minutes = 0, seconds = 0),
    decade  = update(
      x, years = (year(x) - year(x) %% 10), 
      ydays = 1, hours = 0, minutes = 0, seconds = 0)
  )
}
```


### Summarize data by `date_step` in preparation for timeseries plot

```{r}
# choices for `date_step`: "day", "week", month", "quarter", "year", "decade"
get_oceano_var_aoi <- function(
  var, aoi, 
  date_step = c("year", "day", "week", "month", "quarter", "decade"), 
  depth_min = 0, depth_max = 10){
  
  # test values
  # var = "Bottle O2 (ml_L)"; aoi = cinms_ply; date_step = "year"; depth_min = 0; depth_max = 4000
  # var = "Salnty"; aoi = cinms_ply; date_step = "year"; depth_min = 0; depth_max = 1000
  
  d <- eval(parse(text = glue("var_lookup$`{var}`$dataset"))) %>% 
    as.name() %>% eval()
  
  # pts <- get_pts(d)
  pts <- stations
  
  # find stations in aoi
  pts_aoi <- pts %>% 
    mutate(
      x = st_intersects(pts, aoi) %>% as.logical()) %>% 
    filter(x)
  
  # d_var <- d %>% filter(!is.na(eval(parse(text = glue("d$`{var}`")))))
  
  d_summ     <- d %>% filter(!is.na(.data[[var]]))
  d_aoi_summ <- d_summ %>% filter(Sta_ID %in% pts_aoi$Sta_ID)

  # d_test <- d %>% filter(!is.na(`Bottle O2(ml_L)`))
  # d_test_aoi <- d_test %>% filter(Sta_ID %in% pts_aoi$Sta_ID)
  
  # d_aoi_summ <- d_aoi %>% 
  #   filter(!is.na(.data[[var]]))
  
  empty_data_for_var <- ifelse(nrow(d_aoi_summ) == 0, TRUE, FALSE)
  
  if (empty_data_for_var) {
    d_aoi_summ <- d_summ
  }
  if (any(!is.na(.data[[Depthm]]))) {
    d_aoi_summ <- d_aoi_summ %>% 
      filter(Depthm >= depth_min, Depthm < depth_max)
  }
  d_aoi_summ <- d_aoi_summ %>% 
    mutate(Date_Step = update_date(Date, unit = date_step)) %>% 
    group_by(Date_Step) %>% 
    summarize(
      var_n   = n(),
      var_min = min(.data[[var]], na.rm = T),
      var_q10 = quantile(.data[[var]], probs = 0.10, na.rm = T),
      var_avg = mean(.data[[var]], na.rm = T),
      var_q90 = quantile(.data[[var]], 0.90, na.rm = T),
      var_max = max(.data[[var]], na.rm = T),
      var_sd  = sd(.data[[var]], na.rm = T)) %>% 
    rename(Date = Date_Step)

  attr(d_aoi_summ, "labels")    <- eval(parse(text = glue("var_lookup$`{var}`")))
  attr(d_aoi_summ, "date_step") <- date_step
  attr(d_aoi_summ, "date_msg")  <- glue("This dataset was summarized by {date_step}.")
  attr(d_aoi_summ, "aoi") <- ifelse(
    empty_data_for_var, 
    glue("No data were found for {var} in this area of interest. Summaries were conducted across all existing data points."),
    glue("Data for {var} in selected area of interest")
  )
  
  d_aoi_summ
}
```

### Plot time series
see: [plot_metric_timeseries()](https://github.com/noaa-onms/onmsR/blob/2e438a9bdff8ee90b8fc811aafae4520c26049ab/R/spatial.R#L16-L49)

```{r}
plot_timeseries <- function(d) {
  x <- d %>% select(
    Date, 
    `10% quantile` = var_q10,
    `Average`      = var_avg,
    `90% quantile` = var_q90)
  var_attrs <- tibble(
    title = attributes(d)$labels$title,
    var   = attributes(d)$labels$description_abbv)
  xts::xts(x = x %>% select(-Date), order.by = x %>% pull(Date)) %>% 
    dygraph(
      main = var_attrs$title,
      xlab = "Date", ylab = var_attrs$var) %>% # ...) %>%
    dySeries(
      c("10% quantile", "Average", "90% quantile"), 
      label = var_attrs$var, color = "Red") %>%
    dyRangeSelector(fillColor = "#FFFFFF", strokeColor = "#FFFFFF")
}
```

### Get IDW raster without AOI (based on Cruise_ID)

```{r}
get_oceano_var_cruise_raster <- function(cruise_id, var, depth_min, depth_max){
  # test values
  # cruise_id = "2020-01-05-C-33RL"; var = "T_degC"; depth_min = 0; depth_max = 10
  # cruise_id = "1949-03-01-C-31CR"; var = "Bottle O2(ml_L)"; depth_min = 0; depth_max = 1000
  
  d <- eval(parse(text = glue("var_lookup$`{var}`$data_source_name"))) %>% 
    as.name() %>% eval()
  
  d_var <- d %>% 
    filter(
      !is.na(.data[[var]]),
      Depthm >= !!depth_min,
      Depthm < !!depth_max)
  
  cruise_id_choices <- unique(d_var$Cruise_ID) # Cruise_ID values with data for var
  
  d_daily <- d_var %>% 
    filter(Cruise_ID == cruise_id) %>% 
    group_by(Cruise_ID, Date, Sta_ID) %>% 
    summarize(
      var_n   = n(),
      var_avg = mean(.data[[var]], na.rm = T), 
      .groups = "drop") %>% 
    arrange(desc(Date), Cruise_ID)
  d_daily
  
  pts <- stations
  # pts <- get_pts(d)
  
  # find stations in aoi
  pts_aoi <- pts %>% 
    mutate(
      x = st_intersects(pts, aoi) %>% as.logical()) %>% 
    filter(x)
  
  p <- pts %>% 
    left_join(
      d_daily,
      by="Sta_ID") %>% 
    select(Sta_ID, Date, var_avg) %>% 
    filter(!is.na(var_avg))
  mapview(p, zcol="var_avg")
  
  h <- st_convex_hull(st_union(p)) %>% st_as_sf() %>% 
    mutate(one = 1)
  mapview(h)
  r <- raster(as_Spatial(h), res=0.1, crs=4326)
  z <- rasterize(as_Spatial(h), r, "one")
  
  # inverse distance weighted interpolation
  #   https://rspatial.org/raster/analysis/4-interpolation.html
  gs <- gstat(formula=var_avg~1, locations=p)
  idw <- interpolate(z, gs)
  w <- mask(idw, z)
  
  w_tif <- here(glue("data/idw_{var}_{cruise_id}_{depth_min}_{depth_max}"))
  # w_tif <- here("data/_test_idw.tif")
  raster::writeRaster(w, w_tif, overwrite=T)
  w <- raster(w_tif) %>% readAll()
  unlink(w_tif)
  
  w
}
```

# Example timeseries plots

### TEMPERATURE

```{r}
get_oceano_var_aoi("T_degC", cinms_ply, "year", 0, 20) %>% 
  plot_timeseries()
```

### SALINITY

```{r}
get_oceano_var_aoi("Salnty", cinms_ply, "year", 0, 4000) %>% 
  plot_timeseries()
```

### OXYGEN (HYPOXIA)

```{r}
get_oceano_var_aoi("O2Sat", cinms_ply, "year", 0, 2000) %>% 
  plot_timeseries()

# only contains data for 1949-1950
get_oceano_var_aoi("Bottle O2(ml_L)", cinms_ply, "week", 0, 4000) %>% 
  plot_timeseries()
get_oceano_var_aoi("Bottle O2 (µmol/Kg)", cinms_ply, "week", 0, 4000) %>% 
  plot_timeseries()
```

### dic

```{r}
get_oceano_var_aoi("DIC1", cinms_ply, "month", 0, 4000) %>% 
  plot_timeseries()
```

### pH (OA)

#### [ ] TODO: find pH data for OA measurements

```{r}
# need to find pH data:
dic_cast %>% filter(!is.na(pH1 | pH2)) %>% nrow()
bottle_cast %>% filter(!is.na(pH1 | pH2)) %>% nrow()
# get_oceano_var_aoi("pH1", cinms_ply, "week", 0, 4000) %>% 
#   plot_timeseries()
```


# Example IDW rasters of variables for latest (not within AOI)

### TEMPERATURE

```{r}
get_oceano_var_cruise_raster(
  cruise_id = "2020-01-05-C-33RL",
  var       = "T_degC", 
  depth_min = 0,
  depth_max = 10) %>% 
  mapview()
```

### SALINITY

```{r}
get_oceano_var_cruise_raster(
  cruise_id = "2020-01-05-C-33RL",
  var       = "Salnty", 
  depth_min = 0,
  depth_max = 20) %>% 
  mapview()
```

### OXYGEN

```{r}
get_oceano_var_cruise_raster(
  cruise_id = "1949-03-01-C-31CR",
  var       = "Bottle O2(ml_L)", 
  depth_min = 0,
  depth_max = 8000) %>% 
  mapview()
```

### dic

```{r}
get_oceano_var_cruise_raster(
  cruise_id = "1949-03-01-C-31CR",
  var       = "dic1", 
  depth_min = 0,
  depth_max = 8000) %>% 
  mapview()
```


# BIOLOGICAL DATA

## Read whales, seabirds, turtles data from gdrive

### [Bird Mammal Census](https://oceaninformatics.ucsd.edu/datazoo/catalogs/ccelter/datasets/255)

```{r}
bird_mamm_csv <- file.path(dir_data, "/whales-seabirds-turtles/bird-mammal-census/CalCOFI_bird-mammal-census_observations.csv")
transects_csv <- file.path(dir_data, "/whales-seabirds-turtles/bird-mammal-census/CalCOFI_bird-mammal-census_transects.csv")
key_bird_mamm_beh_csv <- file.path(dir_data, "/whales-seabirds-turtles/bird-mammal-census/CalCOFI_bird-mammal-census_behaviorcodes.csv")
key_bird_mamm_sp_csv <- file.path(dir_data, "/whales-seabirds-turtles/bird-mammal-census/CalCOFI_bird-mammal-census_allspecieslist.csv")

key_bird_mamm_beh <- read_csv(key_bird_mamm_beh_csv) 
key_bird_mamm_sp  <- read_csv(key_bird_mamm_sp_csv) 


# bird & mammal counts from CalCOFI cruises
d_bird_mamm <- read_csv(bird_mamm_csv, guess_max = 1000000)

# log of transect datetimes & locations from CalCOFI cruises
d_transects <- read_csv(transects_csv, guess_max = 1000000)

bird_mamm_census <- d_bird_mamm %>% 
  left_join(
    d_transects, 
    by = "GIS key") %>% 
  left_join(
    key_bird_mamm_beh %>% 
      select(Behavior, Behavior_Description = Description), 
    by = "Behavior") %>%
  left_join(
    key_bird_mamm_sp %>% 
      mutate(across(
        c("Bird", "LargeBird", "Fish", "Mammal", "Include", "Unidentified"), 
        as.logical)), 
    by = "Species")
```


# OLD

## Get raster of variable for latest within AOI

```{r}
# test variables
var = "T_degC"; aoi = cinms_ply; depth_min = 0; depth_max = 100

d <- eval(parse(text = glue("var_lookup$`{var}`$data_source_name"))) %>% 
  as.name() %>% eval()

pts <- get_pts(d)

# find stations in aoi
pts_aoi <- pts %>%
  mutate(
    x = st_intersects(pts, aoi) %>% as.logical()) %>%
  filter(x)

d_aoi <- d %>%
  filter(Sta_ID %in% pts_aoi$Sta_ID)

d_aoi_daily <- d_aoi %>%
  filter(
    !is.na(.data[[var]]),
    Depthm >= depth_min,
    Depthm < depth_max) %>%
  group_by(Date, Sta_ID) %>%
  summarize(
    var_n   = n(),
    var_avg = mean(.data[[var]], na.rm = T),
    .groups = "drop") %>%
  arrange(desc(Date))
d_aoi_daily
```

## TODO

summarize:

- [x] OCEAN TEMP: `T_degC`
- [x] salinity: `Salnty`
- [x] OXYGEN (HYPOXIA): `Bottle O2(ml_L)`, `Bottle O2 (µmol/Kg)`
- [ ] dic: 
- [ ] ZOOPLANKTON,
- [ ] ICHTHYOPLANKTON for each station ID

- Create `read_bottle()` in `calcofi4r`
