---
title: "Publish every dataset to ERDDAP"
subtitle: "One config row per dataset_key over the core, served from DuckDB views"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_to_erddap
  workflow_type: publish
  dependency:
    - release_database
  output: data/erddap/datasets_calcofi.xml
  workflow_url: https://calcofi.io/workflows/publish_to-erddap.html
editor_options:
  chunk_output_type: console
---

## Overview

Generate the ERDDAP configuration for **every** dataset in the frozen release,
discovered from `dataset_key`'s presence in the core schema rather than from a
hand-maintained table list.

This replaces `publish_calcofi_to_erddap.qmd`, which was **already broken and
already prescribed this design**. Its own callout said every `parquet` path in it
pointed at a per-dataset table the ingests no longer publish, that it "will fail
until the config is repointed", and that:

> `sample` carries `time`/`latitude`/`longitude`/`depth_min_m`/`depth_max_m`
> directly … **one config row per `dataset_key` over `sample`/`obs` replaces the
> per-table list.**

Two things changed besides the repointing:

**1 · `EDDTableFromDatabase` over DuckDB, not `EDDTableFromParquetFiles`.** ERDDAP
streams filtered results from DuckDB — predicate pushdown, partition pruning, disk
spill — instead of loading whole Parquet files into the JVM heap. That is what
fixed the OOM which killed `ctd_wide` at 4, 5 **and** 6 GB container sizes; see
[the serving benchmark](https://calcofi.io/workflows/bench_erddap_ctd.html).

**2 · Every view is executed locally before its XML is written.** The old config's
failure mode was silent: paths that no longer existed, discovered only when ERDDAP
logged "Bad line(s)". Here each view is run against the real release, its row count
and coordinate coverage recorded, and **a view that returns nothing does not get a
dataset block**.

::: {.callout-note title="What gets served"}
Per `dataset_key`, up to four ERDDAP datasets, each only when it has rows:

| datasetID | grain | source |
|---|---|---|
| `{dataset_key}` | one row per observation (occurrence × measurement) | `obs` + `taxon` + `sample` |
| `{dataset_key}_sample` | one row per sampling event, effort widened onto it | `sample` + `sample_measurement` |
| `{dataset_key}_attribute` | sub-occurrence detail (length/stage bins) | `obs_attribute` + `sample` |
| `{dataset_key}_full` | the full series *before* thinning | a supplemental table |

The old datasetIDs (`calcofi_ctd`, `calcofi_casts`, `calcofi_ctd_thin`, …) are
**not** preserved — these are `dataset_key`-based and self-describing. Deploy is
manual and selective, so the changeover is a decision made at splice time, not
here.
:::

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(DBI, duckdb, dplyr, fs, glue, jsonlite, purrr, readr, stringr,
                 tibble, tidyr, here, knitr, quiet = TRUE)
here <- here::here
options(readr.show_col_types = FALSE)
devtools::load_all(here::here("../calcofi4db"))
source(here("libs/erddap.R"))
source(here("libs/erddap_duckdb.R"))
source(here("libs/publish_netcdf.R"))   # cc_release_version(), cc_release_parquet()

RELEASE <- cc_release_version()
dir_erddap <- here("data/erddap"); dir_create(dir_erddap)

# LOCAL release parquet — the same bytes as the promoted release, read from disk so
# every view below can actually be executed before its config is written.
PQ_LOCAL <- here(glue("data/releases/{RELEASE}/parquet"))

# The path ERDDAP will see. A DuckDB view binds its parquet paths LITERALLY, so the
# deployed .db must be built against the server's path, not this machine's — an
# identity mount. Two .db files are therefore produced: one with local paths for
# validation, one with server paths for deploy.
ERDDAP_ROOT <- Sys.getenv("CALCOFI_ERDDAP_ROOT", "/share/erddap/data")
PQ_SERVER   <- glue("{ERDDAP_ROOT}/release/{RELEASE}/parquet")
ING_SERVER  <- glue("{ERDDAP_ROOT}/ingest")
ING_LOCAL   <- here("data/parquet")

cat(glue("release       : {RELEASE}\n"))
cat(glue("local parquet : {PQ_LOCAL} ({ifelse(dir_exists(PQ_LOCAL), 'present', 'MISSING')})\n"))
cat(glue("server parquet: {PQ_SERVER}\n"))
cat(glue("output        : {dir_erddap}\n"))
stopifnot("local release parquet is required to validate the views" = dir_exists(PQ_LOCAL))
```

```{r}
#| label: connect
con <- dbConnect(duckdb())
for (s in c("SET memory_limit='8GB'", "SET enable_progress_bar=false"))
  try(dbExecute(con, s), silent = TRUE)
q <- function(sql, ...) dbGetQuery(con, glue(sql, ..., .envir = parent.frame()))
pq <- function(root, tbl) if (dir_exists(file.path(PQ_LOCAL, tbl)))
  glue("{root}/{tbl}/**/*.parquet") else glue("{root}/{tbl}.parquet")
```

## Discover what the core holds per dataset

```{r}
#| label: discover
core_tbls <- c("sample", "obs", "obs_attribute", "sample_measurement")
presence <- bind_rows(lapply(core_tbls, function(t) {
  src <- pq(PQ_LOCAL, t)
  q("SELECT '{t}' AS tbl, dataset_key, count(*) AS n
     FROM read_parquet('{src}', hive_partitioning = true, union_by_name = true)
     GROUP BY 1, 2")
}))
pres_wide <- presence |>
  tidyr::pivot_wider(names_from = tbl, values_from = n, values_fill = 0) |>
  arrange(dataset_key)
kable(pres_wide, caption = glue("Core rows per dataset in {RELEASE}"))
```

```{r}
#| label: discover-supplemental
# A `_full` variant is whatever an ingest declares `supplemental: true` in its
# `tables_owned` — the same YAML the netCDF publisher reads. Declared is not the
# same as usable, so each one is checked below rather than trusted.
iy <- read_ingest_yaml(here())
supp <- bind_rows(lapply(names(iy), function(ds) {
  to <- iy[[ds]]$tables_owned
  if (is.null(to)) return(NULL)
  hit <- Filter(function(t) isTRUE(t$supplemental), to)
  if (!length(hit)) return(NULL)
  tibble(dataset_key = ds, table = vapply(hit, function(t) t$table, character(1)),
         note = vapply(hit, function(t) t$note %||% "", character(1)))
}))
supp <- supp |>
  mutate(
    in_release = dir_exists(file.path(PQ_LOCAL, table)) |
                 file_exists(file.path(PQ_LOCAL, glue("{table}.parquet"))),
    in_ingest  = dir_exists(file.path(ING_LOCAL, dataset_key, table)) |
                 file_exists(file.path(ING_LOCAL, dataset_key, glue("{table}.parquet"))))
kable(supp, caption = "Supplemental (pre-thinning) tables declared by the ingests")
```

```{r}
#| label: supplemental-usability
# ERDDAP is a tabular server keyed on time/latitude/longitude. A supplemental table
# that carries neither its own coordinates nor a resolvable link to an event that
# has them cannot be served, no matter that it is published.
supp_check <- bind_rows(lapply(seq_len(nrow(supp)), function(i) {
  r <- supp[i, ]
  root <- if (r$in_release) PQ_LOCAL else file.path(ING_LOCAL, r$dataset_key)
  src  <- if (dir_exists(file.path(root, r$table)))
    glue("{root}/{r$table}/**/*.parquet") else glue("{root}/{r$table}.parquet")
  if (!r$in_release && !r$in_ingest)
    return(tibble(dataset_key = r$dataset_key, table = r$table, rows = NA_real_,
                  has_coords = NA, servable = FALSE, why = "not found locally"))
  cols <- q("DESCRIBE SELECT * FROM read_parquet('{src}', hive_partitioning = true,
             union_by_name = true)")$column_name
  has_coords <- all(c("latitude", "longitude", "datetime") %in% cols)
  n <- q("SELECT count(*) AS n FROM read_parquet('{src}', hive_partitioning = true,
          union_by_name = true)")$n
  # no coordinates of its own: can the rows reach an event that has them?
  linked <- NA_real_
  if (!has_coords) {
    fk <- grep("_uuid$|^sample_key$", cols, value = TRUE)
    fk <- setdiff(fk, grep(glue("^{r$table}"), fk, value = TRUE))
    if (length(fk)) {
      st <- (iy[[r$dataset_key]]$netcdf$sample_type %||%
               q("SELECT sample_type FROM read_parquet('{pq(PQ_LOCAL,\"sample\")}')
                  WHERE dataset_key = '{r$dataset_key}' LIMIT 1")$sample_type)
      linked <- q("
        SELECT count(s.sample_key) AS n FROM (
          SELECT DISTINCT {fk[1]} AS k FROM read_parquet('{src}',
                 hive_partitioning = true, union_by_name = true)) m
        LEFT JOIN (SELECT sample_key FROM read_parquet('{pq(PQ_LOCAL,\"sample\")}')
                   WHERE dataset_key = '{r$dataset_key}') s
          ON s.sample_key = '{r$dataset_key}:{st}:' || m.k")$n
      tot <- q("SELECT count(DISTINCT {fk[1]}) AS n FROM read_parquet('{src}',
                hive_partitioning = true, union_by_name = true)")$n
      linked <- linked / tot
    }
  }
  tibble(dataset_key = r$dataset_key, table = r$table, rows = as.numeric(n),
         has_coords = has_coords,
         servable = has_coords || (!is.na(linked) && linked > 0.99),
         why = if (has_coords) "carries its own time/lat/lon"
               else if (is.na(linked)) "no coordinates and no resolvable event link"
               else glue("{round(100*linked, 1)}% of its events resolve in `sample`"))
}))
kable(supp_check, caption = "Can each supplemental table be served?")
```

::: {.callout-warning title="`mets_measurement` is published but not servable"}
The full ~1-minute METS series carries **no coordinates of its own** — only
`mets_sample_uuid`, `measurement_type`, `measurement_value`, `cruise_key` — and
its event table was never published: the mets ingest emits
`sample`/`obs`/`mets_measurement`, where `sample` holds only the **thinned**
events. So of its **2,366,547** distinct underway events, just **77,795 (3.3%)**
resolve to a `sample` row.

A `calcofi_mets_full` dataset would therefore serve 20.6 M measurements of which
96.7% had no time or position. It is **excluded**, and the fix is upstream: the
mets ingest needs to publish the full underway event table (positions and times
for all 2.37 M records), not only the thinned subset. Filed as a finding rather
than worked around here, because inventing coordinates is worse than serving less.

`obs_ctd_full` has the opposite property — it carries `latitude`/`longitude`/
`datetime`/`cruise_key` denormalized on every row — so `calcofi_ctd-cast_full`
serves fine.
:::

## Build the view definitions

```{r}
#| label: view-sql
# One SQL builder per grain. `root` is substituted twice over: once with local
# paths to VALIDATE, once with the server's paths to DEPLOY.
mt_units <- {
  mt <- read_measurement_type(here("metadata/measurement_type.csv"))
  setNames(as.list(mt$units), mt$measurement_type)
}

sql_obs <- function(ds, root) glue("
  SELECT o.sample_key, o.cruise_key, o.grid_key, s.site_key, s.sample_type,
         o.datetime                  AS time,
         o.latitude::DOUBLE          AS latitude,
         o.longitude::DOUBLE         AS longitude,
         o.depth_min_m::DOUBLE       AS depth,
         o.depth_max_m::DOUBLE       AS depth_max_m,
         o.taxon_key, t.scientific_name, o.life_stage,
         o.measurement_type,
         o.measurement_value::DOUBLE AS measurement_value,
         o.measurement_qual
  FROM read_parquet('{root}/obs/dataset_key={ds}/*.parquet') o
  LEFT JOIN read_parquet('{root}/taxon.parquet') t USING (taxon_key)
  LEFT JOIN (SELECT sample_key, site_key, sample_type
             FROM read_parquet('{root}/sample.parquet')) s USING (sample_key)")

sql_sample <- function(ds, root, eff = character()) {
  # effort widened onto the event: one column per sample_measurement type is far
  # more usable than a long table with no coordinates of its own
  piv <- if (length(eff)) paste0(",\n         ", paste(glue(
    "m.\"{eff}\""), collapse = ",\n         ")) else ""
  jn <- if (length(eff)) {
    sel <- paste(glue("MAX(measurement_value) FILTER (WHERE measurement_type = '{eff}')",
                      "::DOUBLE AS \"{eff}\""), collapse = ",\n             ")
    # paste0 the leading break on AFTER glue: glue's .trim strips a leading newline
    # and indentation, which welded this clause onto the preceding alias
    # (`... ) sLEFT JOIN ...`) and made both effort-bearing datasets fail to bind
    paste0("\n  ", glue("
  LEFT JOIN (SELECT sample_key,
             {sel}
             FROM read_parquet('{root}/sample_measurement.parquet')
             WHERE dataset_key = '{ds}' GROUP BY sample_key) m USING (sample_key)"))
  } else ""
  glue("
  SELECT s.sample_key, s.sample_type, s.parent_sample_key, s.cruise_key,
         s.grid_key, s.site_key, s.order_occ,
         s.datetime            AS time,
         s.latitude::DOUBLE    AS latitude,
         s.longitude::DOUBLE   AS longitude,
         s.depth_min_m::DOUBLE AS depth,
         s.depth_max_m::DOUBLE AS depth_max_m,
         s.tow_type{piv}
  FROM (SELECT sample_key, sample_type, parent_sample_key, cruise_key, grid_key,
               site_key, order_occ, datetime, latitude, longitude, depth_min_m,
               depth_max_m, tow_type
        FROM read_parquet('{root}/sample.parquet')
        WHERE dataset_key = '{ds}') s{jn}")
}

sql_attribute <- function(ds, root) glue("
  SELECT a.sample_key, a.taxon_key, t.scientific_name, a.life_stage,
         a.measurement_type, a.bin_value, a.bin_label, a.count, a.measurement_qual,
         s.cruise_key, s.grid_key, s.site_key,
         s.datetime            AS time,
         s.latitude::DOUBLE    AS latitude,
         s.longitude::DOUBLE   AS longitude,
         s.depth_min_m::DOUBLE AS depth
  FROM (SELECT * FROM read_parquet('{root}/obs_attribute.parquet')
        WHERE dataset_key = '{ds}') a
  LEFT JOIN read_parquet('{root}/taxon.parquet') t USING (taxon_key)
  LEFT JOIN (SELECT sample_key, cruise_key, grid_key, site_key, datetime,
                    latitude, longitude, depth_min_m
             FROM read_parquet('{root}/sample.parquet')) s USING (sample_key)")

sql_full <- function(ds, tbl, root) glue("
  SELECT f.sample_key, f.cruise_key, f.grid_key,
         f.datetime                  AS time,
         f.latitude::DOUBLE          AS latitude,
         f.longitude::DOUBLE         AS longitude,
         f.depth_min_m::DOUBLE       AS depth,
         f.measurement_type,
         f.measurement_value::DOUBLE AS measurement_value,
         f.measurement_qual
  FROM read_parquet('{root}/{tbl}/**/*.parquet', hive_partitioning = true) f
  WHERE f.dataset_key = '{ds}'")
```

```{r}
#| label: plan
ds_all <- sort(unique(presence$dataset_key))
n_of <- function(ds, tbl) {
  v <- pres_wide[[tbl]][pres_wide$dataset_key == ds]
  if (!length(v)) 0 else v
}
eff_of <- function(ds) q("
  SELECT DISTINCT measurement_type FROM read_parquet('{pq(PQ_LOCAL,\"sample_measurement\")}')
  WHERE dataset_key = '{ds}' ORDER BY 1")$measurement_type

cfg <- bind_rows(lapply(ds_all, function(ds) {
  rows <- list()
  if (n_of(ds, "obs") > 0)
    rows <- c(rows, list(tibble(dataset_id = ds, grain = "obs",
                                src_rows = n_of(ds, "obs"))))
  if (n_of(ds, "sample") > 0)
    rows <- c(rows, list(tibble(dataset_id = glue("{ds}_sample"), grain = "sample",
                                src_rows = n_of(ds, "sample"))))
  if (n_of(ds, "obs_attribute") > 0)
    rows <- c(rows, list(tibble(dataset_id = glue("{ds}_attribute"),
                                grain = "obs_attribute",
                                src_rows = n_of(ds, "obs_attribute"))))
  if (!length(rows)) return(NULL)
  bind_rows(rows) |> mutate(dataset_key = ds, .before = 1)
}))
# servable supplemental tables only
cfg <- bind_rows(cfg, supp_check |>
  filter(servable) |>
  transmute(dataset_key, dataset_id = glue("{dataset_key}_full"), grain = "full",
            src_rows = rows, supp_table = table))
cfg <- cfg |> arrange(dataset_key, grain)
kable(cfg, caption = glue("{nrow(cfg)} ERDDAP datasets planned"))
```

## Validate every view against the real release

```{r}
#| label: validate
# `root` is the release parquet; `ing_root` is where an ingest's own outputs live.
# A supplemental table reads from whichever actually holds it, named explicitly
# rather than derived by path arithmetic.
build_sql <- function(r, root, ing_root = ING_LOCAL) {
  supp_root <- function(tbl, ds) {
    in_rel <- any(supp$in_release[supp$table == tbl], na.rm = TRUE)
    if (in_rel) root else glue("{ing_root}/{ds}")
  }
  switch(r$grain,
    obs           = sql_obs(r$dataset_key, root),
    sample        = sql_sample(r$dataset_key, root, eff_of(r$dataset_key)),
    obs_attribute = sql_attribute(r$dataset_key, root),
    full          = sql_full(r$dataset_key, r$supp_table,
                             supp_root(r$supp_table, r$dataset_key)))
}

val <- bind_rows(lapply(seq_len(nrow(cfg)), function(i) {
  r <- cfg[i, ]
  # the `full` grain reads a partitioned multi-GB table; count from a LIMIT-free
  # aggregate but probe coordinates from a sample, so validation stays cheap
  s <- build_sql(r, PQ_LOCAL)
  out <- tryCatch({
    d <- q("SELECT count(*) AS n, count(time) AS n_time, count(latitude) AS n_lat,
                   min(epoch(time)) AS t0, max(epoch(time)) AS t1
            FROM ({s})")
    tibble(dataset_id = r$dataset_id, rows = as.numeric(d$n),
           pct_time = round(100 * d$n_time / pmax(d$n, 1), 1),
           pct_coord = round(100 * d$n_lat / pmax(d$n, 1), 1),
           t_start = as.character(as.POSIXct(d$t0, origin = "1970-01-01", tz = "UTC")),
           t_end   = as.character(as.POSIXct(d$t1, origin = "1970-01-01", tz = "UTC")),
           ok = d$n > 0, err = NA_character_)
  }, error = function(e) tibble(
    dataset_id = r$dataset_id, rows = NA_real_, pct_time = NA_real_,
    pct_coord = NA_real_, t_start = NA_character_, t_end = NA_character_,
    ok = FALSE, err = conditionMessage(e)))
  out
}))
kable(val, caption = "Every view executed against the local release parquet")

# A view that returns nothing, or that errors, must not become a dataset block —
# that is exactly how the previous config came to point at tables that no longer
# existed and was only discovered from ERDDAP's own error log.
if (any(!val$ok)) {
  cat("\nEXCLUDED (empty or failing):\n")
  for (i in which(!val$ok))
    cat(glue("  - {val$dataset_id[i]}: {val$err[i] %||% 'zero rows'}\n"))
}
cfg <- cfg |> semi_join(val |> filter(ok) |> select(dataset_id), by = "dataset_id")
```

```{r}
#| label: coord-warning
low <- val |> filter(ok, pct_coord < 99 | pct_time < 99)
if (nrow(low)) {
  kable(low |> select(dataset_id, rows, pct_time, pct_coord),
        caption = "Served, but with incomplete time or position — ERDDAP will drop these rows from spatial/temporal queries")
}
```

## Build the DuckDB view database

```{r}
#| label: build-db
# DuckDB validates a `read_parquet()` path when the VIEW IS CREATED — it needs the
# schema — so a view bound to the server's path CANNOT be created on this machine.
# That is why `libs/erddap_duckdb.R` documents an identity mount: the .db has to be
# built where the data already sits at the path the views name.
#
# So this produces two artifacts instead of one impossible file:
#   calcofi_local.db     — built and validated here, against the local release
#   build_erddap_db.R    — the same view SQL, to run ON THE SERVER after the sync
build_db <- function(db_path, root, ing_root) {
  if (file_exists(db_path)) file_delete(db_path)
  cx <- dbConnect(duckdb(), dbdir = db_path)
  on.exit(dbDisconnect(cx, shutdown = TRUE))
  for (i in seq_len(nrow(cfg))) {
    r <- cfg[i, ]
    dbExecute(cx, glue("CREATE OR REPLACE VIEW \"{r$dataset_id}\" AS
                        {build_sql(r, root, ing_root)}"))
  }
  writeLines(as.character(dbGetQuery(cx, "SELECT version()")[[1]]),
             file.path(dirname(db_path), "BUILD_VERSION.txt"))
  dbGetQuery(cx, "SELECT view_name FROM duckdb_views() WHERE NOT internal")$view_name
}
v_local <- build_db(file.path(dir_erddap, "calcofi_local.db"), PQ_LOCAL, ING_LOCAL)
cat(glue("validated views: {length(v_local)}\n"))
cat(glue("duckdb engine  : {readLines(file.path(dir_erddap, 'BUILD_VERSION.txt'))[1]}\n"))
stopifnot(setequal(v_local, cfg$dataset_id))
```

```{r}
#| label: emit-server-script
# The server script carries the SQL verbatim, so the deployed views are the ones
# validated above rather than a hand-reimplementation of them.
view_sql <- vapply(seq_len(nrow(cfg)), function(i)
  build_sql(cfg[i, ], "{{PQ}}", "{{ING}}"), character(1))

script <- c(
  "#!/usr/bin/env Rscript",
  "# build_erddap_db.R — GENERATED by publish_to-erddap.qmd. Do not hand-edit.",
  "#",
  glue("# release: {RELEASE}   datasets: {nrow(cfg)}"),
  "#",
  "# Run this ON THE CalCOFI SERVER, after the release parquet has been synced to",
  "# the path below. DuckDB binds a view's parquet paths literally and validates",
  "# them at CREATE time, so the .db must be built where the data actually is.",
  "#",
  "#   docker exec -i rstudio bash -lc \\",
  "#     'cd /share/erddap/data && Rscript build_erddap_db.R'",
  "",
  "suppressMessages({library(DBI); library(duckdb)})",
  glue("PQ  <- Sys.getenv('CALCOFI_ERDDAP_PQ',  '{PQ_SERVER}')"),
  glue("ING <- Sys.getenv('CALCOFI_ERDDAP_ING', '{ING_SERVER}')"),
  glue("DB  <- Sys.getenv('CALCOFI_ERDDAP_DB',  '{ERDDAP_ROOT}/calcofi.db')"),
  "stopifnot('release parquet not found' = dir.exists(PQ))",
  "if (file.exists(DB)) file.remove(DB)",
  "con <- dbConnect(duckdb(), dbdir = DB)",
  "views <- list(", 
  paste0("  `", cfg$dataset_id, "` = ",
         vapply(view_sql, function(x) paste0("r\"(", x, ")\""), character(1)),
         c(rep(",", nrow(cfg) - 1), "")),
  ")",
  "for (nm in names(views)) {",
  "  sql <- gsub('{{PQ}}', PQ, views[[nm]], fixed = TRUE)",
  "  sql <- gsub('{{ING}}', ING, sql, fixed = TRUE)",
  "  dbExecute(con, sprintf('CREATE OR REPLACE VIEW \"%s\" AS %s', nm, sql))",
  "  n <- dbGetQuery(con, sprintf('SELECT count(*) n FROM \"%s\"', nm))$n",
  "  cat(sprintf('%-42s %12s rows\\n', nm, format(n, big.mark = ',')))",
  "}",
  "dbDisconnect(con, shutdown = TRUE)",
  "cat('wrote ', DB, '\\n', sep = '')")

script_path <- file.path(dir_erddap, "build_erddap_db.R")
writeLines(script, script_path)
Sys.chmod(script_path, "0755")
# a generated script that does not parse is worse than none
invisible(parse(script_path))
cat(glue("wrote {basename(script_path)} ({length(script)} lines, parses clean)\n"))
```

## Generate `datasets.xml`

```{r}
#| label: xml
#| results: asis
# Title and summary come from each ingest's `dataset_meta` YAML, which is
# authoritative for dataset metadata (it is what the release `dataset` table is
# built from), so ERDDAP, the database and the schema site cannot disagree.
iy_meta <- function(ds) iy[[ds]]$dataset_meta %||% list()

title_of <- function(r) {
  dm <- iy_meta(r$dataset_key)
  base <- dm$dataset_name %||% r$dataset_key
  switch(r$grain,
    obs           = glue("{base} — observations"),
    sample        = glue("{base} — sampling events"),
    obs_attribute = glue("{base} — length/stage frequency"),
    full          = glue("{base} — full resolution (pre-thinning)"))
}
summary_of <- function(r) {
  dm <- iy_meta(r$dataset_key)
  desc <- gsub("[[:space:]]+", " ", trimws(dm$description %||% ""))
  grain_txt <- switch(r$grain,
    obs = paste("One row per observation: a measurement of one quantity at one",
                "event, taxon and depth. `measurement_type` names the quantity and",
                "`measurement_value` holds it, so filter on measurement_type."),
    sample = paste("One row per sampling event, with event-level effort widened",
                   "into its own columns. Use this for effort and station",
                   "counting — summing effort over the observation table",
                   "double-counts it."),
    obs_attribute = paste("Sub-occurrence detail: length- and stage-frequency bins",
                          "(`bin_value`, `bin_label`, `count`) under each",
                          "occurrence. Coordinates are inherited from the event."),
    full = paste("The full series before adaptive thinning. Much larger than the",
                 "headline table and provided for scan-level work; most analyses",
                 "want the thinned table."))
  paste(c(desc, grain_txt, glue("Source: CalCOFI integrated database release {RELEASE}."),
          if (nzchar(dm$citation_main %||% "")) glue("Citation: {dm$citation_main}.")),
        collapse = " ")
}
# every view is a long table of independent rows carrying their own time and
# position, which is ERDDAP's "Point"; the vertical/trajectory structure is
# expressed in the netCDF products, not here
CDM <- "Point"

blocks <- character(0)
for (i in seq_len(nrow(cfg))) {
  r <- cfg[i, ]
  cols <- q("DESCRIBE SELECT * FROM ({build_sql(r, PQ_LOCAL)})") |>
    transmute(column = column_name, duckdb_type = column_type)
  xml <- erddap_duckdb_dataset_xml(
    staged        = as.data.frame(cols),
    dataset_id    = r$dataset_id,
    title         = title_of(r),
    summary       = summary_of(r),
    source_url    = glue("jdbc:duckdb:{ERDDAP_ROOT}/calcofi.db"),
    table_name    = r$dataset_id,
    # keyed by column name: this lands on the `{ds}_sample` views, whose widened
    # effort columns ARE measurement_type names, and is inert elsewhere
    units_lookup  = mt_units,
    cdm_data_type = CDM,
    global_atts   = list(
      license   = iy_meta(r$dataset_key)$license %||% "CC-BY 4.0",
      infoUrl   = iy_meta(r$dataset_key)$link_calcofi_org %||% "https://calcofi.org",
      db_release = RELEASE, dataset_key = r$dataset_key,
      references = "https://calcofi.io/workflows/publish_to-erddap.html"))
  dir_create(file.path(dir_erddap, r$dataset_id))
  writeLines(xml, file.path(dir_erddap, r$dataset_id, glue("{r$dataset_id}.xml")))
  blocks <- c(blocks, xml)
  cat(glue("\n- `{r$dataset_id}`: {ncol(cols)} cols, ",
           "{val$rows[val$dataset_id == r$dataset_id]} rows\n"))
}
```

```{r}
#| label: assemble
xml_all <- paste0(
  "<!-- CalCOFI ERDDAP datasets — generated by publish_to-erddap.qmd\n",
  glue("     release {RELEASE}, {length(blocks)} datasets, ",
       "{format(Sys.time(), '%Y-%m-%d')}\n"),
  "     EDDTableFromDatabase over DuckDB views; see the notebook for deploy. -->\n",
  paste(blocks, collapse = "\n\n"))
writeLines(xml_all, file.path(dir_erddap, "datasets_calcofi.xml"))
write_csv(cfg |> left_join(val, by = "dataset_id"),
          file.path(dir_erddap, "erddap_plan.csv"), na = "")
cat(glue("wrote {length(blocks)} dataset blocks to data/erddap/datasets_calcofi.xml\n"))
```

## Deploy (manual)

```{r}
#| label: deploy
#| echo: false
#| results: asis
cat(glue('
The notebook writes config and a view database; it does **not** touch the server.

1. **Sync the release parquet** to the identity-mounted path the views name. The
   views bind `{PQ_SERVER}` literally, so this path must match exactly:

   ```bash
   rsync -av --delete data/releases/{RELEASE}/parquet/ \\
     calcofi:{PQ_SERVER}/
   ```

2. **Build the view database on the server.** It cannot be built here: DuckDB
   validates a view\'s parquet paths at `CREATE` time, so the views must be created
   where the data already sits at the path they name. The generated script carries
   the exact SQL validated above.

   ```bash
   rsync -av data/erddap/build_erddap_db.R calcofi:{ERDDAP_ROOT}/
   ssh calcofi "docker exec -i rstudio bash -lc \\
     \'cd {ERDDAP_ROOT} && Rscript build_erddap_db.R\'"
   ```

   It prints a row count per view — a zero there means the sync in step 1 is
   incomplete, and is the cheapest place to catch it.

3. **Splice the blocks** from `data/erddap/datasets_calcofi.xml` into
   `CalCOFI/erddap/content/datasets.xml` at the
   `<!-- add dataset definitions below -->` marker, then commit and push the
   `CalCOFI/erddap` repo (it is bind-mounted into the container).

4. **Reload** per dataset:

   ```bash
   touch {ERDDAP_ROOT}/flag/<datasetID>
   ```

5. **Verify** — a `.das` request is the cheapest end-to-end check:

   ```bash
   curl -s https://erddap.calcofi.io/erddap/tabledap/calcofi_ctd-cast.das | head
   ```

   and check `{ERDDAP_ROOT}/log.txt` for `WARNING: Bad line(s)`.

**Requirements.** The DuckDB JDBC driver must be in ERDDAP\'s `WEB-INF/lib`
(custom image: `CalCOFI/server/erddap/Dockerfile`), and its engine version must be
>= the engine that wrote the views — this build used
**{readLines(file.path(dir_erddap, "BUILD_VERSION.txt"))[1]}**.

**Retiring the old datasetIDs.** The live server currently serves `calcofi_casts`,
`calcofi_ctd_measurement`, `calcofi_ctd_thin`, `calcofi_dic`,
`calcofi_euphausiids`, `calcofi_phytoplankton` and `calcofi_zooplankton`. None of
those IDs appear above. Splicing this config *adds* the new ones; removing the old
ones is a separate edit to `datasets.xml`, deliberately left to whoever owns that
contract.
'))
```

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
