---
title: "Publish Larvae to OBIS"
editor_options: 
  chunk_output_type: console
---

## Goals

- Create a DarwinCore Archive (DwC-A) for CalCOFI ichthyoplankton (fish eggs and larvae) data to publish to OBIS.
- Use the existing CalCOFI database schema and connection.
- Ensure compliance with DarwinCore standards, including event hierarchy and extensions.

### References

DarwinCore:

- [The OBIS Manual](https://manual.obis.org/)
- [Darwin Core Quick Reference Guide](https://dwc.tdwg.org/terms/)
- [Guide for publishing biological survey and monitoring data to GBIF](https://docs.gbif.org/guide-publishing-survey-data/en/)
- [Survey and Monitoring Data Quick-Start Guide: A how-to for updating a Darwin Core dataset using the Humboldt Extension](https://docs.gbif.org/survey-monitoring-quick-start/en/)

CalCOFI:

- [CalCOFI.org: Fish Eggs & Larvae ](https://calcofi.org/data/marine-ecosystem-data/fish-eggs-larvae/)
  - [ERDDAP: CoastWatch - Search 'CalCOFI'](https://coastwatch.pfeg.noaa.gov/erddap/search/index.html?page=1&itemsPerPage=1000&searchFor=CalCOFI)
    - [CalCOFI info](https://oceanview.pfeg.noaa.gov/CalCOFI/calcofi_info.html)
  - [EDI: CALCOFI fish larvae at 66 standard stations, 1966 - ongoing](https://portal.edirepository.org/nis/mapbrowse?scope=edi&identifier=109&revision=4)

```{r setup}
# TODO:
# commit
# - [x] use lazy loading of tbl, so skip `|> collect()`
# - [x] %>% to |>
# - [x] invert events hierarchy: net -> tow (-> tow_type) -> site -> cruise -> ship
# - [x] use UUID, ie drop prefix `urn:uuid:` for eventID, occurrenceID, measurementID
# - [ ] Consider merging larva, larvastage, and larvasize into one occurrence table, since all sharing net_uuid, species_id


librarian::shelf(
  CalCOFI/calcofi4db, 
  DBI, dm, dplyr, DT, glue, here, lubridate, purrr, tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F)

# get database connection
schema <- "dev"
con    <- get_db_con(schema)
```

## Show Database Tables

```{r fig-dm_tbls}
#| fig-cap: "Entity relationship diagram (ERD) of the CalCOFI database. Created using the data models `dm` R package function [`dm_draw()`](https://krlmlr.github.io/dm/reference/dm_draw.html)."

# exclude experimental tables
tbls <- dbListTables(con) |> 
  sort() |> 
  setdiff(
    c("schema_version",
      "grid", "site_seg",
      "bottle", "cast"))

# learn relations from database and draw
dm <- dm_from_con(con, schema = schema, table_names = tbls, learn_keys  = T)
dm_draw(dm, view_type = "all")
```

```{mermaid}
%%| label: fig-calcofi_erd
%%| fig-cap: "Entity relationship diagram (ERD) of the CalCOFI database. Created from image above and prompt to [Claude](https://claude.ai/): 'Generate an ERD mermaid diagram from the image.'"
%%| file: diagrams/calcofi_erd.mmd
```

## Prompt Claude to Generate DarwinCore

PROMPT: Using the ERD above, create a Quarto document with chunks of R code and tidyverse style use of the database connection object `con` to generate a DarwinCore package using the `species.worms_id` (aka WoRMS AphiaID) to populate the species identifiers. Combine biological observation where possible, otherwise compose as seperate datasets, given the different biological observations: egg, eggstage, larva, larvasize, larvastage. Generate DarwinCore with as much of the extended information (i.e.,  `event`, `occurrence`, and `measurementOrFact` ) as possible.
Compose Ecological Metadata Language (EML) metadata for the DarwinCore package, including the `dataset` and `creator` elements. Include prompts where more
information is needed.

```{r stop_eval}
knitr::opts_chunk$set(eval = F)
```

```{r setup}
#| label: load-packages
librarian::shelf(
  DBI, EML, here, lubridate, tidyverse, uuid)

# Assuming database connection is already established as 'con'
# con <- dbConnect(...) 
```

## Data Extraction from Database

First, let's extract all necessary tables from the database:

```{r extract-data}
#| label: extract-tables

# Get simple lazy table references
tbl_ship       <- tbl(con, "ship")
tbl_cruise     <- tbl(con, "cruise")
tbl_site       <- tbl(con, "site")
tbl_tow        <- tbl(con, "tow")
tbl_tow_type   <- tbl(con, "tow_type")
tbl_net        <- tbl(con, "net")
tbl_species    <- tbl(con, "species")
# biological observation tables
tbl_egg         <- tbl(con, "egg")
tbl_egg_stage   <- tbl(con, "egg_stage")
tbl_larva       <- tbl(con, "larva")
tbl_larva_stage <- tbl(con, "larva_stage")
tbl_larva_size  <- tbl(con, "larva_size")
```

## Create Event Hierarchy

DarwinCore uses an event-based model. We'll create a hierarchical event structure:

- cruise
  - site
    - tow
      - net

```{r fig-larvae_hierarchy}
# TODO: create mermaid diagram of fig-larvae_hierarchy
```

Per [3.3.2. Simple nested datasets with Project-level information – Guide for publishing biological survey and monitoring data to GBIF](https://docs.gbif.org/guide-publishing-survey-data/en/#simple-nested-datasets-with-project-level-information)

```{r create-events}
#| label: build-event-hierarchy

# Build complete event hierarchy
q_events <- tbl_net |> 
  left_join(tbl_tow,      by = "tow_uuid") |>
  left_join(tbl_tow_type, by = "tow_type_key") |>
  left_join(tbl_site,     by = "site_uuid") |>
  left_join(tbl_cruise,   by = "cruise_uuid") |>
  left_join(tbl_ship,     by = "ship_key")

# Create event core at different levels ----

flds_to_iso8601 <-  function(flds, output = "datetime") {
  # helper function to convert date/time fields to ISO 8601 format
  # per https://dwc.tdwg.org/terms/ eventDate: add 2007-03-01T13:00:00Z/2008-05-11T15:30:00Z (some time within the interval beginning 1 March 2007 at 1pm UTC and before 11 May 2008 at 3:30pm UTC)
  
  stopifnot(length(flds) %in% c(1,2))
  stopifnot(output %in% c("datetime", "time"))
  
  if (length(flds) == 1 & output == "datetime")
      s <- glue::glue(
        "to_char({flds}, 'YYYY-MM-DD\"T\"HH24:MI:SS\"Z\"')")
  if (length(flds) == 2 & output == "datetime")
      s <- glue::glue(
        "to_char({flds[1]}, 'YYYY-MM-DD\"T\"HH24:MI:SS\"Z\"') || '/' || 
         to_char({flds[2]}, 'YYYY-MM-DD\"T\"HH24:MI:SS\"Z\"')")
  
  if (length(flds) == 1 & output == "time")
    s <- glue::glue(
      "to_char({flds}, 'HH24:MI:SS\"Z\"')")
  if (length(flds) == 2 & output == "time")
    s <- glue::glue(
      "to_char({flds[1]}, 'HH24:MI:SS\"Z\"') || '/' || 
       to_char({flds[2]}, 'HH24:MI:SS\"Z\"')")
  
  as.character(s)
}

# * Cruise level events ----

# get min/max datetimes for cruises
q_cruise_event_date <- q_events |>
  group_by(cruise_uuid) |>
  summarize(
    dtime_min = min(time_start, na.rm = T),
    dtime_max = max(time_start, na.rm = T),
    .groups = "drop") |>
  mutate(
    eventDate = sql(flds_to_iso8601(c("dtime_min", "dtime_max"))) ) |> 
  select(cruise_uuid, eventDate)
    
q_cruise_events <- q_events |>
  distinct(cruise_uuid, ship_name, ship_nodc, date_ym) |>
  mutate(
    eventID          = cruise_uuid,
    # parentEventID  = NA, # no parent for cruise level; NA generated with union_all
    eventRemarks     = paste("Cruise on", ship_name),
    samplingProtocol = "Marine plankton survey cruise",
    sampleSizeValue  = NA_real_,
    sampleSizeUnit   = NA_character_,
    habitat          = "marine",
    eventType        = "cruise") |>
  left_join(
    q_cruise_event_date,
    by = "cruise_uuid") |> 
  select(
    eventID, # parentEventID, 
    eventType, 
    eventDate, samplingProtocol, eventRemarks, habitat)

# Site level events
q_site_events <- q_events |>
  distinct(site_uuid, cruise_uuid, longitude, latitude, line, station, orderocc) |>
  mutate(
    eventID          = site_uuid,
    parentEventID    = cruise_uuid,
    decimalLatitude  = latitude,
    decimalLongitude = longitude,
    eventRemarks     = paste("CalCOFI station", station, "on line", line),
    # per https://proj.org/en/stable/operations/projections/calcofi.html
    locationID       = paste0(line, "_", station),
    eventType        = "site",
    samplingProtocol = "Marine plankton survey station") |> 
  select(
    eventID, parentEventID, eventType, 
    samplingProtocol, eventRemarks, decimalLatitude, decimalLongitude, locationID)

# Tow level events
q_tow_events <- q_events |>
  distinct(
    tow_uuid, site_uuid, 
    tow_number, tow_type_key, tow_type_description = description, 
    time_start) |>
  mutate(
    eventID          = tow_uuid,
    parentEventID    = site_uuid,
    eventDate        = sql(flds_to_iso8601("time_start")),
    eventTime        = sql(flds_to_iso8601("time_start", output = "time")),
    eventRemarks     = paste("Tow", tow_number, "-", tow_type_description),
    eventType        = "tow",
    samplingProtocol = paste(tow_type_key, ":", tow_type_description)) |> 
  select(
    eventID, parentEventID, eventType,
    eventDate, eventTime, 
    samplingProtocol, eventRemarks)

# Net level events (sampling events)
q_net_events_allflds <- q_events |>
  distinct(
    net_uuid, tow_uuid, side, std_haul_factor, vol_sampled_m3, 
    prop_sorted, smallplankton, totalplankton) |>
  mutate(
    eventID          = net_uuid,
    parentEventID    = tow_uuid,
    # TODO: glue to paste
    eventRemarks     = paste(
      "Net sample from side", side, "; only", prop_sorted, "proportion sorted of",
      vol_sampled_m3, "cubic meters sampled"),
    eventType        = "net_sample",
    samplingProtocol = paste("Plankton net tow -", side, "side"),
    sampleSizeValue  = vol_sampled_m3,
    sampleSizeUnit   = "cubic meters")
q_net_events <- q_net_events_allflds |> 
  select(
    eventID, parentEventID, eventType, 
    samplingProtocol, eventRemarks, sampleSizeValue, sampleSizeUnit)

# Combine all event levels
db_version <- tbl(con, "schema_version") |> pull(version)
dataset_id <- paste0(
  # "https://github.com/CalCOFI/calcofi4db/releases/tag/v", 
  "calcofi_db.", db_version,
  "_larvae.", format(Sys.Date(), "%Y-%m-%d"))
# "calcofi_db.1.1.0_larvae.2025-10-09"

q_event_core <- union_all(
  q_cruise_events,
  q_site_events,
  q_tow_events, 
  q_net_events) |> 
  # Add required DwC fields
  mutate(
    basisOfRecord                 = "HumanObservation",
    geodeticDatum                 = "WGS84",
    coordinateUncertaintyInMeters = 1000, # TODO: adjust based on accuracy of GPS/Loran/dead-reckoning over time
    countryCode                   = "US",
    waterBody                     = "California Current",  # LME: http://marineregions.org/mrgid/8549
    datasetID                     = dataset_id)
# TODO: datasetID: permalink to this qmd in calcofi4db release?
```

## Create Occurrence Records

Now let's combine all biological observations into occurrence records:

```{r create-occurrences}
#| label: build-occurrences

# Function to standardize biological observations
create_occurrences <- function(obs_data, obs_type, has_stage = FALSE, has_size = FALSE) {
  
  base_occ <- obs_data |>
    left_join(
      tbl_species,
      by = "species_id") |>
    mutate(
      occurrenceID         = sql("gen_random_uuid()"), # UUIDgenerate(n = n()),    # TODO: set UUID in create_db.qmd
      eventID              = net_uuid,
      scientificName       = scientific_name,
      scientificNameID     = paste0("urn:lsid:marinespecies.org:taxname:", worms_id),
      # taxonID          = paste0("urn:lsid:marinespecies.org:taxname:", worms_id),
      kingdom              = "Animalia",
      occurrenceStatus     = ifelse(tally > 0, "present", "absent"),
      # https://manual.obis.org/FAQ.html#:~:text=What%20is%20the%20difference%20between%20individualCount%2C%20and%20organismQuantity
      organismQuantity     = tally,
      organismQuantityType = "individuals",
      lifeStage            = case_when(
        obs_type == "egg"   ~ "egg",
        obs_type == "larva" ~ "larva",
        TRUE ~ NA_character_),
      # associatedTaxa       = common_name
      preparations         = paste0(obs_type, " sample"))
  
  # Add stage information if present
  if (has_stage)
    base_occ <- base_occ |>
      mutate(
        lifeStage = ifelse(
          !is.na(stage), 
          paste(obs_type, stage, sep = "_"), 
          lifeStage),
        occurrenceRemarks = paste("Stage:", stage))
  
  # Add size class if larvasize
  if (has_size)
    base_occ <- base_occ |>
      mutate(
        occurrenceRemarks = paste("Size class:", length_mm, "mm"))
  
  return(base_occ)
}

# Process each observation type
q_egg_occurrences        <- create_occurrences(tbl_egg,         "egg")
q_eggstage_occurrences   <- create_occurrences(tbl_egg_stage,   "egg",   has_stage = T)
q_larva_occurrences      <- create_occurrences(tbl_larva,       "larva")
q_larvastage_occurrences <- create_occurrences(tbl_larva_stage, "larva", has_stage = T)
q_larvasize_occurrences  <- create_occurrences(tbl_larva_size,  "larva", has_size  = T)

# Combine all occurrences
q_occurrence_extension <- union_all(
  q_egg_occurrences,
  q_eggstage_occurrences,
  q_larva_occurrences,
  q_larvastage_occurrences,
  q_larvasize_occurrences) |>
  select(
    occurrenceID, eventID, 
    scientificName, scientificNameID, # taxonID, 
    kingdom, occurrenceStatus, organismQuantity, organismQuantityType,
    lifeStage, preparations, occurrenceRemarks) |>  #, associatedTaxa)
  # Add additional required/recommended fields
  mutate(
    # TODO: identifiedBy, dateIdentified,identificationReferences
    # identifiedBy = "[NEED IDENTIFIER NAME]", # Prompt: Add taxonomist name
    # dateIdentified = "[NEED IDENTIFICATION DATE]", # Prompt: Add identification date
    # identificationReferences = "[NEED REFERENCES]", # Prompt: Add identification guides used
    modified = sql("CURRENT_DATE"))
```

## Create MeasurementOrFact Extension

Extract all measurements and environmental data:

```{r create-measurements}
#| label: build-measurements

# Environmental measurements from net samples <-> eventID
d_env_measurements <- q_net_events_allflds |>
  select(
    eventID, 
    std_haul_factor, vol_sampled_m3, prop_sorted, smallplankton, totalplankton) |>
  collect() |> 
  pivot_longer(
    cols = c(
      std_haul_factor, vol_sampled_m3, prop_sorted, smallplankton, totalplankton),
    names_to  = "measurementType",
    values_to = "measurementValue") |>
  filter(!is.na(measurementValue)) |>
  mutate(
    measurementID      = UUIDgenerate(n = n()), # TODO: how does UUID relate to occurrenceID?
    measurementUnit    = case_when(
      measurementType == "vol_sampled_m3" ~ "cubic meters",
      measurementType == "std_haul_factor" ~ "dimensionless",
      measurementType == "prop_sorted" ~ "proportion",
      measurementType %in% c("smallplankton", "totalplankton") ~ "grams",
      TRUE ~ NA_character_),
    measurementType    = case_when(
      measurementType == "vol_sampled_m3" ~ "volume filtered",
      measurementType == "std_haul_factor" ~ "standardized haul factor",
      measurementType == "prop_sorted" ~ "proportion of sample sorted",
      measurementType == "smallplankton" ~ "small plankton biomass",
      measurementType == "totalplankton" ~ "total plankton biomass",
      TRUE ~ measurementType),
    measurementMethod  = "https://oceanview.pfeg.noaa.gov/CalCOFI/calcofi_info.html", # TODO: consider revising with full description based on year collected
    measurementRemarks = NA_character_)

# Size measurements from larvasize <-> occurrenceID
d_size_measurements <- q_larvasize_occurrences |>
  filter(!is.na(length_mm)) |>
  mutate(
    measurementID      = sql("gen_random_uuid()"), # UUIDgenerate(n = n()),    # TODO: set UUID in create_db.qmd? relate to occurrenceID?
    measurementType    = "body length",
    measurementValue   = length_mm,
    measurementUnit    = "millimeters",
    measurementMethod  = "https://oceanview.pfeg.noaa.gov/CalCOFI/calcofi_info.html", # TODO: consider revising with full description based on year collected
    measurementRemarks = "Larval body length") |>
  select(
    measurementID, occurrenceID, measurementType, measurementValue,
    measurementUnit, measurementMethod, measurementRemarks) |> 
  collect()

# setdiff(names(d_env_measurements), names(d_size_measurements)) # eventID
# setdiff(names(d_size_measurements), names(d_env_measurements)) # occurrenceID

# Combine all measurements
d_measurementorfact_extension <- bind_rows(
  d_env_measurements, 
  d_size_measurements)
```

## Write DarwinCore Archive Files

```{r write-dwc-files}
#| label: export-dwc

# Create output directory
dir_out <- here("data/darwincore/larvae")
dir.create(dwc_dir, showWarnings = F)

# Write core and extension files
write_csv(q_event_core |> collect(), file.path(dir_out, "event.csv"))
write_csv(q_occurrence_extension |> collect(), file.path(dir_out, "occurrence.csv"))
write_csv(d_measurementorfact_extension, file.path(dir_out, "measurementorfact.csv"))

# Create meta.xml for DwC-A
meta_xml <- '<?xml version="1.0" encoding="UTF-8"?>
<archive xmlns="http://rs.tdwg.org/dwc/text/" 
         xmlns:xsi="http://www.w3.org/2001/XMLSchema-instance" 
         xsi:schemaLocation="http://rs.tdwg.org/dwc/text/ http://rs.tdwg.org/dwc/text/tdwg_dwc_text.xsd">
  <core encoding="UTF-8" fieldsTerminatedBy="," linesTerminatedBy="\\n" 
        fieldsEnclosedBy=\'"\' ignoreHeaderLines="1" rowType="http://rs.tdwg.org/dwc/terms/Event">
    <files>
      <location>event.csv</location>
    </files>
    <id index="0" />
  </core>
  <extension encoding="UTF-8" fieldsTerminatedBy="," linesTerminatedBy="\\n" 
             fieldsEnclosedBy=\'"\' ignoreHeaderLines="1" rowType="http://rs.tdwg.org/dwc/terms/Occurrence">
    <files>
      <location>occurrence.csv</location>
    </files>
    <coreid index="1" />
  </extension>
  <extension encoding="UTF-8" fieldsTerminatedBy="," linesTerminatedBy="\\n" 
             fieldsEnclosedBy=\'"\' ignoreHeaderLines="1" rowType="http://rs.tdwg.org/dwc/terms/MeasurementOrFact">
    <files>
      <location>measurementorfact.csv</location>
    </files>
    <coreid index="1" />
  </extension>
</archive>'

writeLines(meta_xml, file.path(dir_out, "meta.xml"))
```

## Create EML Metadata

```{r create-eml}
#| label: build-eml

# Dataset metadata - PROMPTS FOR REQUIRED INFORMATION
dataset_title    <- "CalCOFI Fish Larvae Tows" # Prompt: Enter dataset title
dataset_abstract <- "Fish larvae counts and standardized counts for eggs captured in CalCOFI icthyoplankton nets (primarily vertical [Calvet or Pairovet], oblique [bongo or ring nets], and surface tows [Manta nets]) . Surface tows are normally standardized to count per 1,000 m3 strained. Oblique tows are normally standardized to count per 10 m2 of surface sampled."
dataset_keywords <- c(
  "atmosphere", "biology", "biosphere", "calcofi", "earth science", "environment", "latitude", "longitude", "ocean", "time") # GCMD Science Keywords (source: https://catalog.data.gov/dataset/calcofi-larvae-counts-positive-tows)
# dataset_uuid <- UUIDgenerate() # Or use existing dataset UUID

# Geographic coverage
geo_coverage <- tbl_site |>
  summarise(
    westBoundingCoordinate  = min(longitude, na.rm = T),
    eastBoundingCoordinate  = max(longitude, na.rm = T),
    northBoundingCoordinate = max(latitude,  na.rm = T),
    southBoundingCoordinate = min(latitude,  na.rm = T)) |> 
  collect()

# Temporal coverage
temp_coverage <- tbl_tow |>
  summarise(
    beginDate = min(time_start, na.rm = TRUE),
    endDate   = max(time_start, na.rm = TRUE)) |> 
  collect()

# Taxonomic coverage
taxa_coverage <- tbl_species |>
  filter(!is.na(worms_id)) |>
  select(scientificName = scientific_name) |>
  collect() |> 
  as.list()

# Create EML document
my_eml <- list(
  packageId = dataset_id,
  system    = "calcofi_db.{version}_{dataset}.{render-date}",
  dataset = list(
    title   = dataset_title,
    creator = list(
      individualName = list(
        givenName = "Ed",
        surName   = "Weber"),
    organizationName      = "NOAA SWFSC",
    electronicMailAddress = "ed.weber@noaa.gov",
      userId = list(
        directory = "https://orcid.org/",
        userId = "0000-0002-0942-434X") ),
    abstract = dataset_abstract,
    keywordSet = list(
      keyword          = dataset_keywords,
      keywordThesaurus = "GCMD Science Keywords"),
    coverage = list(
      geographicCoverage = list(
        geographicDescription = "California Current Large Marine Ecoregion",
        boundingCoordinates   = geo_coverage),
      temporalCoverage = list(
        rangeOfDates = list(
          beginDate = list(calendarDate = as.character(temp_coverage$beginDate)),
          endDate   = list(calendarDate = as.character(temp_coverage$endDate)) )),
      taxonomicCoverage = list(
        generalTaxonomicCoverage = "Marine ichthyoplankton and fish eggs",
        taxonomicClassification = taxa_coverage )),
    contact = list(
      individualName = list(
        givenName = "Ed",
        surName   = "Weber"),
      electronicMailAddress = "ed.weber@noaa.gov"),
    methods = list(
      methodStep = list(
        description = list(
          para = "[SAMPLING METHODS DESCRIPTION]" )), # TODO: sampling methods
      sampling = list(
        studyExtent = list(
          description = list(
            para = "[STUDY EXTENT DESCRIPTION]")),    # TODO: study extent
        samplingDescription = list(
          para = "[SAMPLING DESCRIPTION]" ))),        # TODO: sampling
    project = list(
      title = dataset_title,                          # TODO: project title?
      personnel = list(
        individualName = list(
          givenName = "Ed",                           # TODO: PI?
          surName   = "Weber" ),
        role = "Principal Investigator"),
      funding = list(
        para = "[FUNDING INFORMATION]" ))))          # TODO: Funding

# Write EML
write_eml(my_eml, file.path(dir_out, "eml.xml"))
```

## Data Quality Checks

```{r quality-checks}
#| label: check-data-quality

# Check for missing WoRMS IDs
d_missing_worms <- tbl_species |>
  filter(is.na(worms_id)) |>
  select(species_id, scientific_name) |> 
  collect()

if(nrow(d_missing_worms) > 0) {
  cat("WARNING: The following species are missing WoRMS IDs:\n")
  print(d_missing_worms)
  cat("\nPlease add WoRMS AphiaIDs for these species before finalizing the dataset.\n")
}

# Check for orphan records
d_orphan_occurrences <- q_occurrence_extension |>
  anti_join(q_event_core, by = "eventID") |> 
  collect()

if(nrow(d_orphan_occurrences) > 0) {
  cat("\nWARNING: Found", nrow(d_orphan_occurrences), "occurrence records without matching events.\n")
}

# Summary statistics
n_events   <- q_event_core |> summarize(n = n()) |> pull(n)
n_events   <- q_event_core |> filter(eventType == "net_sample") |> summarize(n = n()) |> pull(n)
n_cruises  <- q_event_core |> filter(eventType == "cruise")     |> summarize(n = n()) |> pull(n)
n_sites    <- q_event_core |> filter(eventType == "site")       |> summarize(n = n()) |> pull(n)
n_tows     <- q_event_core |> filter(eventType == "tow")        |> summarize(n = n()) |> pull(n)
n_nets     <- q_event_core |> filter(eventType == "net_sample") |> summarize(n = n()) |> pull(n)
n_occs     <- q_occurrence_extension |> summarize(n = n()) |> pull(n)
n_species  <- q_occurrence_extension |> distinct(scientificName) |> summarize(n = n()) |> pull(n)
n_measures <- d_measurementorfact_extension |> nrow()

glue("
  === Dataset Summary ===
  Total events:        {n_events}
  - Cruises:           {n_cruises}
  - Sites:             {n_sites}
  - Tows:              {n_tows}
  - Net samples:       {n_nets}
  Total occurrences:   {n_occs}
  Total species:       {n_species}
  Total measurements:  {n_measures}")
```

## Package Creation

```{r create-package}
#| label: create-dwc-archive

# Create DwC-A zip file
zip_file <- file.path(dir_out, paste0("../larvae_", Sys.Date(), ".zip"))
zip(zip_file, 
    files = file.path(
      dir_out, c(
        "event.csv", "occurrence.csv", 
        "measurementorfact.csv", "meta.xml", "eml.xml")))

cat("\nDarwin Core Archive created:", zip_file, "\n")
```

## Questions

- `datasetID`: 
  https://manual.obis.org/examples.html -> datasetID: https://marineinfo.org/id/dataset/6403

## Required Information Prompts

Please provide the following information to complete the DarwinCore package:

### Dataset Information
- [ ] Dataset title
- [ ] Dataset abstract (200-500 words describing the dataset)
- [ ] Keywords (3-5 relevant keywords)
- [ ] Country code (ISO 3166-1 alpha-2)
- [ ] Water body name

### Personnel Information
- [ ] Dataset creator name and ORCID
- [ ] Creator organization
- [ ] Creator email
- [ ] Dataset contact name and email
- [ ] Principal investigator name

### Project Information
- [ ] Project title
- [ ] Funding information (agency and grant numbers)

### Methodology
- [ ] Detailed sampling methods description
- [ ] Study extent description
- [ ] Sampling design and protocols
- [ ] Measurement methods for environmental variables
- [ ] Size measurement methods for larvae
- [ ] Taxonomist name(s) who identified specimens
- [ ] Identification date(s)
- [ ] Identification references used

### Geographic Information
- [ ] Geographic description of the study area
- [ ] GPS accuracy (coordinate uncertainty in meters)

Once this information is provided, update the placeholders marked with [NEED ...] in the code above.
