---
title: "Publish CTD casts to CF NetCDF"
subtitle: "A true CF Discrete-Sampling-Geometry profile dataset — one profile per cast, sensors as variables"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_ctd_cast_to_netcdf
  workflow_type: publish
  dependency:
    - release_database
  provider: calcofi
  dataset: ctd-cast
  output: data/netcdf/ctd-cast.nc
  workflow_url: https://calcofi.io/workflows/publish_ctd-cast_to-netcdf.html
editor_options:
  chunk_output_type: console
---

## Overview

Publish the CalCOFI CTD record as a **CF Discrete-Sampling-Geometry profile
dataset**, downstream of `release_database`.

This is the *simple* end of the publish range, and deliberately so — it is the
counterpart to
[`publish_ichthyo_to-netcdf`](https://calcofi.io/workflows/publish_ichthyo_to-netcdf.html),
which handles the deeply nested case. CTD needs no netCDF-4 groups:

| | ichthyo | **ctd-cast** |
|---|---|---|
| hierarchy | site → tow → net → occurrence → bin | **flat: cast → depth level** |
| CF status | CF where it applies; groups elsewhere | **fully CF (`featureType=Profile`)** |
| structure | netCDF-4 groups + `parent_index` | **contiguous ragged array (`rowSize`)** |

A CTD cast *is* a CF profile: one instance, a vertical series of depth levels,
each sensor its own variable. So this file is genuinely CF-compliant end to end —
no caveat needed — and any CF-aware tool will read it without special handling.

::: {.callout-note title="Why the widening still matters"}
The database stores CTD in long form: one row per cast × depth ×
`measurement_type`, 5.55 M rows over 15 sensor types. A single
`measurement_value` column therefore mixes °C, PSU, mL/L and volts. CF requires
one unit per variable, so each `measurement_type` becomes its own variable with
its own `units` and `long_name`. That widening is what turns a queryable table
into a self-documenting file.
:::

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(DBI, duckdb, dplyr, glue, ncdf4, readr, jsonlite, digest,
                 knitr, tibble, here, quiet = TRUE)
here <- here::here
source(here("libs/publish_netcdf.R"))

DATASET     <- "ctd-cast"
DATASET_KEY <- "calcofi_ctd-cast"
RELEASE     <- cc_release_version()      # PROMOTED release, never a serving tree
PQ          <- cc_release_parquet(RELEASE)
OUT_DIR     <- here("data/netcdf"); dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
OUT_NC      <- file.path(OUT_DIR, glue("{DATASET}.nc"))
PUBLISH     <- identical(Sys.getenv("CALCOFI_PUBLISH"), "true")

cat(glue("release : {RELEASE}\nparquet : {PQ}\noutput  : {OUT_NC}\npublish : {PUBLISH}\n"))
```

```{r}
#| label: connect
con <- dbConnect(duckdb())
for (s in c("INSTALL httpfs", "LOAD httpfs", "SET memory_limit='6GB'",
            "SET enable_progress_bar=false")) try(dbExecute(con, s), silent = TRUE)
q <- function(sql, ...) dbGetQuery(con, glue(sql, ..., .envir = parent.frame()))

dbExecute(con, glue("
  CREATE VIEW smp AS SELECT * FROM read_parquet('{PQ}/sample.parquet')
                     WHERE dataset_key = '{DATASET_KEY}';
  CREATE VIEW obs AS SELECT * FROM read_parquet('{PQ}/obs/dataset_key={DATASET_KEY}/data_0.parquet');"))
```

## Shape and integrity

```{r}
#| label: integrity
shape <- q("
  SELECT (SELECT count(*) FROM smp WHERE sample_type='cast')       AS casts_in_sample,
         (SELECT count(DISTINCT sample_key) FROM obs)              AS casts_with_obs,
         (SELECT count(*) FROM obs)                                AS obs_rows,
         (SELECT count(DISTINCT measurement_type) FROM obs)        AS sensor_types")
kable(shape)

# A profile must have a well-defined depth axis: one value per (cast, depth,
# measurement_type). Duplicates there would silently collapse in the pivot below,
# so check rather than assume.
dupes <- q("
  SELECT count(*) AS n FROM (
    SELECT sample_key, depth_min_m, measurement_type
    FROM obs GROUP BY 1,2,3 HAVING count(*) > 1)")$n
cat(glue("duplicate (cast, depth, measurement_type) triples: {dupes}\n"))
```

```{r}
#| label: pivot
# WIDEN: measurement_type -> one variable each, ordered by (cast, depth) so the
# profiles are contiguous — which is what a CF contiguous ragged array requires.
types <- q("SELECT DISTINCT measurement_type FROM obs ORDER BY 1")$measurement_type
sel <- paste(sprintf(
  "MAX(measurement_value) FILTER (WHERE measurement_type='%s')::DOUBLE AS \"%s\"",
  types, types), collapse = ",\n    ")

prof <- q("
  SELECT o.sample_key                       AS profile_id,
         any_value(s.cruise_key)            AS cruise_key,
         any_value(s.grid_key)              AS grid_key,
         epoch(any_value(s.datetime))::DOUBLE AS time,
         any_value(s.latitude)::DOUBLE      AS latitude,
         any_value(s.longitude)::DOUBLE     AS longitude,
         o.depth_min_m::DOUBLE              AS depth,
         {sel}
  FROM obs o LEFT JOIN smp s ON s.sample_key = o.sample_key
  GROUP BY o.sample_key, o.depth_min_m
  ORDER BY o.sample_key, o.depth_min_m")

cat(glue("wide rows (cast x depth): {nrow(prof)}\n"))
cat(glue("sensor variables        : {length(types)}\n"))
```

## Write the CF profile file

```{r}
#| label: write-nc
mt_meta <- cc_measurement_meta(here())

prof_ids <- unique(prof$profile_id)
row_size <- as.integer(table(factor(prof$profile_id, levels = prof_ids)))
first_ix <- match(prof_ids, prof$profile_id)
stopifnot(sum(row_size) == nrow(prof))

STRLEN <- 64L
FILL   <- 9.969209968386869e36
d_prof <- ncdim_def("profile", "", seq_along(prof_ids), create_dimvar = FALSE)
d_obs  <- ncdim_def("obs",     "", seq_len(nrow(prof)), create_dimvar = FALSE)
d_str  <- ncdim_def("name_strlen", "", seq_len(STRLEN), create_dimvar = FALSE)

v_pid  <- ncvar_def("profile_id", "", list(d_str, d_prof), prec = "char")
v_cru  <- ncvar_def("cruise_key", "", list(d_str, d_prof), prec = "char")
v_grd  <- ncvar_def("grid_key",   "", list(d_str, d_prof), prec = "char")
v_time <- ncvar_def("time", "seconds since 1970-01-01T00:00:00Z", d_prof, prec = "double")
v_lat  <- ncvar_def("latitude",  "degrees_north", d_prof, prec = "double")
v_lon  <- ncvar_def("longitude", "degrees_east",  d_prof, prec = "double")
v_rs   <- ncvar_def("rowSize",   "",              d_prof, prec = "integer")
v_dep  <- ncvar_def("depth", "m", d_obs, prec = "double", missval = FILL)
v_sens <- lapply(types, function(v) ncvar_def(
  v, as.character(mt_meta[[v]]$units %||% ""), d_obs, prec = "double", missval = FILL))

unlink(OUT_NC)
nc <- nc_create(OUT_NC, c(list(v_pid, v_cru, v_grd, v_time, v_lat, v_lon, v_rs, v_dep),
                          v_sens), force_v4 = TRUE)
ncvar_put(nc, v_pid,  prof_ids)
ncvar_put(nc, v_cru,  prof$cruise_key[first_ix])
ncvar_put(nc, v_grd,  prof$grid_key[first_ix])
ncvar_put(nc, v_time, prof$time[first_ix])
ncvar_put(nc, v_lat,  prof$latitude[first_ix])
ncvar_put(nc, v_lon,  prof$longitude[first_ix])
ncvar_put(nc, v_rs,   row_size)
ncvar_put(nc, v_dep,  prof$depth)
for (i in seq_along(types)) ncvar_put(nc, v_sens[[i]], prof[[types[i]]])

# CF DSG attributes — these are what make it a profile dataset rather than a
# table that happens to be in netCDF.
ncatt_put(nc, "profile_id", "cf_role", "profile_id")
ncatt_put(nc, "profile_id", "long_name", "CTD cast identifier")
ncatt_put(nc, "time", "standard_name", "time");      ncatt_put(nc, "time", "axis", "T")
ncatt_put(nc, "latitude",  "standard_name", "latitude");  ncatt_put(nc, "latitude", "axis", "Y")
ncatt_put(nc, "longitude", "standard_name", "longitude"); ncatt_put(nc, "longitude", "axis", "X")
ncatt_put(nc, "depth", "standard_name", "depth"); ncatt_put(nc, "depth", "axis", "Z")
ncatt_put(nc, "depth", "positive", "down")
ncatt_put(nc, "rowSize", "long_name", "Number of observations for this profile")
ncatt_put(nc, "rowSize", "sample_dimension", "obs")   # -> contiguous ragged array
for (v in types) {
  ncatt_put(nc, v, "long_name",
            as.character(mt_meta[[v]]$long_name %||% gsub("_", " ", v)))
  ncatt_put(nc, v, "coordinates", "time latitude longitude depth")
  sn <- mt_meta[[v]]$standard_name %||% NA_character_
  if (!is.na(sn) && nzchar(sn)) ncatt_put(nc, v, "standard_name", sn)
}
globals <- list(
  title = "CalCOFI CTD casts — CF profile dataset",
  summary = paste(
    "CalCOFI CTD profiles as CF Discrete Sampling Geometry: one profile per cast,",
    "depth levels stored as a contiguous ragged array, each canonical sensor its",
    "own variable with its own units. Widened from the normalized long form in the",
    glue("CalCOFI integrated database release {RELEASE}.")),
  featureType = "profile", cdm_data_type = "Profile",
  cdm_profile_variables = "profile_id, time, latitude, longitude, cruise_key, grid_key",
  Conventions = "CF-1.10, COARDS, ACDD-1.3",
  cf_scope = paste("Fully CF: a CTD cast is a CF profile, so this dataset needs no",
                   "extension beyond the standard (contrast the nested net-tow datasets)."),
  institution = "CalCOFI", source = glue("CalCOFI integrated database release {RELEASE}"),
  db_release = RELEASE, dataset_key = DATASET_KEY,
  references = "https://calcofi.io/workflows/publish_ctd-cast_to-netcdf.html",
  license = "CC-BY 4.0", creator_name = "CalCOFI", creator_url = "https://calcofi.io",
  date_created = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"))
for (nm in names(globals)) ncatt_put(nc, 0, nm, globals[[nm]])
nc_close(nc)

cat(glue("wrote {OUT_NC} ({round(file.size(OUT_NC)/1048576, 1)} MB)\n"))
```

## Verify

```{r}
#| label: verify
ncv <- nc_open(OUT_NC)
rs  <- ncvar_get(ncv, "rowSize")
chk <- tibble(
  check = c("featureType", "cf_role set", "profiles", "obs levels",
            "rowSize sums to obs", "sensor variables", "depths per profile (min/med/max)"),
  value = c(
    ncatt_get(ncv, 0, "featureType")$value,
    ncatt_get(ncv, "profile_id", "cf_role")$value,
    as.character(ncv$dim[["profile"]]$len),
    as.character(ncv$dim[["obs"]]$len),
    as.character(sum(rs) == ncv$dim[["obs"]]$len),
    as.character(length(types)),
    paste(min(rs), median(rs), max(rs), sep = " / ")))
nc_close(ncv)
kable(chk)
```

## Publish

```{r}
#| label: publish
#| eval: !expr PUBLISH
plan <- cc_netcdf_plan(OUT_NC, DATASET, RELEASE)
man  <- cc_netcdf_manifest(plan, DATASET, RELEASE, RELEASE, file.size(OUT_NC),
                           source_tables = c("sample", "obs"),
                           cf_scope = globals$cf_scope)
cat(glue("sha256: {substr(plan$sha256,1,16)}...  identical_to: {plan$identical_to}  upload: {plan$upload}\n"))
jsonlite::write_json(man, file.path(OUT_DIR, "manifest_ctd-cast.json"),
                     auto_unbox = TRUE, pretty = TRUE)
```

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
