---
title: "Publish every program dataset to EDI"
subtitle: "EML + data entities from the frozen release; PASTA evaluate always, upload gated"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_to_edi
  workflow_type: publish
  dependency:
    - release_database
  output: data/edi/manifest.csv
  workflow_url: https://calcofi.io/workflows/publish_to-edi.html
  description: >
    Publishes CalCOFI's own program datasets — the ones with no existing
    archive of record (bottle, CTD casts, underway TSG/meteorology) — as EDI
    data packages: the release's own EML 2.2 document (`build_eml()`) paired
    with each core table's rows for that dataset as CSV entities, evaluated
    against EDI's PASTA staging environment on every run that has credentials,
    and created/updated as a real package only under an explicit flag.
editor_options:
  chunk_output_type: console
---

## Overview

One EDI data package per dataset, generic over `dataset_key` — the same shape
as `publish_to-netcdf.qmd` and `publish_to-erddap.qmd`: read the frozen release,
write files, publish only under an explicit flag.

**Scope** (plan `2026-09-05 CalCOFI.io as a dataset catalog` § D-6, Decision 24):
EDI scope `edi` (an open namespace any registered account publishes under, not
an organisational registration — CCE-LTER's own `knb-lter-cce` packages stay
theirs), one package per dataset, owned by a CalCOFI EDI account. This notebook
defaults to the three program datasets that have **no existing archive of
record** — `calcofi_bottle`, `calcofi_ctd-cast`, `calcofi_mets` — because every
other CalCOFI dataset already has a home (`swfsc_ichthyo` on OBIS; the nine
CCE-LTER-adjacent datasets already live in `knb-lter-cce` and are sourced from
there, never republished).

**The non-interference rule holds generically, not just by default list.** A
requested `dataset_key` is refused — reported, not published — when its own
`link_data_source` is itself an EDI/PASTA package, or its record already
carries a `kind = "archive"` distribution on `portal %in% c("edi",
"knb-lter-cce")`. Republishing a provider's own package under a CalCOFI-owned
one would fork the record OBIS's non-interference rule (plan D-6/D-8) already
established for `publish_to-obis.qmd`.

**What becomes an entity, per table, and why** (a table is whatever
`datasets.json`'s record lists in `tables[]` for that dataset — not a hardcoded
list, so a schema change is picked up automatically):

| classification | rule | example (this run) |
|---|---|---|
| `dataTable`, CSV | the table carries a `dataset_key` column and is not `supplemental` | `sample`, `obs`, `sample_measurement` |
| `otherEntity`, whole parquet | no `dataset_key` column — a shared vocabulary/reference table, too small and too shared to duplicate a filtered copy of honestly | `measurement_type` |
| excluded (noted, not entitied) | the catalog marks the table `supplemental` — the full-resolution scan tables (`obs_ctd_full`, `obs_mets_full`) are hundreds of millions of rows partitioned by `cruise_key`, not `dataset_key`; no single file is "this dataset's slice" and enumerating every cruise partition would add 100+ entities to a package meant to be readable | `obs_ctd_full` (ctd-cast), `obs_mets_full` (mets) |

The exclusion is recorded in the EML's own `additionalMetadata` (never silent)
and reported in the plan table below with measured sizes.

**Evaluate always, upload gated.** `EDIutils::evaluate_data_package()` runs
against `env = "staging"` on every render that has EDI credentials
(`EDI_KEY`, or `EDI_USER` + `EDI_PASS`) — evaluating is non-destructive and
does not mint anything, so there is no reason to gate it behind a flag the way
`create_data_package()` / `update_data_package()` are (env = `"production"`,
only under `CALCOFI_PUBLISH_EDI = true`). Without credentials the notebook
**says so and skips cleanly** rather than prompting or failing.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(DBI, duckdb, dplyr, glue, jsonlite, digest, readr, tibble,
                 knitr, here, quiet = TRUE)
here <- here::here
options(readr.show_col_types = FALSE)
devtools::load_all(here::here("../calcofi4db"))
source(here("libs/edi_entities.R"))

# Staging by default (plan § D-6: evaluate against staging first) — and, at the
# time this notebook was written, datasets.json/eml/ exist only in the staging
# release (v2026.09.05); a promoted release will carry them once R0/E1 ship.
RELEASE_PREFIX  <- Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake-staging/releases")
RELEASE_VERSION <- edi_resolve_version(RELEASE_PREFIX, Sys.getenv("CALCOFI_RELEASE_VERSION", ""))
BASE_HTTPS      <- "https://storage.googleapis.com/calcofi-db"

# the three program datasets with no existing archive of record (plan § D-6);
# override to iterate on others (the non-interference check still applies)
DATASET_KEYS <- Filter(nzchar, trimws(strsplit(
  Sys.getenv("CALCOFI_DATASETS", "calcofi_bottle,calcofi_ctd-cast,calcofi_mets"), ",")[[1]]))

PUBLISH_EDI    <- identical(Sys.getenv("CALCOFI_PUBLISH_EDI"), "true")   # opt-in create/update
creds          <- edi_has_credentials()
NO_CREDS       <- !creds$available   # separate booleans: `!expr !x` confuses the YAML chunk-option parser
NO_PUBLISH_EDI <- !PUBLISH_EDI

OUT_DIR <- here("data/edi"); dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
PKG_REGISTRY_PATH <- here("metadata/edi_packages.csv")

cat(glue("release prefix   : {RELEASE_PREFIX}\n"))
cat(glue("release version  : {RELEASE_VERSION}\n"))
cat(glue("datasets         : {paste(DATASET_KEYS, collapse=', ')}\n"))
cat(glue("EDI credentials  : {creds$available} ({creds$method %||% 'none'})\n"))
cat(glue("CALCOFI_PUBLISH_EDI : {PUBLISH_EDI} (create/update only when true)\n"))
```

```{r}
#| label: fetch-release-json
u <- function(f) glue("{BASE_HTTPS}/{RELEASE_PREFIX}/{RELEASE_VERSION}/{f}")
datasets_json <- edi_read_json(u("datasets.json"))
catalog       <- edi_read_json(u("catalog.json"))
meta_json     <- edi_read_json(u("metadata.json"))
coverage_json <- edi_read_json(u("coverage.json"))
release_block <- datasets_json[["release"]] %||% list(version = RELEASE_VERSION)

sidecars <- read_dataset_sidecars(here("metadata"))
gear     <- read_gear_registry(here("metadata/gear.csv"))
pkg_registry <- edi_read_package_registry(PKG_REGISTRY_PATH)

meta_has_col <- function(table, col) paste0(table, ".", col) %in% names(meta_json[["columns"]] %||% list())
cat_entry <- function(table) Find(function(t) identical(t[["name"]], table), catalog[["tables"]] %||% list())
```

## The plan — every requested dataset, before anything is written

```{r}
#| label: plan
plan_rows <- list()
records <- list()
skip <- list()

for (key in DATASET_KEYS) {
  rec <- edi_dataset_record(datasets_json, key)
  if (is.null(rec)) { skip[[key]] <- "not in datasets.json for this release"; next }
  if (!identical(rec[["visibility"]], "public")) { skip[[key]] <- glue("visibility = {rec[['visibility']]}"); next }
  chk <- edi_non_interference_check(rec, sidecars[[key]])
  if (chk$blocked) { skip[[key]] <- paste("non-interference:", paste(chk$reasons, collapse = "; ")); next }
  records[[key]] <- rec
  for (tb in as.character(unlist(rec[["tables"]]))) {
    cls <- edi_classify_table(tb, cat_entry(tb), meta_has_col(tb, "dataset_key"))
    plan_rows[[length(plan_rows) + 1]] <- tibble(dataset_key = key, table = tb, class = cls$class, reason = cls$reason)
  }
}

if (length(skip))
  cat("refused (non-interference or not eligible):\n",
     paste(sprintf("  - %s: %s", names(skip), unlist(skip)), collapse = "\n"), "\n\n")

plan_tbl <- if (length(plan_rows)) bind_rows(plan_rows) else
  tibble(dataset_key = character(), table = character(), class = character(), reason = character())
kable(plan_tbl, caption = glue("EDI entity plan for {RELEASE_VERSION} — {length(records)} dataset(s)"))
```

## Build EML, export entities

```{r}
#| label: connect
con <- dbConnect(duckdb())
for (s in c("INSTALL httpfs", "LOAD httpfs", "SET enable_progress_bar=false"))
  try(dbExecute(con, s), silent = TRUE)
```

```{r}
#| label: export
#| results: asis

manifest_rows <- list()

for (key in names(records)) {
  rec <- records[[key]]
  doc <- build_eml(rec, sidecar = sidecars[[key]], meta = meta_json, coverage = coverage_json,
                   release = release_block, gear = gear)

  pkg_dir <- file.path(OUT_DIR, key, glue("{key}_{RELEASE_VERSION}"))
  dir.create(pkg_dir, recursive = TRUE, showWarnings = FALSE)

  rows <- plan_tbl |> filter(dataset_key == key)
  hashes <- character(); bytes_total <- 0
  n_csv <- 0L; n_other <- 0L; n_excl <- 0L

  for (i in seq_len(nrow(rows))) {
    tb <- rows$table[i]; cls <- rows$class[i]
    if (cls == "excluded_supplemental") {
      doc <- edi_note_excluded_table(doc, tb, rows$reason[i])
      n_excl <- n_excl + 1L
      cat(glue("- `{key}`/`{tb}`: excluded — {rows$reason[i]}\n"))
      next
    }
    if (cls == "other_ref") {
      obj <- edi_first_object(catalog, tb, BASE_HTTPS)
      if (is.null(obj)) { cat(glue("- `{key}`/`{tb}`: NOTE no catalog object found, skipping\n")); next }
      doc <- edi_add_other_entity(doc, tb, rows$reason[i], basename(obj$path), obj$bytes, obj$sha256, obj$url)
      hashes <- c(hashes, obj$sha256); bytes_total <- bytes_total + obj$bytes; n_other <- n_other + 1L
      cat(glue("- `{key}`/`{tb}`: otherEntity -> {basename(obj$path)} ({fmt_mb0(obj$bytes)})\n"))
      next
    }
    # cls == "csv": write this dataset's rows for `tb` to a local CSV
    csv_path <- file.path(pkg_dir, glue("{tb}.csv"))
    plan_i <- edi_table_read_plan(catalog, tb, key)
    from_sql <- if (plan_i$mode == "partition")
      glue("read_parquet('{plan_i$url}')") else
      glue("read_parquet([{paste(sprintf(\"'%s'\", plan_i$urls), collapse=', ')}]) {plan_i$filter_sql}")
    dbExecute(con, glue(
      "COPY (SELECT * FROM {from_sql}) TO '{csv_path}' (FORMAT CSV, HEADER, DELIMITER ',', NULLSTR '')"))
    csv_bytes  <- file.size(csv_path)
    csv_sha256 <- digest::digest(csv_path, algo = "sha256", file = TRUE)
    n_rows     <- dbGetQuery(con, glue("SELECT COUNT(*) n FROM read_csv_auto('{csv_path}')"))$n
    doc <- edi_rewrite_datatable_physical(doc, tb, basename(csv_path), csv_bytes, csv_sha256)
    hashes <- c(hashes, csv_sha256); bytes_total <- bytes_total + csv_bytes; n_csv <- n_csv + 1L
    cat(glue("- `{key}`/`{tb}`: {format(n_rows, big.mark=',')} rows -> `{basename(csv_path)}` ({fmt_mb0(csv_bytes)})\n"))
  }

  eml_path <- file.path(pkg_dir, glue("{key}.xml"))
  EML::write_eml(doc, eml_path)
  eml_sha256 <- digest::digest(eml_path, algo = "sha256", file = TRUE)
  hashes <- c(hashes, eml_sha256); bytes_total <- bytes_total + file.size(eml_path)

  chk <- check_eml(doc, path = eml_path, record = rec)
  bad <- chk |> filter(level == "error", !exempt)
  if (nrow(bad)) cat(glue("  EML check: {nrow(bad)} blocking finding(s) — see below\n"))
  print(kable(chk |> filter(finding != "ok"), caption = glue("check_eml(): {key}")))

  content_hash <- edi_content_hash(hashes)
  pkg_id <- edi_package_id_for(pkg_registry, key)
  jsonlite::write_json(list(
    dataset_key = key, version = RELEASE_VERSION, content_hash = content_hash,
    package_id = if (is.na(pkg_id)) NULL else pkg_id, n_csv = n_csv, n_other_ref = n_other,
    n_excluded = n_excl, bytes_total = bytes_total,
    evaluated_utc = NULL, uploaded_utc = NULL),
    file.path(pkg_dir, "manifest.json"), auto_unbox = TRUE, pretty = TRUE, null = "null")

  manifest_rows[[key]] <- edi_manifest_row(key, RELEASE_VERSION, content_hash,
                                           n_csv = n_csv, n_other_ref = n_other, n_excluded = n_excl,
                                           bytes_total = bytes_total, package_id = pkg_id)
}
```

```{r}
#| label: manifest
manifest_tbl <- if (length(manifest_rows)) bind_rows(manifest_rows) else
  edi_manifest_row(character(), character(), character())[0, ]
write_csv(manifest_tbl, file.path(OUT_DIR, "manifest.csv"), na = "")
kable(manifest_tbl, caption = "data/edi/manifest.csv")
```

::: {.callout-note title="Measured, `calcofi_bottle`"}
`sample` and `sample_measurement` are shared, single-file tables (all 16
datasets' rows in one object) filtered here to `dataset_key = 'calcofi_bottle'`;
`obs` is already partitioned by `dataset_key`, so its object *is* this
dataset's rows with no filter needed. `measurement_type` (a 200-row vocabulary
table with no `dataset_key` column) is named whole as an `otherEntity` rather
than duplicated per dataset. See the `export` chunk's printed sizes above for
this run's measured byte counts and row counts.
:::

## Evaluate against EDI's PASTA staging environment (always, when credentials exist)

```{r}
#| label: evaluate
#| eval: !expr creds$available

librarian::shelf(EDIutils, quiet = TRUE)
if (identical(creds$method, "key")) {
  EDIutils::login(key = Sys.getenv("EDI_KEY"))
} else {
  EDIutils::login(userId = Sys.getenv("EDI_USER"), userPass = Sys.getenv("EDI_PASS"))
}

# EDI's evaluate/create/update fetch each entity from its EML
# physical/distribution/online/url over the open web — a LOCAL file is not
# enough. Stage this run's entities + EML at a public, deterministic URL under
# the same staging bucket the release itself uses, so nothing here depends on a
# host that could disappear between this render and PASTA's fetch.
edi_stage_prefix <- glue("ducklake-staging/edi/{{key}}/{{key}}_{RELEASE_VERSION}")
evaluate_reports <- list()
for (key in names(records)) {
  pkg_dir <- file.path(OUT_DIR, key, glue("{key}_{RELEASE_VERSION}"))
  files <- list.files(pkg_dir, full.names = TRUE)
  stage_prefix <- glue(edi_stage_prefix, key = key)
  for (f in files) put_gcs_file(f, glue("gs://calcofi-db/{stage_prefix}/{basename(f)}"))
  staged_url <- function(f) glue("{BASE_HTTPS}/{stage_prefix}/{basename(f)}")

  doc <- EML::read_eml(file.path(pkg_dir, glue("{key}.xml")))
  for (i in seq_along(doc$dataset$dataTable %||% list()))
    doc$dataset$dataTable[[i]]$physical$distribution <-
      list(online = list(`function` = "download", url = staged_url(doc$dataset$dataTable[[i]]$physical$objectName)))
  for (i in seq_along(doc$dataset$otherEntity %||% list()))
    doc$dataset$otherEntity[[i]]$physical$distribution <-
      list(online = list(`function` = "download", url = staged_url(doc$dataset$otherEntity[[i]]$physical$objectName)))
  eml_staged_path <- file.path(pkg_dir, glue("{key}.staged.xml"))
  EML::write_eml(doc, eml_staged_path)
  put_gcs_file(eml_staged_path, glue("gs://calcofi-db/{stage_prefix}/{key}.xml"))

  tx <- tryCatch(
    EDIutils::evaluate_data_package(eml = eml_staged_path, env = "staging"),
    error = function(e) { message(glue("evaluate failed for {key}: {conditionMessage(e)}")); NULL })
  if (is.null(tx)) next
  EDIutils::check_status_evaluate(tx, env = "staging")
  rpt <- tryCatch(EDIutils::read_evaluate_report_summary(tx, with_exceptions = FALSE, env = "staging"),
                  error = function(e) conditionMessage(e))
  evaluated_utc <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  writeLines(as.character(rpt), file.path(pkg_dir, "evaluate_report.txt"))
  evaluate_reports[[key]] <- list(transaction = tx, summary = rpt, evaluated_utc = evaluated_utc)
  cat(glue("- `{key}`: evaluate transaction `{tx}` — report written to evaluate_report.txt\n"))
}
EDIutils::logout()
```

```{r}
#| label: evaluate-skip
#| eval: !expr NO_CREDS
cat("EDI_USER/EDI_PASS (or EDI_KEY) are not set — evaluate_data_package() was not run.\n",
   "Set one of those and re-render to evaluate against EDI's staging environment.\n")
```

## Publish — create or update the real EDI package (gated)

```{r}
#| label: publish
#| eval: !expr PUBLISH_EDI

# Only reached when CALCOFI_PUBLISH_EDI=true AND (for the create/update call
# itself) EDI credentials are present — never run as part of this repo's own
# CI/staging checks, and never targeting anything but env = "production".
stopifnot("CALCOFI_PUBLISH_EDI=true requires EDI credentials" = creds$available)
librarian::shelf(EDIutils, quiet = TRUE)
if (identical(creds$method, "key")) EDIutils::login(key = Sys.getenv("EDI_KEY")) else
  EDIutils::login(userId = Sys.getenv("EDI_USER"), userPass = Sys.getenv("EDI_PASS"))

publish_rows <- list()
for (key in names(records)) {
  pkg_dir <- file.path(OUT_DIR, key, glue("{key}_{RELEASE_VERSION}"))
  eml_staged_path <- file.path(pkg_dir, glue("{key}.staged.xml"))
  stopifnot("run the `evaluate` chunk first (it stages entities + rewrites physical URLs)" = file.exists(eml_staged_path))
  existing <- edi_package_id_for(pkg_registry, key)
  uploaded_utc <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  tx <- if (is.na(existing))
    EDIutils::create_data_package(eml = eml_staged_path, env = "production") else
    EDIutils::update_data_package(eml = eml_staged_path, env = "production")
  ok <- if (is.na(existing)) EDIutils::check_status_create(tx, env = "production") else
    EDIutils::check_status_update(tx, env = "production")
  # EDIutils' create_data_package() example names its transaction
  # "create_<timestamp>__<scope.id.rev>" (?EDIutils::create_data_package); the
  # update case is not documented as explicitly, so this is a best-effort parse
  # — UNVERIFIED against a real transaction (this repo has never held EDI
  # credentials). Confirm the pattern on the first real run and, if it does not
  # match, read the package id back with EDIutils::list_data_package_identifiers()
  # / read_data_package_report_summary(tx) instead of trusting this regex.
  pkg_id <- sub("^(create|update)_[0-9]+__?", "", tx)
  publish_rows[[key]] <- tibble(dataset_key = key, package_id = pkg_id, uploaded_utc = uploaded_utc, ok = ok)
  cat(glue("- `{key}`: {if (is.na(existing)) 'create' else 'update'} -> `{pkg_id}` (status ok={ok})\n"))
}
EDIutils::logout()

if (length(publish_rows)) {
  new_rows <- bind_rows(publish_rows) |>
    mutate(scope = "edi", identifier = sub("^edi\\.([0-9]+)\\..*$", "\\1", package_id),
          revision = sub("^edi\\.[0-9]+\\.([0-9]+)$", "\\1", package_id), env = "production",
          created_utc = uploaded_utc, updated_utc = uploaded_utc) |>
    select(dataset_key, scope, identifier, revision, env, package_id, created_utc, updated_utc)
  pkg_registry <- bind_rows(pkg_registry |> filter(!dataset_key %in% new_rows$dataset_key), new_rows)
  write_csv(pkg_registry, PKG_REGISTRY_PATH, na = "")
}
```

```{r}
#| label: publish-skip
#| eval: !expr NO_PUBLISH_EDI
cat("CALCOFI_PUBLISH_EDI is not `true` — create_data_package()/update_data_package() were not run.\n",
   "Set CALCOFI_PUBLISH_EDI=true (with EDI credentials) to mint or revise a real EDI package.\n")
```

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
