---
title: "Publish every program dataset to EDI"
subtitle: "EML + data entities from the frozen release; rebuilt only when a dataset changed; PASTA evaluate always, upload gated"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_to_edi
  workflow_type: publish
  dependency:
    - test_release
  output: data/edi/manifest.csv
  workflow_url: https://calcofi.io/workflows/publish_to-edi.html
  description: >
    Publishes CalCOFI's own program datasets — the ones with no existing
    archive of record (bottle, CTD casts, underway TSG/meteorology) — as EDI
    data packages: the release's own EML 2.2 document (`build_eml()`) paired
    with each core table's rows for that dataset as CSV entities. Runs after
    every promoted release but rebuilds a package only when that dataset's
    rows, metadata or this code changed; evaluates against EDI's PASTA staging
    environment when credentials exist; creates/updates the real package only
    under an explicit flag, and says which packages EDI is behind on.
editor_options:
  chunk_output_type: console
---

## Overview

One EDI data package per dataset, generic over `dataset_key` — the same shape
as `publish_to-netcdf.qmd` and `publish_to-erddap.qmd`: read the frozen release,
write files, publish only under an explicit flag.

**Scope** (plan `2026-09-05 CalCOFI.io as a dataset catalog` § D-6, Decision 24):
EDI scope `edi` (an open namespace any registered account publishes under, not
an organisational registration — CCE-LTER's own `knb-lter-cce` packages stay
theirs), one package per dataset, owned by a CalCOFI EDI account. This notebook
defaults to the three program datasets that have **no existing archive of
record** — `calcofi_bottle`, `calcofi_ctd-cast`, `calcofi_mets` — because every
other CalCOFI dataset already has a home (`swfsc_ichthyo` on OBIS; `calcofi_dic`
at NCEI; the nine CCE-LTER-adjacent datasets already live in `knb-lter-cce` and
are sourced from there, never republished).

**The non-interference rule holds generically, not just by default list.** A
requested `dataset_key` is refused — reported, not published — when its own
`link_data_source` is itself an EDI/PASTA package, or its record already
carries a `kind = "archive"` distribution on `portal %in% c("edi",
"knb-lter-cce")`. Republishing a provider's own package under a CalCOFI-owned
one would fork the record OBIS's non-interference rule (plan D-6/D-8) already
established for `publish_to-obis.qmd`.

**What becomes an entity, per table, and why** (a table is whatever
`datasets.json`'s record lists in `tables[]` for that dataset — not a hardcoded
list, so a schema change is picked up automatically):

| classification | rule | example (this run) |
|---|---|---|
| `dataTable`, CSV | the table carries a `dataset_key` column and is not `supplemental` | `sample`, `obs`, `sample_measurement` |
| `otherEntity`, whole parquet | no `dataset_key` column — a shared vocabulary/reference table, too small and too shared to duplicate a filtered copy of honestly | `measurement_type` |
| excluded (noted, not entitied) | the catalog marks the table `supplemental` — the full-resolution scan tables (`obs_ctd_full`, `obs_mets_full`) are hundreds of millions of rows partitioned by `cruise_key`, not `dataset_key`; no single file is "this dataset's slice" and enumerating every cruise partition would add 100+ entities to a package meant to be readable | `obs_ctd_full` (ctd-cast), `obs_mets_full` (mets) |

The exclusion is recorded in the EML's own `additionalMetadata` (never silent)
and reported in the plan table below with measured sizes.

**Rebuilt only when the dataset changed.** The notebook runs after every
promoted release (it depends on `test_release`), but most releases leave most
datasets alone, and a rebuild would re-export gigabytes of CSV and — worse — ask
EDI for a new revision of data that did not change. So each package carries an
input fingerprint (`calcofi4db::publish_fingerprint()`): the dataset's rows in
every table it exports, identified by the release's own row signatures rather
than by bytes or version; its catalog record and the registries the EML reads;
and this notebook's code. An unchanged fingerprint reuses the package built from
an earlier release (its `version` stays the release it was built from, and
`checked_version` records the release it was confirmed against); anything else
rebuilds, and the plan says which input moved.

**Evaluate always, upload gated, and say when EDI is behind.**
`EDIutils::evaluate_data_package()` runs against `env = "staging"` on every
render that has EDI credentials (`EDI_KEY`, or `EDI_USER` + `EDI_PASS`) for
each package whose bytes it has not yet evaluated — evaluating is
non-destructive and does not mint anything. `create_data_package()` /
`update_data_package()` (env = `"production"`) run only under
`CALCOFI_PUBLISH_EDI = true`, and only for a package whose `content_hash`
differs from the one `metadata/edi_packages.csv` records as deposited. That same
comparison is the *Upload due* table at the end, which `publish_status.qmd`
gathers with the other portals'.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(DBI, duckdb, dplyr, glue, jsonlite, digest, readr, tibble,
                 knitr, here, quiet = TRUE)
here <- here::here
options(readr.show_col_types = FALSE)
devtools::load_all(here::here("../calcofi4db"))
source(here("libs/edi_entities.R"))

# the promoted release by default; a STAGING run (CALCOFI_RELEASE_PREFIX=
# ducklake-staging/releases) reads the staging release, writes under
# data/edi-staging/ and stages nothing to GCS, so it can never be mistaken for, or
# overwrite, what was built from the promoted one (the OBIS publisher's rule)
RELEASE_PREFIX  <- Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake/releases")
STAGING         <- grepl("staging", RELEASE_PREFIX, fixed = TRUE)
RELEASE_VERSION <- edi_resolve_version(RELEASE_PREFIX, Sys.getenv("CALCOFI_RELEASE_VERSION", ""))
BASE_HTTPS      <- "https://storage.googleapis.com/calcofi-db"

# the three program datasets with no existing archive of record (plan § D-6);
# override to iterate on others (the non-interference check still applies)
DATASET_KEYS <- Filter(nzchar, trimws(strsplit(
  Sys.getenv("CALCOFI_DATASETS", "calcofi_bottle,calcofi_ctd-cast,calcofi_mets"), ",")[[1]]))

PUBLISH_EDI    <- identical(Sys.getenv("CALCOFI_PUBLISH_EDI"), "true")   # opt-in create/update
creds          <- edi_has_credentials()
# separate booleans: `!expr !x` confuses the YAML chunk-option parser
NO_CREDS       <- !creds$available
NO_PUBLISH_EDI <- !PUBLISH_EDI
DO_EVALUATE    <- creds$available && !STAGING   # a staging run never talks to EDI
DO_PUBLISH_EDI <- PUBLISH_EDI && !STAGING

OUT_DIR <- here(if (STAGING) "data/edi-staging" else "data/edi")
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
PKG_REGISTRY_PATH <- here("metadata/edi_packages.csv")
# where a package is staged for review, and where its EML says each entity downloads from
EDI_STAGE_ROOT <- "publish/edi"
staged_url <- function(key, version, file)
  glue("{BASE_HTTPS}/{EDI_STAGE_ROOT}/{key}/{key}_{version}/{file}")
# per-dataset row signatures of shared release objects, cached by content_hash
# (true forever: objects are content-addressed); shared by every publisher
SIG_CACHE <- here("data/publish/object_signatures.csv")
# the code the packages are a function of: a change here rebuilds them
CODE_FILES <- c(here("publish_to-edi.qmd"), here("libs/edi_entities.R"),
                here("../calcofi4db/R/eml.R"))

cat(glue("release prefix   : {RELEASE_PREFIX}{if (STAGING) ' (STAGING)' else ''}\n"))
cat(glue("release version  : {RELEASE_VERSION}\n"))
cat(glue("datasets         : {paste(DATASET_KEYS, collapse=', ')}\n"))
cat(glue("output           : {OUT_DIR}\n"))
cat(glue("EDI credentials  : {creds$available} ({creds$method %||% 'none'})\n"))
cat(glue("CALCOFI_PUBLISH_EDI : {PUBLISH_EDI} (create/update only when true)\n"))
```

```{r}
#| label: fetch-release-json
u <- function(f) glue("{BASE_HTTPS}/{RELEASE_PREFIX}/{RELEASE_VERSION}/{f}")
datasets_json <- edi_read_json(u("datasets.json"))
catalog       <- edi_read_json(u("catalog.json"))
meta_json     <- edi_read_json(u("metadata.json"))
coverage_json <- edi_read_json(u("coverage.json"))
release_block <- datasets_json[["release"]] %||% list(version = RELEASE_VERSION)

sidecars <- read_dataset_sidecars(here("metadata"))
gear     <- read_gear_registry(here("metadata/gear.csv"))
pkg_registry <- edi_read_package_registry(PKG_REGISTRY_PATH)

meta_has_col <- function(table, col) paste0(table, ".", col) %in% names(meta_json[["columns"]] %||% list())
cat_entry <- function(table) Find(function(t) identical(t[["name"]], table), catalog[["tables"]] %||% list())

# table -> the datasets that read it: a table only one dataset reads is signed by its
# own content_hash and never scanned
owners <- list()
for (r in datasets_json[["datasets"]] %||% list())
  for (tb in as.character(unlist(r[["tables"]])))
    owners[[tb]] <- c(owners[[tb]], r[["dataset_key"]])
```

## The plan — every requested dataset, before anything is written

```{r}
#| label: plan
plan_rows <- list()
records <- list()
skip <- list()

for (key in DATASET_KEYS) {
  rec <- edi_dataset_record(datasets_json, key)
  if (is.null(rec)) { skip[[key]] <- "not in datasets.json for this release"; next }
  if (!identical(rec[["visibility"]], "public")) { skip[[key]] <- glue("visibility = {rec[['visibility']]}"); next }
  chk <- edi_non_interference_check(rec, sidecars[[key]])
  if (chk$blocked) { skip[[key]] <- paste("non-interference:", paste(chk$reasons, collapse = "; ")); next }
  records[[key]] <- rec
  for (tb in as.character(unlist(rec[["tables"]]))) {
    cls <- edi_classify_table(tb, cat_entry(tb), meta_has_col(tb, "dataset_key"))
    plan_rows[[length(plan_rows) + 1]] <- tibble(dataset_key = key, table = tb, class = cls$class, reason = cls$reason)
  }
}

if (length(skip))
  cat("refused (non-interference or not eligible):\n",
     paste(sprintf("  - %s: %s", names(skip), unlist(skip)), collapse = "\n"), "\n\n")

plan_tbl <- if (length(plan_rows)) bind_rows(plan_rows) else
  tibble(dataset_key = character(), table = character(), class = character(), reason = character())
kable(plan_tbl, caption = glue("EDI entity plan for {RELEASE_VERSION} — {length(records)} dataset(s)"))
```

## What changed since each package was built

```{r}
#| label: connect
# get_duckdb_con() honours CALCOFI_DUCKDB_MEMORY_LIMIT / CALCOFI_DUCKDB_THREADS and
# spills to tempdir(), so a publisher sharing the machine with an ingest stays bounded
con <- get_duckdb_con()
for (s in c("INSTALL httpfs", "LOAD httpfs", "SET enable_progress_bar=false"))
  try(dbExecute(con, s), silent = TRUE)
# the CSV entities are hashed, so their row order must be a function of the release, not
# of the scheduler: DuckDB keeps a scan's insertion order when this is on (its default,
# pinned here so a future default cannot change the bytes), and the release objects are
# themselves written under a total ORDER BY
dbExecute(con, "SET preserve_insertion_order = true")
```

```{r}
#| label: fingerprint
# every table a package is a function of: the CSV entities and the whole-object
# references (an excluded supplemental contributes nothing but its name)
entity_tables <- function(key)
  plan_tbl$table[plan_tbl$dataset_key == key & plan_tbl$class != "excluded_supplemental"]
sigs <- publish_object_signatures(con, catalog, unique(unlist(lapply(names(records), entity_tables))),
                                  owners = owners, cache = SIG_CACHE, base_url = BASE_HTTPS)
code_parts <- publish_code_parts(CODE_FILES)

fps <- list(); decisions <- list(); priors <- list()
for (key in names(records)) {
  tbls <- entity_tables(key)
  cols <- meta_json[["columns"]] %||% list()
  taxa <- Filter(function(t) key %in% vapply(t[["datasets"]] %||% list(), function(d) d[["dataset_key"]] %||% "", ""),
                 coverage_json[["taxa"]] %||% list())
  fps[[key]] <- publish_fingerprint(
    data     = publish_data_parts(sigs, key, tbls),
    metadata = list(
      record   = publish_record_digest(records[[key]]),
      sidecar  = sidecars[[key]],
      tables   = (meta_json[["tables"]] %||% list())[intersect(tbls, names(meta_json[["tables"]]))],
      columns  = cols[sub("\\..*$", "", names(cols)) %in% tbls],
      taxa     = taxa,
      gear     = gear,
      excluded = plan_tbl$table[plan_tbl$dataset_key == key & plan_tbl$class == "excluded_supplemental"]),
    code     = code_parts)
  priors[[key]] <- edi_latest_package(OUT_DIR, key)
  pm <- priors[[key]]$manifest
  outputs <- if (is.null(pm)) character() else
    file.path(priors[[key]]$dir, c(glue("{key}.xml"), as.character(unlist(pm[["files"]]))))
  decisions[[key]] <- publish_decide(fps[[key]], pm[["input_fingerprint"]], outputs)
  if (!is.null(pm) && is.null(pm[["input_fingerprint"]]))
    decisions[[key]]$reason <- glue("the {pm$version} build predates input fingerprints")
}

kable(tibble(
  dataset_key = names(decisions),
  previous    = vapply(names(decisions), function(k) priors[[k]]$manifest$version %||% "—", ""),
  action      = vapply(decisions, `[[`, "", "action"),
  why         = vapply(decisions, `[[`, "", "reason")),
  caption = glue("reuse or rebuild, against {RELEASE_VERSION}"))
```

## Build EML, export entities

```{r}
#| label: export
#| results: asis

manifest_rows <- list()
pkg_dirs <- list()

for (key in names(records)) {
  rec <- records[[key]]
  up  <- edi_uploaded_for(pkg_registry, key)

  # ---- reuse: the dataset did not change since an earlier release's build ----
  if (decisions[[key]]$action == "reuse") {
    pkg_dir <- priors[[key]]$dir
    pm <- priors[[key]]$manifest
    eml_path <- file.path(pkg_dir, glue("{key}.xml"))
    # the schema check is cheap and the one that let v2026.09.06's packages through, so a
    # reused package is re-validated; the record-level findings were stored at build time
    # and still hold, because a changed record would have changed the fingerprint
    v <- EML::eml_validate(eml_path)
    if (!isTRUE(as.logical(v)))
      stop(glue("{key}: reused EML fails EML::eml_validate(): ",
                "{paste(head(as.character(attr(v, 'errors')), 3), collapse = ' | ')}"))
    pm$checked_version <- RELEASE_VERSION
    jsonlite::write_json(pm, file.path(pkg_dir, "manifest.json"), auto_unbox = TRUE, pretty = TRUE,
                         null = "null")
    pkg_dirs[[key]] <- pkg_dir
    manifest_rows[[key]] <- edi_manifest_row(
      key, pm$version, pm$content_hash, n_csv = pm$n_csv, n_other_ref = pm$n_other_ref,
      n_excluded = pm$n_excluded, bytes_total = pm$bytes_total,
      package_id = pm$package_id %||% NA_character_, evaluated_utc = pm$evaluated_utc %||% NA_character_,
      uploaded_utc = up$uploaded_utc, checked_version = RELEASE_VERSION, action = "reuse",
      changed = "", input_fingerprint = fps[[key]]$hash,
      eml_blocking = pm$eml_blocking %||% "", uploaded_hash = up$uploaded_hash,
      uploaded_version = up$uploaded_version)
    cat(glue("- `{key}`: unchanged since {pm$version} — package reused ({pm$n_csv} CSV, ",
             "{fmt_mb0(pm$bytes_total)}, content_hash {substr(pm$content_hash, 1, 12)})\n"))
    next
  }

  # ---- build ----------------------------------------------------------------
  doc <- build_eml(rec, sidecar = sidecars[[key]], meta = meta_json, coverage = coverage_json,
                   release = release_block, gear = gear)
  pkg_dir <- file.path(OUT_DIR, key, glue("{key}_{RELEASE_VERSION}"))
  unlink(pkg_dir, recursive = TRUE)
  dir.create(pkg_dir, recursive = TRUE, showWarnings = FALSE)
  cat(glue("- `{key}`: rebuilding — {decisions[[key]]$reason}\n"))

  rows <- plan_tbl |> filter(dataset_key == key)
  hashes <- character(); bytes_total <- 0; files <- character()
  n_csv <- 0L; n_other <- 0L; n_excl <- 0L

  for (i in seq_len(nrow(rows))) {
    tb <- rows$table[i]; cls <- rows$class[i]
    if (cls == "excluded_supplemental") {
      doc <- edi_note_excluded_table(doc, tb, rows$reason[i])
      n_excl <- n_excl + 1L
      cat(glue("  - `{tb}`: excluded — {rows$reason[i]}\n"))
      next
    }
    if (cls == "other_ref") {
      obj <- edi_first_object(catalog, tb, BASE_HTTPS)
      if (is.null(obj)) { cat(glue("  - `{tb}`: NOTE no catalog object found, skipping\n")); next }
      doc <- edi_add_other_entity(doc, tb, rows$reason[i], basename(obj$path), obj$bytes, obj$sha256, obj$url)
      hashes <- c(hashes, obj$sha256); bytes_total <- bytes_total + obj$bytes; n_other <- n_other + 1L
      cat(glue("  - `{tb}`: otherEntity -> {basename(obj$path)} ({fmt_mb0(obj$bytes)})\n"))
      next
    }
    # cls == "csv": write this dataset's rows for `tb` to a local CSV
    csv_path <- file.path(pkg_dir, glue("{tb}.csv"))
    plan_i <- edi_table_read_plan(catalog, tb, key)
    from_sql <- if (plan_i$mode == "partition")
      glue("read_parquet('{plan_i$url}')") else
      glue("read_parquet([{paste(sprintf(\"'%s'\", plan_i$urls), collapse=', ')}]) {plan_i$filter_sql}")
    dbExecute(con, glue(
      "COPY (SELECT * FROM {from_sql}) TO '{csv_path}' (FORMAT CSV, HEADER, DELIMITER ',', NULLSTR '')"))
    csv_bytes  <- file.size(csv_path)
    csv_sha256 <- digest::digest(csv_path, algo = "sha256", file = TRUE)
    n_rows     <- dbGetQuery(con, glue("SELECT COUNT(*) n FROM read_csv_auto('{csv_path}')"))$n
    # the EML names the entity's public download URL: the staged copy PASTA fetches
    doc <- edi_rewrite_datatable_physical(doc, tb, basename(csv_path), csv_bytes, csv_sha256,
                                          url = staged_url(key, RELEASE_VERSION, basename(csv_path)))
    hashes <- c(hashes, csv_sha256); bytes_total <- bytes_total + csv_bytes; n_csv <- n_csv + 1L
    files <- c(files, basename(csv_path))
    cat(glue("  - `{tb}`: {format(n_rows, big.mark=',')} rows -> `{basename(csv_path)}` ({fmt_mb0(csv_bytes)})\n"))
  }

  eml_path <- file.path(pkg_dir, glue("{key}.xml"))
  EML::write_eml(doc, eml_path)
  eml_sha256 <- digest::digest(eml_path, algo = "sha256", file = TRUE)
  hashes <- c(hashes, eml_sha256); bytes_total <- bytes_total + file.size(eml_path)

  # a schema-invalid document is a bug in this code, so it stops the render; a gap in the
  # record (no licence, no creator) builds and stages the package for review but blocks
  # the deposit, and is reported until the record is filled. An open question exempts a
  # finding from failing the RELEASE (check_eml()'s rule), never from blocking a deposit:
  # a package with no licence must not reach EDI because someone asked what it should be
  chk <- check_eml(doc, path = eml_path, record = rec)
  if (any(chk$finding == "invalid_eml"))
    stop(glue("{key}: EML fails EML::eml_validate(): {chk$detail[chk$finding == 'invalid_eml'][1]}"))
  errs <- chk[chk$level == "error", , drop = FALSE]
  blocking <- ifelse(errs$exempt & !is.na(errs$question),
                     glue("{errs$finding} (open: {errs$question})"), errs$finding)
  if (length(blocking)) cat(glue("  - EML check: blocks deposit — {paste(blocking, collapse = ', ')}\n"))
  print(kable(chk |> filter(finding != "ok"), caption = glue("check_eml(): {key}")))

  content_hash <- edi_content_hash(hashes)
  pkg_id <- edi_package_id_for(pkg_registry, key)
  jsonlite::write_json(list(
    dataset_key = key, version = RELEASE_VERSION, checked_version = RELEASE_VERSION,
    content_hash = content_hash, input_fingerprint = list(hash = fps[[key]]$hash, parts = as.list(fps[[key]]$parts)),
    files = files, package_id = if (is.na(pkg_id)) NULL else pkg_id, n_csv = n_csv,
    n_other_ref = n_other, n_excluded = n_excl, bytes_total = bytes_total,
    eml_findings = chk |> filter(finding != "ok") |> select(finding, level, exempt, detail),
    eml_blocking = paste(blocking, collapse = ","),
    evaluated_utc = NULL, evaluated_hash = NULL),
    file.path(pkg_dir, "manifest.json"), auto_unbox = TRUE, pretty = TRUE, null = "null")

  # the build before this one is superseded; it is reproducible from its frozen release
  old <- setdiff(Sys.glob(file.path(OUT_DIR, key, glue("{key}_v*"))), pkg_dir)
  if (length(old)) { unlink(old, recursive = TRUE); cat(glue("  - removed superseded {paste(basename(old), collapse = ', ')}\n")) }

  pkg_dirs[[key]] <- pkg_dir
  manifest_rows[[key]] <- edi_manifest_row(
    key, RELEASE_VERSION, content_hash, n_csv = n_csv, n_other_ref = n_other, n_excluded = n_excl,
    bytes_total = bytes_total, package_id = pkg_id, uploaded_utc = up$uploaded_utc,
    checked_version = RELEASE_VERSION, action = "build",
    changed = paste(decisions[[key]]$changed, collapse = ","), input_fingerprint = fps[[key]]$hash,
    eml_blocking = paste(blocking, collapse = ","), uploaded_hash = up$uploaded_hash,
    uploaded_version = up$uploaded_version)
}
```

```{r}
#| label: manifest
manifest_tbl <- if (length(manifest_rows)) bind_rows(manifest_rows) else
  edi_manifest_row(character(), character(), character())[0, ]
write_csv(manifest_tbl, file.path(OUT_DIR, "manifest.csv"), na = "")
kable(manifest_tbl |> select(dataset_key, version, checked_version, action, content_hash,
                             eml_blocking, upload_status),
      caption = glue("{basename(OUT_DIR)}/manifest.csv"))
```

## Stage the packages where a reviewer can see them

Before any `evaluate`/`create` at EDI, each package (the CSV entities, the EML, its
`manifest.json`) is copied to a public, deterministic address —
`gs://calcofi-db/publish/edi/{dataset_key}/{dataset_key}_{version}/`, `version` being the
release it was built from — so the provider and the dataset page
(calcofi.io/datasets/{dataset_key}/, *Archives & portals*, "built, not deposited") can
inspect it, and so PASTA can fetch each entity from the URL its EML names. An unchanged
file is not re-uploaded (`put_gcs_file()` compares md5s); a superseded version's folder
is removed once the new one is staged. A staging run stages nothing.

```{r}
#| label: stage
if (!STAGING) {
  for (key in names(pkg_dirs)) {
    pkg_dir <- pkg_dirs[[key]]
    files <- list.files(pkg_dir, full.names = TRUE)
    if (!length(files)) next
    stage_prefix <- glue("{EDI_STAGE_ROOT}/{key}/{basename(pkg_dir)}")
    for (f in files) put_gcs_file(f, glue("gs://calcofi-db/{stage_prefix}/{basename(f)}"))
    cat(glue("staged {key}: {length(files)} file(s) at https://storage.googleapis.com/calcofi-db/{stage_prefix}/"), "\n")
    staged <- unique(sub("^(.*/[^/]+_v[0-9.]+)/.*$", "\\1",
                         list_gcs_files("calcofi-db", prefix = glue("{EDI_STAGE_ROOT}/{key}/"))$name))
    for (p in setdiff(staged[grepl(glue("/{key}_v"), staged, fixed = TRUE)], stage_prefix)) {
      delete_gcs_prefix(glue("{p}/"), bucket = "calcofi-db", dry_run = FALSE)
      cat(glue("  removed superseded gs://calcofi-db/{p}/"), "\n")
    }
  }
  put_gcs_file(file.path(OUT_DIR, "manifest.csv"), "gs://calcofi-db/publish/edi/manifest.csv")
} else cat("staging run: nothing staged to GCS\n")
```

::: {.callout-note title="Measured, `calcofi_bottle`"}
`sample` and `sample_measurement` are shared, single-file tables (all 16
datasets' rows in one object) filtered here to `dataset_key = 'calcofi_bottle'`;
`obs` is already partitioned by `dataset_key`, so its object *is* this
dataset's rows with no filter needed. `measurement_type` (a 200-row vocabulary
table with no `dataset_key` column) is named whole as an `otherEntity` rather
than duplicated per dataset. See the `export` chunk's printed sizes above for
this run's measured byte counts and row counts.
:::

## Evaluate against EDI's PASTA staging environment (always, when credentials exist)

```{r}
#| label: evaluate
#| eval: !expr DO_EVALUATE

librarian::shelf(EDIutils, quiet = TRUE)
if (identical(creds$method, "key")) {
  EDIutils::login(key = Sys.getenv("EDI_KEY"))
} else {
  EDIutils::login(userId = Sys.getenv("EDI_USER"), userPass = Sys.getenv("EDI_PASS"))
}

# EDI's evaluate/create/update fetch each entity from its EML
# physical/distribution/online/url over the open web; the `export` chunk wrote those
# URLs as the staged copies above. A package whose bytes were already evaluated is
# not evaluated again (PASTA would re-fetch gigabytes to say the same thing).
evaluate_reports <- list()
for (key in names(pkg_dirs)) {
  pkg_dir <- pkg_dirs[[key]]
  man_path <- file.path(pkg_dir, "manifest.json")
  pm <- jsonlite::fromJSON(man_path, simplifyVector = FALSE)
  if (identical(pm$evaluated_hash %||% "", pm$content_hash)) {
    cat(glue("- `{key}`: these bytes were evaluated {pm$evaluated_utc} — not re-evaluated\n")); next
  }
  eml_path <- file.path(pkg_dir, glue("{key}.xml"))
  tx <- tryCatch(
    EDIutils::evaluate_data_package(eml = eml_path, env = "staging"),
    error = function(e) { message(glue("evaluate failed for {key}: {conditionMessage(e)}")); NULL })
  if (is.null(tx)) next
  EDIutils::check_status_evaluate(tx, env = "staging")
  rpt <- tryCatch(EDIutils::read_evaluate_report_summary(tx, with_exceptions = FALSE, env = "staging"),
                  error = function(e) conditionMessage(e))
  pm$evaluated_utc  <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  pm$evaluated_hash <- pm$content_hash
  writeLines(as.character(rpt), file.path(pkg_dir, "evaluate_report.txt"))
  jsonlite::write_json(pm, man_path, auto_unbox = TRUE, pretty = TRUE, null = "null")
  evaluate_reports[[key]] <- list(transaction = tx, summary = rpt, evaluated_utc = pm$evaluated_utc)
  cat(glue("- `{key}`: evaluate transaction `{tx}` — report written to evaluate_report.txt\n"))
}
EDIutils::logout()
```

```{r}
#| label: evaluate-skip
#| eval: !expr NO_CREDS
cat("EDI_USER/EDI_PASS (or EDI_KEY) are not set — evaluate_data_package() was not run.\n",
   "Set one of those and re-render to evaluate against EDI's staging environment.\n")
```

## Publish — create or update the real EDI package (gated)

```{r}
#| label: publish
#| eval: !expr DO_PUBLISH_EDI

# Only reached when CALCOFI_PUBLISH_EDI=true AND (for the create/update call
# itself) EDI credentials are present — never run as part of this repo's own
# CI/staging checks, and never targeting anything but env = "production".
stopifnot("CALCOFI_PUBLISH_EDI=true requires EDI credentials" = creds$available)
librarian::shelf(EDIutils, quiet = TRUE)
if (identical(creds$method, "key")) EDIutils::login(key = Sys.getenv("EDI_KEY")) else
  EDIutils::login(userId = Sys.getenv("EDI_USER"), userPass = Sys.getenv("EDI_PASS"))

# only a package whose bytes differ from EDI's copy, and whose record is complete
due <- manifest_tbl |> filter(needs_upload, !nzchar(coalesce(eml_blocking, "")))
held <- manifest_tbl |> filter(needs_upload, nzchar(coalesce(eml_blocking, "")))
for (i in seq_len(nrow(held)))
  cat(glue("- `{held$dataset_key[i]}`: NOT deposited — the record is incomplete ({held$eml_blocking[i]})\n"))

publish_rows <- list()
for (key in due$dataset_key) {
  pkg_dir  <- pkg_dirs[[key]]
  eml_path <- file.path(pkg_dir, glue("{key}.xml"))
  existing <- edi_package_id_for(pkg_registry, key)
  uploaded_utc <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  tx <- if (is.na(existing))
    EDIutils::create_data_package(eml = eml_path, env = "production") else
    EDIutils::update_data_package(eml = eml_path, env = "production")
  ok <- if (is.na(existing)) EDIutils::check_status_create(tx, env = "production") else
    EDIutils::check_status_update(tx, env = "production")
  # EDIutils' create_data_package() example names its transaction
  # "create_<timestamp>__<scope.id.rev>" (?EDIutils::create_data_package); the
  # update case is not documented as explicitly, so this is a best-effort parse
  # — UNVERIFIED against a real transaction (this repo has never held EDI
  # credentials). Confirm the pattern on the first real run and, if it does not
  # match, read the package id back with EDIutils::list_data_package_identifiers()
  # / read_data_package_report_summary(tx) instead of trusting this regex.
  pkg_id <- sub("^(create|update)_[0-9]+__?", "", tx)
  row_i <- manifest_tbl[manifest_tbl$dataset_key == key, ]
  publish_rows[[key]] <- tibble(dataset_key = key, package_id = pkg_id, uploaded_utc = uploaded_utc,
                                ok = ok, content_hash = row_i$content_hash, built_from = row_i$version)
  cat(glue("- `{key}`: {if (is.na(existing)) 'create' else 'update'} -> `{pkg_id}` (status ok={ok})\n"))
}
EDIutils::logout()

if (length(publish_rows)) {
  new_rows <- bind_rows(publish_rows) |>
    filter(ok) |>
    mutate(scope = "edi", identifier = sub("^edi\\.([0-9]+)\\..*$", "\\1", package_id),
          revision = sub("^edi\\.[0-9]+\\.([0-9]+)$", "\\1", package_id), env = "production",
          created_utc = uploaded_utc, updated_utc = uploaded_utc) |>
    select(all_of(EDI_PACKAGES_COLS))
  # a revision keeps the package's first created_utc
  new_rows$created_utc <- coalesce(pkg_registry$created_utc[match(new_rows$dataset_key, pkg_registry$dataset_key)],
                                   new_rows$created_utc)
  pkg_registry <- bind_rows(pkg_registry |> filter(!dataset_key %in% new_rows$dataset_key), new_rows)
  write_csv(pkg_registry, PKG_REGISTRY_PATH, na = "")
}
```

```{r}
#| label: publish-skip
#| eval: !expr NO_PUBLISH_EDI
cat("CALCOFI_PUBLISH_EDI is not `true` — create_data_package()/update_data_package() were not run.\n",
   "Set CALCOFI_PUBLISH_EDI=true (with EDI credentials) to mint or revise a real EDI package.\n")
```

## Upload due

A package needs a fresh deposit when its `content_hash` is not the one
`metadata/edi_packages.csv` records for EDI's copy — never deposited, or changed
since. `eml_blocking` names what the record still lacks before it may be deposited.

```{r}
#| label: upload-due
due_tbl <- manifest_tbl |>
  mutate(staged = if (STAGING) NA_character_ else
           as.character(staged_url(dataset_key, version, ""))) |>
  select(dataset_key, built_from = version, checked_version, upload_status, needs_upload,
         eml_blocking, staged)
kable(due_tbl, caption = glue("EDI: {sum(due_tbl$needs_upload)} of {nrow(due_tbl)} package(s) due for a deposit"))
```

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
