---
title: "Publish CTD casts to CF NetCDF"
subtitle: "Two CF Discrete-Sampling-Geometry profile files — the thinned headline record and the full-resolution scans"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_ctd_cast_to_netcdf
  workflow_type: publish
  dependency:
    - release_database
  provider: calcofi
  dataset: ctd-cast
  output: data/netcdf/ctd-cast.nc
  workflow_url: https://calcofi.io/workflows/publish_ctd-cast_to-netcdf.html
editor_options:
  chunk_output_type: console
---

## Overview

Publish the CalCOFI CTD record as a **CF Discrete-Sampling-Geometry profile
dataset**, downstream of `release_database`.

This is the *simple* end of the publish range, and deliberately so — it is the
counterpart to
[`publish_ichthyo_to-netcdf`](https://calcofi.io/workflows/publish_ichthyo_to-netcdf.html),
which handles the deeply nested case. CTD needs no netCDF-4 groups:

| | ichthyo | **ctd-cast** |
|---|---|---|
| hierarchy | site → tow → net → occurrence → bin | **flat: cast → depth level** |
| CF status | CF where it applies; groups elsewhere | **fully CF (`featureType=Profile`)** |
| structure | netCDF-4 groups + `parent_index` | **contiguous ragged array (`rowSize`)** |

A CTD cast *is* a CF profile: one instance, a vertical series of depth levels,
each sensor its own variable. So this file is genuinely CF-compliant end to end —
no caveat needed — and any CF-aware tool will read it without special handling.

::: {.callout-note title="Why the widening still matters"}
The database stores CTD in long form: one row per cast × depth ×
`measurement_type`, 5.55 M rows over 15 sensor types. A single
`measurement_value` column therefore mixes °C, PSU, mL/L and volts. CF requires
one unit per variable, so each `measurement_type` becomes its own variable with
its own `units` and `long_name`. That widening is what turns a queryable table
into a self-documenting file.
:::

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(DBI, duckdb, dplyr, glue, ncdf4, readr, jsonlite, digest,
                 knitr, tibble, here, quiet = TRUE)
here <- here::here
source(here("libs/publish_netcdf.R"))

DATASET     <- "ctd-cast"
DATASET_KEY <- "calcofi_ctd-cast"
RELEASE     <- cc_release_version()      # PROMOTED release, never a serving tree
PQ          <- cc_release_parquet(RELEASE)
OUT_DIR     <- here("data/netcdf"); dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
OUT_NC      <- file.path(OUT_DIR, glue("{DATASET}.nc"))
PUBLISH     <- identical(Sys.getenv("CALCOFI_PUBLISH"), "true")

cat(glue("release : {RELEASE}\nparquet : {PQ}\noutput  : {OUT_NC}\npublish : {PUBLISH}\n"))
```

```{r}
#| label: connect
con <- dbConnect(duckdb())
for (s in c("INSTALL httpfs", "LOAD httpfs", "SET memory_limit='6GB'",
            "SET enable_progress_bar=false")) try(dbExecute(con, s), silent = TRUE)
q <- function(sql, ...) dbGetQuery(con, glue(sql, ..., .envir = parent.frame()))

dbExecute(con, glue("
  CREATE VIEW smp AS SELECT * FROM read_parquet('{PQ}/sample.parquet')
                     WHERE dataset_key = '{DATASET_KEY}';
  CREATE VIEW obs AS SELECT * FROM read_parquet('{PQ}/obs/dataset_key={DATASET_KEY}/data_0.parquet');"))
```

## Shape and integrity

```{r}
#| label: integrity
shape <- q("
  SELECT (SELECT count(*) FROM smp WHERE sample_type='cast')       AS casts_in_sample,
         (SELECT count(DISTINCT sample_key) FROM obs)              AS casts_with_obs,
         (SELECT count(*) FROM obs)                                AS obs_rows,
         (SELECT count(DISTINCT measurement_type) FROM obs)        AS sensor_types")
kable(shape)

# A profile must have a well-defined depth axis: one value per (cast, depth,
# measurement_type). Duplicates there would silently collapse in the pivot below,
# so check rather than assume.
dupes <- q("
  SELECT count(*) AS n FROM (
    SELECT sample_key, depth_min_m, measurement_type
    FROM obs GROUP BY 1,2,3 HAVING count(*) > 1)")$n
cat(glue("duplicate (cast, depth, measurement_type) triples: {dupes}\n"))
```

```{r}
#| label: pivot
# WIDEN: measurement_type -> one variable each, ordered by (cast, depth) so the
# profiles are contiguous — which is what a CF contiguous ragged array requires.
types <- q("SELECT DISTINCT measurement_type FROM obs ORDER BY 1")$measurement_type
sel <- paste(sprintf(
  "MAX(measurement_value) FILTER (WHERE measurement_type='%s')::DOUBLE AS \"%s\"",
  types, types), collapse = ",\n    ")

prof <- q("
  SELECT o.sample_key                       AS profile_id,
         any_value(s.cruise_key)            AS cruise_key,
         any_value(s.grid_key)              AS grid_key,
         epoch(any_value(s.datetime))::DOUBLE AS time,
         any_value(s.latitude)::DOUBLE      AS latitude,
         any_value(s.longitude)::DOUBLE     AS longitude,
         o.depth_min_m::DOUBLE              AS depth,
         {sel}
  FROM obs o LEFT JOIN smp s ON s.sample_key = o.sample_key
  GROUP BY o.sample_key, o.depth_min_m
  ORDER BY o.sample_key, o.depth_min_m")

cat(glue("wide rows (cast x depth): {nrow(prof)}\n"))
cat(glue("sensor variables        : {length(types)}\n"))
```

## Write the CF profile file

```{r}
#| label: write-nc
mt_meta <- cc_measurement_meta(here())

prof_ids <- unique(prof$profile_id)
row_size <- as.integer(table(factor(prof$profile_id, levels = prof_ids)))
first_ix <- match(prof_ids, prof$profile_id)
stopifnot(sum(row_size) == nrow(prof))

STRLEN <- 64L
FILL   <- 9.969209968386869e36
d_prof <- ncdim_def("profile", "", seq_along(prof_ids), create_dimvar = FALSE)
d_obs  <- ncdim_def("obs",     "", seq_len(nrow(prof)), create_dimvar = FALSE)
d_str  <- ncdim_def("name_strlen", "", seq_len(STRLEN), create_dimvar = FALSE)

v_pid  <- ncvar_def("profile_id", "", list(d_str, d_prof), prec = "char")
v_cru  <- ncvar_def("cruise_key", "", list(d_str, d_prof), prec = "char")
v_grd  <- ncvar_def("grid_key",   "", list(d_str, d_prof), prec = "char")
v_time <- ncvar_def("time", "seconds since 1970-01-01T00:00:00Z", d_prof, prec = "double")
v_lat  <- ncvar_def("latitude",  "degrees_north", d_prof, prec = "double")
v_lon  <- ncvar_def("longitude", "degrees_east",  d_prof, prec = "double")
v_rs   <- ncvar_def("rowSize",   "",              d_prof, prec = "integer")
v_dep  <- ncvar_def("depth", "m", d_obs, prec = "double", missval = FILL)
v_sens <- lapply(types, function(v) ncvar_def(
  v, as.character(mt_meta[[v]]$units %||% ""), d_obs, prec = "double", missval = FILL))

unlink(OUT_NC)
nc <- nc_create(OUT_NC, c(list(v_pid, v_cru, v_grd, v_time, v_lat, v_lon, v_rs, v_dep),
                          v_sens), force_v4 = TRUE)
ncvar_put(nc, v_pid,  prof_ids)
ncvar_put(nc, v_cru,  prof$cruise_key[first_ix])
ncvar_put(nc, v_grd,  prof$grid_key[first_ix])
ncvar_put(nc, v_time, prof$time[first_ix])
ncvar_put(nc, v_lat,  prof$latitude[first_ix])
ncvar_put(nc, v_lon,  prof$longitude[first_ix])
ncvar_put(nc, v_rs,   row_size)
ncvar_put(nc, v_dep,  prof$depth)
for (i in seq_along(types)) ncvar_put(nc, v_sens[[i]], prof[[types[i]]])

# CF DSG attributes — these are what make it a profile dataset rather than a
# table that happens to be in netCDF.
ncatt_put(nc, "profile_id", "cf_role", "profile_id")
ncatt_put(nc, "profile_id", "long_name", "CTD cast identifier")
ncatt_put(nc, "time", "standard_name", "time");      ncatt_put(nc, "time", "axis", "T")
ncatt_put(nc, "latitude",  "standard_name", "latitude");  ncatt_put(nc, "latitude", "axis", "Y")
ncatt_put(nc, "longitude", "standard_name", "longitude"); ncatt_put(nc, "longitude", "axis", "X")
ncatt_put(nc, "depth", "standard_name", "depth"); ncatt_put(nc, "depth", "axis", "Z")
ncatt_put(nc, "depth", "positive", "down")
ncatt_put(nc, "rowSize", "long_name", "Number of observations for this profile")
ncatt_put(nc, "rowSize", "sample_dimension", "obs")   # -> contiguous ragged array
for (v in types) {
  ncatt_put(nc, v, "long_name",
            as.character(mt_meta[[v]]$long_name %||% gsub("_", " ", v)))
  ncatt_put(nc, v, "coordinates", "time latitude longitude depth")
  sn <- mt_meta[[v]]$standard_name %||% NA_character_
  if (!is.na(sn) && nzchar(sn)) ncatt_put(nc, v, "standard_name", sn)
}
globals <- list(
  title = "CalCOFI CTD casts — CF profile dataset",
  summary = paste(
    "CalCOFI CTD profiles as CF Discrete Sampling Geometry: one profile per cast,",
    "depth levels stored as a contiguous ragged array, each canonical sensor its",
    "own variable with its own units. Widened from the normalized long form in the",
    glue("CalCOFI integrated database release {RELEASE}.")),
  featureType = "profile", cdm_data_type = "Profile",
  cdm_profile_variables = "profile_id, time, latitude, longitude, cruise_key, grid_key",
  Conventions = "CF-1.10, COARDS, ACDD-1.3",
  cf_scope = paste("Fully CF: a CTD cast is a CF profile, so this dataset needs no",
                   "extension beyond the standard (contrast the nested net-tow datasets)."),
  institution = "CalCOFI", source = glue("CalCOFI integrated database release {RELEASE}"),
  db_release = RELEASE, dataset_key = DATASET_KEY,
  references = "https://calcofi.io/workflows/publish_ctd-cast_to-netcdf.html",
  license = "CC-BY 4.0", creator_name = "CalCOFI", creator_url = "https://calcofi.io",
  date_created = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"))
for (nm in names(globals)) ncatt_put(nc, 0, nm, globals[[nm]])
nc_close(nc)

cat(glue("wrote {OUT_NC} ({round(file.size(OUT_NC)/1048576, 1)} MB)\n"))
```

## Verify

```{r}
#| label: verify
ncv <- nc_open(OUT_NC)
rs  <- ncvar_get(ncv, "rowSize")
chk <- tibble(
  check = c("featureType", "cf_role set", "profiles", "obs levels",
            "rowSize sums to obs", "sensor variables", "depths per profile (min/med/max)"),
  value = c(
    ncatt_get(ncv, 0, "featureType")$value,
    ncatt_get(ncv, "profile_id", "cf_role")$value,
    as.character(ncv$dim[["profile"]]$len),
    as.character(ncv$dim[["obs"]]$len),
    as.character(sum(rs) == ncv$dim[["obs"]]$len),
    as.character(length(types)),
    paste(min(rs), median(rs), max(rs), sep = " / ")))
nc_close(ncv)
kable(chk)
```


## Full-resolution companion — `ctd-cast_full.nc`

The file above is the **adaptively-thinned** record (5.55 M long rows → 434 k depth
levels, 7,175 profiles). `obs_ctd_full` is ~216 M long rows across 96 cruise
partitions and 32 sensor types, flagged *supplemental* in the release precisely
because of its size.

It is a **superset in two dimensions, not one**: every depth scan rather than a
thinned subset, *and* all **14,336** casts — both the down- and up-cast of each
station occupation — where the thinned record keeps one direction per occupation.
So the profile count roughly doubles for a reason unrelated to resolution. Both
files state this in their own global attributes, because "full resolution" alone
would imply the same profiles sampled more densely, which is not what this is.

Two things make this build different, and both are consequences of scale:

::: {.callout-important title="Why this is chunked rather than pivoted in one query"}
This is the table that **OOM'd ERDDAP at every container size tested** — 4, 5 and
6 GB (see [the serving benchmark](https://calcofi.io/workflows/bench_erddap_ctd.html)).
Materializing it here would hit the same wall, so it is processed **one cruise
partition at a time** and written into the file incrementally with
`ncvar_put(start=, count=)`. Peak memory is one cruise (~59 k levels × 32
variables ≈ 15 MB), not the whole table.

Also, `read_parquet('.../obs_ctd_full/**/*.parquet')` **404s over HTTPS**:
expanding a glob needs a directory listing and object storage has none. Partitions
are therefore enumerated from the XML listing API (`cc_release_partitions()`) and
passed as an explicit URL vector.
:::

```{r}
#| label: full-partitions
parts <- cc_release_partitions("obs_ctd_full", RELEASE)
cat(glue("partitions: {length(parts)}\n"))

# obs_ctd_full already carries latitude/longitude/datetime, so unlike the thinned
# path no join back to `sample` is needed.
full_types <- dbGetQuery(con, glue(
  "SELECT DISTINCT measurement_type FROM read_parquet('{parts[1]}') ORDER BY 1"))$measurement_type
cat(glue("sensor types: {length(full_types)}\n"))
```

```{r}
#| label: full-pass1
# PASS 1 — size the dimensions. Cheap: scans two columns per partition rather
# than pivoting. netCDF needs dimension lengths at creation time, and a wrong
# guess means rewriting a multi-GB file.
pass1 <- lapply(seq_along(parts), function(i) {
  r <- dbGetQuery(con, glue("
    SELECT count(*) AS n_lev, count(DISTINCT sample_key) AS n_prof FROM (
      SELECT DISTINCT sample_key, depth_min_m FROM read_parquet('{parts[i]}'))"))
  if (i %% 24 == 0) message(glue("  pass1 {i}/{length(parts)}"))
  r
})
p1 <- do.call(rbind, pass1)
N_PROF <- sum(p1$n_prof); N_LEV <- sum(p1$n_lev)
tibble(partitions = length(parts), profiles = N_PROF, depth_levels = N_LEV,
       `mean levels/profile` = round(N_LEV / N_PROF)) |> kable()
```

```{r}
#| label: full-write
FULL_NC <- file.path(OUT_DIR, "ctd-cast_full.nc")
sel_full <- paste(sprintf(
  "MAX(measurement_value) FILTER (WHERE measurement_type='%s')::DOUBLE AS \"%s\"",
  full_types, full_types), collapse = ",\n    ")

d_prof_f <- ncdim_def("profile", "", seq_len(N_PROF), create_dimvar = FALSE)
d_obs_f  <- ncdim_def("obs",     "", seq_len(N_LEV),  create_dimvar = FALSE)
d_str_f  <- ncdim_def("name_strlen", "", seq_len(STRLEN), create_dimvar = FALSE)

vf_pid  <- ncvar_def("profile_id", "", list(d_str_f, d_prof_f), prec = "char")
vf_cru  <- ncvar_def("cruise_key", "", list(d_str_f, d_prof_f), prec = "char")
vf_time <- ncvar_def("time", "seconds since 1970-01-01T00:00:00Z", d_prof_f, prec = "double")
vf_lat  <- ncvar_def("latitude",  "degrees_north", d_prof_f, prec = "double")
vf_lon  <- ncvar_def("longitude", "degrees_east",  d_prof_f, prec = "double")
vf_rs   <- ncvar_def("rowSize",   "",              d_prof_f, prec = "integer")
vf_dep  <- ncvar_def("depth", "m", d_obs_f, prec = "double", missval = FILL)
vf_sens <- lapply(full_types, function(v) ncvar_def(
  v, as.character(mt_meta[[v]]$units %||% ""), d_obs_f, prec = "double", missval = FILL))

unlink(FULL_NC)
ncf <- nc_create(FULL_NC, c(list(vf_pid, vf_cru, vf_time, vf_lat, vf_lon, vf_rs, vf_dep),
                            vf_sens), force_v4 = TRUE)

# PASS 2 — one cruise at a time, appended at the running offsets. Nothing larger
# than a single cruise is ever held in memory.
off_p <- 1L; off_o <- 1L
for (i in seq_along(parts)) {
  w <- dbGetQuery(con, glue("
    SELECT sample_key AS profile_id,
           any_value(cruise_key)              AS cruise_key,
           epoch(any_value(datetime))::DOUBLE AS time,
           any_value(latitude)::DOUBLE        AS latitude,
           any_value(longitude)::DOUBLE       AS longitude,
           depth_min_m::DOUBLE                AS depth,
           {sel_full}
    FROM read_parquet('{parts[i]}')
    GROUP BY sample_key, depth_min_m
    ORDER BY sample_key, depth_min_m"))
  if (!nrow(w)) next
  ids <- unique(w$profile_id)
  rs  <- as.integer(table(factor(w$profile_id, levels = ids)))
  fx  <- match(ids, w$profile_id)

  ncvar_put(ncf, vf_pid,  ids,               start = c(1, off_p), count = c(STRLEN, length(ids)))
  ncvar_put(ncf, vf_cru,  w$cruise_key[fx],  start = c(1, off_p), count = c(STRLEN, length(ids)))
  ncvar_put(ncf, vf_time, w$time[fx],        start = off_p, count = length(ids))
  ncvar_put(ncf, vf_lat,  w$latitude[fx],    start = off_p, count = length(ids))
  ncvar_put(ncf, vf_lon,  w$longitude[fx],   start = off_p, count = length(ids))
  ncvar_put(ncf, vf_rs,   rs,                start = off_p, count = length(ids))
  ncvar_put(ncf, vf_dep,  w$depth,           start = off_o, count = nrow(w))
  for (k in seq_along(full_types))
    ncvar_put(ncf, vf_sens[[k]], w[[full_types[k]]], start = off_o, count = nrow(w))

  off_p <- off_p + length(ids); off_o <- off_o + nrow(w)
  if (i %% 12 == 0) message(glue("  wrote {i}/{length(parts)} — {off_p-1} profiles, {off_o-1} levels"))
  rm(w); invisible(gc(FALSE))
}
# the pass-1 sizing must match exactly, or the file has trailing fill values that
# look like real (missing) data
stopifnot(off_p - 1L == N_PROF, off_o - 1L == N_LEV)

ncatt_put(ncf, "profile_id", "cf_role", "profile_id")
ncatt_put(ncf, "profile_id", "long_name", "CTD cast identifier")
ncatt_put(ncf, "time", "standard_name", "time");     ncatt_put(ncf, "time", "axis", "T")
ncatt_put(ncf, "latitude",  "standard_name", "latitude");  ncatt_put(ncf, "latitude", "axis", "Y")
ncatt_put(ncf, "longitude", "standard_name", "longitude"); ncatt_put(ncf, "longitude", "axis", "X")
ncatt_put(ncf, "depth", "standard_name", "depth"); ncatt_put(ncf, "depth", "axis", "Z")
ncatt_put(ncf, "depth", "positive", "down")
ncatt_put(ncf, "rowSize", "long_name", "Number of observations for this profile")
ncatt_put(ncf, "rowSize", "sample_dimension", "obs")
for (v in full_types) {
  ncatt_put(ncf, v, "long_name", as.character(mt_meta[[v]]$long_name %||% gsub("_", " ", v)))
  ncatt_put(ncf, v, "coordinates", "time latitude longitude depth")
}
globals_full <- globals
globals_full$title <- "CalCOFI CTD casts, full resolution — CF profile dataset"
globals_full$summary <- paste(
  "Full-resolution CalCOFI CTD scans as CF Discrete Sampling Geometry: one profile",
  "per cast, every recorded depth level, each sensor its own variable.",
  "NOTE this file differs from the thinned companion (ctd-cast.nc) in TWO ways, not",
  "one: it keeps every depth scan rather than an adaptively-thinned subset, AND it",
  "covers all 14,336 casts including both the down- and up-cast of each station",
  "occupation, where the thinned record keeps a single direction (7,175). It is",
  "therefore a superset, not a higher-resolution rendering of the same profiles.",
  "Most analyses want the thinned file; use this one when individual scans or the",
  "second cast direction matter.")
globals_full$comment <- paste(
  "profiles here (14,336) are cast-direction resolved; the thinned companion",
  "reports 7,175 because it retains one direction per occupation.")
globals_full$source_table <- "obs_ctd_full (supplemental)"
for (nm in names(globals_full)) ncatt_put(ncf, 0, nm, globals_full[[nm]])
nc_close(ncf)
cat(glue("wrote {FULL_NC} ({round(file.size(FULL_NC)/1048576, 1)} MB)\n"))
```

```{r}
#| label: full-verify
ncv <- nc_open(FULL_NC)
rs  <- ncvar_get(ncv, "rowSize")
tibble(
  check = c("featureType", "profiles", "depth levels", "rowSize sums to obs",
            "sensor variables", "levels per profile (min/med/max)"),
  value = c(ncatt_get(ncv, 0, "featureType")$value,
            as.character(ncv$dim[["profile"]]$len),
            as.character(ncv$dim[["obs"]]$len),
            as.character(sum(rs) == ncv$dim[["obs"]]$len),
            as.character(length(full_types)),
            paste(min(rs), median(rs), max(rs), sep = " / "))) |> kable()
nc_close(ncv)
```

## Publish

```{r}
#| label: publish
#| eval: !expr PUBLISH
# Two products, published as SEPARATE datasets: they are the same casts at
# different resolutions, and a user choosing between them should see two entries
# with their own sizes and provenance rather than one entry with a hidden variant.
publish_one <- function(nc_path, ds, src_tables, scope) {
  plan <- cc_netcdf_plan(nc_path, ds, RELEASE)
  man  <- cc_netcdf_manifest(plan, ds, RELEASE, RELEASE, file.size(nc_path),
                             source_tables = src_tables, cf_scope = scope)
  cat(glue("{ds}: sha256 {substr(plan$sha256,1,12)}… identical_to={plan$identical_to} upload={plan$upload}\n"))
  res <- cc_netcdf_publish(nc_path, ds, RELEASE, plan, man)
  cat(glue("  -> {res$url}\n"))
  invisible(res)
}
publish_one(OUT_NC,  DATASET,           c("sample", "obs"),  globals$cf_scope)
publish_one(FULL_NC, "ctd-cast_full",   "obs_ctd_full",      globals_full$cf_scope)
```

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
