---
title: "Portal status — what needs a fresh upload"
subtitle: "Every publisher's output against what each portal holds, after every release"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_status
  workflow_type: publish
  dependency:
    - publish_to_edi
    - publish_to_obis
    - publish_to_netcdf
    - publish_to_erddap
  output: data/publish/portal_status.csv
  workflow_url: https://calcofi.io/workflows/publish_status.html
  description: >
    The caboose of the publishers: after publish_to-edi, -obis, -netcdf and
    -erddap have run against a promoted release, compares each dataset's built
    package, archive or file with the copy its portal holds, and lists what
    needs a fresh upload — never deposited, or changed since the last deposit —
    with what still blocks it and how to do it.
editor_options:
  chunk_output_type: console
---

## Why this page exists

The four publishers run after every promoted release, and each rebuilds a dataset's
output only when that dataset's rows, metadata or the publisher's code changed
(`calcofi4db::publish_fingerprint()`). Two of the four portals then take a deliberate,
manual step — an EDI deposit and an OBIS-USA IPT upload — and the other two publish
from the pipeline when a flag says so. Whether a portal is behind is therefore a
comparison, not a memory: the bytes built now against the bytes the portal was last
given (`calcofi4db::publish_upload_status()`). This page makes that comparison for every
dataset on every portal, and its first table is the to-do list.

| portal | the copy it holds is recorded in | an upload is |
|---|---|---|
| EDI | `metadata/edi_packages.csv` (`content_hash`, `built_from`) | `CALCOFI_PUBLISH_EDI=true` + EDI credentials, re-rendering `publish_to-edi.qmd`; refused while the record lacks a required field |
| OBIS (via the OBIS-USA IPT) | each archive's `{dataset_key}_manifest.json` (`uploaded_hash`) | manual, after the provider agrees (Decision 21) — [`docs/portals.qmd` § OBIS](https://calcofi.io/docs/portals.html#obis) |
| netCDF (`calcofi-files-public/netcdf/`) | the site's `{dataset}/manifests.json` | `CALCOFI_PUBLISH=true`, re-rendering `publish_to-netcdf.qmd`; unchanged bytes are never re-uploaded |
| ERDDAP (`erddap.calcofi.io`) | `data/erddap/deploy_status.json` | automatic unless `CALCOFI_DEPLOY=false` / `CALCOFI_ERDDAP_DEPLOY=false` |

## Setup

```{r}
#| label: setup
#| message: false
librarian::shelf(dplyr, glue, jsonlite, knitr, purrr, readr, tibble, here, quiet = TRUE)
here <- here::here
options(readr.show_col_types = FALSE)
devtools::load_all(here::here("../calcofi4db"))

RELEASE_PREFIX <- Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake/releases")
STAGING        <- grepl("staging", RELEASE_PREFIX, fixed = TRUE)
BASE_HTTPS     <- "https://storage.googleapis.com/calcofi-db"
RELEASE <- Sys.getenv("CALCOFI_RELEASE_VERSION", "")
if (!nzchar(RELEASE))
  RELEASE <- trimws(readLines(glue("{BASE_HTTPS}/{RELEASE_PREFIX}/latest.txt"), warn = FALSE)[1])
NETCDF_SITE <- "https://storage.calcofi.io/calcofi-files-public/netcdf"
OUT <- here("data/publish/portal_status.csv")
dir.create(dirname(OUT), recursive = TRUE, showWarnings = FALSE)

read_json_or_null <- function(p) tryCatch(jsonlite::fromJSON(p, simplifyVector = FALSE),
                                          error = function(e) NULL)
cat(glue("release : {RELEASE}{if (STAGING) ' (STAGING)' else ''}"), "\n")
```

## EDI

```{r}
#| label: edi
edi_csv <- here(if (STAGING) "data/edi-staging" else "data/edi", "manifest.csv")
edi <- if (file.exists(edi_csv)) read_csv(edi_csv, col_types = cols(.default = "c"), na = "") else NULL
edi_rows <- if (is.null(edi) || !nrow(edi)) tibble() else edi |>
  transmute(
    portal = "edi", dataset_key, built_from = version, checked_version,
    upload_status, needs_upload = as.logical(needs_upload),
    blocked_by = if_else(nzchar(coalesce(eml_blocking, "")),
                         paste("record incomplete:", eml_blocking), ""),
    how  = "CALCOFI_PUBLISH_EDI=true + EDI credentials, re-render publish_to-edi.qmd",
    link = glue("{BASE_HTTPS}/publish/edi/{dataset_key}/{dataset_key}_{version}/"))
if (nrow(edi_rows)) kable(edi_rows |> select(dataset_key, built_from, upload_status, blocked_by)) else
  cat("no EDI manifest yet\n")
```

## OBIS

```{r}
#| label: obis
obis_dir <- here(if (STAGING) "data/darwincore-staging" else "data/darwincore")
mans <- Sys.glob(file.path(obis_dir, "*_manifest.json"))
obis_rows <- map_dfr(mans, function(p) {
  m <- read_json_or_null(p)
  if (is.null(m)) return(NULL)
  st <- publish_upload_status(m$content_hash %||% NA_character_, m$uploaded_hash %||% NA_character_)
  # a resource can be on OBIS from before manifests stamped `uploaded_hash`
  # (swfsc_ichthyo, 2026-04): it is not "never uploaded", its copy is just unverifiable
  on_portal <- !is.null(m$ipt_resource) || !is.null(m$obis_dataset_id)
  if (on_portal && is.null(m$uploaded_hash))
    st$upload_status <- "on OBIS; the uploaded copy's hash was never recorded"
  tibble(
    portal = "obis", dataset_key = m$dataset_key, built_from = m$version,
    checked_version = m$checked_version %||% m$version,
    upload_status = st$upload_status, needs_upload = st$needs_upload,
    blocked_by = if (!on_portal) "provider agreement (Decision 21); no IPT resource yet" else
      if (is.null(m$uploaded_hash)) "stamp uploaded_utc + uploaded_hash in the manifest at the next upload" else "",
    how  = "manual OBIS-USA IPT upload — docs/portals.qmd § OBIS",
    link = glue("{BASE_HTTPS}/publish/dwca/{m$dataset_key}/{m$archive}"))
})
# a candidate that failed its checks has no manifest, and is listed rather than dropped
plan_csv <- file.path(obis_dir, "publish_plan.csv")
if (file.exists(plan_csv)) {
  cand <- read_csv(plan_csv, col_types = cols(.default = "c"))$dataset_key
  missing <- setdiff(cand, obis_rows$dataset_key)
  if (length(missing)) obis_rows <- bind_rows(obis_rows, tibble(
    portal = "obis", dataset_key = missing, built_from = NA, checked_version = NA,
    upload_status = "not built", needs_upload = FALSE,
    blocked_by = "no archive: failed dwc_check() — see publish_to-obis.html",
    how = "", link = ""))
}
if (nrow(obis_rows)) kable(obis_rows |> select(dataset_key, built_from, upload_status, blocked_by)) else
  cat("no Darwin Core archives built yet\n")
```

## netCDF

A file counts as published for this release when the site's `manifests.json` lists
the release, and its bytes are online when any listed release carries the local
file's sha256 — an unchanged file is never uploaded twice, only listed.

```{r}
#| label: netcdf
sides <- Sys.glob(here("data/netcdf/*.fingerprint.json"))
nc_rows <- map_dfr(sides, function(p) {
  s  <- read_json_or_null(p)
  ds <- sub("\\.fingerprint\\.json$", "", basename(p))
  nc <- sub("\\.fingerprint\\.json$", ".nc", p)
  if (is.null(s) || !file.exists(nc)) return(NULL)
  rel  <- read_json_or_null(glue("{NETCDF_SITE}/{ds}/manifests.json"))$releases %||% list()
  sha  <- digest::digest(nc, algo = "sha256", file = TRUE)
  online <- any(vapply(rel, function(r) identical(r$sha256, sha), logical(1)))
  listed <- any(vapply(rel, function(r) identical(r$version, RELEASE), logical(1)))
  status <- if (online && listed) "current" else if (online) "bytes online; release not listed" else
    "not online"
  tibble(portal = "netcdf", dataset_key = ds, built_from = s$built_from, checked_version = RELEASE,
         upload_status = status, needs_upload = !(online && listed), blocked_by = "",
         how  = "CALCOFI_PUBLISH=true, re-render publish_to-netcdf.qmd (+ scripts/build_netcdf_index.R)",
         link = glue("{NETCDF_SITE}/{ds}/"))
})
if (nrow(nc_rows)) kable(nc_rows |> select(dataset_key, built_from, upload_status)) else
  cat("no netCDF fingerprints yet — publish_to-netcdf.qmd has not run since change detection\n")
```

## ERDDAP

```{r}
#| label: erddap
es <- read_json_or_null(here("data/erddap/deploy_status.json"))
erddap_rows <- if (is.null(es)) tibble() else {
  served <- es$deployed_release %||% NA_character_
  tibble(portal = "erddap", dataset_key = "(all)", built_from = RELEASE, checked_version = RELEASE,
         upload_status = if (identical(served, RELEASE)) "current" else
           glue("serves {served %||% 'an unknown release'}"),
         needs_upload = !identical(served, RELEASE), blocked_by = "",
         how  = "re-render publish_to-erddap.qmd without CALCOFI_DEPLOY=false",
         link = "https://erddap.calcofi.io/erddap/")
}
if (nrow(erddap_rows)) kable(erddap_rows |> select(upload_status, how)) else
  cat("no ERDDAP deploy status yet — publish_to-erddap.qmd has not run since it began recording one\n")
```

## What needs a fresh upload

```{r}
#| label: status
status <- bind_rows(edi_rows, obis_rows, nc_rows, erddap_rows) |>
  mutate(release = RELEASE, .before = 1) |>
  arrange(desc(needs_upload), portal, dataset_key)
write_csv(status, OUT, na = "")
if (!STAGING) put_gcs_file(OUT, "gs://calcofi-db/publish/portal_status.csv")

due <- status |> filter(needs_upload)
cat(glue("{nrow(due)} of {nrow(status)} portal copies need a fresh upload for {RELEASE}",
         "{if (nrow(due)) paste0(' (', paste(sort(unique(due$portal)), collapse = ', '), ')') else ''}"), "\n")
if (nrow(due))
  kable(due |> select(portal, dataset_key, built_from, upload_status, blocked_by, how),
        caption = "due: never deposited, or changed since the last deposit")
```

```{r}
#| label: status-all
kable(status |> select(portal, dataset_key, built_from, checked_version, upload_status, blocked_by),
      caption = glue("every portal copy, {RELEASE} (data/publish/portal_status.csv)"))
```
