---
title: "Publish every biological dataset to OBIS"
subtitle: "One Darwin Core Archive per dataset, generated from the core and the registries"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_to_obis
  workflow_type: publish
  dependency:
    - release_database
  output: data/darwincore/publish_plan.csv
  workflow_url: https://calcofi.io/workflows/publish_to-obis.html
  description: >
    Builds one Darwin Core Archive (Event core + Occurrence + eMoF + meta.xml +
    the release's own eml.xml) per biological dataset in the frozen release,
    generically from the core and the metadata registries. Replaces
    publish_ichthyo_to-obis.qmd, which read the swfsc_ichthyo source tables
    directly and left nine other biological datasets with no OBIS route. Writes
    files only: the upload to the OBIS-USA IPT stays a deliberate manual step.
execute:
  echo: true
  warning: false
editor_options:
  chunk_output_type: console
---

## Overview

Build one **Darwin Core Archive** per biological dataset in the frozen release —
`event.csv` (Event core) + `occurrence.csv` + `extendedMeasurementOrFact.csv` +
`meta.xml` + the release's own `eml/{dataset_key}.xml` — validate it, and write it
to `data/darwincore/{dataset_key}/{dataset_key}_{version}.zip` with a manifest.

This notebook **writes files only**. The upload to the OBIS-USA IPT is a
deliberate manual act with Ben's login, gated on a content change (Decision 10;
`docs/portals.qmd` § OBIS has the steps). Nothing here talks to a portal except
the read-only duplicate check below.

### Why it is generic now

`publish_ichthyo_to-obis.qmd` was the last per-dataset publisher. It read the
`swfsc_ichthyo` **source** tables (`ichthyo`, `net`, `tow`, `site`, `species`,
`lookup`) and hand-built the Event / Occurrence / eMoF triple, which is why nine
other biological datasets had no OBIS route at all — and why, since the core
consolidation retired those tables from the release, it can no longer run.

Everything it hand-built now exists generically, and the vocabulary ids it was
missing live in the registries the release already publishes:

| Darwin Core | from the core | vocabulary |
|---|---|---|
| **Event core** `eventID` · `parentEventID` · `eventType` · `eventDate` · `decimalLatitude/Longitude` · `minimum`/`maximumDepthInMeters` · `locationID` · `samplingProtocol` · `sampleSizeValue`/`Unit` · `geodeticDatum` · `datasetID` | `sample.sample_key` · `sample.parent_sample_key` (the row's `cruise_key` for a root) · `sample.sample_type` · `sample.datetime` · `sample.latitude`/`longitude` · `sample.depth_min_m`/`depth_max_m` · `sample.site_key` · `sample.tow_type` · `sample_measurement.volume_sampled` · the release CRS · `dataset_key` | `gear.csv` `dwc_samplingProtocol` (+ NERC **L22**) |
| **Occurrence** `occurrenceID` · `eventID` · `basisOfRecord` · `occurrenceStatus` · `scientificName` · `scientificNameID` · `taxonID` · `taxonRank` · `kingdom`…`family` · `vernacularName` · `lifeStage` · `individualCount` · `organismQuantity` + `organismQuantityType` · `occurrenceRemarks` | md5 of `obs_bio`'s natural grain · `obs_bio.sample_key` · `HumanObservation` · `obs_bio.value` · `taxon.scientific_name` · WoRMS LSID of `taxon.worms_id` · `obs_bio.taxon_key` · `taxon.rank` · `taxon` lineage · `taxon.common_name` · `obs_bio.life_stage` · `obs_bio.value` where the unit is a count · `density_per_10m2` / `density_per_1000m3` | `life_stage.csv` `dwc_lifeStage` (+ NERC **S11**) |
| **eMoF** `measurementType` · `measurementTypeID` · `measurementValue` · `measurementUnit` · `measurementUnitID` · `measurementRemarks` | `sample_measurement` (event grain) · `obs_attribute` (occurrence grain) · `obs_env` on the dataset's own events | `measurement_type.csv` `nerc_p01` / `units_nerc_p06` — **empty where no exact concept exists, never invented** |
| `meta.xml` | generated from the term map (`calcofi4db::dwc_term_map()`) | — |
| `eml.xml` | the release's `eml/{dataset_key}.xml` (`build_eml()`, D-8) | — |

The logic lives in **`calcofi4db::dwc_*()`** (`R/dwc.R`), not in this notebook, so
`devtools::test()` asserts the exact DwC rows against a synthetic core and this
file cannot drift from what the tests pin.

### `occurrenceStatus` is emitted honestly

::: {.callout-important title="An absence is a claim about a protocol, not about a table"}
`calcofi4db::dwc_absence_rule()` measures which rule each dataset falls under:

* **`zeros_recorded`** — the dataset has zero-valued `obs_bio` rows, so a sample
  examined and found empty of a taxon is already in the release. Those rows become
  `occurrenceStatus = "absent"`, and nothing is derived.
* **`positive_only`** — the dataset has **no** zero rows: a surveyed-empty sample
  simply has no row. Every row it does have is `present`, and an absence can only be
  *derived*, from `sample_root` minus the positives — which is true only if the
  protocol sorted every sample for the dataset's whole vocabulary.

That last claim is about the protocol, not about the data, so deriving absences is
never the default. For `swfsc_ichthyo` it would be plainly false: the ichthyoplankton
protocol identifies each specimen to the lowest taxon possible, so 963 observed taxa
× 61,104 sorted stations is 58 M absences nobody ever asserted. `dwc_occurrence()`
refuses to emit them without `absences = "sample_root"` and a `max_absences` the
caller raised on purpose.

The two positive-only datasets where the claim may genuinely hold —
`cce-lter_euphausiids` (BTEDB stages 37 species in every sample) and
`sio_mesopelagic-fish` (90 taxa, 102 tows) — are **questions for their providers**,
not something this notebook decides. Until a provider says yes, they publish
presence-only, which is what OBIS assumes of an archive with no `absent` rows.
:::

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(DBI, duckdb, dplyr, glue, jsonlite, knitr, readr, tibble,
                 here, curl, quiet = TRUE)
here <- here::here
devtools::load_all(here::here("../calcofi4db"))

# a STAGING run (CALCOFI_RELEASE_PREFIX=ducklake-staging/releases) reads the
# staging release and writes under data/darwincore-staging/, so it can never be
# mistaken for, or overwrite, what was published from the promoted one
RELEASE_PREFIX <- Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake/releases")
STAGING        <- grepl("staging", RELEASE_PREFIX, fixed = TRUE)
BASE_HTTPS     <- "https://storage.googleapis.com/calcofi-db"
RELEASES_URL   <- glue("{BASE_HTTPS}/{RELEASE_PREFIX}")

RELEASE <- Sys.getenv("CALCOFI_RELEASE_VERSION", "")
if (!nzchar(RELEASE))
  RELEASE <- trimws(readLines(glue("{RELEASES_URL}/latest.txt"), warn = FALSE)[1])

OUT_DIR <- here(if (STAGING) "data/darwincore-staging" else "data/darwincore")
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# comma-separated dataset_keys to restrict a run while iterating; empty = all
ONLY <- Filter(nzchar, trimws(strsplit(Sys.getenv("CALCOFI_DATASETS", ""), ",")[[1]]))
# the duplicate check asks api.obis.org; off-line it is skipped, never faked
NET  <- !identical(Sys.getenv("CALCOFI_OFFLINE"), "true")

cat(glue("release  : {RELEASE} (prefix {RELEASE_PREFIX})"), "\n")
cat(glue("output   : {OUT_DIR}"), "\n")
cat(glue("network  : {NET} (set CALCOFI_OFFLINE=true to skip the OBIS duplicate check)"), "\n")
if (length(ONLY)) cat(glue("restricted to: {paste(ONLY, collapse = ', ')}"), "\n")
```

```{r}
#| label: connect
# the release catalog resolves every table to its content-addressed objects —
# never concatenate releases/{v}/parquet/… by hand (CLAUDE.md § content-addressed)
rel_catalog <- jsonlite::fromJSON(glue("{RELEASES_URL}/{RELEASE}/catalog.json"),
                                  simplifyVector = FALSE)
rel_urls <- function(table) {
  src <- calcofi4r::cc_release_sources(rel_catalog, table)
  sf <- src$single_file
  # `single_file` is NA for a table with no whole-table twin, and as.character(NA)
  # is the STRING "NA" — which nzchar() happily accepts and read_parquet() then
  # resolves to a file called "NA". Test is.na() first, always.
  if (!is.null(sf) && length(sf) == 1 && !is.na(sf) && nzchar(as.character(sf)))
    return(as.character(sf))
  as.character(src$urls)
}
url_list <- function(u) if (length(u) == 1) glue("'{u}'") else
  glue("[{paste0(\"'\", u, \"'\", collapse = ', ')}]")
read_pq <- function(table, hive = FALSE)
  glue("read_parquet({url_list(rel_urls(table))}",
       "{if (hive) ', hive_partitioning = true' else ''}, union_by_name = true)")

con <- dbConnect(duckdb())
for (s in c("INSTALL httpfs", "LOAD httpfs", "SET memory_limit='8GB'",
            "SET enable_progress_bar=false")) try(dbExecute(con, s), silent = TRUE)

# `sample` is read WITHOUT geom: the exporter needs lat/lon numerics, and a
# CRS-tagged geometry column would pull in the spatial extension for nothing.
t0 <- Sys.time()
dbExecute(con, glue("
  CREATE TABLE sample AS
    SELECT sample_key, sample_type, parent_sample_key, root_sample_key, dataset_key,
           grid_key, site_key, cruise_key, order_occ, latitude, longitude, datetime,
           depth_min_m, depth_max_m, tow_type
    FROM {read_pq('sample')};
  CREATE TABLE obs_bio           AS SELECT * FROM {read_pq('obs_bio')};
  CREATE TABLE obs_attribute     AS SELECT * FROM {read_pq('obs_attribute')};
  CREATE TABLE sample_measurement AS SELECT * FROM {read_pq('sample_measurement')};
  CREATE TABLE taxon             AS SELECT * FROM {read_pq('taxon')};
  CREATE TABLE sample_root       AS SELECT * FROM {read_pq('sample_root')};
  CREATE TABLE cruise            AS SELECT * FROM {read_pq('cruise')};
  CREATE TABLE dataset           AS SELECT * FROM {read_pq('dataset')};"))
cat(glue("materialized the core in {round(difftime(Sys.time(), t0, units = 'secs'))}s"), "\n")
q <- function(sql, ...) dbGetQuery(con, glue(sql, ..., .envir = parent.frame()))
```

```{r}
#| label: registries
# gear.csv + life_stage.csv + measurement_type.csv — the single sources of truth
# for samplingProtocol, lifeStage and the P01/P06 ids. An id is filled ONLY on an
# exact match; an empty cell means "no concept says exactly this".
reg <- dwc_registries(here("metadata"))
cat(glue("gear            : {nrow(reg$gear)} codes, ",
         "{sum(!is.na(reg$gear$nerc_l22))} with an L22 device id"), "\n")
cat(glue("life stage      : {nrow(reg$life_stage)} values, ",
         "{sum(!is.na(reg$life_stage$nerc_s11))} with an S11 concept"), "\n")
cat(glue("measurement type: {nrow(reg$measurement_type)} types, ",
         "{sum(!is.na(reg$measurement_type$nerc_p01))} with a P01, ",
         "{sum(!is.na(reg$measurement_type$units_nerc_p06))} with a P06 unit"), "\n")
```

```{r}
#| label: eml
# the release's own EML, one document per dataset (D-8). A staging release may not
# carry eml/ yet; the archive is still built and the missing document is reported,
# never substituted with strings typed here.
EML_DIR <- file.path(OUT_DIR, "_eml", RELEASE)
dir.create(EML_DIR, recursive = TRUE, showWarnings = FALSE)
eml_path_of <- function(k) {
  local <- file.path(EML_DIR, paste0(k, ".xml"))
  if (file.exists(local)) return(local)
  url <- glue("{RELEASES_URL}/{RELEASE}/eml/{k}.xml")
  ok <- tryCatch({ curl::curl_download(url, local, quiet = TRUE); TRUE },
                 error = function(e) FALSE)
  if (ok && file.exists(local)) local else NA_character_
}
```

## The plan — every dataset, before anything is written

```{r}
#| label: plan
# Decision 21 is MEASURED, not listed: a dataset publishes when it has obs_bio rows
# and at least one observed taxon carries a WoRMS id. cce-lter_picoplankton-bacteria
# (flow-cytometry groups, environmental realm) and sio_pic-zooplankton (no taxa) have
# no obs_bio rows at all, so they fall out by construction rather than by exclusion.
cand_all <- dwc_datasets(con)
cand <- if (length(ONLY)) cand_all[cand_all$dataset_key %in% ONLY, , drop = FALSE] else cand_all

names_of <- q("SELECT provider || '_' || dataset AS dataset_key, dataset_name FROM dataset")
plan <- cand |>
  left_join(names_of, by = "dataset_key") |>
  mutate(
    n_events = vapply(dataset_key, function(k)
      q("SELECT COUNT(*) n FROM sample WHERE dataset_key = '{k}'")$n, numeric(1)),
    gear_codes = vapply(dataset_key, function(k)
      nrow(dataset_gear(reg$gear, k)), numeric(1)),
    eml = vapply(dataset_key, function(k) !is.na(eml_path_of(k)), logical(1))) |>
  select(dataset_key, dataset_name, n_events, n_obs, n_taxa, n_worms, n_no_worms,
         n_no_taxon, absence_rule, gear_codes, eml)

write_csv(plan, file.path(OUT_DIR, "publish_plan.csv"), na = "")
kable(plan, caption = glue("Darwin Core plan for {RELEASE} — {nrow(plan)} datasets"))
```

```{r}
#| label: obs-env-for-candidates
# eMoF's third grain is "the env rows sitting on THIS dataset's own events". obs_env
# is 25 M rows across 84 objects hive-partitioned by **measurement_type**, not by
# dataset, so which datasets it covers cannot be read off the object names — one
# projection scan of `dataset_key` answers it in ~13 s, and only the candidates' rows
# are then materialized. Today that is zero rows (every obs_env dataset is
# environmental: bottle, ctd-cast, dic, mets, picoplankton-bacteria), which the count
# below STATES rather than assumes — a bio dataset that starts recording its own CTD
# would light up here with no code change.
env_by_ds <- q("SELECT dataset_key, COUNT(*) AS n FROM {read_pq('obs_env', hive = TRUE)}
                GROUP BY 1 ORDER BY 1")
env_cand <- intersect(env_by_ds$dataset_key, cand$dataset_key)
cat(glue("obs_env covers: {paste(env_by_ds$dataset_key, collapse = ', ')}"), "\n")
if (length(env_cand)) {
  keys <- paste0("'", env_cand, "'", collapse = ", ")
  dbExecute(con, glue("CREATE TABLE obs_env AS SELECT * FROM
                       {read_pq('obs_env', hive = TRUE)} WHERE dataset_key IN ({keys})"))
  cat(glue("\nenv rows on a candidate's own events: ",
           "{q('SELECT COUNT(*) n FROM obs_env')$n} ({paste(env_cand, collapse = ', ')})"), "\n")
} else {
  cat("\nno candidate dataset has obs_env rows — the eMoF env grain is empty\n")
}
```

```{r}
#| label: plan-notes
bio_all <- q("SELECT DISTINCT dataset_key FROM obs_bio ORDER BY 1")$dataset_key
no_worms <- setdiff(bio_all, cand_all$dataset_key)
if (length(no_worms))
  cat(glue("biological in the core but no taxon resolves to WoRMS ",
           "(no IPT resource, Decision 21): {paste(no_worms, collapse = ', ')}"), "\n")
env_only <- setdiff(q("SELECT DISTINCT dataset_key FROM sample")$dataset_key, bio_all)
cat(glue("\nno obs_bio rows at all (environmental or taxon-free, nothing to publish ",
         "to OBIS): {paste(sort(env_only), collapse = ', ')}"), "\n")
if (any(!plan$eml))
  cat(glue("\nno eml/{{dataset_key}}.xml in {RELEASE} for: ",
           "{paste(plan$dataset_key[!plan$eml], collapse = ', ')} — ",
           "build it with build_eml_catalog() + write_eml_files() ",
           "(release_database.qmd s 3c) before an upload"), "\n")
```

## Duplicates first — what OBIS already holds for these sources

::: {.callout-warning title="A provider's own record is never duplicated"}
Before any upload, the OBIS records that already cover the same source are listed
and resolved **with their owners** (Decision 21): a historical CalCOFI record is
retired or cross-referenced with its owner's agreement, and a dataset whose
authority is CCE-LTER or NOAA is never republished by CalCOFI without that
provider's yes. This table is for Ben and the providers; it blocks nothing here,
because nothing here uploads.
:::

```{r}
#| label: obis-duplicates
# the curated rows first: metadata/distribution.csv is the record of what CalCOFI
# already published where, and it is never guessed from a search
dist <- read_distribution_registry(here("metadata/distribution.csv"))
known <- dist |>
  filter(portal %in% c("obis", "ipt")) |>
  select(dataset_key, portal, id, url, title, status, notes)
if (nrow(known)) kable(known, caption = "curated OBIS / IPT distributions (distribution.csv)")

obis_search <- function(terms) {
  # OBIS's dataset search, read-only. The CalCOFI record is NOT findable by the
  # word "CalCOFI" (measured, 2026-09-05 — that is why distribution.csv carries
  # its id), so institution and taxon words are what is asked here.
  out <- lapply(terms, function(tm) {
    u <- glue("https://api.obis.org/v3/dataset?q={utils::URLencode(tm, reserved = TRUE)}&size=25")
    j <- tryCatch(jsonlite::fromJSON(u, simplifyVector = FALSE), error = function(e) NULL)
    r <- if (is.null(j)) list() else j$results
    if (!length(r)) return(NULL)
    tibble(
      term    = tm,
      obis_id = vapply(r, function(x) x$id %||% NA_character_, ""),
      title   = vapply(r, function(x) x$title %||% NA_character_, ""),
      owner   = vapply(r, function(x)
        paste(unlist(lapply(x$institutes %||% list(), function(i) i$name)), collapse = "; "), ""),
      records = vapply(r, function(x) as.numeric(x$records %||% NA), numeric(1)),
      published = vapply(r, function(x) as.character(x$published %||% NA), ""))
  })
  bind_rows(out)
}

if (NET) {
  hits <- obis_search(c("CalCOFI",
                        "California Cooperative Oceanic Fisheries Investigations",
                        "Southwest Fisheries Science Center",
                        "Scripps Institution of Oceanography",
                        "California Current Ecosystem LTER"))
  # OBIS's `q` is a loose full-text OR, so "California Current euphausiid" returns
  # Happywhale killer whales. The candidate list is therefore narrowed to records
  # whose OWN title or institute names one of the organizations behind these
  # datasets — a filter on what came back, never a claim that nothing else exists.
  who <- "calcofi|cooperative oceanic|southwest fisheries|scripps|california current ecosystem|cce.?lter|farallon|dungeness"
  hits <- hits |>
    distinct(obis_id, .keep_all = TRUE) |>
    filter(grepl(who, tolower(paste(title, owner)))) |>
    arrange(desc(records))
  if (nrow(hits)) {
    kable(hits |> select(-term), caption = paste(
      "OBIS datasets whose title or institute names one of these datasets' sources —",
      "resolve each with its owner before an upload (a provider's own record is never",
      "republished). Not exhaustive: OBIS's text search does not match every record,",
      "which is why metadata/distribution.csv holds the ids we know."))
  } else cat("no OBIS dataset named one of these sources in its title or institute\n")
} else {
  cat("CALCOFI_OFFLINE=true — the OBIS duplicate check was skipped (not 'no duplicates')\n")
}
```

## Build every archive

```{r}
#| label: build
results <- list()
findings <- list()

for (k in plan$dataset_key) {
  cat(glue("\n--- {k} ---"), "\n")
  d_dir <- file.path(OUT_DIR, k)

  ev <- dwc_event(con, k, gear = reg$gear, measurement_type = reg$measurement_type)
  oc <- dwc_occurrence(con, k, life_stage = reg$life_stage,
                       measurement_type = reg$measurement_type)
  mf <- dwc_emof(con, k, occurrence = oc, measurement_type = reg$measurement_type,
                 env = TRUE)

  chk <- dwc_check(ev, oc, mf, dataset_key = k)
  findings[[k]] <- chk
  print(chk[, c("finding", "level", "n", "detail")], row.names = FALSE)

  if (any(chk$level == "error")) {
    # a failing dataset gets NO zip: a broken archive at OBIS is worse than a
    # missing one, and the finding is what says why
    cat(glue("SKIPPED — {sum(chk$level == 'error')} error finding(s); no archive written"), "\n")
    results[[k]] <- tibble(dataset_key = k, status = "failed checks",
                           n_event = nrow(ev), n_occurrence = nrow(oc), n_emof = nrow(mf),
                           content_hash = NA_character_, archive = NA_character_)
    next
  }

  ipt <- dist |> filter(dataset_key == k, portal == "ipt") |> pull(id)
  obs_id <- dist |> filter(dataset_key == k, portal == "obis") |> pull(id)
  a <- dwc_archive(
    d_dir, ev, oc, mf,
    eml_path        = eml_path_of(k),
    dataset_key     = k,
    version         = RELEASE,
    ipt_resource    = if (length(ipt)) ipt[1] else NULL,
    obis_dataset_id = if (length(obs_id)) obs_id[1] else NULL)
  cat(glue("wrote {basename(a$zip)} ({round(file.size(a$zip) / 1e6, 1)} MB), ",
           "content_hash {substr(a$content_hash, 1, 12)}"), "\n")
  results[[k]] <- tibble(dataset_key = k, status = "built",
                         n_event = nrow(ev), n_occurrence = nrow(oc), n_emof = nrow(mf),
                         content_hash = a$content_hash, archive = basename(a$zip))
}
res <- bind_rows(results)
```

```{r}
#| label: build-summary
kable(res, caption = glue("Darwin Core archives written for {RELEASE}"))
fin <- bind_rows(findings)
if (nrow(fin)) kable(fin[fin$finding != "ok", ],
                     caption = "every finding, by dataset (error = no archive written)")
```

## Registration status — when is an upload due?

```{r}
#| label: manifests
mans <- Sys.glob(file.path(OUT_DIR, "*_manifest.json"))
if (length(mans)) {
  st <- bind_rows(lapply(mans, dwc_manifest_status))
  kable(st, caption = paste(
    "the manifest each archive carries: `built, not uploaded` until the IPT copy is",
    "made, then `published (vX)` while the uploaded bytes are these bytes and",
    "`stale — data changed in vY` once they are not. This is what `registrations[]`",
    "in datasets.json reads."))
} else cat("no manifests yet\n")
```

## Upload — a deliberate manual step

::: {.callout-note title="What this notebook does NOT do"}
It does not upload. The OBIS-USA IPT (`ipt-obis.gbif.us`) holds CalCOFI's
resources under Ben's login, and an upload is made only after the dataset's
provider has agreed (Decision 21) and only when the manifest's `content_hash`
differs from the published copy's. The steps are in
[`docs/portals.qmd` § OBIS](https://calcofi.io/docs/portals.html#obis); in short:
create or open the resource, upload the zip's five files as the source, map the
Event core + the two extensions from `meta.xml`, publish a new version, then
record the resulting OBIS dataset id in `metadata/distribution.csv` and stamp
`uploaded_utc` / `uploaded_hash` into the archive's `{dataset_key}_manifest.json`.
:::

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
cat("done\n")
```
