---
title: "CalCOFI CTD QA/QC Protocol"
subtitle: "What every check does, where its threshold came from, and what it cannot see"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    code-fold: true
    code-tools: true
calcofi:
  target_name: qc_protocol
  workflow_type: reference
  # every pipeline notebook must declare its output: build_targets_list() uses it
  # as the target's file, and a missing key fails the whole run at resolution time
  output: _output/qc_protocol.html
  workflow_url: https://calcofi.io/workflows/qc_protocol.html
editor_options:
  chunk_output_type: console
---

## What this document is

The reference for CalCOFI CTD quality control: every check that runs, stated
plainly enough that a data manager can agree or disagree with it without reading
SQL, and specific enough that a reviewer looking at a flagged cast knows what the
flag means.

**It is generated from the rules themselves.** The per-rule sections below are
assembled at render time from `metadata/qc_rules/` — `rules.csv` for the index,
`sql/*.sql` for the check and for the reasoning its author recorded in the file's
header comment. Nothing here is a separate description that could drift from the
code: change a threshold and this page changes with it, and a rule that runs
without a rationale [fails this document's build](#coverage).

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
librarian::shelf(
  calcofi4db, dplyr, DT, glue, here, knitr, purrr, readr, stringr,
  quiet = T)

source(here("libs/build_qc_protocol.R"))

dir_rules <- here("metadata/qc_rules")
rules_all <- qc_read_rules(dir_rules, active_only = FALSE)
rules_act <- filter(rules_all, active)

say <- function(...) cat(..., "\n", sep = "")
say(glue("rules: {nrow(rules_all)} registered, {nrow(rules_act)} active, ",
         "{nrow(rules_all) - nrow(rules_act)} parked"))
say(glue("SQL files: {length(unique(na.omit(rules_all$sql_file)))}"))
```

### How to read a rule

Each rule declares four things that decide how a result should be read.

| field | what it changes |
|---|---|
| **severity** | `error` means the pipeline is broken and the finding is a defect in *our* processing. `warning` means the data looks wrong and a human must decide. No rule drops or rewrites a value. |
| **target** | `sample` (one row per cast), `obs` (the thinned, published profile) or `obs_ctd_full` (the full-resolution scans, supplemental). A rule on `obs` cannot see what thinning removed. |
| **scope** | `all` runs over the whole dataset. `cruise` rules read the 212M-row `obs_ctd_full` and are run one cruise at a time — they are a *sample*, never a sweep, and the app is where the remaining cruises get covered. |
| **requires** | measurement types that must be present. If one is absent the rule reports **`skip`**, never `pass`. |

That last row is the single most important convention here.

Every rule returns at least `subject_key` (what is flagged — the unit of review)
and `detail` (one sentence naming the problem). A rule whose finding concerns a
particular **scan** rather than a whole cast also returns `depth_min_m` and
`measurement_type`. Those two are a contract rather than a habit: they are what
lets a reviewer click a finding in `ctd-qaqc` and land on the right profile at the
right depth, without the app needing to know anything about the rule that produced
it. The two cast-level rules below (`ctd_cast_deeper_than_seafloor`,
`ctd_cast_position_vs_station`) correctly return neither — they are about the cast,
not about a reading.

::: {.callout-important title="`skip` is not `pass`, and this is not hypothetical"}
A rule whose input is missing returns zero rows — which is indistinguishable from
"the data is clean" unless something says otherwise. The three bottle-vs-sensor
calibration rules did exactly that against release `v2026.07.30`, which predates
the change making the bottle reference types canonical: they reported no findings
while comparing nothing at all.

The engine therefore treats an unmet precondition as `skip` with a stated reason,
and the same principle governs reference data — `calcofi4db::qc_stage_reference()`
leaves a missing input as a **missing table** so its rules error visibly, rather
than creating an empty one whose rules would quietly pass.
:::

Results carry one of five statuses:

| status | meaning |
|---|---|
| `pass` | ran, found nothing |
| `flag` | ran, found something a human should look at (`severity = warning`) |
| `FAIL` | ran, found something that should not be possible (`severity = error`) |
| `skip` | **did not run** — a precondition was unmet |
| `ERROR` | could not run (missing reference table, bad SQL) |

## Where these rules came from

Two sources, deliberately combined: the checks CalCOFI has been running for
decades, and the checks that only make sense on CTD data.

### Mining the CalCOFI hydrographic master

The program's working database is an MS Access file carrying the bottle record
from 1949, and — more valuably — **155 saved queries**, a self-documenting field
dictionary, an instrument/accuracy registry, a 280-row dated change log, and a
fitted seasonal climatology. Three decades of institutional QA logic lived there
and nowhere else.

It was extracted reproducibly on macOS/Linux (`libs/extract_accdb.R`,
`scripts/extract_accdb.sh`) with **two engines on purpose**: `mdbtools` for bulk
table export, and Jackcess for the saved-query SQL, which is authoritative.
`mdb-queries` reconstructs the stored parse tree only partially — it silently drops
JOIN clauses, GROUP BY, HAVING and aliases. The output looks plausible and is
wrong: a LEFT JOIN find-unmatched query degrades into a cross join, which inverts
the meaning of the QA check it encodes. Both engines are run and diffed so a
regression in either surfaces instead of silently corrupting a ported rule.

Of the 155 queries, only one family is standing QA: the `TQ -` / `TR -` / `TV -`
tests and the `Anomalies` family. The rest are corrections, materializations and
derived products.

```{r}
#| label: triage

read_csv(here("metadata/calcofi/hydro-master/query_triage.csv"),
         show_col_types = FALSE) |>
  count(category, sort = TRUE) |>
  mutate(`what it is` = recode(category,
    validate           = "QA checks — the family ported here",
    `derived-product`  = "computed products (MLD, nutricline, anomalies)",
    `correction-history` = "one-off UPDATE/DELETE repairs, dated and initialled",
    materialization    = "make-table steps feeding other queries",
    `export-or-info`   = "extracts for outside consumers",
    unclassified       = "not confidently categorized")) |>
  kable(caption = "The 155 Access saved queries, triaged")
```

**A faithful port is not always the right port.** Two of the Access checks could
not be carried over as written, and both are documented on the rules that replaced
them: `TQ - BottomDepth_Vs_AvgBottomDepth` compares the ship's echosounder to the
station average, but CTD casts carry no reported bottom depth at all
(`bottom_depth` exists in `sample_measurement` for 33,363 **bottle** casts and for
**0** of 14,336 CTD casts) — so the ported form tests position plausibility against
bathymetry instead, which is a different question. And
`TQ - StationNameChecker` compares `Rpt_Line > "76.6"` as a **string**, so line 100
sorts below line 76.6 and has been silently excluded from that check for twenty
years; whether to port the bug or the intent is question `hydro_master_09`.

### The climatology, and why two properties are excluded

`HarmCoeffBottle` holds a fitted annual harmonic per station × standard depth ×
property — mean, amplitude, frequency, phase, and a residual standard deviation.
That turns "is this value plausible" from a fixed global range into a *local,
seasonal* question, which nothing in the pipeline could ask before.

The providers have not confirmed the fitting procedure (question
`hydro_master_12`), so the reconstruction form was **determined empirically**:
candidate forms were scored against the 200,640 bottle observations the
coefficients were derived from.

| form | RMSE | vs mean-only |
|---|---|---|
| `Mean` (null) | 1.269 | — |
| **`Mean + Ampl·sin(Freq·(doy − Phase))`** | **0.949** | **+25.2%** |
| `Mean + Ampl·cos(Freq·(doy − Phase))` | 2.113 | −66.4% |
| `Mean + Ampl·cos(Freq·doy − Phase_deg)` | 1.800 | −41.8% |
| `Mean + Ampl·sin(2π(doy − Phase)/365)` | 1.474 | −16.1% |

Every cosine variant is *worse than using the mean alone*. Two independent checks
agree with the winner: physically, the seasonal signal is strongest at the surface
(42% RMSE reduction at 0 m) and decays monotonically to ~5% by 150–500 m, which is
what seasonal heating does and what a wrong form would not show; statistically,
residuals divided by `StDev` are near-standard-normal (mean −0.06, sd 1.14),
confirming `StDev` is the residual scale a z-score needs.

Each property was then scored the same way, and **only those that passed are used**:

| property | RMSE gain | sd(z) | verdict |
|---|---|---|---|
| Temp | 25.2% | 1.14 | use |
| SiO3 | 11.6% | 1.12 | use |
| NO3 | 10.3% | 1.40 | use |
| PO4 | 9.7% | 1.26 | use |
| O2 | 9.3% | 1.29 | use |
| Sig | 2.4% | 4.80 | **exclude** |
| Sal | 0.5% | 6.73 | **exclude** |
| NO2 | 5.0% | 5.4e12 | **exclude** |
| Chla | 4.3% | 1.2e11 | **exclude** |

Salinity and sigma-theta are excluded on evidence, not taste. A seasonal harmonic
explains almost none of their variance here — Southern California salinity is
advection-driven, not seasonally heated — and the tabulated `StDev` is 5–7× smaller
than the actual residual, so it is measuring something other than residual spread.
A z-score built on it would be wildly over-dispersed and would flag thousands of
ordinary values. NO2 and Chla fail differently: a handful of station × depth cells
carry a near-zero `StDev`, which makes z explode (one nitrite cell produced
z = 3.5e12), so they are left out rather than patched.

### Reference data the rules join against

None of this is part of the release. It is the QC engine's *input*, staged onto a
connection by `calcofi4db::qc_stage_reference()`.

```{r}
#| label: reference_inputs

tibble::tribble(
  ~table,                  ~source,                                        ~role,
  "measurement_type",      "metadata/measurement_type.csv",                "valid_min/valid_max per type — the registry, not a release, because it moves ahead of releases",
  "measurement_qual",      "metadata/measurement_qual.csv",                "the quality-code vocabulary, recovered from the Access field documentation",
  "climatology_harmonic",  "hydro-master/reference/climatology_harmonic.csv", "fitted seasonal expectation per station x depth x property",
  "station",               "hydro-master/reference/station.csv",           "the 75 current standard stations and their average bottom depth",
  "standard_depth",        "hydro-master/reference/standard_depth.csv",    "the standard depth grid",
  "sample_seafloor",       "GEBCO 2025 crop (apps/ctd-viz)",               "seafloor depth at each cast position — the only bottom-depth reference CTD casts have") |>
  kable(caption = "QC reference inputs, none of which ship in the release")
```

### The quality-code vocabulary

`measurement_qual` was a verbatim pass-through for years: carried through every
table, never interpreted, with no vocabulary anywhere in the repo. There is one
now, recovered from the hydrographic master's own field documentation — but it is
**incomplete, and the gaps are blockers**.

```{r}
#| label: qual_vocab

read_csv(here("metadata/measurement_qual.csv"),
         col_types = cols(qual_code = col_character(), .default = col_guess())) |>
  select(code = qual_code, label, description, documented = is_documented,
         n_observed) |>
  arrange(desc(documented), code) |>
  datatable(
    caption  = "metadata/measurement_qual.csv — the controlled vocabulary",
    rownames = FALSE,
    options  = list(pageLength = 12, scrollX = TRUE, dom = "tip")) |>
  formatStyle("documented",
              color = styleEqual(c(TRUE, FALSE), c("inherit", "#d03b3b")))
```

Three questions are open with the providers, two of them blockers: what the
undocumented single-digit codes mean (`hydro_master_02`), whether a NULL flag means
"good" or "never assessed" (`hydro_master_05`), and whether `S_qual`'s **253
distinct values** in the bottle master — clustered at 256–271, which looks like bit
8 of a bitmask — indicate a bitmask or a corrupted column (`hydro_master_02b`).
Every other quality column has 3–10 distinct values.

**Nothing in this protocol acts on a flag.** Acting on one whose semantics are
unconfirmed would be guessing with released data.

```{r}
#| label: rules_index

rules_all |>
  transmute(rule_key, active, severity, type = rule_type, target,
            scope = coalesce(scope, "all"), description) |>
  arrange(desc(active), type, rule_key) |>
  datatable(
    caption  = "The registry: metadata/qc_rules/rules.csv",
    rownames = FALSE,
    options  = list(pageLength = 20, scrollX = TRUE, dom = "tip")) |>
  formatStyle("active",
              color = styleEqual(c(TRUE, FALSE), c("inherit", "#6e7781")))
```

# The rules {#the-rules}

Grouped by what kind of question each asks. Rules that share a query — the three
bottle-vs-sensor calibration checks, the two spike checks — show the SQL once, under
the first of them.

```{r}
#| label: rules_active
#| output: asis

md_active <- qc_protocol_markdown(
  arrange(rules_act, rule_type, rule_key), group_by = "rule_type")
cat(md_active)
```

# Parked rules

Registered, documented, and deliberately **not running**. They are kept in the
registry rather than deleted so that the reason survives — a rule that quietly
disappears looks like a rule that was never considered.

```{r}
#| label: rules_parked
#| output: asis

rules_off <- filter(rules_all, !active)
md_parked <- if (nrow(rules_off))
  qc_protocol_markdown(rules_off, group_by = NULL) else
    "*No parked rules.*\n"
cat(md_parked)
```

# Standards and prior art

This protocol is deliberately small and CalCOFI-specific. It is not a
reimplementation of the community standards, and where it diverges from them that
is worth stating.

### QARTOD

The IOOS *Manual for Real-Time Quality Control of In-situ Temperature and Salinity
Observations* (and its dissolved-oxygen and ocean-optics companions) is the
community reference for profile QC, and its `1` / `2` / `3` / `4` / `9` flags
(pass / not evaluated / suspect / fail / missing) are what an ERDDAP consumer
expects to find. Several tests here have direct QARTOD counterparts — gross range
(QARTOD test 6), spike (9), rate of change (10), flat line (11), density inversion
(15) — though the thresholds are ours and derived from this dataset rather than
adopted from the manual.

CalCOFI CTD is published to `erddap.calcofi.io`, where the native codes currently
pass through uninterpreted, and one of them (`6` = "data OK but taken from CTD
sensor") is meaningless for a CTD dataset. **Emitting QARTOD flags alongside the
native codes is question `calcofi_ctd-cast_11`**, and is the natural home for the
results of these rules once the flag semantics above are settled.

### `oce` — Dan Kelley

The reference R package for oceanographic data, and the closest prior art for the
profile rules here: `read.ctd()`, `ctdTrim()` (removes the soak and the upcast),
`despike()`, and `ctdDecimate()` (bins a profile to standard depths). Anyone
implementing these checks should read the *oce* CTD vignette and
*Analysis of Oceanographic Data* alongside this document.

Two deliberate differences. `ctdTrim()` discards the upcast; we keep it, in
`obs_ctd_full`, precisely so the down/up comparison is available as a check — the
disagreement is signal about sensor lag and fouling, not noise to be trimmed.
And `ctdDecimate()` resamples onto a regular grid by interpolation, whereas
`ctd_thin` *selects real scans* and never interpolates, so a thinned value is
always a measurement that was actually taken.

### `castr` — Jean-Olivier Irisson

Cast-oriented profile utilities: smoothing, despiking, and detection of the mixed
layer depth, deep chlorophyll maximum and thermocline. Relevant twice over — as a
comparison for the despiking approach, and because the Access master's `MLD_Sigma`
and `NutClineDepth` tables are exactly the derived products `castr` computes. Those
two tables are committed under
`metadata/calcofi/hydro-master/reference/` unused, so that any reimplementation can
be validated against the Access-era numbers rather than trusted.

::: {.callout-warning title="The CRAN package named `CTD` is not oceanographic"}
`CTD` on CRAN is "Connecting The Dots" — a weighted-graph pattern-discovery method
for metabolomics. It has nothing to do with conductivity-temperature-depth
profiles. It is an easy citation to make by name-matching and a hard one to notice
afterwards, so: do not cite it here.
:::

# What is not checked

Stated so that a clean run is not over-read.

- **The headline variables carry no quality flags at all.** Source flags attach to
  the component sensors (`Temp1Q`, `Salt1Q`, `Ox1Q`, `Ox2Q`); the canonical types
  are the *averages* of them, so the flag is lost in the mean. Only 8 of 27
  published types carry any flag, over 0.05% of rows. How a sensor flag should
  qualify an average of two is question `calcofi_ctd-cast_09` — a blocker.
- **Range coverage is partial.** 18 of the 27 published types have a declared
  `valid_min`/`valid_max`; the rest — mostly raw voltages and derived quantities
  with no agreed bound — are never range-checked.
- **The profile rules are sampled, not swept.** Spike, loop-edit and up/down
  disagreement run one cruise at a time against `obs_ctd_full`; a whole-archive
  sweep is the `ctd-qaqc` app's job, not this document's.
- **Dual-sensor disagreement is parked.** The thresholds per property are not
  agreed, and a rule with a guessed threshold is worse than a rule that says it is
  not running.
- **No rule drops, rewrites or interpolates a value.** Every result is a report.
  Two blocker questions remain open with the data providers, and acting on them
  before they are answered would be guessing with released data.

# Coverage {#coverage}

The build fails if a rule that runs is not described here, or carries no rationale
in its SQL header. A protocol that silently omits a check is worse than no
protocol, because it reads as completeness.

```{r}
#| label: coverage_check

qc_protocol_check(rules_all, paste(md_active, md_parked))

say(glue("active rules documented : {nrow(rules_act)}/{nrow(rules_act)}"))
say(glue("parked rules documented : {nrow(rules_off)}/{nrow(rules_off)}"))
say(glue("every active rule carries a rationale in its SQL header: yes"))
```

## Where results appear

- **[`ctd-qaqc`](https://github.com/CalCOFI/apps/tree/main/ctd-qaqc)** — runs this
  registry interactively against a release, one rule or one cruise at a time, and
  records review verdicts.
- **[`ingest_calcofi_ctd-cast.qmd`](https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html#findings)**
  — runs the same registry over the parquet each ingest writes, so the notebook
  reports the condition of the data it just published.

Both call `calcofi4db::qc_run_all()`. There is one engine and one registry: a
number in the app and a number in the notebook cannot disagree.

## Session Info

```{r}
#| label: session_info
#| code-fold: true

sessioninfo::session_info()
```
