---
title: "Questions for Data Providers"
editor_options:
  chunk_output_type: console
---

## Overview

Aggregates the per-dataset `metadata/{provider}/{dataset}/questions.csv` files
(seeded during `/explore-dataset` and `/generate-metadata`, rendered inside each
`ingest_*`/`publish_*` workflow) into **one draft email per provider** for the
data team to review and send. Nothing is sent — this produces copy-paste drafts.

Two statuses go out, and they ask for different things:

- **`open`** — a question we cannot answer ourselves.
- **`proposed`** — we have already built or reasoned an answer and want it
  **confirmed**. `proposed_answer` goes into the email, so the provider is
  approving a solution rather than being handed a problem.

`answered` and `wontfix` are excluded.

```{r setup}
#| message: false
librarian::shelf(
  calcofi4db, dplyr, glue, here, purrr, readr, stringr, knitr, quiet = T)
options(readr.show_col_types = FALSE)
```

## Collect Questions

```{r collect}
q_files <- Sys.glob(here("metadata/*/*/questions.csv"))

# read_questions() enforces the controlled vocabulary, so a registry that
# invents a status fails HERE rather than quietly dropping out of the email —
# which is what a hand-written `filter(status %in% ...)` used to do.
questions <- map_dfr(q_files, function(f) {
  rel   <- str_replace(f, "^.*/metadata/", "")        # provider/dataset/questions.csv
  parts <- str_split(rel, "/", simplify = TRUE)
  read_questions(f) |>
    mutate(provider = parts[1], dataset = parts[2], .before = 1)
})

# dataset metadata comes from each ingest's `calcofi:` YAML — metadata/dataset.csv
# is deprecated and had already drifted from the notebooks
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  select(provider, dataset, dataset_name, pi_names)

open_q <- questions |>
  filter(status %in% c("open", "proposed")) |>
  left_join(d_dataset, by = c("provider", "dataset")) |>
  mutate(
    priority     = factor(priority, question_priorities()),
    dataset_name = coalesce(dataset_name, glue("{provider}/{dataset}")))

cat(glue(
  "{nrow(open_q)} questions to send across ",
  "{n_distinct(open_q$provider)} providers ",
  "({sum(open_q$priority == 'blocker', na.rm = TRUE)} blockers, ",
  "{sum(open_q$status == 'proposed')} awaiting confirmation)"), "\n")
```

## Summary

```{r summary}
open_q |>
  count(provider, dataset, status, priority, .drop = FALSE) |>
  filter(n > 0) |>
  arrange(provider, dataset, status, priority) |>
  kable()
```

## Draft Emails

One draft per provider. Recipients are drawn from `pi_names` in each ingest's
`calcofi.dataset_meta` YAML; fill in the bracketed placeholders before sending.

```{r emails}
#| results: asis

providers <- sort(unique(open_q$provider))

for (prov in providers) {
  pq <- open_q |> filter(provider == prov) |> arrange(dataset, priority)

  recipients <- pq$pi_names |>
    na.omit() |> unique() |>
    str_split(";\\s*") |> unlist() |> unique() |> str_trim()
  recipients <- recipients[recipients != ""]
  to_line <- if (length(recipients)) paste(recipients, collapse = ", ") else
    "[data provider contact]"

  n_block <- sum(pq$priority == "blocker", na.rm = TRUE)

  cat(glue("\n### {prov} ({nrow(pq)} questions, {n_block} blockers)\n\n"))
  cat("```\n")
  cat(glue("To: {to_line}\n"))
  cat(glue("Subject: CalCOFI data integration — follow-up questions on ",
           "{paste(sort(unique(pq$dataset_name)), collapse = ', ')}\n\n"))
  cat("Hi,\n\n")
  cat(str_wrap(glue(
    "As we integrate your data into the CalCOFI database we have a few ",
    "follow-up questions. Answers will improve how the data are matched, ",
    "documented, and published. Questions marked [BLOCKER] are gating ",
    "publication. Items marked [PROPOSED] already have an answer from us — ",
    "we only need you to confirm or correct it. Thank you!"), width = 78), "\n")

  for (ds in sort(unique(pq$dataset))) {
    dq <- pq |> filter(dataset == ds)
    cat(glue("\n\n{unique(dq$dataset_name)}:\n"))
    for (i in seq_len(nrow(dq))) {
      tag <- paste0(
        if (!is.na(dq$priority[i]) && dq$priority[i] == "blocker") " [BLOCKER]" else "",
        if (dq$status[i] == "proposed") " [PROPOSED]" else "")
      field <- if (!is.na(dq$related_field[i]) && nzchar(dq$related_field[i]))
        glue(" (re: {dq$related_field[i]})") else ""
      # the durable label, not the loop counter: "Q15" in the email must be the
      # same Q15 the notebook and the protocol document cite
      cat(glue("\n  {dq$label[i]}.{tag} {dq$question[i]}{field}\n"))
      if (!is.na(dq$context[i]) && nzchar(dq$context[i]))
        cat(str_wrap(glue("      Context: {dq$context[i]}"),
                     width = 76, exdent = 9), "\n")
      if (!is.na(dq$proposed_answer[i]) && nzchar(dq$proposed_answer[i]))
        cat(str_wrap(glue("      We propose: {dq$proposed_answer[i]}"),
                     width = 76, exdent = 9), "\n")
    }
  }
  cat("\n\nBest regards,\nThe CalCOFI data team\n")
  cat("```\n\n")
}
```
