---
title: "Release CalCOFI Database"
execute:
  echo: true
  warning: false
editor_options:
  chunk_output_type: console
format:
  html:
    code-fold: true
editor:
  markdown:
    wrap: 72
---

## Overview {.unnumbered}

**Goal**: Create a frozen (immutable) release of the CalCOFI integrated
database by assembling all ingest parquet outputs. This is the "caboose"
notebook that always runs last, after all ingest notebooks complete.

**Upstream notebooks** (must complete first):

1.  `ingest_swfsc_ichthyo.qmd` — ichthyo tables + taxonomy
2.  `ingest_calcofi_bottle.qmd` — bottle/cast tables + cruise bridge
3.  `ingest_calcofi_ctd-cast.qmd` — CTD tables

```{mermaid}
%%| label: fig-workflow
%%| fig-cap: "Pipeline: ingest notebooks produce parquet → release caboose assembles frozen release"

flowchart LR
    subgraph ingest["Ingest Notebooks"]
        i1["ichthyo<br/>12 tables"]
        i2["bottle<br/>5 tables"]
        i3["ctd-cast<br/>4 tables"]
    end
    subgraph assembly["Assembly (this notebook)"]
        a["Load all parquet<br/>into in-memory DuckDB"]
    end
    subgraph release["Frozen Release"]
        f["Clean parquet<br/>+ catalog.json"]
    end
    i1 --> a
    i2 --> a
    i3 --> a
    a --> f

    style ingest fill:#e3f2fd,stroke:#1565c0
    style assembly fill:#fff3e0,stroke:#ef6c00
    style release fill:#e8f4e8,stroke:#2e7d32
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
# cleanup_gcs_obsolete(dry_run = F)
librarian::shelf(
  CalCOFI / calcofi4db,
  DBI,
  dm,
  dplyr,
  DT,
  fs,
  glue,
  here,
  jsonlite,
  purrr,
  tibble,
  quiet = T
)
options(DT.options = list(scrollX = TRUE))

# release version
release_version <- format(Sys.Date(), "v%Y.%m.%d")
message(glue("Release version: {release_version}"))
```

## Assemble from Ingest Outputs

Load all per-dataset parquet outputs into an in-memory DuckDB. Each
ingest notebook writes parquet to `data/parquet/{dataset}/` and uploads
to `gs://calcofi-db/ingest/{dataset}/`.

```{r}
#| label: assemble_working

con_wdl <- get_duckdb_con(":memory:")
load_duckdb_extension(con_wdl, "spatial")

# load all ingest parquet outputs into one DuckDB
parquet_dirs <- c(
  here("data/parquet/swfsc_ichthyo"),
  here("data/parquet/calcofi_bottle"),
  here("data/parquet/calcofi_ctd-cast"),
  here("data/parquet/calcofi_dic")
)

load_stats <- purrr::map_dfr(parquet_dirs, function(pd) {
  load_prior_tables(con = con_wdl, parquet_dir = pd)
})

load_stats |>
  datatable(caption = "Assembled tables from ingest parquet outputs")
```

## Validate

Cross-dataset validation to ensure data integrity before freezing.

```{r}
#| label: validate

# grid_key integrity: casts.grid_key should all be in grid.grid_key
tbls <- DBI::dbListTables(con_wdl)

if (all(c("casts", "grid") %in% tbls)) {
  if (
    "grid_key" %in%
      DBI::dbListFields(con_wdl, "casts") &&
      "grid_key" %in% DBI::dbListFields(con_wdl, "grid")
  ) {
    grid_orphans <- dbGetQuery(
      con_wdl,
      "SELECT COUNT(*) AS n FROM casts c
       WHERE c.grid_key IS NOT NULL
         AND c.grid_key NOT IN (SELECT grid_key FROM grid)"
    )$n
    message(glue("Grid key orphans in casts: {grid_orphans}"))
  }
}

# ship PK uniqueness
if ("ship" %in% tbls) {
  ship_dups <- dbGetQuery(
    con_wdl,
    "SELECT ship_key, COUNT(*) AS n FROM ship
     GROUP BY ship_key HAVING COUNT(*) > 1"
  )
  if (nrow(ship_dups) > 0) {
    warning(glue("Duplicate ship_key values: {nrow(ship_dups)}"))
  } else {
    message("ship_key: all unique")
  }
}

# cruise PK uniqueness
if ("cruise" %in% tbls) {
  cruise_dups <- dbGetQuery(
    con_wdl,
    "SELECT cruise_key, COUNT(*) AS n FROM cruise
     GROUP BY cruise_key HAVING COUNT(*) > 1"
  )
  if (nrow(cruise_dups) > 0) {
    warning(glue("Duplicate cruise_key values: {nrow(cruise_dups)}"))
  } else {
    message("cruise_key: all unique")
  }
}

# cruise bridge coverage
if ("casts" %in% tbls) {
  bridge_stats <- dbGetQuery(
    con_wdl,
    "SELECT
      COUNT(*) AS total_casts,
      SUM(CASE WHEN ship_key IS NOT NULL THEN 1 ELSE 0 END) AS with_ship_key,
      SUM(CASE WHEN cruise_key IS NOT NULL THEN 1 ELSE 0 END) AS with_cruise_key
    FROM casts"
  )
  bridge_stats |> datatable(caption = "Cruise bridge coverage")
}

# run standard release validation
validation <- validate_for_release(con_wdl)
if (validation$passed) {
  message("Release validation passed!")
} else {
  cat("Validation FAILED:\n")
  cat(paste("-", validation$errors, collapse = "\n"))
}
```

## Show Combined Schema

```{r}
#| label: combined_schema
#| fig-width: 12
#| fig-height: 10

# build dm from connection (exclude internal tables)
schema_tbls <- setdiff(tbls, c("_meta", "_sp_update"))
# exclude views
schema_tbls <- schema_tbls[
  !schema_tbls %in% c("casts_derived", "ctd_cast_derived")
]

d <- dm::dm_from_con(con_wdl, table_names = schema_tbls, learn_keys = FALSE)

# merge per-dataset relationships.json files
rels_paths <- c(
  here("data/parquet/swfsc_ichthyo/relationships.json"),
  here("data/parquet/calcofi_bottle/relationships.json"),
  here("data/parquet/calcofi_ctd-cast/relationships.json"),
  here("data/parquet/calcofi_dic/relationships.json")
)
rels_paths <- rels_paths[file.exists(rels_paths)]

dir_frozen <- here(glue("data/releases/{release_version}"))
dir.create(dir_frozen, recursive = TRUE, showWarnings = FALSE)
rels_merged_path <- file.path(dir_frozen, "relationships.json")

if (length(rels_paths) > 0) {
  merge_relationships_json(rels_paths, rels_merged_path)

  # apply merged PKs/FKs to dm
  d <- read_relationships_json(rels_merged_path, dm = d)

  # add cross-dataset FKs (bottle casts → ichthyo cruise/ship)
  d <- tryCatch(
    dm::dm_add_fk(d, casts, cruise_key, cruise),
    error = function(e) d
  )
  d <- tryCatch(dm::dm_add_fk(d, casts, ship_key, ship), error = function(e) d)
  d <- tryCatch(dm::dm_add_fk(d, casts, grid_key, grid), error = function(e) d)

  # re-write merged file with cross-dataset FKs included
  build_relationships_json(
    dm = d,
    output_dir = dir_frozen,
    provider = "calcofi",
    dataset = "calcofi-db-release"
  )
}

# set table colors by group
ichthyo_tbls <- intersect(
  c("cruise", "ship", "site", "tow", "net"),
  schema_tbls
)
species_tbls <- intersect(
  c("ichthyo", "species", "lookup", "taxon", "taxa_rank"),
  schema_tbls
)
spatial_tbls <- intersect(c("grid", "segment"), schema_tbls)
bottle_tbls <- intersect(
  c(
    "casts",
    "bottle",
    "bottle_measurement",
    "cast_condition",
    "measurement_type"
  ),
  schema_tbls
)
ctd_tbls <- intersect(
  c("ctd_cast", "ctd_measurement", "ctd_summary"),
  schema_tbls
)
dic_tbls <- intersect(
  c("dic_sample", "dic_measurement", "dic_measurement_summary"),
  schema_tbls
)
meta_tbls <- intersect(c("dataset"), schema_tbls)

if (length(ichthyo_tbls) > 0) {
  d <- d |> dm::dm_set_colors(lightblue = !!ichthyo_tbls)
}
if (length(species_tbls) > 0) {
  d <- d |> dm::dm_set_colors(lightyellow = !!species_tbls)
}
if (length(spatial_tbls) > 0) {
  d <- d |> dm::dm_set_colors(lightgreen = !!spatial_tbls)
}
if (length(bottle_tbls) > 0) {
  d <- d |> dm::dm_set_colors(pink = !!bottle_tbls)
}
if (length(ctd_tbls) > 0) {
  d <- d |> dm::dm_set_colors(lavender = !!ctd_tbls)
}
if (length(dic_tbls) > 0) {
  d <- d |> dm::dm_set_colors(lightsalmon = !!dic_tbls)
}
if (length(meta_tbls) > 0) {
  d <- d |> dm::dm_set_colors(white = !!meta_tbls)
}

dm::dm_draw(d, rankdir = "LR", view_type = "all")
```

## Create Frozen Release

Strip provenance columns and export clean parquet files for public
access. See
[Frozen DuckLake](https://ducklake.select/2025/10/24/frozen-ducklake/)
pattern.

```{r}
#| label: freeze_release

dir_frozen <- here(glue("data/releases/{release_version}"))
message(glue("Creating frozen release: {release_version}"))

# get tables to freeze (exclude internal and view tables)
tables_to_freeze <- schema_tbls

# create frozen parquet (strips provenance)
# partition large tables by cruise_key for efficient reads
freeze_stats <- write_parquet_outputs(
  con = con_wdl,
  output_dir = file.path(dir_frozen, "parquet"),
  tables = tables_to_freeze,
  partition_by = list(
    ctd_measurement = "cruise_key",
    ctd_summary = "cruise_key"
  ),
  strip_provenance = TRUE,
  compression = "zstd"
)

freeze_stats |>
  datatable(caption = glue("Frozen release {release_version} statistics"))
```

## Release Notes

```{r}
#| label: release_notes
#| results: asis

# build release notes
tables_list <- paste0(
  "- ",
  freeze_stats$table,
  " (",
  format(freeze_stats$rows, big.mark = ","),
  " rows)"
)

release_notes <- paste0(
  "# CalCOFI Database Release ",
  release_version,
  "\n\n",
  "**Release Date**: ",
  Sys.Date(),
  "\n\n",
  "## Tables Included\n\n",
  paste(tables_list, collapse = "\n"),
  "\n\n",
  "## Total\n\n",
  "- **Tables**: ",
  nrow(freeze_stats),
  "\n",
  "- **Total Rows**: ",
  format(sum(freeze_stats$rows), big.mark = ","),
  "\n",
  "- **Total Size**: ",
  round(sum(freeze_stats$file_size) / 1024 / 1024, 1),
  " MB\n\n",
  "## Data Sources\n\n",
  "- `ingest_swfsc_ichthyo.qmd` - Ichthyo tables (cruise, ship, site, tow, net, species, ichthyo, grid, segment, lookup, taxon, taxa_rank)\n",
  "- `ingest_calcofi_bottle.qmd` - Bottle/cast tables (casts, bottle, bottle_measurement, cast_condition, measurement_type)\n",
  "- `ingest_calcofi_ctd-cast.qmd` - CTD tables (ctd_cast, ctd_measurement, ctd_summary, measurement_type)\n",
  "- `ingest_calcofi_dic.qmd` - DIC/alkalinity tables (dic_sample, dic_measurement, dic_measurement_summary, dataset)\n\n",
  "## Cross-Dataset Integration\n\n",
  "- **Ship matching**: Reconciled ship codes between bottle casts and swfsc ship reference\n",
  "- **Cruise bridge**: Derived cruise_key (YYMMKK) for bottle casts via ship_key + datetime\n",
  "- **Taxonomy**: Standardized species with WoRMS AphiaID, ITIS TSN, GBIF backbone key\n",
  "- **Taxon hierarchy**: Built taxon + taxa_rank tables from WoRMS/ITIS classification\n\n",
  "## Access\n\n",
  "Parquet files can be queried directly from GCS:\n\n",
  "```r\n",
  "library(duckdb)\n",
  "con <- dbConnect(duckdb())\n",
  "dbExecute(con, 'INSTALL httpfs; LOAD httpfs;')\n",
  "dbGetQuery(con, \"\n",
  "  SELECT * FROM read_parquet(\n",
  "    'https://storage.googleapis.com/calcofi-db/ducklake/releases/",
  release_version,
  "/parquet/ichthyo.parquet')\n",
  "  LIMIT 10\")\n",
  "```\n\n",
  "Or use calcofi4r:\n\n",
  "```r\n",
  "library(calcofi4r)\n",
  "con <- cc_get_db(version = '",
  release_version,
  "')\n",
  "```\n"
)

writeLines(release_notes, file.path(dir_frozen, "RELEASE_NOTES.md"))
message(glue(
  "Release notes written to {file.path(dir_frozen, 'RELEASE_NOTES.md')}"
))
cat(release_notes)
```

## Upload Frozen Release to GCS

```{r}
#| label: upload_frozen

# upload frozen release to GCS
upload_frozen_release(
  release_dir = dir_frozen,
  version = release_version,
  set_latest = TRUE
)
```

## Cleanup

```{r}
#| label: cleanup

# close in-memory DuckDB connection
close_duckdb(con_wdl)
message("Assembly DuckDB connection closed")

# summary
message(glue("\n=== Summary ==="))
message(glue("Frozen release: {release_version} created at {dir_frozen}"))
message(glue("Tables: {nrow(freeze_stats)}"))
message(glue("Total rows: {format(sum(freeze_stats$rows), big.mark = ',')}"))
```

::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
