---
title: "Release CalCOFI Database"
calcofi:
  target_name: release_database
  workflow_type: release
  dependency:
    - auto
  output: data/releases
execute:
  echo: true
  message: true
  warning: true
editor_options:
  chunk_output_type: console
format:
  html:
    code-fold: true
editor:
  markdown:
    wrap: 72
---

## Overview {.unnumbered}

**Goal**: Create a frozen (immutable) release of the CalCOFI integrated
database by assembling all ingest parquet outputs. This is the "caboose"
notebook that always runs last, after all ingest notebooks complete.

**Upstream notebooks** are auto-discovered from `calcofi:` YAML
frontmatter in each `.qmd`. All workflows with `workflow_type: ingest`
or `spatial` feed into this release notebook via `dependency: [auto]`
in `_targets.R`.

```{mermaid}
%%| label: fig-workflow
%%| fig-cap: "Pipeline: ingest notebooks produce parquet → release caboose assembles frozen release"

flowchart LR
    subgraph ingest["Ingest Notebooks"]
        i1["ichthyo"]
        i2["bottle"]
        i3["ctd-cast"]
        i4["dic"]
        i5["inverts"]
        i6["spatial"]
    end
    subgraph assembly["Assembly (this notebook)"]
        a["Load all parquet<br/>into in-memory DuckDB"]
    end
    subgraph release["Frozen Release"]
        f["Clean parquet<br/>+ manifest.json"]
    end
    i1 --> a
    i2 --> a
    i3 --> a
    i4 --> a
    i5 --> a
    i6 --> a
    a --> f

    style ingest fill:#e3f2fd,stroke:#1565c0
    style assembly fill:#fff3e0,stroke:#ef6c00
    style release fill:#e8f4e8,stroke:#2e7d32
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
# cleanup_gcs_obsolete(dry_run = F)
librarian::shelf(
  CalCOFI / calcofi4db,
  CalCOFI / calcofi4r,
  DBI,
  dplyr,
  DT,
  fs,
  glue,
  here,
  jsonlite,
  purrr,
  tibble,
  quiet = T
)
options(DT.options = list(scrollX = TRUE))

# release version
release_version <- format(Sys.Date(), "v%Y.%m.%d")
message(glue("Release version: {release_version}"))
```

## Assemble from Ingest Outputs

Create VIEWs on local parquet files from each ingest (zero-copy).
For tables appearing in multiple ingests, use the canonical (first) source.

```{r}
#| label: assemble_working

con_wdl <- get_duckdb_con(":memory:")
load_duckdb_extension(con_wdl, "spatial")

# auto-discover table registry from all ingest manifests
registry <- build_release_table_registry(here())

# use only canonical, non-supplemental tables
reg_canon <- registry |>
  filter(canonical, !supplemental)

message(glue(
  "{nrow(reg_canon)} canonical tables from ",
  "{length(unique(reg_canon$ingest))} ingests"))

# create VIEWs on local parquet for each canonical table
# _new delta tables handled separately for merging
all_geom_tables <- c("grid", "site", "segment", "casts", "ctd_cast", "_spatial")
main_tables <- reg_canon |> filter(!grepl("_new$", table))
new_tables  <- registry |> filter(grepl("_new$", table))

load_stats <- purrr::map_dfr(
  split(main_tables, seq_len(nrow(main_tables))),
  function(row) {
    load_prior_tables(
      con         = con_wdl,
      parquet_dir = row$parquet_dir,
      tables      = row$table,
      geom_tables = all_geom_tables,
      as_view     = TRUE
    )
  })

# merge {table}_new additions into their base tables
# driven by calcofi.modifies in YAML frontmatter
if (nrow(new_tables) > 0) {
  # group _new tables by their base table
  base_names <- unique(sub("_new$", "", new_tables$table))

  for (base_tbl in base_names) {
    delta_rows <- new_tables |> filter(table == paste0(base_tbl, "_new"))

    # replace VIEW with TABLE for this base table (so we can INSERT)
    base_src <- main_tables |> filter(table == base_tbl)
    if (nrow(base_src) > 0) {
      dbExecute(con_wdl, glue("DROP VIEW IF EXISTS {base_tbl}"))
      load_prior_tables(
        con = con_wdl, parquet_dir = base_src$parquet_dir[1],
        tables = base_tbl, geom_tables = all_geom_tables)

      # get PK column for dedup
      pk_col <- dbGetQuery(con_wdl, glue(
        "SELECT column_name FROM information_schema.columns
         WHERE table_name = '{base_tbl}'
         ORDER BY ordinal_position LIMIT 1"))$column_name

      for (j in seq_len(nrow(delta_rows))) {
        dr <- delta_rows[j, ]
        pq_path <- file.path(dr$parquet_dir, paste0(base_tbl, "_new.parquet"))
        if (file.exists(pq_path)) {
          dbExecute(con_wdl, glue(
            "INSERT INTO {base_tbl}
             SELECT * FROM read_parquet('{pq_path}')
             WHERE {pk_col} NOT IN (SELECT {pk_col} FROM {base_tbl})"))
          n_new <- dbGetQuery(con_wdl, glue(
            "SELECT COUNT(*) AS n FROM read_parquet('{pq_path}')"))$n
          message(glue("Merged {n_new} {base_tbl} addition(s) from {dr$ingest}"))
        }
      }
    }
  }
}

load_stats |>
  datatable(caption = "Assembled tables (VIEWs on local parquet)")
```

## Scan Manifests for Mismatches

```{r}
#| label: scan_mismatches

# scan all ingest manifests for unresolved mismatches
all_manifests <- list.files(
  "data/parquet", "manifest.json",
  recursive = TRUE, full.names = TRUE)

all_mismatches <- purrr::compact(lapply(all_manifests, function(mf) {
  m <- jsonlite::read_json(mf)
  if (is.null(m$mismatches)) return(NULL)
  dataset <- basename(dirname(mf))
  purrr::imap_dfr(m$mismatches, function(items, category) {
    if (length(items) == 0) return(NULL)
    purrr::map_dfr(items, function(x) {
      # replace NULL values with NA so as_tibble works
      x[vapply(x, is.null, logical(1))] <- NA
      as_tibble(x)
    }) |>
      mutate(dataset = dataset, category = category, .before = 1)
  })
}))

if (length(all_mismatches) > 0) {
  d_mismatches <- bind_rows(all_mismatches)
  message(glue("{nrow(d_mismatches)} unresolved mismatch(es) across manifests"))
  d_mismatches |>
    datatable(caption = "Unresolved mismatches (from manifest.json)")
} else {
  message("No unresolved mismatches found across manifests")
}
```

## Validate

Cross-dataset validation to ensure data integrity before freezing.

```{r}
#| label: validate

# grid_key integrity: casts.grid_key should all be in grid.grid_key
tbls <- DBI::dbListTables(con_wdl)

if (all(c("casts", "grid") %in% tbls)) {
  # use information_schema to check columns (avoids GEOMETRY type issues)
  casts_cols_wdl <- dbGetQuery(
    con_wdl,
    "SELECT column_name FROM information_schema.columns
     WHERE table_name = 'casts'"
  )$column_name
  grid_cols_wdl <- dbGetQuery(
    con_wdl,
    "SELECT column_name FROM information_schema.columns
     WHERE table_name = 'grid'"
  )$column_name
  if ("grid_key" %in% casts_cols_wdl && "grid_key" %in% grid_cols_wdl) {
    grid_orphans <- dbGetQuery(
      con_wdl,
      "SELECT COUNT(*) AS n FROM casts c
       WHERE c.grid_key IS NOT NULL
         AND c.grid_key NOT IN (SELECT grid_key FROM grid)"
    )$n
    message(glue("Grid key orphans in casts: {grid_orphans}"))
    # Grid key orphans in casts: 0
  }
}

# ship PK uniqueness
if ("ship" %in% tbls) {
  ship_dups <- dbGetQuery(
    con_wdl,
    "SELECT ship_key, COUNT(*) AS n FROM ship
     GROUP BY ship_key HAVING COUNT(*) > 1"
  )
  if (nrow(ship_dups) > 0) {
    warning(glue("Duplicate ship_key values: {nrow(ship_dups)}"))
  } else {
    message("ship_key: all unique")
  }
}
# ship_key: all unique

# cruise PK uniqueness
if ("cruise" %in% tbls) {
  cruise_dups <- dbGetQuery(
    con_wdl,
    "SELECT cruise_key, COUNT(*) AS n FROM cruise
     GROUP BY cruise_key HAVING COUNT(*) > 1"
  )
  if (nrow(cruise_dups) > 0) {
    warning(glue("Duplicate cruise_key values: {nrow(cruise_dups)}"))
  } else {
    message("cruise_key: all unique")
  }
}
# cruise_key: all unique

# cruise bridge coverage
if ("casts" %in% tbls) {
  bridge_stats <- dbGetQuery(
    con_wdl,
    "SELECT
      COUNT(*) AS total_casts,
      SUM(CASE WHEN ship_key IS NOT NULL THEN 1 ELSE 0 END) AS with_ship_key,
      SUM(CASE WHEN cruise_key IS NOT NULL THEN 1 ELSE 0 END) AS with_cruise_key
    FROM casts"
  )
  bridge_stats |> datatable(caption = "Cruise bridge coverage")
}

# cruise_key format validation (YYYY-MM-NODC)
if ("cruise" %in% tbls) {
  bad_ck <- dbGetQuery(
    con_wdl,
    "SELECT cruise_key FROM cruise
     WHERE cruise_key IS NOT NULL
       AND NOT regexp_matches(cruise_key, '^\\d{4}-\\d{2}-.+$')"
  )
  if (nrow(bad_ck) > 0) {
    warning(glue("cruise_key format violations: {nrow(bad_ck)} rows"))
  } else {
    message("cruise_key: all match YYYY-MM-NODC format")
  }
}
# Warning message: cruise_key format violations: 1 rows
# cruise_key: 2019-07-

# site_key format validation (NNN.N NNN.N)
for (tbl_name in intersect(c("site", "casts", "ctd_cast"), tbls)) {
  tbl_cols <- dbGetQuery(
    con_wdl,
    glue(
      "SELECT column_name FROM information_schema.columns
     WHERE table_name = '{tbl_name}'"
    )
  )$column_name
  if ("site_key" %in% tbl_cols) {
    bad_sk <- dbGetQuery(
      con_wdl,
      glue(
        "SELECT COUNT(*) AS n FROM {tbl_name}
         WHERE site_key IS NOT NULL
           AND NOT regexp_matches(site_key, '^\\d{{3}}\\.\\d \\d{{3}}\\.\\d$')"
      )
    )$n
    if (bad_sk > 0) {
      warning(glue("site_key format violations in {tbl_name}: {bad_sk} rows"))
    } else {
      message(glue("site_key in {tbl_name}: all match NNN.N NNN.N format"))
    }
  }
}
# site_key in casts: all match NNN.N NNN.N format
# Warning message: site_key format violations in site: 982 rows

# cruise_summary table for Shiny app
if (all(c("cruise", "ship") %in% tbls)) {
  # detect site_key vs sta_key for backward compat with old parquets
  ctd_site_col <- ifelse(
    "site_key" %in%
      dbGetQuery(
        con_wdl,
        "SELECT column_name FROM information_schema.columns
       WHERE table_name = 'ctd_cast'"
      )$column_name,
    "site_key",
    "sta_key"
  )

  dbExecute(
    con_wdl,
    glue(
      "
    CREATE OR REPLACE TABLE cruise_summary AS
    SELECT
      cr.cruise_key,
      EXTRACT(YEAR FROM cr.date_ym)::INTEGER  AS year,
      EXTRACT(MONTH FROM cr.date_ym)::INTEGER AS month,
      s.ship_name,
      s.ship_nodc,
      COALESCE((SELECT COUNT(DISTINCT site_key) FROM site
                WHERE cruise_key = cr.cruise_key), 0) AS ichthyo,
      COALESCE((SELECT COUNT(DISTINCT site_key) FROM casts
                WHERE cruise_key = cr.cruise_key), 0) AS bottle,
      COALESCE((SELECT COUNT(DISTINCT {ctd_site_col}) FROM ctd_cast
                WHERE cruise_key = cr.cruise_key), 0) AS ctd_cast,
      COALESCE((SELECT COUNT(DISTINCT ds.site_key)
                FROM dic_sample ds
                JOIN casts c ON ds.cast_id = c.cast_id
                WHERE c.cruise_key = cr.cruise_key), 0) AS dic
    FROM cruise cr
    JOIN ship s ON cr.ship_key = s.ship_key
    ORDER BY year DESC, month DESC"
    )
  )
  n_cs <- dbGetQuery(con_wdl, "SELECT COUNT(*) AS n FROM cruise_summary")$n
  message(glue("Created cruise_summary table: {n_cs} rows"))
}
# Created cruise_summary table: 691 rows
tbl(con_wdl, "cruise_summary") |>
  collect() |>
  datatable(caption = "cruise_summary preview")

# run standard release validation (wrapped in tryCatch for GEOMETRY compat)
tryCatch(
  {
    validation <- validate_for_release(con_wdl)
    if (validation$passed) {
      message("Release validation passed!")
    } else {
      cat("Validation FAILED:\n")
      cat(paste("-", validation$errors, collapse = "\n"))
    }
  },
  error = function(e) {
    message(glue("validate_for_release skipped: {e$message}"))
  }
)
```

## Show Combined Schema

```{r}

# dir_frozen used later; define early so ERD can reference it
dir_frozen <- here(glue("data/releases/{release_version}"))
dir.create(dir_frozen, recursive = TRUE, showWarnings = FALSE)

erd <- cc_erd(con_wdl, layout = "elk")
plot(erd)

erd <- cc_erd(
  con_wdl,
  colors = list(
    lightblue = c("cruise", "ship", "site", "tow", "net"),
    lightyellow = c("ichthyo", "species", "lookup", "taxon", "taxa_rank"),
    lightgreen = c("grid", "segment"),
    pink = c(
      "casts",
      "bottle",
      "bottle_measurement",
      "cast_condition",
      "measurement_type"
    ),
    lavender = c("ctd_cast", "ctd_measurement", "ctd_summary"),
    lightsalmon = c("dic_sample", "dic_measurement", "dic_summary"),
    white = c("dataset")
  )
)
plot(erd)

```

```{r}
#| label: combined_schema
#| fig-width: 12
#| fig-height: 10

# exclude internal tables and views
schema_tbls <- setdiff(
  tbls,
  c("_meta", "_sp_update", "casts_derived", "ctd_cast_derived"))

# merge per-dataset relationships.json files
rels_paths <- c(
  here("data/parquet/swfsc_ichthyo/relationships.json"),
  here("data/parquet/calcofi_bottle/relationships.json"),
  here("data/parquet/calcofi_ctd-cast/relationships.json"),
  here("data/parquet/calcofi_dic/relationships.json")
)
rels_paths <- rels_paths[file.exists(rels_paths)]

dir_frozen <- here(glue("data/releases/{release_version}"))
dir.create(dir_frozen, recursive = TRUE, showWarnings = FALSE)
rels_merged_path <- file.path(dir_frozen, "relationships.json")

if (length(rels_paths) > 0) {
  merge_relationships_json(rels_paths, rels_merged_path)

  # add cross-dataset FKs (bottle casts → ichthyo cruise/ship/grid)
  rels_merged <- jsonlite::fromJSON(
    rels_merged_path, simplifyVector = FALSE)
  cross_fks <- list(
    list(table = "casts", column = "cruise_key",
         ref_table = "cruise", ref_column = "cruise_key"),
    list(table = "casts", column = "ship_key",
         ref_table = "ship", ref_column = "ship_key"),
    list(table = "casts", column = "grid_key",
         ref_table = "grid", ref_column = "grid_key"))
  rels_merged$foreign_keys <- c(
    rels_merged$foreign_keys, cross_fks)
  jsonlite::write_json(
    rels_merged, rels_merged_path,
    auto_unbox = TRUE, pretty = TRUE, null = "null")
}

# render color-coded ERD (cc_erd handles GEOMETRY columns natively)
cc_erd(
  con_wdl,
  tables   = schema_tbls,
  rels_path = rels_merged_path,
  colors = list(
    lightblue   = c("cruise", "ship", "site", "tow", "net"),
    lightyellow = c("ichthyo", "species", "lookup",
                    "taxon", "taxa_rank"),
    lightgreen  = c("grid", "segment"),
    pink        = c("casts", "bottle", "bottle_measurement",
                    "cast_condition", "measurement_type"),
    lavender    = c("ctd_cast", "ctd_measurement", "ctd_summary"),
    lightsalmon = c("dic_sample", "dic_measurement",
                    "dic_summary"),
    lightcyan   = c("_spatial", "_spatial_attr"),
    white       = "dataset"))
```

## Create Frozen Release

Strip provenance columns and export clean parquet files for public
access. See
[Frozen DuckLake](https://ducklake.select/2025/10/24/frozen-ducklake/)
pattern.

```{r}
#| label: freeze_release

dir_frozen <- here(glue("data/releases/{release_version}"))
dir_frozen_pq <- file.path(dir_frozen, "parquet")
dir.create(dir_frozen_pq, recursive = TRUE, showWarnings = FALSE)
message(glue("Creating frozen release: {release_version}"))

# only cruise_summary is derived in this notebook — export locally
# all other tables are GCS-copied from ingest/ (including provenance columns)
derived_tables <- "cruise_summary"
if (nrow(new_tables) > 0) {
  # tables with _new additions need local merge + export
  merged_base <- unique(sub("_new$", "", new_tables$table))
  derived_tables <- c(derived_tables, merged_base)
}

# export only derived/merged tables to local parquet
export_parquet(con_wdl, "cruise_summary",
  file.path(dir_frozen_pq, "cruise_summary.parquet"), compression = "zstd")
message("Exported cruise_summary.parquet")

# export merged tables (e.g., ship with _new additions)
for (tbl in setdiff(derived_tables, "cruise_summary")) {
  export_parquet(con_wdl, tbl,
    file.path(dir_frozen_pq, paste0(tbl, ".parquet")), compression = "zstd")
  message(glue("Exported {tbl}.parquet (merged)"))
}

# build freeze stats from registry (auto-discovered)
# exclude _new delta tables (intermediate) and supplemental
freeze_stats <- reg_canon |>
  filter(!supplemental, !grepl("_new$", table)) |>
  select(table, rows, partitioned, gcs_prefix)

# merged tables (from _new additions) → mark as derived (gcs_prefix = NA → upload from local)
if (nrow(new_tables) > 0) {
  merged_base <- unique(sub("_new$", "", new_tables$table))
  freeze_stats <- freeze_stats |>
    mutate(gcs_prefix = if_else(table %in% merged_base, NA_character_, gcs_prefix))
}

# add derived tables (cruise_summary, etc.)
for (dt in derived_tables) {
  if (!dt %in% freeze_stats$table) {
    n <- dbGetQuery(con_wdl, glue("SELECT COUNT(*) AS n FROM {dt}"))$n
    freeze_stats <- freeze_stats |>
      bind_rows(tibble(
        table = dt, rows = n, partitioned = FALSE, gcs_prefix = NA_character_))
  }
}

freeze_stats |>
  datatable(caption = glue("Frozen release {release_version} — {nrow(freeze_stats)} tables"))
```

## Release Notes

```{r}
#| label: release_notes
#| results: asis

# build release notes
tables_list <- paste0(
  "- ",
  freeze_stats$table,
  " (",
  format(freeze_stats$rows, big.mark = ","),
  " rows)"
)

release_notes <- paste0(
  "# CalCOFI Database Release ",
  release_version,
  "\n\n",
  "**Release Date**: ",
  Sys.Date(),
  "\n\n",
  "## Tables Included\n\n",
  paste(tables_list, collapse = "\n"),
  "\n\n",
  "## Total\n\n",
  "- **Tables**: ",
  nrow(freeze_stats),
  "\n",
  "- **Total Rows**: ",
  format(sum(freeze_stats$rows, na.rm = TRUE), big.mark = ","),
  "\n\n",
  "## Data Sources\n\n",
  "- `ingest_swfsc_ichthyo.qmd` - Ichthyo tables (cruise, ship, site, tow, net, species, ichthyo, grid, segment, lookup, taxon, taxa_rank)\n",
  "- `ingest_calcofi_bottle.qmd` - Bottle/cast tables (casts, bottle, bottle_measurement, cast_condition, measurement_type)\n",
  "- `ingest_calcofi_ctd-cast.qmd` - CTD tables (ctd_cast, ctd_measurement, ctd_summary, measurement_type)\n",
  "- `ingest_calcofi_dic.qmd` - DIC/alkalinity tables (dic_sample, dic_measurement, dic_summary, dataset)\n\n",
  "## Cross-Dataset Integration\n\n",
  "- **Ship matching**: Reconciled ship codes between bottle casts and swfsc ship reference\n",
  "- **Cruise bridge**: Derived cruise_key (YYYY-MM-NODC) for bottle casts via ship matching + datetime\n",
  "- **Taxonomy**: Standardized species with WoRMS AphiaID, ITIS TSN, GBIF backbone key\n",
  "- **Taxon hierarchy**: Built taxon + taxa_rank tables from WoRMS/ITIS classification\n\n",
  "## Access\n\n",
  "Parquet files can be queried directly from GCS:\n\n",
  "```r\n",
  "library(duckdb)\n",
  "con <- dbConnect(duckdb())\n",
  "dbExecute(con, 'INSTALL httpfs; LOAD httpfs;')\n",
  "dbGetQuery(con, \"\n",
  "  SELECT * FROM read_parquet(\n",
  "    'https://storage.googleapis.com/calcofi-db/ducklake/releases/",
  release_version,
  "/parquet/ichthyo.parquet')\n",
  "  LIMIT 10\")\n",
  "```\n\n",
  "Or use calcofi4r:\n\n",
  "```r\n",
  "library(calcofi4r)\n",
  "con <- cc_get_db(version = '",
  release_version,
  "')\n",
  "```\n"
)

writeLines(release_notes, file.path(dir_frozen, "RELEASE_NOTES.md"))
message(glue(
  "Release notes written to {file.path(dir_frozen, 'RELEASE_NOTES.md')}"
))
cat(release_notes)
```

## Upload Frozen Release to GCS

```{r}
#| label: upload_frozen

gcs_bucket  <- "calcofi-db"
gcs_release <- glue("ducklake/releases/{release_version}")
gcloud      <- find_gcloud()

# 1. GCS server-side copy for ingest tables (auto-discovered from registry)
copy_rows <- freeze_stats |> filter(!is.na(gcs_prefix))
message(glue("Copying {nrow(copy_rows)} tables from ingest/ to releases/ on GCS..."))

for (i in seq_len(nrow(copy_rows))) {
  tbl  <- copy_rows$table[i]
  pfx  <- copy_rows$gcs_prefix[i]
  part <- copy_rows$partitioned[i]

  if (part) {
    # partitioned: copy directory
    src <- glue("gs://{gcs_bucket}/{pfx}/{tbl}")
    dst <- glue("gs://{gcs_bucket}/{gcs_release}/parquet/{tbl}")
    system2(gcloud, c("storage", "cp", "-r",
      paste0(src, "/*"), dst), stdout = TRUE, stderr = TRUE)
  } else {
    src <- glue("gs://{gcs_bucket}/{pfx}/{tbl}.parquet")
    dst <- glue("gs://{gcs_bucket}/{gcs_release}/parquet/{tbl}.parquet")
    system2(gcloud, c("storage", "cp",
      src, dst), stdout = TRUE, stderr = TRUE)
  }
  message(glue("  {tbl}: copied from {pfx}"))
}

# 2. upload derived tables from local parquet (cruise_summary)
derived_local <- list.files(dir_frozen_pq, pattern = "[.]parquet$",
                            full.names = TRUE)
for (pq in derived_local) {
  tbl <- tools::file_path_sans_ext(basename(pq))
  gcs_path <- glue("gs://{gcs_bucket}/{gcs_release}/parquet/{tbl}.parquet")
  put_gcs_file(pq, gcs_path)
  message(glue("  {tbl}: uploaded (derived)"))
}

# 3. build and upload catalog.json (needed by cc_get_db())
tables_df <- freeze_stats |>
  select(name = table, rows, partitioned)

catalog <- list(
  version      = release_version,
  release_date = as.character(Sys.Date()),
  total_rows   = sum(tables_df$rows, na.rm = TRUE),
  total_size   = 0,
  tables       = tables_df)
catalog_path <- file.path(dir_frozen, "catalog.json")
jsonlite::write_json(catalog, catalog_path, auto_unbox = TRUE, pretty = TRUE)
put_gcs_file(catalog_path,
  glue("gs://{gcs_bucket}/{gcs_release}/catalog.json"))

# upload RELEASE_NOTES.md
notes_path <- file.path(dir_frozen, "RELEASE_NOTES.md")
if (file.exists(notes_path))
  put_gcs_file(notes_path,
    glue("gs://{gcs_bucket}/{gcs_release}/RELEASE_NOTES.md"))

# upload relationships.json
rels_json <- file.path(dir_frozen, "relationships.json")
if (file.exists(rels_json))
  put_gcs_file(rels_json,
    glue("gs://{gcs_bucket}/{gcs_release}/relationships.json"))

# 4. update versions.json and latest.txt
# discover all releases from GCS and rebuild versions.json
gcs_ls <- system2(gcloud, c("storage", "ls",
  glue("gs://{gcs_bucket}/ducklake/releases/")),
  stdout = TRUE, stderr = TRUE)
release_vers <- regmatches(gcs_ls,
  regexpr("v[0-9]{4}[.][0-9]{2}[.]*[0-9]*", gcs_ls))
https_base <- glue("https://storage.googleapis.com/{gcs_bucket}/ducklake/releases")

all_versions <- purrr::compact(lapply(release_vers, function(v) {
  tryCatch({
    cat_data <- jsonlite::fromJSON(glue("{https_base}/{v}/catalog.json"))
    list(
      version      = cat_data$version,
      release_date = cat_data$release_date %||% NA_character_,
      tables       = if (is.data.frame(cat_data$tables)) nrow(cat_data$tables)
                     else length(cat_data$tables),
      total_rows   = as.integer(cat_data$total_rows %||% 0),
      size_mb      = round((cat_data$total_size %||% 0) / 1024 / 1024, 1))
  }, error = function(e) NULL)
}))
all_versions <- all_versions[order(
  sapply(all_versions, `[[`, "version"), decreasing = TRUE)]

versions_local <- tempfile(fileext = ".json")
jsonlite::write_json(list(versions = all_versions), versions_local,
  auto_unbox = TRUE, pretty = TRUE)
put_gcs_file(versions_local,
  glue("gs://{gcs_bucket}/ducklake/releases/versions.json"))

# update latest.txt
latest_local <- tempfile()
writeLines(release_version, latest_local)
put_gcs_file(latest_local,
  glue("gs://{gcs_bucket}/ducklake/releases/latest.txt"))

message(glue(
  "Release {release_version} published ({length(all_versions)} versions tracked)"))
```

## Cleanup

```{r}
#| label: cleanup

# close in-memory DuckDB connection
close_duckdb(con_wdl)
message("Assembly DuckDB connection closed")

# summary
message(glue("\n=== Summary ==="))
message(glue("Frozen release: {release_version} created at {dir_frozen}"))
message(glue("Tables: {nrow(freeze_stats)}"))
message(glue("Total rows: {format(sum(freeze_stats$rows, na.rm = TRUE), big.mark = ',')}"))
```

::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
