---
title: "Sync Google Drive to GCS"
editor_options:
  chunk_output_type: console
---

## Overview

Mirrors the entire Google Drive `data-public/` folder to
`gs://calcofi-files-public/_sync/`. Uses hash-based comparison
(CRC32C) to skip unchanged files and logs all actions to GCS.

This is a standalone utility — not a pipeline target. Run manually
or schedule as a cron job for daily backup.

**Buckets**:

- `calcofi-files-public/_sync/` — living mirror of GD (this notebook)
- `calcofi-files-public/archive/` — immutable timestamped snapshots
  (created by ingest QMDs)

The `_sync/` mirror deletes stale GCS files, but `archive/` provides
the reproducibility safety net — nothing is lost permanently.

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))

librarian::shelf(
  DT,
  fs,
  glue,
  here,
  quiet = T
)
options(DT.options = list(scrollX = TRUE))

gd_dir     <- path_expand("~/My Drive/projects/calcofi/data-public")
gcs_bucket <- "calcofi-files-public"
gcs_prefix <- "_sync"
```

## Scan Google Drive

```{r}
#| label: scan_gd

gd_files <- dir_ls(gd_dir, recurse = TRUE, type = "file")
gd_rel   <- sub(paste0("^", normalizePath(gd_dir), "/?"), "", gd_files)

cat(glue("{length(gd_files)} files in Google Drive data-public/"), "\n")
cat(glue("Total size: {sum(file_size(gd_files))}"), "\n\n")

# summary by top-level folder
tibble::tibble(path = gd_rel) |>
  dplyr::mutate(folder = sub("/.*", "", path)) |>
  dplyr::count(folder, name = "files") |>
  dplyr::arrange(dplyr::desc(files)) |>
  datatable(caption = "Files per top-level folder in Google Drive")
```

## Sync to GCS

```{r}
#| label: sync

results <- sync_to_gcs(
  local_dir    = gd_dir,
  gcs_prefix   = gcs_prefix,
  bucket       = gcs_bucket,
  delete_stale = TRUE,
  log_to_gcs   = TRUE,
  exclude      = c(".DS_Store", "*.tmp", "~$*"))
```

## Summary

```{r}
#| label: summary

results |>
  dplyr::count(action, name = "files") |>
  datatable(caption = "Sync action summary")

# show non-skipped actions
results |>
  dplyr::filter(action != "skipped") |>
  dplyr::mutate(size = fs::as_fs_bytes(size)) |>
  datatable(
    caption = "Files uploaded or deleted",
    rownames = FALSE,
    filter = "top")
```
