---
title: "Test Release — gate latest.txt on query suite"
calcofi:
  target_name: test_release
  workflow_type: test
  dependency:
    - release_database
  output: _output/test_release.html
execute:
  echo: true
  message: true
  warning: true
editor_options:
  chunk_output_type: console
format:
  html:
    code-fold: true
editor:
  markdown:
    wrap: 72
---

## Overview {.unnumbered}

**Goal**: Promote a freshly-built release to `latest.txt` only when every
pre-baked query in [CalCOFI/db-query](https://github.com/CalCOFI/db-query) runs
cleanly against it.

Pipeline position: this notebook runs after `release_database.qmd` has
uploaded the parquet, `metadata.json`, `relationships.json`, `erd.mmd`, and
`catalog.json` to `gs://calcofi-db/ducklake/releases/{release_version}/`,
but **before** `latest.txt` is updated. Every `.md` query in the
[`_queries/`](https://github.com/CalCOFI/db-query/tree/main/_queries) folder
is rendered with its YAML defaults, executed against the just-uploaded
release on GCS, and reported below. If every query returns without error,
`latest.txt` is written; otherwise the target fails and consumer apps
keep pointing at the previously-promoted version.

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
librarian::shelf(
  CalCOFI / calcofi4db,
  DBI,
  duckdb,
  dplyr,
  DT,
  fs,
  glue,
  here,
  jsonlite,
  purrr,
  tibble,
  yaml,
  quiet = T)

# which release to test? read the just-uploaded version from the local
# release directory (most recent date-stamped folder under data/releases/)
releases_dir <- here("data/releases")
release_dirs <- list.dirs(releases_dir, recursive = FALSE)
release_dirs <- release_dirs[grepl("v[0-9]{4}[.][0-9]{2}[.]?[0-9]*$",
                                   basename(release_dirs))]
stopifnot(length(release_dirs) > 0)
release_version <- basename(release_dirs[which.max(file.mtime(release_dirs))])
message(glue("Testing release: {release_version}"))

# where the pre-baked queries live
queries_dir <- here("../db-query/_queries")
if (!dir.exists(queries_dir)) {
  stop(glue(
    "Could not find {queries_dir}. Clone CalCOFI/db-query as a sibling of ",
    "this repo, or set queries_dir manually."))
}
message(glue("Query repo: {queries_dir}"))

# fresh in-memory DuckDB with httpfs + spatial — every query reads parquet
# from GCS over httpfs, mirroring what the query app's DuckDB-WASM does
con_test <- get_duckdb_con(":memory:")
load_duckdb_extension(con_test, "httpfs")
load_duckdb_extension(con_test, "spatial")
```

## Render the query templates

The query app uses Handlebars at runtime. For tests we replicate the
subset of Handlebars features actually used in `_queries/*.md`:

- `{{var}}` — simple substitution
- `{{{var}}}` — raw substitution (no escaping; same effect here)
- `{{sqlesc var}}` — escape single quotes for SQL string literals
- `{{#if var}}...{{/if}}` — include block when `var` is truthy

```{r}
#| label: render_helpers

sqlesc <- function(x) gsub("'", "''", as.character(x %||% ""), fixed = TRUE)

# truthy test mirrors Handlebars: non-NULL, non-NA, non-empty string,
# non-zero number, non-FALSE logical
is_truthy <- function(val) {
  if (is.null(val))           return(FALSE)
  if (length(val) == 0)       return(FALSE)
  if (is.logical(val))        return(isTRUE(val))
  if (is.character(val))      return(any(!is.na(val) & nzchar(val)))
  if (is.numeric(val))        return(any(!is.na(val) & val != 0))
  !is.na(val[[1]])
}

# resolve innermost {{#if X}}...{{/if}} repeatedly until none remain.
# [\\s\\S] (not .) so the body spans newlines; (?!\\{\\{#if) guard forces
# "innermost only" so we strip from the inside out.
render_if_blocks <- function(txt, params) {
  pat <- "\\{\\{#if\\s+([A-Za-z_][A-Za-z0-9_]*)\\}\\}((?:(?!\\{\\{#if)[\\s\\S])*?)\\{\\{/if\\}\\}"
  repeat {
    m <- regexpr(pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    parts <- regmatches(full, regexec(pat, full, perl = TRUE))[[1]]
    var_name <- parts[2]
    body     <- parts[3]
    replacement <- if (is_truthy(params[[var_name]])) body else ""
    txt <- paste0(
      substr(txt, 1, m - 1),
      replacement,
      substr(txt, m + attr(m, "match.length"), nchar(txt)))
  }
  txt
}

render_template <- function(template, params) {
  txt <- template

  # 1. {{#if X}}...{{/if}} blocks
  txt <- render_if_blocks(txt, params)

  # 2. {{sqlesc VAR}}
  helper_pat <- "\\{\\{\\s*sqlesc\\s+([A-Za-z_][A-Za-z0-9_]*)\\s*\\}\\}"
  repeat {
    m <- regexpr(helper_pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    name <- sub(helper_pat, "\\1", full, perl = TRUE)
    txt <- sub(full, sqlesc(params[[name]]), txt, fixed = TRUE)
  }

  # 3. {{{VAR}}} (triple-brace, raw) and {{VAR}} (simple) — handle the
  #    triple-brace first so the {{VAR}} pass doesn't eat the inner pair
  triple_pat <- "\\{\\{\\{\\s*([A-Za-z_][A-Za-z0-9_]*)\\s*\\}\\}\\}"
  repeat {
    m <- regexpr(triple_pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    name <- sub(triple_pat, "\\1", full, perl = TRUE)
    val <- as.character(params[[name]] %||% "")
    txt <- sub(full, val, txt, fixed = TRUE)
  }
  single_pat <- "\\{\\{\\s*([A-Za-z_][A-Za-z0-9_]*)\\s*\\}\\}"
  repeat {
    m <- regexpr(single_pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    name <- sub(single_pat, "\\1", full, perl = TRUE)
    val <- as.character(params[[name]] %||% "")
    txt <- sub(full, val, txt, fixed = TRUE)
  }

  # query-app convention: literal __VERSION__ placeholder -> release version,
  # mirroring _includes/form-field.html ({% replace "__VERSION__",
  # site.default_version %}). Used in free-form defaults like sql-shell/shell.md.
  if (!is.null(params$version)) {
    txt <- gsub("__VERSION__", as.character(params$version), txt, fixed = TRUE)
  }

  txt
}

# parse YAML front-matter + body from a query .md file
parse_query_md <- function(path) {
  raw <- readLines(path, warn = FALSE)
  delims <- which(trimws(raw) == "---")
  if (length(delims) < 2) {
    stop(glue("Missing YAML delimiters in {path}"))
  }
  yaml_text <- paste(raw[(delims[1] + 1):(delims[2] - 1)], collapse = "\n")
  meta <- yaml::yaml.load(yaml_text)
  body <- paste(raw[(delims[2] + 1):length(raw)], collapse = "\n")
  list(meta = meta, body = body, path = path)
}

# build the params list for a query: each parameter's `default`, with
# `version` overridden to the release under test
defaults_for <- function(meta, release_version) {
  params <- list(version = release_version)
  if (!is.null(meta$parameters)) {
    for (name in names(meta$parameters)) {
      spec <- meta$parameters[[name]]
      params[[name]] <- spec$default
    }
  }
  params$version <- release_version
  params
}
```

## Discover and execute every query

```{r}
#| label: execute

query_files <- list.files(queries_dir, pattern = "[.]md$",
                          recursive = TRUE, full.names = TRUE)
message(glue("Found {length(query_files)} query files"))

results <- purrr::map_dfr(query_files, function(qpath) {
  rel  <- sub(paste0(normalizePath(queries_dir), "/"), "",
              normalizePath(qpath), fixed = TRUE)
  q    <- parse_query_md(qpath)
  meta <- q$meta

  # sql_builder queries are JS-only — flag them; the query app exercises
  # them in the browser
  if (!is.null(meta$sql_builder)) {
    return(tibble::tibble(
      query    = rel,
      label    = meta$label %||% NA_character_,
      status   = "skip",
      reason   = paste0("sql_builder=", meta$sql_builder),
      rows     = NA_integer_,
      ms       = NA_real_,
      error    = NA_character_))
  }

  if (is.null(meta$sql)) {
    return(tibble::tibble(
      query = rel, label = meta$label %||% NA_character_,
      status = "skip", reason = "no inline sql",
      rows = NA_integer_, ms = NA_real_, error = NA_character_))
  }

  params <- defaults_for(meta, release_version)
  sql    <- render_template(meta$sql, params)

  t0  <- Sys.time()
  out <- tryCatch(
    DBI::dbGetQuery(con_test, sql),
    error = function(e) e)
  elapsed_ms <- as.numeric(difftime(Sys.time(), t0, units = "secs")) * 1000

  if (inherits(out, "error")) {
    tibble::tibble(
      query = rel, label = meta$label %||% NA_character_,
      status = "fail", reason = NA_character_,
      rows   = NA_integer_, ms = round(elapsed_ms, 1),
      error  = conditionMessage(out))
  } else {
    tibble::tibble(
      query = rel, label = meta$label %||% NA_character_,
      status = "pass", reason = NA_character_,
      rows   = nrow(out), ms = round(elapsed_ms, 1),
      error  = NA_character_)
  }
})

# status badges for the report
status_badge <- function(s) {
  color <- c(pass = "#2e7d32", fail = "#c62828", skip = "#757575")[s]
  glue("<span style='background:{color};color:#fff;padding:2px 8px;",
       "border-radius:10px;font-size:11px;'>{toupper(s)}</span>")
}

results |>
  mutate(status_html = status_badge(status)) |>
  select(status_html, query, label, rows, ms, reason, error) |>
  DT::datatable(
    escape    = FALSE,
    rownames  = FALSE,
    options   = list(pageLength = 20, order = list(list(0, "asc"))),
    colnames  = c("", "query", "label", "rows", "ms", "skip reason", "error"))
```

## Consumer contract queries

The `_queries/*.md` suite covers the query app, but the `sql_builder` (match) and
the reproducible-download / app SQL live in calcofi4r + the Shiny apps and are
otherwise only exercised in the browser — so a schema change (a renamed column, a
retired table) slips past the gate and breaks downloads/apps in production. These
**contract** queries pin the exact shapes those consumers depend on (bio↔env
match, station rollup, sample/cruise counts) plus core-integrity asserts, run
server-side against the frozen release, and feed the same promote gate.

```{r}
#| label: consumer_contract

rel_base <- glue(
  "https://storage.googleapis.com/calcofi-db/ducklake/releases/{release_version}/parquet")
rp <- function(t) glue("read_parquet('{rel_base}/{t}.parquet')")

# name | expect ('nonzero' rows for a consumer read, or 'zero' for an integrity
# assert) | SQL returning a single column `n`
contract <- tibble::tribble(
  ~name, ~expect, ~sql,
  "consumer: match env (bottle temperature)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs')}
     WHERE realm='env' AND dataset_key='calcofi_bottle' AND measurement_type='temperature'"),
  "consumer: match/download bio (ichthyo abundance x taxon x effort)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs')} o
       JOIN {rp('taxon')} t ON t.taxon_key = o.taxon_key
       LEFT JOIN {rp('sample_measurement')} shf ON shf.sample_key=o.sample_key AND shf.measurement_type='std_haul_factor'
     WHERE o.realm='bio' AND o.dataset_key='swfsc_ichthyo' AND o.measurement_type='abundance'"),
  "integrity: every non-null obs.taxon_key resolves to taxon", "zero", glue(
    "SELECT count(*) n FROM {rp('obs')} o
     WHERE o.taxon_key IS NOT NULL
       AND NOT EXISTS (SELECT 1 FROM {rp('taxon')} t WHERE t.taxon_key=o.taxon_key)"),
  "consumer: station coverage rollup (obs GROUP BY grid,dataset)", "nonzero", glue(
    "SELECT count(*) n FROM (
       SELECT grid_key, dataset_key, count(*), count(DISTINCT sample_key), count(DISTINCT cruise_key)
       FROM {rp('obs')} WHERE grid_key IS NOT NULL GROUP BY 1,2)"),
  "consumer: sample event grain (bottle cast)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('sample')} WHERE dataset_key='calcofi_bottle' AND sample_type='cast'"),
  "consumer: cruise enriched per-dataset counts", "nonzero", glue(
    "SELECT count(*) n FROM {rp('cruise')} WHERE ichthyo IS NOT NULL"),
  "consumer: obs single-file readable over HTTPS (browser match.js)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs')} WHERE realm='bio' LIMIT 1"),
  "contract: obs.measurement_type all in registry", "zero", glue(
    "SELECT count(*) n FROM {rp('obs')}
     WHERE measurement_type NOT IN (SELECT measurement_type FROM {rp('measurement_type')})"),
  "contract: obs.sample_key resolves in sample", "zero", glue(
    "SELECT count(*) n FROM {rp('obs')} WHERE sample_key NOT IN (SELECT sample_key FROM {rp('sample')})"),
  "contract: obs.hex_id present where lat/lng", "zero", glue(
    "SELECT count(*) n FROM {rp('obs')} WHERE hex_id IS NULL AND latitude IS NOT NULL AND longitude IS NOT NULL"))

contract_res <- purrr::pmap_dfr(contract, function(name, expect, sql) {
  t0  <- Sys.time()
  out <- tryCatch(DBI::dbGetQuery(con_test, sql), error = function(e) e)
  ms  <- as.numeric(difftime(Sys.time(), t0, units = "secs")) * 1000
  if (inherits(out, "error"))
    return(tibble::tibble(query = name, label = "consumer-contract", status = "fail",
      reason = NA_character_, rows = NA_integer_, ms = round(ms, 1), error = conditionMessage(out)))
  n  <- as.numeric(out$n[1])
  ok <- if (expect == "zero") n == 0 else n > 0
  tibble::tibble(query = name, label = "consumer-contract",
    status = if (ok) "pass" else "fail", reason = NA_character_,
    rows = as.integer(n), ms = round(ms, 1),
    error = if (ok) NA_character_ else glue("expected {expect}, got n={n}"))
})

# fold into the same results table the promote gate reads
results <- dplyr::bind_rows(results, contract_res)

contract_res |>
  mutate(status_html = status_badge(status)) |>
  select(status_html, query, rows, ms, error) |>
  DT::datatable(escape = FALSE, rownames = FALSE,
    options = list(pageLength = 20), colnames = c("", "contract query", "n", "ms", "error"))
```

## Save results sidecar

```{r}
#| label: save_results

results_path <- file.path(here("data/releases"), release_version,
                          "test_results.json")
jsonlite::write_json(
  list(
    release_version = release_version,
    tested_at       = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    n_pass          = sum(results$status == "pass"),
    n_fail          = sum(results$status == "fail"),
    n_skip          = sum(results$status == "skip"),
    results         = results),
  results_path,
  auto_unbox = TRUE, pretty = TRUE, null = "null")

gcs_bucket  <- "calcofi-db"
gcs_release <- glue("ducklake/releases/{release_version}")
put_gcs_file(results_path,
  glue("gs://{gcs_bucket}/{gcs_release}/test_results.json"))
message(glue("test_results.json uploaded for {release_version}"))
```

## Promote (gated on all-pass)

`latest.txt` is written **only** if no query failed. Skipped queries
(JS-only `sql_builder`) do not block promotion — they're exercised in
the browser by the query app.

```{r}
#| label: promote

n_fail <- sum(results$status == "fail")
n_pass <- sum(results$status == "pass")
n_skip <- sum(results$status == "skip")

if (n_fail > 0) {
  failures <- results |> filter(status == "fail")
  for (i in seq_len(nrow(failures))) {
    message(glue("FAIL [{failures$query[i]}]: {failures$error[i]}"))
  }
  stop(glue(
    "{n_fail} query(ies) failed against {release_version}. ",
    "latest.txt NOT updated."))
}

# all green — promote
latest_local <- tempfile()
writeLines(release_version, latest_local)
put_gcs_file(latest_local,
  glue("gs://{gcs_bucket}/ducklake/releases/latest.txt"))
message(glue(
  "PROMOTED {release_version}: {n_pass} passed, {n_skip} skipped, ",
  "0 failed. latest.txt is now {release_version}."))

# nudge the query site to adopt the new default_version now (previously a
# 6-hourly cron poll): dispatch CalCOFI/db-query's bump-default-version workflow,
# which mirrors latest.txt -> _config.yml and triggers a Pages redeploy. Gated
# here on the all-pass promotion, so the site never points at a release whose
# queries haven't been validated. Non-fatal — a missing gh just means the bump
# waits for a manual `gh workflow run`.
gh_bin <- Sys.which("gh")
if (nzchar(gh_bin)) {
  disp <- system2(gh_bin,
    c("workflow", "run", "bump-default-version.yml", "--repo", "CalCOFI/db-query"),
    stdout = TRUE, stderr = TRUE)
  if (!identical(attr(disp, "status") %||% 0L, 0L))
    warning(glue("query bump dispatch failed: {paste(disp, collapse = '; ')}"))
  else
    message("dispatched CalCOFI/db-query bump-default-version workflow")
} else {
  message("gh not found; CalCOFI/db-query default_version needs a manual bump dispatch")
}

# also refresh the station data portal: rebuild its prebuilt per-station coverage
# JSON (public/data/*.json) from the new release's ingest parquet, then redeploy.
# Same gating + non-fatal pattern as the db-query bump above.
if (nzchar(gh_bin)) {
  disp2 <- system2(gh_bin,
    c("workflow", "run", "refresh.yml", "--repo", "CalCOFI/db-viz-station"),
    stdout = TRUE, stderr = TRUE)
  if (!identical(attr(disp2, "status") %||% 0L, 0L))
    warning(glue("station-portal refresh dispatch failed: {paste(disp2, collapse = '; ')}"))
  else
    message("dispatched CalCOFI/db-viz-station refresh workflow")
}
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con_test)
```

::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
