---
title: "Test Release — gate latest.txt on query suite"
calcofi:
  target_name: test_release
  workflow_type: test
  dependency:
    - release_database
  output: _output/test_release.html
execute:
  echo: true
  message: true
  warning: true
editor_options:
  chunk_output_type: console
format:
  html:
    code-fold: true
editor:
  markdown:
    wrap: 72
---

## Overview {.unnumbered}

**Goal**: Promote a freshly-built release to `latest.txt` only when every
pre-baked query in [CalCOFI/db-query](https://github.com/CalCOFI/db-query) runs
cleanly against it.

Pipeline position: this notebook runs after `release_database.qmd` has
uploaded the parquet, `metadata.json`, `relationships.json`, `erd.mmd`, and
`catalog.json` to `gs://calcofi-db/ducklake/releases/{release_version}/`,
but **before** `latest.txt` is updated. Every `.md` query in the
[`_queries/`](https://github.com/CalCOFI/db-query/tree/main/_queries) folder
is rendered with its YAML defaults, executed against the just-uploaded
release on GCS, and reported below. If every query returns without error,
`latest.txt` is written; otherwise the target fails and consumer apps
keep pointing at the previously-promoted version.

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
librarian::shelf(
  CalCOFI / calcofi4db,
  DBI,
  duckdb,
  dplyr,
  DT,
  fs,
  glue,
  here,
  jsonlite,
  purrr,
  tibble,
  yaml,
  quiet = T)

# which release to test? read the just-uploaded version from the local
# release directory (most recent date-stamped folder). A STAGING run
# (CALCOFI_RELEASE_PREFIX=ducklake-staging/releases) keeps its sidecars under
# data/releases-staging/ — the same rule release_database.qmd applies — so the
# version MUST be picked there: until 2026-09-04 this always read data/releases/,
# and the 2026-08-28 staging run "passed" by testing the promoted v2026.08.25
# instead of the staging catalog (no test_results.json ever landed under
# data/releases-staging/). A false green is worse than a failure.
release_prefix <- Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake/releases")
staging        <- release_prefix != "ducklake/releases"
releases_dir   <- here(if (staging) "data/releases-staging" else "data/releases")
if (staging) message(glue("STAGING run: testing the release under {releases_dir}"))
release_dirs <- list.dirs(releases_dir, recursive = FALSE)
release_dirs <- release_dirs[grepl("v[0-9]{4}[.][0-9]{2}[.]?[0-9]*$",
                                   basename(release_dirs))]
stopifnot(length(release_dirs) > 0)
release_version <- basename(release_dirs[which.max(file.mtime(release_dirs))])
message(glue("Testing release: {release_version}"))

# where the pre-baked queries live
queries_dir <- here("../db-query/_queries")
if (!dir.exists(queries_dir)) {
  stop(glue(
    "Could not find {queries_dir}. Clone CalCOFI/db-query as a sibling of ",
    "this repo, or set queries_dir manually."))
}
message(glue("Query repo: {queries_dir}"))

# fresh in-memory DuckDB with httpfs + spatial — every query reads parquet
# from GCS over httpfs, mirroring what the query app's DuckDB-WASM does
con_test <- get_duckdb_con(":memory:")
load_duckdb_extension(con_test, "httpfs")
load_duckdb_extension(con_test, "spatial")
```

## Render the query templates

The query app uses Handlebars at runtime. For tests we replicate the
subset of Handlebars features actually used in `_queries/*.md`:

- `{{var}}` — simple substitution
- `{{{var}}}` — raw substitution (no escaping; same effect here)
- `{{sqlesc var}}` — escape single quotes for SQL string literals
- `{{#if var}}...{{/if}}` — include block when `var` is truthy

```{r}
#| label: render_helpers

sqlesc <- function(x) gsub("'", "''", as.character(x %||% ""), fixed = TRUE)

# truthy test mirrors Handlebars: non-NULL, non-NA, non-empty string,
# non-zero number, non-FALSE logical
is_truthy <- function(val) {
  if (is.null(val))           return(FALSE)
  if (length(val) == 0)       return(FALSE)
  if (is.logical(val))        return(isTRUE(val))
  if (is.character(val))      return(any(!is.na(val) & nzchar(val)))
  if (is.numeric(val))        return(any(!is.na(val) & val != 0))
  !is.na(val[[1]])
}

# resolve innermost {{#if X}}...{{/if}} repeatedly until none remain.
# [\\s\\S] (not .) so the body spans newlines; (?!\\{\\{#if) guard forces
# "innermost only" so we strip from the inside out.
render_if_blocks <- function(txt, params) {
  pat <- "\\{\\{#if\\s+([A-Za-z_][A-Za-z0-9_]*)\\}\\}((?:(?!\\{\\{#if)[\\s\\S])*?)\\{\\{/if\\}\\}"
  repeat {
    m <- regexpr(pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    parts <- regmatches(full, regexec(pat, full, perl = TRUE))[[1]]
    var_name <- parts[2]
    body     <- parts[3]
    replacement <- if (is_truthy(params[[var_name]])) body else ""
    txt <- paste0(
      substr(txt, 1, m - 1),
      replacement,
      substr(txt, m + attr(m, "match.length"), nchar(txt)))
  }
  txt
}

render_template <- function(template, params) {
  txt <- template

  # 1. {{#if X}}...{{/if}} blocks
  txt <- render_if_blocks(txt, params)

  # 2. {{sqlesc VAR}}
  helper_pat <- "\\{\\{\\s*sqlesc\\s+([A-Za-z_][A-Za-z0-9_]*)\\s*\\}\\}"
  repeat {
    m <- regexpr(helper_pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    name <- sub(helper_pat, "\\1", full, perl = TRUE)
    txt <- sub(full, sqlesc(params[[name]]), txt, fixed = TRUE)
  }

  # 3. {{{VAR}}} (triple-brace, raw) and {{VAR}} (simple) — handle the
  #    triple-brace first so the {{VAR}} pass doesn't eat the inner pair
  triple_pat <- "\\{\\{\\{\\s*([A-Za-z_][A-Za-z0-9_]*)\\s*\\}\\}\\}"
  repeat {
    m <- regexpr(triple_pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    name <- sub(triple_pat, "\\1", full, perl = TRUE)
    val <- as.character(params[[name]] %||% "")
    txt <- sub(full, val, txt, fixed = TRUE)
  }
  single_pat <- "\\{\\{\\s*([A-Za-z_][A-Za-z0-9_]*)\\s*\\}\\}"
  repeat {
    m <- regexpr(single_pat, txt, perl = TRUE)
    if (m == -1) break
    full <- regmatches(txt, m)
    name <- sub(single_pat, "\\1", full, perl = TRUE)
    val <- as.character(params[[name]] %||% "")
    txt <- sub(full, val, txt, fixed = TRUE)
  }

  # query-app convention: literal __VERSION__ placeholder -> release version,
  # mirroring _includes/form-field.html ({% replace "__VERSION__",
  # site.default_version %}). Used in free-form defaults like sql-shell/shell.md.
  if (!is.null(params$version)) {
    txt <- gsub("__VERSION__", as.character(params$version), txt, fixed = TRUE)
    # query-app convention since the content-addressed releases: __TBL:obs__ (or
    # __TBL:obs:dataset_key=calcofi_bottle__ for one partition) -> the
    # read_parquet(...) resolved through that version's catalog, mirroring
    # db-query lib/release.js. Never a hand-built releases/{v}/parquet path.
    tbl_pat <- "__TBL:([A-Za-z0-9_]+)(?::([A-Za-z0-9_]+)=([^_]+(?:_[^_]+)*?))?__"
    repeat {
      m <- regexpr(tbl_pat, txt, perl = TRUE)
      if (m == -1) break
      full <- regmatches(txt, m)
      tb   <- sub(tbl_pat, "\\1", full, perl = TRUE)
      col  <- sub(tbl_pat, "\\2", full, perl = TRUE)
      val  <- sub(tbl_pat, "\\3", full, perl = TRUE)
      txt <- sub(full, tbl_sql(as.character(params$version), tb,
                               if (nzchar(col)) col else NULL, if (nzchar(val)) val else NULL),
                 txt, fixed = TRUE)
    }
  }

  txt
}

# read_parquet(...) for a table (or one partition of it) in a release, through
# the catalog; one catalog fetch per version. objects[].path is bucket-relative,
# so the staging prefix resolves the same way. A catalog VIEW (`obs` over
# obs_bio + obs_env since D-S1, calcofi4db 3.31.0) expands to its SQL over the
# objects of the tables it reads, parenthesised — exactly what db-query's
# readParquetFor() does for `__TBL:obs__`, so this suite exercises the view the
# browser will run; a partition token on a view becomes a WHERE on it.
.tbl_cats <- new.env(parent = emptyenv())
tbl_sql <- function(version, table, col = NULL, val = NULL) {
  if (is.null(.tbl_cats[[version]])) {
    pre <- Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake/releases")
    .tbl_cats[[version]] <- jsonlite::fromJSON(
      glue("https://storage.googleapis.com/calcofi-db/{pre}/{version}/catalog.json"),
      simplifyVector = FALSE)
  }
  cat_ <- .tbl_cats[[version]]
  if (table %in% names(calcofi4r::cc_catalog_views(cat_))) {
    v <- paste0("(", calcofi4r::cc_view_sql(cat_, table, function(t) tbl_sql(version, t)), ")")
    if (!is.null(col)) v <- glue("(SELECT * FROM {v} WHERE \"{col}\" = '{val}')")
    return(v)
  }
  src <- calcofi4r::cc_release_sources(cat_, table)
  if (!is.null(col)) {
    keep <- grepl(glue("/{col}={val}/"), src$urls, fixed = TRUE)
    if (isTRUE(src$canonical) && any(keep)) src$urls <- src$urls[keep]
    else if (!isTRUE(src$canonical))          # legacy layout: the partition's own path
      src$urls <- sub("/\\*\\*/\\*\\.parquet$", glue("/{col}={val}/*.parquet"), src$urls)
  }
  calcofi4r::cc_read_parquet_sql(src)
}

# parse YAML front-matter + body from a query .md file
parse_query_md <- function(path) {
  raw <- readLines(path, warn = FALSE)
  delims <- which(trimws(raw) == "---")
  if (length(delims) < 2) {
    stop(glue("Missing YAML delimiters in {path}"))
  }
  yaml_text <- paste(raw[(delims[1] + 1):(delims[2] - 1)], collapse = "\n")
  meta <- yaml::yaml.load(yaml_text)
  body <- paste(raw[(delims[2] + 1):length(raw)], collapse = "\n")
  list(meta = meta, body = body, path = path)
}

# build the params list for a query: each parameter's `default`, with
# `version` overridden to the release under test
defaults_for <- function(meta, release_version) {
  params <- list(version = release_version)
  if (!is.null(meta$parameters)) {
    for (name in names(meta$parameters)) {
      spec <- meta$parameters[[name]]
      params[[name]] <- spec$default
    }
  }
  params$version <- release_version
  params
}
```

## Discover and execute every query

```{r}
#| label: execute

query_files <- list.files(queries_dir, pattern = "[.]md$",
                          recursive = TRUE, full.names = TRUE)
message(glue("Found {length(query_files)} query files"))

results <- purrr::map_dfr(query_files, function(qpath) {
  rel  <- sub(paste0(normalizePath(queries_dir), "/"), "",
              normalizePath(qpath), fixed = TRUE)
  q    <- parse_query_md(qpath)
  meta <- q$meta

  # sql_builder queries are JS-only — flag them; the query app exercises
  # them in the browser
  if (!is.null(meta$sql_builder)) {
    return(tibble::tibble(
      query    = rel,
      label    = meta$label %||% NA_character_,
      status   = "skip",
      reason   = paste0("sql_builder=", meta$sql_builder),
      rows     = NA_integer_,
      ms       = NA_real_,
      error    = NA_character_))
  }

  if (is.null(meta$sql)) {
    return(tibble::tibble(
      query = rel, label = meta$label %||% NA_character_,
      status = "skip", reason = "no inline sql",
      rows = NA_integer_, ms = NA_real_, error = NA_character_))
  }

  params <- defaults_for(meta, release_version)
  sql    <- render_template(meta$sql, params)

  t0  <- Sys.time()
  out <- tryCatch(
    DBI::dbGetQuery(con_test, sql),
    error = function(e) e)
  elapsed_ms <- as.numeric(difftime(Sys.time(), t0, units = "secs")) * 1000

  if (inherits(out, "error")) {
    tibble::tibble(
      query = rel, label = meta$label %||% NA_character_,
      status = "fail", reason = NA_character_,
      rows   = NA_integer_, ms = round(elapsed_ms, 1),
      error  = conditionMessage(out))
  } else {
    tibble::tibble(
      query = rel, label = meta$label %||% NA_character_,
      status = "pass", reason = NA_character_,
      rows   = nrow(out), ms = round(elapsed_ms, 1),
      error  = NA_character_)
  }
})

# status badges for the report
status_badge <- function(s) {
  color <- c(pass = "#2e7d32", fail = "#c62828", skip = "#757575")[s]
  glue("<span style='background:{color};color:#fff;padding:2px 8px;",
       "border-radius:10px;font-size:11px;'>{toupper(s)}</span>")
}

results |>
  mutate(status_html = status_badge(status)) |>
  select(status_html, query, label, rows, ms, reason, error) |>
  DT::datatable(
    escape    = FALSE,
    rownames  = FALSE,
    options   = list(pageLength = 20, order = list(list(0, "asc"))),
    colnames  = c("", "query", "label", "rows", "ms", "skip reason", "error"))
```

## Consumer contract queries

The `_queries/*.md` suite covers the query app, but the `sql_builder` (match) and
the reproducible-download / app SQL live in calcofi4r + the Shiny apps and are
otherwise only exercised in the browser — so a schema change (a renamed column, a
retired table) slips past the gate and breaks downloads/apps in production. These
**contract** queries pin the exact shapes those consumers depend on (bio↔env
match, station rollup, sample/cruise counts) plus core-integrity asserts, run
server-side against the frozen release, and feed the same promote gate.

```{r}
#| label: consumer_contract

# tables resolved through this release's catalog (content-addressed since
# v2026.09; objects[].path is bucket-relative, so the staging prefix works too),
# exactly as calcofi4r::cc_get_db() and every migrated consumer resolve them
rel_prefix <- Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake/releases")
rel_cat <- jsonlite::fromJSON(
  glue("https://storage.googleapis.com/calcofi-db/{rel_prefix}/{release_version}/catalog.json"),
  simplifyVector = FALSE)
rp <- function(t) calcofi4r::cc_read_parquet_sql(calcofi4r::cc_release_sources(rel_cat, t))
# the catalog's `obs` VIEW over obs_bio + obs_env (D-S1, calcofi4db 3.31.0), as cc_get_db() and
# db-query serve `obs` from this release on: its SQL over the pair's objects, parenthesised so it
# stands wherever rp('obs') stood. Every contract row that reads obs runs twice — against the
# deprecated table's own objects (rp) and against the view (rpv) — until the objects are dropped.
rpv <- function(t) paste0("(", calcofi4r::cc_view_sql(rel_cat, t, rp), ")")
# a partitioned table through its single-file twin (obs publishes one): the table's own column order
rp1 <- function(t) calcofi4r::cc_read_parquet_sql(calcofi4r::cc_release_sources(rel_cat, t), prefer_single_file = TRUE)
has_obs_view <- "obs" %in% names(calcofi4r::cc_catalog_views(rel_cat))
stopifnot(has_obs_view)   # the catalog must carry the view: consumers read obs through it

# the rows that read obs, parameterised on HOW obs is read (R = rp or rpv) and tagged
obs_rows <- function(R, tag) tibble::tribble(
  ~name, ~expect, ~sql,
  glue("consumer: match env (bottle temperature) [{tag}]"), "nonzero", glue(
    "SELECT count(*) n FROM {R('obs')}
     WHERE realm='env' AND dataset_key='calcofi_bottle' AND measurement_type='temperature'"),
  glue("consumer: match/download bio (ichthyo abundance x taxon x effort) [{tag}]"), "nonzero", glue(
    "SELECT count(*) n FROM {R('obs')} o
       JOIN {rp('taxon')} t ON t.taxon_key = o.taxon_key
       LEFT JOIN {rp('sample_measurement')} shf ON shf.sample_key=o.sample_key AND shf.measurement_type='std_haul_factor'
     WHERE o.realm='bio' AND o.dataset_key='swfsc_ichthyo' AND o.measurement_type='abundance'"),
  glue("integrity: every non-null obs.taxon_key resolves to taxon [{tag}]"), "zero", glue(
    "SELECT count(*) n FROM {R('obs')} o
     WHERE o.taxon_key IS NOT NULL
       AND NOT EXISTS (SELECT 1 FROM {rp('taxon')} t WHERE t.taxon_key=o.taxon_key)"),
  # db-viz-hex::get_sp() picks taxa by worms_id and walks the WoRMS hierarchy for
  # children. Seabirds and marine mammals key `itis:` (WoRMS bird taxonomy lags),
  # and nothing populated their worms_id COLUMN — so every one of them returned
  # ZERO rows: 59,858 of 64,956 Farallon obs, 92.2% of the dataset, with no error
  # anywhere. Resolving a taxon and making it REACHABLE are different things;
  # assert the second.
  glue("consumer: seabird/mammal taxa reachable by worms_id (db-viz-hex get_sp) [{tag}]"), "nonzero", glue(
    "SELECT count(*) n FROM {R('obs')} o
       JOIN {rp('taxon')} t ON t.taxon_key = o.taxon_key
     WHERE o.dataset_key='farallon_bird-mammal' AND t.worms_id IS NOT NULL"),
  # `o.taxon_key IS NOT NULL` is load-bearing, not defensive: 1,316 of this
  # dataset's 66,272 obs carry NO taxon at all — species codes the source marks
  # Include=0 (fish, and non-target sightings) — and a LEFT JOIN reports those as
  # a NULL worms_id too. Without the filter this asserts "every observation has a
  # taxon", which is a different (and false) claim than the one intended.
  glue("integrity: no farallon obs left unreachable by worms_id [{tag}]"), "zero", glue(
    "SELECT count(*) n FROM {R('obs')} o
       LEFT JOIN {rp('taxon')} t ON t.taxon_key = o.taxon_key
     WHERE o.dataset_key='farallon_bird-mammal'
       AND o.taxon_key IS NOT NULL AND t.worms_id IS NULL"),
  glue("consumer: station coverage rollup (obs GROUP BY grid,dataset) [{tag}]"), "nonzero", glue(
    "SELECT count(*) n FROM (
       SELECT grid_key, dataset_key, count(*), count(DISTINCT sample_key), count(DISTINCT cruise_key)
       FROM {R('obs')} WHERE grid_key IS NOT NULL GROUP BY 1,2)"),
  glue("consumer: obs readable over HTTPS as the browser reads it (db-query match.js) [{tag}]"), "nonzero", glue(
    "SELECT count(*) n FROM {R('obs')} WHERE realm='bio' LIMIT 1"),
  glue("contract: obs.measurement_type all in registry [{tag}]"), "zero", glue(
    "SELECT count(*) n FROM {R('obs')}
     WHERE measurement_type NOT IN (SELECT measurement_type FROM {rp('measurement_type')})"),
  glue("contract: obs.sample_key resolves in sample [{tag}]"), "zero", glue(
    "SELECT count(*) n FROM {R('obs')} WHERE sample_key NOT IN (SELECT sample_key FROM {rp('sample')})"),
  glue("contract: obs.hex_id present where lat/lng [{tag}]"), "zero", glue(
    "SELECT count(*) n FROM {R('obs')} WHERE hex_id IS NULL AND latitude IS NOT NULL AND longitude IS NOT NULL"))

# the same contract, read from the pair DIRECTLY under its own names (value, no realm) — what a
# migrated consumer writes; obs_bio carries the effort inline, so the bio row needs no join
pair_rows <- tibble::tribble(
  ~name, ~expect, ~sql,
  "consumer: match env (bottle temperature) [obs_env]", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs_env')}
     WHERE dataset_key='calcofi_bottle' AND measurement_type='temperature' AND value IS NOT NULL"),
  "consumer: match/download bio (ichthyo abundance x taxon x effort) [obs_bio]", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')} o JOIN {rp('taxon')} t ON t.taxon_key = o.taxon_key
     WHERE o.dataset_key='swfsc_ichthyo' AND o.measurement_type='abundance' AND o.std_haul_factor IS NOT NULL"),
  "integrity: every non-null obs_bio.taxon_key resolves to taxon [obs_bio]", "zero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')} o
     WHERE o.taxon_key IS NOT NULL AND NOT EXISTS (SELECT 1 FROM {rp('taxon')} t WHERE t.taxon_key=o.taxon_key)"),
  "consumer: seabird/mammal taxa reachable by worms_id [obs_bio]", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')} o JOIN {rp('taxon')} t ON t.taxon_key = o.taxon_key
     WHERE o.dataset_key='farallon_bird-mammal' AND t.worms_id IS NOT NULL"),
  "integrity: no farallon obs_bio left unreachable by worms_id [obs_bio]", "zero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')} o LEFT JOIN {rp('taxon')} t ON t.taxon_key = o.taxon_key
     WHERE o.dataset_key='farallon_bird-mammal' AND o.taxon_key IS NOT NULL AND t.worms_id IS NULL"),
  "consumer: station coverage rollup (obs_bio + obs_env GROUP BY grid,dataset) [pair]", "nonzero", glue(
    "SELECT count(*) n FROM (
       SELECT grid_key, dataset_key, count(*), count(DISTINCT sample_key), count(DISTINCT cruise_key)
       FROM (SELECT grid_key, dataset_key, sample_key, cruise_key FROM {rp('obs_bio')}
             UNION ALL SELECT grid_key, dataset_key, sample_key, cruise_key FROM {rp('obs_env')})
       WHERE grid_key IS NOT NULL GROUP BY 1,2)"),
  "consumer: obs_bio single object readable over HTTPS (Explorer, db-query) [obs_bio]", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')} LIMIT 1"),
  "contract: pair measurement_type all in registry [pair]", "zero", glue(
    "SELECT count(*) n FROM (SELECT measurement_type FROM {rp('obs_bio')} UNION ALL SELECT measurement_type FROM {rp('obs_env')})
     WHERE measurement_type NOT IN (SELECT measurement_type FROM {rp('measurement_type')})"),
  "contract: pair sample_key resolves in sample [pair]", "zero", glue(
    "SELECT count(*) n FROM (SELECT sample_key FROM {rp('obs_bio')} UNION ALL SELECT sample_key FROM {rp('obs_env')})
     WHERE sample_key NOT IN (SELECT sample_key FROM {rp('sample')})"),
  "contract: pair hex_id present where lat/lng [pair]", "zero", glue(
    "SELECT count(*) n FROM (SELECT hex_id, latitude, longitude FROM {rp('obs_bio')} UNION ALL SELECT hex_id, latitude, longitude FROM {rp('obs_env')})
     WHERE hex_id IS NULL AND latitude IS NOT NULL AND longitude IS NOT NULL"),
  # the view IS obs: same rows per realm (the gate release_database.qmd ran per dataset with a signature)
  "integrity: the obs view reproduces obs row counts (bio = obs_bio, env = obs_env)", "zero", glue(
    "SELECT abs((SELECT count(*) FROM {rp('obs')} WHERE realm='bio') - (SELECT count(*) FROM {rp('obs_bio')}))
          + abs((SELECT count(*) FROM {rp('obs')} WHERE realm='env') - (SELECT count(*) FROM {rp('obs_env')}))
          + abs((SELECT count(*) FROM {rp('obs')}) - (SELECT count(*) FROM {rpv('obs')})) AS n"),
  # against the single-file twin, which carries the TABLE's column order; a read over the hive
  # partition list appends the partition column (dataset_key) last, which is a reader artefact
  "integrity: the obs view has obs's 18 columns in order", "zero", glue(
    "SELECT count(*) n FROM (
       SELECT row_number() OVER () AS i, column_name FROM (DESCRIBE SELECT * FROM {rpv('obs')})) v
     FULL JOIN (
       SELECT row_number() OVER () AS i, column_name FROM (DESCRIBE SELECT * FROM {rp1('obs')})) o USING (i)
     WHERE v.column_name IS DISTINCT FROM o.column_name"))

# name | expect ('nonzero' rows for a consumer read, or 'zero' for an integrity
# assert) | SQL returning a single column `n`
contract <- tibble::tribble(
  ~name, ~expect, ~sql,
  # The spatial tables had NO contract coverage at all until 2026-08-03, which is
  # why renaming _spatial -> spatial and _spatial_attr -> spatial_attribute could
  # not have been caught by the gate that exists to catch exactly that. These read
  # them the way a consumer does — including the JOIN KEY, since `id` is per-layer
  # and joining on it alone silently mixes layers.
  "consumer: spatial layer with attributes (joined on spatial_key)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('spatial')} s
       JOIN {rp('spatial_attribute')} a USING (spatial_key)
     WHERE s.layer = 'Marine Protected Areas'"),
  "integrity: every spatial_attribute row resolves to a spatial feature", "zero", glue(
    "SELECT count(*) n FROM {rp('spatial_attribute')} a
     WHERE NOT EXISTS (SELECT 1 FROM {rp('spatial')} s WHERE s.spatial_key = a.spatial_key)"),
  "integrity: spatial_key is unique", "zero", glue(
    "SELECT count(*) n FROM (SELECT spatial_key FROM {rp('spatial')}
                             GROUP BY 1 HAVING count(*) > 1)"),
  # a sample-to-polygon join is the thing consumers actually do, and it ERRORS
  # outright when CRS tags disagree — which they did until v2026.08.03
  "consumer: sample joins spatial (CRS tags must agree)", "nonzero", glue(
    "SELECT count(*) n FROM (SELECT geom FROM {rp('sample')}
                             WHERE geom IS NOT NULL LIMIT 20000) s
       JOIN {rp('spatial')} sp ON ST_Intersects(sp.geom, s.geom)"),
  # browser-shaped objects (Explorer plan D4, Phase 1): the app fetches these whole and aggregates in
  # the browser; every consumer inherits the exact polygon membership and the D8 densities
  "consumer: obs_bio carries the D8 densities for the default taxon (sardine larvae per 10 m2)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')}
     WHERE taxon_key = 'worms:217452' AND life_stage = 'larva' AND density_per_10m2 IS NOT NULL"),
  "integrity: obs_bio effort_class is the D8 vocabulary", "zero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')}
     WHERE effort_class NOT IN ('count_with_effort', 'raw_count_no_effort', 'density_as_published', 'other_unit')"),
  "integrity: obs_bio never derives a density without effort", "zero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')}
     WHERE effort_class = 'raw_count_no_effort' AND (density_per_10m2 IS NOT NULL OR density_per_1000m3 IS NOT NULL)"),
  "consumer: one env variable is one object (temperature, quality-filtered)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs_env')} WHERE measurement_type = 'temperature' AND qual_ok"),
  "integrity: every sample_spatial root resolves to sample_root", "zero", glue(
    "SELECT count(*) n FROM {rp('sample_spatial')} ss
     WHERE NOT EXISTS (SELECT 1 FROM {rp('sample_root')} r WHERE r.root_id = ss.root_id)"),
  "integrity: sample_spatial (root_id, spatial_key) is unique", "zero", glue(
    "SELECT count(*) n FROM (SELECT root_id, spatial_key FROM {rp('sample_spatial')}
                             GROUP BY 1, 2 HAVING count(*) > 1)"),
  "consumer: a region summary joins obs_bio to sample_spatial on root_id", "nonzero", glue(
    "SELECT count(*) n FROM {rp('obs_bio')} o JOIN {rp('sample_spatial')} s USING (root_id)
     WHERE s.layer = 'National Marine Sanctuaries'"),
  "integrity: no non-finite coordinates in sample", "zero", glue(
    "SELECT count(*) n FROM {rp('sample')}
     WHERE isnan(latitude) OR isnan(longitude) OR isinf(latitude) OR isinf(longitude)"),
  # (the rows that read obs live in obs_rows() above: once against the deprecated table's
  #  own objects, once against the catalog view; pair_rows() reads obs_bio / obs_env directly)
  # an itis:-keyed taxon must KEEP its itis: key once worms_id is filled — the
  # cross-reference is not a licence to re-key onto the lagging authority
  "integrity: seabirds still key itis:, not worms:", "nonzero", glue(
    "SELECT count(*) n FROM {rp('dataset_taxon')} dt
     WHERE dt.dataset_key='farallon_bird-mammal' AND dt.taxon_key LIKE 'itis:%'"),
  # taxonomic_status was the literal string "accepted" on all 2,090 taxa, stamped
  # rather than fetched. A single distinct value means it has regressed to that.
  "integrity: taxonomic_status is fetched, not stamped", "nonzero", glue(
    "SELECT count(*) n FROM (
       SELECT 1 FROM {rp('taxon')} WHERE status_checked IS NOT NULL LIMIT 1)"),
  "consumer: sample event grain (bottle cast)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('sample')} WHERE dataset_key='calcofi_bottle' AND sample_type='cast'"),
  "consumer: cruise enriched per-dataset counts", "nonzero", glue(
    "SELECT count(*) n FROM {rp('cruise')} WHERE ichthyo IS NOT NULL"),
  # the one baseline every anomaly subtracts (2026-08-31): read the way ctd-transects and the
  # Explorer's Sections lens do — the cast's month-matched cell for the CTD headline series
  "consumer: climatology month-matched cell (line 90 st 60, July, 0 m, temperature_ave)", "nonzero", glue(
    "SELECT count(*) n FROM {rp('climatology')}
     WHERE dataset_key = 'calcofi_ctd-cast' AND measurement_type = 'temperature_ave'
       AND grid_key = 'st60-ln90' AND month = 7 AND depth_bin = 0 AND n_cruises >= 3"),
  "integrity: climatology carries one window, stamped on every row", "zero", glue(
    "SELECT count(*) - 1 AS n FROM (SELECT DISTINCT clim_yr_min, clim_yr_max FROM {rp('climatology')})"),
  "integrity: climatology depth_bin is a 10 m floor bin within 0-500", "zero", glue(
    "SELECT count(*) n FROM {rp('climatology')} WHERE depth_bin % 10 <> 0 OR depth_bin < 0 OR depth_bin > 500"),
  # station_uuid (WS-B / Ed Weber's ask, calcofi4db >= 3.32.0): the SWFSC station
  # occupation each event belongs to. Read the way a consumer would — obs -> its
  # sample -> the ichthyo site sample.station_uuid names — and require a real
  # majority of bottle casts to resolve (measured 78.0% of 35,644 at v2026.08.25;
  # 25,000 is a conservative floor a schema regression would trip well before).
  "consumer: obs -> sample.station_uuid resolves to an ichthyo site (bottle)", "nonzero", glue(
    "SELECT CASE WHEN count(DISTINCT s.sample_key) >= 25000
                 THEN count(DISTINCT s.sample_key) ELSE 0 END AS n
     FROM {rp('obs')} o
     JOIN {rp('sample')} s ON s.sample_key = o.sample_key
     JOIN {rp('sample')} i ON i.source_uuid = s.station_uuid
     WHERE o.dataset_key = 'calcofi_bottle' AND s.station_uuid IS NOT NULL
       AND i.dataset_key = 'swfsc_ichthyo' AND i.sample_type = 'site'"))

# obs three ways — the deprecated table's own objects, the catalog view, the pair directly — so the
# contract covers every path a consumer can take through the deprecation window (D-S1)
contract <- dplyr::bind_rows(contract, obs_rows(rp, "obs"), obs_rows(rpv, "obs view"), pair_rows)

contract_res <- purrr::pmap_dfr(contract, function(name, expect, sql) {
  t0  <- Sys.time()
  out <- tryCatch(DBI::dbGetQuery(con_test, sql), error = function(e) e)
  ms  <- as.numeric(difftime(Sys.time(), t0, units = "secs")) * 1000
  if (inherits(out, "error"))
    return(tibble::tibble(query = name, label = "consumer-contract", status = "fail",
      reason = NA_character_, rows = NA_integer_, ms = round(ms, 1), error = conditionMessage(out)))
  n  <- as.numeric(out$n[1])
  ok <- if (expect == "zero") n == 0 else n > 0
  tibble::tibble(query = name, label = "consumer-contract",
    status = if (ok) "pass" else "fail", reason = NA_character_,
    rows = as.integer(n), ms = round(ms, 1),
    error = if (ok) NA_character_ else glue("expected {expect}, got n={n}"))
})

# fold into the same results table the promote gate reads
results <- dplyr::bind_rows(results, contract_res)

contract_res |>
  mutate(status_html = status_badge(status)) |>
  select(status_html, query, rows, ms, error) |>
  DT::datatable(escape = FALSE, rownames = FALSE,
    options = list(pageLength = 20), colnames = c("", "contract query", "n", "ms", "error"))
```

## Save results sidecar

```{r}
#| label: save_results

results_path <- file.path(releases_dir, release_version, "test_results.json")
jsonlite::write_json(
  list(
    release_version = release_version,
    tested_at       = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    n_pass          = sum(results$status == "pass"),
    n_fail          = sum(results$status == "fail"),
    n_skip          = sum(results$status == "skip"),
    results         = results),
  results_path,
  auto_unbox = TRUE, pretty = TRUE, null = "null")

gcs_bucket  <- "calcofi-db"
gcs_release <- glue("ducklake/releases/{release_version}")
put_gcs_file(results_path,
  glue("gs://{gcs_bucket}/{gcs_release}/test_results.json"))
message(glue("test_results.json uploaded for {release_version}"))
```

## Promote (gated on all-pass)

`latest.txt` is written **only** if no query failed. Skipped queries
(JS-only `sql_builder`) do not block promotion — they're exercised in
the browser by the query app.

```{r}
#| label: promote

n_fail <- sum(results$status == "fail")
n_pass <- sum(results$status == "pass")
n_skip <- sum(results$status == "skip")

if (n_fail > 0) {
  failures <- results |> filter(status == "fail")
  for (i in seq_len(nrow(failures))) {
    message(glue("FAIL [{failures$query[i]}]: {failures$error[i]}"))
  }
  stop(glue(
    "{n_fail} query(ies) failed against {release_version}. ",
    "latest.txt NOT updated."))
}

# All green — but a green suite is NOT sufficient to promote, and on 2026-08-14
# that distinction cost an hour of broken consumers. `release_database` died at
# `upload_frozen` with the parquet uploaded and the JSON sidecars not; this suite
# then passed 28/28 against that parquet — correctly, the data was fine — and
# moved `latest.txt` to a release with no catalog.json, the file `cc_get_db()`
# opens. The queries test the DATA; they cannot see whether the release is
# READABLE, because they never open the catalog.
#
# promote_release() answers the second question first: it refuses to move the
# pointer unless catalog.json / metadata.json / relationships.json are all
# present, and writes the object with Cache-Control: no-cache so the change
# reaches consumers immediately instead of up to an hour later.
promote_release(release_version, bucket = gcs_bucket,
                prefix = Sys.getenv("CALCOFI_RELEASE_PREFIX", "ducklake/releases"))
message(glue(
  "PROMOTED {release_version}: {n_pass} passed, {n_skip} skipped, ",
  "0 failed. latest.txt is now {release_version}."))

# re-publish this version's RELEASE_NOTES.md now that catalog.json (with its
# total size) and test_results.json exist and latest.txt points here — the
# appendix carries the validation result and the "promoted" mark. Notes are not
# data, so a failure here is a warning, never an un-promotion.
tryCatch(
  publish_release_notes(
    release_version, here::here("RELEASES.md"), releases_dir,
    bucket = gcs_bucket,
    pkg_versions = c(
      calcofi4db = as.character(packageVersion("calcofi4db")),
      calcofi4r  = as.character(packageVersion("calcofi4r")))),
  error = function(e) warning(glue("release notes not re-published: {conditionMessage(e)}")))

# regenerate the browsable release index so the public listing cannot lag the
# bucket. `versions.json` and `latest.txt` are written by the release itself and
# were always current, but the HTML at
# storage.calcofi.io/calcofi-db/ducklake/releases/ is a STATIC page built by
# scripts/build_release_index.R -- which nothing called. It was last run by hand
# on 2026-07-29, so it advertised "16 releases, latest v2026.07.17" while the
# bucket held 20 and had long since promoted past it. Data consumers were
# unaffected (they read latest.txt); it was humans who were misinformed.
#
# Runs AFTER the promotion above, because the index reads latest.txt to mark the
# "latest" badge. Non-fatal for the same reason as the dispatch below: a stale
# index is a documentation bug, not a reason to fail a validated release.
idx_script <- here::here("scripts", "build_release_index.R")
if (file.exists(idx_script)) {
  idx_rc <- system2(file.path(R.home("bin"), "Rscript"), idx_script,
                    stdout = TRUE, stderr = TRUE)
  if ((attr(idx_rc, "status") %||% 0L) != 0) {
    warning(glue("release index rebuild FAILED (latest.txt is still correct): ",
                 "{paste(utils::tail(idx_rc, 5), collapse = ' | ')}"))
  } else {
    message("release index rebuilt")
  }
}

# nudge the query site to adopt the new default_version now (previously a
# 6-hourly cron poll): dispatch CalCOFI/db-query's bump-default-version workflow,
# which mirrors latest.txt -> _config.yml and triggers a Pages redeploy. Gated
# here on the all-pass promotion, so the site never points at a release whose
# queries haven't been validated. Non-fatal — a missing gh just means the bump
# waits for a manual `gh workflow run`.
# Declared as a table rather than one if-block per consumer: the first two were
# copy-pasted and had already drifted (only one warned when `gh` was missing), and
# each new static consumer would add another copy of the same eight lines.
gh_dispatch <- tribble(
  ~repo,                      ~workflow,                     ~what,
  "CalCOFI/db-query",         "bump-default-version.yml",    "query-site default_version bump",
  "CalCOFI/db-viz-station",   "refresh.yml",                 "station-portal coverage refresh",
  # rebuilds the per-(line, cruise) section shards from the new release
  "CalCOFI/ctd-transects",    "refresh.yml",                 "CTD transect sections refresh")

gh_bin <- Sys.which("gh")
if (!nzchar(gh_bin)) {
  message(glue(
    "gh not found; these need a manual `gh workflow run`: ",
    "{paste(gh_dispatch$repo, collapse = ', ')}"))
} else {
  pwalk(gh_dispatch, function(repo, workflow, what) {
    disp <- system2(gh_bin,
      c("workflow", "run", workflow, "--repo", repo),
      stdout = TRUE, stderr = TRUE)
    if (!identical(attr(disp, "status") %||% 0L, 0L))
      warning(glue("{what} dispatch failed: {paste(disp, collapse = '; ')}"))
    else
      message(glue("dispatched {repo} {workflow} ({what})"))
  })
}

# Server-side consumers (db-viz-hex, db-viz-cruise, the h3t tile API) are NOT
# deployed here. They are their own target — `deploy_consumers`, which depends on
# this one — so that "the release shipped but consumers were never updated" is a
# state visible in tar_visnetwork()/tar_outdated() rather than a silent branch
# inside this chunk, and so the deploy can be re-run without re-running the query
# suite. The dispatches above stay here because they are instantaneous and target
# GitHub-hosted consumers that redeploy themselves.

```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con_test)
```

::: {.callout-caution collapse="true"}

## Session Info

```{r session_info}
devtools::session_info()
```

:::
