---
title: "Ingest {{dataset_name}}"
editor_options:
  chunk_output_type: console
---

## Overview

Ingest **{{dataset_name}}** from {{provider_name}} into the CalCOFI database.

- **Provider**: `{{provider}}`
- **Dataset**: `{{dataset}}`
- **Source**: {{source_description}}

### Data Flow

```{mermaid}
graph LR
  A[Source CSVs] --> B[DuckDB Wrangling]
  B --> C[Validate & Clean]
  C --> D[Parquet Export]
  D --> E[GCS Archive]
  E --> F[Release Database]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db,
  CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, googledrive,
  here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr,
  tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# define paths
dataset_name <- "{{dataset_name}}"
provider     <- "{{provider}}"
dataset      <- "{{dataset}}"
dir_label    <- glue("{provider}_{dataset}")
dir_data     <- "~/My Drive/projects/calcofi/data-public"
overwrite    <- TRUE

dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

# clean slate if overwrite
if (overwrite) {
  if (file_exists(db_path))     file_delete(db_path)
  if (dir_exists(dir_parquet))  dir_delete(dir_parquet)
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

```{r}
#| label: read-source

# read CSV files with metadata redefinition
d <- read_csv_files(
  provider     = provider,
  dataset      = dataset,
  dir_data     = glue("{dir_data}/{{provider}}/{{dataset}}"),
  metadata_dir = here(glue("metadata/{provider}/{dataset}")),
  gcs_archive  = NULL)  # set to timestamp string if reading from GCS
```

## Check Data Integrity

```{r}
#| label: check-integrity

integrity <- check_data_integrity(d)
render_integrity_message(integrity)
```

## Source Files

```{r}
#| label: show-source
show_source_files(d)
```

## Tables and Fields to Ingest

```{r}
#| label: show-tables
show_tables_redefine(d)
```

```{r}
#| label: show-fields
show_fields_redefine(d)
```

## Load Tables into Database

```{r}
#| label: ingest-dataset

ingest_dataset(con, d)
```

## Establish Primary Keys
<!-- {{pk_section}} -->

```{r}
#| label: TODO-primary-keys

# TODO: choose PK strategy for this dataset
# option A: sequential IDs
# assign_sequential_ids(con, "{{dataset}}_data", "{{dataset}}_id", sort_cols = c("..."))

# option B: deterministic UUIDs (for datasets with natural composite keys)
# assign_deterministic_uuids(con, "{{dataset}}_data", "{{dataset}}_uuid",
#   key_cols = c("..."))

# option C: source IDs (if provider includes stable identifiers)
# (no action needed, use source PKs directly)
```

<!-- {{taxonomy_section_start}} -->
## Standardize Taxonomy

```{r}
#| label: TODO-taxonomy

# TODO: uncomment if dataset has species/taxonomy data
# standardize_species_local(
#   con,
#   spp_db_path  = "~/My Drive/projects/calcofi/data-public/MarineSensitivity/spp.duckdb",
#   species_tbl  = "species")

# build_taxon_hierarchy(
#   con,
#   spp_db_path  = "~/My Drive/projects/calcofi/data-public/MarineSensitivity/spp.duckdb",
#   species_tbl  = "species")
```
<!-- {{taxonomy_section_end}} -->

<!-- {{pivot_section_start}} -->
## Pivot Measurements

```{r}
#| label: TODO-pivot

# TODO: if dataset has wide-format measurements, pivot to long format
# e.g., pivot columns like temp, salinity, oxygen → measurement_type + value
#
# d_long <- dbGetQuery(con, "SELECT * FROM {{dataset}}_wide") |>
#   pivot_longer(
#     cols      = c(...),  # measurement columns
#     names_to  = "measurement_type",
#     values_to = "measurement_value") |>
#   filter(!is.na(measurement_value))
#
# dbWriteTable(con, "{{dataset}}_measurement", d_long, overwrite = TRUE)
```
<!-- {{pivot_section_end}} -->

<!-- {{measurement_summary_section_start}} -->
## Summarize Replicate Measurements

Aggregate replicate measurements at each unique position in time and
space (station + date + depth) into mean and standard deviation.
Filters out `NaN`, `-Inf`, `Inf` values. See `ctd_summary` in
`ingest_calcofi_ctd-cast.qmd` and `dic_summary` in
`ingest_calcofi_dic.qmd` for production examples.

```{r}
#| label: TODO-measurement-summary

# TODO: adjust table names and grouping columns for this dataset
# dbExecute(
#   con,
#   "CREATE OR REPLACE TABLE {{dataset}}_measurement_summary AS
#    SELECT
#      site_key,
#      datetime_utc,
#      depth_m,
#      measurement_type,
#      AVG(measurement_value)    AS avg,
#      CASE
#        WHEN COUNT(*) = 1 THEN 0
#        ELSE COALESCE(STDDEV_SAMP(measurement_value), 0)
#      END                       AS stddev,
#      COUNT(*)                  AS n_obs
#    FROM {{dataset}}_measurement
#    WHERE NOT isnan(measurement_value)
#      AND isfinite(measurement_value)
#    GROUP BY site_key, datetime_utc, depth_m, measurement_type"
# )
```
<!-- {{measurement_summary_section_end}} -->

<!-- {{cross_dataset_section_start}} -->
## Cross-Dataset Integration

```{r}
#| label: load-prior-tables

# load shared reference tables from prior ingest
load_prior_tables(
  con,
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables      = c("ship", "cruise", "grid"),
  geom_tables = c("grid"))
```

```{r}
#| label: TODO-cross-dataset

# TODO: match this dataset's records to existing cruises/casts
# derive_cruise_key_on_casts(con, ship_renames_csv = here("metadata/..."))
```
<!-- {{cross_dataset_section_end}} -->

## Schema Documentation

Define PKs, FKs, and color-code tables for the ER diagram.
Color scheme: lightblue = new dataset tables, lightyellow = amended
reference tables (e.g. measurement_type), white = shared metadata.

```{r}
#| label: schema

# TODO: define PK/FK relationships for this dataset
{{dataset}}_rels <- list(
  primary_keys = list(
    {{dataset}}_measurement = "{{dataset}}_measurement_id",
    measurement_type        = "measurement_type"),
  foreign_keys = list(
    list(table = "{{dataset}}_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))
    # add more FKs as needed

{{dataset}}_tables <- c(
  "{{dataset}}_sample", "{{dataset}}_measurement",
  "{{dataset}}_measurement_summary",
  "measurement_type", "dataset")
cc_erd(
  con,
  tables = {{dataset}}_tables,
  rels   = {{dataset}}_rels,
  colors = list(
    lightblue   = c("{{dataset}}_sample", "{{dataset}}_measurement",
                    "{{dataset}}_measurement_summary"),
    lightyellow = "measurement_type",
    white       = "dataset"))
```

### Write relationships.json

Write PK/FK sidecar for use in `release_database.qmd` schema merging.

```{r}
#| label: write-relationships

build_relationships_json(
  rels       = {{dataset}}_rels,
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset
)
```

## Add Spatial

```{r}
#| label: TODO-spatial

# TODO: add geometry column from lat/lon (adjust table and column names)
# add_point_geom(con, "{{spatial_table}}", lon_col = "lon_dec", lat_col = "lat_dec")
# assign_grid_key(con, "{{spatial_table}}")
```

## Load Dataset Metadata

Register this dataset in the `dataset` reference table, which tracks
citations, source URLs, and CalCOFI.org landing pages for all ingested
datasets. Ensure `metadata/dataset.csv` has a row for this dataset
before running.

```{r}
#| label: load-dataset-metadata

d_dataset <- read_csv(here("metadata/dataset.csv"))
dbWriteTable(con, "dataset", d_dataset, overwrite = TRUE)
cat(glue("dataset: {nrow(d_dataset)} datasets registered"), "\n")
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
if (length(results$warnings) > 0)
  cat("Warnings:\n", paste("-", results$warnings, collapse = "\n"), "\n")
```

## Enforce Column Types

```{r}
#| label: enforce-types

enforce_column_types(con)
```

## Data Preview

```{r}
#| label: preview

preview_tables(con)
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

write_parquet_outputs(con, dir_parquet)
```

## Write Metadata

```{r}
#| label: write-metadata

build_metadata_json(con, dir_parquet)
build_relationships_json(con, dir_parquet)
```

## Upload to GCS Archive

```{r}
#| label: upload-gcs

sync_to_gcs(dir_parquet, glue("ingest/{dir_label}"))
```

## Cleanup

```{r}
#| label: cleanup

dbDisconnect(con, shutdown = TRUE)
```
