---
title: "Ingest {{dataset_name}}"
calcofi:
  # follow-up questions for the data provider live here and render in the
  # "Questions for Data Providers" section below; questions_email.qmd aggregates
  # them per provider from metadata/{provider}/{dataset}/questions.csv
  questions_file: metadata/{{provider}}/{{dataset}}/questions.csv
editor_options:
  chunk_output_type: console
---

## Overview

Ingest **{{dataset_name}}** from {{provider_name}} into the CalCOFI database.

- **Provider**: `{{provider}}`
- **Dataset**: `{{dataset}}`
- **Source**: {{source_description}}

### Data Flow

```{mermaid}
graph LR
  A[Source CSVs] --> B[DuckDB Wrangling]
  B --> C[Validate & Clean]
  C --> D[Parquet Export]
  D --> E[GCS Archive]
  E --> F[Release Database]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db,
  CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue, googledrive,
  here, htmltools, janitor, jsonlite, knitr,
  lubridate, purrr, readr, sf, stringr,
  tibble, tidyr, units,
  quiet = T)
options(readr.show_col_types = F)
options(DT.options = list(scrollX = TRUE))

# common ingest settings (overwrite, dir_data)
source(here("libs/ingest.R"))

# define paths
dataset_name <- "{{dataset_name}}"
provider     <- "{{provider}}"
dataset      <- "{{dataset}}"
dir_label    <- glue("{provider}_{dataset}")

dir_parquet  <- here(glue("data/parquet/{dir_label}"))
db_path      <- here(glue("data/wrangling/{dir_label}.duckdb"))

# clean slate if overwrite
if (overwrite) {
  if (file_exists(db_path))     file_delete(db_path)
  if (dir_exists(dir_parquet))  dir_delete(dir_parquet)
}
dir_create(dirname(db_path))
con <- get_duckdb_con(db_path)
load_duckdb_extension(con, "spatial")
```

## Read Source Data

```{r}
#| label: read-source

# read CSV files with metadata redefinition and archive to GCS
d <- read_csv_files(
  provider     = provider,
  dataset      = dataset,
  dir_data     = dir_data,
  metadata_dir = here("metadata"),
  sync_archive = TRUE)  # archives source CSVs to gs://calcofi-files-public/archive/
```

## Check Data Integrity

```{r}
#| label: check-integrity

integrity <- check_data_integrity(d)
render_integrity_message(integrity)
```

## Source Files

```{r}
#| label: show-source
show_source_files(d)
```

## Tables and Fields to Ingest

```{r}
#| label: show-tables
show_tables_redefine(d)
```

```{r}
#| label: show-fields
show_fields_redefine(d)
```

## Load Tables into Database

```{r}
#| label: ingest-dataset

ingest_dataset(con, d)
```

## Establish Primary Keys
<!-- {{pk_section}} -->

```{r}
#| label: TODO-primary-keys

# TODO: choose PK strategy for this dataset
# option A: sequential IDs
# assign_sequential_ids(con, "{{dataset}}_data", "{{dataset}}_id", sort_cols = c("..."))

# option B: deterministic UUIDs (for datasets with natural composite keys)
# assign_deterministic_uuids(con, "{{dataset}}_data", "{{dataset}}_uuid",
#   key_cols = c("..."))

# option C: source IDs (if provider includes stable identifiers)
# (no action needed, use source PKs directly)
```

<!-- {{taxonomy_section_start}} -->
## Standardize Taxonomy

Delete this section if the dataset carries no taxa. Otherwise: the vocabulary is
**declared here** and **resolved by the package** (taxon plan D1-D6). There is no
per-dataset arm inside `calcofi4db` — a column you rename or drop below fails at
`append_dataset_taxon()` rather than becoming an `NA` at release. Copy the shape
from `ingest_farallon_bird-mammal.qmd`, which is the worked example.

```{r}
#| label: TODO-taxonomy

# registries. taxon_override.csv is read whole (every helper filters to ds_key
# itself); measurement_taxon.csv is filtered to THIS dataset.
mt_taxon <- read_csv(here("metadata/measurement_taxon.csv"),
                     col_types = cols(worms_id = "i", itis_id = "i",
                                      bin_value = "d", .default = "c")) |>
  filter(dataset_key == ds_key)
tx_over  <- read_csv(here("metadata/taxon_override.csv"), show_col_types = FALSE)

# 1. DECLARE. One row per local taxon. `ds_taxa_code` is the code `obs` stores,
#    VERBATIM (never cleaned — rewriting it orphans every observation of that
#    taxon); `ds_scientific_name` is the source's name and the lookup query;
#    `worms_id` / `itis_id` / `gbif_id` / `rank` are what THE SOURCE supplied and
#    are stored together as `ds_source_json`. Unknown columns error.
#    `ds_prefix` defaults to `dataset_key`; pass "calcofi" only for the shared
#    CalCOFI species list.
# d_vocab <- dbGetQuery(con, "
#   SELECT CAST(<code> AS VARCHAR) AS ds_taxa_code,
#          <name>                  AS ds_scientific_name,
#          <label>                 AS ds_common_name,
#          CAST(<aphia> AS INTEGER) AS worms_id
#   FROM <source_taxon_table> ORDER BY 1")
# n_staged <- append_dataset_taxon(con, ds_key, d_vocab)

# 2. RESOLVE against the authorities. xref BEFORE lineage: the lineage fetch must
#    ask about the accepted id, and the CLASS it returns is what decides whether
#    the key is `itis:` (Aves with an accepted TSN) or `worms:`. Both are cached
#    under metadata/, so a re-run costs no API calls.
# ensure_taxon_xref(con, mt_taxon, tx_over,
#                   cache_csv = here("metadata/taxon_xref.csv"))
# ensure_taxon_lineage(con, mt_taxon, tx_over,
#                      cache_csv = here("metadata/taxon_lineage.csv"))

# 3. BUILD the shared references. Groups come from the registry, never from code.
# n_taxon    <- build_taxon_reference(con, mt_taxon, tx_over)
# n_ds_taxon <- resolve_dataset_taxon(con, mt_taxon, tx_over)
# n_tx_group <- build_taxon_group(
#   con, read_taxon_group_rules(here("metadata/taxon_group.csv")))

# 4. ASSERT, before append_obs(). `codes` is every code the observations
#    reference; `allow` names the dataset-local keys ONE AT A TIME, with a
#    comment, for genuinely non-taxonomic classes (an operational bioclass, an
#    undefined source code) — never as a pattern.
# obs_codes <- dbGetQuery(con, "SELECT DISTINCT <code> FROM <observation_table>")[[1]]
# chk_taxon <- check_dataset_taxon(con, ds_key, codes = obs_codes,
#                                  allow = character())
```
<!-- {{taxonomy_section_end}} -->

<!-- {{pivot_section_start}} -->
## Pivot Measurements

```{r}
#| label: TODO-pivot

# TODO: if dataset has wide-format measurements, pivot to long format
# e.g., pivot columns like temp, salinity, oxygen → measurement_type + value
#
# d_long <- dbGetQuery(con, "SELECT * FROM {{dataset}}_wide") |>
#   pivot_longer(
#     cols      = c(...),  # measurement columns
#     names_to  = "measurement_type",
#     values_to = "measurement_value") |>
#   filter(!is.na(measurement_value))
#
# dbWriteTable(con, "{{dataset}}_measurement", d_long, overwrite = TRUE)
```
<!-- {{pivot_section_end}} -->

<!-- {{bounds_section_start}} -->
## Check Declared Bounds

Every long-format measurement table is checked against the `valid_min` /
`valid_max` declared in `metadata/measurement_type.csv`. This section is **not
optional** — it is where an impossible value is caught while the provider can
still be asked about it.

The check reports two findings, and the second is usually the bigger one:

* **`out_of_range`** — a bound is declared and the data breaks it. Almost always
  an unconverted sentinel (`-99`, `-999`) or a scaling error.
* **`undeclared`** — no bound is declared, so nothing was checked. At
  v2026.08.07 this was 73 of 98 (dataset, type) pairs and 67% of all `obs` rows,
  which is why a release-time guard alone is not enough: it can only enforce what
  someone declared during an ingest.

Every row that is not `ok` gets resolved one of two ways, and neither is
"leave it":

1. **Declare a bound** with `declare_measurement_bounds()` when the plausible
   range is known (`register_measurement_types()` only *appends*, so it cannot
   set a bound on a type that already exists). Bounds are *generous* — impossible, not merely unusual. A
   one-sided `valid_min = 0` for a count or an abundance is the common case and
   is worth declaring even when there is no meaningful ceiling.
2. **Ask the provider** — add a row to
   `metadata/{{provider}}/{{dataset}}/questions.csv` with the `finding` column as
   its `context`. Use `status = proposed` and put the bound you intend to apply
   in `proposed_answer`, so the provider confirms a decision rather than being
   handed a problem.

```{r}
#| label: bounds-check

d_bounds <- check_measurement_bounds(
  con,
  tbl = "{{dataset}}_measurement",
  mt  = here("metadata/measurement_type.csv"))

bounds_datatable(d_bounds)
```

```{r}
#| label: bounds-findings

# the copy-paste source for a questions.csv `context` cell
d_bounds |>
  filter(status != "ok") |>
  select(measurement_type, status, finding) |>
  dt(caption = "Bounds findings to declare or ask about",
     fname   = "{{dataset}}_bounds_findings")
```

```{r}
#| label: bounds-enforce

# TODO: enable once the bounds are agreed. DELETE rather than flag, because in a
# long-format table a row IS an assertion that a value was measured — a -10 pH
# left in place silently corrupts every mean, minimum and anomaly downstream, and
# there is no in-band way to mark it as not-a-value.
#
# drop_out_of_bounds(
#   con, tbl = "{{dataset}}_measurement",
#   mt = here("metadata/measurement_type.csv"))
```
<!-- {{bounds_section_end}} -->

<!-- {{measurement_summary_section_start}} -->
## Summarize Replicate Measurements

Aggregate replicate measurements at each unique position in time and
space (station + date + depth) into mean and standard deviation.
Filters out `NaN`, `-Inf`, `Inf` values. See `ctd_summary` in
`ingest_calcofi_ctd-cast.qmd` and `dic_summary` in
`ingest_calcofi_dic.qmd` for production examples.

```{r}
#| label: TODO-measurement-summary

# TODO: adjust table names and grouping columns for this dataset
# dbExecute(
#   con,
#   "CREATE OR REPLACE TABLE {{dataset}}_measurement_summary AS
#    SELECT
#      site_key,
#      datetime_utc,
#      depth_m,
#      measurement_type,
#      AVG(measurement_value)    AS avg,
#      CASE
#        WHEN COUNT(*) = 1 THEN 0
#        ELSE COALESCE(STDDEV_SAMP(measurement_value), 0)
#      END                       AS stddev,
#      COUNT(*)                  AS n_obs
#    FROM {{dataset}}_measurement
#    WHERE NOT isnan(measurement_value)
#      AND isfinite(measurement_value)
#    GROUP BY site_key, datetime_utc, depth_m, measurement_type"
# )
```
<!-- {{measurement_summary_section_end}} -->

<!-- {{cross_dataset_section_start}} -->
## Cross-Dataset Integration

```{r}
#| label: load-prior-tables

# load shared reference tables from prior ingest
load_prior_tables(
  con,
  parquet_dir = here("data/parquet/swfsc_ichthyo"),
  tables      = c("ship", "cruise", "grid"),
  geom_tables = c("grid"))
```

```{r}
#| label: TODO-cross-dataset

# TODO: match this dataset's records to existing cruises/casts.
#
# (a) records that carry ship + datetime -> derive cruise_key:
# derive_cruise_key_on_casts(con, ship_renames_csv = here("metadata/ship_renames.csv"))
#
# (b) records lacking a cast_id/cruise FK -> match by site_key + datetime window,
#     then to the nearest bottle by depth (resolves issue #47; see ingest_calcofi_dic.qmd):
# match_by_site_datetime(
#   con, data_tbl = "{{dataset}}_sample", ref_tbl = "casts",
#   fk_col = "cast_id", ref_pk = "cast_id",
#   key_col = "site_key", datetime_col = "datetime_start_utc", window_days = 3)
# match_nearest_by_depth(
#   con, data_tbl = "{{dataset}}_sample", ref_tbl = "bottle",
#   fk_col = "bottle_id", ref_pk = "bottle_id", parent_fk = "cast_id",
#   axis_col = "depth_m", tolerance = 1.0)
# both return matched/unmatched counts — report the match rate for validation.
```
<!-- {{cross_dataset_section_end}} -->

## Schema Documentation

Define PKs, FKs, and color-code tables for the ER diagram.
Color scheme: lightblue = new dataset tables, lightyellow = amended
reference tables (e.g. measurement_type), white = shared metadata.

```{r}
#| label: schema

# TODO: define PK/FK relationships for this dataset
{{dataset}}_rels <- list(
  primary_keys = list(
    {{dataset}}_measurement = "{{dataset}}_measurement_id",
    measurement_type        = "measurement_type"),
  foreign_keys = list(
    list(table = "{{dataset}}_measurement", column = "measurement_type",
         ref_table = "measurement_type", ref_column = "measurement_type")))
    # add more FKs as needed

{{dataset}}_tables <- c(
  "{{dataset}}_sample", "{{dataset}}_measurement",
  "{{dataset}}_measurement_summary",
  "measurement_type", "dataset")
cc_erd(
  con,
  tables = {{dataset}}_tables,
  rels   = {{dataset}}_rels,
  colors = list(
    lightblue   = c("{{dataset}}_sample", "{{dataset}}_measurement",
                    "{{dataset}}_measurement_summary"),
    lightyellow = "measurement_type",
    white       = "dataset"))
```

### Write relationships.json

Write PK/FK sidecar for use in `release_database.qmd` schema merging.

```{r}
#| label: write-relationships

build_relationships_json(
  rels       = {{dataset}}_rels,
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset
)
```

## Add Spatial

```{r}
#| label: TODO-spatial

# TODO: add geometry column from canonical lat/lon, then assign the CalCOFI grid
# cell. Use canonical names from metadata/field_dictionary.csv (latitude/longitude,
# NOT lat_dec/lon_dec). For datasets with coordinates but no site_key, this also
# provides the grid_key join path (no station+date match needed).
# add_point_geom(con, "{{spatial_table}}", lon_col = "longitude", lat_col = "latitude")
# assign_grid_key(con, "{{spatial_table}}")
```

## Load Dataset Metadata

Build the `dataset` reference row from this notebook's own
`calcofi.dataset_meta` YAML block (citations, source URLs, CalCOFI.org landing
page). `metadata/dataset.csv` is deprecated — the YAML is authoritative and
cannot drift from the pipeline.

```{r}
#| label: load-dataset-metadata

this_provider <- provider; this_dataset <- dataset  # also column names
d_dataset <- ingest_yaml_to_dataset_df(read_ingest_yaml(here())) |>
  filter(provider == this_provider, dataset == this_dataset)
stopifnot("this dataset must appear in the ingest YAML registry" = nrow(d_dataset) == 1)
dbWriteTable(con, "dataset", as.data.frame(d_dataset), overwrite = TRUE)
```

## Questions for Data Providers

Follow-up questions for the provider, tracked in
`metadata/{provider}/{dataset}/questions.csv`. Seed new questions during
`/explore-dataset` and `/generate-metadata`; `questions_email.qmd` aggregates the
`open` and `proposed` ones into a draft email per provider. A `blocker` question
still `open` fails `/validate-ingest --strict`.

```{r}
#| label: provider-questions

# one validated read + render for every ingest: the vocabulary and the column
# order live in calcofi4db, not in 16 hand-written factor() calls
questions_datatable(
  here(cc$questions_file),
  caption = "Questions for the {{provider}} {{dataset}} data providers (ranked)")
```

## Validate

```{r}
#| label: validate

results <- validate_for_release(con, checks = "all", strict = FALSE)
cat("Validation:", ifelse(results$passed, "PASSED", "FAILED"), "\n")
if (length(results$errors) > 0)
  cat("Errors:\n", paste("-", results$errors, collapse = "\n"), "\n")
if (length(results$warnings) > 0)
  cat("Warnings:\n", paste("-", results$warnings, collapse = "\n"), "\n")
```

## Enforce Column Types

```{r}
#| label: enforce-types

enforce_column_types(con)
```

## Data Preview

```{r}
#| label: preview

preview_tables(con)
```

## Emit Core Tables

Project this dataset's per-dataset tables into the shared **core** family
(`sample` / `obs` / `obs_attribute` / `sample_measurement`) via the `calcofi4db` model
helpers, then expose the per-dataset tables as compat VIEWs. See `RUNBOOK.md`
(§3b) and `design_env-bio-consolidation.md`. `sample_key` is namespaced
`dataset_key:sample_type:id`.

```{r}
#| label: emit-core

build_grid_reference(con)  # idempotent shared grid reference

# 1. sample dimension (adjacency list) — one row per physical event
append_sample(con, glue("
  SELECT '{provider}_{dataset}:{sample_type}:' || CAST({id_col} AS VARCHAR) AS sample_key,
         '{sample_type}' AS sample_type, NULL::VARCHAR AS parent_sample_key,
         '{provider}_{dataset}:{sample_type}:' || CAST({id_col} AS VARCHAR) AS root_sample_key,
         '{provider}_{dataset}' AS dataset_key, grid_key, cruise_key,
         latitude, longitude, CAST(datetime_start_utc AS TIMESTAMP) AS datetime,
         depth_min_m, depth_max_m
  FROM {event_tbl}"))

# 2. obs — occurrence headline (bio abundance = the count); 3. obs_attribute (bins);
#    4. sample_measurement (event-level effort). Fill the SELECTs per dataset.
append_obs(con, "<headline occurrence SELECT yielding the canonical obs columns>")
# append_obs_attribute(con, "<bin/count SELECT>")            # only if size/stage detail
# append_sample_measurement(con, "<event-effort SELECT>")  # only if per-event effort

# 5. per-dataset tables become VIEWs over the core (edit per dataset)
# dbExecute(con, "CREATE OR REPLACE VIEW {dataset}_measurement AS
#   SELECT ... FROM obs WHERE dataset_key = '{provider}_{dataset}'")
```

## Write Parquet Outputs

```{r}
#| label: write-parquet

write_parquet_outputs(con, dir_parquet)
```

## Write Metadata

```{r}
#| label: write-metadata

metadata_path <- build_metadata_json(
  con        = con,
  d_tbls_rd  = d$d_tbls_rd,
  d_flds_rd  = d$d_flds_rd,
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset,
  workflow_url = glue(
    "https://calcofi.io/workflows/ingest_{dir_label}.html"))

build_relationships_json(
  rels       = {{dataset}}_rels,
  output_dir = dir_parquet,
  provider   = provider,
  dataset    = dataset)
```

```{r}
#| label: show_metadata_json

listviewer::jsonedit(
  jsonlite::fromJSON(metadata_path, simplifyVector = FALSE),
  mode = "view")
```

```{r}
#| label: show_relationships_json

listviewer::jsonedit(
  jsonlite::fromJSON(
    file.path(dir_parquet, "relationships.json"),
    simplifyVector = FALSE),
  mode = "view")
```

## Upload to GCS

```{r}
#| label: upload-gcs

sync_to_gcs(
  local_dir  = dir_parquet,
  gcs_prefix = glue("ingest/{dir_label}"),
  bucket     = "calcofi-db")
```

## Cleanup

```{r}
#| label: cleanup

close_duckdb(con)
cat(glue("parquet outputs written to: {dir_parquet}"), "\n")
```
