---
title: "Publish {{dataset_name}} to OBIS"
editor_options:
  chunk_output_type: console
---

## Overview

Publish **{{dataset_name}}** to the Ocean Biodiversity Information System (OBIS) as a Darwin Core Archive (DwC-A).

- **Source ingest**: `{{source_ingest}}`
- **Portal**: OBIS (https://obis.org)
- **Format**: Darwin Core Archive (Event Core + Occurrence + eMoF)

### Event Hierarchy

```{mermaid}
graph TD
  A[Cruise Event] --> B[Station/Site Event]
  B --> C[Sample Event]
  C --> D[Occurrence Records]
  C --> E[eMoF Measurements]
```

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db,
  CalCOFI/calcofi4r,
  DBI, dplyr, DT, EML, fs, glue,
  here, htmltools, lubridate, purrr, readr,
  sf, stringr, tibble, tidyr, uuid, xml2,
  quiet = T)
options(readr.show_col_types = F)

# output paths
dataset      <- "{{dataset}}"
dir_dwca     <- here(glue("data/darwincore/{dataset}"))
dir_create(dir_dwca)
```

## Load Source Data

```{r}
#| label: load-data

# load from release parquet or ingest parquet
con <- get_duckdb_con(":memory:")
load_prior_tables(
  con,
  parquet_dir = here("data/parquet/{{source_ingest}}"),
  geom_tables = c("grid", "site"))

# TODO: list tables and preview
dbListTables(con)
```

## Build Event Table

```{r}
#| label: TODO-events

# TODO: build hierarchical events following DwC Event Core
# required fields: eventID, parentEventID, eventDate,
#   decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters,
#   geodeticDatum, samplingProtocol, sampleSizeValue, sampleSizeUnit

# cruise-level events
# d_events_cruise <- dbGetQuery(con, "
#   SELECT DISTINCT
#     cruise_key                          AS eventID,
#     NULL                                AS parentEventID,
#     ...
#   FROM cruise")

# station/sample-level events
# d_events_sample <- dbGetQuery(con, "
#   SELECT DISTINCT
#     ...                                 AS eventID,
#     cruise_key                          AS parentEventID,
#     ...
#   FROM ...")

# combine
# d_events <- bind_rows(d_events_cruise, d_events_sample)
```

## Build Occurrence Table

```{r}
#| label: TODO-occurrences

# TODO: build occurrence records
# required fields: occurrenceID, eventID, scientificName,
#   scientificNameID (WoRMS LSID: urn:lsid:marinespecies.org:taxname:{AphiaID}),
#   kingdom, occurrenceStatus, basisOfRecord

# d_occurrences <- dbGetQuery(con, "
#   SELECT
#     ...                                 AS occurrenceID,
#     ...                                 AS eventID,
#     scientific_name                     AS scientificName,
#     CONCAT('urn:lsid:marinespecies.org:taxname:', worms_id) AS scientificNameID,
#     'Animalia'                          AS kingdom,
#     'present'                           AS occurrenceStatus,
#     'HumanObservation'                  AS basisOfRecord
#   FROM ...")
```

## Build ExtendedMeasurementOrFact

```{r}
#| label: TODO-emof

# TODO: build eMoF records with NERC vocabulary URIs
# required fields: measurementID, occurrenceID (or eventID),
#   measurementType, measurementTypeID (NERC P01),
#   measurementValue, measurementUnit, measurementUnitID (NERC P06)

# example NERC URIs:
# abundance: http://vocab.nerc.ac.uk/collection/P01/current/SDBIOL01/
# count:     http://vocab.nerc.ac.uk/collection/P01/current/OCOUNT01/
# length:    http://vocab.nerc.ac.uk/collection/P01/current/OBSINDLX/
# per m²:    http://vocab.nerc.ac.uk/collection/P06/current/UPMS/
# mm:        http://vocab.nerc.ac.uk/collection/P06/current/UXMM/

# d_emof <- dbGetQuery(con, "
#   SELECT
#     ...   AS measurementID,
#     ...   AS occurrenceID,
#     ...   AS measurementType,
#     ...   AS measurementTypeID,
#     ...   AS measurementValue,
#     ...   AS measurementUnit,
#     ...   AS measurementUnitID
#   FROM ...")
```

## Write DwC-A Files

```{r}
#| label: write-csvs

# write_csv(d_events,      path(dir_dwca, "event.csv"))
# write_csv(d_occurrences, path(dir_dwca, "occurrence.csv"))
# write_csv(d_emof,        path(dir_dwca, "extendedmeasurementorfact.csv"))
```

## Create meta.xml

```{r}
#| label: TODO-meta-xml

# TODO: create meta.xml mapping columns to DwC terms
# follow pattern from publish_ichthyo_to_obis.qmd
# key namespaces:
#   core: http://rs.tdwg.org/dwc/terms/ (Event)
#   extension 1: http://rs.tdwg.org/dwc/terms/ (Occurrence)
#   extension 2: http://rs.iobis.org/obis/terms/ (eMoF)
```

## Create EML Metadata

```{r}
#| label: TODO-eml

# TODO: build EML metadata for the DwC-A
# required elements:
#   title, abstract, creator, contact,
#   geographicCoverage, temporalCoverage, taxonomicCoverage,
#   intellectualRights, methods

# eml <- list(
#   dataset = list(
#     title    = "{{dataset_name}}",
#     abstract = list(para = "..."),
#     creator  = list(
#       individualName = list(givenName = "...", surName = "..."),
#       organizationName = "CalCOFI",
#       electronicMailAddress = "..."
#     ),
#     contact = list(...),
#     coverage = list(
#       geographicCoverage = list(
#         geographicDescription = "CalCOFI sampling grid, Southern California Bight",
#         boundingCoordinates = list(
#           westBoundingCoordinate  = -135,
#           eastBoundingCoordinate  = -117,
#           northBoundingCoordinate = 51,
#           southBoundingCoordinate = 23)),
#       temporalCoverage = list(
#         rangeOfDates = list(
#           beginDate = list(calendarDate = "..."),
#           endDate   = list(calendarDate = "...")))),
#     intellectualRights = list(para = "CC-BY 4.0")
#   ))
```

## Data Quality Checks

```{r}
#| label: TODO-quality-checks

# TODO: validate DwC-A structure
# - every occurrence has a valid eventID in events table
# - every eMoF has a valid occurrenceID or eventID
# - no orphan events (events without occurrences)
# - scientificNameID format: urn:lsid:marinespecies.org:taxname:{digits}
# - coordinates within expected range
# - eventDate format: ISO 8601
```

## Package DwC Archive

```{r}
#| label: package-archive

# zip_path <- path(dir_dwca, glue("{dataset}_dwca.zip"))
# zip_files <- c("event.csv", "occurrence.csv",
#                "extendedmeasurementorfact.csv", "meta.xml", "eml.xml")
# withr::with_dir(dir_dwca, zip(zip_path, zip_files))
# cat("DwC-A created:", zip_path, "\n")
```

## Validate with obistools

```{r}
#| label: TODO-validate-obis

# librarian::shelf(obistools, quiet = T)
# obistools::check_eventids(d_events)
# obistools::check_fields(d_occurrences)
# obistools::check_scientificname_and_ids(d_occurrences)
```

## Upload to GCS

```{r}
#| label: upload-gcs

# sync_to_gcs(dir_dwca, glue("publish/{dataset}_obis"))
```

## Cleanup

```{r}
#| label: cleanup

dbDisconnect(con, shutdown = TRUE)
```
