---
title: "Publish {{dataset_name}} to ERDDAP"
editor_options:
  chunk_output_type: console
---

## Overview

Publish **{{dataset_name}}** to ERDDAP for public access via OPeNDAP/WMS/WFS.

- **Source ingest**: `{{source_ingest}}`
- **Portal**: ERDDAP (https://coastwatch.pfeg.noaa.gov/erddap/)
- **Format**: NCCSV (NetCDF-compatible CSV with CF metadata)

## Setup

```{r}
#| label: setup

devtools::load_all(here::here("../calcofi4db"))
devtools::load_all(here::here("../calcofi4r"))
librarian::shelf(
  CalCOFI/calcofi4db,
  CalCOFI/calcofi4r,
  DBI, dplyr, DT, fs, glue,
  here, lubridate, readr, stringr,
  tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F)

# output paths
dataset    <- "{{dataset}}"
dir_erddap <- here(glue("data/erddap/{dataset}"))
dir_create(dir_erddap)
```

## Load Source Data

```{r}
#| label: load-data

# load from release parquet
con <- get_duckdb_con(":memory:")
load_prior_tables(
  con,
  parquet_dir = here("data/parquet/{{source_ingest}}"),
  geom_tables = c("grid", "site"))

dbListTables(con)
```

## Subset and Filter

```{r}
#| label: TODO-subset

# TODO: extract the target subset from source data
# e.g., for cephalopods: filter ichthyo by species list
# e.g., for CTD: select measurement columns and reshape

# d_erddap <- dbGetQuery(con, "
#   SELECT
#     ...
#   FROM ...
#   WHERE ...")
```

## Format for ERDDAP

```{r}
#| label: TODO-format

# TODO: format columns for ERDDAP compatibility
# - time: ISO 8601 format (YYYY-MM-DDTHH:MM:SSZ)
# - lat/lon: decimal degrees, WGS84
# - depth: positive down (meters)
# - missing values: NaN (not NA)
# - column names: camelCase or CF standard names

# d_erddap <- d_erddap |>
#   mutate(
#     time      = format(datetime_utc, "%Y-%m-%dT%H:%M:%SZ"),
#     latitude  = lat_dec,
#     longitude = lon_dec,
#     depth     = depth_m) |>
#   mutate(across(where(is.numeric), ~ifelse(is.na(.), NaN, .)))
```

## Generate NCCSV Metadata

```{r}
#| label: TODO-nccsv-global

# TODO: define NCCSV global attributes
# see: https://coastwatch.pfeg.noaa.gov/erddap/download/NCCSV.html

nccsv_global <- list(
  `*GLOBAL*,Conventions`           = '"COARDS, CF-1.6, ACDD-1.3, NCCSV-1.2"',
  `*GLOBAL*,cdm_data_type`        = '"Point"',
  `*GLOBAL*,creator_email`        = '"calcofi@ucsd.edu"',
  `*GLOBAL*,creator_name`         = '"CalCOFI"',
  `*GLOBAL*,creator_url`          = '"https://calcofi.org"',
  `*GLOBAL*,history`              = glue('"Created {Sys.Date()} from CalCOFI database"'),
  `*GLOBAL*,infoUrl`              = '"https://calcofi.org"',
  `*GLOBAL*,institution`          = '"CalCOFI"',
  `*GLOBAL*,keywords`             = '"CalCOFI, {{dataset}}, oceanography"',
  `*GLOBAL*,license`              = '"CC-BY 4.0"',
  `*GLOBAL*,sourceUrl`            = '"https://calcofi.org"',
  `*GLOBAL*,standard_name_vocabulary` = '"CF Standard Name Table v79"',
  `*GLOBAL*,summary`              = '"{{dataset_name}} from CalCOFI cruises"',
  `*GLOBAL*,title`                = '"{{dataset_name}}"'
)
```

## Generate Variable Attributes

```{r}
#| label: TODO-nccsv-variables

# TODO: define per-variable attributes
# required: long_name, units
# recommended: standard_name (CF), colorBarMinimum, colorBarMaximum

# nccsv_vars <- list(
#   `time,*DATA_TYPE*`        = "String",
#   `time,long_name`          = '"Time"',
#   `time,standard_name`      = '"time"',
#   `time,units`              = '"yyyy-MM-dd\'T\'HH:mm:ssZ"',
#
#   `latitude,*DATA_TYPE*`    = "double",
#   `latitude,long_name`      = '"Latitude"',
#   `latitude,standard_name`  = '"latitude"',
#   `latitude,units`          = '"degrees_north"',
#
#   `longitude,*DATA_TYPE*`   = "double",
#   `longitude,long_name`     = '"Longitude"',
#   `longitude,standard_name` = '"longitude"',
#   `longitude,units`         = '"degrees_east"',
#
#   `depth,*DATA_TYPE*`       = "double",
#   `depth,long_name`         = '"Depth"',
#   `depth,standard_name`     = '"depth"',
#   `depth,units`             = '"m"',
#   `depth,positive`          = '"down"'
# )
```

## Write NCCSV File

```{r}
#| label: TODO-write-nccsv

# nccsv_path <- path(dir_erddap, glue("{dataset}.csv"))
#
# # write header
# header_lines <- c(
#   paste(names(nccsv_global), nccsv_global, sep = ","),
#   "",
#   paste(names(nccsv_vars), nccsv_vars, sep = ","),
#   "",
#   "*END_METADATA*"
# )
# writeLines(header_lines, nccsv_path)
#
# # append data
# write_csv(d_erddap, nccsv_path, append = TRUE, na = "NaN")
# cat("NCCSV file written:", nccsv_path, "\n")
```

## Generate datasets.xml Snippet

```{r}
#| label: TODO-datasets-xml

# TODO: generate the <dataset> element for ERDDAP's datasets.xml
# this snippet gets sent to the ERDDAP admin for inclusion

# cat('
# <dataset type="EDDTableFromNccsvFiles" datasetID="calcofi_{{dataset}}"
#          active="true">
#   <reloadEveryNMinutes>10080</reloadEveryNMinutes>
#   <fileDir>/path/to/data/</fileDir>
#   <fileNameRegex>.*\\.csv</fileNameRegex>
#   <recursive>false</recursive>
#   <addAttributes>
#     <att name="title">{{dataset_name}}</att>
#     <att name="institution">CalCOFI</att>
#   </addAttributes>
# </dataset>
# ')
```

## Validate

```{r}
#| label: TODO-validate

# TODO: validate NCCSV format
# - check CF standard names exist in CF table
# - check coordinate conventions (lat: -90 to 90, lon: -180 to 180)
# - check time format is ISO 8601
# - check no NA values (should be NaN for numeric)
# - check data types match declared *DATA_TYPE*
```

## Upload to GCS

```{r}
#| label: upload-gcs

# sync_to_gcs(dir_erddap, glue("publish/{dataset}_erddap"))
```

## Cleanup

```{r}
#| label: cleanup

dbDisconnect(con, shutdown = TRUE)
```
