---
title: "Publish {{dataset_name}} to CF NetCDF"
subtitle: "One self-documenting file preserving the sampling hierarchy"
author: "CalCOFI"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    df-print: kable
calcofi:
  target_name: publish_{{dataset}}_to_netcdf
  workflow_type: publish
  dependency:
    - release_database
  provider: {{provider}}
  dataset: {{dataset}}
  output: data/netcdf/{{dataset}}.nc
  workflow_url: https://calcofi.io/workflows/publish_{{dataset}}_to-netcdf.html
editor_options:
  chunk_output_type: console
---

## Overview

Publish **{{dataset_name}}** as a single self-documenting netCDF-4 file,
downstream of `release_database`.

Two things distinguish this from a table dump:

**1 · The hierarchy is preserved, not flattened.** {{hierarchy_prose}} Effort is
measured at the event level; flatten the hierarchy and each event's effort value
repeats onto every child row, so a naive `SUM()` silently over-counts. netCDF-4
groups store each level once and link children to parents by explicit index.

**2 · Variables are widened for CF.** The database is normalized — one row per
observation with `measurement_type` naming the quantity and `measurement_value`
holding the number — so one column mixes quantities with different units. CF
requires one unit and one `standard_name` per variable, so each
`measurement_type` becomes its own variable with units from
`metadata/measurement_type.csv`.

::: {.callout-note title="CF scope — state this honestly in every dataset"}
CF covers profile and trajectory geometries cleanly. For a nested
event → occurrence → bin hierarchy there is **no CF standard**. Say so in the
`cf_scope` global attribute so a consumer is never misled into expecting full CF
compliance. Do not claim "CF-compliant" for a structure CF does not define.
:::

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(DBI, duckdb, dplyr, glue, ncdf4, readr, jsonlite, digest,
                 knitr, tibble, here, quiet = TRUE)
here <- here::here
source(here("libs/publish_netcdf.R"))

DATASET     <- "{{dataset}}"           # file/URL identity (no provider prefix)
DATASET_KEY <- "{{provider}}_{{dataset}}"   # provenance stamp in the database
RELEASE     <- cc_release_version()    # the PROMOTED release, never a local tree
PQ          <- cc_release_parquet(RELEASE)
OUT_DIR     <- here("data/netcdf"); dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
OUT_NC      <- file.path(OUT_DIR, glue("{DATASET}.nc"))
PUBLISH     <- identical(Sys.getenv("CALCOFI_PUBLISH"), "true")   # opt-in upload

cat(glue("release : {RELEASE}\nparquet : {PQ}\noutput  : {OUT_NC}\npublish : {PUBLISH}\n"))
```

```{r}
#| label: connect
con <- dbConnect(duckdb())
for (s in c("INSTALL httpfs", "LOAD httpfs", "SET memory_limit='6GB'",
            "SET enable_progress_bar=false")) try(dbExecute(con, s), silent = TRUE)

# ALWAYS read the frozen release over HTTPS. Reading a local/serving tree is how
# the first published NetCDF shipped a month-old snapshot without anyone noticing.
q <- function(sql, ...) dbGetQuery(con, glue(sql, ..., .envir = parent.frame()))
dbExecute(con, glue("
  CREATE VIEW smp  AS SELECT * FROM read_parquet('{PQ}/sample.parquet')
                      WHERE dataset_key = '{DATASET_KEY}';
  CREATE VIEW obs  AS SELECT * FROM read_parquet('{PQ}/obs/dataset_key={DATASET_KEY}/data_0.parquet');
  CREATE VIEW oatt AS SELECT * FROM read_parquet('{PQ}/obs_attribute.parquet')
                      WHERE dataset_key = '{DATASET_KEY}';
  CREATE VIEW smea AS SELECT * FROM read_parquet('{PQ}/sample_measurement.parquet')
                      WHERE dataset_key = '{DATASET_KEY}';
  CREATE VIEW tax  AS SELECT * FROM read_parquet('{PQ}/taxon.parquet');"))
```

## Data integrity — check BEFORE writing

The release is the source of truth but is not assumed clean. Run these first so a
structural problem surfaces here rather than as a silently wrong hierarchy inside
a published product.

```{r}
#| label: integrity
# `obs` has no unique NATURAL key. Verify before relying on any composite: in
# swfsc_ichthyo, (sample_key, taxon_key, life_stage) repeats for 3 rows with
# DIFFERENT values, so joining children on it would fan out. obs_id is unique.
dup_obs <- q("
  SELECT sample_key, taxon_key, life_stage, count(*) AS n,
         string_agg(measurement_value::VARCHAR, ' | ') AS values
  FROM obs GROUP BY 1,2,3 HAVING count(*) > 1 ORDER BY n DESC")

# Children that match no parent are ORPHANS. Keep them with a missing parent
# index — silently discarding measured data is worse than admitting it is
# unparented — and report the count.
orphan_n <- q("
  SELECT count(*) AS n FROM oatt a
  LEFT JOIN (SELECT DISTINCT sample_key, taxon_key, life_stage FROM obs) o
    ON  o.sample_key = a.sample_key AND o.taxon_key = a.taxon_key
    AND coalesce(o.life_stage,'') = coalesce(a.life_stage,'')
  WHERE o.sample_key IS NULL")$n

cat(glue("duplicate occurrence keys : {nrow(dup_obs)}\norphan attribute rows     : {orphan_n}\n"))
if (nrow(dup_obs)) kable(dup_obs)
```

## Build the levels

Adapt this section per dataset — it is the reason these are notebooks rather than
one generic script. Each level is a data.frame ordered so children are
contiguous; each child carries a 1-based `parent_index` into its parent.

```{r}
#| label: levels
# {{levels_prose}}
# Example (net-tow shape):
#   site <- q("SELECT ... FROM smp WHERE sample_type='site' ORDER BY sample_key")
#   tow  <- q("SELECT ... FROM smp WHERE sample_type='tow'  ORDER BY sample_key")
#   net  <- q("SELECT ... FROM smp WHERE sample_type='net'  ORDER BY sample_key")

# WIDEN event-level effort into variables on its level — this is what makes
# double-counting structurally impossible rather than merely documented.
eff_types <- q("SELECT DISTINCT measurement_type FROM smea ORDER BY 1")$measurement_type
eff_sel <- paste(sprintf(
  "MAX(measurement_value) FILTER (WHERE measurement_type='%s') AS \"%s\"",
  eff_types, eff_types), collapse = ",\n    ")
# event <- event |> left_join(q("SELECT sample_key, {eff_sel} FROM smea GROUP BY sample_key"),
#                             by = "sample_key")
```

```{r}
#| label: parent-index
# match() on the parent's ORDERED key vector turns "these rows share a value"
# into an explicit, checkable pointer.
# child$parent_index <- match(child$parent_sample_key, parent$sample_key)
# stopifnot(!anyNA(child$parent_index))
```

## Write the netCDF-4 file

```{r}
#| label: write-nc
mt_meta <- cc_measurement_meta(here())

# ncdf4 exposes NO group API, but a slash-separated variable name creates a REAL
# netCDF-4 group — verified with ncdump AND h5dump, not by reading it back
# through ncdf4 itself. nc_level_vars()/nc_level_put() rely on this.
# d_<level> <- ncdim_def("<level>_n", "", seq_len(nrow(<level>)), create_dimvar = FALSE)
# v_<level> <- nc_level_vars("<level>", drop_key(<level>), d_<level>,
#                            d_<parent>, <level>$parent_index, mt_meta)

CF_SCOPE <- paste(
  "CF-1.10 where CF applies (coordinates, units, standard names, time).",
  "The {{hierarchy_short}} hierarchy uses netCDF-4 groups with explicit",
  "parent_index links; CF defines no feature type for this nesting.")
# globals: title, summary, Conventions, cf_scope, institution, source,
#          db_release, dataset_key, license, date_created, and the anomaly counts
#          (n_duplicate_occurrence_keys, n_orphan_attributes) so nothing found in
#          the integrity step is lost between the notebook and the file.
```

## Verify the written file

A published file that merely *exists* is not evidence it is correct. Re-open it
and confirm the hierarchy survived.

```{r}
#| label: verify
# ncv <- nc_open(OUT_NC)
#   - groups present
#   - each level's dimension length matches its data.frame
#   - effort variable length == event count (NOT child count)
#   - parent_index within 1..nrow(parent)
# nc_close(ncv)
```

```{r}
#| label: double-count-demo
# Show the payoff numerically rather than asserting it: effort summed at the
# right level vs repeated onto children.
# eff_once <- sum(event$volume_sampled, na.rm = TRUE)
# eff_flat <- sum(event$volume_sampled[child$parent_index], na.rm = TRUE)
```

## Publish

Versioned by DB release, bytes written once: if this build is byte-identical to
an earlier release the folder still gets a manifest and index, but the file is not
re-uploaded — `storage.calcofi.io` 302s the release-scoped `.nc` URL to the
canonical object.

```{r}
#| label: publish
#| eval: !expr PUBLISH
plan <- cc_netcdf_plan(OUT_NC, DATASET, RELEASE)
man  <- cc_netcdf_manifest(plan, DATASET, RELEASE, RELEASE, file.size(OUT_NC),
                           source_tables = c("sample","obs","obs_attribute",
                                             "sample_measurement","taxon"),
                           cf_scope = CF_SCOPE)
jsonlite::write_json(man, file.path(OUT_DIR, "manifest.json"),
                     auto_unbox = TRUE, pretty = TRUE)
```

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
